### 2018-01-04 RNARedPrint-Bioinfo2018.tex
Initial commit

\PassOptionsToPackage{utf8}{inputenc}
\documentclass{bioinfo}
\copyrightyear{2015} \pubyear{2015}

\access{Advance Access Publication Date: Day Month Year}
\appnotes{Manuscript Category}

%%% PACKAGES
\usepackage{booktabs} % for much better looking tables
\usepackage{array} % for better arrays (eg matrices) in maths
\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
  \let\itemize\compactitem
  \let\enditemize\endcompactitem
  \let\enumerate\compactenum
  \let\endenumerate\endcompactenum
  \let\description\compactdesc
  \let\enddescription\endcompactdesc
  \pltopsep=1pt
  \plitemsep=1pt
  \plparsep=1pt  
\usepackage{hyperref} 
\usepackage{xspace} 
\usepackage{amsmath,amssymb} 
\usepackage{bm}
\usepackage{verbatim}
\usepackage{longtable}
\usepackage[vlined]{algorithm2e}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Rolf's includegraphicstop
\makeatletter
\newsavebox{\@alignepsbox}
\newlength{\@aligneps}
\newcommand{\includegraphicstop}[2][]{%
\sbox{\@alignepsbox}{\includegraphics[#1]{#2}}%
\setlength{\@aligneps}{-\ht\@alignepsbox}%
\addtolength{\@aligneps}{2ex}%
\raisebox{\@aligneps}{\usebox{\@alignepsbox}}}
\makeatother

\makeatletter
\let\oldlt\longtable
\let\endoldlt\endlongtable
\def\longtable{\@ifnextchar[\longtable@i \longtable@ii}
\def\longtable@i[#1]{\begin{figure}[t]
\onecolumn
\begin{minipage}{0.5\textwidth}
\oldlt[#1]
}
\def\longtable@ii{\begin{figure}[t]
\onecolumn
\begin{minipage}{0.5\textwidth}
\oldlt
}
\def\endlongtable{\endoldlt
\end{minipage}
\twocolumn
\end{figure}}
\makeatother

%%%%%%%%%%%%%%%%% Theorems %%%%%%%%%%%%%%%%%%%%%%%%%

 \newtheorem{theorem}{Theoreme}
 \newtheorem{definition}[theorem]{Definition}
 \newtheorem{remark}[theorem]{Remark}
 \newtheorem{corollary}[theorem]{Corollary}
 \newtheorem{lemma}[theorem]{Lemma}
 \newtheorem{proposition}[theorem]{Proposition}

\newtheorem{observation}[theorem]{Observation}

%\newtheorem{algorithm}{Algorithm}
\newtheorem{axiom}{Axiom}
\newtheorem{hypothesis}{Working Hypothesis}

%%% macros for notation in DP framework
\newcommand{\network}{\mathcal{N}}
\newcommand{\dom}{\operatorname{dom}} %domain
\newcommand{\val}{a} % valuation aka assignment
\newcommand{\vars}{\operatorname{vars}}
\newcommand{\dep}{\operatorname{dep}}
\newcommand{\energy}[1]{\operatorname{e}_{#1}}
\newcommand{\numberof}{\operatorname{\#}}
\newcommand{\partfun}[1]{\mathcal{Z}_{#1}}
\newcommand{\separator}[2]{\operatorname{sep}(#1,#2)}
\newcommand{\difference}[2]{\operatorname{diff}(#1 \rightarrow #2)}
\newcommand{\real}{\mathbb{R}}
\newcommand{\genmarg}[1]{(\!|\!#1\!|\!)}
\newcommand{\gencomb}[1]{\langle\!|#1|\!\rangle}
\newcommand{\Message}[2]{m_{#1\rightarrow #2}}

\newcommand{\energyModel}{{\cal M}}
\newcommand{\structureElements}{{\cal SE}}
\newcommand{\powerSet}[1]{2^{#1}}
\newcommand{\underConstruction}[1]{{\LARGE$\triangle$\Large\!\!\!\!!}$\quad$\textcolor{red}{#1}}
\newcommand{\argmin}{\operatorname*{arg\,min}}
\newcommand{\objective}{{\mathbb{F}}}

%\newcommand{\assignments}{\operatorname{as}}
\newcommand{\assignments}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\R}{\mathcal{R}}
\renewcommand{\S}{\mathcal{S}}
\newcommand{\X}{\mathcal{X}}
\newcommand{\Y}{\mathcal{Y}}

\newcommand{\width}{w}

\newcommand{\sample}{\texttt{Sample}}
\newcommand{\elim}[2]{\operatorname{elim}(#1,#2)}
\newcommand{\edgesToR}{E^r_T}

\newcommand{\phitotal}{\phi_{\operatorname{m}}}

\newcommand{\Ebp}[2]{E^{\textrm{bp}}_{#1}(#2)}
\newcommand{\Ehp}[1]{E^{\textrm{hp}}(#1)}
\newcommand{\Eint}[1]{E^{\textrm{int}}(#1)}

\newcommand{\Def}[1]{{\bf #1}}

\newcommand{\TargetE}{E^{\star}}

\newcommand{\TODO}[1]{\Final{\begin{mdframed}[leftline=false,rightline=false,linecolor=red!70!black,linewidth=1.5pt]\color{red!70!black}\sf TODO: #1\end{mdframed}}}
\newcommand{\parHead}[1]{\Final{\paragraph{#1}}}

\newcommand{\Final}[1]{#1}
%% Uncomment the line below for ``Final'' version
\renewcommand{\Final}[1]{}

\newcommand{\Design}[1]{{\sf Designs}^{\star}(#1)}
\newcommand{\NumDesign}{\ensuremath{\#}{\sf Designs}\xspace}
\newcommand{\IS}[1]{{\sf IndSets}(#1)}
\newcommand{\Nuc}[1]{{\sf #1}}
\newcommand{\Ab}{\Nuc{A}}
\newcommand{\Cb}{\Nuc{C}}
\newcommand{\Gb}{\Nuc{G}}
\newcommand{\Ub}{\Nuc{U}}

\newcommand{\Software}[1]{{\ttfamily #1}}

\newcommand{\ourprog}{\Software{RNARedPrint}}

\setlength{\parskip}{.2em}

%%% end macro defs

\begin{document}
\firstpage{1}

\subtitle{Subject Section}

\title[short Title]{Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures}
\author[Hammer, Ponty, Wang and Will]{Stefan Hammer\,$^{\text{\sfb 1,2,3}}$, Yann Ponty\,$^{\text{\sfb 4,5,}\star}$, Wei Wang\,$^{\text{\sfb 4}}$ and Sebastian Will\,$^{\text{\sfb 2}}$}
\address{$^{\text{\sf 1}}$University Leipzig, Department of Computer Science and Interdisciplinary Center for Bioinformatics, 04107 Leipzig, Germany; 
$^{\text{\sf 2}}$University of Vienna, Faculty of Chemistry, Department of Theoretical Chemistry, 1090 Vienna, Austria;
$^{\text{\sf 3}}$University of Vienna, Faculty of Computer Science, Research Group Bioinformatics and
Computational Biology, 1090 Vienna, Austria;
$^{\text{\sf 4}}$CNRS UMR 7161 LIX, Ecole Polytechnique, Bat. Turing, 91120 Palaiseau, France;
and $^{\text{\sf 5}}$ AMIBio team, Inria Saclay, Bat Alan Turing, 91120 Palaiseau, France}

\corresp{$^\ast$To whom correspondence should be addressed.}

\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

\editor{Associate Editor: XXXXXXX}

\abstract{\textbf{Motivation:} The design of multi-stable RNA molecules has important applications in biology, medicine, and biotechnology. Synthetic design approaches largely benefit from effective in-silico methods, which can tremendously impact their cost and feasibility. \\
\textbf{Results:} Here, we revisit a central ingredient of most in-silico design methods: the sampling of sequences for multi-target design. We establish the $\#${\sf P}-hardness of uniform sampling, and introduce \ourprog{}, a tree decomposition-based algorithm for efficient, fixed parameter tractable sampling. By modeling the problem as a constraint network, \ourprog{} supports generic Boltzmann-weighted sampling for arbitrary additive RNA energy models; this enables generating designs meeting specific goals like expected free energies or \Gb\Cb-content. Finally, we empirically study general properties of the approach and generate biologically relevant multi-target Boltzmann-weighted designs for a common design benchmark. In particular, we show significant improvements over seed sequences generated by uniform sampling, the previously best available sampling strategy for multi-target design.\\
\textbf{Availability:} \url{https://github.com/yannponty/RNARedPrint}\\
\textbf{Contact:} Our software is freely available at: \href{yann.ponty@lix.polytechnique.fr}{yann.ponty@lix.polytechnique.fr}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

\maketitle

\section{Introduction}
\parHead{Design, applications and motivation for multiple design.}Synthetic biology endeavors the engineering of artificial biological
systems, promising broad applications in biology, biotechnology and
medicine. Centrally, this requires the design of biological
macromolecules with highly specific properties and programmable functions.
In particular, RNAs present themselves as well-suited tools for
rational design targeting specific functions~\cite{Kushwaha2016}. RNA function is tightly
coupled to the formation of secondary structure, as well as changes in
base pairing propensities and the accessibility of regions, e.g. by
burying or exposing interaction sites~\cite{Rodrigo2014}. At the same time, the
thermodynamics of RNA secondary structure is well understood and its prediction is
computationally tractable~\cite{McCaskill1990}. Thus,  structure can serve as effective
proxy within rational design approaches, ultimately targeting catalytic~\cite{Zhang2013} or regulatory~\cite{Rodrigo2014} functions.

\parHead{Motivating multiple RNA design.} The function of many RNAs
depends on their selective folding into one or several alternative
conformations. Classic examples include riboswitches, which
notoriously adopt different stable structures upon binding a specific
ligand. Riboswitches have been a popular application of rational
design~\cite{Wachsmuth2013,Domin2017}, partly motivated by their capacity to
act as biosensors~\cite{Findeiss2017}. At the co-transcriptional
level, certain RNA families feature alternative, evolutionarily
conserved, transient structures~\cite{Zhu2013}, which facilitate the
ultimate adoption of functional structures at full elongation.  More
generally, simultaneous compatibility to multiple structures
is a relevant design objective for engineering kinetically controlled RNAs, finally targeting prescribed folding pathways. Thus,
modern applications of RNA design often target (the simultaneous
stability of) multiple structures, additionally aiming at other
features, such as specific \Gb\Cb-content~\cite{Reinharz2013} or the
presence/absence of functionally relevant motifs (either anywhere or
at specific positions)~\cite{Zhou2013}; these objectives motivate flexible
computational design methods.

\parHead{On the importance of sampling for design.}
Many computational methods for RNA design rely on similar overall
strategies: initially generating one or several \Def{seed} sequences
and optimizing them subsequently. In many cases, the seed quality was
found to be critical for the empirical performance (and solution
quality) of RNA design methods~\cite{Levin2012}. For instance, random
seed generation improves the prospect of subsequent optimizations, helping
to overcome local optima of the objective function, and induce better
sequence diversity across designs~\cite{Reinharz2013}.  For
single-target approaches, \Software{INFO-RNA}~\cite{Busch2006} made
significant improvements mainly by starting its local search from the
minimum energy sequence for the target structure instead of (uniform)
random sequences for the early \Software{RNAinverse}
algorithm~\cite{Hofacker1994}. This strategy was later shown to result
in unrealistically high \Gb\Cb-contents in designed sequences. To
address this issue, \Software{IncaRNAtion}~\cite{Reinharz2013}
controls the \Gb\Cb-content through an adaptive sampling strategy.

\parHead{Specificities and similarities of multi-target design.}
Specifically, for multi-target design, virtually all available methods~\cite{Lyngsoe2012,HoenerzuSiederdissen2013,Taneda2015,Hammer2017} follow the same overall generation/optimization scheme.
%
Facing the complex sequence constraints induced by multiple targets, early methods such as \Software{Frnakenstein}~\cite{Lyngsoe2012} and \Software{Modena}~\cite{Taneda2015} did not attempt to solve sequence generation systematically, but rely on \emph{ad-hoc} sampling strategies. 
%
Recently, the \Software{RNAdesign} approach~\cite{HoenerzuSiederdissen2013}, coupled with powerful local search within \Software{RNAblueprint}~\cite{Hammer2017}, solved the problem of sampling seeds from the uniform distribution for multiple targets. These methods adopt a graph coloring perspective, initially decomposing the graph hierarchically using various decomposition algorithms, and \Def{precomputing} the number of valid sequences within each subgraph. The decomposition is then reinterpreted as a decision tree to perform a \Def{stochastic backtrack}, inspired by Ding and Lawrence~\cite{Ding2003}, where individual assignments are chosen with probabilities derived from the subsolution counts, leading to uniform sampling. The overall complexity of \Software{RNAdesign} grows like $\Theta(4^{\gamma})$, where $\gamma$ is a parameter whose value is bounded only by the length of the designed RNA.

\TODO{Develop questions (relevance, difficulty...). Maybe tease results?}
\parHead{Motivation.} The exponential time and space requirements of the \Software{RNAdesign} method already raise the question of the \Def{complexity of (uniform) sampling for multi-target design}. Since stochastic backtrack can be performed in linear time per sample, the method is dominated by the precomputation step, which requires counting valid designs. Thus, we focus on the question: \emph{Is there a polynomial-time algorithm to count valid multi-target designs?} In Section~\ref{sec:counting}, we answer in the negative, showing that there exists no such algorithm unless ${\sf P}={\sf NP}$. Our result relies on a surprising bijection (up to a trivial symmetry) between valid sequences  and independent sets of a bipartite graph, being the object of recent breakthroughs in approximate counting complexity~\cite{Bulatov2013,Cai2016}.
The hardness of counting (and conjectured hardness of sampling) does not preclude, however, practically applicable algorithms for counting and sampling. In particular, we wish to extend the flexibility of multi-structure design, leading to the following questions: \emph{How to sample, in a principled way, from a Boltzmann distribution based on expressive energy models? How to enforce additional constraints, such as the \Gb\Cb-content, complex sequence constraints, or the energy of individual structures?}

\begin{figure*}[t]
{\centering\includegraphics[width=.8\textwidth]{Workflow}\\}
\caption{General outline of \ourprog{} for base pair-based energy models. From a set of target secondary structures (i), base-pairs are merged (ii) into a compatibility graph (iii) and transformed into a tree decomposition (iv). The tree is then used to compute the partition function, followed by a Boltzmann sampling of valid sequences (v). An adaptive scheme learns weights to achieve targeted energies and $\Gb\Cb$-content, leading to the production of suitable designs (vi).}
\label{fig:workflow}
\end{figure*}

To answer these questions, we introduce a generic framework (illustrated in~Fig.~\ref{fig:workflow}) enabling efficient Boltzmann-weighted sampling over RNA sequences with multiple target structures (Section~\ref{sec:FPT}). Guided by a \Def{tree decomposition} of the network, we devise dynamic programming to compute partition functions and sample sequences from the Boltzmann distribution%
% (Subsection~\ref{sec:PF})
. We show that these algorithms are \Def{Fixed-Parameter Tractable} for the \Def{treewidth}%
% Subsection~\ref{sec:complexity}
%
; in practice, we limit this parameter by using state-of-the-art tree decomposition algorithms. 
%Uniform sampling is handled as a special case of Boltzmann sampling, where each valid sequence receives energy zero and---consequently---computing partition functions specializes to counting.
By modeling sequences as variable assignments in a weighted constraint
network, we support arbitrary multi-ary constraints and thus
arbitrarily complex energy models,
notably subsuming all commonly
used RNA energy models%
%(Subsection~\ref{sec:energy_models})
.  Moreover, we describe an \Def{adaptive
  sampling} strategy to control the free energies of the individual
target structures and the $\Gb\Cb$-content%
% (SubSection~\ref{sec:multiBoltzmann})
. Notably, the sampling procedure is based on a simplified RNA energy model, taking only the most important energy contributions into account, which crucially limits computation time and space; at the same time, our approach targets realistic RNA energies in the well-accepted Turner RNA energy model. This high accuracy enables efficiently generating biologically relevant multi-target designs in our final application of our overall strategy to a large set of multi-target RNA design instances from a representative benchmark Section~\ref{sec:results}).

%In this work, we present a generic approach to multi-target sequence sampling that enables Boltzmann-weighted sampling (subsuming uniform sampling) for a wide class of objective functions that can even make use of the commonly used ``full-featured'' RNA energy models (like the Turner model~\cite{turner}). Due to its high expressivity the same framework allows combining objective functions with various hard and soft constraints, which can accommodate various desirable multi-target design scenarios.

%To provide a simple concrete example of multi-target design, consider designing a single sequence where all target structures are as stable as possible.
%That is, one asks for the sequence $S$ that minimizes the energy differences of the target structures to the minimum energy for this sequence among all possible structures; one aims to minimize
%   \begin{equation}
%\label{eq:design-objective}
%\sum_{\ell\in[1,k]} E(S;R_\ell) - E_0(S),
%\end{equation}
%over all sequences $S$, where $E_0(S):=\min_R E(S;R)$.
%\TODO{even Taneda's objective is significantly more complex; how much do we want to talk about this...}
%
%The proposed Boltzmann-weighted sampling approach can support this design procedure by sampling start structures from the Boltzmann distribution over
%\begin{equation}
%\label{eq:sampling-objective} \sum_{\ell\in[1,k]} w_\ell E(S;R_\ell)
%\end{equation}
%or extensions thereof, e.g.~for controlling base frequencies. Suitable weights $w_\ell$ to accommodate the design objective can be approximated automatically after the machinery for Boltzmann sampling over such functions is set up.
%
%\TODO{Rewrite (from here to "Contributions"): sounds too negative}
%Notably, by Eq.~(\ref{eq:sampling-objective}) we do not suggest to sample over the energy difference of Eq.~$(\ref{eq:design-objective})$. The term $\min_R E(S;R)$, which---for each sequence---minimizes over the entire structure space, precludes efficient sampling even by the proposed approach.
%
%Already sampling over Eq.~(\ref{eq:sampling-objective}) strictly extends the only previously existing approach to controlled sampling from single to multi-target sampling (in complete analogy to IncaRNAtion only after GC content is controlled, which we present later) and subsumes uniform sampling for multi-target design.
%
%Naturally, also more complex objective functions are desirable for RNA design and have been discussed in the literature; e.g. the ensemble defect~\cite{Zadeh:2010}. Typically, more complex objective functions make design computationally much more challenging. Similarily, also the proposed sampling framework supports only a limited class of objective functions directly and efficiently---namely ``only'' all additive energy functions. Nevertheless, one can safely assume that the design under various more realistic objective functions profits from controlled sampling as long as the sampling is directed towards the final objective function.

%\TODO{Recycled material: To be merged with intro later}
%{\it 
%This graph coloring perspective was the basis for the uniform sampling method of \Software{RNADesign} / \Software{RNAblueprint}, which decomposes the graph hierarchically, counts the solutions in subgraphs, and finally samples solutions. Essentially, an \Def{ear decomposition} is used as a decision tree, where decisions are made according to the subsolution counts in order to guarantee uniform sampling.
%
%However, highly desirable objectives of RNA design, such as a control over the energy of individual structures, require one to depart from the uniform generation, and therefore generalize this setting. First, instead of subsolution counts, one needs partition functions for guiding the sampling procedure. Second, in the case of realistic energy models %like Eq.~(\ref{eq:sampling-objective}) are based on realistic RNA energy models 
%(e.g.~the Turner model), the atomic energy contributions depend on bases at more than two sequence positions---i.e. they cannot be captured by a strategy that only accounts for base-pairs.
%%
%
%
%These considerations motivate our sampling strategy, where we:
%\begin{itemize}
%\item Model RNA sequences as assignments over a constraint network. The network consists of one variable per sequence position; constraints and energy contributions are modeled as functions of arbitrary arity over the variables, such that the total energy is the sum of all function values;
%\item Translate the dependency hyper-graph into a binary dependency graph by inserting cliques (Appendix), such that a tree decomposition with small treewidth can be computed by a generic tree decomposition algorithm;
%\item Compute the partition functions by specializing the generic ``Cluster Tree Elimination'' (CTE) dynamic programming algorithm, based on the cluster tree decomposition induced by the constraint network; 
%
%\item Sample sequences are from the Boltzmann-distribution guided by the decomposition tree and the sub-partition functions.
%\end{itemize}
%}

\section{Definitions and problem statement}
\label{sec:problem-statement}

An \Def{RNA sequence $S$} is a word over the \Def{nucleotides
  alphabet} $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}$; let $\S_n$ denote the set of
sequences of length $n$. An \Def{RNA (secondary) structure of length
  $n$} is a set of \Def{base pairs} $(i,j)$, where $1\leq i<j\leq
n$. We call an RNA structure $R$ \Def{non-crossing}, iff it does not
contain any two different base pairs $(i,j)$ and $(i',j')$ such that
$i\leq i'\leq j \leq j'.$ \Def{Valid base pair} must pair bases from
the $\B:=\left\{\{\Ab,\Cb\},\{\Gb,\Cb\},\{\Gb,\Ub\}\right\}.$
Consequently, $S$ is \Def{valid} for $R$, iff
$(S_i,S_j)\in \B$ for all $(i,j)\in R$.
%
%\TODO{Also, we need to decide whether base-pairs are oriented or not ($(x,y)$ versus $\{x,y\}$). Right now, we are inconsistent in the definitions above. SW: i would prefer $(i,j)$ with $i<j$; } %% where is this inconsistent?

Given is a fixed set of target RNA structures
$\R:=\{R_1, \dots, R_k\}$ for sequences of length $n$. $\R$ induces a
\Def{compatibility graph}
$G_{\R}$ with nodes $\{1,\dots,n\}$ and edges $\bigcup_{\ell\in[1,k]} R_\ell$. Valid
sequences for $\R$ can then be seen as specific colorings of $G_{\R}$,
using the four colors in $\Sigma$, such that the colors assigned to
adjacent vertices of $G_{\R}$ constitute valid base pairs.
%
We define the energy of a sequence $S$ based on the set of structures
$\R$ as $E_\R(S) \in \mathbb{R}\cup\{\infty\}$. In
our setting, the energy $E_\R(S)$ is additively composed of
the energies of the single RNA structures in an RNA energy model, as
well as sequence dependent features like \Gb\Cb-content.  Furthermore note
that $E_\R(S)$ is finite iff $S$ is valid for each structure
$R_1,\dots,R_k$.

At the core of this work, we study the computation of partition
functions over sequences.\smallskip\\
\textbf{Central problem (Partition function over sequences).}
  Given an energy function $E$ and a set $\R$ of structures of length $n$, compute the
  partition function
  \begin{equation}
    \label{eq:mainproblem}
    \partfun{E_\R}(S) = \sum_{S\in\S_n} \exp(-\beta E_\R(S)),
  \end{equation}
  where $\beta$ denotes the inverse pseudo-temperature, and $\S_n$ the set of sequences of length $n$.

%We solve the problem efficiently for any fixed parameter $w$, whose value reflects the intrinsic complexity of the structures and energy function $E_\R$.
%
As will be explicited in future sections, our approach relies on breaking down the energy function $E_\R(S)$ into additive components, each depending on only few sequence positions. Given $\R$, we express $E_\R(S)$
as the sum of energy contributions $f(S)$ over a set $\F$ (of
functions $f:\S_n\to\real$), s.t.~$E_\R(S)=\sum_{f\in\F} f(S)$. This
captures realistic RNA energy models, while bounding the
dependencies to sequence positions introduced by each single $f$. Formally, define the \Def{dependencies} $\dep(f)$ of $f$ as the minimal set of sequence positions $\I\subseteq\{1,\dots,n\}$, where $f(S)=f(S')$ for all sequences $S$ and $S'$ agreeing at all positions in $\I$.

The \Def{dependency graph} of the set $\F$ of functions (on
sequences of length $n$) is the \Def{hypergraph}
$G_\F=(\{1,\dots,n\} ,\{\dep(f)\mid f\in \F\})$. 
Our algorithms will critically rely on a \Def{tree decomposition} of the dependency graph, which we define below.
\begin{definition}[Tree decomposition and width]
  \label{def:treedecomp}
  Given an hypergraph $G=(X, E)$,a \Def{tree decomposition} is a
  pair $(T,\chi)$, where $T$ is an unrooted tree/forest, and
  $\chi(v)\subseteq X$ is the set of vertices from $G$ assigned to the node $v\in T$, such
  that
\begin{enumerate}
\item each $x\in X$ occurs in at least one $\chi(v)$;
\item for each $x\in X$, the set $\{ v \mid x \in \chi(v) \}$ induces a connected subtree of $T$.
\item for every hyper-edge $e\in E$ of $G$, there is a node $v\in T$, such that $e\subseteq\chi(v)$.
\end{enumerate}
The \Def{width} of a tree decomposition $(T,\chi)$ is defined as
$\width(T,\chi) = \min_{u\in T} |\chi(u)| - 1 $. The \Def{treewidth}
of $G$ is the smallest width in any tree decomposition of $G$.
\end{definition}

\section{Counting designs for multiple structures is \#{\sf P}-hard}\label{sec:counting}

%%% reformulate later
%We define the following notations: $\real^\infty:=\real\cup\{\infty\}$, $[i,j]:=\{i,\ldots,j\},$ \ldots
Now we turn to the complexity of $\NumDesign(G)$, the problem of computing the number of valid sequences for a given compatibility graph $G=(V,E)$. Note that this problem corresponds to the partition function problem in a simple $(0/\infty)$-valued base pair models,with $\beta\neq 1$. As previously noted~\cite{Flamm2001}, a set of target structures admits a valid design iff its compatibility graph is bipartite, which can be tested in linear time.
Moreover, without $\{\Gb, \Ub\}$ base pairs, any connected component  $C\in {\rm CC}(G)$ is entirely determined by the assignment of a single of its nucleotides. The number of valid designs is thus simply $4^{\#{\rm CC}(G)}$, where $\#{\rm CC}(G)$ is the \Def{number of connected components}.

The introduction of $\{\Gb, \Ub\}$ base pairs radically changes this picture, and we show that valid designs for a set of structures cannot be counted in polynomial time, unless ${\sf\# P}={\sf FP}$. The latter equality would, in particular, imply the classic ${\sf P}={\sf NP}$, and solving $\NumDesign(G)$ is polynomial time is therefore probably difficult.

To establish that claim, we consider instances $G=(V_1\cup V_2, E)$ that are  strongly connected and bipartite ($(V_1\times V_2) \cap E = \varnothing$), noting that hardness over restricted instances implies hardness of the general problem. Moreover remark that, as observed in Subsec.~\ref{sec:complexity}, assigning a nucleotide to a position $u\in V$ constrains the parity ($\{\Ab,\Gb\}$ or $\{\Cb,\Ub\}$) of all positions in the connected component of $u$. For this reason, we restrict our attention to the counting of valid designs \emph{up to trivial} $(\Ab\leftrightarrow \Cb/\Gb\leftrightarrow \Ub)$ symmetry, by constraining the positions in $V_{1}$ (resp. $V_2$) to only $\Ab$ and $\Gb$ (resp. $\Cb$ and $\Ub$). Let $\Design{G}$ denote the subset of all designs for $G$ under this constraint, noting that $\NumDesign(G) = 2\cdot|\Design{G}|$.

We remind that an \Def{independent set} of $G=(V,E)$ is a subset $V'\subseteq V$ of nodes that are not connected by any edge in $E$. Let $\IS{G}$ denote the set of all independent sets in $G$. 
Now, consider the function $\Psi: \Design{G} \to \IS{G}$ defined by $ \Psi(f) := \left\{v\in V\mid f(v)\in\{\Ab,\Cb\}\right\}.$

\begin{proposition}
 $\Psi$ is a bijection between $\Design{G}$ and $\IS{G}$.
\end{proposition}
\begin{proof}
Let us establish the injectivity of  $\Psi$, i.e. that $\Psi(f)\neq\Psi(f')$ for all $f\neq f'$.
If $f\neq f'$, then there exists a node $v\in V$ such that $f(v)\neq f'(v)$. 
Assume that $v\in V_1$, and remind that the only nucleotides allowed in $V_1$ are $\Ab$ and $\Gb$. Since $f(v)\neq f'(v)$, then we have $\{f(v),f'(v)\}=\{\Ab,\Gb\}$
and we conclude that $\Psi(f)$ differs from $\Psi(f')$ at least with respect to its inclusion/exclusion of $v$.

We turn to the surjectivity of $\Psi$, i.e. the existence of a preimage for each element $S\in \IS{G}$. Let us consider the function $f$ defined as
\begin{align}
 \forall v_1\in V_1, f(v_1)&= \begin{cases} \Ab & \text{if }v_1\in S\\ \Gb & \text{if }v_1\notin S\end{cases}& 
 \forall v_2\in V_2, f(v_2)&= \begin{cases} \Cb & \text{if }v_2\in S\\ \Ub & \text{if }v_2\notin S\end{cases}
\end{align}
One easily verifies that $\Psi(f) = S$.

We are then left to determine if $f$ is a valid design for $G$, i.e. if for each $(v, v') \in E$ one has  $\{f(v),f(v')\}\in \B.$ Since $G$ is bipartite, any edge in $E$ involves two nodes $v_1\in V_1$ and $v_2\in V_2$. Remark that, among all the possible assignments $f(v_1)$ and $f(v_2)$, the only invalid combination of nucleotides is $(f(v_1),f(v_2)) = (\Ab,\Cb)$. However, such nucleotides are assigned to positions that are in the independent set $S$, and therefore cannot be adjacent. We conclude that $\Psi$ is surjective, and thus bijective.
\end{proof}

Now we can build on the connection between the two problems to obtain complexity results for \NumDesign. Counting independent sets in bipartite graphs ($\#{\sf BIS}$) is indeed a well-studied \#{\sf P}-hard problem~\cite{Ge2012}, from which we immediately conclude:
\begin{corollary}
  \NumDesign is $\#{\sf P}$-hard
\end{corollary}
\begin{proof}
  Note that $\#{\sf BIS}$ is also \#{\sf P}-hard on connected graphs, as the number of independent sets for a disconnected graph $G$ is given by $|\IS{G}|=\prod_{cc\in CC(G)} |\IS{cc}|$. Thus any efficient algorithm for $\#{\sf BIS}$ on connected instances provides an efficient algorithm for general graphs.

  Let us now hypothesize the existence of a polynomial-time algorithm $\mathcal{A}$ for \NumDesign over strongly-connected graphs $G$. Now consider the polynomial-time algorithm $\mathcal{A}'$ that first executes $\mathcal{A}$ on $G$ to produce $\NumDesign(G)$, and returns $\NumDesign(G)/2=|\Design{G}| = |\IS{G}|$. We conclude that $\mathcal{A}'$ solves $\#{\sf BIS}$ in polynomial-time. This means that $\NumDesign$  is at least as hard as $\#{\sf BIS}$, thus does not admit a polynomial time exact algorithm unless $\#{\sf P}={\sf FP}$.
\end{proof}

\section{An FPT algorithm for sampling Boltzmann-weighted designs}
\label{sec:FPT}

For our algorithmic description, we translate the concepts of
Section~\ref{sec:problem-statement} to the formalism of constraint networks, here
specialized as RNA design network. This allows us to base our
algorithm on cluster tree elimination (CTE) by Rina
Dechter~\cite{Dechter2013}.
%
In the RNA design network, sequences $S\in\S_n$ are isomorphically
modeled as assignments $\val_S$ of bases to $n$ variables, each associated to
a sequence position. The functions $f\in\F$ of
Section~\ref{sec:problem-statement}, are interpreted as functions on the network's variables.

\begin{definition}[RNA design network]
An \Def{RNA design network} is a tuple $\network=(\X,\F)$ such that:
\begin{itemize}
\item $\X$ is a set of \Def{variables} $\{x_1,\dots,x_n\}$ associated with the sequence positions
\item $\F$ is a set of \Def{functions} $f:\S_n\to\real$, each function $f\in \F$ being associated with the set $\vars(f)=\{x_i\mid i\in\dep(f) \}\subset X$ of variables.
\end{itemize}
\end{definition}

A \Def{(partial) assignment} is a function $\val: \X' \to \Sigma$
that maps the variables $\X'\subseteq \X$ to bases; we call $\X'$ the \Def{domain of the assignment
  $\val$}, written $\dom(\val)$; $\val$ is \Def{total} iff $\dom(\val)=\X.$
%
For assignments $\val$ and functions $f$, we define the
special notation $f[a]$ to \Def{evaluate $f$ for $\val$}; i.e.{}
$f[\val] := f(S),$ for any sequence $S$ where $S_i=\val(i)$ for all
$i\in\dep(f).$ Then, the \Def{energy $\energy{\network}(S)$ of a
  sequence} $S$ in a network $\network$ is defined as sum of the
values of all functions in $f\in\F$ evaluated for $\val_S$, i.e.
$\energy{\network}(S) := \sum_{f\in F} f[\val_S].$

The network energy $\energy{\network}(S)$ corresponds to the energy in
Eq.~$(\ref{eq:mainproblem}),$ where this energy is modeled as sum of
the functions in $\F$. Consequently, $\partfun{\R}$ of
Eq.~$(\ref{eq:mainproblem})$ is modeled as network partition function
$\partfun{\network} := \sum_{S}\exp(-\beta\energy{\network}(S)) = \sum_{S}\prod_{f\in F} \exp( -\beta\cdot
f[\val_S] ).$

% For a given RNA design network, one is typically interested in the \Def{minimum free energy sequence} $\energy{\network}:=\min_{S}\energy{\network}(S)$, the \Def{enumeration of valid sequences} $\numberof{\network}$ and the \Def{partition function} of the network $\network$ as \begin{equation}
%  \partfun{\network;\beta} := \sum_{S}\prod_{f\in F} \exp( -\beta\cdot f[\val_S] ).
%  \end{equation}

\subsection{Partition function and Boltzmann sampling through stochastic backtrack}\label{sec:PF}
The minimum energy, counting, and partition function
over RNA design network can be computed by dynamic programming based
on a tree decomposition of the network's dependency graph
(i.e. cluster tree elimination).
%This yields fixed parameter tractable (FPT) algorithms, based on the treewidth.
We focus on the efficient computation of the partition
function. %Observe that counting is a special case of the partition function and can be obtained by setting $\beta =0$.%; moreover, efficient energy minimization can be obtained by a simple $(+,\times)\to (\min,+)$ algebraic substitution and dropping the transformation of energy contributions to Boltzmann weights.

%\begin{figure}[t]
%{\centering\scalebox{11}{\fbox{$\phantom{5689}$}}\\}
%
%  \caption{Figure illustrating RNA network to hypergraph to tree decomposition}
%\end{figure}

Let us now introduce some definitions. A \Def{cluster tree} for the network $\network=(\X,\F)$ is a tuple $(T,\chi,\phi)$, where $(T,\chi)$ is a tree decomposition of $G_\F$, and $\phi(v)$ represents a set of functions $f$, each uniquely assigned to a node $v\in T$; $\vars(f)\subseteq\chi(v)$ and $\phi(v)\cap \phi(v')=\varnothing$ for all $v\neq v'$.
For two nodes $v$ and $u$ of the cluster tree, define their \Def{separator} as $\separator{u}{v} := \chi(u)\cap\chi(v)$; moreover, we define the \Def{difference variables} from $u$ to $v$ by $\difference{u}{v}:=\chi(v) - \separator{u}{v}$.

Denote by $\assignments(\Y)$ the set of all assignments of the variables in $\Y$; moreover, given non-overlapping assignments $\val'$ and $\val''$, we write $\val'\mid\val''$ to denote the assignment defined as $\val'$ and $\val''$ respectively on $\dom(\val')$ and $\dom(\val'')$. 
Finally we assume, w.l.o.g., that all variable difference sets $\difference{u}{v}$ are restricted to a single variable: for any given cluster tree, an equivalent (in term of treewidth) cluster tree can always be obtained by inserting at most $\Theta(|\X|)$ additional clusters.

Let us now consider the \Def{computation of the partition function}.
Given is the RNA design network $\network=(\X,\F)$ and its cluster
tree decomposition $(T,\chi,\phi)$.  W.l.o.g, we assume that $T$ is
connected and contains a dedicated node $r$, with $\chi(r)=\varnothing$
and $\phi(r)=\varnothing$, added as a virtual root
$r$ connected to a node in each connected
component of $T$.  Now, we consider the set of directed edges
$\edgesToR{}$ of $T$ oriented to $r$; define $T_r(u)$ as the induced
subtree of $u$. Algorithm~\ref{alg:pf} computes the partition function by
passing messages along these directed edges $u\to v$ (i.e. always from
some child $u$ to its parent $v$). Each message is a function on
variables $\vars(m)\subseteq \X$ to $\real\cup\{\infty\}$. The message
from $u$ to $v$ represents the partition functions of the subtree of
$u$ for all possible assignments to the variables
$\vars(m)=\separator{u}{v}$. Its correctness can be shown by induction over $T$ (Appendix~\ref{appsec:correctness}).
After running Alg.~\ref{alg:pf}, 
multiplying the 0-ary messages sent to the root $r$ yields the total partition function:
\begin{math}
  \partfun{\network} = \prod_{(u\to r)\in T} m_{u\to r}[\varnothing].
\end{math}

\begin{algorithm}[t]
 \KwData{Cluster tree $(T,\chi,\phi)$}
 \KwResult{Messages $\Message{u}{v}$ for all $(u\to v)\in T$; i.e.~partition functions of the subtrees of all $v$ for all possible assignments of variables $\separator{u}{v}$.}
 \For{$u\to v\in T$ in postorder}{
  \For{$\val\in\assignments(\separator{u}{v})$}{
    $\Message{u}{v}[\val] \gets 0$\;
    \For{$\val'\in\assignments(\difference{u}{v})$}{
     $p \gets$ product( $exp(-\beta f[\val\mid \val'])$ for $f\in \phi(u)$ )\;
     $p \gets p\ \times\ $product( $\Message{w}{u}\left[\val\mid \val'\right]$ for $(w\to u)\in T$ )\;
     $\Message{u}{v}[\val] \gets \Message{u}{v}[\val] + p$\; 
   }
  } 
  \Return {$m$}\;
  }
 \caption{FPT computation of the partition function using
   dynamic programming (CTE). }\label{alg:pf}
\end{algorithm}

\SetKwProg{Fn}{Function}{}{}
 \SetKwFunction{Sample}{$\sample$}
 \SetKwFunction{Random}{UnifRand}

The partition functions can then direct a \Def{stochastic backtrack} to achieve \Def{Boltzmann sampling of sequences}, such that one samples from the Boltzmann distribution of a given desgin network $\network$. The sampling algorithm assumes that the cluster tree was expanded and the messages $\Message{u}{v}$ for the edges in $\edgesToR{}$ are already generated by Algorithm \ref{alg:pf} for the expanded cluster tree.
%
Algorithm~\ref{alg:sampling} defines the recursive procedure $\sample(u,\val)$, which returns an  assignment of all variables in the subtree rooted at $u$ from the Boltzmann-distribution.
Called on $r$ and the empty assignment, the procedure returns a total assignment corresponding to a Boltzmann-distributed random sequence.

%\begin{algorithm}[Boltzmann weighted sampling]
%\label{alg:sampling}
% \ \\
%  Define $\sample(u,\val)$:
%  \begin{enumerate}[1)]
%  \item Let $\Delta:=\chi(u)-\dom(\val)$ be the set of free variables for $u$.
%  \item For all assignments $\val'\in\assignments(\Delta)$, compute
%  $$\partfun{\val'} := \prod_{f\in \phitotal(u)} \exp(-\beta f[\val'|\val]).$$
%  \item Choose one of the assignments $\val'\in\assignments(\Delta)$ with probability $\partfun{\val'}/\partfun{u}$, where $\partfun{u}$ is the sum of the $\partfun{\val'}$.
%  \item For all edges $(v',u)$ in $\edgesToR{}$:\quad update $\val':=\sample(v',\val'|\val)$.
%  \item Return $\val'$.
%  \end{enumerate}
%\end{algorithm}

\subsection{Computational complexity of the multiple target sampling algorithm}\label{sec:complexity}

%The complexity of the proposed sampling algorithm depends on the treewidth of the dependency graph $G_\F$.
% The treewidth $w$ of a cluster tree $(T,\chi,\phi)$, $T=(V,E)$, is $\max_{v\in V} |\chi(v)| - 1$, i.e.~the maximum number of variables in any of its clusters minus 1.

First, we observed that the computation time of tree decomposition
(\Software{GreedyFillIn}, implemented in \Software{LibTW}~\cite{Dijk2006}) for multi-target
sampling is negligible compared to Alg.~\ref{alg:pf} (Appendix~\ref{appsec:treedecomp} and \ref{appsec:dependency-cliques});
consequently, we discuss complexities without the tree decomposition
step.

For our analysis, we define the \emph{maximum separator size} $s$ as
$\max_{u,v\in V} | \separator{u}{v} |$ and denote the maximum size of
$\difference{u}{v}$ over $(u,v)\in\edgesToR{}$ as $D$.  Generally,
without specific optimizations, running Alg.~\ref{alg:pf} requires
$\mathcal{O}((|\F|+|V|)\cdot 4^{w+1})$ and $\mathcal{O}(|V|\cdot4^s)$ space (Appendix~\ref{appsec:algcomplexity});
Alg.~\ref{alg:sampling} would require $\mathcal{O}((|\F|+|V|)\cdot 4^D)$ per
sample on arbitrary tree decompositions (Appendix~\ref{appsec:algcomplexity}). W.l.o.g. we assume that
$D=1$; note that tree decompositions can generally be transformed,
such that $\difference{u}{v}\leq 1$.
%
Moreover, the size of $\F$ is linearly bounded: for $k$
input structures for sequences of length $n$, the energy function is
expressed by $\mathcal{O}(n\,k)$ functions. Finally, the number of cluster
tree nodes is in $O(n)$, such that $|\F|+|V| \in \mathcal{O}(n\,k)$.

\begin{algorithm}
 \KwData{Node $u$, partial assignment $\val\in\assignments(\separator{u}{v})$;\newline
 Cluster tree $(T,\chi,\phi)$ and partition functions $\Message{u'}{v'}[\val']$, $\forall (u'\to v')\in T$ and $\val'\in\assignments(\separator{u'}{v'})$.}
 \KwResult{Boltzmann-distributed random assignment for the subtree rooted at $u$, under the assumption of a partial assignment $\val$.}
 \Fn{\Sample$(u,\val;T,\chi,\phi,m)$}{
   $r \gets \Random(\Message{u}{v}[\val])$\;
   \For{$\val'\in\assignments(\difference{u}{v})$}{

     $p \gets$ product( $exp(-\beta f[\val\mid \val'])$ for $f\in \phi(u)$ )\;
     $p \gets p\ \times\ $product( $\Message{w}{u}\left[\val\mid \val'\right]$ for $(w\to u)\in T$ )\;
%
%     $p \gets 1$\;
%     \For{$f\in \phi(v)$}{
%  	      $p \gets p \times \exp(-\beta f[\val\mid \val'])$\; 
%     }
%     \For{$v\to w\in T$}{
%  	      $p \gets p \times \Message{v}{w}\left[\val\mid \val'\right]$\; 
%     }
  	  $r \gets r - p$\; 
  	  \If{$r<0$}{
  	    $a_{\rm res} \gets a\mid a'$\;
  	  \For{$(v\to w)\in T$}{
  	      $a_{\rm res} \gets a_{\rm res} \mid \Sample(v,a\mid a';T,\chi,\phi,m)$\; 
     }
  	  \Return {$a_{\rm res}$}
  	  \;}
 }
 }
 \caption{Stochastic backtrack algorithm for assignments in the Boltzmann distribution.}\label{alg:sampling}
\end{algorithm}
%For example, for design to a single non-crossing target in the nearest neighbor energy model, the complexity of Algs.~\ref{alg:pf} and \ref{alg:sampling} degrades to linear time (as reported for IncaRNAtion), since there are only $O(n)$ many functions and nodes (in the sequence length $n$). Moreover, the dependencies are tree-like owing to the tree-like non-crossing structure; this implies a constantly bounded maximum treewidth.

As already noted by Flamm \emph{et al}~\cite{Flamm2001}, the
requirement of canonical base pairing induces a bi-partition of each connected
component of the compatibility graph, such that the variables of one
partition are restricted to values $\{\Ab,\Gb\}$, but for other
partition, to values $\{\Cb,\Ub\}$. This suggests a strong
optimization of our algorithms, which reduces the basis in the
exponential factor from 4 to 2. The price for this reduction is a
binary choice for each of $c$ connected component. For each of the
$2^c$ combinations, the partition functions can be computed separately
in $O(n\,m\, 2^{w+1})$ time. This strategy is applied only if
$c< w+1$. 
%% SW: This could be optimized further by applying the strategy only to (optimally) selected components
Finally, we compute the sums of the partition functions for all possible combinations.

\begin{theorem}[Partition function and sampling complexity]
  For sequence length $n$, $k$ target structures, treewidth $w$ and
  a compatibility graph having $c$ connected components, $t$ sequences
  are sampled from the Boltzmann distribution  in
  %$O( n\, k \min(2^{w+c+1},4^{w+1}) + t\, n\, k )$ time.
  $\mathcal{O}( 2^d\, n\, k  + t\, n\, k )$ time, where $d:=\min(w+c+1,2(w+1))$.
\end{theorem}

\subsection{Encoding sequence features, constraints, and energy
  models.}\label{sec:energy_models}

The functions in $\F$ allow expressing complex features of the
sequences alone, e.g. rewarding or penalizing specific sequence
motifs, as well as features depending on the target structures.
Naturally, hard
constraints, which enforce or forbid features, can also be expressed by functions that
assign infinite penalties to invalid partial assignments. More concretely, we briefly 
discuss the modeling of RNA energy models in our
framework.

% To illustrate the expressivity of the framework, we outline the construction of function sets that express two different energy models for the energies $E(S;R_\ell)$ for encoding
% \begin{equation}
%   \label{eq:nussinov-network-energy}
%   \energy{\network}(S) = \sum_{\ell\in[1,k]} w_\ell E(S;R_\ell)
% \end{equation}
In simple \Def{base pair-based energy models}, energy is defined as the sum of base pair (pseudo)energies. If base pair energies $\Ebp{k}{i,j,x,y}$ (where
$i$ and $j$ are sequence positions, $x$ and $y$ are bases in $\Sigma$)
are given for each target structure $\ell$, s.t.
$ E(S;R_\ell) := \sum_{(i,j)\in R_\ell} \Ebp{k}{i,j,S_i,S_j}$, we
encode the network energy by the set of functions
$f$ for each base pair $(i,j)\in R_\ell$ of each input structure
$R_\ell$ that evaluate to $w_\ell \Ebp{k}{i,j,\val(x_i),\val(x_j)}$
under assignment $\val$.

% \paragraph{Nearest-neighbor model.}
% Let us now consider a simplified version of the nearest-neighbor model, where the free energy contribution  $\Ehp{x,y,s}$ of a hairpin only depends on the closing base pair $(x,y)$, and the contribution $\Eint{x,y,x',y',s}$ of an interior loop depends on its opening/closing bases pairs $(x,y)/(x',y)$, and its length $s$. Moreover, the energy of a multi loops with $s$ inner base pairs and $t$ unpaired base pairs is approximated as $a+b\,s+c\,t$, where $a,b,c$ are predefined constants.

% Again, the energy is expressed by adding functions to $F$ for each $R_\ell$ and base pair $(i,j)\in R_\ell$:
% \begin{itemize}
%   \item if $(i,j)$ closes a hairpin in $R_\ell$, add $f$, s.t.~$f[\val_S]=\Ehp{S_i,S_j,j-i}$
%   \item if $(i,j)$ closes an interior loop, bulge or stack with inner base pair $(i_1,j_1)$ in $R_\ell$, add $f$, s.t.~$f[\val_S]=\Ehp{S_i,S_j,S_k,S_l,i_1-i+j_1-j}$
%   \item if $(i,j)$ closes a multiloop add $f$, s.t.~$f[\val_S]=a$.
%   \item if $(i,j)$ is inner base pair of a multiloop of $R_\ell$ add $f$, s.t.~$f[\val_S]=b$.
% \end{itemize}
% Finally add for each $R_\ell$ and unpaired base $i$ in a multi loop of $R_\ell$, the function $f$, $f[\val_S]=c$.

More complex \Def{Loop-based}
 energy models ---e.g.~the Turner model,
which among others adds energy terms for special loops and dangling
ends---are encoded as straightforward extensions. An interesting
stripped-down variant of the nearest neighbor model is the
\Def{stacking energy model}. This model assigns non-zero energy
contributions only to stacks, i.e. interior loops with closing base
pair $(i,j)$ and inner base pair $(i+1,j-1)$.

The arity of the introduced functions provides an important bound on the
treewidth of the network (and therefore computational
complexity). Thus, it is noteworthy that the base pair energy model
requires only binary functions; the stacking model, only quarternary
dependencies. This arity is increased in a few cases by the commonly
used Turner 2004 model~\cite{Turner2009} for encoding tabulated
special hairpin and interior loop contributions, which depend on up to
nine bases for the interior loops with a total of 5 unpaired bases
(``2x3'' interior loops)---all other energy contributions (like
dangling ends) still depend on at most four bases of the sequence.

\subsection{Extension to multidimensional Boltzmann sampling}\label{sec:multiBoltzmann}
The flexibility of our framework supports an advanced sampling technique, named multidimensional Boltzmann sampling~\cite{Bodini2010} to (probabilistically) enforce additional constraints.
This technique was previously used to control the $\Gb\Cb$-content~\cite{Waldispuehl2011,Reinharz2013} and dinucleotide content~\cite{Zhang2013} of sampled RNA sequences; it enables explicit control of the free energies $(\TargetE_1,\ldots,\TargetE_k)$ of the single targets. %We use multidimensional Boltzmann sampling to generate sequences having predetermined targeted energies $(\TargetE_1,\ldots,\TargetE_k)$ for the $k$ structures.

Multidimensional Boltzmann sampling requires the ability to \Def{sample from a weighted distribution} over the set of compatible sequences, where the probability of a sequence $S$ with free energies $(E_1,\ldots,E_k)$ for its target structures is
$\mathbb{P}(S\mid \pmb{\pi}) = \frac{\prod_{\ell=1}^{k} \pi_i^{-E_i}}{\partfun{\pmb{\pi}}},$
where $\pmb{\pi}:=(\pi_1\cdots\pi_k)$ is a vector of positive real-valued \Def{weights}, and $\partfun{\pmb{\pi}}$ is the weighted partition function. Such a distribution can be induced by a simple modification of the functions described in Sec.~\ref{sec:energy_models}, where any energy function $E(\val)$ for a structure $\ell$ is replaced by $E'(\val):= \log(\pi_\ell)\, E(\val)/\beta$. The probability of a sequence $S$ is thus proportional to 
$ \prod_{\ell=1}^{k} e^{-\log(\pi_\ell)\, E_i} = \prod_{\ell=1}^{k} \pi_i^{-E_i}. $

One then needs to \Def{learn a weights vector} $\pmb{\pi}$ such that, on average, the targeted energies are achieved by a random sequences in the weighted distribution, \emph{i.e.} such that  $\mathbb{E}(E_\ell(S)\mid \pmb{\pi})=\TargetE_\ell$,  $\forall\ell\in[1,k]$.
The expected value of $E_\ell$ is always decreasing for increasing weights $\pi_\ell$ (see App.~\ref{sec:weight_derivatives}). More generally, computing a suitable $\pmb{\pi}$ can be restated as a convex optimization problem, and be efficiently solved using a wide array of methods~\cite{Denise2010,Bendkowski2017}. 
In practice, we use a simple heuristics which starts from an initial weight vector $\pmb{\pi}^{[0]}:=(exp(\beta),\dots,\exp(\beta))$ and, at each iteration, generates a sample $\mathcal{S}$ of sequences. The expected value of an energy $E_\ell$ is estimated as $\hat\mu_\ell(\mathcal{S}) = \sum_{S\in\mathcal{S}}E_\ell(S)/|\mathcal{S}|$, and the weights are updated at the $t$-th iteration by %  $\pi_\ell^{[t+1]} = \pi_\ell^{[t]}\cdot\TargetE_\ell/\hat\mu_\ell(\mathcal{S})$. 
$\pi_\ell^{[t+1]} = \pi_\ell^{[t]}\cdot \gamma^{\hat\mu_\ell(\mathcal{S})-\TargetE_\ell}$.
While heuristic in nature, this basic iteration was elected in our initial version of \ourprog{} because of its reasonable empirical behavior.

A further \Def{rejection step} is applied to the generated structures to retain only sequences whose energy for each structure $R_\ell$ belongs to $[\TargetE_\ell\cdot(1-\varepsilon),\TargetE_\ell\cdot(1+\varepsilon)]$, for $\varepsilon\ge 0$ some predefined \Def{tolerance}. The rationale for a rejection approach follows from: 
i)\emph{Enacting an exact control over the energies would  be technically hard and costly.} Indeed, controlling the energies through dynamic programming would require explicit convolution products, generalizing Cupal \emph{et al}~\cite{Cupal1996}, inducing additional $\Theta(n^{2k})$ time and $\Theta(n^k)$ space overheads;
ii) \emph{Induced distributions are typically concentrated.} Intuitively, unless sequences are fully constrained individual energy terms are independent enough so that their sum is concentrated around its mean -- the targeted energy (cf Fig.~\ref{fig:energydist}). 
%Indeed, as shown in Figure~\ref{fig:shifting_mean}, empirical joint distributions in the energies have good fits towards Normal multidimensional distributions with linear (co-)variances. 
For base pair-based energy models and special compatibility graph (paths, cycles\ldots) this property rigorously follows from analytic combinatorics, see Bender-Richmond-Williamson~\cite{Bender1983} and Drmota~\cite{Drmota1997}. In such cases, the expected number of rejections before reaching the targeted energies grows like $\Theta(1)$ when $\varepsilon\ge 1/\sqrt{n}$, and $\Theta(n^{k/2})$ when $\varepsilon=0$. The \Gb\Cb-content of designs can also be controlled, jointly and in a similar way, as done in \Software{IncaRNAtion}~\cite{Reinharz2013}.

\section{Results}\label{sec:results}

\subsection{Implementation targeting Turner energies and GC-content}
We implemented the Boltzmann sampling approach (Algorithms
\ref{alg:pf} and \ref{alg:sampling}), performing sampling for given
target structures and weights $\pi_1,\dots,\pi_n$; moreover on top,
multi-dimensional Boltzmann sampling (see
Section~\ref{sec:multiBoltzmann}) to target specific energies and
\Gb\Cb-content.  Our tool \ourprog{} evaluates energies according to the base
pair energy model or the stacking energy model, using parameters which
were trained to approximate Turner energies
(Appendix~\ref{appsec:modelparameters}).
%
%\paragraph{Implementation of multi-dimensional Boltzmann sampling}
%As suggested in Section~\ref{sec:multiBoltzmann}, we heuristically optimize weights to target specific energies and GC content in an iterative procedure (based on the core sampling algorithm).
%Specifically, energy weights are initialized as $w_\ell=e^{\frac{1}{RT}}$; the gc weight as $w_{gc}=1$. For each iteration, $\psi\cdot n$ sequences are sampled with the current weights. Then, we determine the mean energies $E(S;R_\ell)$ and the mean GC content over the sampled sequences. For the next iteration, the weights are updated by the rules $w_\ell = w_\ell \gamma^{mean(E(S;R_\ell))-E_\ell}$ and $w_{gc} = w_{gc}\frac{GC_{target}}{mean(GC)}$. During this procedure, we collect   in a set $P$ all the sequences exhibiting energies within $[E_1-\delta,E_1+\delta]\times [E_2-\delta,E_2+\delta]\ldots$ and and GC-content within $[GC_{target}-\idelta',GC_{target}+\delta']$. 
%If $|P| > n$ or an maximum amount of iterations is reached, the sequences in $P$ are returned.
%

To capture the realistic Turner model $E_{\mathcal{T}}$, we exploit its very good correlation (supp. Fig~\ref{fig:training-cor}) with our simple stacking-based model $E_{\mathcal{S}}$. Namely, we observe a structure-specific affine dependency between these Turner and stacking energy models, so that $E_{\mathcal{T}}(S_\ell;R_\ell) \approx \gamma_\ell\, E_{\mathcal{S};R_\ell}(S_\ell) + \delta_\ell$ for each structure $R_\ell$. We learn the $(\gamma_\ell,\delta_\ell)$ parameters from a set of sequences generated with homogenous weights $w=e^{\beta}$, tuning only the \Gb\Cb-content to a predetermined target frequency.  Finally, we adjust the targeted energy of our stacking model to $E_{\mathcal{S}}^{\star} = (E_{\mathcal{T}}^{\star}- \delta_\ell)/\gamma_\ell$.

To illustrate our above strategy, we sampled $n=1\,000$ sequences
targeting ${\sf GC}_{target}=0.5$ and different Turner energies for
the three structure targets of an example
instance. Figure~\ref{fig:energydist}A illustrates how the Turner
energy distributions of the shown target structures can be accurately
shifted to prescribed target energies; for comparison, we plot the
respective energy distribution of uniformly and Boltzmann sampled
sequences. Figure~\ref{fig:energydist}B summarizes the targeting accuracy for the single structure energies over a larger set of instances
from the \texttt{Modena} benchmark. We emphasize that only the simple stacking energies
are directly targeted (at $\pm$10\% tolerance, expect for a few hard
instances). Due to our linear adjustment, we still achieve well
controlled energies in the ultimately relevant Turner energy model.
\begin{figure*}[t]
  \begin{center}
    {\sf \bfseries A}\includegraphicstop[width=0.57\textwidth]{Figs/energy_distribution}
    {\sf \bfseries B}\includegraphicstop[width=0.32\textwidth]{Figs/offset}
  \end{center}
  \caption{%
  	Targeting specific energies using multi-dimensional Boltzmann sampling. (A) Turner energy distributions for three target structures (right) while targeting $(-40,-40,-20)$ and $(-20,-20,-20)$ free energy vectors. For comparison, we show the distribution of uniform and Boltzmann samples; respectively associated with homogenous weights $1$ and $e^\beta$. (B) Accuracy of targeting over all target structures of the \Software{Modena} benchmark instances \texttt{RNAtabupath}, \texttt{3str} and \texttt{4str}. Shown is the relative deviation of targeted and achieved energies in the simple stacking model and the Turner model.
%    ($\delta=0.05, \delta'=0.1, \gamma=1.1, \psi=20, n=1000$)
}
  \label{fig:energydist}
\end{figure*}
\subsection{Generating high-quality seeds for further optimization}
We empirically study the capacity of \ourprog{} to improve the generation of seed sequences for subsequent local optimizations, showing an important application in biologically relevant scenario.

For a multi-target design instance, individual reasonable target energies are estimated from averaging the energy over samples at relatively high weights $e^{\beta}$, controlling the \Gb\Cb-content at 50\%. For each instance of the \Software{Modena} benchmark~\cite{Taneda2015}, we estimated such target energies and subsequently generated $1\,000$ seed sequences at these energies.
As its rationale, this procedure is tailored to produce sequences with similar Turner energy that favor the stability of the target structures having moderate \Gb\Cb-content; all these properties are desirable objectives for RNA design.
All sampled sequences are evaluated under the multi-stable design objective function of \cite{Hammer2017}:
\vspace{-12pt}
\begin{equation}
  \label{eq:blueprintobjective}
  f(S) = \frac{1}{k} \sum_\ell^{k} (E(S, R_\ell) - G(S))
   + \xi \frac{2}{k(k-1)} \sum\limits_{\ell<j}^k|E(S,R_\ell) - E(S,R_j)|
\end{equation}
\vspace{-12pt}

Using our strategy, we generated at least 1\,000 seeds per instance of the subsets \texttt{RNAtabupath}, \texttt{3str}, and \texttt{4str}, of the \Software{Modena} benchmark set
 with 2,3, and 4 structures. The results are compared, in terms of their value for the objective function of Eq.~\eqref{eq:blueprintobjective}, to seed sequences uniformly sampled using \Software{RNAblueprint}~\cite{Hammer2017}  (Table~\ref{tab:benchmark-results}, seeds). Our first analysis reveals that Boltzmann-weighted sequences constitute better seeds, associated with better values ($\sim 69\%$ improvement on average), compared to uniform seeds for \emph{all instances} of our benchmark (Supp. section~\ref{sec:validity}).

Moreover, for each generated seed sequence, we optimized the objective function $(\ref{eq:blueprintobjective})$ through an adaptive  greedy walk consisting of 500 steps. At each step, we resampled (uniformly at random) the positions of a randomly selected component in the compatibility graph, and accepted the modification only if it resulted in a gain, as described in \Software{RNAblueprint}~\cite{Hammer2017}.
%
Once again, for all instances, we observe a positive improvement of the quality of Boltzmann designs over uniform ones, suggesting a superior potential for optimization ($\sim 32\%$ avg improvement, Table~\ref{tab:benchmark-results}, optimized). This improvement is however more modest than in the absence of optimization, but we attribute this relative decay to the uniform nature of the resampling strategy, which somehow ends up canceling the benefits of the initial Boltzmann distribution, by uniformizing the sequence over longer runs. This suggests an alternative resampling strategy based on Boltzmann-distributed steps.

\begin{table}[t]
\centering
\medskip
\begin{tabular}{@{}>{\bf}l@{\quad}>{\tt}l@{\quad}@{\quad}c@{\quad}c@{\quad}c@{}}
             &   Dataset   & {\bfseries\ourprog{}} & \textbf{Uniform} & \textbf{Improvement} \\\toprule
  Seeds      & RNAtabupath & 21.67 ($\pm$4.38) & 37.74 ($\pm$6.45) & 73\%\\
             & 3str & 18.09 ($\pm$3.98) & 30.49 ($\pm$5.41) & 71\%\\
             & 4str & 19.94 ($\pm$3.84) & 32.29 ($\pm$5.24) & 63\%\\\midrule

  Optimized  & RNAtabupath & 5.84 ($\pm$1.31) & 7.95 ($\pm$1.76) & 28\%\\
             & 3str & 5.08 ($\pm$1.10) & 7.04 ($\pm$1.52) & 31\%\\
             & 4str & 8.77($\pm$1.48) & 13.13 ($\pm$2.13) & 37\% \\ \bottomrule
\end{tabular}\\[1em]

\caption{Comparison of the multi-stable design objective function (
  Eq.~(\ref{eq:blueprintobjective}) of sequences sampled by \ourprog{}
  vs.\ uniform samples for three benchmark sets; before (seeds) and
  after optimization (optimized). We report the average scores with
  standard deviations. Smaller scores are better; the final column
  reports our improvements over uniform sampling.}
\label{tab:benchmark-results}

\vspace{-12pt}
\end{table}

\section{Conclusion}
Motivated by the---here established---hardness of the problem, we introduced a general framework and efficient algorithms for design of RNA sequences to multiple target structures with fine-tuned properties. Our method combines an FPT stochastic sampling algorithm with multi-dimensional Boltzmann sampling over distributions controlled by expressive RNA energy models.
Compared to the previously best available sampling method (uniform sampling), this approach generated significantly better seed sequences for instances of a common multi-target design benchmark.

The presented method enables new possibilities for sequence generation in the field of RNA sequence design by enforcing additional constraints, like the \Gb\Cb-content, while controlling the energy of  multiple target structures. Thus it presents a major advance over previously used ad-hoc sampling and even uniform sampling strategies. We have shown the practicality of such controlled sequence generation and studied its use for multi-target RNA design. Moreover, our framework is already equipped to include more complex sequence constraints, including mandatory/forbidden motifs at specific positions or anywhere in the designed sequences. It could also support negative design principles, for instance by penalizing a set of alternative helices/structures. However, such extensions require delicate calibration, and will be the object of future work.

\Final{
Extensions: 
\begin{enumerate}
  \item What if the input graph is not bipartite? \\
  Extract a minimal subset of edges such that new graph is bipartite!\\ 
  $\to $ Graph bipartization (cf implementation by Hüffner, JGAA 2009)
  \item Is the optimization version in the \emph{weighted base-pair model} already {\sf NP}-hard?
  \item Does it help to know that there are cliques of size $k$ during the tree decomposition?\\
  If anything, it provides a lower bound on tree decomposition (helps branch and bound algorithms)
  \item What if some degree of freedom is allowed in the structure definition?\\ Can we still use tree decomposition?\\
  This looks like a very reasonable first step towards the definition of an {\em optimal} DP decomposition for RNA folding (with Pseudoknots) or RNA-RNA interactions\ldots
  \item Structures can also be penalized, but they must be specified explicitly.\\
  However, any competing structure must consist of helices, and it is possible to penalize each of the $\Theta(n^2)$ helices of any given length $L$. What would be the impact on treewidth?
  \item More general energy  models $\to$ Turner. (Pseudoknots)
  For this description, we assume that all input structures are non-crossing, while the framework would otherwise support crossing input structures as well as according energy models (e.g. the Hotknots model~\cite{Ren2005}) by considering special pseudoknot loops.
\end{enumerate}\label{lbl:theend}
}

\newpage 
\section*{Acknowledgements}
YP is supported by the French/Austrian RNALands project (ANR-14-CE34-0011), jointly supported by the French {\em Agence Nationale de la Recherche} and the Austrian FWF.
SH is supported within the
German Network for Bioinformatics Infrastructure (de.NBI) by the
German Federal Ministry of Education and Research (BMBF; support code
031A538B) and the project RiboNets acknowledged by financial support of the Future and Emerging Technologies (FET) programme within the Seventh Framework Programme for Research of the European Commission, under FET-Open grant number: 323987. 
All authors wish to express their gratitude to Leonid Chindelevitch for very accurately pointing out that the bipartite nature of the dependency graph directly implies a reduced complexity for our dynamic programming scheme. We also thank Arie Koster for fruitful discussions regarding the practical computation of the tree decomposition.

\bibliographystyle{natbib}
%\bibliographystyle{achemnat}
%\bibliographystyle{plainnat}
%\bibliographystyle{abbrv}
%\bibliographystyle{bioinformatics}
%
%\bibliographystyle{plain}
%
\bibliography{biblio}

\newpage

\appendix
{\centering \relsize{+4}Supplementary material\\%
%{\relsize{-1}\BackupTitle}\\
}

\section{Approximate counting and random generation}
In fact, not only is $\#{\sf BIS}$ a reference problem in counting complexity, but it is also a landmark problem with respect to the complexity of approximate counting problems. In this context, it is the representative for a class of $\#{\sf BIS}$-hard problems~\cite{Bulatov2013} that are easier to approximate than $\# {\sf SAT}$, yet are widely believed not to admit any Fully Polynomial-time Randomized Approximation Scheme. Recent results reveal a surprising dichotomy in the behavior of $\#{\sf BIS}$: it admits a Fully Polynomial-Time Approximation Scheme (FPTAS) for graphs of max degree $\le 5$~\cite{Weitz2006}, but is as hard to approximate as the general $\#{\sf BIS}$ problem on graphs of degree $\ge 6$~\cite{Cai2016}. In other words, there is a clear threshold, in term of the max degree, separating (relatively) easy instances from really hard ones.

Additionally, let us note that, from the classic Vizing Theorem, any bipartite graph $G$ having maximum degree $\Delta$ can be decomposed in polynomial time in exactly $\Delta$ matchings. Any such matching can be reinterpreted as a secondary structure, possibly with crossing interactions (\Def{pseudoknots}). These results have two immediate consequences for the pseudoknotted version of the multiple design counting problem.
\begin{corollary}[as follows from~\cite{Weitz2006}]The number of designs compatible with $m\le 5$ pseudoknotted RNA structures can be approximated within any fixed ratio by a deterministic polynomial-time algorithm.
\end{corollary}
\begin{corollary}[as follows from~\cite{Cai2016}]
  As soon as the number of pseudoknotted RNA structures strictly exceeds $5$, \NumDesign is as hard to approximate as {\#{\sf BIS}}.
\end{corollary}

It is worth noting that the $\#{\sf P}$ hardness of \NumDesign does not immediately imply the hardness of generating a valid design uniformly at random, as demonstrated constructively by Jerrum, Valiant and Vazirani~\cite{Jerrum1986}. However, in the same work, the authors establish a strong connection between the complexity of approximate counting and the uniform random generation. Namely, they showed that, for problems associated with self-reducible relations, approximate counting is equally as hard as (almost) uniform random generation. We conjecture that the (almost) uniform sampling of sequences from multiple structures with pseudoknots is in fact \#{\sf BIS}-hard as soon as the number of input structures strictly exceeds $5$, as indicated by Goldberg \emph{et al.}~\cite{Goldberg2004}, motivating even further our parameterized approach.

\section{Tree decomposition for RNA design instances in practice}
\label{appsec:treedecomp}

For studying the typically expected treewidths and tree decomposition
run times in multi-target design instances, we consider five sets of
multi-target RNA design instances of different complexity. Our first
set consists of the Modena benchmark instances.

In addition, we generated four sets of instances of increasing
complexity. The instances of the sets RF3, RF4, RF5, and RF6, each
respectively specify 3,4,5, and 6 target structures for sequence
length 100.  For each instance (100 instances per set), we generated a
set of $k$ ($k=3,\dots,6$) compatible structures as follows
\begin{itemize}
\item generate a random sequence of length 100
\item compute its minimum free energy structure (ViennaRNA package)
\item add the new structure to the instances if the resulting compatibility graph is bipartite
\item repeat until $k$ structures are collected
\end{itemize}
For each instance, we generated the dependency graphs in the base pair
model and in the stacking model. Then, we performed tree decomposition
(using strategy ``GreedyFillIn'' of LibTW~\cite{Dijk2006}) on each dependency
graph. The obtained treewidths are reported in
Fig.~\ref{fig:td-widths}, while Fig.~\ref{fig:td-times} shows the
corresponding run-times of the tree decomposition.
Figure~\ref{fig:td-example} shows the tree decompositions for an example
instance from set RF3.

\begin{figure}
  \hspace*{2cm}\textbf{Base pair model}\hspace{4.75cm}\textbf{Stacking model}\\[-28pt]
  \centering\includegraphics[width=0.9\textwidth]{Figs/td-widths}
  \vspace*{-24pt}
  \caption{Treewidths for multi-target RNA design instances of
    different complexity. Distributions of treewidths shown as boxplots for the base pair (left) and stacking model (right).}
  \label{fig:td-widths}
\end{figure}

\begin{figure}
  \hspace*{2cm}\textbf{Base pair model}\hspace{4.75cm}\textbf{Stacking model}\\[-28pt]
  \centering
  \includegraphics[width=0.9\textwidth]{Figs/td-times}
  \vspace*{-24pt}
  \caption{Computation time of tree decompositions for
    multi-target RNA design instances of different complexity.
    Distributions of times (in ms/instance) shown as boxplots for the base pair (left) and stacking model (right).}
  \label{fig:td-times}
\end{figure}

\begin{figure}[t!]
  %% the figure shows benchmark_inputs/RNAfold/3str/f3.100.0.inp
  \centering
  \verbatiminput{Figs/f3_100_0.txt}
  \includegraphicstop[width=0.6\textwidth]{Figs/td-example-basepair}%
  \includegraphicstop[width=0.35\textwidth]{Figs/td-example-stacking}
  \caption{Example instance from RF3 (top) with its tree decompositions in the base pair (left) and stacking model (right). The respective treewidths are 2 and 12.}
  \label{fig:td-example}
\end{figure}

\section{Expressing higher arity functions as cliques in the dependency graph}
\label{appsec:dependency-cliques}

Our implementation relies on external tools for the tree decomposition. Therefore it is not trivial that the weight functions (i.e. terms of the energy function) can be captured within the tree decompositions returned by a specific tool. In other words, it is not immediately clear how one can construct a cluster tree from an arbitrary tree decomposition for a given network.

Fortunately, we can show that, as long as the network the arguments $\vars(f)$ of a function $f$ are materialized by a clique within the network $\network$, then there exists at least one node in the tree decomposition that contains $\vars(f)$, thus allowing the evaluation of $f$.
\begin{lemma}\label{lem:cliques}
Let $G=(V,E)$ be an undirected graph, and  $T,\chi$ be a tree decomposition for $G$. For each clique $C\subset V$, there exists a node $n\in T$ such that $C\subset\chi(v)$.
\end{lemma}
\begin{proof}
  \newcommand{\TransVars}[1]{{\rm tvars}(#1)}
  \newcommand{\Children}[1]{{\rm Children}(#1)}
  For the sake of simplicity, let us consider $T$ as rooted on an arbitrary node, inducing an orientation, and denote by $\TransVars{n}\in V$ the set of vertices found in a node $n$ or its descendants, namely
    $$\TransVars{n} = \chi(n) \bigcup_{n'\in\Children{n}} \TransVars{n'}.$$

  Now consider (one of) the deepest node(s) $n\in T$, such that $C\subset\TransVars{v}$. We are going to prove that $C\subset\chi(n)$, using contradiction, by showing that $C\not\subset\chi(n)$ implies that $T$ is not a tree decomposition.

  Indeed, let us assume that $C\not\subset\chi(n)$, then there exists a node $v'\notin\chi(n)$ whose presence in $\TransVars{n}$ stems from its presence in some descendant of $n$. Let us denote by $n'\neq n$ the closest descendant of $n$ such that $v'\in\chi(n')$. Now, consider a node $v\in C$ such that $v\in \TransVars{v'}$ and $v\notin \TransVars{v'}$. Such a node always exists, otherwise $n'$, and not $n$, would be the deepest node such that $C\subset \TransVars{v'}$. From the definition of the tree decomposition, we know that neither $n'$ nor its descendants may contain $v$. On the other hand, none of the parents or siblings of $n'$ may contain $v'$. It follows that is no node $n''\in T$ whose vertex set $\chi(v'')$ includes, at the same time, $v$ and $v'$. Since both $v$ and $v'$ belong to a clique, one has $\{v,v'\}\in E$. The absence of a node in $T$ capturing an edge in $E$ contradicts our initial assumption that $T$ is tree decomposition for $G$.

  We conclude that $n$ is such that $C\not\subset\chi(n)$.
\end{proof}

This result implies that classic tree decomposition algorithms and, more importantly, implementations can be directly re-used to produce decompositions that capture energy models of arbitrary complexity, functions of arbitrary arity. Indeed, it suffices to add a clique, involving the parameters of the function, and Lemma~\ref{lem:cliques} guarantees that the tree decomposition will feature one node to which the function can be associated.

\section{Correctness of the FPT partition function algorithm}
\label{appsec:correctness}

\begin{theorem}[Correctness of Alg.~\ref{alg:pf}]
  \label{the:pfalgo-correctness}
  As computed by Alg.~\ref{alg:pf} for cluster tree $(T,\chi,\phi)$,
  the messages $\Message{u}{v}$, for all edges $u\to v\in T$, yield
  the partition functions of subtree of $u$ for the partial
  assignments $\val\in\assignments(\separator{u}{v})$, i.e. the
  messages satisfy
  \begin{equation}
   \Message{u}{v}[\val] = \sum_{\val'\in\assignments(\chi(T_r(u))-\separator{u}{v})} \quad
   \prod_{f\in\phi(T_r(u))} \exp(-\beta f[\val'|\val]),\label{eq:pfalgo-correct}
 \end{equation}
 where $\chi(T_r(u))$ denotes all $\chi$-assigned variables of nodes in $T_r(u)$;
 respectively $\phi(T_r(u))$; all $\phi$-assigned functions.
%
\end{theorem}

\begin{proof}
  Note that in more concise notation, Alg.~\ref{alg:pf} computes
  messages such that
\begin{equation}
  \Message{u}{v}[\val] := \sum_{\val'\in\assignments(\difference{u}{v})}\quad
  \prod_{f\in \phi(u) } \exp(-\beta f[\val'|\val]) \prod_{(w\to u) \in T} m_{w\to u}[\val'|\val].\label{eq:messages}
\end{equation}

Proof by induction on $T$. If $u$ is a leaf, $\chi(r) = \chi(T_r(u))$,
there are no messages sent to $u$, and $\phi(u) = \phi(T_r(u));$ implying
Eq.~$(\ref{eq:pfalgo-correct}).$
%
Otherwise, since the algorithm traverses edges in postorder, $u$ received from its children
$w_1,\dots,w_q$ the messages $m_{w_1\to u}, \dots, m_{w_q\to u}$, which satisfy Eq.~$(\ref{eq:pfalgo-correct})$ (induction hypothesis). Let $\val\in\assignments(\separator{u}{v})$; then, $\Message{u}{v}[\val]$ is computed by the algorithm according to Eq.~$(\ref{eq:messages})$. We rewrite as follows
\begin{align*}
  & \sum_{\val'\in\assignments(\difference{u}{v})}\quad
    \prod_{f\in \phi(u) } \exp(-\beta f[\val'|\val])
    \prod_{(w\to u) \in T} m_{w\to u}[\val'|\val]\\
  & = \sum_{\val'\in\assignments(\chi(u)-\separator{u}{v})}\quad
    \prod_{f\in \phi(u) } \exp(-\beta f[\val'|\val])
    \prod_{i=1}^q m_{w_i\to u}[\val'|\val]\\
  & =_{IH}
    \sum_{\val'\in\assignments(\chi(u)-\separator{u}{v})}\quad
    \prod_{f\in \phi(u) } \exp(-\beta f[\val'|\val])
    \prod_{i=1}^q \sum_{\val''\in\assignments(\chi(T_r(w_i))-\separator{w_i}{u})} \quad
    \prod_{f\in\phi(T_r(w_i))} \exp(-\beta f[\val'' | \val' | \val]) \\
& =_{*}
    \sum_{\val'\in\assignments(\chi(u)-\separator{u}{v})}\quad
  \sum_{\val''\in\assignments(\bigcup_{i=1}^q\chi(T_r(w_i))-\separator{w_i}{u})} \quad
  \prod_{f\in \phi(u) } \exp(-\beta f[\val'|\val])
  \prod_{f\in\phi(T_r(w_i))} \exp(-\beta f[\val'' | \val' | \val]) \\
  & =_{(**)} \sum_{\val'\in\assignments(\chi(T_r(u))-\separator{u}{v})}\quad
    \prod_{f\in \phi(T_r(u)) } \exp(-\beta f[\val'|\val])\\
\end{align*}
To see (*) and (**), we observe:
\begin{itemize}
\item The sets $\chi(u)-\separator{u}{v}$ and
  $\chi(T_r(w_i))-\separator{w_i}{u}$ are all disjoint due to
  Def.~\ref{def:treedecomp}, property 2. First, this property implies
  that any shared variable between the subtrees of $w_i$ and $w_j$
  must be in $\chi(w_i)$, $\chi(w_j)$ and $\chi(u)$, thus the
  variables of $\chi(T_r(w_i))-\separator{w_i}{u}$ are
  disjoint. Second, if a variable $\chi(T_r(w_i))$ occurs in
  $\chi(u)$, it must occur in $\chi(w_i)$ and consequently in $\separator{u}{v}$.
\item The union of the sets $\chi(u)-\separator{u}{v}$ and
  $\chi(T_r(w_i))-\separator{w_i}{u}$ is $\chi(T_r(u))-\separator{u}{v}).$
\end{itemize}

\end{proof}

\section{General complexity of the partition function computation by cluster tree elimination and generation of samples}
\label{appsec:algcomplexity}

\begin{proposition}
  \label{prop:general-complexity}
Given a cluster tree $(T,\chi,\phi)$, $T=(V,E)$ of the RNA design network $\network=(\X,\F)$ with treewidth $w$ and maximum separator size $s$, computing the partition function by Algorithm~\ref{alg:pf} takes $O((|F|+|V|)\cdot 4^{w+1})$ time and $O(|V| 4^s)$ space (for storing all messages). The sampling step has time complexity of $O((|F|+|V|)\cdot 4^D)$ per sample.
\end{proposition}

\begin{proof}[Proposition~\ref{prop:general-complexity}]
Let $d_u$ denote the degree of node $u$ in $T$, $s_u$ is the size of the separator between 
$u$ and its parent in $T$ rooted at $r$. For each node in the cluster tree, Alg.~\ref{alg:pf} computes one message by combining $(|\phi(u)|+d_u-1)$ functions, each time enumerating $4^{|\chi(u)|}$ combinations; Alg.~\ref{alg:sampling} computes $4^{|\chi{u}|-s_u}$ partition functions each time combining $(|\phi(u)|+d_u-1)$ functions.  $4^{|\chi(u)|}$ is bound by $4^{w+1}$ and $4^{|\chi{u}|-s_u}$ by $4^D$; moreover $\sum_{u\in V} (|\phi(u)|+d_u-1) = |F|+|V|-1$.
\end{proof}

\section{Parameters for the base pair and the stacking energy model}
\label{appsec:modelparameters}

We trained parameters for two RNA energy models to approximate the
Turner energy model, as implemented in the \Software{ViennaRNA}
package.  In the \Def{base pair model}, the total energy of a sequence $S$
for an RNA structure $R$ is given as sum of base pair energies, where
we consider six types of base pairs distinguishing by the bases A-U,
C-G or G-U (symmetrically) and between stacked and non-stacked base
pairs; here, we consider base pairs $(i,j)\in R$ \emph{stacked} iff
$(i+1,j-1)\in R$, otherwise \emph{non-stacked}. In the \Def{stacking model},
our features are defined by the stacks, i.e. pairs $(i,j)$ and
$(i+1,j-1)$ which both occur in $R$; we distinguish 18 types based on
$S_i,S_j,S_{i+1},S_{j-1}$ (i.e. all combinations that allow canonical
base pairs; the configurations $S_i,S_j,S_{i+1},S_{j-1}$ and
$S_{i+1},S_{j-1},S_j,S_i$ are symmetric).

Both models describe the energy assigned to a pair of sequence and
structure in linear dependency of the number of features and their
weights. We can thus train weights for linear predictors of the Turner
energy in both models.

For this purpose, we generated 5000 uniform random RNA sequences of
random lengths between 100 and 200. For each sample, we predict the
minimum free energy and corresponding structure using the ViennaRNA
package; then, we count the distinguished features (i.e., base pair or
stack types). The parameters are estimated fitting linear models
without intercept (\texttt{R} function \texttt{lm}). For both models,
\texttt{R} reports an adjusted R-squared value of 0.99. The resulting
parameters are reported in Tables~\ref{tab:basepairmodel} and
\ref{tab:stackingmodel}.

For validating the trained parameters, we generate a second
independent test set of random RNA sequences in the same
way. Figure~\ref{fig:training-cor} shows correlation plots for the
trained parameters in the base pair and stacking models for predicting
the Turner energies in the test set with respective correlations of
$0.95$ an $0.94$.

\begin{figure}
  \centering
  A)\includegraphicstop[width=0.4\textwidth,trim=0 0 0 50,clip]{Figs/basepaircor}
  B)\includegraphicstop[width=0.4\textwidth,trim=0 0 0 50,clip]{Figs/stackingcor}
  \caption{Validation of the trained parameters for the A) base pair
    model B) stacking model for predicting energies of the
    independently sampled sequences in the test set. We show
    correlation plots against the Turner energies reported by the
    ViennaRNA package.}
  \label{fig:training-cor}
\end{figure}

\begin{table}
  \centering
  \caption{Trained weights for the base pair energy model.} 
  \label{tab:basepairmodel}
  \begin{tabular}{c@{\quad}c@{\quad}c@{\quad}|@{\quad}c@{\quad}c@{\quad}c}
    \multicolumn{3}{c}{non-stacked} & \multicolumn{3}{c}{stacked}\\ 
    AU      & CG       & GU      & AU       & GC       & GU \\\hline
    1.26630 & -0.09070 & 0.78566 & -0.52309 & -2.10208 & -0.88474
  \end{tabular}
\end{table}

% GC_IN = -0.09070;
% AU_IN = 1.26630;
% GU_IN = 0.78566;
% GC_TERM = -2.10208;
% AU_TERM = -0.52309;
% GU_TERM = -0.88474;

\begin{table}
  \centering
  \caption{Trained weights for the stacking energy model (the rows specify $S_i,S_j$; the columns, $S_{i+1},S_{j-1}$; we do not show the symmetric weights).}
  \label{tab:stackingmodel}
  \begin{tabular}{c@{\quad}|@{\quad}c@{\quad}c@{\quad}c@{\quad}c@{\quad}c@{\quad}c@{\quad}c@{\quad}c}
       & AU & CG & GC & GU & UA & UG \\\hline
    AU &
-0.18826 &
-1.13291 &
-1.09787 &
-0.38606 &
-0.26510 &
-0.62086
    \\
    CG &
-1.11752 &
-2.23740 &
-1.89434 &
-1.22942 &
-1.10548 &
-1.44085
    \\
    GU &
-0.55066 &
-1.26209 &
-1.58478 &
-0.72185 &
-0.49625 &
-0.68876
    \\
  \end{tabular}
\end{table}

\section{Monotonicity of the partial derivatives within weight calibration}
\label{sec:weight_derivatives}
%
In weighted distributions, one witnesses a fairly predictable impact of the variation of any weight $\pi_\ell$ over the expected value $\mathbb{E}(E_\ell\mid \pmb{\pi})$, $\pmb{\pi}:= (\pi_1\cdots\pi_k)$, of the free energy $E_\ell$ of structure $\ell$.  Let us  first remind the probability of a sequence $S$ in the weighted distribution
\begin{align*} 
  \mathcal{B}_{\pmb{\pi}}(S) &= \prod_{i=1}^{k} \pi_i^{-E_i(S)}, &
  \partfun{\pmb{\pi}}&=\sum_{S'}\mathcal{B}_{\pmb{\pi}}(S') &
    \text{and}& &
  \mathbb{P}(S\mid \pmb{\pi}) &= \frac{\mathcal{B}_{\pmb{\pi}}(S)}{\partfun{\pmb{\pi}}}.
  \end{align*}

Remind also the definition of the expectation of $E_\ell(S)$, for $S$ a Boltzmann-distributed sequence:
$$\mathbb{E}(E_\ell\mid \pmb{\pi}) = \sum_S E_\ell(S)\cdot \mathbb{P}(S\mid \pmb{\pi}).$$

We first remark that the partial derivative of $\mathcal{B}$ yields
\begin{align*} 
  \frac{\partial \mathcal{B}_{\pmb{\pi}}(S)}{\partial \pi_\ell} = -E_\ell(S)\cdot \pi_\ell^{-E_\ell(S)-1}\prod_{\substack{i=1\\i\neq \ell}}^{k} \pi_i^{-E_i(S)} = \frac{-E_\ell(S)}{\pi_\ell}\cdot\mathcal{B}_{\pmb{\pi}}(S) = -E_\ell(S)\cdot \mathbb{P}(S\mid \pmb{\pi})\cdot \frac{\partfun{\pmb{\pi}}}{\pi_\ell }
\end{align*}
while the partial derivative of $\mathcal{Z}$ gives 
\begin{align*} 
  \frac{\partial \partfun{\pmb{\pi}}(S)}{\partial \pi_\ell} = \sum_{S'} -E_\ell(S')\cdot \mathbb{P}(S'\mid \pmb{\pi})\cdot \frac{\partfun{\pmb{\pi}}}{\pi_\ell }= -\mathbb{E}(E_\ell\mid \pmb{\pi})\cdot \frac{\partfun{\pmb{\pi}}}{\pi_\ell}.
\end{align*}
From these expressions, we conclude that
\begin{align*} 
  \frac{\partial \mathbb{E}(E_\ell\mid \pmb{\pi})}{\partial \pi_\ell} &= \sum_S E_\ell(S)\cdot\frac{\partial \mathbb{P}(S\mid \pmb{\pi})}{\partial \pi_\ell}\\
  %
  & = \sum_S E_\ell(S)\left(\frac{\frac{\partial \mathcal{B}_{\pmb{\pi}}(S)}{\partial \pi_\ell}}{\mathcal{Z}_{\pmb{\pi}}} - \frac{\frac{\partial \mathcal{Z}_{\pmb{\pi}}}{\partial \pi_\ell}\cdot\mathcal{B}_{\pmb{\pi}}(S)}{\mathcal{Z}_{\pmb{\pi}}^2}\right)\\
  %
  & = \sum_S E_\ell(S)\left(\frac{-E_\ell(S)\cdot \mathbb{P}(S\mid \pmb{\pi})\cdot \frac{\mathcal{Z}_{\pmb{\pi}}}{\pi_\ell }}{\mathcal{Z}_{\pmb{\pi}}} + \frac{\mathbb{E}(E_\ell\mid \pmb{\pi})\cdot \frac{\mathcal{Z}_{\pmb{\pi}}}{\pi_\ell}\cdot\mathcal{B}_{\pmb{\pi}}(S)}{\mathcal{Z}_{\pmb{\pi}}^2}\right)\\
  %
  & = \sum_S - \frac{E_\ell(S)^2\cdot\mathbb{P}(S\mid \pmb{\pi})}{\pi_\ell} + \frac{\mathbb{E}(E_\ell\mid \pmb{\pi})}{\pi_\ell}\sum_S E_\ell(S)\cdot \mathbb{P}(S\mid \pmb{\pi})\\
  & = -\frac{\mathbb{E}(E_\ell^2\mid \pmb{\pi}) - \mathbb{E}(E_\ell\mid \pmb{\pi})^2}{\pi_\ell}
\end{align*}
In the numerator of the above expression, one recognizes the variance of the distribution of $E_\ell$.
Remark that a variance is always non-negative and, in our case, also strictly positive for $\pmb{\pi}$, $\pi_\ell>0$, as soon as there exist at least two distinct sequences $S$ and $S'$ such that $E_\ell(S)\neq E_\ell(S')$.
Moreover, weights are always positive so the partial derivative in $\pi_\ell$ is always positive. Ultimately, it is always possible to increase (resp. decrease) the expected free energy of a structure by simply decreasing (resp. increasing) its weight, supporting our weight optimization procedure.
\newpage
\section{Detailed result of quality analysis}\label{sec:validity}
N/A values correspond to data that could not be obtained by the time of this submission for two main reasons: In the case of the initial sampling (left columns), they correspond to instances which, in conjunction with an expressive energy model, resulted in very high tree-width, leading to unreasonable memory requirements incompatible with our available computing resources; The case of missing values after optimization  (right columns), they indicate situations where the initial optimization took too long ($\approx$ 1 day) and was killed.
\input{table}

\end{document}

### 2018-01-04 RNARedPrint-Bioinfo2018.tex
Update on Overleaf.

@@
 \begin{figure*}[t]
-{\centering\includegraphics[width=.8\textwidth]{Workflow}\\}
+%{\centering\includegraphics[width=.8\textwidth]{Figs/Workflow}\\}
 \caption{General outline of \ourprog{} for base pair-based energy models.
@@
   \begin{center}
-    {\sf \bfseries A}\includegraphicstop[width=0.57\textwidth]{Figs/energy_distribution}
-    {\sf \bfseries B}\includegraphicstop[width=0.32\textwidth]{Figs/offset}
+   % {\sf \bfseries A}\includegraphicstop[width=0.57\textwidth]{Figs/energy_distribution}
+    %{\sf \bfseries B}\includegraphicstop[width=0.32\textwidth]{Figs/offset}
   \end{center}
@@
   \hspace*{2cm}\textbf{Base pair model}\hspace{4.75cm}\textbf{Stacking model}\\[-28pt]
-  \centering\includegraphics[width=0.9\textwidth]{Figs/td-widths}
+  \centering%\includegraphics[width=0.9\textwidth]{Figs/td-widths}
   \vspace*{-24pt}
@@
   \centering
-  \includegraphics[width=0.9\textwidth]{Figs/td-times}
+  %\includegraphics[width=0.9\textwidth]{Figs/td-times}
   \vspace*{-24pt}
@@
   \verbatiminput{Figs/f3_100_0.txt}
-  \includegraphicstop[width=0.6\textwidth]{Figs/td-example-basepair}%
-  \includegraphicstop[width=0.35\textwidth]{Figs/td-example-stacking}
+  %\includegraphicstop[width=0.6\textwidth]{Figs/td-example-basepair}%
+  %\includegraphicstop[width=0.35\textwidth]{Figs/td-example-stacking}
   \caption{Example instance from RF3 (top) with its tree decompositions in the base pair (left) and stacking model (right). The respective treewidths are 2 and 12.}
@@
   \centering
-  A)\includegraphicstop[width=0.4\textwidth,trim=0 0 0 50,clip]{Figs/basepaircor}
-  B)\includegraphicstop[width=0.4\textwidth,trim=0 0 0 50,clip]{Figs/stackingcor}
+  A)%\includegraphicstop[width=0.4\textwidth,trim=0 0 0 50,clip]{Figs/basepaircor}
+  B)%\includegraphicstop[width=0.4\textwidth,trim=0 0 0 50,clip]{Figs/stackingcor}
   \caption{Validation of the trained parameters for the A) base pair

### 2018-02-13 response_letter.txt
Add 'response letter'

----------------------- REVIEW 1 ---------------------
PAPER: 139
TITLE: Fixed-Parameter Tractable Sampling for RNA Design with Multiple
Target Structures
AUTHORS: Stefan Hammer, Yann Ponty, Wei Wang and Sebastian Will

----------- Overall evaluation -----------
The paper presents a framework based on a basic method in
Parameterized Complexity to handle some aspects, which are argued to
be computational bottlenecks, of the design of multi-stable RNA
molecules. The central problem formulated in this paper concerns the
computation of a partition function over sequences. The precise
definition of this problem was not at all clear to me when I read the
paper for the first time (see page 3), given that there seems to be
overload of notations (in particular, S is given as argument but also
appears in the sum). Generally, the definitions given in this paper
are mostly intuitive/vague rather than mathematically
precise. However, I did understand the problems that are actually
solved subsequently from their specific explanations and
solutions. Thus, the intent of the authors can be deciphered if
desired. Given that the paper is not submitted to a TCS conference, it
might be that this writing style is in fact more appropriate, and
hence I do not know whether it should be criticized.

Firstly, the authors consider a problem where we are given a graph, and we
need to assign a letter (A,G,C or U) to each vertex so that every two
adjacent vertices u,v are assigned one of the following configurations:
{A,C}, {G,C} or {G,U}. The authors show that the problem of counting the
number of solutions to this problem is #P-hard by a simple reduction from
Independent Set on bipartite graphs.

Secondly, the authors present an algorithm for sampling
Boltzmann-weighted designs. Here, they solve a more general problem
than only the counting problem above, but can compute a partition
function depending on the assignments. The result is a parameterized
algorithm with respect to the treewidth of the input graph (Section
4.1). The design of this algorithm is completely straightforward. The
authors further suggest an optimization of their algorithm that is
based on a simple yet nice observation (from a previous paper) that
speeds-up the computation. The authors use their own notation rather
than standard notation in Parameterized Complexity. In general, the
presentation and proofs of the algorithms are quite cumbersome, making
them appear more difficult than they truly are. There also seems to be
several typos in the presentation (e.g., how can we have a|a' in the
entry m_{w\rightarrow u}?). Nevertheless, the ideas of the algorithms
themselves are correct. Next, the computation is further extended by
enriching the functions computed at individual nodes to suit
multidimensional sampling.

The algorithms above were implemented, and the authors conducted several
experiments to evaluate their approach. I have carefully read this
section, but given that this subject is completely outside my area of
expertise, I cannot comment on this. It seems to me that this part has
taken most of the efforts invested in this project. The authors also argue
here that in practice, the treewidth of the graphs relevant to their
applications will be small.

Overall, the acceptance of this paper depends on whether the biological
motivation behind it is reasonable and of interest. The technique used is
a straightforward application of a known method in Parameterized
Complexity (DP over tree decomposition for counting), and the hardness
proof is also quite simple. However, I greatly appreciate seeing this
technique applied in a new context, as well as seeing such algorithms
implemented and used in practice. In this sense, the paper is indeed quite
novel. I therefore believe that, if the biological motivation is good
enough, this paper can greatly advance relations between bioinformatics
and parameterized analysis. Although I disliked the presentation (I would
have preferred much more formal, precise definitions and arguments,
preferably also using standard notations), I have carefully read the
proofs themselves and the results in the paper are correct. I therefore
recommend acceptance, but with very low confidence.

-------------------------------------------------------------------------------------------
Main Reviewer's comment: My sub-reviewer is an expert in parameterized
complexity rather than bioinformatics. She originally set the confidence
to 2, stating that it is up to the other reviewers to decide if the
application is important in bioinformatics. My opinion is that the
application (sampling for multi-target RNA design) is an important, highly
motivated problem in structural RNAomics. Furthermore, the fact that the
authors implemented their algorithm as a tool and made their code
available via GitHub is appreciated.  Therefore, I raised the confidence
score.

----------------------- REVIEW 2 ---------------------
PAPER: 139
TITLE: Fixed-Parameter Tractable Sampling for RNA Design with Multiple
Target Structures
AUTHORS: Stefan Hammer, Yann Ponty, Wei Wang and Sebastian Will

----------- Overall evaluation -----------
The paper presents an algorithm for sampling RNA sequences with the
aim of obtaining a sequence whose energy in a given set of secondary
structures is close to a desired vector of target energies.

The algorithm is based on dynamic programming on a tree decomposition
of a graph that is built from the constraints by the given secondary
structures. The main and most difficult part of the algorithm is to
fill a dynamic programming table for computing the partition functions
of the given energies. Afterwards, the sampling is performed using the
entries of the dynamic programming table.

In addition to this algorithm a negative result about the computation
of the number of valid RNA sequences for a given set of secondary
structures is given, stating that this number cannot be computed in
polynomial time unless #P=P.

The approach to use dynamic programming on a dependency graph in the
context of sampling seems novel. The algorithm itself works more or
less as expected. The authors describe one improvement over brute-force DP
that is based on the fact that the constraint graph obtained from the
structures is bipartite.

The experimental part of the paper demonstrates that the sampling
based on the new approach leads to better results than uniform
sampling both in terms of the energy distribution of the sample
population itself but also if the samples are used as starting
solutions in a local search algorithm. A small caveat is that the
paper does not compare with any of the previous approaches for RNA
Design but in my opinion, the sampling procedure itself is
sufficiently important to make the experiments meaningful.

Overall, I recommend to accept the paper. The topic and the treatment
with dependency networks make it a timely contribution. The paper is
reasonably well-written.

Specific suggestions for improvements:

-Page 1: "The overall complexity ... grows like Theta(4^gamma) where gamma
is a parameter whose value is bounded only by the length" This seems
rather strange because if gamma is the overall length, then 4^gamma can be
trivially obtained by enumerating all sequences and then computing the
energies for each sequence. Maybe it is supposed to mean that in the
worst-case gamma can be the length of the sequence? But this is also true
for the treewidth making this remark maybe a bit unfair. I suggest to be
more precise and explain roughly what gamma is.

>>>>> Elaborate that gamma=n is worst case and in practice much smaller due to ear decomp (similar to tree width)

-Page 2 "in a principled way" Could you provide some further explanation
what "principled" means here?

>>>>> --> YANN; something like 'generic'?

-Page 3 The definition of RNA structure allows that one position i takes
part in more than one base pair (e.g. {(1,2),(1,3)} is a set of base pairs
and thus a secondary structure according to the definition).

>>>>> add this restriction (we do not need degree>1, even if it does not harm)

-Page 3 "{A,C}" -> "{A,U}"

-Page 3 "(Si,Sj)" -> "{Si,Sj}"

-Page 3 Please highlight a bit better when the compatability graph is of
interest and when the dependency graph is of interest. It seems that the
compatability graph is only interesting in Section 3?

-Page 3 It seems strange that the partition function depends on S while S
is used in the sum. Should it be "Z_E_R(n)" or "Z_E_R(S_n)"?

>>>> Embarassing typo: this should be Z_E_R(\R)

-Page 3 "as the minimal set" Is it necessarily unique? If not, "as a
minimal set".

-Page 3 "an hypergraph" -> "a hypergraph"

-Page 4 "base pair models" -> "base pair model"

-Page 4 "ins polynomial time" -> "in polynomial time"

-Page 4 "strongly connected" Why "strongly" is the graph directed? There
is no discussion of weak vs. strong connected components.

-Page 4,5 two missing "\qed"s

-Page 5 In the definition of "f[a]" one needs that "dep(f)" contains "dom(A)"

>>>> CHECK

-Page 5 The difference variables should probably be defined only for
adjacent bags of the tree decomposition? Otherwise, I don't see how the
claim about the one-variable difference is true.

-Page 5 "Denote by A(Y)" What is Y?

-Page 6 In Line 6 of the algorithm, I don't see how m_w->u has an entry
for "[a|a']" as "a" may contain some new variables.

-Page 6 "Its correctness" -> "The correctness of the algorithm"

-Page 6 Running times are not reported here but it seems crucial to obtain
good tree decompositions, I am not sure whether greedy fill-in is the best
strategy. A more recent approach is for example Flow-Cutter by Ben
Strasser (github.com/kit-algo/flow-cutter-pace17).

-Page 6 "requires O(..." + "time"

-Page 7 For Theorem 1, I was wondering whether it is really necessary to
consider the total number of connected components. It seems that only the
maximum number of components per bag is crucial?

-Page 7 "we encode the network energy ..." I was a bit confused here. What
is w_l? Is it possible to give the formula?

-Page 8 Should it "ln" instead of "log"?

>>>>> YES

-Page 8 What is "gamma" in "gamma^{mu_l}"?

-Page 8 "expected number grows like Theta(1)" This sounds a bit awkward as
Theta(1) doesn't really grow. How about "expected number is Theta(1)"?

>>>> REFORMULATE

-Page 9 In the definition of f(S), the first sum should be from l=1?

-Page 14 Lemma 1 is a well-known fact about tree decompositions. It does
not need a proof and should be found in some surveys about treewidth.

>>>>> AGREED (Discuss with Yann: drop proof and cite some text book)

----------------------- REVIEW 3 ---------------------
PAPER: 139
TITLE: Fixed-Parameter Tractable Sampling for RNA Design with Multiple
Target Structures
AUTHORS: Stefan Hammer, Yann Ponty, Wei Wang and Sebastian Will

----------- Overall evaluation -----------
The manuscript by Hammer et al. describes an general theoretical framework
for designing RNA sequences with multiple target secondary structures.
First, the authors show that the problem of counting the number of
possible designs is #P-hard. Then, they develop an algorithm for
calculating the partition function, and sampling design instances from
this distribution. The complexity of these algorithms is determined by the
tree-width of the constraint graph.

The method is timely and nicely expands recent contributions in the same
area. Moreover, I believe that the nature of the work and theoretical
results presented in this paper should be of interest for the broad RECOMB
community.

Comments:

* There are a couple of shortcuts in the notations that occasionally
distracted me (although it was correct). For instance, in definition 1,
point 3, e is an edge and X(v) is a set of vertice. Or in the definition
of E_R(S) the dependency between F and R is not explicited. Similarly,
definition 2 introduces the RNA design networks but the definition stays
at a relatively abstract level. If the object is associated with RNA, why
not using a specialized terminology (e.g. nucleotide instead of
variables)?

>>>>> Notation could be adapted; e.g. "nucleotide instead of
variables" would work for this paper (but not for all possible extensions)

* The experimental validation is limited, but it is (to my point-of-view)
not an issue as the core of the paper are the theoretical results. If
possible, it would be useful to expand this benchmark. For instance, by
including a discussion using known multi-stable RNAs and comparing real
and artificial sequences. It would be interesting to characterize which
parameters increase the difficulty of the design, or determine a priori
the range of possible target energies.

>>>>> Don't we already have artificial vs. real; can we write some discussion?

* This is an open question, but I was wondering how the authors could use
their algorithm to control the energy barrier between two target
structures. For instance, one could identify a structure near the saddle
point and force the design to assign a high energy to the latter (and by
constrast low energies to the targets).

Controling the energy barrier between two stable structure An important
feature when designing multi-stable RNA

>>>>> INTERESTING IDEA; rather for design than sampling. We could try to optimize 
the direct path barrier in this way (for producing decoys...)

* It was not clear to me if the authors implemented dangling ends?

>>>>> NO; if possible, reformulate the resepctive sentence

Typos:
* Eq 1: Z is a function of S but S is in the sum as well...
* Sec. 2, start of paragraph 2: "Given is"?

### 2018-02-13 RNARedPrint-Bioinfo2018.tex
Some editing of formal notation; re-include figures

--- 2018-01-04 RNARedPrint-Bioinfo2018.tex
@@
 
+\usepackage{stmaryrd}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
- \newtheorem{theorem}{Theoreme}
+ \newtheorem{theorem}{Theorem}
  \newtheorem{definition}[theorem]{Definition}
@@
 
-\newcommand{\Def}[1]{{\bf #1}}
+\newcommand{\Def}[1]{{\it #1}}
 
@@
 
+\newcommand{\GCb}{\Gb\Cb}
+
 \newcommand{\Software}[1]{{\ttfamily #1}}
@@
 \newcommand{\ourprog}{\Software{RNARedPrint}}
+
+\newcommand{\evalfor}[2]{#1\llbracket{}#2\rrbracket{}}
+\newcommand{\substitute}[2]{#1\!\mid\!\mid\! #2}
+
+\renewcommand{\gets}{:=}
 
@@
 
-\subtitle{Subject Section}
-
-\title[short Title]{Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures}
+\subtitle{Research Article}
+
+\title[Efficient Sampling for Multi-Target RNA Design]{Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures}
 \author[Hammer, Ponty, Wang and Will]{Stefan Hammer\,$^{\text{\sfb 1,2,3}}$, Yann Ponty\,$^{\text{\sfb 4,5,}\star}$, Wei Wang\,$^{\text{\sfb 4}}$ and Sebastian Will\,$^{\text{\sfb 2}}$}
@@
 We establish the $\#${\sf P}-hardness of uniform sampling, and introduce \ourprog{}, a tree decomposition-based algorithm for efficient, fixed parameter tractable sampling.
-By modeling the problem as a constraint network, \ourprog{} supports generic Boltzmann-weighted sampling for arbitrary additive RNA energy models; this enables generating designs meeting specific goals like expected free energies or \Gb\Cb-content.
+By modeling the problem as a constraint network, \ourprog{} supports generic Boltzmann-weighted sampling for arbitrary additive RNA energy models; this enables generating designs meeting specific goals like expected free energies or \GCb-content.
 Finally, we empirically study general properties of the approach and generate biologically relevant multi-target Boltzmann-weighted designs for a common design benchmark.
@@
 stability of) multiple structures, additionally aiming at other
-features, such as specific \Gb\Cb-content~\cite{Reinharz2013} or the
+features, such as specific \GCb-content~\cite{Reinharz2013} or the
 presence/absence of functionally relevant motifs (either anywhere or
@@
 algorithm~\cite{Hofacker1994}. This strategy was later shown to result
-in unrealistically high \Gb\Cb-contents in designed sequences. To
+in unrealistically high \GCb-contents in designed sequences. To
 address this issue, \Software{IncaRNAtion}~\cite{Reinharz2013}
-controls the \Gb\Cb-content through an adaptive sampling strategy.
+controls the \GCb-content through an adaptive sampling strategy.
 
@@
 In particular, we wish to extend the flexibility of multi-structure design, leading to the following questions: \emph{How to sample, in a principled way, from a Boltzmann distribution based on expressive energy models?
-How to enforce additional constraints, such as the \Gb\Cb-content, complex sequence constraints, or the energy of individual structures?}
+How to enforce additional constraints, such as the \GCb-content, complex sequence constraints, or the energy of individual structures?}
 
 \begin{figure*}[t]
-{\centering\includegraphics[width=.8\textwidth]{Workflow}\\}
+{\centering\includegraphics[width=.8\textwidth]{Figs/Workflow}\\}
 \caption{General outline of \ourprog{} for base pair-based energy models.
@@
 The tree is then used to compute the partition function, followed by a Boltzmann sampling of valid sequences (v).
-An adaptive scheme learns weights to achieve targeted energies and $\Gb\Cb$-content, leading to the production of suitable designs (vi).}
+An adaptive scheme learns weights to achieve targeted energies and \GCb-content, leading to the production of suitable designs (vi).}
 \label{fig:workflow}
@@
   sampling} strategy to control the free energies of the individual
-target structures and the $\Gb\Cb$-content%
+target structures and the \GCb-content%
 % (SubSection~\ref{sec:multiBoltzmann})
@@
 the energies of the single RNA structures in an RNA energy model, as
-well as sequence dependent features like \Gb\Cb-content.  Furthermore note
+well as sequence dependent features like \GCb-content.  Furthermore note
 that $E_\R(S)$ is finite iff $S$ is valid for each structure
@@
     \label{eq:mainproblem}
-    \partfun{E_\R}(S) = \sum_{S\in\S_n} \exp(-\beta E_\R(S)),
+    \partfun{E_\R} = \sum_{S\in\S_n} \exp(-\beta E_\R(S)),
   \end{equation}
@@
 %
-As will be explicited in future sections, our approach relies on breaking down the energy function $E_\R(S)$ into additive components, each depending on only few sequence positions.
+As will be explicited in subsequent sections, our approach relies on breaking down the energy function $E_\R(S)$ into additive components, each depending on only few sequence positions.
 Given $\R$, we express $E_\R(S)$
@@
   \label{def:treedecomp}
-  Given an hypergraph $G=(X, E)$,a \Def{tree decomposition} is a
+  Given a hypergraph $G=(X, E)$, a \Def{tree decomposition} is a
   pair $(T,\chi)$, where $T$ is an unrooted tree/forest, and
@@
 For assignments $\val$ and functions $f$, we define the
-special notation $f[a]$ to \Def{evaluate $f$ for $\val$}; i.e.{}
-$f[\val] := f(S),$ for any sequence $S$ where $S_i=\val(i)$ for all
+special notation $\evalfor{f}{\val}$ to \Def{evaluate $f$ for $\val$}; i.e.{}
+$\evalfor{f}{\val} := f(S),$ for any sequence $S$ where $S_i=\val(i)$ for all
 $i\in\dep(f).$ Then, the \Def{energy $\energy{\network}(S)$ of a
@@
 values of all functions in $f\in\F$ evaluated for $\val_S$, i.e.
-$\energy{\network}(S) := \sum_{f\in F} f[\val_S].$
+$\energy{\network}(S) := \sum_{f\in F} \evalfor{f}{\val_S}.$
 
@@
 $\partfun{\network} := \sum_{S}\exp(-\beta\energy{\network}(S)) = \sum_{S}\prod_{f\in F} \exp( -\beta\cdot
-f[\val_S] ).$
+\evalfor{f}{\val_S} ).$
 
 % For a given RNA design network, one is typically interested in the \Def{minimum free energy sequence} $\energy{\network}:=\min_{S}\energy{\network}(S)$, the \Def{enumeration of valid sequences} $\numberof{\network}$ and the \Def{partition function} of the network $\network$ as \begin{equation}
-%  \partfun{\network;\beta} := \sum_{S}\prod_{f\in F} \exp( -\beta\cdot f[\val_S] ).
+%  \partfun{\network;\beta} := \sum_{S}\prod_{f\in F} \exp( -\beta\cdot \evalfor{f}{\val_S} ).
 %  \end{equation}
@@
 
-Denote by $\assignments(\Y)$ the set of all assignments of the variables in $\Y$; moreover, given non-overlapping assignments $\val'$ and $\val''$, we write $\val'\mid\val''$ to denote the assignment defined as $\val'$ and $\val''$ respectively on $\dom(\val')$ and $\dom(\val'')$.
+Denote by $\assignments(\Y)$ the set of all assignments of the variables in $\Y$; moreover, given non-overlapping assignments $\val'$ and $\val''$, we write $\substitute{\val'}{\val''}$ to denote the assignment defined as $\val'$ and $\val''$ respectively on $\dom(\val')$ and $\dom(\val'')$.
 Finally we assume, w.l.o.g., that all variable difference sets $\difference{u}{v}$ are restricted to a single variable: for any given cluster tree, an equivalent (in term of treewidth) cluster tree can always be obtained by inserting at most $\Theta(|\X|)$ additional clusters.
@@
 \begin{math}
-  \partfun{\network} = \prod_{(u\to r)\in T} m_{u\to r}[\varnothing].
+  \partfun{\network} = \prod_{(u\to{}r)\in T} \evalfor{\Message{u}{r}}{\varnothing}.
 \end{math}
@@
  \KwData{Cluster tree $(T,\chi,\phi)$}
- \KwResult{Messages $\Message{u}{v}$ for all $(u\to v)\in T$; i.e.~partition functions of the subtrees of all $v$ for all possible assignments of variables $\separator{u}{v}$.}
- \For{$u\to v\in T$ in postorder}{
+ \KwResult{Messages $\Message{u}{v}$ for all $(u\to{}v)\in T$; i.e.~partition functions of the subtrees of all $v$ for all possible assignments of variables $\separator{u}{v}$.}
+ \For{$u\to{}v\in T$ in postorder}{
   \For{$\val\in\assignments(\separator{u}{v})$}{
-    $\Message{u}{v}[\val] \gets 0$\;
+    $\evalfor{\Message{u}{v}}{\val} \gets 0$\;
     \For{$\val'\in\assignments(\difference{u}{v})$}{
-     $p \gets$ product( $exp(-\beta f[\val\mid \val'])$ for $f\in \phi(u)$ )\;
-     $p \gets p\ \times\ $product( $\Message{w}{u}\left[\val\mid \val'\right]$ for $(w\to u)\in T$ )\;
-     $\Message{u}{v}[\val] \gets \Message{u}{v}[\val] + p$\; 
+     $p \gets$ product( $exp(-\beta \evalfor{f}{\substitute{\val}{\val'}})$ for $f\in \phi(u)$ )\;
+     $p \gets p\ \cdot\ $product( $\evalfor{\Message{w}{u}}{\substitute{\val}{\val'}}$ for $(w\to{}u)\in T$ )\;
+     $\evalfor{\Message{u}{v}}{\val} \gets \evalfor{\Message{u}{v}}{\val} + p$\; 
    }
@@
 %  \item For all assignments $\val'\in\assignments(\Delta)$, compute
-%  $$\partfun{\val'} := \prod_{f\in \phitotal(u)} \exp(-\beta f[\val'|\val]).$$
+%  $$\partfun{\val'} := \prod_{f\in \phitotal(u)} \exp(-\beta \evalfor{f}{\substitute{val'}{\val}}).$$
 %  \item Choose one of the assignments $\val'\in\assignments(\Delta)$ with probability $\partfun{\val'}/\partfun{u}$, where $\partfun{u}$ is the sum of the $\partfun{\val'}$.
-%  \item For all edges $(v',u)$ in $\edgesToR{}$:\quad update $\val':=\sample(v',\val'|\val)$.
+%  \item For all edges $(v',u)$ in $\edgesToR{}$:\quad update $\val':=\sample(v',\substitute{\val'}{\val})$.
 %  \item Return $\val'$.
@@
  \KwData{Node $u$, partial assignment $\val\in\assignments(\separator{u}{v})$;\newline
- Cluster tree $(T,\chi,\phi)$ and partition functions $\Message{u'}{v'}[\val']$, $\forall (u'\to v')\in T$ and $\val'\in\assignments(\separator{u'}{v'})$.}
+ Cluster tree $(T,\chi,\phi)$ and partition functions $\Message{u'}{v'}[\val']$, $\forall (u'\to{}v')\in T$ and $\val'\in\assignments(\separator{u'}{v'})$.}
  \KwResult{Boltzmann-distributed random assignment for the subtree rooted at $u$, under the assumption of a partial assignment $\val$.}
  \Fn{\Sample$(u,\val;T,\chi,\phi,m)$}{
-   $r \gets \Random(\Message{u}{v}[\val])$\;
+   $r \gets \Random(\evalfor{\Message{u}{v}}{\val})$\;
    \For{$\val'\in\assignments(\difference{u}{v})$}{
 
-     $p \gets$ product( $exp(-\beta f[\val\mid \val'])$ for $f\in \phi(u)$ )\;
-     $p \gets p\ \times\ $product( $\Message{w}{u}\left[\val\mid \val'\right]$ for $(w\to u)\in T$ )\;
+     $p \gets$ product( $exp(-\beta \evalfor{f}{\substitute{\val}{\val'}})$ for $f\in \phi(u)$ )\;
+     $p \gets p\ \cdot\ $product( $\evalfor{\Message{w}{u}}{\substitute{\val}{\val'}}$ for $(w\to{}u)\in T$ )\;
 %
@@
 %     \For{$f\in \phi(v)$}{
-%  	      $p \gets p \times \exp(-\beta f[\val\mid \val'])$\; 
+%  	      $p \gets p \cdot \exp(-\beta \evalfor{f}{\substitute{\val}{\val'}})$\; 
 %     }
-%     \For{$v\to w\in T$}{
-%  	      $p \gets p \times \Message{v}{w}\left[\val\mid \val'\right]$\; 
+%     \For{$v\to{}w\in T$}{
+%  	      $p \gets p \cdot \evalfor{\Message{v}{w}}{\substitute{\val}{\val'}}$\; 
 %     }
@@
   	  \If{$r<0$}{
-  	    $a_{\rm res} \gets a\mid a'$\;
-  	  \For{$(v\to w)\in T$}{
-  	      $a_{\rm res} \gets a_{\rm res} \mid \Sample(v,a\mid a';T,\chi,\phi,m)$\; 
+  	    $a_{\rm res} \gets \substitute{\val}{\val'}$\;
+  	  \For{$(v\to{}w)\in T$}{
+  	      $a_{\rm res} \gets a_{\rm res} \mid \Sample(v,\substitute{\val}{\val'};T,\chi,\phi,m)$\; 
      }
@@
 % \begin{itemize}
-%   \item if $(i,j)$ closes a hairpin in $R_\ell$, add $f$, s.t.~$f[\val_S]=\Ehp{S_i,S_j,j-i}$
-%   \item if $(i,j)$ closes an interior loop, bulge or stack with inner base pair $(i_1,j_1)$ in $R_\ell$, add $f$, s.t.~$f[\val_S]=\Ehp{S_i,S_j,S_k,S_l,i_1-i+j_1-j}$
-%   \item if $(i,j)$ closes a multiloop add $f$, s.t.~$f[\val_S]=a$.
-%   \item if $(i,j)$ is inner base pair of a multiloop of $R_\ell$ add $f$, s.t.~$f[\val_S]=b$.
+%   \item if $(i,j)$ closes a hairpin in $R_\ell$, add $f$, s.t.~$\evalfor{f}{\val_S}=\Ehp{S_i,S_j,j-i}$
+%   \item if $(i,j)$ closes an interior loop, bulge or stack with inner base pair $(i_1,j_1)$ in $R_\ell$, add $f$, s.t.~$\evalfor{f}{\val_S}=\Ehp{S_i,S_j,S_k,S_l,i_1-i+j_1-j}$
+%   \item if $(i,j)$ closes a multiloop add $f$, s.t.~$\evalfor{f}{\val_S}=a$.
+%   \item if $(i,j)$ is inner base pair of a multiloop of $R_\ell$ add $f$, s.t.~$\evalfor{f}{\val_S}=b$.
 % \end{itemize}
-% Finally add for each $R_\ell$ and unpaired base $i$ in a multi loop of $R_\ell$, the function $f$, $f[\val_S]=c$.
+% Finally add for each $R_\ell$ and unpaired base $i$ in a multi loop of $R_\ell$, the function $f$, $\evalfor{f}{\val_S}=c$.
 
@@
 The flexibility of our framework supports an advanced sampling technique, named multidimensional Boltzmann sampling~\cite{Bodini2010} to (probabilistically) enforce additional constraints.
-This technique was previously used to control the $\Gb\Cb$-content~\cite{Waldispuehl2011,Reinharz2013} and dinucleotide content~\cite{Zhang2013} of sampled RNA sequences; it enables explicit control of the free energies $(\TargetE_1,\ldots,\TargetE_k)$ of the single targets.
+This technique was previously used to control the \GCb-content~\cite{Waldispuehl2011,Reinharz2013} and dinucleotide content~\cite{Zhang2013} of sampled RNA sequences; it enables explicit control of the free energies $(\TargetE_1,\ldots,\TargetE_k)$ of the single targets.
 %We use multidimensional Boltzmann sampling to generate sequences having predetermined targeted energies $(\TargetE_1,\ldots,\TargetE_k)$ for the $k$ structures.
@@
 In such cases, the expected number of rejections before reaching the targeted energies grows like $\Theta(1)$ when $\varepsilon\ge 1/\sqrt{n}$, and $\Theta(n^{k/2})$ when $\varepsilon=0$.
-The \Gb\Cb-content of designs can also be controlled, jointly and in a similar way, as done in \Software{IncaRNAtion}~\cite{Reinharz2013}.
+The \GCb-content of designs can also be controlled, jointly and in a similar way, as done in \Software{IncaRNAtion}~\cite{Reinharz2013}.
 
@@
 Section~\ref{sec:multiBoltzmann}) to target specific energies and
-\Gb\Cb-content.  Our tool \ourprog{} evaluates energies according to the base
+\GCb-content.  Our tool \ourprog{} evaluates energies according to the base
 pair energy model or the stacking energy model, using parameters which
@@
 Namely, we observe a structure-specific affine dependency between these Turner and stacking energy models, so that $E_{\mathcal{T}}(S_\ell;R_\ell) \approx \gamma_\ell\, E_{\mathcal{S};R_\ell}(S_\ell) + \delta_\ell$ for each structure $R_\ell$.
-We learn the $(\gamma_\ell,\delta_\ell)$ parameters from a set of sequences generated with homogenous weights $w=e^{\beta}$, tuning only the \Gb\Cb-content to a predetermined target frequency.
+We learn the $(\gamma_\ell,\delta_\ell)$ parameters from a set of sequences generated with homogenous weights $w=e^{\beta}$, tuning only the \GCb-content to a predetermined target frequency.
 Finally, we adjust the targeted energy of our stacking model to $E_{\mathcal{S}}^{\star} = (E_{\mathcal{T}}^{\star}- \delta_\ell)/\gamma_\ell$.
@@
 
-For a multi-target design instance, individual reasonable target energies are estimated from averaging the energy over samples at relatively high weights $e^{\beta}$, controlling the \Gb\Cb-content at 50\%.
+For a multi-target design instance, individual reasonable target energies are estimated from averaging the energy over samples at relatively high weights $e^{\beta}$, controlling the \GCb-content at 50\%.
 For each instance of the \Software{Modena} benchmark~\cite{Taneda2015}, we estimated such target energies and subsequently generated $1\,000$ seed sequences at these energies.
-As its rationale, this procedure is tailored to produce sequences with similar Turner energy that favor the stability of the target structures having moderate \Gb\Cb-content; all these properties are desirable objectives for RNA design.
+As its rationale, this procedure is tailored to produce sequences with similar Turner energy that favor the stability of the target structures having moderate \GCb-content; all these properties are desirable objectives for RNA design.
 All sampled sequences are evaluated under the multi-stable design objective function of \cite{Hammer2017}:
 \vspace{-12pt}
-\begin{equation}
+\begin{align}
   \label{eq:blueprintobjective}
-  f(S) = \frac{1}{k} \sum_\ell^{k} (E(S, R_\ell) - G(S))
-   + \xi \frac{2}{k(k-1)} \sum\limits_{\ell<j}^k|E(S,R_\ell) - E(S,R_j)|
-\end{equation}
-\vspace{-12pt}
-
-Using our strategy, we generated at least 1\,000 seeds per instance of the subsets \texttt{RNAtabupath}, \texttt{3str}, and \texttt{4str}, of the \Software{Modena} benchmark set
- with 2,3, and 4 structures.
-The results are compared, in terms of their value for the objective function of Eq.~\eqref{eq:blueprintobjective}, to seed sequences uniformly sampled using \Software{RNAblueprint}~\cite{Hammer2017}  (Table~\ref{tab:benchmark-results}, seeds).
-Our first analysis reveals that Boltzmann-weighted sequences constitute better seeds, associated with better values ($\sim 69\%$ improvement on average), compared to uniform seeds for \emph{all instances} of our benchmark (Supp.
-section~\ref{sec:validity}).
+  f(S)= \quad & \frac{1}{k} \sum_\ell^{k} (E(S, R_\ell) - G(S))\notag\\
+   & +\ \xi \frac{2}{k(k-1)} \sum\limits_{\ell<j}^k|E(S,R_\ell) - E(S,R_j)|.
+\end{align}
+
+Using our strategy, we generated at least 1\,000 seeds per instance of
+the subsets \texttt{RNAtabupath}, \texttt{3str}, and \texttt{4str}, of
+the \Software{Modena} benchmark set with 2,3, and 4 structures. The
+results are compared, in terms of their value for the objective
+function of Eq.~\eqref{eq:blueprintobjective}, to seed sequences
+uniformly sampled using \Software{RNAblueprint}~\cite{Hammer2017}
+(Table~\ref{tab:benchmark-results}, seeds). Our first analysis reveals
+that Boltzmann-weighted sequences constitute better seeds, associated
+with better values ($\sim\! 69\%$ improvement on average), compared to
+uniform seeds for \emph{all instances} of our benchmark
+(Supp. section~\ref{sec:validity}).
 
@@
 
-The presented method enables new possibilities for sequence generation in the field of RNA sequence design by enforcing additional constraints, like the \Gb\Cb-content, while controlling the energy of  multiple target structures.
+The presented method enables new possibilities for sequence generation in the field of RNA sequence design by enforcing additional constraints, like the \GCb-content, while controlling the energy of  multiple target structures.
 Thus it presents a major advance over previously used ad-hoc sampling and even uniform sampling strategies.
@@
   Extract a minimal subset of edges such that new graph is bipartite!\\ 
-  $\to $ Graph bipartization (cf implementation by Hüffner, JGAA 2009)
+  $\to{}$ Graph bipartization (cf implementation by Hüffner, JGAA 2009)
   \item Is the optimization version in the \emph{weighted base-pair model} already {\sf NP}-hard?
@@
 \newpage
+\onecolumn
 
@@
 \begin{figure}
-  \hspace*{2cm}\textbf{Base pair model}\hspace{4.75cm}\textbf{Stacking model}\\[-28pt]
+  %\hspace*{2cm}\textbf{Base pair model}\hspace{4.75cm}\textbf{Stacking model}\\[-28pt]
   \centering\includegraphics[width=0.9\textwidth]{Figs/td-widths}
@@
 \begin{figure}
-  \hspace*{2cm}\textbf{Base pair model}\hspace{4.75cm}\textbf{Stacking model}\\[-28pt]
+  %\hspace*{2cm}\textbf{Base pair model}\hspace{4.75cm}\textbf{Stacking model}\\[-28pt]
   \centering
@@
   As computed by Alg.~\ref{alg:pf} for cluster tree $(T,\chi,\phi)$,
-  the messages $\Message{u}{v}$, for all edges $u\to v\in T$, yield
+  the messages $\Message{u}{v}$, for all edges $u\to{}v\in T$, yield
   the partition functions of subtree of $u$ for the partial
@@
   \begin{equation}
-   \Message{u}{v}[\val] = \sum_{\val'\in\assignments(\chi(T_r(u))-\separator{u}{v})} \quad
-   \prod_{f\in\phi(T_r(u))} \exp(-\beta f[\val'|\val]),\label{eq:pfalgo-correct}
+   \evalfor{\Message{u}{v}}{\val} = \sum_{\val'\in\assignments(\chi(T_r(u))-\separator{u}{v})} \quad
+   \prod_{f\in\phi(T_r(u))} \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}}),\label{eq:pfalgo-correct}
  \end{equation}
@@
 \begin{equation}
-  \Message{u}{v}[\val] := \sum_{\val'\in\assignments(\difference{u}{v})}\quad
-  \prod_{f\in \phi(u) } \exp(-\beta f[\val'|\val]) \prod_{(w\to u) \in T} m_{w\to u}[\val'|\val].\label{eq:messages}
+  \evalfor{\Message{u}{v}}{\val} := \sum_{\val'\in\assignments(\difference{u}{v})}\quad
+  \prod_{f\in \phi(u) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}}) \prod_{(w\to{}u) \in T} \evalfor{\Message{w}{u}}{\substitute{\val'}{\val}}.\label{eq:messages}
 \end{equation}
@@
 Otherwise, since the algorithm traverses edges in postorder, $u$ received from its children
-$w_1,\dots,w_q$ the messages $m_{w_1\to u}, \dots, m_{w_q\to u}$, which satisfy Eq.~$(\ref{eq:pfalgo-correct})$ (induction hypothesis).
-Let $\val\in\assignments(\separator{u}{v})$; then, $\Message{u}{v}[\val]$ is computed by the algorithm according to Eq.~$(\ref{eq:messages})$.
+$w_1,\dots,w_q$ the messages $\Message{w_1}{u}, \dots, \Message{w_q}{u}$, which satisfy Eq.~$(\ref{eq:pfalgo-correct})$ (induction hypothesis).
+Let $\val\in\assignments(\separator{u}{v})$; then, $\evalfor{\Message{u}{v}}{\val}$ is computed by the algorithm according to Eq.~$(\ref{eq:messages})$.
 We rewrite as follows
@@
   & \sum_{\val'\in\assignments(\difference{u}{v})}\quad
-    \prod_{f\in \phi(u) } \exp(-\beta f[\val'|\val])
-    \prod_{(w\to u) \in T} m_{w\to u}[\val'|\val]\\
+    \prod_{f\in \phi(u) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}})
+    \prod_{(w\to{}u) \in T} \evalfor{\Message{w}{u}}{\substitute{\val'}{\val}}\\
   & = \sum_{\val'\in\assignments(\chi(u)-\separator{u}{v})}\quad
-    \prod_{f\in \phi(u) } \exp(-\beta f[\val'|\val])
-    \prod_{i=1}^q m_{w_i\to u}[\val'|\val]\\
+    \prod_{f\in \phi(u) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}})
+    \prod_{i=1}^q \evalfor{\Message{w_i}{u}}{\substitute{\val'}{\val}}\\
   & =_{IH}
     \sum_{\val'\in\assignments(\chi(u)-\separator{u}{v})}\quad
-    \prod_{f\in \phi(u) } \exp(-\beta f[\val'|\val])
+    \prod_{f\in \phi(u) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}})
     \prod_{i=1}^q \sum_{\val''\in\assignments(\chi(T_r(w_i))-\separator{w_i}{u})} \quad
-    \prod_{f\in\phi(T_r(w_i))} \exp(-\beta f[\val'' | \val' | \val]) \\
+    \prod_{f\in\phi(T_r(w_i))} \exp(-\beta \evalfor{f}{\substitute{\val''}{\substitute{\val'}{\val}}}) \\
 & =_{*}
@@
   \sum_{\val''\in\assignments(\bigcup_{i=1}^q\chi(T_r(w_i))-\separator{w_i}{u})} \quad
-  \prod_{f\in \phi(u) } \exp(-\beta f[\val'|\val])
-  \prod_{f\in\phi(T_r(w_i))} \exp(-\beta f[\val'' | \val' | \val]) \\
+  \prod_{f\in \phi(u) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}})
+  \prod_{f\in\phi(T_r(w_i))} \exp(-\beta \evalfor{f}{\substitute{\val''}{\val'} | \val}) \\
   & =_{(**)} \sum_{\val'\in\assignments(\chi(T_r(u))-\separator{u}{v})}\quad
-    \prod_{f\in \phi(T_r(u)) } \exp(-\beta f[\val'|\val])\\
+    \prod_{f\in \phi(T_r(u)) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}})\\
 \end{align*}

### 2018-02-13 RNARedPrint-Bioinfo2018.tex
Shorten acknowledgments etc (minor)

@@
 For comparison, we show the distribution of uniform and Boltzmann samples; respectively associated with homogenous weights $1$ and $e^\beta$.
-(B) Accuracy of targeting over all target structures of the \Software{Modena} benchmark instances \texttt{RNAtabupath}, \texttt{3str} and \texttt{4str}.
+(B) Accuracy of targeting over all target structures of the \Software{Modena} benchmark instances \texttt{RNAtabupath} (2str), \texttt{3str} and \texttt{4str}.
 Shown is the relative deviation of targeted and achieved energies in the simple stacking model and the Turner model.
@@
 Using our strategy, we generated at least 1\,000 seeds per instance of
-the subsets \texttt{RNAtabupath}, \texttt{3str}, and \texttt{4str}, of
+the subsets \texttt{RNAtabupath} (2str), \texttt{3str}, and \texttt{4str}, of
 the \Software{Modena} benchmark set with 2,3, and 4 structures. The
@@
 \begin{tabular}{@{}>{\bf}l@{\quad}>{\tt}l@{\quad}@{\quad}c@{\quad}c@{\quad}c@{}}
-             &   Dataset   & {\bfseries\ourprog{}} & \textbf{Uniform} & \textbf{Improvement} \\\toprule
-  Seeds      & RNAtabupath & 21.67 ($\pm$4.38) & 37.74 ($\pm$6.45) & 73\%\\
+             &   \textbf{\textrm{Dataset}}   & {\bfseries\ourprog{}} & \textbf{Uniform} & \textbf{Improvement} \\\toprule
+  Seeds      & 2str & 21.67 ($\pm$4.38) & 37.74 ($\pm$6.45) & 73\%\\
              & 3str & 18.09 ($\pm$3.98) & 30.49 ($\pm$5.41) & 71\%\\
@@
 
-  Optimized  & RNAtabupath & 5.84 ($\pm$1.31) & 7.95 ($\pm$1.76) & 28\%\\
+  Optimized  & 2str & 5.84 ($\pm$1.31) & 7.95 ($\pm$1.76) & 28\%\\
              & 3str & 5.08 ($\pm$1.10) & 7.04 ($\pm$1.52) & 31\%\\
@@
 \section*{Acknowledgements}
-YP is supported by the French/Austrian RNALands project (ANR-14-CE34-0011), jointly supported by the French {\em Agence Nationale de la Recherche} and the Austrian FWF.
-SH is supported within the
-German Network for Bioinformatics Infrastructure (de.NBI) by the
-German Federal Ministry of Education and Research (BMBF; support code
-031A538B) and the project RiboNets acknowledged by financial support of the Future and Emerging Technologies (FET) programme within the Seventh Framework Programme for Research of the European Commission, under FET-Open grant number: 323987.
-All authors wish to express their gratitude to Leonid Chindelevitch for very accurately pointing out that the bipartite nature of the dependency graph directly implies a reduced complexity for our dynamic programming scheme.
-We also thank Arie Koster for fruitful discussions regarding the practical computation of the tree decomposition.
+YP is supported 
+%jointly supported 
+by
+% the French 
+{\em Agence
+  Nationale de la Recherche} and the 
+%Austrian
+FWF (ANR-14-CE34-0011; project RNALands).  SH is supported by the
+German Federal Ministry of Education and Research (BMBF support code
+031A538B; de.NBI: German Network for Bioinformatics Infrastructure)
+and the
+% acknowledged by
+%financial support 
+of the Future and Emerging Technologies programme
+% within the Seventh Framework Programme for Research of the
+%European Commission, under 
+(FET-Open grant 323987; project RiboNets).
+%
+We thank Leonid Chindelevitch for pointing out the potential for
+complexity reductions due to the bipartiteness of the compatibility
+graph; moreover, Arie Koster for discussing the practical computation
+of the tree decomposition.
 

### 2018-02-13 RNARedPrint-Bioinfo2018.tex
Minor?

--- 2018-01-04 RNARedPrint-Bioinfo2018.tex
@@
 
-\title[short Title]{Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures}
+\title[Parameterized multi-target RNA design ]{Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures}
 \author[Hammer, Ponty, Wang and Will]{Stefan Hammer\,$^{\text{\sfb 1,2,3}}$, Yann Ponty\,$^{\text{\sfb 4,5,}\star}$, Wei Wang\,$^{\text{\sfb 4}}$ and Sebastian Will\,$^{\text{\sfb 2}}$}
@@
 
-\begin{theorem}[Partition function and sampling complexity]
+\begin{theorem}[Complexities]
   For sequence length $n$, $k$ target structures, treewidth $w$ and

### 2018-02-13 RNARedPrint-Bioinfo2018.tex

--- 2018-02-13 RNARedPrint-Bioinfo2018.tex
@@
 
-\makeatletter
-\let\oldlt\longtable
-\let\endoldlt\endlongtable
-\def\longtable{\@ifnextchar[\longtable@i \longtable@ii}
-\def\longtable@i[#1]{\begin{figure}[t]
-\onecolumn
-\begin{minipage}{0.5\textwidth}
-\oldlt[#1]
-}
-\def\longtable@ii{\begin{figure}[t]
-\onecolumn
-\begin{minipage}{0.5\textwidth}
-\oldlt
-}
-\def\endlongtable{\endoldlt
-\end{minipage}
-\twocolumn
-\end{figure}}
-\makeatother
+%\makeatletter
+%\let\oldlt\longtable
+%\let\endoldlt\endlongtable
+%\def\longtable{\@ifnextchar[\longtable@i \longtable@ii}
+%\def\longtable@i[#1]{\begin{figure}[t]
+%\onecolumn
+%\begin{minipage}{0.5\textwidth}
+%\oldlt[#1]
+%}
+%\def\longtable@ii{\begin{figure}[t]
+%\onecolumn
+%\begin{minipage}{0.5\textwidth}
+%\oldlt
+%}
+%\def\endlongtable{\endoldlt
+%\end{minipage}
+%\twocolumn
+%\end{figure}}
+%\makeatother
 
@@
 }
-
+\relsize{+1}
 \section{Approximate counting and random generation}

### 2018-02-13 response_letter.txt
Some editing of formal notation; re-include figures
Edits due to comments of reviewer 2

@@
 
->>>>> Elaborate that gamma=n is worst case and in practice much smaller due to ear decomp (similar to tree width)
+>>>>> TODO: Elaborate that gamma=n is worst case and in practice much
+      smaller due to ear decomp (similar to tree width)
 
@@
 
->>>>> --> YANN; something like 'generic'?
+>>>>> TODO --> YANN; something like 'generic'?
 
@@
 
->>>>> add this restriction (we do not need degree>1, even if it does not harm)
+>>>>> TODO add this restriction (we do not need degree>1, even if it does not harm)
 
@@
 
->>>> Embarassing typo: this should be Z_E_R(\R)
+>>>> DONE. Embarassing typo
 
@@
 minimal set".
+DONE
 
 -Page 3 "an hypergraph" -> "a hypergraph"
+DONE
 
 -Page 4 "base pair models" -> "base pair model"
+DONE
 
 -Page 4 "ins polynomial time" -> "in polynomial time"
+DONE
 
@@
 is no discussion of weak vs. strong connected components.
+DONE
 
@@
 -Page 5 In the definition of "f[a]" one needs that "dep(f)" contains "dom(A)"
-
->>>> CHECK
+DONE
 
@@
 claim about the one-variable difference is true.
+DONE
 
 -Page 5 "Denote by A(Y)" What is Y?
+DONE
 
@@
 for "[a|a']" as "a" may contain some new variables.
+TODO
 
 -Page 6 "Its correctness" -> "The correctness of the algorithm"
+DONE
 
@@
 
+IGNORE this for now.
+
 -Page 6 "requires O(..." + "time"
+DONE
 
@@
 
+TODO: obviously, we confused the reviewer---
+MOREOVER: strategy should be optimized. Instead of fixing all polarities in the beginning,
+it suffices to fix the polarities simultaneously in one bag....
+
 -Page 7 "we encode the network energy ..." I was a bit confused here. What
 is w_l? Is it possible to give the formula?
+>>>>>> CHECK
 
 -Page 8 Should it "ln" instead of "log"?
-
->>>>> YES
+DONE
 
 -Page 8 What is "gamma" in "gamma^{mu_l}"?
+DONE
 
@@
 Theta(1) doesn't really grow. How about "expected number is Theta(1)"?
-
->>>> REFORMULATE
+DONE
 
 -Page 9 In the definition of f(S), the first sum should be from l=1?
+DONE
 
--Page 14 Lemma 1 is a well-known fact about tree decompositions. It does
-not need a proof and should be found in some surveys about treewidth.
-
->>>>> AGREED (Discuss with Yann: drop proof and cite some text book)
+-Page 14 Lemma 1 is a well-known fact about tree decompositions. It
+does not need a proof and should be found in some surveys about
+treewidth.  >>>> -> YANN: add that this is a folklore results, which
+we show for the sake of completeness
 

### 2018-02-13 RNARedPrint-Bioinfo2018.tex
Edits due to comments of reviewer 2

--- 2018-02-13 RNARedPrint-Bioinfo2018.tex
@@
 
+\usepackage{xcolor}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\newcommand{\TODO}[1]{\Final{\begin{mdframed}[leftline=false,rightline=false,linecolor=red!70!black,linewidth=1.5pt]\color{red!70!black}\sf TODO: #1\end{mdframed}}}
+\newcommand{\TODO}[1]{\textcolor{red!70!black}{\textbf{TODO: #1}}}
+
 \newcommand{\parHead}[1]{\Final{\paragraph{#1}}}
@@
 
-\TODO{Develop questions (relevance, difficulty...). Maybe tease results?}
 \parHead{Motivation.} The exponential time and space requirements of the \Software{RNAdesign} method already raise the question of the \Def{complexity of (uniform) sampling for multi-target design}.
@@
 %over all sequences $S$, where $E_0(S):=\min_R E(S;R)$.
-%\TODO{even Taneda's objective is significantly more complex; how much do we want to talk about this...}
 %
@@
 dependencies to sequence positions introduced by each single $f$.
-Formally, define the \Def{dependencies} $\dep(f)$ of $f$ as the minimal set of sequence positions $\I\subseteq\{1,\dots,n\}$, where $f(S)=f(S')$ for all sequences $S$ and $S'$ agreeing at all positions in $\I$.
+Formally, define the \Def{dependencies} $\dep(f)$ of $f$ as a minimal set of sequence positions $\I\subseteq\{1,\dots,n\}$, where $f(S)=f(S')$ for all sequences $S$ and $S'$ agreeing at all positions in $\I$.
 
@@
 Now we turn to the complexity of $\NumDesign(G)$, the problem of computing the number of valid sequences for a given compatibility graph $G=(V,E)$.
-Note that this problem corresponds to the partition function problem in a simple $(0/\infty)$-valued base pair models,with $\beta\neq 1$.
+Note that this problem corresponds to the partition function problem in a simple $(0/\infty)$-valued base pair model, with $\beta\neq 1$.
 As previously noted~\cite{Flamm2001}, a set of target structures admits a valid design iff its compatibility graph is bipartite, which can be tested in linear time.
@@
 
-To establish that claim, we consider instances $G=(V_1\cup V_2, E)$ that are  strongly connected and bipartite ($(V_1\times V_2) \cap E = \varnothing$), noting that hardness over restricted instances implies hardness of the general problem.
+To establish that claim, we consider instances $G=(V_1\cup V_2, E)$ that are connected and bipartite ($(V_1\times V_2) \cap E = \varnothing$), noting that hardness over restricted instances implies hardness of the general problem.
 Moreover remark that, as observed in Subsec.~\ref{sec:complexity}, assigning a nucleotide to a position $u\in V$ constrains the parity ($\{\Ab,\Gb\}$ or $\{\Cb,\Ub\}$) of all positions in the connected component of $u$.
@@
 %
-For assignments $\val$ and functions $f$, we define the
-special notation $\evalfor{f}{\val}$ to \Def{evaluate $f$ for $\val$}; i.e.{}
-$\evalfor{f}{\val} := f(S),$ for any sequence $S$ where $S_i=\val(i)$ for all
-$i\in\dep(f).$ Then, the \Def{energy $\energy{\network}(S)$ of a
-  sequence} $S$ in a network $\network$ is defined as sum of the
-values of all functions in $f\in\F$ evaluated for $\val_S$, i.e.
+For assignments $\val$ and functions $f$, $\dep(f)\subseteq\dom(\val)$,
+we define the special notation $\evalfor{f}{\val}$ to \Def{evaluate
+  $f$ for $\val$}; i.e.{} $\evalfor{f}{\val} := f(S),$ for any
+sequence $S$ where $S_i=\val(i)$ for all $i\in\dep(f).$ Then, the
+\Def{energy $\energy{\network}(S)$ of a sequence} $S$ in a network
+$\network$ is defined as sum of the values of all functions in
+$f\in\F$ evaluated for $\val_S$, i.e.
 $\energy{\network}(S) := \sum_{f\in F} \evalfor{f}{\val_S}.$
@@
 
-Let us now introduce some definitions.
-A \Def{cluster tree} for the network $\network=(\X,\F)$ is a tuple $(T,\chi,\phi)$, where $(T,\chi)$ is a tree decomposition of $G_\F$, and $\phi(v)$ represents a set of functions $f$, each uniquely assigned to a node $v\in T$; $\vars(f)\subseteq\chi(v)$ and $\phi(v)\cap \phi(v')=\varnothing$ for all $v\neq v'$.
-For two nodes $v$ and $u$ of the cluster tree, define their \Def{separator} as $\separator{u}{v} := \chi(u)\cap\chi(v)$; moreover, we define the \Def{difference variables} from $u$ to $v$ by $\difference{u}{v}:=\chi(v) - \separator{u}{v}$.
-
-Denote by $\assignments(\Y)$ the set of all assignments of the variables in $\Y$; moreover, given non-overlapping assignments $\val'$ and $\val''$, we write $\substitute{\val'}{\val''}$ to denote the assignment defined as $\val'$ and $\val''$ respectively on $\dom(\val')$ and $\dom(\val'')$.
-Finally we assume, w.l.o.g., that all variable difference sets $\difference{u}{v}$ are restricted to a single variable: for any given cluster tree, an equivalent (in term of treewidth) cluster tree can always be obtained by inserting at most $\Theta(|\X|)$ additional clusters.
+Let us now introduce some definitions. A \Def{cluster tree} for the
+network $\network=(\X,\F)$ is a tuple $(T,\chi,\phi)$, where
+$(T,\chi)$ is a tree decomposition of $G_\F$, and $\phi(v)$ represents
+a set of functions $f$, each uniquely assigned to a node $v\in T$;
+$\vars(f)\subseteq\chi(v)$ and $\phi(v)\cap \phi(v')=\varnothing$ for
+all $v\neq v'$.  For two nodes $v$ and $u$ of the cluster tree, define
+their \Def{separator} as $\separator{u}{v} := \chi(u)\cap\chi(v)$;
+moreover, we define the \Def{difference variables} from $u$ to an
+adjacent $v$ by $\difference{u}{v}:=\chi(v) - \separator{u}{v}$.
+
+For a set $Y$ of variables, denote the set of all assignments of the
+variables in $\Y$ by $\assignments(\Y)$; moreover, given
+non-overlapping assignments $\val'$ and $\val''$, we write
+$\substitute{\val'}{\val''}$ to denote the assignment defined as
+$\val'$ and $\val''$ respectively on $\dom(\val')$ and $\dom(\val'')$.
+Finally we assume, w.l.o.g., that all variable difference sets
+$\difference{u}{v}$ are restricted to a single variable: for any given
+cluster tree, an equivalent (in term of treewidth) cluster tree can
+always be obtained by inserting at most $\Theta(|\X|)$ additional
+clusters.
 
@@
 $u$ for all possible assignments to the variables
-$\vars(m)=\separator{u}{v}$. Its correctness can be shown by induction over $T$ (Appendix~\ref{appsec:correctness}).
+$\vars(m)=\separator{u}{v}$. The correctness of the algorithm can be shown by induction over $T$ (Appendix~\ref{appsec:correctness}).
 After running Alg.~\ref{alg:pf}, 
@@
 without specific optimizations, running Alg.~\ref{alg:pf} requires
-$\mathcal{O}((|\F|+|V|)\cdot 4^{w+1})$ and $\mathcal{O}(|V|\cdot4^s)$ space (Appendix~\ref{appsec:algcomplexity});
+$\mathcal{O}((|\F|+|V|)\cdot 4^{w+1})$ time and $\mathcal{O}(|V|\cdot4^s)$ space (Appendix~\ref{appsec:algcomplexity});
 Alg.~\ref{alg:sampling} would require $\mathcal{O}((|\F|+|V|)\cdot 4^D)$ per
@@
 where $\pmb{\pi}:=(\pi_1\cdots\pi_k)$ is a vector of positive real-valued \Def{weights}, and $\partfun{\pmb{\pi}}$ is the weighted partition function.
-Such a distribution can be induced by a simple modification of the functions described in Sec.~\ref{sec:energy_models}, where any energy function $E(\val)$ for a structure $\ell$ is replaced by $E'(\val):= \log(\pi_\ell)\, E(\val)/\beta$.
+Such a distribution can be induced by a simple modification of the functions described in Sec.~\ref{sec:energy_models}, where any energy function $E(\val)$ for a structure $\ell$ is replaced by $E'(\val):= \ln(\pi_\ell)\, E(\val)/\beta$.
 The probability of a sequence $S$ is thus proportional to
-$ \prod_{\ell=1}^{k} e^{-\log(\pi_\ell)\, E_i} = \prod_{\ell=1}^{k} \pi_i^{-E_i}. $
+$ \prod_{\ell=1}^{k} e^{-\ln(\pi_\ell)\, E_i} = \prod_{\ell=1}^{k} \pi_i^{-E_i}. $
 
@@
 The expected value of an energy $E_\ell$ is estimated as $\hat\mu_\ell(\mathcal{S}) = \sum_{S\in\mathcal{S}}E_\ell(S)/|\mathcal{S}|$, and the weights are updated at the $t$-th iteration by %  $\pi_\ell^{[t+1]} = \pi_\ell^{[t]}\cdot\TargetE_\ell/\hat\mu_\ell(\mathcal{S})$.
-$\pi_\ell^{[t+1]} = \pi_\ell^{[t]}\cdot \gamma^{\hat\mu_\ell(\mathcal{S})-\TargetE_\ell}$.
-While heuristic in nature, this basic iteration was elected in our initial version of \ourprog{} because of its reasonable empirical behavior.
+$\pi_\ell^{[t+1]} = \pi_\ell^{[t]}\cdot \gamma^{\hat\mu_\ell(\mathcal{S})-\TargetE_\ell}$. In practice, the constant $\gamma>1$ is chosen empirically to achieve effective optimization. 
+While heuristic in nature, this basic iteration was elected in our initial version of \ourprog{} because of its reasonable empirical behavior (choosing $\gamma=1.2$).
 
@@
 %Indeed, as shown in Figure~\ref{fig:shifting_mean}, empirical joint distributions in the energies have good fits towards Normal multidimensional distributions with linear (co-)variances. 
-For base pair-based energy models and special compatibility graph (paths, cycles\ldots) this property rigorously follows from analytic combinatorics, see Bender-Richmond-Williamson~\cite{Bender1983} and Drmota~\cite{Drmota1997}.
-In such cases, the expected number of rejections before reaching the targeted energies grows like $\Theta(1)$ when $\varepsilon\ge 1/\sqrt{n}$, and $\Theta(n^{k/2})$ when $\varepsilon=0$.
-The \GCb-content of designs can also be controlled, jointly and in a similar way, as done in \Software{IncaRNAtion}~\cite{Reinharz2013}.
+For base pair-based energy models and special compatibility graph
+(paths, cycles\ldots) this property rigorously follows from analytic
+combinatorics, see Bender-Richmond-Williamson~\cite{Bender1983} and
+Drmota~\cite{Drmota1997}. In such cases, the expected number of
+rejections before reaching the targeted energies remains constant when
+$\varepsilon\ge 1/\sqrt{n}$, and $\Theta(n^{k/2})$ when
+$\varepsilon=0$. The \GCb-content of designs can also be controlled,
+jointly and in a similar way, as done in
+\Software{IncaRNAtion}~\cite{Reinharz2013}.
 
@@
 \vspace{-12pt}
+% \begin{align}
+%   f(S)= \quad & \frac{1}{k} \sum_{\ell=1}^{k} (E(S, R_\ell) - G(S))\notag\\
+%    & +\ \xi \frac{2}{k(k-1)} \sum\limits_{1\leq\ell<j\leq k}|E(S,R_\ell) - E(S,R_j)|.
+% \end{align}
 \begin{align}
   \label{eq:blueprintobjective}
-  f(S)= \quad & \frac{1}{k} \sum_\ell^{k} (E(S, R_\ell) - G(S))\notag\\
-   & +\ \xi \frac{2}{k(k-1)} \sum\limits_{\ell<j}^k|E(S,R_\ell) - E(S,R_j)|.
+    f(S)= \quad & \frac{1}{k} \sum_{\ell=1}^{k} (E(S, R_\ell) - G(S))\notag\\
+   & +\ 0.5\frac{1}{\binom{k}{2}} \sum\limits_{1\leq\ell<j\leq k}|E(S,R_\ell) - E(S,R_j)|.
 \end{align}

### 2018-02-14 RNARedPrint-Bioinfo2018.tex
More response
Moved appendix

--- 2018-02-13 RNARedPrint-Bioinfo2018.tex
@@
 \begin{itemize}
-\item generate a random sequence of length 100
-\item compute its minimum free energy structure (ViennaRNA package)
-\item add the new structure to the instances if the resulting compatibility graph is bipartite
-\item repeat until $k$ structures are collected
+\item Generate a random sequence of length 100;
+\item Compute its minimum free energy structure (ViennaRNA package);
+\item Add the new structure to the instances if the resulting compatibility graph is bipartite;
+\item Repeat until $k$ structures are collected.
 \end{itemize}

### 2018-02-14 response_letter.txt

@@
 does not need a proof and should be found in some surveys about
-treewidth.  >>>> -> YANN: add that this is a folklore results, which
-we show for the sake of completeness
+treewidth. 
+DONE
 
@@
 
->>>>> Don't we already have artificial vs. real; can we write some discussion?
+TODO Discuss:
+Adenin-Riboswitch (Schwalbe). Details aus Stefans Diss: 3-state/4-state riboswitch.
+Experimentell ueberpruefte Designs gibts bisher nur fuer 2-state.
 
@@
 >>>>> NO; if possible, reformulate the resepctive sentence
+CHECK
 
@@
 * Sec. 2, start of paragraph 2: "Given is"?
+DONE

### 2018-02-14 RNARedPrint-Bioinfo2018.tex

--- 2018-02-13 RNARedPrint-Bioinfo2018.tex
@@
 
-Given is a fixed set of target RNA structures
+We consider a fixed set of target RNA structures
 $\R:=\{R_1, \dots, R_k\}$ for sequences of length $n$. $\R$ induces a

### 2018-02-15 RNARedPrint-Bioinfo2018.tex
formatting stuffs

@@
 
-\newcommand{\Def}[1]{{\it #1}}
+\newcommand{\Def}[1]{{\bfseries #1}}
 
@@
 \item each $x\in X$ occurs in at least one $\chi(v)$;
-\item for each $x\in X$, the set $\{ v \mid x \in \chi(v) \}$ induces a connected subtree of $T$.
-\item for every hyper-edge $e\in E$ of $G$, there is a node $v\in T$, such that $e\subseteq\chi(v)$.
+\item  $\forall x\in X$, $\{ v \mid x \in \chi(v) \}$ induces a connected subtree of $T$;
+\item $\forall e\in E$ of $G$, there is a node $v\in T$, such that $e\subseteq\chi(v)$.
 \end{enumerate}
@@
 \end{definition}
-
-\section{Counting designs for multiple structures is \#{\sf P}-hard}\label{sec:counting}
+\section{Counting valid designs is \#{\sf P}-hard}\label{sec:counting}
 
@@
  Cluster tree $(T,\chi,\phi)$ and partition functions $\Message{u'}{v'}[\val']$, $\forall (u'\to{}v')\in T$ and $\val'\in\assignments(\separator{u'}{v'})$.}
- \KwResult{Boltzmann-distributed random assignment for the subtree rooted at $u$, under the assumption of a partial assignment $\val$.}
+ \KwResult{Boltzmann-distributed random assignment for the subtree rooted at $u$, assuming a partial assignment $\val$.}
  \Fn{\Sample$(u,\val;T,\chi,\phi,m)$}{
@@
 
-\subsection{Encoding sequence features, constraints, and energy
+\subsection{Sequence features, constraints, and energy
   models.}\label{sec:energy_models}
@@
 ii) \emph{Induced distributions are typically concentrated.} Intuitively, unless sequences are fully constrained individual energy terms are independent enough so that their sum is concentrated around its mean -- the targeted energy (cf Fig.~\ref{fig:energydist}).
-%Indeed, as shown in Figure~\ref{fig:shifting_mean}, empirical joint distributions in the energies have good fits towards Normal multidimensional distributions with linear (co-)variances. 
+%Indeed, as shown in Fig.~\ref{fig:shifting_mean}, empirical joint distributions in the energies have good fits towards Normal multidimensional distributions with linear (co-)variances. 
 For base pair-based energy models and special compatibility graph
@@
 
-\subsection{Implementation targeting Turner energies and GC-content}
+\subsection{Targeting Turner energies and GC-content}
 We implemented the Boltzmann sampling approach (Algorithms
@@
 the three structure targets of an example
-instance. Figure~\ref{fig:energydist}A illustrates how the Turner
+instance. Fig.~\ref{fig:energydist}A illustrates how the Turner
 energy distributions of the shown target structures can be accurately
@@
 respective energy distribution of uniformly and Boltzmann sampled
-sequences. Figure~\ref{fig:energydist}B summarizes the targeting accuracy for the single structure energies over a larger set of instances
+sequences. Fig.~\ref{fig:energydist}B summarizes the targeting accuracy for the single structure energies over a larger set of instances
 from the \texttt{Modena} benchmark. We emphasize that only the simple stacking energies
@@
 All sampled sequences are evaluated under the multi-stable design objective function of \cite{Hammer2017}:
-\vspace{-12pt}
 % \begin{align}
@@
 
-\newpage 
 \section*{Acknowledgements}
@@
 corresponding run-times of the tree decomposition.
-Figure~\ref{fig:td-example} shows the tree decompositions for an example
+Fig.~\ref{fig:td-example} shows the tree decompositions for an example
 instance from set RF3.
@@
 independent test set of random RNA sequences in the same
-way. Figure~\ref{fig:training-cor} shows correlation plots for the
+way. Fig.~\ref{fig:training-cor} shows correlation plots for the
 trained parameters in the base pair and stacking models for predicting

### 2018-02-15 RNARedPrint-Bioinfo2018.tex
Cosmetic changes

@@
 In particular, we show significant improvements over seed sequences generated by uniform sampling, the previously best available sampling strategy for multi-target design.\\
-\textbf{Availability:} \url{https://github.com/yannponty/RNARedPrint}\\
-\textbf{Contact:} Our software is freely available at: \href{yann.ponty@lix.polytechnique.fr}{yann.ponty@lix.polytechnique.fr}\\
+\textbf{Availability:} Our software is freely available at: \url{https://github.com/yannponty/RNARedPrint}\\
+\textbf{Contact:} \href{yann.ponty@lix.polytechnique.fr}{yann.ponty@lix.polytechnique.fr}\\
 \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
@@
 In particular, RNAs present themselves as well-suited tools for
-rational design targeting specific functions~\cite{Kushwaha2016}. RNA function is tightly
+rational design targeting specific functions~\citep{Kushwaha2016}. RNA function is tightly
 coupled to the formation of secondary structure, as well as changes in
 base pairing propensities and the accessibility of regions, e.g. by
-burying or exposing interaction sites~\cite{Rodrigo2014}. At the same time, the
+burying or exposing interaction sites~\citep{Rodrigo2014}. At the same time, the
 thermodynamics of RNA secondary structure is well understood and its prediction is
-computationally tractable~\cite{McCaskill1990}. Thus,  structure can serve as effective
-proxy within rational design approaches, ultimately targeting catalytic~\cite{Zhang2013} or regulatory~\cite{Rodrigo2014} functions.
+computationally tractable~\citep{McCaskill1990}. Thus,  structure can serve as effective
+proxy within rational design approaches, ultimately targeting catalytic~\citep{Zhang2013} or regulatory~\citep{Rodrigo2014} functions.
 
@@
 ligand. Riboswitches have been a popular application of rational
-design~\cite{Wachsmuth2013,Domin2017}, partly motivated by their capacity to
-act as biosensors~\cite{Findeiss2017}. At the co-transcriptional
+design~\citep{Wachsmuth2013,Domin2017}, partly motivated by their capacity to
+act as biosensors~\citep{Findeiss2017}. At the co-transcriptional
 level, certain RNA families feature alternative, evolutionarily
-conserved, transient structures~\cite{Zhu2013}, which facilitate the
+conserved, transient structures~\citep{Zhu2013}, which facilitate the
 ultimate adoption of functional structures at full elongation.  More
@@
 stability of) multiple structures, additionally aiming at other
-features, such as specific \GCb-content~\cite{Reinharz2013} or the
+features, such as specific \GCb-content~\citep{Reinharz2013} or the
 presence/absence of functionally relevant motifs (either anywhere or
-at specific positions)~\cite{Zhou2013}; these objectives motivate flexible
+at specific positions)~\citep{Zhou2013}; these objectives motivate flexible
 computational design methods.
@@
 found to be critical for the empirical performance (and solution
-quality) of RNA design methods~\cite{Levin2012}. For instance, random
+quality) of RNA design methods~\citep{Levin2012}. For instance, random
 seed generation improves the prospect of subsequent optimizations, helping
 to overcome local optima of the objective function, and induce better
-sequence diversity across designs~\cite{Reinharz2013}.  For
-single-target approaches, \Software{INFO-RNA}~\cite{Busch2006} made
+sequence diversity across designs~\citep{Reinharz2013}.  For
+single-target approaches, \Software{INFO-RNA}~\citep{Busch2006} made
 significant improvements mainly by starting its local search from the
@@
 random sequences for the early \Software{RNAinverse}
-algorithm~\cite{Hofacker1994}. This strategy was later shown to result
+algorithm~\citep{Hofacker1994}. This strategy was later shown to result
 in unrealistically high \GCb-contents in designed sequences. To
-address this issue, \Software{IncaRNAtion}~\cite{Reinharz2013}
+address this issue, \Software{IncaRNAtion}~\citep{Reinharz2013}
 controls the \GCb-content through an adaptive sampling strategy.
@@
 \parHead{Specificities and similarities of multi-target design.}
-Specifically, for multi-target design, virtually all available methods~\cite{Lyngsoe2012,HoenerzuSiederdissen2013,Taneda2015,Hammer2017} follow the same overall generation/optimization scheme.
-%
-Facing the complex sequence constraints induced by multiple targets, early methods such as \Software{Frnakenstein}~\cite{Lyngsoe2012} and \Software{Modena}~\cite{Taneda2015} did not attempt to solve sequence generation systematically, but rely on \emph{ad-hoc} sampling strategies.
-%
-Recently, the \Software{RNAdesign} approach~\cite{HoenerzuSiederdissen2013}, coupled with powerful local search within \Software{RNAblueprint}~\cite{Hammer2017}, solved the problem of sampling seeds from the uniform distribution for multiple targets.
+Specifically, for multi-target design, virtually all available methods~\citep{Lyngsoe2012,HoenerzuSiederdissen2013,Taneda2015,Hammer2017} follow the same overall generation/optimization scheme.
+%
+Facing the complex sequence constraints induced by multiple targets, early methods such as \Software{Frnakenstein}~\citep{Lyngsoe2012} and \Software{Modena}~\citep{Taneda2015} did not attempt to solve sequence generation systematically, but rely on \emph{ad-hoc} sampling strategies.
+%
+Recently, the \Software{RNAdesign} approach~\citep{HoenerzuSiederdissen2013}, coupled with powerful local search within \Software{RNAblueprint}~\citep{Hammer2017}, solved the problem of sampling seeds from the uniform distribution for multiple targets.
 These methods adopt a graph coloring perspective, initially decomposing the graph hierarchically using various decomposition algorithms, and \Def{precomputing} the number of valid sequences within each subgraph.
-The decomposition is then reinterpreted as a decision tree to perform a \Def{stochastic backtrack}, inspired by Ding and Lawrence~\cite{Ding2003}, where individual assignments are chosen with probabilities derived from the subsolution counts, leading to uniform sampling.
+The decomposition is then reinterpreted as a decision tree to perform a \Def{stochastic backtrack}, inspired by Ding and Lawrence~\citep{Ding2003}, where individual assignments are chosen with probabilities derived from the subsolution counts, leading to uniform sampling.
 The overall complexity of \Software{RNAdesign} grows like $\Theta(4^{\gamma})$, where $\gamma$ is a parameter whose value is bounded only by the length of the designed RNA.
@@
 Thus, we focus on the question: \emph{Is there a polynomial-time algorithm to count valid multi-target designs?} In Section~\ref{sec:counting}, we answer in the negative, showing that there exists no such algorithm unless ${\sf P}={\sf NP}$.
-Our result relies on a surprising bijection (up to a trivial symmetry) between valid sequences  and independent sets of a bipartite graph, being the object of recent breakthroughs in approximate counting complexity~\cite{Bulatov2013,Cai2016}.
+Our result relies on a surprising bijection (up to a trivial symmetry) between valid sequences  and independent sets of a bipartite graph, being the object of recent breakthroughs in approximate counting complexity~\citep{Bulatov2013,Cai2016}.
 The hardness of counting (and conjectured hardness of sampling) does not preclude, however, practically applicable algorithms for counting and sampling.
@@
 
-%In this work, we present a generic approach to multi-target sequence sampling that enables Boltzmann-weighted sampling (subsuming uniform sampling) for a wide class of objective functions that can even make use of the commonly used ``full-featured'' RNA energy models (like the Turner model~\cite{turner}).
+%In this work, we present a generic approach to multi-target sequence sampling that enables Boltzmann-weighted sampling (subsuming uniform sampling) for a wide class of objective functions that can even make use of the commonly used ``full-featured'' RNA energy models (like the Turner model~\citep{turner}).
 Due to its high expressivity the same framework allows combining objective functions with various hard and soft constraints, which can accommodate various desirable multi-target design scenarios.
@@
 %Naturally, also more complex objective functions are desirable for RNA design and have been discussed in the literature; e.g.
-the ensemble defect~\cite{Zadeh:2010}.
+the ensemble defect~\citep{Zadeh:2010}.
 Typically, more complex objective functions make design computationally much more challenging.
@@
 %
-As will be explicited in subsequent sections, our approach relies on breaking down the energy function $E_\R(S)$ into additive components, each depending on only few sequence positions.
+As will be explicit in subsequent sections, our approach relies on breaking down the energy function $E_\R(S)$ into additive components, each depending on only few sequence positions.
 Given $\R$, we express $E_\R(S)$
@@
 Note that this problem corresponds to the partition function problem in a simple $(0/\infty)$-valued base pair model, with $\beta\neq 1$.
-As previously noted~\cite{Flamm2001}, a set of target structures admits a valid design iff its compatibility graph is bipartite, which can be tested in linear time.
+As previously noted~\citep{Flamm2001}, a set of target structures admits a valid design iff its compatibility graph is bipartite, which can be tested in linear time.
 Moreover, without $\{\Gb, \Ub\}$ base pairs, any connected component  $C\in {\rm CC}(G)$ is entirely determined by the assignment of a single of its nucleotides.
@@
 We remind that an \Def{independent set} of $G=(V,E)$ is a subset $V'\subseteq V$ of nodes that are not connected by any edge in $E$. Let $\IS{G}$ denote the set of all independent sets in $G$.
-Now, consider the function $\Psi: \Design{G} \to \IS{G}$ defined by $ \Psi(f) := \left\{v\in V\mid f(v)\in\{\Ab,\Cb\}\right\}.$
 
 \begin{proposition}
- $\Psi$ is a bijection between $\Design{G}$ and $\IS{G}$.
+ $\Design{G}$ is in bijection with $\IS{G}$.
 \end{proposition}
 \begin{proof}
+Consider the function $\Psi: \Design{G} \to \IS{G}$ defined by $ \Psi(f) := \left\{v\in V\mid f(v)\in\{\Ab,\Cb\}\right\}.$
+
 Let us establish the injectivity of  $\Psi$, i.e. that $\Psi(f)\neq\Psi(f')$ for all $f\neq f'$.
@@
 \begin{align}
- \forall v_1\in V_1, f(v_1)&= \begin{cases} \Ab & \text{if }v_1\in S\\ \Gb & \text{if }v_1\notin S\end{cases}& 
+ \forall v_1\in V_1, f(v_1)&= \begin{cases} \Ab & \text{if }v_1\in S\\ \Gb & \text{if }v_1\notin S\end{cases}\\
  \forall v_2\in V_2, f(v_2)&= \begin{cases} \Cb & \text{if }v_2\in S\\ \Ub & \text{if }v_2\notin S\end{cases}
@@
 Now we can build on the connection between the two problems to obtain complexity results for \NumDesign.
-Counting independent sets in bipartite graphs ($\#{\sf BIS}$) is indeed a well-studied \#{\sf P}-hard problem~\cite{Ge2012}, from which we immediately conclude:
+Counting independent sets in bipartite graphs ($\#{\sf BIS}$) is indeed a well-studied \#{\sf P}-hard problem~\citep{Ge2012}, from which we immediately conclude:
 \begin{corollary}
@@
 
-\section{An FPT algorithm for sampling Boltzmann-weighted designs}
+\section{An FPT algorithm for the partition function and sampling of Boltzmann-weighted designs}
 \label{sec:FPT}
@@
 algorithm on cluster tree elimination (CTE) by Rina
-Dechter~\cite{Dechter2013}.
+Dechter~\citep{Dechter2013}.
 %
@@
 $u$ for all possible assignments to the variables
-$\vars(m)=\separator{u}{v}$. The correctness of the algorithm can be shown by induction over $T$ (Appendix~\ref{appsec:correctness}).
+$\vars(m)=\separator{u}{v}$. The correctness of the algorithm can be shown by induction over $T$ (Supp. Mat.~\ref{appsec:correctness}).
 After running Alg.~\ref{alg:pf}, 
@@
 
-The partition functions can then direct a \Def{stochastic backtrack} to achieve \Def{Boltzmann sampling of sequences}, such that one samples from the Boltzmann distribution of a given desgin network $\network$.
+The partition functions can then direct a \Def{stochastic backtrack} to achieve \Def{Boltzmann sampling of sequences}, such that one samples from the Boltzmann distribution of a given design network $\network$.
 The sampling algorithm assumes that the cluster tree was expanded and the messages $\Message{u}{v}$ for the edges in $\edgesToR{}$ are already generated by Algorithm \ref{alg:pf} for the expanded cluster tree.
@@
 First, we observed that the computation time of tree decomposition
-(\Software{GreedyFillIn}, implemented in \Software{LibTW}~\cite{Dijk2006}) for multi-target
-sampling is negligible compared to Alg.~\ref{alg:pf} (Appendix~\ref{appsec:treedecomp} and \ref{appsec:dependency-cliques});
-consequently, we discuss complexities without the tree decomposition
-step.
+(\Software{GreedyFillIn}, implemented in \Software{LibTW}~\citep{Dijk2006}) for multi-target
+sampling is negligible compared to Alg.~\ref{alg:pf} (Supp. Mat.~\ref{appsec:treedecomp} and \ref{appsec:dependency-cliques});
+we will thus omit the tree decomposition from our complexity analysis.
 
@@
 $\max_{u,v\in V} | \separator{u}{v} |$ and denote the maximum size of
-$\difference{u}{v}$ over $(u,v)\in\edgesToR{}$ as $D$.  Generally,
-without specific optimizations, running Alg.~\ref{alg:pf} requires
-$\mathcal{O}((|\F|+|V|)\cdot 4^{w+1})$ time and $\mathcal{O}(|V|\cdot4^s)$ space (Appendix~\ref{appsec:algcomplexity});
+$\difference{u}{v}$ over $(u,v)\in\edgesToR{}$ as $D$.  In the absence
+of specific optimizations, running Alg.~\ref{alg:pf} requires
+$\mathcal{O}((|\F|+|V|)\cdot 4^{w+1})$ time and $\mathcal{O}(|V|\cdot4^s)$ space (Supp. Mat.~\ref{appsec:algcomplexity});
 Alg.~\ref{alg:sampling} would require $\mathcal{O}((|\F|+|V|)\cdot 4^D)$ per
-sample on arbitrary tree decompositions (Appendix~\ref{appsec:algcomplexity}). W.l.o.g. we assume that
+sample on arbitrary tree decompositions (Supp. Mat.~\ref{appsec:algcomplexity}). W.l.o.g. we assume that
 $D=1$; note that tree decompositions can generally be transformed,
@@
 
-As already noted by Flamm \emph{et al}~\cite{Flamm2001}, the
+As already noted by \citet{Flamm2001}, the
 requirement of canonical base pairing induces a bi-partition of each connected
@@
 dependencies. This arity is increased in a few cases by the commonly
-used Turner 2004 model~\cite{Turner2009} for encoding tabulated
+used Turner 2004 model~\citep{Turner2009} for encoding tabulated
 special hairpin and interior loop contributions, which depend on up to
@@
 \subsection{Extension to multidimensional Boltzmann sampling}\label{sec:multiBoltzmann}
-The flexibility of our framework supports an advanced sampling technique, named multidimensional Boltzmann sampling~\cite{Bodini2010} to (probabilistically) enforce additional constraints.
-This technique was previously used to control the \GCb-content~\cite{Waldispuehl2011,Reinharz2013} and dinucleotide content~\cite{Zhang2013} of sampled RNA sequences; it enables explicit control of the free energies $(\TargetE_1,\ldots,\TargetE_k)$ of the single targets.
+The flexibility of our framework supports an advanced sampling technique, named multidimensional Boltzmann sampling~\citep{Bodini2010} to (probabilistically) enforce additional constraints.
+This technique was previously used to control the \GCb-content~\citep{Waldispuehl2011,Reinharz2013} and dinucleotide content~\citep{Zhang2013} of sampled RNA sequences; it enables explicit control of the free energies $(\TargetE_1,\ldots,\TargetE_k)$ of the single targets.
 %We use multidimensional Boltzmann sampling to generate sequences having predetermined targeted energies $(\TargetE_1,\ldots,\TargetE_k)$ for the $k$ structures.
@@
 One then needs to \Def{learn a weights vector} $\pmb{\pi}$ such that, on average, the targeted energies are achieved by a random sequences in the weighted distribution, \emph{i.e.} such that  $\mathbb{E}(E_\ell(S)\mid \pmb{\pi})=\TargetE_\ell$,  $\forall\ell\in[1,k]$.
-The expected value of $E_\ell$ is always decreasing for increasing weights $\pi_\ell$ (see App.~\ref{sec:weight_derivatives}).
-More generally, computing a suitable $\pmb{\pi}$ can be restated as a convex optimization problem, and be efficiently solved using a wide array of methods~\cite{Denise2010,Bendkowski2017}.
-In practice, we use a simple heuristics which starts from an initial weight vector $\pmb{\pi}^{[0]}:=(exp(\beta),\dots,\exp(\beta))$ and, at each iteration, generates a sample $\mathcal{S}$ of sequences.
+The expected value of $E_\ell$ is always decreasing for increasing weights $\pi_\ell$ (see Supp.
+Mat.~\ref{sec:weight_derivatives}).
+More generally, computing a suitable $\pmb{\pi}$ can be restated as a convex optimization problem, and be efficiently solved using a wide array of methods~\citep{Denise2010,Bendkowski2017}.
+In practice, we use a simple heuristics which starts from an initial weight vector $\pmb{\pi}^{[0]}:=(e^\beta,\dots,e^\beta)$ and, at each iteration, generates a sample $\mathcal{S}$ of sequences.
 The expected value of an energy $E_\ell$ is estimated as $\hat\mu_\ell(\mathcal{S}) = \sum_{S\in\mathcal{S}}E_\ell(S)/|\mathcal{S}|$, and the weights are updated at the $t$-th iteration by %  $\pi_\ell^{[t+1]} = \pi_\ell^{[t]}\cdot\TargetE_\ell/\hat\mu_\ell(\mathcal{S})$.
@@
 The rationale for a rejection approach follows from:
-i)\emph{Enacting an exact control over the energies would  be technically hard and costly.} Indeed, controlling the energies through dynamic programming would require explicit convolution products, generalizing Cupal \emph{et al}~\cite{Cupal1996}, inducing additional $\Theta(n^{2k})$ time and $\Theta(n^k)$ space overheads;
+i) \emph{Enacting an exact control over the energies would  be technically hard and costly.} Indeed, controlling the energies through dynamic programming would require explicit convolution products, generalizing~\citet{Cupal1996}, inducing additional $\Theta(n^{2k})$ time and $\Theta(n^k)$ space overheads;
 ii) \emph{Induced distributions are typically concentrated.} Intuitively, unless sequences are fully constrained individual energy terms are independent enough so that their sum is concentrated around its mean -- the targeted energy (cf Fig.~\ref{fig:energydist}).
@@
 (paths, cycles\ldots) this property rigorously follows from analytic
-combinatorics, see Bender-Richmond-Williamson~\cite{Bender1983} and
-Drmota~\cite{Drmota1997}. In such cases, the expected number of
+combinatorics, see Bender-Richmond-Williamson~\citep{Bender1983} and
+Drmota~\citep{Drmota1997}. In such cases, the expected number of
 rejections before reaching the targeted energies remains constant when
@@
 jointly and in a similar way, as done in
-\Software{IncaRNAtion}~\cite{Reinharz2013}.
+\Software{IncaRNAtion}~\citep{Reinharz2013}.
 
@@
 were trained to approximate Turner energies
-(Appendix~\ref{appsec:modelparameters}).
+(Supp. Mat.~\ref{appsec:modelparameters}).
 %
@@
 
-For a multi-target design instance, individual reasonable target energies are estimated from averaging the energy over samples at relatively high weights $e^{\beta}$, controlling the \GCb-content at 50\%.
-For each instance of the \Software{Modena} benchmark~\cite{Taneda2015}, we estimated such target energies and subsequently generated $1\,000$ seed sequences at these energies.
+For a multi-target design instance, individual reasonable target energies are estimated from averaging the energy over samples at relatively high weights $e^{\beta}$, setting the targeted \GCb-content to 50\%.
+For each instance of the \Software{Modena} benchmark~\citep{Taneda2015}, we estimated such target energies and subsequently generated $1\,000$ seed sequences at these energies.
 As its rationale, this procedure is tailored to produce sequences with similar Turner energy that favor the stability of the target structures having moderate \GCb-content; all these properties are desirable objectives for RNA design.
-All sampled sequences are evaluated under the multi-stable design objective function of \cite{Hammer2017}:
+All sampled sequences are evaluated under the multi-stable design objective function of \citet{Hammer2017}:
 % \begin{align}
@@
 function of Eq.~\eqref{eq:blueprintobjective}, to seed sequences
-uniformly sampled using \Software{RNAblueprint}~\cite{Hammer2017}
+uniformly sampled using \Software{RNAblueprint}~\citep{Hammer2017}
 (Table~\ref{tab:benchmark-results}, seeds). Our first analysis reveals
@@
 uniform seeds for \emph{all instances} of our benchmark
-(Supp. section~\ref{sec:validity}).
+(Supp. Mat.~\ref{sec:validity}).
 
 Moreover, for each generated seed sequence, we optimized the objective function $(\ref{eq:blueprintobjective})$ through an adaptive  greedy walk consisting of 500 steps.
-At each step, we resampled (uniformly at random) the positions of a randomly selected component in the compatibility graph, and accepted the modification only if it resulted in a gain, as described in \Software{RNAblueprint}~\cite{Hammer2017}.
+At each step, we resampled (uniformly at random) the positions of a randomly selected component in the compatibility graph, and accepted the modification only if it resulted in a gain, as described in \Software{RNAblueprint}~\citep{Hammer2017}.
 %
@@
 \label{tab:benchmark-results}
-
-\vspace{-12pt}
 \end{table}
@@
   For this description, we assume that all input structures are non-crossing, while the framework would otherwise support crossing input structures as well as according energy models (e.g.
-the Hotknots model~\cite{Ren2005}) by considering special pseudoknot loops.
+the Hotknots model~\citep{Ren2005}) by considering special pseudoknot loops.
 \end{enumerate}\label{lbl:theend}
@@
 In fact, not only is $\#{\sf BIS}$ a reference problem in counting complexity, but it is also a landmark problem with respect to the complexity of approximate counting problems.
-In this context, it is the representative for a class of $\#{\sf BIS}$-hard problems~\cite{Bulatov2013} that are easier to approximate than $\# {\sf SAT}$, yet are widely believed not to admit any Fully Polynomial-time Randomized Approximation Scheme.
-Recent results reveal a surprising dichotomy in the behavior of $\#{\sf BIS}$: it admits a Fully Polynomial-Time Approximation Scheme (FPTAS) for graphs of max degree $\le 5$~\cite{Weitz2006}, but is as hard to approximate as the general $\#{\sf BIS}$ problem on graphs of degree $\ge 6$~\cite{Cai2016}.
+In this context, it is the representative for a class of $\#{\sf BIS}$-hard problems~\citep{Bulatov2013} that are easier to approximate than $\# {\sf SAT}$, yet are widely believed not to admit any Fully Polynomial-time Randomized Approximation Scheme.
+Recent results reveal a surprising dichotomy in the behavior of $\#{\sf BIS}$: it admits a Fully Polynomial-Time Approximation Scheme (FPTAS) for graphs of max degree $\le 5$~\citep{Weitz2006}, but is as hard to approximate as the general $\#{\sf BIS}$ problem on graphs of degree $\ge 6$~\citep{Cai2016}.
 In other words, there is a clear threshold, in term of the max degree, separating (relatively) easy instances from really hard ones.
@@
 These results have two immediate consequences for the pseudoknotted version of the multiple design counting problem.
-\begin{corollary}[as follows from~\cite{Weitz2006}]The number of designs compatible with $m\le 5$ pseudoknotted RNA structures can be approximated within any fixed ratio by a deterministic polynomial-time algorithm.
+\begin{corollary}[as follows from~\citep{Weitz2006}]The number of designs compatible with $m\le 5$ pseudoknotted RNA structures can be approximated within any fixed ratio by a deterministic polynomial-time algorithm.
 \end{corollary}
-\begin{corollary}[as follows from~\cite{Cai2016}]
+\begin{corollary}[as follows from~\citep{Cai2016}]
   As soon as the number of pseudoknotted RNA structures strictly exceeds $5$, \NumDesign is as hard to approximate as {\#{\sf BIS}}.
@@
 
-It is worth noting that the $\#{\sf P}$ hardness of \NumDesign does not immediately imply the hardness of generating a valid design uniformly at random, as demonstrated constructively by Jerrum, Valiant and Vazirani~\cite{Jerrum1986}.
+It is worth noting that the $\#{\sf P}$ hardness of \NumDesign does not immediately imply the hardness of generating a valid design uniformly at random, as demonstrated constructively by Jerrum, Valiant and Vazirani~\citep{Jerrum1986}.
 However, in the same work, the authors establish a strong connection between the complexity of approximate counting and the uniform random generation.
 Namely, they showed that, for problems associated with self-reducible relations, approximate counting is equally as hard as (almost) uniform random generation.
-We conjecture that the (almost) uniform sampling of sequences from multiple structures with pseudoknots is in fact \#{\sf BIS}-hard as soon as the number of input structures strictly exceeds $5$, as indicated by Goldberg \emph{et al.}~\cite{Goldberg2004}, motivating even further our parameterized approach.
+We conjecture that the (almost) uniform sampling of sequences from multiple structures with pseudoknots is in fact \#{\sf BIS}-hard as soon as the number of input structures strictly exceeds $5$, as indicated by~\citet{Goldberg2004}, motivating even further our parameterized approach.
 
@@
 model and in the stacking model. Then, we performed tree decomposition
-(using strategy ``GreedyFillIn'' of LibTW~\cite{Dijk2006}) on each dependency
+(using strategy ``GreedyFillIn'' of LibTW~\citep{Dijk2006}) on each dependency
 graph. The obtained treewidths are reported in
@@
   \centering\includegraphics[width=0.9\textwidth]{Figs/td-widths}
-  \vspace*{-24pt}
   \caption{Treewidths for multi-target RNA design instances of
@@
   \includegraphics[width=0.9\textwidth]{Figs/td-times}
-  \vspace*{-24pt}
   \caption{Computation time of tree decompositions for

### 2018-02-15 response_letter.txt
... and more

@@
 
->>>>> TODO: Elaborate that gamma=n is worst case and in practice much
-      smaller due to ear decomp (similar to tree width)
+CHANGED (DONE).
 
@@
 
->>>>> TODO --> YANN; something like 'generic'?
+CHANGED: "How to sample, generically, from a Boltzmann distribution..."
 
@@
 
->>>>> TODO add this restriction (we do not need degree>1, even if it does not harm)
+DONE (in a formal way)
 
 -Page 3 "{A,C}" -> "{A,U}"
+DONE
 
 -Page 3 "(Si,Sj)" -> "{Si,Sj}"
+DONE
 
@@
 
+DONE: added explanation "While we typically consider
+additional dependencies (in the dependency graph), the edges in the
+compatibility graph pose constraints on the possible bases due to
+canonical base pairing, which are present in all considered settings."
+
 -Page 3 It seems strange that the partition function depends on S while S
@@
 
->>>> DONE. Embarassing typo
+DONE.
 
@@
 for "[a|a']" as "a" may contain some new variables.
-TODO
+TODO: the point is that those new variables are ignored when evaluating the message, since the message does not depend on them. This is likely to confuse many readers:(
 
@@
 it suffices to fix the polarities simultaneously in one bag....
+
+REWRITE and present a better optimization strategy due to the base pairing constraints (compatibility)
 

### 2018-02-15 RNARedPrint-Bioinfo2018.tex
... and more

--- 2018-02-14 RNARedPrint-Bioinfo2018.tex
@@
 %
-Recently, the \Software{RNAdesign} approach~\cite{HoenerzuSiederdissen2013}, coupled with powerful local search within \Software{RNAblueprint}~\cite{Hammer2017}, solved the problem of sampling seeds from the uniform distribution for multiple targets.
-These methods adopt a graph coloring perspective, initially decomposing the graph hierarchically using various decomposition algorithms, and \Def{precomputing} the number of valid sequences within each subgraph.
-The decomposition is then reinterpreted as a decision tree to perform a \Def{stochastic backtrack}, inspired by Ding and Lawrence~\cite{Ding2003}, where individual assignments are chosen with probabilities derived from the subsolution counts, leading to uniform sampling.
-The overall complexity of \Software{RNAdesign} grows like $\Theta(4^{\gamma})$, where $\gamma$ is a parameter whose value is bounded only by the length of the designed RNA.
+Recently, the \Software{RNAdesign}
+approach~\cite{HoenerzuSiederdissen2013}, coupled with powerful local
+search within \Software{RNAblueprint}~\cite{Hammer2017}, solved the
+problem of sampling seeds from the uniform distribution for multiple
+targets. These methods adopt a graph coloring perspective, initially
+decomposing the graph hierarchically using various decomposition
+algorithms, and \Def{precomputing} the number of valid sequences
+within each subgraph. The decomposition is then reinterpreted as a
+decision tree to perform a \Def{stochastic backtrack}, inspired by
+Ding and Lawrence~\cite{Ding2003}. Uniform sampling is achieved by
+choosing individual assignments with probabilities derived from the
+subsolution counts. The overall complexity of \Software{RNAdesign}
+grows like $\Theta(4^{\gamma})$, where the parameter $\gamma$ is
+bounded by the length of the designed RNA; typically $\gamma$ is kept
+much lower by the decomposition.
 
@@
 The hardness of counting (and conjectured hardness of sampling) does not preclude, however, practically applicable algorithms for counting and sampling.
-In particular, we wish to extend the flexibility of multi-structure design, leading to the following questions: \emph{How to sample, in a principled way, from a Boltzmann distribution based on expressive energy models?
+In particular, we wish to extend the flexibility of multi-structure design, leading to the following questions: \emph{How to sample, generically, from a Boltzmann distribution based on expressive energy models?
 How to enforce additional constraints, such as the \GCb-content, complex sequence constraints, or the energy of individual structures?}
@@
   alphabet} $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}$; let $\S_n$ denote the set of
-sequences of length $n$. An \Def{RNA (secondary) structure of length
-  $n$} is a set of \Def{base pairs} $(i,j)$, where $1\leq i<j\leq
-n$. We call an RNA structure $R$ \Def{non-crossing}, iff it does not
-contain any two different base pairs $(i,j)$ and $(i',j')$ such that
-$i\leq i'\leq j \leq j'.$ \Def{Valid base pair} must pair bases from
-the $\B:=\left\{\{\Ab,\Cb\},\{\Gb,\Cb\},\{\Gb,\Ub\}\right\}.$
-Consequently, $S$ is \Def{valid} for $R$, iff
-$(S_i,S_j)\in \B$ for all $(i,j)\in R$.
-%
-%\TODO{Also, we need to decide whether base-pairs are oriented or not ($(x,y)$ versus $\{x,y\}$).
-Right now, we are inconsistent in the definitions above.
-SW: i would prefer $(i,j)$ with $i<j$; } %% where is this inconsistent?
+sequences of length $n$. An \Def{RNA (secondary) structure $R$ of
+  length $n$} is a set of \Def{base pairs} $(i,j)$, where
+$1\leq i<j\leq n$, where for $(i,j), (i',j')\in R$: $i\neq j'$ and ($i=i'$ iff
+$j=j'$) (``degree $\leq$ 1''). We call an RNA structure $R$ \Def{non-crossing}, iff it does
+not contain any two different base pairs $(i,j)$ and $(i',j')$ such
+that $i\leq i'\leq j \leq j'.$ \Def{Valid base pair} must pair bases
+from the $\B:=\left\{\{\Ab,\Ub\},\{\Gb,\Cb\},\{\Gb,\Ub\}\right\}.$
+Consequently, $S$ is \Def{valid} for $R$, iff $\{S_i,S_j\}\in \B$ for
+all $(i,j)\in R$.
+%
+% \TODO{Also, we need to decide whether base-pairs are oriented or not
+% ($(x,y)$ versus $\{x,y\}$). Right now, we are inconsistent in the
+% definitions above. SW: i would prefer $(i,j)$ with $i<j$;
+% } %% where is this inconsistent?
 
@@
 $\R:=\{R_1, \dots, R_k\}$ for sequences of length $n$. $\R$ induces a
-\Def{compatibility graph}
-$G_{\R}$ with nodes $\{1,\dots,n\}$ and edges $\bigcup_{\ell\in[1,k]} R_\ell$. Valid
-sequences for $\R$ can then be seen as specific colorings of $G_{\R}$,
-using the four colors in $\Sigma$, such that the colors assigned to
-adjacent vertices of $G_{\R}$ constitute valid base pairs.
+\Def{compatibility graph} $G_{\R}$ with nodes $\{1,\dots,n\}$ and
+edges $S\bigcup_{\ell\in[1,k]} R_\ell$. While we typically consider
+additional dependencies (in the dependency graph), the edges in the
+compatibility graph pose constraints on the possible bases due to
+canonical base pairing, which are present in all considered settings.
+
+Valid sequences for $\R$ can then be seen as specific colorings of
+$G_{\R}$, using the four colors in $\Sigma$, such that the colors
+assigned to adjacent vertices of $G_{\R}$ constitute valid base pairs.
 %
@@
 \begin{align}
- \forall v_1\in V_1, f(v_1)&= \begin{cases} \Ab & \text{if }v_1\in S\\ \Gb & \text{if }v_1\notin S\end{cases}& 
- \forall v_2\in V_2, f(v_2)&= \begin{cases} \Cb & \text{if }v_2\in S\\ \Ub & \text{if }v_2\notin S\end{cases}
+ &\forall v_1\in V_1,v_2\in V_2:\notag\\& f(v_1) = \begin{cases} \Ab & \text{if }v_1\in S\\ \Gb & \text{if }v_1\notin S\end{cases} \text{\quad and\quad} 
+ f(v_2) = \begin{cases} \Cb & \text{if }v_2\in S\\ \Ub & \text{if }v_2\notin S\end{cases}
 \end{align}
@@
 \begin{itemize}
-\item Generate a random sequence of length 100;
-\item Compute its minimum free energy structure (ViennaRNA package);
-\item Add the new structure to the instances if the resulting compatibility graph is bipartite;
-\item Repeat until $k$ structures are collected.
+\item generate a random sequence of length 100
+\item compute its minimum free energy structure (ViennaRNA package)
+\item add the new structure to the instances if the resulting compatibility graph is bipartite
+\item repeat until $k$ structures are collected
 \end{itemize}

### 2018-02-15 RNARedPrint-Bioinfo2018.tex

--- 2018-02-15 RNARedPrint-Bioinfo2018.tex
@@
 %
-Recently, the \Software{RNAdesign} approach~\citep{HoenerzuSiederdissen2013}, coupled with powerful local search within \Software{RNAblueprint}~\citep{Hammer2017}, solved the problem of sampling seeds from the uniform distribution for multiple targets.
-These methods adopt a graph coloring perspective, initially decomposing the graph hierarchically using various decomposition algorithms, and \Def{precomputing} the number of valid sequences within each subgraph.
-The decomposition is then reinterpreted as a decision tree to perform a \Def{stochastic backtrack}, inspired by Ding and Lawrence~\citep{Ding2003}, where individual assignments are chosen with probabilities derived from the subsolution counts, leading to uniform sampling.
-The overall complexity of \Software{RNAdesign} grows like $\Theta(4^{\gamma})$, where $\gamma$ is a parameter whose value is bounded only by the length of the designed RNA.
+Recently, the \Software{RNAdesign}
+approach~\citep{HoenerzuSiederdissen2013}, coupled with powerful local
+search within \Software{RNAblueprint}~\citep{Hammer2017}, solved the
+problem of sampling seeds from the uniform distribution for multiple
+targets. These methods adopt a graph coloring perspective, initially
+decomposing the graph hierarchically using various decomposition
+algorithms, and \Def{precomputing} the number of valid sequences
+within each subgraph. The decomposition is then reinterpreted as a
+decision tree to perform a \Def{stochastic backtrack}, inspired by
+Ding and Lawrence~\citep{Ding2003}. Uniform sampling is achieved by
+choosing individual assignments with probabilities derived from the
+subsolution counts. The overall complexity of \Software{RNAdesign}
+grows like $\Theta(4^{\gamma})$, where the parameter $\gamma$ is
+bounded by the length of the designed RNA; typically $\gamma$ is kept
+much lower by the decomposition.
 
@@
 \begin{align}
- \forall v_1\in V_1, f(v_1)&= \begin{cases} \Ab & \text{if }v_1\in S\\ \Gb & \text{if }v_1\notin S\end{cases}\\
- \forall v_2\in V_2, f(v_2)&= \begin{cases} \Cb & \text{if }v_2\in S\\ \Ub & \text{if }v_2\notin S\end{cases}
+ &\forall v_1\in V_1,v_2\in V_2:\notag\\& f(v_1) = \begin{cases} \Ab & \text{if }v_1\in S\\ \Gb & \text{if }v_1\notin S\end{cases} \text{\quad and\quad} 
+ f(v_2) = \begin{cases} \Cb & \text{if }v_2\in S\\ \Ub & \text{if }v_2\notin S\end{cases}
 \end{align}

### 2018-02-15 RNARedPrint-Bioinfo2018.tex
cleaned up biblio + acknowledgements

@@
 (paths, cycles\ldots) this property rigorously follows from analytic
-combinatorics, see Bender-Richmond-Williamson~\citep{Bender1983} and
-Drmota~\citep{Drmota1997}. In such cases, the expected number of
+combinatorics, see \citet{Bender1983} and
+\citet{Drmota1997}. In such cases, the expected number of
 rejections before reaching the targeted energies remains constant when
@@
 To illustrate our above strategy, we sampled $n=1\,000$ sequences
-targeting ${\sf GC}_{target}=0.5$ and different Turner energies for
+targeting ${\sf GC}\%=0.5$ and different Turner energies for
 the three structure targets of an example
@@
 %jointly supported 
-by
-% the French 
+by the
+% French 
 {\em Agence
@@
 % acknowledged by
-%financial support 
-of the Future and Emerging Technologies programme
+%financial support of the 
+Future and Emerging Technologies programme
 % within the Seventh Framework Programme for Research of the
@@
 %
-We thank Leonid Chindelevitch for pointing out the potential for
-complexity reductions due to the bipartiteness of the compatibility
-graph; moreover, Arie Koster for discussing the practical computation
-of the tree decomposition.
+We thank Leonid Chindelevitch for suggesting using the stable sets analogy to optimize our FPT algorithm, and Arie Koster for practical recommendations 
+for computing tree decompositions.
 

### 2018-02-15 RNARedPrint-Bioinfo2018.tex
Added PKs cosmetically

@@
 \\
-\textbf{Results:} Here, we revisit a central ingredient of most in-silico design methods: the sampling of sequences for multi-target design.
+\textbf{Results:} Here, we revisit a central ingredient of most in-silico design methods: the sampling of sequences for the design of multi-target structures, possibly including pseudoknots.
 We establish the $\#${\sf P}-hardness of uniform sampling, and introduce \ourprog{}, a tree decomposition-based algorithm for efficient, fixed parameter tractable sampling.
-By modeling the problem as a constraint network, \ourprog{} supports generic Boltzmann-weighted sampling for arbitrary additive RNA energy models; this enables generating designs meeting specific goals like expected free energies or \GCb-content.
+By modeling the problem as a constraint network, \ourprog{} supports generic Boltzmann-weighted sampling for arbitrary additive RNA energy models; this enables the generation of RNA sequences meeting specific goals like expected free energies or \GCb-content.
 Finally, we empirically study general properties of the approach and generate biologically relevant multi-target Boltzmann-weighted designs for a common design benchmark.
-In particular, we show significant improvements over seed sequences generated by uniform sampling, the previously best available sampling strategy for multi-target design.\\
+Our results show significant improvements over seed sequences generated by uniform sampling, the previously best available sampling strategy for multi-target design.\\
 \textbf{Availability:} Our software is freely available at: \url{https://github.com/yannponty/RNARedPrint}\\

### 2018-02-15 RNARedPrint-Bioinfo2018.tex
Reactivate figures; address question about w_\ell

--- 2018-02-15 RNARedPrint-Bioinfo2018.tex
@@
 % \end{equation}
-In simple \Def{base pair-based energy models}, energy is defined as the sum of base pair (pseudo)energies. If base pair energies $\Ebp{k}{i,j,x,y}$ (where
-$i$ and $j$ are sequence positions, $x$ and $y$ are bases in $\Sigma$)
-are given for each target structure $\ell$, s.t.
-$ E(S;R_\ell) := \sum_{(i,j)\in R_\ell} \Ebp{k}{i,j,S_i,S_j}$, we
-encode the network energy by the set of functions
-$f$ for each base pair $(i,j)\in R_\ell$ of each input structure
-$R_\ell$ that evaluate to $w_\ell \Ebp{k}{i,j,\val(x_i),\val(x_j)}$
-under assignment $\val$.
+In simple \Def{base pair-based energy models}, energy is defined as
+the sum of base pair (pseudo-)energies. If base pair energies
+$\Ebp{k}{i,j,x,y}$ (where $i$ and $j$ are sequence positions, $x$ and
+$y$ are bases in $\Sigma$) are given for each target structure $\ell$,
+s.t.  $ E(S;R_\ell) := \sum_{(i,j)\in R_\ell} \Ebp{k}{i,j,S_i,S_j}$,
+we encode the network energy by the set of functions $f$ for each base
+pair $(i,j)\in R_\ell$ of each input structure $R_\ell$ that evaluate
+to $\ln(\pi_\ell) \Ebp{k}{i,j,\val(x_i),\val(x_j)}$ under assignment $\val$;
+here, $\pi_\ell>0$ is a weight that controls the
+influence of structure $\ell$ on the sampling (as elaborated later).
 
@@
   \centering
-  A)%\includegraphicstop[width=0.4\textwidth,trim=0 0 0 50,clip]{Figs/basepaircor}
-  B)%\includegraphicstop[width=0.4\textwidth,trim=0 0 0 50,clip]{Figs/stackingcor}
+  A)\includegraphicstop[width=0.4\textwidth,trim=0 0 0 50,clip]{Figs/basepaircor}
+  B)\includegraphicstop[width=0.4\textwidth,trim=0 0 0 50,clip]{Figs/stackingcor}
   \caption{Validation of the trained parameters for the A) base pair

### 2018-02-15 response_letter.txt
Todos

@@
+########################################
+MAJOR open issues:
+
+* Emphasize: we can do pseudoknots (like Blueprint)!
+  already in abstract by now
+
+* improve optimization by connected components of the compatibility graph
+
+* simplify
+   -- (sequence) positions/nucleotides instead of variables/values
+   -- (partial) sequences instead of (partial) assignments
+   -- evaluation of functions under assignments; maybe provide example---rethink if
+      the abstraction is necessary (especially if separator to parent is always size 1)
+   -- potentially simplify since separator to parent is always size 1
+
+* TODO discuss (for the question of artificial vs. natural / reviewer 3):
+  -- Adenin-Riboswitch (Schwalbe). Details aus Stefans Diss: 3-state/4-state riboswitch.
+     Experimentally verified designs were done only for 2-state.
+
+* RESULTS
+    -- three additional pseudoknotsets (for Table 1). 
+    -- linear correlation plot for Turner-energies fitted to simple energies (stacking?) 
+         (additionally with PKs!?)
+    -- shifting of targets (Figure 2) with PKs
+    -- time benchmark; run time for 1000 instances
+
+########################################
+
 ----------------------- REVIEW 1 ---------------------
@@
 
-REWRITE and present a better optimization strategy due to the base pairing constraints (compatibility)
+TODO: rewrite and present a better optimization strategy due to the base pairing constraints (compatibility)
 
@@
 is w_l? Is it possible to give the formula?
->>>>>> CHECK
+DONE
 
@@
 
->>>>> NO; if possible, reformulate the resepctive sentence
-CHECK
+>>>>> NO, but it is covered by the framework; if possible, reformulate the
+resepctive sentence CHECK
 

### 2018-02-19 RNARedPrint-Bioinfo2018.tex
Replace term 'compatibility graph' by 'base pair dependency graph'

@@
 \caption{General outline of \ourprog{} for base pair-based energy models.
-From a set of target secondary structures (i), base-pairs are merged (ii) into a compatibility graph (iii) and transformed into a tree decomposition (iv).
+From a set of target secondary structures (i), base-pairs are merged (ii) into a (base pair) dependency graph (iii) and transformed into a tree decomposition (iv).
 The tree is then used to compute the partition function, followed by a Boltzmann sampling of valid sequences (v).
@@
 $\R:=\{R_1, \dots, R_k\}$ for sequences of length $n$. $\R$ induces a
-\Def{compatibility graph} $G_{\R}$ with nodes $\{1,\dots,n\}$ and
-edges $S\bigcup_{\ell\in[1,k]} R_\ell$. While we typically consider
-additional dependencies (in the dependency graph), the edges in the
-compatibility graph pose constraints on the possible bases due to
-canonical base pairing, which are present in all considered settings.
+\Def{base pair dependency graph} $G_{\R}$ with nodes $\{1,\dots,n\}$
+and edges $S\bigcup_{\ell\in[1,k]} R_\ell$, which describe the minimal
+dependencies present in all relevant settings due to the requirement
+of canonical base pairing.
 
@@
 %We define the following notations: $\real^\infty:=\real\cup\{\infty\}$, $[i,j]:=\{i,\ldots,j\},$ \ldots
-Now we turn to the complexity of $\NumDesign(G)$, the problem of computing the number of valid sequences for a given compatibility graph $G=(V,E)$.
+Now we turn to the complexity of $\NumDesign(G)$, the problem of computing the number of valid sequences for a given base pair dependency graph $G=(V,E)$.
 Note that this problem corresponds to the partition function problem in a simple $(0/\infty)$-valued base pair model, with $\beta\neq 1$.
-As previously noted~\citep{Flamm2001}, a set of target structures admits a valid design iff its compatibility graph is bipartite, which can be tested in linear time.
+As previously noted~\citep{Flamm2001}, a set of target structures admits a valid design iff its base pair dependency graph is bipartite, which can be tested in linear time.
 Moreover, without $\{\Gb, \Ub\}$ base pairs, any connected component  $C\in {\rm CC}(G)$ is entirely determined by the assignment of a single of its nucleotides.
@@
   	  \For{$(v\to{}w)\in T$}{
-  	      $a_{\rm res} \gets a_{\rm res} \mid \Sample(v,\substitute{\val}{\val'};T,\chi,\phi,m)$\; 
+  	      $a_{\rm res} \gets \substitute{a_{\rm res}}{\Sample(v,\substitute{\val}{\val'};T,\chi,\phi,m)}$\; 
      }
@@
 requirement of canonical base pairing induces a bi-partition of each connected
-component of the compatibility graph, such that the variables of one
+component of the base pair dependency graph, such that the variables of one
 partition are restricted to values $\{\Ab,\Gb\}$, but for other
@@
   For sequence length $n$, $k$ target structures, treewidth $w$ and
-  a compatibility graph having $c$ connected components, $t$ sequences
+  a base pair dependency graph having $c$ connected components, $t$ sequences
   are sampled from the Boltzmann distribution  in
@@
 %Indeed, as shown in Fig.~\ref{fig:shifting_mean}, empirical joint distributions in the energies have good fits towards Normal multidimensional distributions with linear (co-)variances. 
-For base pair-based energy models and special compatibility graph
+For base pair-based energy models and special base pair dependency graph
 (paths, cycles\ldots) this property rigorously follows from analytic
@@
 Moreover, for each generated seed sequence, we optimized the objective function $(\ref{eq:blueprintobjective})$ through an adaptive  greedy walk consisting of 500 steps.
-At each step, we resampled (uniformly at random) the positions of a randomly selected component in the compatibility graph, and accepted the modification only if it resulted in a gain, as described in \Software{RNAblueprint}~\citep{Hammer2017}.
+At each step, we resampled (uniformly at random) the positions of a randomly selected component in the base pair dependency graph, and accepted the modification only if it resulted in a gain, as described in \Software{RNAblueprint}~\citep{Hammer2017}.
 %
@@
 \item Compute its minimum free energy structure (ViennaRNA package);
-\item Add the new structure to the instances if the resulting compatibility graph is bipartite;
+\item Add the new structure to the instances if the resulting base pair dependency graph is bipartite;
 \item Repeat until $k$ structures are collected.

### 2018-02-19 RNARedPrint-Bioinfo2018.tex
Use problem-specific terminology in place of assignments and variables etc

@@
 \newcommand{\network}{\mathcal{N}}
-\newcommand{\dom}{\operatorname{dom}} %domain
-\newcommand{\val}{a} % valuation aka assignment
-\newcommand{\vars}{\operatorname{vars}}
+\newcommand{\val}{\bar S} % valuation aka assignment
 \newcommand{\dep}{\operatorname{dep}}
@@
 
-%\newcommand{\assignments}{\operatorname{as}}
-\newcommand{\assignments}{\mathcal{A}}
+\newcommand{\partseqs}{\mathcal{P\!S}}
 \newcommand{\B}{\mathcal{B}}
@@
 \newcommand{\evalfor}[2]{#1\llbracket{}#2\rrbracket{}}
-\newcommand{\substitute}[2]{#1\!\mid\!\mid\! #2}
+\newcommand{\substitute}[2]{#1\!\oplus\!#2}
 
@@
 Ding and Lawrence~\citep{Ding2003}. Uniform sampling is achieved by
-choosing individual assignments with probabilities derived from the
-subsolution counts. The overall complexity of \Software{RNAdesign}
-grows like $\Theta(4^{\gamma})$, where the parameter $\gamma$ is
-bounded by the length of the designed RNA; typically $\gamma$ is kept
-much lower by the decomposition.
+choosing individual nucleotide assignments with probabilities derived
+from the subsolution counts. The overall complexity of
+\Software{RNAdesign} grows like $\Theta(4^{\gamma})$, where the
+parameter $\gamma$ is bounded by the length of the designed RNA;
+typically $\gamma$ is kept much lower by the decomposition.
 
@@
 %Uniform sampling is handled as a special case of Boltzmann sampling, where each valid sequence receives energy zero and---consequently---computing partition functions specializes to counting.
-By modeling sequences as variable assignments in a weighted constraint
+By evaluating (partial) sequences in a weighted constraint
 network, we support arbitrary multi-ary constraints and thus
@@
 %\item Model RNA sequences as assignments over a constraint network.
-The network consists of one variable per sequence position; constraints and energy contributions are modeled as functions of arbitrary arity over the variables, such that the total energy is the sum of all function values;
+The network consists of one position per sequence position; constraints and energy contributions are modeled as functions of arbitrary arity over the positions, such that the total energy is the sum of all function values;
 %\item Translate the dependency hyper-graph into a binary dependency graph by inserting cliques (Appendix), such that a tree decomposition with small treewidth can be computed by a generic tree decomposition algorithm;
@@
   length $n$} is a set of \Def{base pairs} $(i,j)$, where
-$1\leq i<j\leq n$, where for $(i,j), (i',j')\in R$: $i\neq j'$ and ($i=i'$ iff
-$j=j'$) (``degree $\leq$ 1''). We call an RNA structure $R$ \Def{non-crossing}, iff it does
-not contain any two different base pairs $(i,j)$ and $(i',j')$ such
-that $i\leq i'\leq j \leq j'.$ \Def{Valid base pair} must pair bases
-from the $\B:=\left\{\{\Ab,\Ub\},\{\Gb,\Cb\},\{\Gb,\Ub\}\right\}.$
+$1\leq i<j\leq n$, where for all different $(i,j), (i',j')\in R$:
+$\{i,j\}\cap\{i',j'\}=\emptyset$ (``degree $\leq$ 1''). We call an RNA
+structure $R$ \Def{non-crossing}, iff it does not contain any two
+different base pairs $(i,j)$ and $(i',j')$ such that
+$i\leq i'\leq j \leq j'.$ \Def{Valid base pair} must pair bases from
+the $\B:=\left\{\{\Ab,\Ub\},\{\Gb,\Cb\},\{\Gb,\Ub\}\right\}.$
 Consequently, $S$ is \Def{valid} for $R$, iff $\{S_i,S_j\}\in \B$ for
 all $(i,j)\in R$.
-%
-% \TODO{Also, we need to decide whether base-pairs are oriented or not
-% ($(x,y)$ versus $\{x,y\}$). Right now, we are inconsistent in the
-% definitions above. SW: i would prefer $(i,j)$ with $i<j$;
-% } %% where is this inconsistent?
 
@@
 
-The \Def{dependency graph} of the set $\F$ of functions (on
-sequences of length $n$) is the \Def{hypergraph}
+The \Def{dependency graph} induced by the set $\F$ of functions (on
+sequences of length $n$) is the hypergraph
 $G_\F=(\{1,\dots,n\} ,\{\dep(f)\mid f\in \F\})$. 
@@
 %
-In the RNA design network, sequences $S\in\S_n$ are isomorphically
-modeled as assignments $\val_S$ of bases to $n$ variables, each associated to
-a sequence position. The functions $f\in\F$ of
-Section~\ref{sec:problem-statement}, are interpreted as functions on the network's variables.
-
-\begin{definition}[RNA design network]
-An \Def{RNA design network} is a tuple $\network=(\X,\F)$ such that:
+In the RNA design network, (partially determined) RNA sequences
+replace the more general concept of (partial) assignments in
+constraint networks. Partially determined RNA sequences, called
+\Def{partial sequences}, are words $\val$ over the alphabet
+$\Sigma\cup\{?\}$ equivalently representing the set ${\mathcal S}(\val)$
+of RNA sequences, where for positions $1\leq i\leq n$,
+$\val_i\in\Sigma$ implies $S_i=\val_i$ for all $S\in{\mathcal
+  S}(\val)$. The positions $1\leq i\leq n$, where $\val_i\in\Sigma$,
+are called \Def{determined} by $\val$ and form its \Def{domain}.
+%
+Since the functions $f\in\F$ of Section~\ref{sec:problem-statement}
+depend on only the subset $\dep(f)$ of sequence positions, one can
+evaluate them for partial sequences $\val$ that determine (at least)
+the nucleotides at all positions in $\dep(f)$. Thus for functions $f$
+and partial sequences $\val$ that determine $\dep(f)$, we write
+$\evalfor{f}{\val}$ to \Def{evaluate $f$ for $\val$}; i.e.{}
+$\evalfor{f}{\val} := f(S),$ for any sequence
+$S\in{\mathcal S}(\val)$.
+
+\begin{definition}
+An \Def{RNA design network} is a tuple $\network=(\X,\F)$, where
 \begin{itemize}
-\item $\X$ is a set of \Def{variables} $\{x_1,\dots,x_n\}$ associated with the sequence positions
-\item $\F$ is a set of \Def{functions} $f:\S_n\to\real$, each function $f\in \F$ being associated with the set $\vars(f)=\{x_i\mid i\in\dep(f) \}\subset X$ of variables.
+\item[--]\ $\X$ is the set of sequence positions $1,\dots,n$
+\item[--]\ $\F$ is a set of \Def{functions} $f:\S_n\to\real$
 \end{itemize}
@@
 
-A \Def{(partial) assignment} is a function $\val: \X' \to \Sigma$
-that maps the variables $\X'\subseteq \X$ to bases; we call $\X'$ the \Def{domain of the assignment
-  $\val$}, written $\dom(\val)$; $\val$ is \Def{total} iff $\dom(\val)=\X.$
-%
-For assignments $\val$ and functions $f$, $\dep(f)\subseteq\dom(\val)$,
-we define the special notation $\evalfor{f}{\val}$ to \Def{evaluate
-  $f$ for $\val$}; i.e.{} $\evalfor{f}{\val} := f(S),$ for any
-sequence $S$ where $S_i=\val(i)$ for all $i\in\dep(f).$ Then, the
-\Def{energy $\energy{\network}(S)$ of a sequence} $S$ in a network
+%
+The \Def{energy $\energy{\network}(S)$ of a sequence} $S$ in a network
 $\network$ is defined as sum of the values of all functions in
@@
 
-Let us now introduce some definitions. A \Def{cluster tree} for the
+We require additional definitions: A \Def{cluster tree} for the
 network $\network=(\X,\F)$ is a tuple $(T,\chi,\phi)$, where
@@
 a set of functions $f$, each uniquely assigned to a node $v\in T$;
-$\vars(f)\subseteq\chi(v)$ and $\phi(v)\cap \phi(v')=\varnothing$ for
+$\dep(f)\subseteq\chi(v)$ and $\phi(v)\cap \phi(v')=\varnothing$ for
 all $v\neq v'$.  For two nodes $v$ and $u$ of the cluster tree, define
 their \Def{separator} as $\separator{u}{v} := \chi(u)\cap\chi(v)$;
-moreover, we define the \Def{difference variables} from $u$ to an
+moreover, we define the \Def{difference positions} from $u$ to an
 adjacent $v$ by $\difference{u}{v}:=\chi(v) - \separator{u}{v}$.
 
-For a set $Y$ of variables, denote the set of all assignments of the
-variables in $\Y$ by $\assignments(\Y)$; moreover, given
-non-overlapping assignments $\val'$ and $\val''$, we write
-$\substitute{\val'}{\val''}$ to denote the assignment defined as
-$\val'$ and $\val''$ respectively on $\dom(\val')$ and $\dom(\val'')$.
-Finally we assume, w.l.o.g., that all variable difference sets
-$\difference{u}{v}$ are restricted to a single variable: for any given
+For a set $\Y$ of sequence positions, write $\partseqs(\Y)$ to
+denote the set of all partial sequences that determine exactly the positions
+in $\Y$; moreover, given partial sequences $\val$ and
+$\val'$, we define the \Def{combined partial sequence $\substitute{\val'}{\val''}$} such that
+$$
+(\substitute{\val}{\val'})_i :=
+\begin{cases}
+  \val'_i & \text{if } \val'_i\in \Sigma\\
+  \val_i & \text{otherwise}
+\end{cases}
+$$
+
+Finally we assume, w.l.o.g., that all position difference sets
+$\difference{u}{v}$ are singleton: for any given
 cluster tree, an equivalent (in term of treewidth) cluster tree can
@@
 tree decomposition $(T,\chi,\phi)$.  W.l.o.g, we assume that $T$ is
-connected and contains a dedicated node $r$, with $\chi(r)=\varnothing$
-and $\phi(r)=\varnothing$, added as a virtual root
-$r$ connected to a node in each connected
-component of $T$.  Now, we consider the set of directed edges
-$\edgesToR{}$ of $T$ oriented to $r$; define $T_r(u)$ as the induced
-subtree of $u$. Algorithm~\ref{alg:pf} computes the partition function by
-passing messages along these directed edges $u\to v$ (i.e. always from
-some child $u$ to its parent $v$). Each message is a function on
-variables $\vars(m)\subseteq \X$ to $\real\cup\{\infty\}$. The message
-from $u$ to $v$ represents the partition functions of the subtree of
-$u$ for all possible assignments to the variables
-$\vars(m)=\separator{u}{v}$. The correctness of the algorithm can be shown by induction over $T$ (Supp. Mat.~\ref{appsec:correctness}).
-After running Alg.~\ref{alg:pf}, 
-multiplying the 0-ary messages sent to the root $r$ yields the total partition function:
+connected and contains a dedicated node $r$, with
+$\chi(r)=\varnothing$ and $\phi(r)=\varnothing$, added as a virtual
+root $r$ connected to a node in each connected component of $T$.  Now,
+we consider the set of directed edges $\edgesToR{}$ of $T$ oriented to
+$r$; define $T_r(u)$ as the induced subtree of
+$u$. Algorithm~\ref{alg:pf} computes the partition function by passing
+messages along these directed edges $u\to v$ (i.e. always from some
+child $u$ to its parent $v$). Each message is a function that depends
+on the positions $\dep(m)\subseteq \X$ and yields a partition function
+in $\real\cup\{\infty\}$. The message from $u$ to $v$ represents the
+partition functions of the subtree of $u$ for all possible partial
+sequences in $\partseqs(\separator{u}{v})$. Induction over $T$ lets us show
+the correctness of the algorithm
+(Supp. Mat.~\ref{appsec:correctness}).  After running
+Alg.~\ref{alg:pf}, multiplying the 0-ary messages sent to the root $r$
+yields the total partition function:
 \begin{math}
@@
 \begin{algorithm}[t]
- \KwData{Cluster tree $(T,\chi,\phi)$}
- \KwResult{Messages $\Message{u}{v}$ for all $(u\to{}v)\in T$; i.e.~partition functions of the subtrees of all $v$ for all possible assignments of variables $\separator{u}{v}$.}
+  \KwData{Cluster tree $(T,\chi,\phi)$} \KwResult{Messages
+    $\Message{u}{v}$ for all $(u\to{}v)\in T$; i.e.~partition
+    functions of the subtrees of all $v$ for all possible partial
+    sequences determining exactly the positions $\separator{u}{v}$.}
  \For{$u\to{}v\in T$ in postorder}{
-  \For{$\val\in\assignments(\separator{u}{v})$}{
+  \For{$\val\in\partseqs(\separator{u}{v})$}{
     $\evalfor{\Message{u}{v}}{\val} \gets 0$\;
-    \For{$\val'\in\assignments(\difference{u}{v})$}{
+    \For{$\val'\in\partseqs(\difference{u}{v})$}{
      $p \gets$ product( $exp(-\beta \evalfor{f}{\substitute{\val}{\val'}})$ for $f\in \phi(u)$ )\;
@@
 %
-Algorithm~\ref{alg:sampling} defines the recursive procedure $\sample(u,\val)$, which returns an  assignment of all variables in the subtree rooted at $u$ from the Boltzmann-distribution.
-Called on $r$ and the empty assignment, the procedure returns a total assignment corresponding to a Boltzmann-distributed random sequence.
+Algorithm~\ref{alg:sampling} defines the recursive procedure
+$\sample(u,\val)$, which returns a partial sequence that determines
+the positions of the subtree rooted at $u$ from the
+Boltzmann-distribution.  Called on $r$ and the empty partial sequence,
+the procedure returns a Boltzmann-distributed random sequence.
 
@@
 %  \begin{enumerate}[1)]
-%  \item Let $\Delta:=\chi(u)-\dom(\val)$ be the set of free variables for $u$.
+%  \item Let $\Delta:=\chi(u)-\dom(\val)$ be the set of free positions for $u$.
 %  \item For all assignments $\val'\in\assignments(\Delta)$, compute
@@
 %The complexity of the proposed sampling algorithm depends on the treewidth of the dependency graph $G_\F$.
-% The treewidth $w$ of a cluster tree $(T,\chi,\phi)$, $T=(V,E)$, is $\max_{v\in V} |\chi(v)| - 1$, i.e.~the maximum number of variables in any of its clusters minus 1.
+% The treewidth $w$ of a cluster tree $(T,\chi,\phi)$, $T=(V,E)$, is $\max_{v\in V} |\chi(v)| - 1$, i.e.~the maximum number of positions in any of its clusters minus 1.
 
@@
 \begin{algorithm}
- \KwData{Node $u$, partial assignment $\val\in\assignments(\separator{u}{v})$;\newline
- Cluster tree $(T,\chi,\phi)$ and partition functions $\Message{u'}{v'}[\val']$, $\forall (u'\to{}v')\in T$ and $\val'\in\assignments(\separator{u'}{v'})$.}
- \KwResult{Boltzmann-distributed random assignment for the subtree rooted at $u$, assuming a partial assignment $\val$.}
+ \KwData{Node $u$, partial sequence $\val\in\partseqs(\separator{u}{v})$;\newline
+ Cluster tree $(T,\chi,\phi)$ and partition functions $\Message{u'}{v'}[\val']$, $\forall (u'\to{}v')\in T$ and $\val'\in\partseqs(\separator{u'}{v'})$.}
+ \KwResult{Boltzmann-distributed random partial sequence for the subtree rooted at $u$, specializing a partial sequence $\val$.}
  \Fn{\Sample$(u,\val;T,\chi,\phi,m)$}{
    $r \gets \Random(\evalfor{\Message{u}{v}}{\val})$\;
-   \For{$\val'\in\assignments(\difference{u}{v})$}{
+   \For{$\val'\in\partseqs(\difference{u}{v})$}{
 
@@
  }
- \caption{Stochastic backtrack algorithm for assignments in the Boltzmann distribution.}\label{alg:sampling}
+ \caption{Stochastic backtrack algorithm for partial sequences in the Boltzmann distribution.}\label{alg:sampling}
 \end{algorithm}
@@
 requirement of canonical base pairing induces a bi-partition of each connected
-component of the base pair dependency graph, such that the variables of one
-partition are restricted to values $\{\Ab,\Gb\}$, but for other
-partition, to values $\{\Cb,\Ub\}$. This suggests a strong
+component of the base pair dependency graph, such that the positions of one
+partition are restricted to nucleotides $\{\Ab,\Gb\}$, but for other
+partition, to nucleotides $\{\Cb,\Ub\}$. This suggests a strong
 optimization of our algorithms, which reduces the basis in the
@@
 motifs, as well as features depending on the target structures.
-Naturally, hard
-constraints, which enforce or forbid features, can also be expressed by functions that
-assign infinite penalties to invalid partial assignments. More concretely, we briefly 
-discuss the modeling of RNA energy models in our
-framework.
+Naturally, hard constraints, which enforce or forbid features, can
+also be expressed by functions that assign infinite penalties to
+invalid (partial) sequences. More concretely, we briefly discuss the
+modeling of RNA energy models in our framework.
 
@@
 pair $(i,j)\in R_\ell$ of each input structure $R_\ell$ that evaluate
-to $\ln(\pi_\ell) \Ebp{k}{i,j,\val(x_i),\val(x_j)}$ under assignment $\val$;
+to $\ln(\pi_\ell) \Ebp{k}{i,j,\val(x_i),\val(x_j)}$ under partial sequence $\val$;
 here, $\pi_\ell>0$ is a weight that controls the
@@
 
-Fortunately, we can show that, as long as the network the arguments $\vars(f)$ of a function $f$ are materialized by a clique within the network $\network$, then there exists at least one node in the tree decomposition that contains $\vars(f)$, thus allowing the evaluation of $f$.
+Crucially, one can show that, as long as the network the arguments $\dep(f)$ of a function $f$ are materialized by a clique within the network $\network$, then there exists at least one node in the tree decomposition that contains $\dep(f)$, thus allowing the evaluation of $f$.
 \begin{lemma}\label{lem:cliques}
@@
   the messages $\Message{u}{v}$, for all edges $u\to{}v\in T$, yield
-  the partition functions of subtree of $u$ for the partial
-  assignments $\val\in\assignments(\separator{u}{v})$, i.e. the
-  messages satisfy
+  the partition functions of subtree of $u$ for the partial sequences
+  $\val\in\partseqs(\separator{u}{v})$, i.e. the messages satisfy
   \begin{equation}
-   \evalfor{\Message{u}{v}}{\val} = \sum_{\val'\in\assignments(\chi(T_r(u))-\separator{u}{v})} \quad
+   \evalfor{\Message{u}{v}}{\val} = \sum_{\val'\in\partseqs(\chi(T_r(u))-\separator{u}{v})} \quad
    \prod_{f\in\phi(T_r(u))} \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}}),\label{eq:pfalgo-correct}
  \end{equation}
- where $\chi(T_r(u))$ denotes all $\chi$-assigned variables of nodes in $T_r(u)$;
+ where $\chi(T_r(u))$ denotes all $\chi$-assigned positions of nodes in $T_r(u)$;
  respectively $\phi(T_r(u))$; all $\phi$-assigned functions.
@@
 \begin{equation}
-  \evalfor{\Message{u}{v}}{\val} := \sum_{\val'\in\assignments(\difference{u}{v})}\quad
+  \evalfor{\Message{u}{v}}{\val} := \sum_{\val'\in\partseqs(\difference{u}{v})}\quad
   \prod_{f\in \phi(u) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}}) \prod_{(w\to{}u) \in T} \evalfor{\Message{w}{u}}{\substitute{\val'}{\val}}.\label{eq:messages}
@@
 $w_1,\dots,w_q$ the messages $\Message{w_1}{u}, \dots, \Message{w_q}{u}$, which satisfy Eq.~$(\ref{eq:pfalgo-correct})$ (induction hypothesis).
-Let $\val\in\assignments(\separator{u}{v})$; then, $\evalfor{\Message{u}{v}}{\val}$ is computed by the algorithm according to Eq.~$(\ref{eq:messages})$.
+Let $\val\in\partseqs(\separator{u}{v})$; then, $\evalfor{\Message{u}{v}}{\val}$ is computed by the algorithm according to Eq.~$(\ref{eq:messages})$.
 We rewrite as follows
 \begin{align*}
-  & \sum_{\val'\in\assignments(\difference{u}{v})}\quad
+  & \sum_{\val'\in\partseqs(\difference{u}{v})}\quad
     \prod_{f\in \phi(u) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}})
     \prod_{(w\to{}u) \in T} \evalfor{\Message{w}{u}}{\substitute{\val'}{\val}}\\
-  & = \sum_{\val'\in\assignments(\chi(u)-\separator{u}{v})}\quad
+  & = \sum_{\val'\in\partseqs(\chi(u)-\separator{u}{v})}\quad
     \prod_{f\in \phi(u) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}})
@@
   & =_{IH}
-    \sum_{\val'\in\assignments(\chi(u)-\separator{u}{v})}\quad
+    \sum_{\val'\in\partseqs(\chi(u)-\separator{u}{v})}\quad
     \prod_{f\in \phi(u) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}})
-    \prod_{i=1}^q \sum_{\val''\in\assignments(\chi(T_r(w_i))-\separator{w_i}{u})} \quad
+    \prod_{i=1}^q \sum_{\val''\in\partseqs(\chi(T_r(w_i))-\separator{w_i}{u})} \quad
     \prod_{f\in\phi(T_r(w_i))} \exp(-\beta \evalfor{f}{\substitute{\val''}{\substitute{\val'}{\val}}}) \\
 & =_{*}
-    \sum_{\val'\in\assignments(\chi(u)-\separator{u}{v})}\quad
-  \sum_{\val''\in\assignments(\bigcup_{i=1}^q\chi(T_r(w_i))-\separator{w_i}{u})} \quad
+    \sum_{\val'\in\partseqs(\chi(u)-\separator{u}{v})}\quad
+  \sum_{\val''\in\partseqs(\bigcup_{i=1}^q\chi(T_r(w_i))-\separator{w_i}{u})} \quad
   \prod_{f\in \phi(u) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}})
   \prod_{f\in\phi(T_r(w_i))} \exp(-\beta \evalfor{f}{\substitute{\val''}{\val'} | \val}) \\
-  & =_{(**)} \sum_{\val'\in\assignments(\chi(T_r(u))-\separator{u}{v})}\quad
+  & =_{(**)} \sum_{\val'\in\partseqs(\chi(T_r(u))-\separator{u}{v})}\quad
     \prod_{f\in \phi(T_r(u)) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}})\\
@@
   Def.~\ref{def:treedecomp}, property 2. First, this property implies
-  that any shared variable between the subtrees of $w_i$ and $w_j$
+  that any shared position between the subtrees of $w_i$ and $w_j$
   must be in $\chi(w_i)$, $\chi(w_j)$ and $\chi(u)$, thus the
-  variables of $\chi(T_r(w_i))-\separator{w_i}{u}$ are
-  disjoint. Second, if a variable $\chi(T_r(w_i))$ occurs in
+  positions of $\chi(T_r(w_i))-\separator{w_i}{u}$ are
+  disjoint. Second, if a position $\chi(T_r(w_i))$ occurs in
   $\chi(u)$, it must occur in $\chi(w_i)$ and consequently in $\separator{u}{v}$.

### 2018-02-19 response_letter.txt
Update todos

@@
 * improve optimization by connected components of the compatibility graph
-
-* simplify
-   -- (sequence) positions/nucleotides instead of variables/values
-   -- (partial) sequences instead of (partial) assignments
-   -- evaluation of functions under assignments; maybe provide example---rethink if
-      the abstraction is necessary (especially if separator to parent is always size 1)
-   -- potentially simplify since separator to parent is always size 1
 

### 2018-02-19 RNARedPrint-Bioinfo2018.tex
More about pseudoknots; optimization run interpretation

@@
 $1\leq i<j\leq n$, where for all different $(i,j), (i',j')\in R$:
-$\{i,j\}\cap\{i',j'\}=\emptyset$ (``degree $\leq$ 1''). We call an RNA
-structure $R$ \Def{non-crossing}, iff it does not contain any two
-different base pairs $(i,j)$ and $(i',j')$ such that
-$i\leq i'\leq j \leq j'.$ \Def{Valid base pair} must pair bases from
-the $\B:=\left\{\{\Ab,\Ub\},\{\Gb,\Cb\},\{\Gb,\Ub\}\right\}.$
+$\{i,j\}\cap\{i',j'\}=\emptyset$ (``degree $\leq$ 1'').
+% 
+% We call an RNA structure $R$ \Def{non-crossing}, iff it does not
+% contain any two different base pairs $(i,j)$ and $(i',j')$ such that
+% $i\leq i'\leq j \leq j'.$
+% 
+\Def{Valid base pair} must pair bases from the
+$\B:=\left\{\{\Ab,\Ub\},\{\Gb,\Cb\},\{\Gb,\Ub\}\right\}.$
 Consequently, $S$ is \Def{valid} for $R$, iff $\{S_i,S_j\}\in \B$ for
@@
 %
-As will be explicit in subsequent sections, our approach relies on breaking down the energy function $E_\R(S)$ into additive components, each depending on only few sequence positions.
-Given $\R$, we express $E_\R(S)$
-as the sum of energy contributions $f(S)$ over a set $\F$ (of
-functions $f:\S_n\to\real$), s.t.~$E_\R(S)=\sum_{f\in\F} f(S)$. This
-captures realistic RNA energy models, while bounding the
-dependencies to sequence positions introduced by each single $f$.
-Formally, define the \Def{dependencies} $\dep(f)$ of $f$ as a minimal set of sequence positions $\I\subseteq\{1,\dots,n\}$, where $f(S)=f(S')$ for all sequences $S$ and $S'$ agreeing at all positions in $\I$.
+  As will be explicit in subsequent sections, our approach relies on
+  breaking down the energy function $E_\R(S)$ into additive
+  components, each depending on only few sequence positions. Given
+  $\R$, we express $E_\R(S)$ as the sum of energy contributions $f(S)$
+  over a set $\F$ (of functions $f:\S_n\to\real$),
+  s.t.~$E_\R(S)=\sum_{f\in\F} f(S)$. This captures realistic RNA
+  energy models---including nearest neighbor models, e.g.~\citet{Turner2009}, and even
+  pseudoknot models, e.g. \citet{Andronescu2010}, while bounding the dependencies to
+  sequence positions introduced by each single $f$. Formally, define
+  the \Def{dependencies} $\dep(f)$ of $f$ as a minimal set of sequence
+  positions $\I\subseteq\{1,\dots,n\}$, where $f(S)=f(S')$ for all
+  sequences $S$ and $S'$ agreeing at all positions in $\I$.
 
@@
 Given is the RNA design network $\network=(\X,\F)$ and its cluster
-tree decomposition $(T,\chi,\phi)$.  W.l.o.g, we assume that $T$ is
+tree decomposition $(T,\chi,\phi)$.  W.l.o.g., we assume that $T$ is
 connected and contains a dedicated node $r$, with
@@
 Algorithm~\ref{alg:sampling} defines the recursive procedure
-$\sample(u,\val)$, which returns a partial sequence that determines
-the positions of the subtree rooted at $u$ from the
-Boltzmann-distribution.  Called on $r$ and the empty partial sequence,
-the procedure returns a Boltzmann-distributed random sequence.
+$\sample(u,\val)$, which returns---randomly drawn from the
+Boltzmann distribution---a partial sequence that determines all
+sequence positions in the subtree rooted at $u$.  Called on $r$ and
+the empty partial sequence, which does not determine any positions,
+the procedure samples a random sequence from the Boltzmann
+distribution.
 
@@
 Moreover, for each generated seed sequence, we optimized the objective function $(\ref{eq:blueprintobjective})$ through an adaptive  greedy walk consisting of 500 steps.
-At each step, we resampled (uniformly at random) the positions of a randomly selected component in the base pair dependency graph, and accepted the modification only if it resulted in a gain, as described in \Software{RNAblueprint}~\citep{Hammer2017}.
-%
-Once again, for all instances, we observe a positive improvement of the quality of Boltzmann designs over uniform ones, suggesting a superior potential for optimization ($\sim 32\%$ avg improvement, Table~\ref{tab:benchmark-results}, optimized).
-This improvement is however more modest than in the absence of optimization, but we attribute this relative decay to the uniform nature of the resampling strategy, which somehow ends up canceling the benefits of the initial Boltzmann distribution, by uniformizing the sequence over longer runs.
-This suggests an alternative resampling strategy based on Boltzmann-distributed steps.
+At each step, we resampled (uniformly random) the positions of a randomly selected component in the base pair dependency graph, and accepted the modification only if it resulted in a gain, as described in \Software{RNAblueprint}~\citep{Hammer2017}.
+%
+Once again, for all instances, we observe a positive improvement of
+the quality of Boltzmann designs over uniform ones, suggesting a
+superior potential for optimization ($\sim 32\%$ avg improvement,
+Table~\ref{tab:benchmark-results}, optimized). Notably, our subsequent
+optimization runs, which are directly inherited from the uniformly
+sampling RNABlueprint software, partially level the advantages of
+Boltzmann sampling. In future work, we hope to improve this aspect by
+exploiting Boltzmann sampling even during the optimization run.
 

### 2018-02-21 RNARedPrint-Bioinfo2018.tex
Added complexity improvement Moved #P hardness result

--- 2018-02-19 RNARedPrint-Bioinfo2018.tex
@@
 \usepackage{xspace} 
+%\usepackage{geometry} 
 \usepackage{amsmath,amssymb} 
@@
 
+\usepackage{xifthen}
 \usepackage{stmaryrd}
@@
 \usepackage{xcolor}
+
+\hypersetup{
+    bookmarks=false,         % show bookmarks bar?
+    unicode=false,          % non-Latin characters in Acrobat’s bookmarks
+    pdftoolbar=true,        % show Acrobat’s toolbar?
+    pdfmenubar=true,        % show Acrobat’s menu?
+    pdffitwindow=true,     % window fit to page when opened
+    pdfstartview={FitH},    % fits the width of the page to the window
+    colorlinks=true,       % false: boxed links; true: colored links
+    linkcolor=red!70!black,          % color of internal links (change box color with linkbordercolor)
+    citecolor=blue!50!black,        % color of links to bibliography
+    filecolor=magenta,      % color of file links
+    urlcolor=green!50!black           % color of external links
+}
 
@@
 \newtheorem{hypothesis}{Working Hypothesis}
+\renewenvironment{proof}[1][]{\noindent \em Proof\ifthenelse{\equal{#1}{}}{}{ (#1)}:~}{}
 
@@
 \caption{General outline of \ourprog{} for base pair-based energy models.
-From a set of target secondary structures (i), base-pairs are merged (ii) into a (base pair) dependency graph (iii) and transformed into a tree decomposition (iv).
+From a set of target secondary structures (i), base-pairs are merged (ii) into a compatibility graph (iii) and transformed into a tree decomposition (iv).
 The tree is then used to compute the partition function, followed by a Boltzmann sampling of valid sequences (v).
@@
 $\R:=\{R_1, \dots, R_k\}$ for sequences of length $n$. $\R$ induces a
-\Def{base pair dependency graph} $G_{\R}$ with nodes $\{1,\dots,n\}$
-and edges $S\bigcup_{\ell\in[1,k]} R_\ell$, which describe the minimal
-dependencies present in all relevant settings due to the requirement
-of canonical base pairing.
+\Def{compatibility graph} $G_{\R}$ with nodes $\{1,\dots,n\}$ and
+edges $S\bigcup_{\ell\in[1,k]} R_\ell$. While we typically consider
+additional dependencies (in the dependency graph), the edges in the
+compatibility graph pose constraints on the possible bases due to
+canonical base pairing, which are present in all considered settings.
 
@@
 \end{definition}
-\section{Counting valid designs is \#{\sf P}-hard}\label{sec:counting}
-
-%%% reformulate later
-%We define the following notations: $\real^\infty:=\real\cup\{\infty\}$, $[i,j]:=\{i,\ldots,j\},$ \ldots
-Now we turn to the complexity of $\NumDesign(G)$, the problem of computing the number of valid sequences for a given base pair dependency graph $G=(V,E)$.
-Note that this problem corresponds to the partition function problem in a simple $(0/\infty)$-valued base pair model, with $\beta\neq 1$.
-As previously noted~\citep{Flamm2001}, a set of target structures admits a valid design iff its base pair dependency graph is bipartite, which can be tested in linear time.
-Moreover, without $\{\Gb, \Ub\}$ base pairs, any connected component  $C\in {\rm CC}(G)$ is entirely determined by the assignment of a single of its nucleotides.
-The number of valid designs is thus simply $4^{\#{\rm CC}(G)}$, where $\#{\rm CC}(G)$ is the \Def{number of connected components}.
-
-The introduction of $\{\Gb, \Ub\}$ base pairs radically changes this picture, and we show that valid designs for a set of structures cannot be counted in polynomial time, unless ${\sf\# P}={\sf FP}$.
-The latter equality would, in particular, imply the classic ${\sf P}={\sf NP}$, and solving $\NumDesign(G)$ is polynomial time is therefore probably difficult.
-
-To establish that claim, we consider instances $G=(V_1\cup V_2, E)$ that are connected and bipartite ($(V_1\times V_2) \cap E = \varnothing$), noting that hardness over restricted instances implies hardness of the general problem.
-Moreover remark that, as observed in Subsec.~\ref{sec:complexity}, assigning a nucleotide to a position $u\in V$ constrains the parity ($\{\Ab,\Gb\}$ or $\{\Cb,\Ub\}$) of all positions in the connected component of $u$.
-For this reason, we restrict our attention to the counting of valid designs \emph{up to trivial} $(\Ab\leftrightarrow \Cb/\Gb\leftrightarrow \Ub)$ symmetry, by constraining the positions in $V_{1}$ (resp.
-$V_2$) to only $\Ab$ and $\Gb$ (resp.
-$\Cb$ and $\Ub$).
-Let $\Design{G}$ denote the subset of all designs for $G$ under this constraint, noting that $\NumDesign(G) = 2\cdot|\Design{G}|$.
-
-We remind that an \Def{independent set} of $G=(V,E)$ is a subset $V'\subseteq V$ of nodes that are not connected by any edge in $E$. Let $\IS{G}$ denote the set of all independent sets in $G$.
-
-\begin{proposition}
- $\Design{G}$ is in bijection with $\IS{G}$.
-\end{proposition}
-\begin{proof}
-Consider the function $\Psi: \Design{G} \to \IS{G}$ defined by $ \Psi(f) := \left\{v\in V\mid f(v)\in\{\Ab,\Cb\}\right\}.$
-
-Let us establish the injectivity of  $\Psi$, i.e. that $\Psi(f)\neq\Psi(f')$ for all $f\neq f'$.
-If $f\neq f'$, then there exists a node $v\in V$ such that $f(v)\neq f'(v)$. 
-Assume that $v\in V_1$, and remind that the only nucleotides allowed in $V_1$ are $\Ab$ and $\Gb$. Since $f(v)\neq f'(v)$, then we have $\{f(v),f'(v)\}=\{\Ab,\Gb\}$
-and we conclude that $\Psi(f)$ differs from $\Psi(f')$ at least with respect to its inclusion/exclusion of $v$.
-
-We turn to the surjectivity of $\Psi$, i.e. the existence of a preimage for each element $S\in \IS{G}$. Let us consider the function $f$ defined as
-\begin{align}
- &\forall v_1\in V_1,v_2\in V_2:\notag\\& f(v_1) = \begin{cases} \Ab & \text{if }v_1\in S\\ \Gb & \text{if }v_1\notin S\end{cases} \text{\quad and\quad} 
- f(v_2) = \begin{cases} \Cb & \text{if }v_2\in S\\ \Ub & \text{if }v_2\notin S\end{cases}
-\end{align}
-One easily verifies that $\Psi(f) = S$.
-
-We are then left to determine if $f$ is a valid design for $G$, i.e.
-if for each $(v, v') \in E$ one has  $\{f(v),f(v')\}\in \B.$ Since $G$ is bipartite, any edge in $E$ involves two nodes $v_1\in V_1$ and $v_2\in V_2$.
-Remark that, among all the possible assignments $f(v_1)$ and $f(v_2)$, the only invalid combination of nucleotides is $(f(v_1),f(v_2)) = (\Ab,\Cb)$.
-However, such nucleotides are assigned to positions that are in the independent set $S$, and therefore cannot be adjacent.
-We conclude that $\Psi$ is surjective, and thus bijective.
-\end{proof}
-
-Now we can build on the connection between the two problems to obtain complexity results for \NumDesign.
-Counting independent sets in bipartite graphs ($\#{\sf BIS}$) is indeed a well-studied \#{\sf P}-hard problem~\citep{Ge2012}, from which we immediately conclude:
-\begin{corollary}
-  \NumDesign is $\#{\sf P}$-hard
-\end{corollary}
-\begin{proof}
-  Note that $\#{\sf BIS}$ is also \#{\sf P}-hard on connected graphs, as the number of independent sets for a disconnected graph $G$ is given by $|\IS{G}|=\prod_{cc\in CC(G)} |\IS{cc}|$.
-Thus any efficient algorithm for $\#{\sf BIS}$ on connected instances provides an efficient algorithm for general graphs.
-
-  Let us now hypothesize the existence of a polynomial-time algorithm $\mathcal{A}$ for \NumDesign over strongly-connected graphs $G$.
-Now consider the polynomial-time algorithm $\mathcal{A}'$ that first executes $\mathcal{A}$ on $G$ to produce $\NumDesign(G)$, and returns $\NumDesign(G)/2=|\Design{G}| = |\IS{G}|$.
-We conclude that $\mathcal{A}'$ solves $\#{\sf BIS}$ in polynomial-time.
-This means that $\NumDesign$  is at least as hard as $\#{\sf BIS}$, thus does not admit a polynomial time exact algorithm unless $\#{\sf P}={\sf FP}$.
-\end{proof}
 
@@
 specialized as RNA design network. This allows us to base our
-algorithm on cluster tree elimination (CTE) by Rina
-Dechter~\citep{Dechter2013}.
+algorithm on the cluster tree elimination (CTE) of~\citet{Dechter2013}.
 %
@@
 a sequence position. The functions $f\in\F$ of
-Section~\ref{sec:problem-statement}, are interpreted as functions on the network's variables.
+Section~\ref{sec:problem-statement}, are interpreted as functions on the network variables.
 
@@
   	  \For{$(v\to{}w)\in T$}{
-  	      $a_{\rm res} \gets \substitute{a_{\rm res}}{\Sample(v,\substitute{\val}{\val'};T,\chi,\phi,m)}$\; 
+  	      $a_{\rm res} \gets a_{\rm res} \mid \Sample(v,\substitute{\val}{\val'};T,\chi,\phi,m)$\; 
      }
@@
 
-As already noted by \citet{Flamm2001}, the
-requirement of canonical base pairing induces a bi-partition of each connected
-component of the base pair dependency graph, such that the variables of one
-partition are restricted to values $\{\Ab,\Gb\}$, but for other
-partition, to values $\{\Cb,\Ub\}$. This suggests a strong
-optimization of our algorithms, which reduces the basis in the
-exponential factor from 4 to 2. The price for this reduction is a
-binary choice for each of $c$ connected component. For each of the
-$2^c$ combinations, the partition functions can be computed separately
-in $O(n\,m\, 2^{w+1})$ time. This strategy is applied only if
-$c< w+1$. 
-%% SW: This could be optimized further by applying the strategy only to (optimally) selected components
-Finally, we compute the sums of the partition functions for all possible combinations.
-
-\begin{theorem}[Complexities]
+\begin{theorem}[Complexities]\label{th:complexities}
   For sequence length $n$, $k$ target structures, treewidth $w$ and
-  a base pair dependency graph having $c$ connected components, $t$ sequences
-  are sampled from the Boltzmann distribution  in
+  a compatibility graph having $c$ connected components, $t$ sequences
+  are generated from the Boltzmann distribution  in
+  $O( n\, k \, 4^{w+1} + t\, n\, k )$ time.
   %$O( n\, k \min(2^{w+c+1},4^{w+1}) + t\, n\, k )$ time.
-  $\mathcal{O}( 2^d\, n\, k  + t\, n\, k )$ time, where $d:=\min(w+c+1,2(w+1))$.
+  %$\mathcal{O}( 2^d\, n\, k  + t\, n\, k )$ time, where $d:=\min(w+c+1,2(w+1))$.
 \end{theorem}
+As shown in Supp. Mat.~\ref{sec:improvedComplexity}, the complexity of the precomputation can be further improved to 
+$\mathcal{O}(n\,k\,2^{w+1}\,2^{c})$, where $c$ is the maximum number of connected components represented in a node of the tree decomposition ($c\le w+1$).
 
@@
 % \end{equation}
-In simple \Def{base pair-based energy models}, energy is defined as
-the sum of base pair (pseudo-)energies. If base pair energies
-$\Ebp{k}{i,j,x,y}$ (where $i$ and $j$ are sequence positions, $x$ and
-$y$ are bases in $\Sigma$) are given for each target structure $\ell$,
-s.t.  $ E(S;R_\ell) := \sum_{(i,j)\in R_\ell} \Ebp{k}{i,j,S_i,S_j}$,
-we encode the network energy by the set of functions $f$ for each base
-pair $(i,j)\in R_\ell$ of each input structure $R_\ell$ that evaluate
-to $\ln(\pi_\ell) \Ebp{k}{i,j,\val(x_i),\val(x_j)}$ under assignment $\val$;
-here, $\pi_\ell>0$ is a weight that controls the
-influence of structure $\ell$ on the sampling (as elaborated later).
+In simple \Def{base pair-based energy models}, energy is defined as the sum of base pair (pseudo)energies. If base pair energies $\Ebp{k}{i,j,x,y}$ (where
+$i$ and $j$ are sequence positions, $x$ and $y$ are bases in $\Sigma$)
+are given for each target structure $\ell$, s.t.
+$ E(S;R_\ell) := \sum_{(i,j)\in R_\ell} \Ebp{k}{i,j,S_i,S_j}$, we
+encode the network energy by the set of functions
+$f$ for each base pair $(i,j)\in R_\ell$ of each input structure
+$R_\ell$ that evaluate to $w_\ell \Ebp{k}{i,j,\val(x_i),\val(x_j)}$
+under assignment $\val$.
 
@@
 While heuristic in nature, this basic iteration was elected in our initial version of \ourprog{} because of its reasonable empirical behavior (choosing $\gamma=1.2$).
+Now, consider Given an elected vertex $v$ of a connected component, say the first node encountered in a rooted
 
@@
 %Indeed, as shown in Fig.~\ref{fig:shifting_mean}, empirical joint distributions in the energies have good fits towards Normal multidimensional distributions with linear (co-)variances. 
-For base pair-based energy models and special base pair dependency graph
+For base pair-based energy models and special compatibility graph
 (paths, cycles\ldots) this property rigorously follows from analytic
@@
 Moreover, for each generated seed sequence, we optimized the objective function $(\ref{eq:blueprintobjective})$ through an adaptive  greedy walk consisting of 500 steps.
-At each step, we resampled (uniformly at random) the positions of a randomly selected component in the base pair dependency graph, and accepted the modification only if it resulted in a gain, as described in \Software{RNAblueprint}~\citep{Hammer2017}.
+At each step, we resampled (uniformly at random) the positions of a randomly selected component in the compatibility graph, and accepted the modification only if it resulted in a gain, as described in \Software{RNAblueprint}~\citep{Hammer2017}.
 %
@@
 \end{table}
+
+\section{Counting valid designs is \#{\sf P}-hard}\label{sec:counting}
+
+%%% reformulate later
+%We define the following notations: $\real^\infty:=\real\cup\{\infty\}$, $[i,j]:=\{i,\ldots,j\},$ \ldots
+Now we turn to the complexity of $\NumDesign(G)$, the problem of computing the number of valid sequences for a given compatibility graph $G=(V,E)$.
+Note that this problem corresponds to the partition function problem in a simple $(0/\infty)$-valued base pair model, with $\beta\neq 1$.
+As previously noted~\citep{Flamm2001}, a set of target structures admits a valid design iff its compatibility graph is bipartite, which can be tested in linear time.
+Moreover, without $\{\Gb, \Ub\}$ base pairs, any connected component  $C\in {\rm CC}(G)$ is entirely determined by the assignment of a single of its nucleotides.
+The number of valid designs is thus simply $4^{\#{\rm CC}(G)}$, where $\#{\rm CC}(G)$ is the \Def{number of connected components}.
+
+The introduction of $\{\Gb, \Ub\}$ base pairs radically changes this picture, and we show that valid designs for a set of structures cannot be counted in polynomial time, unless ${\sf\# P}={\sf FP}$.
+The latter equality would, in particular, imply the classic ${\sf P}={\sf NP}$, and solving $\NumDesign(G)$ is polynomial time is therefore probably difficult.
+
+To establish that claim, we consider instances $G=(V_1\cup V_2, E)$ that are connected and bipartite ($(V_1\times V_2) \cap E = \varnothing$), noting that hardness over restricted instances implies hardness of the general problem.
+Moreover remark that, as observed in Subsec.~\ref{sec:complexity}, assigning a nucleotide to a position $u\in V$ constrains the parity ($\{\Ab,\Gb\}$ or $\{\Cb,\Ub\}$) of all positions in the connected component of $u$.
+For this reason, we restrict our attention to the counting of valid designs \emph{up to trivial} $(\Ab\leftrightarrow \Cb/\Gb\leftrightarrow \Ub)$ symmetry, by constraining the positions in $V_{1}$ (resp.
+$V_2$) to only $\Ab$ and $\Gb$ (resp.
+$\Cb$ and $\Ub$).
+Let $\Design{G}$ denote the subset of all designs for $G$ under this constraint, noting that $\NumDesign(G) = 2\cdot|\Design{G}|$.
+
+We remind that an \Def{independent set} of $G=(V,E)$ is a subset $V'\subseteq V$ of nodes that are not connected by any edge in $E$. Let $\IS{G}$ denote the set of all independent sets in $G$.
+
+\begin{proposition}
+ $\Design{G}$ is in bijection with $\IS{G}$.
+\end{proposition}
+\begin{proof}
+Consider the function $\Psi: \Design{G} \to \IS{G}$ defined by $ \Psi(f) := \left\{v\in V\mid f(v)\in\{\Ab,\Cb\}\right\}.$
+
+Let us establish the injectivity of  $\Psi$, i.e. that $\Psi(f)\neq\Psi(f')$ for all $f\neq f'$.
+If $f\neq f'$, then there exists a node $v\in V$ such that $f(v)\neq f'(v)$. 
+Assume that $v\in V_1$, and remind that the only nucleotides allowed in $V_1$ are $\Ab$ and $\Gb$. Since $f(v)\neq f'(v)$, then we have $\{f(v),f'(v)\}=\{\Ab,\Gb\}$
+and we conclude that $\Psi(f)$ differs from $\Psi(f')$ at least with respect to its inclusion/exclusion of $v$.
+
+We turn to the surjectivity of $\Psi$, i.e. the existence of a preimage for each element $S\in \IS{G}$. Let us consider the function $f$ defined as
+\begin{align}
+ &\forall v_1\in V_1,v_2\in V_2:\notag\\& f(v_1) = \begin{cases} \Ab & \text{if }v_1\in S\\ \Gb & \text{if }v_1\notin S\end{cases} \text{\quad and\quad} 
+ f(v_2) = \begin{cases} \Cb & \text{if }v_2\in S\\ \Ub & \text{if }v_2\notin S\end{cases}
+\end{align}
+One easily verifies that $\Psi(f) = S$.
+
+We are then left to determine if $f$ is a valid design for $G$, i.e.
+if for each $(v, v') \in E$ one has  $\{f(v),f(v')\}\in \B.$ Since $G$ is bipartite, any edge in $E$ involves two nodes $v_1\in V_1$ and $v_2\in V_2$.
+Remark that, among all the possible assignments $f(v_1)$ and $f(v_2)$, the only invalid combination of nucleotides is $(f(v_1),f(v_2)) = (\Ab,\Cb)$.
+However, such nucleotides are assigned to positions that are in the independent set $S$, and therefore cannot be adjacent.
+We conclude that $\Psi$ is surjective, and thus bijective.
+\end{proof}
+
+Now we can build on the connection between the two problems to obtain complexity results for \NumDesign.
+Counting independent sets in bipartite graphs ($\#{\sf BIS}$) is indeed a well-studied \#{\sf P}-hard problem~\citep{Ge2012}, from which we immediately conclude:
+\begin{corollary}
+  \NumDesign is $\#{\sf P}$-hard
+\end{corollary}
+\begin{proof}
+  Note that $\#{\sf BIS}$ is also \#{\sf P}-hard on connected graphs, as the number of independent sets for a disconnected graph $G$ is given by $|\IS{G}|=\prod_{cc\in CC(G)} |\IS{cc}|$.
+Thus any efficient algorithm for $\#{\sf BIS}$ on connected instances provides an efficient algorithm for general graphs.
+
+  Let us now hypothesize the existence of a polynomial-time algorithm $\mathcal{A}$ for \NumDesign over strongly-connected graphs $G$.
+Consider the (polynomial-time) algorithm $\mathcal{A}'$ that first executes $\mathcal{A}$ on $G$ to produce $\NumDesign(G)$, and returns $\NumDesign(G)/2=|\Design{G}| = |\IS{G}|$.
+Clearly $\mathcal{A}'$ solves $\#{\sf BIS}$ in polynomial-time.
+This means that $\NumDesign$  is at least as hard as $\#{\sf BIS}$, thus does not admit a polynomial time exact algorithm unless $\#{\sf P}={\sf FP}$.
+\end{proof}
 
@@
 \newpage
+%\newgeometry{margin=2cm}
 \onecolumn
@@
 \item Compute its minimum free energy structure (ViennaRNA package);
-\item Add the new structure to the instances if the resulting base pair dependency graph is bipartite;
+\item Add the new structure to the instances if the resulting compatibility graph is bipartite;
 \item Repeat until $k$ structures are collected.
@@
 
+\section{Exploiting the bipartite nature of compatibility graphs to reduce the complexity}\label{sec:improvedComplexity}
+
+As already noted by \citet{Flamm2001}, the
+requirement of canonical base pairing induces a bi-partition of each connected
+component within the compatibility graph, such that the nucleotides assigned to the two set of nodes in the partition are 
+restricted to values $\{\Ab,\Gb\}$ and $\{\Cb,\Ub\}$ respectively.
+It follows that, within each connected component of the compatibility graph, the assignment of any position drastically restricts the domain of the remaining positions, \emph{i.e.} they may now be assigned to only 2 out of the original 4 nucleotides.
+\newcommand{\Parity}[1]{p_{#1}}
+To exploit this property, and further reduce the exponential factor of the complexity, we introduce the notion of \Def{parity} of a vertex/position, whereby vertices of \Def{even} (resp.
+\Def{odd}) polarities may only be assigned values in $\{\Ab,\Gb\}$ (resp.
+$\{\Cb,\Ub\}$).
+Neighbors of an odd vertex must be even and vice versa, so the choice of a parity for any single vertex caracterizes the parity of all vertices in a bipartite connected subgraph.
+
+Let $CC=\{cc_1,cc_2,\ldots\}$ denote the connected components of the compatibility graph, we consider a modified version of the cluster tree, obtained by:
+\begin{enumerate}
+\item Electing, in each connected component $cc_i$, the first position $v_i^\star$ from the component encountered from the root in a pre-order traversal of the cluster tree;\label{step:election}
+\item Introducing new variables $\mathcal{P}=\{\Parity{1},\ldots,\Parity{|CC|}\}$, representing the parity of elected vertices;
+\item Adding, to each node $x$ in the tree decomposition, the parity variables of connected components that are represented in the node, \emph{i.e.} such that at least one of the positions in the connected component is in $\vars(x)$; \label{step:addparities}
+\item Introducing, in the tree decomposition, a chain of intermediate nodes to fix cases where more than a single (parity) variable is introduced between a parent and its child.\label{step:intermediates}
+\end{enumerate}
+Note that, since connected components correspond to connected subtrees in the tree decomposition, then adding those variables preserves the properties of a tree decomposition.
+Also, each node in the former cluster tree introduces at most one new variable compared to its parent, so the elected node is uniquely defined by the rule stated in Step~\ref{step:election}.
+Finally, we note thatthe parity of a variable $v\in cc_i$ is entirely characterized by the parity $\Parity{i}$ of $v_i^\star$, and the parity of the distance between $v$ and $v_i^\star$ in the compatibility graph.
+Namely, the parity of $v$ is $\Parity{i}$ if the distance between $v$ and $v^{\star}_i$ is even, and opposite to $\Parity{i}$ if odd.
+It follows that, thanks to the new variables introduced in Step~\ref{step:addparities}, the domain of each of the original variables can be reduced to two alternatives instead of the original four.
+
+The computation of Alg.~\ref{alg:pf} on the new cluster tree, considering reduced assignments for position variables and additional odd/even assignments for parity variables, has complexity in $\mathcal{O}(n\,k\,2^{w+1}\,2^{c})$, where $n$ is the RNA length, $k$ is the number of structures, $w$ is the tree-width of the unmodified cluster tree, and $c$ is the maximum number of connected components represented in a node.
+It should be noted that $c\le \min(|CC|,w+1)$, so that the new complexity cannot be larger than the one stated in Th.~\ref{th:complexities}.
+Also, note that the number of nodes introduced by Step~\ref{step:intermediates} is bounded by $|CC|$, so that the introduction of these intermediate nodes does not impact the worst-case complexity.
+For the same reason, the complexity of the stochastic backtrack remains unchanged.
+
 \section{Parameters for the base pair and the stacking energy model}

### 2018-02-21 RNARedPrint-Bioinfo2018.tex

--- 2018-02-19 RNARedPrint-Bioinfo2018.tex
@@
 \end{definition}
-\section{Counting valid designs is \#{\sf P}-hard}\label{sec:counting}
-
-%%% reformulate later
-%We define the following notations: $\real^\infty:=\real\cup\{\infty\}$, $[i,j]:=\{i,\ldots,j\},$ \ldots
-Now we turn to the complexity of $\NumDesign(G)$, the problem of computing the number of valid sequences for a given base pair dependency graph $G=(V,E)$.
-Note that this problem corresponds to the partition function problem in a simple $(0/\infty)$-valued base pair model, with $\beta\neq 1$.
-As previously noted~\citep{Flamm2001}, a set of target structures admits a valid design iff its base pair dependency graph is bipartite, which can be tested in linear time.
-Moreover, without $\{\Gb, \Ub\}$ base pairs, any connected component  $C\in {\rm CC}(G)$ is entirely determined by the assignment of a single of its nucleotides.
-The number of valid designs is thus simply $4^{\#{\rm CC}(G)}$, where $\#{\rm CC}(G)$ is the \Def{number of connected components}.
-
-The introduction of $\{\Gb, \Ub\}$ base pairs radically changes this picture, and we show that valid designs for a set of structures cannot be counted in polynomial time, unless ${\sf\# P}={\sf FP}$.
-The latter equality would, in particular, imply the classic ${\sf P}={\sf NP}$, and solving $\NumDesign(G)$ is polynomial time is therefore probably difficult.
-
-To establish that claim, we consider instances $G=(V_1\cup V_2, E)$ that are connected and bipartite ($(V_1\times V_2) \cap E = \varnothing$), noting that hardness over restricted instances implies hardness of the general problem.
-Moreover remark that, as observed in Subsec.~\ref{sec:complexity}, assigning a nucleotide to a position $u\in V$ constrains the parity ($\{\Ab,\Gb\}$ or $\{\Cb,\Ub\}$) of all positions in the connected component of $u$.
-For this reason, we restrict our attention to the counting of valid designs \emph{up to trivial} $(\Ab\leftrightarrow \Cb/\Gb\leftrightarrow \Ub)$ symmetry, by constraining the positions in $V_{1}$ (resp.
-$V_2$) to only $\Ab$ and $\Gb$ (resp.
-$\Cb$ and $\Ub$).
-Let $\Design{G}$ denote the subset of all designs for $G$ under this constraint, noting that $\NumDesign(G) = 2\cdot|\Design{G}|$.
-
-We remind that an \Def{independent set} of $G=(V,E)$ is a subset $V'\subseteq V$ of nodes that are not connected by any edge in $E$. Let $\IS{G}$ denote the set of all independent sets in $G$.
-
-\begin{proposition}
- $\Design{G}$ is in bijection with $\IS{G}$.
-\end{proposition}
-\begin{proof}
-Consider the function $\Psi: \Design{G} \to \IS{G}$ defined by $ \Psi(f) := \left\{v\in V\mid f(v)\in\{\Ab,\Cb\}\right\}.$
-
-Let us establish the injectivity of  $\Psi$, i.e. that $\Psi(f)\neq\Psi(f')$ for all $f\neq f'$.
-If $f\neq f'$, then there exists a node $v\in V$ such that $f(v)\neq f'(v)$.
-Assume that $v\in V_1$, and remind that the only nucleotides allowed in $V_1$ are $\Ab$ and $\Gb$. Since $f(v)\neq f'(v)$, then we have $\{f(v),f'(v)\}=\{\Ab,\Gb\}$
-and we conclude that $\Psi(f)$ differs from $\Psi(f')$ at least with respect to its inclusion/exclusion of $v$.
-
-We turn to the surjectivity of $\Psi$, i.e. the existence of a preimage for each element $S\in \IS{G}$. Let us consider the function $f$ defined as
-\begin{align}
- &\forall v_1\in V_1,v_2\in V_2:\notag\\& f(v_1) = \begin{cases} \Ab & \text{if }v_1\in S\\ \Gb & \text{if }v_1\notin S\end{cases} \text{\quad and\quad}
- f(v_2) = \begin{cases} \Cb & \text{if }v_2\in S\\ \Ub & \text{if }v_2\notin S\end{cases}
-\end{align}
-One easily verifies that $\Psi(f) = S$.
-
-We are then left to determine if $f$ is a valid design for $G$, i.e.
-if for each $(v, v') \in E$ one has  $\{f(v),f(v')\}\in \B.$ Since $G$ is bipartite, any edge in $E$ involves two nodes $v_1\in V_1$ and $v_2\in V_2$.
-Remark that, among all the possible assignments $f(v_1)$ and $f(v_2)$, the only invalid combination of nucleotides is $(f(v_1),f(v_2)) = (\Ab,\Cb)$.
-However, such nucleotides are assigned to positions that are in the independent set $S$, and therefore cannot be adjacent.
-We conclude that $\Psi$ is surjective, and thus bijective.
-\end{proof}
-
-Now we can build on the connection between the two problems to obtain complexity results for \NumDesign.
-Counting independent sets in bipartite graphs ($\#{\sf BIS}$) is indeed a well-studied \#{\sf P}-hard problem~\citep{Ge2012}, from which we immediately conclude:
-\begin{corollary}
-  \NumDesign is $\#{\sf P}$-hard
-\end{corollary}
-\begin{proof}
-  Note that $\#{\sf BIS}$ is also \#{\sf P}-hard on connected graphs, as the number of independent sets for a disconnected graph $G$ is given by $|\IS{G}|=\prod_{cc\in CC(G)} |\IS{cc}|$.
-Thus any efficient algorithm for $\#{\sf BIS}$ on connected instances provides an efficient algorithm for general graphs.
-
-  Let us now hypothesize the existence of a polynomial-time algorithm $\mathcal{A}$ for \NumDesign over strongly-connected graphs $G$.
-Now consider the polynomial-time algorithm $\mathcal{A}'$ that first executes $\mathcal{A}$ on $G$ to produce $\NumDesign(G)$, and returns $\NumDesign(G)/2=|\Design{G}| = |\IS{G}|$.
-We conclude that $\mathcal{A}'$ solves $\#{\sf BIS}$ in polynomial-time.
-This means that $\NumDesign$  is at least as hard as $\#{\sf BIS}$, thus does not admit a polynomial time exact algorithm unless $\#{\sf P}={\sf FP}$.
-\end{proof}
 
@@
 
-As already noted by \citet{Flamm2001}, the
-requirement of canonical base pairing induces a bi-partition of each connected
-component of the base pair dependency graph, such that the positions of one
-partition are restricted to nucleotides $\{\Ab,\Gb\}$, but for other
-partition, to nucleotides $\{\Cb,\Ub\}$. This suggests a strong
-optimization of our algorithms, which reduces the basis in the
-exponential factor from 4 to 2. The price for this reduction is a
-binary choice for each of $c$ connected component. For each of the
-$2^c$ combinations, the partition functions can be computed separately
-in $O(n\,m\, 2^{w+1})$ time. This strategy is applied only if
-$c< w+1$.
-%% SW: This could be optimized further by applying the strategy only to (optimally) selected components
-Finally, we compute the sums of the partition functions for all possible combinations.
-
-\begin{theorem}[Complexities]
+\begin{theorem}[Complexities]\label{th:complexities}
   For sequence length $n$, $k$ target structures, treewidth $w$ and
   a base pair dependency graph having $c$ connected components, $t$ sequences
-  are sampled from the Boltzmann distribution  in
+  are generated from the Boltzmann distribution  in
+  $O( n\, k \, 4^{w+1} + t\, n\, k )$ time.
   %$O( n\, k \min(2^{w+c+1},4^{w+1}) + t\, n\, k )$ time.
-  $\mathcal{O}( 2^d\, n\, k  + t\, n\, k )$ time, where $d:=\min(w+c+1,2(w+1))$.
+  %$\mathcal{O}( 2^d\, n\, k  + t\, n\, k )$ time, where $d:=\min(w+c+1,2(w+1))$.
 \end{theorem}
+As shown in Supp. Mat.~\ref{sec:improvedComplexity}, the complexity of the precomputation can be further improved to 
+$\mathcal{O}(n\,k\,2^{w+1}\,2^{c})$, where $c$ is the maximum number of connected components represented in a node of the tree decomposition ($c\le w+1$).
 
@@
 
+\section{Counting valid designs is \#{\sf P}-hard}\label{sec:counting}
+
+%%% reformulate later
+%We define the following notations: $\real^\infty:=\real\cup\{\infty\}$, $[i,j]:=\{i,\ldots,j\},$ \ldots
+Now we turn to the complexity of $\NumDesign(G)$, the problem of computing the number of valid sequences for a given compatibility graph $G=(V,E)$.
+Note that this problem corresponds to the partition function problem in a simple $(0/\infty)$-valued base pair model, with $\beta\neq 1$.
+As previously noted~\citep{Flamm2001}, a set of target structures admits a valid design iff its compatibility graph is bipartite, which can be tested in linear time.
+Moreover, without $\{\Gb, \Ub\}$ base pairs, any connected component  $C\in {\rm CC}(G)$ is entirely determined by the assignment of a single of its nucleotides.
+The number of valid designs is thus simply $4^{\#{\rm CC}(G)}$, where $\#{\rm CC}(G)$ is the \Def{number of connected components}.
+
+The introduction of $\{\Gb, \Ub\}$ base pairs radically changes this picture, and we show that valid designs for a set of structures cannot be counted in polynomial time, unless ${\sf\# P}={\sf FP}$.
+The latter equality would, in particular, imply the classic ${\sf P}={\sf NP}$, and solving $\NumDesign(G)$ is polynomial time is therefore probably difficult.
+
+To establish that claim, we consider instances $G=(V_1\cup V_2, E)$ that are connected and bipartite ($(V_1\times V_2) \cap E = \varnothing$), noting that hardness over restricted instances implies hardness of the general problem.
+Moreover remark that, as observed in Subsec.~\ref{sec:complexity}, assigning a nucleotide to a position $u\in V$ constrains the parity ($\{\Ab,\Gb\}$ or $\{\Cb,\Ub\}$) of all positions in the connected component of $u$.
+For this reason, we restrict our attention to the counting of valid designs \emph{up to trivial} $(\Ab\leftrightarrow \Cb/\Gb\leftrightarrow \Ub)$ symmetry, by constraining the positions in $V_{1}$ (resp.
+$V_2$) to only $\Ab$ and $\Gb$ (resp.
+$\Cb$ and $\Ub$).
+Let $\Design{G}$ denote the subset of all designs for $G$ under this constraint, noting that $\NumDesign(G) = 2\cdot|\Design{G}|$.
+
+We remind that an \Def{independent set} of $G=(V,E)$ is a subset $V'\subseteq V$ of nodes that are not connected by any edge in $E$. Let $\IS{G}$ denote the set of all independent sets in $G$.
+
+\begin{proposition}
+ $\Design{G}$ is in bijection with $\IS{G}$.
+\end{proposition}
+\begin{proof}
+Consider the function $\Psi: \Design{G} \to \IS{G}$ defined by $ \Psi(f) := \left\{v\in V\mid f(v)\in\{\Ab,\Cb\}\right\}.$
+
+Let us establish the injectivity of  $\Psi$, i.e. that $\Psi(f)\neq\Psi(f')$ for all $f\neq f'$.
+If $f\neq f'$, then there exists a node $v\in V$ such that $f(v)\neq f'(v)$. 
+Assume that $v\in V_1$, and remind that the only nucleotides allowed in $V_1$ are $\Ab$ and $\Gb$. Since $f(v)\neq f'(v)$, then we have $\{f(v),f'(v)\}=\{\Ab,\Gb\}$
+and we conclude that $\Psi(f)$ differs from $\Psi(f')$ at least with respect to its inclusion/exclusion of $v$.
+
+We turn to the surjectivity of $\Psi$, i.e. the existence of a preimage for each element $S\in \IS{G}$. Let us consider the function $f$ defined as
+\begin{align}
+ &\forall v_1\in V_1,v_2\in V_2:\notag\\& f(v_1) = \begin{cases} \Ab & \text{if }v_1\in S\\ \Gb & \text{if }v_1\notin S\end{cases} \text{\quad and\quad} 
+ f(v_2) = \begin{cases} \Cb & \text{if }v_2\in S\\ \Ub & \text{if }v_2\notin S\end{cases}
+\end{align}
+One easily verifies that $\Psi(f) = S$.
+
+We are then left to determine if $f$ is a valid design for $G$, i.e.
+if for each $(v, v') \in E$ one has  $\{f(v),f(v')\}\in \B.$ Since $G$ is bipartite, any edge in $E$ involves two nodes $v_1\in V_1$ and $v_2\in V_2$.
+Remark that, among all the possible assignments $f(v_1)$ and $f(v_2)$, the only invalid combination of nucleotides is $(f(v_1),f(v_2)) = (\Ab,\Cb)$.
+However, such nucleotides are assigned to positions that are in the independent set $S$, and therefore cannot be adjacent.
+We conclude that $\Psi$ is surjective, and thus bijective.
+\end{proof}
+
+Now we can build on the connection between the two problems to obtain complexity results for \NumDesign.
+Counting independent sets in bipartite graphs ($\#{\sf BIS}$) is indeed a well-studied \#{\sf P}-hard problem~\citep{Ge2012}, from which we immediately conclude:
+\begin{corollary}
+  \NumDesign is $\#{\sf P}$-hard
+\end{corollary}
+\begin{proof}
+  Note that $\#{\sf BIS}$ is also \#{\sf P}-hard on connected graphs, as the number of independent sets for a disconnected graph $G$ is given by $|\IS{G}|=\prod_{cc\in CC(G)} |\IS{cc}|$.
+Thus any efficient algorithm for $\#{\sf BIS}$ on connected instances provides an efficient algorithm for general graphs.
+
+  Let us now hypothesize the existence of a polynomial-time algorithm $\mathcal{A}$ for \NumDesign over strongly-connected graphs $G$.
+Consider the (polynomial-time) algorithm $\mathcal{A}'$ that first executes $\mathcal{A}$ on $G$ to produce $\NumDesign(G)$, and returns $\NumDesign(G)/2=|\Design{G}| = |\IS{G}|$.
+Clearly $\mathcal{A}'$ solves $\#{\sf BIS}$ in polynomial-time.
+This means that $\NumDesign$  is at least as hard as $\#{\sf BIS}$, thus does not admit a polynomial time exact algorithm unless $\#{\sf P}={\sf FP}$.
+\end{proof}
+
 \section{Conclusion}
@@
 
+\section{Exploiting the bipartite nature of compatibility graphs to reduce the complexity}\label{sec:improvedComplexity}
+
+As already noted by \citet{Flamm2001}, the
+requirement of canonical base pairing induces a bi-partition of each connected
+component within the compatibility graph, such that the nucleotides assigned to the two set of nodes in the partition are 
+restricted to values $\{\Ab,\Gb\}$ and $\{\Cb,\Ub\}$ respectively.
+It follows that, within each connected component of the compatibility graph, the assignment of any position drastically restricts the domain of the remaining positions, \emph{i.e.} they may now be assigned to only 2 out of the original 4 nucleotides.
+\newcommand{\Parity}[1]{p_{#1}}
+To exploit this property, and further reduce the exponential factor of the complexity, we introduce the notion of \Def{parity} of a vertex/position, whereby vertices of \Def{even} (resp.
+\Def{odd}) polarities may only be assigned values in $\{\Ab,\Gb\}$ (resp.
+$\{\Cb,\Ub\}$).
+Neighbors of an odd vertex must be even and vice versa, so the choice of a parity for any single vertex caracterizes the parity of all vertices in a bipartite connected subgraph.
+
+Let $CC=\{cc_1,cc_2,\ldots\}$ denote the connected components of the compatibility graph, we consider a modified version of the cluster tree, obtained by:
+\begin{enumerate}
+\item Electing, in each connected component $cc_i$, the first position $v_i^\star$ from the component encountered from the root in a pre-order traversal of the cluster tree;\label{step:election}
+\item Introducing new variables $\mathcal{P}=\{\Parity{1},\ldots,\Parity{|CC|}\}$, representing the parity of elected vertices;
+\item Adding, to each node $x$ in the tree decomposition, the parity variables of connected components that are represented in the node, \emph{i.e.} such that at least one of the positions in the connected component is in $\chi(x)$; \label{step:addparities}
+\item Introducing, in the tree decomposition, a chain of intermediate nodes to fix cases where more than a single (parity) variable is introduced between a parent and its child.\label{step:intermediates}
+\end{enumerate}
+Note that, since connected components correspond to connected subtrees in the tree decomposition, then adding those variables preserves the properties of a tree decomposition.
+Also, each node in the former cluster tree introduces at most one new variable compared to its parent, so the elected node is uniquely defined by the rule stated in Step~\ref{step:election}.
+Finally, we note thatthe parity of a variable $v\in cc_i$ is entirely characterized by the parity $\Parity{i}$ of $v_i^\star$, and the parity of the distance between $v$ and $v_i^\star$ in the compatibility graph.
+Namely, the parity of $v$ is $\Parity{i}$ if the distance between $v$ and $v^{\star}_i$ is even, and opposite to $\Parity{i}$ if odd.
+It follows that, thanks to the new variables introduced in Step~\ref{step:addparities}, the domain of each of the original variables can be reduced to two alternatives instead of the original four.
+
+The computation of Alg.~\ref{alg:pf} on the new cluster tree, considering reduced assignments for position variables and additional odd/even assignments for parity variables, has complexity in $\mathcal{O}(n\,k\,2^{w+1}\,2^{c})$, where $n$ is the RNA length, $k$ is the number of structures, $w$ is the tree-width of the unmodified cluster tree, and $c$ is the maximum number of connected components represented in a node.
+It should be noted that $c\le \min(|CC|,w+1)$, so that the new complexity cannot be larger than the one stated in Th.~\ref{th:complexities}.
+Also, note that the number of nodes introduced by Step~\ref{step:intermediates} is bounded by $|CC|$, so that the introduction of these intermediate nodes does not impact the worst-case complexity.
+For the same reason, the complexity of the stochastic backtrack remains unchanged.
+
 \section{Parameters for the base pair and the stacking energy model}
@@
 
-\begin{table}
+\begin{table}[b]
   \centering

### 2018-02-27 RNARedPrint-Bioinfo2018.tex
some edits (plus comments)

@@
 \usepackage{array} % for better arrays (eg matrices) in maths
-\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
-  \let\itemize\compactitem
-  \let\enditemize\endcompactitem
-  \let\enumerate\compactenum
-  \let\endenumerate\endcompactenum
-  \let\description\compactdesc
-  \let\enddescription\endcompactdesc
-  \pltopsep=1pt
-  \plitemsep=1pt
-  \plparsep=1pt  
+% \usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
+%   \let\itemize\compactitem
+%   \let\enditemize\endcompactitem
+%   \let\enumerate\compactenum
+%   \let\endenumerate\endcompactenum
+%   \let\description\compactdesc
+%   \let\enddescription\endcompactdesc
+%   \pltopsep=1pt
+%   \plitemsep=1pt
+%   \plparsep=1pt  
 \usepackage{hyperref} 
@@
 \abstract{\textbf{Motivation:} The design of multi-stable RNA molecules has important applications in biology, medicine, and biotechnology.
-Synthetic design approaches largely benefit from effective in-silico methods, which can tremendously impact their cost and feasibility.
+Synthetic design approaches profit strongly from effective in-silico methods, which can tremendously impact their cost and feasibility.
 \\
-\textbf{Results:} Here, we revisit a central ingredient of most in-silico design methods: the sampling of sequences for the design of multi-target structures, possibly including pseudoknots.
-We establish the $\#${\sf P}-hardness of uniform sampling, and introduce \ourprog{}, a tree decomposition-based algorithm for efficient, fixed parameter tractable sampling.
-By modeling the problem as a constraint network, \ourprog{} supports generic Boltzmann-weighted sampling for arbitrary additive RNA energy models; this enables the generation of RNA sequences meeting specific goals like expected free energies or \GCb-content.
-Finally, we empirically study general properties of the approach and generate biologically relevant multi-target Boltzmann-weighted designs for a common design benchmark.
-Our results show significant improvements over seed sequences generated by uniform sampling, the previously best available sampling strategy for multi-target design.\\
+  \textbf{Results:} We revisit a central ingredient of most in-silico
+  design methods: the sampling of sequences for the design of
+  multi-target structures, possibly including pseudoknots. For this
+  task, we present the efficient, tree decomposition-based algorithm
+  \ourprog{}. Our fixed parameter tractable approach is underpinned by
+  establishing the $\#${\sf P}-hardness of uniform sampling. Modeling
+  the problem as a constraint network, \ourprog{} supports generic
+  Boltzmann-weighted sampling for arbitrary additive RNA energy
+  models; this enables the generation of RNA sequences meeting
+  specific goals like expected free energies or \GCb-content. Finally,
+  we empirically study general properties of the approach and generate
+  biologically relevant multi-target Boltzmann-weighted designs for a
+  common design benchmark. Generating seed sequences with \ourprog{}, we demonstrate significant improvements over the previously best multi-target sampling strategy (uniform sampling).\\
   \textbf{Availability:} Our software is freely available at: \url{https://github.com/yannponty/RNARedPrint}\\
   \textbf{Contact:} \href{yann.ponty@lix.polytechnique.fr}{yann.ponty@lix.polytechnique.fr}\\
-\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
-online.}
+  \textbf{Supplementary information:} Supplementary data are available
+  at \textit{Bioinformatics} online.}
 
@@
 ligand. Riboswitches have been a popular application of rational
-design~\citep{Wachsmuth2013,Domin2017}, partly motivated by their capacity to
-act as biosensors~\citep{Findeiss2017}. At the co-transcriptional
-level, certain RNA families feature alternative, evolutionarily
-conserved, transient structures~\citep{Zhu2013}, which facilitate the
-ultimate adoption of functional structures at full elongation.  More
-generally, simultaneous compatibility to multiple structures
-is a relevant design objective for engineering kinetically controlled RNAs, finally targeting prescribed folding pathways. Thus,
-modern applications of RNA design often target (the simultaneous
-stability of) multiple structures, additionally aiming at other
-features, such as specific \GCb-content~\citep{Reinharz2013} or the
-presence/absence of functionally relevant motifs (either anywhere or
-at specific positions)~\citep{Zhou2013}; these objectives motivate flexible
+design~\citep{Wachsmuth2013,Domin2017}, partly motivated by their
+capacity to act as biosensors~\citep{Findeiss2017}. At the
+co-transcriptional level, certain RNA families feature alternative,
+evolutionarily conserved, transient structures~\citep{Zhu2013}, which
+facilitate the ultimate adoption of functional structures at full
+elongation.  More generally, simultaneous compatibility to multiple
+structures is a relevant design objective for engineering kinetically
+controlled RNAs, finally targeting prescribed folding pathways. Thus,
+modern applications of RNA design often target multiple structures,
+additionally aiming at other features, such as specific
+\GCb-content~\citep{Reinharz2013} or the presence/absence of
+functionally relevant motifs (either anywhere or at specific
+positions)~\citep{Zhou2013}; these objectives motivate flexible
 computational design methods.
@@
 and optimizing them subsequently. In many cases, the seed quality was
-found to be critical for the empirical performance (and solution
-quality) of RNA design methods~\citep{Levin2012}. For instance, random
-seed generation improves the prospect of subsequent optimizations, helping
-to overcome local optima of the objective function, and induce better
-sequence diversity across designs~\citep{Reinharz2013}.  For
-single-target approaches, \Software{INFO-RNA}~\citep{Busch2006} made
-significant improvements mainly by starting its local search from the
-minimum energy sequence for the target structure instead of (uniform)
-random sequences for the early \Software{RNAinverse}
-algorithm~\citep{Hofacker1994}. This strategy was later shown to result
-in unrealistically high \GCb-contents in designed sequences. To
-address this issue, \Software{IncaRNAtion}~\citep{Reinharz2013}
-controls the \GCb-content through an adaptive sampling strategy.
+found to be critical for the empirical performance of RNA design
+methods~\citep{Levin2012}. For instance, random seed generation
+improves the prospect of subsequent optimizations, helping to overcome
+local optima of the objective function, and increases the diversity
+across designs~\citep{Reinharz2013}.  For single-target approaches,
+\Software{INFO-RNA}~\citep{Busch2006} made significant improvements
+mainly by starting its local search from the minimum energy sequence
+for the target structure instead of (uniform) random sequences for the
+early \Software{RNAinverse} algorithm~\citep{Hofacker1994}. This
+strategy was later shown to result in unrealistically high
+\GCb-contents in designed sequences. To address this issue,
+\Software{IncaRNAtion}~\citep{Reinharz2013} controls the \GCb-content
+through an adaptive sampling strategy.
 
@@
 parameter $\gamma$ is bounded by the length of the designed RNA;
-typically $\gamma$ is kept much lower by the decomposition.
+typically, the decomposition strategy achieves much lower $\gamma$.
 
@@
 The hardness of counting (and conjectured hardness of sampling) does not preclude, however, practically applicable algorithms for counting and sampling.
-In particular, we wish to extend the flexibility of multi-structure design, leading to the following questions: \emph{How to sample, generically, from a Boltzmann distribution based on expressive energy models?
+In particular, we wish to extend the flexibility of multi-structure design, leading to the following questions: \emph{How to sample, generally, from a Boltzmann distribution based on expressive energy models?
 How to enforce additional constraints, such as the \GCb-content, complex sequence constraints, or the energy of individual structures?}
@@
 % (Subsection~\ref{sec:PF})
-. We show that these algorithms are \Def{Fixed-Parameter Tractable} for the \Def{treewidth}%
+. We show that these algorithms are \Def{fixed-parameter tractable} for the \Def{treewidth}%
 % Subsection~\ref{sec:complexity}
@@
 % (SubSection~\ref{sec:multiBoltzmann})
-.
-Notably, the sampling procedure is based on a simplified RNA energy model, taking only the most important energy contributions into account, which crucially limits computation time and space; at the same time, our approach targets realistic RNA energies in the well-accepted Turner RNA energy model.
-This high accuracy enables efficiently generating biologically relevant multi-target designs in our final application of our overall strategy to a large set of multi-target RNA design instances from a representative benchmark Section~\ref{sec:results}).
+. %
+We observe that sampling based on less complex RNA energy models
+(taking only the most important energy contributions into account)
+still allows targeting realistic RNA energies in the well-accepted
+Turner RNA energy model. The resulting combination of efficiency and
+high accuracy finally enables generating biologically relevant
+multi-target designs in our final application of our overall strategy
+to a large set of multi-target RNA design instances from a
+representative benchmark Section~\ref{sec:results}).
 
@@
 % 
-\Def{Valid base pair} must pair bases from the
+\Def{Valid base pair} must pair bases from 
 $\B:=\left\{\{\Ab,\Ub\},\{\Gb,\Cb\},\{\Gb,\Ub\}\right\}.$
@@
 \Def{base pair dependency graph} $G_{\R}$ with nodes $\{1,\dots,n\}$
-and edges $S\bigcup_{\ell\in[1,k]} R_\ell$, which describe the minimal
+and edges $\bigcup_{\ell\in[1,k]} R_\ell$, which describe the minimal
 dependencies present in all relevant settings due to the requirement
@@
 
-Valid sequences for $\R$ can then be seen as specific colorings of
-$G_{\R}$, using the four colors in $\Sigma$, such that the colors
-assigned to adjacent vertices of $G_{\R}$ constitute valid base pairs.
+One can interpret the valid sequences for $\R$ as colorings of
+$G_{\R}$ in a slightly modified graph coloring variant, where the
+colors (from $\Sigma$) assigned to adjacent vertices of $G_{\R}$ must
+constitute valid base pairs.
 %
@@
 %
-  As will be explicit in subsequent sections, our approach relies on
+  As we elaborate in subsequent sections, our approach relies on
   breaking down the energy function $E_\R(S)$ into additive
@@
   s.t.~$E_\R(S)=\sum_{f\in\F} f(S)$. This captures realistic RNA
-  energy models---including nearest neighbor models, e.g.~\citet{Turner2009}, and even
-  pseudoknot models, e.g. \citet{Andronescu2010}, while bounding the dependencies to
+  energy models---including nearest neighbor models,
+  e.g.~\citet{Turner2009}, and even pseudoknot models,
+  e.g. \citet{Andronescu2010}, while bounding the dependencies to
   sequence positions introduced by each single $f$. Formally, define
-  the \Def{dependencies} $\dep(f)$ of $f$ as a minimal set of sequence
+  the \Def{dependencies} $\dep(f)$ of $f$ as the minimum set of sequence
   positions $\I\subseteq\{1,\dots,n\}$, where $f(S)=f(S')$ for all
-  sequences $S$ and $S'$ agreeing at all positions in $\I$.
-
-The \Def{dependency graph} induced by the set $\F$ of functions (on
-sequences of length $n$) is the hypergraph
-$G_\F=(\{1,\dots,n\} ,\{\dep(f)\mid f\in \F\})$.
-Our algorithms will critically rely on a \Def{tree decomposition} of the dependency graph, which we define below.
+  sequences $S$ and $S'$ that agree at the positions in $\I$.
+
+  Each set $\F$ of functions (on sequences of length $n$) induces a
+  \Def{dependency graph} on sequence positions, namely the hypergraph
+  $G_\F=(\{1,\dots,n\} ,\{\dep(f)\mid f\in \F\})$.  Our algorithms
+  will critically rely on a \Def{tree decomposition} of the dependency
+  graph, which we define below.
 \begin{definition}[Tree decomposition and width]
   \label{def:treedecomp}
-  Given a hypergraph $G=(X, E)$, a \Def{tree decomposition} is a
-  pair $(T,\chi)$, where $T$ is an unrooted tree/forest, and
-  $\chi(v)\subseteq X$ is the set of vertices from $G$ assigned to the node $v\in T$, such
-  that
+  Let $G=(X, E)$ be a hypergraph. A \Def{tree decomposition} of $G$ is
+  a pair $(T,\chi)$, where $T$ is an unrooted tree/forest, and (for
+  each $v\in T$) $\chi(v)\subseteq X$ is a set of vertices assigned to
+  the node tree $v\in T$, such that
 \begin{enumerate}
 \item each $x\in X$ occurs in at least one $\chi(v)$;
-\item  $\forall x\in X$, $\{ v \mid x \in \chi(v) \}$ induces a connected subtree of $T$;
-\item $\forall e\in E$ of $G$, there is a node $v\in T$, such that $e\subseteq\chi(v)$.
+\item for all $x\in X$, $\{ v \mid x \in \chi(v) \}$ induces a connected subtree of $T$;
+\item for all $e\in E$, there is a node $v\in T$, such that $e\subseteq\chi(v)$.
 \end{enumerate}
@@
 $\width(T,\chi) = \min_{u\in T} |\chi(u)| - 1 $. The \Def{treewidth}
-of $G$ is the smallest width in any tree decomposition of $G$.
+of $G$ is the smallest width of any tree decomposition of $G$.
 \end{definition}
@@
 replace the more general concept of (partial) assignments in
-constraint networks. Partially determined RNA sequences, called
+constraint networks. Partially determined RNA sequences, for short
 \Def{partial sequences}, are words $\val$ over the alphabet
@@
 \begin{definition}
-An \Def{RNA design network} is a tuple $\network=(\X,\F)$, where
+An \Def{RNA design network} (for sequences of length $n$) is a tuple $\network=(\X,\F)$, where\vspace{-6pt}
 \begin{itemize}
-\item[--]\ $\X$ is the set of sequence positions $1,\dots,n$
-\item[--]\ $\F$ is a set of \Def{functions} $f:\S_n\to\real$
+\item $\X$ is the set of sequence positions $1,\dots,n$
+\item $\F$ is a set of \Def{functions} $f:\S_n\to\real$
 \end{itemize}
@@
 $\network$ is defined as sum of the values of all functions in
-$f\in\F$ evaluated for $\val_S$, i.e.
-$\energy{\network}(S) := \sum_{f\in F} \evalfor{f}{\val_S}.$
+$f\in\F$ evaluated for $S$, i.e.
+$\energy{\network}(S) := \sum_{f\in F} \evalfor{f}{S}.$
 
@@
 $\partfun{\network} := \sum_{S}\exp(-\beta\energy{\network}(S)) = \sum_{S}\prod_{f\in F} \exp( -\beta\cdot
-\evalfor{f}{\val_S} ).$
+\evalfor{f}{S} ).$
 
@@
 denote the set of all partial sequences that determine exactly the positions
-in $\Y$; moreover, given partial sequences $\val$ and
+in $\Y$; furthermore, given partial sequences $\val$ and
 $\val'$, we define the \Def{combined partial sequence $\substitute{\val'}{\val''}$} such that
@@
 
-First, we observed that the computation time of tree decomposition
-(\Software{GreedyFillIn}, implemented in \Software{LibTW}~\citep{Dijk2006}) for multi-target
-sampling is negligible compared to Alg.~\ref{alg:pf} (Supp. Mat.~\ref{appsec:treedecomp} and \ref{appsec:dependency-cliques});
-we will thus omit the tree decomposition from our complexity analysis.
-
-For our analysis, we define the \emph{maximum separator size} $s$ as
+Note that in the following complexity analysis, we omit time and space for computing the
+tree decomposition itself, since we observed that the computation time
+of tree decomposition (\Software{GreedyFillIn}, implemented in
+\Software{LibTW}~by \citet{Dijk2006}) for multi-target sampling is
+negligible compared to Alg.~\ref{alg:pf}
+(Supp. Mat.~\ref{appsec:treedecomp} and
+\ref{appsec:dependency-cliques}).
+
+We define the \emph{maximum separator size} $s$ as
 $\max_{u,v\in V} | \separator{u}{v} |$ and denote the maximum size of
@@
 of specific optimizations, running Alg.~\ref{alg:pf} requires
-$\mathcal{O}((|\F|+|V|)\cdot 4^{w+1})$ time and $\mathcal{O}(|V|\cdot4^s)$ space (Supp. Mat.~\ref{appsec:algcomplexity});
-Alg.~\ref{alg:sampling} would require $\mathcal{O}((|\F|+|V|)\cdot 4^D)$ per
-sample on arbitrary tree decompositions (Supp. Mat.~\ref{appsec:algcomplexity}). W.l.o.g. we assume that
+$\mathcal{O}((|\F|+|V|)\cdot 4^{w+1})$ time and
+$\mathcal{O}(|V|\cdot4^s)$ space
+(Supp. Mat.~\ref{appsec:algcomplexity}); Alg.~\ref{alg:sampling} would
+require $\mathcal{O}((|\F|+|V|)\cdot 4^D)$ per sample on arbitrary
+tree decompositions
+(Supp. Mat.~\ref{appsec:algcomplexity}). W.l.o.g. we assume that
 $D=1$; note that tree decompositions can generally be transformed,
@@
 motifs, as well as features depending on the target structures.
-Naturally, hard constraints, which enforce or forbid features, can
-also be expressed by functions that assign infinite penalties to
-invalid (partial) sequences. More concretely, we briefly discuss the
-modeling of RNA energy models in our framework.
+Furthermore, constraints, which enforce or forbid features, are
+naturally expressed by assigning infinite penalties to invalid
+(partial) sequences. Finally, but less obviously, the framework
+captures various RNA energy models with bounded dependencies, which we
+describe briefly.
 
@@
 While heuristic in nature, this basic iteration was elected in our initial version of \ourprog{} because of its reasonable empirical behavior (choosing $\gamma=1.2$).
-Now, consider Given an elected vertex $v$ of a connected component, say the first node encountered in a rooted
+
+\textcolor{red}{\bf RED ALERT! CHECK: IS THERE SOMETHING MISSING?}
 
 A further \Def{rejection step} is applied to the generated structures to retain only sequences whose energy for each structure $R_\ell$ belongs to $[\TargetE_\ell\cdot(1-\varepsilon),\TargetE_\ell\cdot(1+\varepsilon)]$, for $\varepsilon\ge 0$ some predefined \Def{tolerance}.
-The rationale for a rejection approach follows from:
+The rejection approach is justified by the following considerations:
 i) \emph{Enacting an exact control over the energies would  be technically hard and costly.} Indeed, controlling the energies through dynamic programming would require explicit convolution products, generalizing~\citet{Cupal1996}, inducing additional $\Theta(n^{2k})$ time and $\Theta(n^k)$ space overheads;
@@
 To illustrate our above strategy, we sampled $n=1\,000$ sequences
-targeting ${\sf GC}\%=0.5$ and different Turner energies for
-the three structure targets of an example
-instance. Fig.~\ref{fig:energydist}A illustrates how the Turner
-energy distributions of the shown target structures can be accurately
-shifted to prescribed target energies; for comparison, we plot the
-respective energy distribution of uniformly and Boltzmann sampled
-sequences. Fig.~\ref{fig:energydist}B summarizes the targeting accuracy for the single structure energies over a larger set of instances
-from the \texttt{Modena} benchmark. We emphasize that only the simple stacking energies
-are directly targeted (at $\pm$10\% tolerance, expect for a few hard
-instances). Due to our linear adjustment, we still achieve well
-controlled energies in the ultimately relevant Turner energy model.
+targeting ${\sf GC}\%=0.5$ and different Turner energies for the three
+structure targets of an example instance. Fig.~\ref{fig:energydist}A
+illustrates how the Turner energy distributions of the shown target
+structures can be accurately shifted to prescribed target energies;
+for comparison, we plot the respective energy distribution of
+uniformly and Boltzmann sampled sequences. Fig.~\ref{fig:energydist}B
+summarizes the targeting accuracy for the single structure energies
+over a larger set of instances from the \texttt{Modena} benchmark. We
+emphasize that only the simple stacking energies are directly targeted
+(at $\pm$10\% tolerance, expect for a few hard instances). Due to our
+linear adjustment, we finally achieve well-controlled energies in the
+realistic Turner energy model.
 \begin{figure*}[t]
@@
 
-For a multi-target design instance, individual reasonable target energies are estimated from averaging the energy over samples at relatively high weights $e^{\beta}$, setting the targeted \GCb-content to 50\%.
-For each instance of the \Software{Modena} benchmark~\citep{Taneda2015}, we estimated such target energies and subsequently generated $1\,000$ seed sequences at these energies.
-As its rationale, this procedure is tailored to produce sequences with similar Turner energy that favor the stability of the target structures having moderate \GCb-content; all these properties are desirable objectives for RNA design.
-All sampled sequences are evaluated under the multi-stable design objective function of \citet{Hammer2017}:
+For a multi-target design instance, individual reasonable target
+energies are estimated from averaging the energy over samples at
+relatively high weights $e^{\beta}$, setting the targeted \GCb-content
+to 50\%. For each instance of the \Software{Modena}
+benchmark~\citep{Taneda2015}, we estimated such target energies and
+subsequently generated $1\,000$ seed sequences at these energies.
+Our procedure is tailored to produce sequences with similar Turner
+energy that favor the stability of the target structures having
+moderate \GCb-content; all these properties are desirable objectives
+for RNA design.  All sampled sequences are evaluated under the
+multi-stable design objective function of \citet{Hammer2017}:
 % \begin{align}
@@
 (Table~\ref{tab:benchmark-results}, seeds). Our first analysis reveals
-that Boltzmann-weighted sequences constitute better seeds, associated
+that Boltzmann-sampled sequences constitute better seeds, associated
 with better values ($\sim\! 69\%$ improvement on average), compared to
@@
 %We define the following notations: $\real^\infty:=\real\cup\{\infty\}$, $[i,j]:=\{i,\ldots,j\},$ \ldots
-Now we turn to the complexity of $\NumDesign(G)$, the problem of computing the number of valid sequences for a given compatibility graph $G=(V,E)$.
+Finally, we turn to the complexity of $\NumDesign(G)$, the problem of computing the number of valid sequences for a given compatibility graph $G=(V,E)$.
 Note that this problem corresponds to the partition function problem in a simple $(0/\infty)$-valued base pair model, with $\beta\neq 1$.
@@
 \section{Conclusion}
-Motivated by the---here established---hardness of the problem, we introduced a general framework and efficient algorithms for design of RNA sequences to multiple target structures with fine-tuned properties.
-Our method combines an FPT stochastic sampling algorithm with multi-dimensional Boltzmann sampling over distributions controlled by expressive RNA energy models.
-Compared to the previously best available sampling method (uniform sampling), this approach generated significantly better seed sequences for instances of a common multi-target design benchmark.
-
-The presented method enables new possibilities for sequence generation in the field of RNA sequence design by enforcing additional constraints, like the \GCb-content, while controlling the energy of  multiple target structures.
-Thus it presents a major advance over previously used ad-hoc sampling and even uniform sampling strategies.
-We have shown the practicality of such controlled sequence generation and studied its use for multi-target RNA design.
-Moreover, our framework is already equipped to include more complex sequence constraints, including mandatory/forbidden motifs at specific positions or anywhere in the designed sequences.
-It could also support negative design principles, for instance by penalizing a set of alternative helices/structures.
-However, such extensions require delicate calibration, and will be the object of future work.
+Motivated by the---here established---hardness of the problem, we
+introduced a general framework and efficient algorithms for design of
+RNA sequences to multiple target structures with fine-tuned
+properties. Our method combines an FPT stochastic sampling algorithm
+with multi-dimensional Boltzmann sampling over distributions
+controlled by expressive RNA energy models.  Compared to the
+previously best available sampling method (uniform sampling), this
+approach generated significantly better seed sequences for instances
+of a common multi-target design benchmark.
+
+The presented method enables new possibilities for sequence generation
+in the field of RNA sequence design by enforcing additional
+constraints, like the \GCb-content, while controlling the energy of
+multiple target structures. Thus it presents a major advance over
+previously applied ad-hoc sampling and even efficient uniform sampling
+strategies. We have shown the practicality of such controlled sequence
+generation and studied its use for multi-target RNA design. Moreover,
+our framework is equipped to include more complex sequence
+constraints, including mandatory/forbidden motifs at specific
+positions or anywhere in the designed sequences. Furthermore, the
+method can support even negative design principles, for instance by
+penalizing a set of alternative helices/structures. Based on this
+generality, we envision our framework---in extensions, which still
+require delicate elaboration---to support various complex RNA design
+scenarios far beyond the sketched applications.
 
@@
 However, in the same work, the authors establish a strong connection between the complexity of approximate counting and the uniform random generation.
-Namely, they showed that, for problems associated with self-reducible relations, approximate counting is equally as hard as (almost) uniform random generation.
+Namely, they showed that, for problems associated with self-reducible relations, approximate counting is equally hard as (almost) uniform random generation.
 We conjecture that the (almost) uniform sampling of sequences from multiple structures with pseudoknots is in fact \#{\sf BIS}-hard as soon as the number of input structures strictly exceeds $5$, as indicated by~\citet{Goldberg2004}, motivating even further our parameterized approach.
@@
 \begin{proof}
-  \newcommand{\TransVars}[1]{{\rm tvars}(#1)}
+  \newcommand{\TransVars}[1]{{\rm tverts}(#1)}
   \newcommand{\Children}[1]{{\rm Children}(#1)}
@@
 
+\textcolor{red}{\textbf{SW: Introducing variables now conflicts with
+    the specialization of constraint networks to RNA design
+    networks. Luckily, there should be an even easier way to describe
+    the complexity reduction without introducing new variables; ideally let's discuss}}
+
 Let $CC=\{cc_1,cc_2,\ldots\}$ denote the connected components of the compatibility graph, we consider a modified version of the cluster tree, obtained by:
@@
 Also, each node in the former cluster tree introduces at most one new variable compared to its parent, so the elected node is uniquely defined by the rule stated in Step~\ref{step:election}.
-Finally, we note thatthe parity of a variable $v\in cc_i$ is entirely characterized by the parity $\Parity{i}$ of $v_i^\star$, and the parity of the distance between $v$ and $v_i^\star$ in the compatibility graph.
+Finally, we note that the parity of a variable $v\in cc_i$ is entirely characterized by the parity $\Parity{i}$ of $v_i^\star$, and the parity of the distance between $v$ and $v_i^\star$ in the compatibility graph.
 Namely, the parity of $v$ is $\Parity{i}$ if the distance between $v$ and $v^{\star}_i$ is even, and opposite to $\Parity{i}$ if odd.

### 2018-02-28 RNARedPrint-Bioinfo2018.tex
Split off supplement and cross reference

@@
 %%% PACKAGES
+
+\usepackage{xr}
+\externaldocument{supplement}
+
 \usepackage{booktabs} % for much better looking tables
@@
 
-\newpage
-%\newgeometry{margin=2cm}
-\onecolumn
-
-\appendix
-{\centering \relsize{+4}Supplementary material\\%
-%{\relsize{-1}\BackupTitle}\\
-}
-\relsize{+1}
-\section{Approximate counting and random generation}
-In fact, not only is $\#{\sf BIS}$ a reference problem in counting complexity, but it is also a landmark problem with respect to the complexity of approximate counting problems.
-In this context, it is the representative for a class of $\#{\sf BIS}$-hard problems~\citep{Bulatov2013} that are easier to approximate than $\# {\sf SAT}$, yet are widely believed not to admit any Fully Polynomial-time Randomized Approximation Scheme.
-Recent results reveal a surprising dichotomy in the behavior of $\#{\sf BIS}$: it admits a Fully Polynomial-Time Approximation Scheme (FPTAS) for graphs of max degree $\le 5$~\citep{Weitz2006}, but is as hard to approximate as the general $\#{\sf BIS}$ problem on graphs of degree $\ge 6$~\citep{Cai2016}.
-In other words, there is a clear threshold, in term of the max degree, separating (relatively) easy instances from really hard ones.
-
-Additionally, let us note that, from the classic Vizing Theorem, any bipartite graph $G$ having maximum degree $\Delta$ can be decomposed in polynomial time in exactly $\Delta$ matchings.
-Any such matching can be reinterpreted as a secondary structure, possibly with crossing interactions (\Def{pseudoknots}).
-These results have two immediate consequences for the pseudoknotted version of the multiple design counting problem.
-\begin{corollary}[as follows from~\citep{Weitz2006}]The number of designs compatible with $m\le 5$ pseudoknotted RNA structures can be approximated within any fixed ratio by a deterministic polynomial-time algorithm.
-\end{corollary}
-\begin{corollary}[as follows from~\citep{Cai2016}]
-  As soon as the number of pseudoknotted RNA structures strictly exceeds $5$, \NumDesign is as hard to approximate as {\#{\sf BIS}}.
-\end{corollary}
-
-It is worth noting that the $\#{\sf P}$ hardness of \NumDesign does not immediately imply the hardness of generating a valid design uniformly at random, as demonstrated constructively by Jerrum, Valiant and Vazirani~\citep{Jerrum1986}.
-However, in the same work, the authors establish a strong connection between the complexity of approximate counting and the uniform random generation.
-Namely, they showed that, for problems associated with self-reducible relations, approximate counting is equally hard as (almost) uniform random generation.
-We conjecture that the (almost) uniform sampling of sequences from multiple structures with pseudoknots is in fact \#{\sf BIS}-hard as soon as the number of input structures strictly exceeds $5$, as indicated by~\citet{Goldberg2004}, motivating even further our parameterized approach.
-
-\section{Tree decomposition for RNA design instances in practice}
-\label{appsec:treedecomp}
-
-For studying the typically expected treewidths and tree decomposition
-run times in multi-target design instances, we consider five sets of
-multi-target RNA design instances of different complexity. Our first
-set consists of the Modena benchmark instances.
-
-In addition, we generated four sets of instances of increasing
-complexity. The instances of the sets RF3, RF4, RF5, and RF6, each
-respectively specify 3,4,5, and 6 target structures for sequence
-length 100.  For each instance (100 instances per set), we generated a
-set of $k$ ($k=3,\dots,6$) compatible structures as follows
-\begin{itemize}
-\item Generate a random sequence of length 100;
-\item Compute its minimum free energy structure (ViennaRNA package);
-\item Add the new structure to the instances if the resulting base pair dependency graph is bipartite;
-\item Repeat until $k$ structures are collected.
-\end{itemize}
-For each instance, we generated the dependency graphs in the base pair
-model and in the stacking model. Then, we performed tree decomposition
-(using strategy ``GreedyFillIn'' of LibTW~\citep{Dijk2006}) on each dependency
-graph. The obtained treewidths are reported in
-Fig.~\ref{fig:td-widths}, while Fig.~\ref{fig:td-times} shows the
-corresponding run-times of the tree decomposition.
-Fig.~\ref{fig:td-example} shows the tree decompositions for an example
-instance from set RF3.
-
-\begin{figure}
-  %\hspace*{2cm}\textbf{Base pair model}\hspace{4.75cm}\textbf{Stacking model}\\[-28pt]
-  \centering\includegraphics[width=0.9\textwidth]{Figs/td-widths}
-  \caption{Treewidths for multi-target RNA design instances of
-    different complexity. Distributions of treewidths shown as boxplots for the base pair (left) and stacking model (right).}
-  \label{fig:td-widths}
-\end{figure}
-
-\begin{figure}
-  %\hspace*{2cm}\textbf{Base pair model}\hspace{4.75cm}\textbf{Stacking model}\\[-28pt]
-  \centering
-  \includegraphics[width=0.9\textwidth]{Figs/td-times}
-  \caption{Computation time of tree decompositions for
-    multi-target RNA design instances of different complexity.
-    Distributions of times (in ms/instance) shown as boxplots for the base pair (left) and stacking model (right).}
-  \label{fig:td-times}
-\end{figure}
-
-\begin{figure}[t!]
-  %% the figure shows benchmark_inputs/RNAfold/3str/f3.100.0.inp
-  \centering
-  \verbatiminput{Figs/f3_100_0.txt}
-  \includegraphicstop[width=0.6\textwidth]{Figs/td-example-basepair}%
-  \includegraphicstop[width=0.35\textwidth]{Figs/td-example-stacking}
-  \caption{Example instance from RF3 (top) with its tree decompositions in the base pair (left) and stacking model (right). The respective treewidths are 2 and 12.}
-  \label{fig:td-example}
-\end{figure}
-
-\section{Expressing higher arity functions as cliques in the dependency graph}
-\label{appsec:dependency-cliques}
-
-Our implementation relies on external tools for the tree decomposition.
-Therefore it is not trivial that the weight functions (i.e.
-terms of the energy function) can be captured within the tree decompositions returned by a specific tool.
-In other words, it is not immediately clear how one can construct a cluster tree from an arbitrary tree decomposition for a given network.
-
-Crucially, one can show that, as long as the network the arguments $\dep(f)$ of a function $f$ are materialized by a clique within the network $\network$, then there exists at least one node in the tree decomposition that contains $\dep(f)$, thus allowing the evaluation of $f$.
-\begin{lemma}\label{lem:cliques}
-Let $G=(V,E)$ be an undirected graph, and  $T,\chi$ be a tree decomposition for $G$. For each clique $C\subset V$, there exists a node $n\in T$ such that $C\subset\chi(v)$.
-\end{lemma}
-\begin{proof}
-  \newcommand{\TransVars}[1]{{\rm tverts}(#1)}
-  \newcommand{\Children}[1]{{\rm Children}(#1)}
-  For the sake of simplicity, let us consider $T$ as rooted on an arbitrary node, inducing an orientation, and denote by $\TransVars{n}\in V$ the set of vertices found in a node $n$ or its descendants, namely
-    $$\TransVars{n} = \chi(n) \bigcup_{n'\in\Children{n}} \TransVars{n'}.$$
-
-  Now consider (one of) the deepest node(s) $n\in T$, such that $C\subset\TransVars{v}$.
-We are going to prove that $C\subset\chi(n)$, using contradiction, by showing that $C\not\subset\chi(n)$ implies that $T$ is not a tree decomposition.
-
-  Indeed, let us assume that $C\not\subset\chi(n)$, then there exists a node $v'\notin\chi(n)$ whose presence in $\TransVars{n}$ stems from its presence in some descendant of $n$.
-Let us denote by $n'\neq n$ the closest descendant of $n$ such that $v'\in\chi(n')$.
-Now, consider a node $v\in C$ such that $v\in \TransVars{v'}$ and $v\notin \TransVars{v'}$.
-Such a node always exists, otherwise $n'$, and not $n$, would be the deepest node such that $C\subset \TransVars{v'}$.
-From the definition of the tree decomposition, we know that neither $n'$ nor its descendants may contain $v$.
-On the other hand, none of the parents or siblings of $n'$ may contain $v'$.
-It follows that is no node $n''\in T$ whose vertex set $\chi(v'')$ includes, at the same time, $v$ and $v'$.
-Since both $v$ and $v'$ belong to a clique, one has $\{v,v'\}\in E$.
-The absence of a node in $T$ capturing an edge in $E$ contradicts our initial assumption that $T$ is tree decomposition for $G$.
-
-  We conclude that $n$ is such that $C\not\subset\chi(n)$.
-\end{proof}
-
-This result implies that classic tree decomposition algorithms and, more importantly, implementations can be directly re-used to produce decompositions that capture energy models of arbitrary complexity, functions of arbitrary arity.
-Indeed, it suffices to add a clique, involving the parameters of the function, and Lemma~\ref{lem:cliques} guarantees that the tree decomposition will feature one node to which the function can be associated.
-
-\section{Correctness of the FPT partition function algorithm}
-\label{appsec:correctness}
-
-\begin{theorem}[Correctness of Alg.~\ref{alg:pf}]
-  \label{the:pfalgo-correctness}
-  As computed by Alg.~\ref{alg:pf} for cluster tree $(T,\chi,\phi)$,
-  the messages $\Message{u}{v}$, for all edges $u\to{}v\in T$, yield
-  the partition functions of subtree of $u$ for the partial sequences
-  $\val\in\partseqs(\separator{u}{v})$, i.e. the messages satisfy
-  \begin{equation}
-   \evalfor{\Message{u}{v}}{\val} = \sum_{\val'\in\partseqs(\chi(T_r(u))-\separator{u}{v})} \quad
-   \prod_{f\in\phi(T_r(u))} \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}}),\label{eq:pfalgo-correct}
- \end{equation}
- where $\chi(T_r(u))$ denotes all $\chi$-assigned positions of nodes in $T_r(u)$;
- respectively $\phi(T_r(u))$; all $\phi$-assigned functions.
-%
-\end{theorem}
-
-\begin{proof}
-  Note that in more concise notation, Alg.~\ref{alg:pf} computes
-  messages such that
-\begin{equation}
-  \evalfor{\Message{u}{v}}{\val} := \sum_{\val'\in\partseqs(\difference{u}{v})}\quad
-  \prod_{f\in \phi(u) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}}) \prod_{(w\to{}u) \in T} \evalfor{\Message{w}{u}}{\substitute{\val'}{\val}}.\label{eq:messages}
-\end{equation}
-
-Proof by induction on $T$. If $u$ is a leaf, $\chi(r) = \chi(T_r(u))$,
-there are no messages sent to $u$, and $\phi(u) = \phi(T_r(u));$ implying
-Eq.~$(\ref{eq:pfalgo-correct}).$
-%
-Otherwise, since the algorithm traverses edges in postorder, $u$ received from its children
-$w_1,\dots,w_q$ the messages $\Message{w_1}{u}, \dots, \Message{w_q}{u}$, which satisfy Eq.~$(\ref{eq:pfalgo-correct})$ (induction hypothesis).
-Let $\val\in\partseqs(\separator{u}{v})$; then, $\evalfor{\Message{u}{v}}{\val}$ is computed by the algorithm according to Eq.~$(\ref{eq:messages})$.
-We rewrite as follows
-\begin{align*}
-  & \sum_{\val'\in\partseqs(\difference{u}{v})}\quad
-    \prod_{f\in \phi(u) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}})
-    \prod_{(w\to{}u) \in T} \evalfor{\Message{w}{u}}{\substitute{\val'}{\val}}\\
-  & = \sum_{\val'\in\partseqs(\chi(u)-\separator{u}{v})}\quad
-    \prod_{f\in \phi(u) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}})
-    \prod_{i=1}^q \evalfor{\Message{w_i}{u}}{\substitute{\val'}{\val}}\\
-  & =_{IH}
-    \sum_{\val'\in\partseqs(\chi(u)-\separator{u}{v})}\quad
-    \prod_{f\in \phi(u) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}})
-    \prod_{i=1}^q \sum_{\val''\in\partseqs(\chi(T_r(w_i))-\separator{w_i}{u})} \quad
-    \prod_{f\in\phi(T_r(w_i))} \exp(-\beta \evalfor{f}{\substitute{\val''}{\substitute{\val'}{\val}}}) \\
-& =_{*}
-    \sum_{\val'\in\partseqs(\chi(u)-\separator{u}{v})}\quad
-  \sum_{\val''\in\partseqs(\bigcup_{i=1}^q\chi(T_r(w_i))-\separator{w_i}{u})} \quad
-  \prod_{f\in \phi(u) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}})
-  \prod_{f\in\phi(T_r(w_i))} \exp(-\beta \evalfor{f}{\substitute{\val''}{\val'} | \val}) \\
-  & =_{(**)} \sum_{\val'\in\partseqs(\chi(T_r(u))-\separator{u}{v})}\quad
-    \prod_{f\in \phi(T_r(u)) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}})\\
-\end{align*}
-To see (*) and (**), we observe:
-\begin{itemize}
-\item The sets $\chi(u)-\separator{u}{v}$ and
-  $\chi(T_r(w_i))-\separator{w_i}{u}$ are all disjoint due to
-  Def.~\ref{def:treedecomp}, property 2. First, this property implies
-  that any shared position between the subtrees of $w_i$ and $w_j$
-  must be in $\chi(w_i)$, $\chi(w_j)$ and $\chi(u)$, thus the
-  positions of $\chi(T_r(w_i))-\separator{w_i}{u}$ are
-  disjoint. Second, if a position $\chi(T_r(w_i))$ occurs in
-  $\chi(u)$, it must occur in $\chi(w_i)$ and consequently in $\separator{u}{v}$.
-\item The union of the sets $\chi(u)-\separator{u}{v}$ and
-  $\chi(T_r(w_i))-\separator{w_i}{u}$ is $\chi(T_r(u))-\separator{u}{v}).$
-\end{itemize}
-
-\end{proof}
-
-\section{General complexity of the partition function computation by cluster tree elimination and generation of samples}
-\label{appsec:algcomplexity}
-
-\begin{proposition}
-  \label{prop:general-complexity}
-Given a cluster tree $(T,\chi,\phi)$, $T=(V,E)$ of the RNA design network $\network=(\X,\F)$ with treewidth $w$ and maximum separator size $s$, computing the partition function by Algorithm~\ref{alg:pf} takes $O((|F|+|V|)\cdot 4^{w+1})$ time and $O(|V| 4^s)$ space (for storing all messages).
-The sampling step has time complexity of $O((|F|+|V|)\cdot 4^D)$ per sample.
-\end{proposition}
-
-\begin{proof}[Proposition~\ref{prop:general-complexity}]
-Let $d_u$ denote the degree of node $u$ in $T$, $s_u$ is the size of the separator between
-$u$ and its parent in $T$ rooted at $r$.
-For each node in the cluster tree, Alg.~\ref{alg:pf} computes one message by combining $(|\phi(u)|+d_u-1)$ functions, each time enumerating $4^{|\chi(u)|}$ combinations; Alg.~\ref{alg:sampling} computes $4^{|\chi{u}|-s_u}$ partition functions each time combining $(|\phi(u)|+d_u-1)$ functions.
-$4^{|\chi(u)|}$ is bound by $4^{w+1}$ and $4^{|\chi{u}|-s_u}$ by $4^D$; moreover $\sum_{u\in V} (|\phi(u)|+d_u-1) = |F|+|V|-1$.
-\end{proof}
-
-\section{Exploiting the bipartite nature of compatibility graphs to reduce the complexity}\label{sec:improvedComplexity}
-
-As already noted by \citet{Flamm2001}, the
-requirement of canonical base pairing induces a bi-partition of each connected
-component within the compatibility graph, such that the nucleotides assigned to the two set of nodes in the partition are 
-restricted to values $\{\Ab,\Gb\}$ and $\{\Cb,\Ub\}$ respectively.
-It follows that, within each connected component of the compatibility graph, the assignment of any position drastically restricts the domain of the remaining positions, \emph{i.e.} they may now be assigned to only 2 out of the original 4 nucleotides.
-\newcommand{\Parity}[1]{p_{#1}}
-To exploit this property, and further reduce the exponential factor of the complexity, we introduce the notion of \Def{parity} of a vertex/position, whereby vertices of \Def{even} (resp.
-\Def{odd}) polarities may only be assigned values in $\{\Ab,\Gb\}$ (resp.
-$\{\Cb,\Ub\}$).
-Neighbors of an odd vertex must be even and vice versa, so the choice of a parity for any single vertex caracterizes the parity of all vertices in a bipartite connected subgraph.
-
-\textcolor{red}{\textbf{SW: Introducing variables now conflicts with
-    the specialization of constraint networks to RNA design
-    networks. Luckily, there should be an even easier way to describe
-    the complexity reduction without introducing new variables; ideally let's discuss}}
-
-Let $CC=\{cc_1,cc_2,\ldots\}$ denote the connected components of the compatibility graph, we consider a modified version of the cluster tree, obtained by:
-\begin{enumerate}
-\item Electing, in each connected component $cc_i$, the first position $v_i^\star$ from the component encountered from the root in a pre-order traversal of the cluster tree;\label{step:election}
-\item Introducing new variables $\mathcal{P}=\{\Parity{1},\ldots,\Parity{|CC|}\}$, representing the parity of elected vertices;
-\item Adding, to each node $x$ in the tree decomposition, the parity variables of connected components that are represented in the node, \emph{i.e.} such that at least one of the positions in the connected component is in $\chi(x)$; \label{step:addparities}
-\item Introducing, in the tree decomposition, a chain of intermediate nodes to fix cases where more than a single (parity) variable is introduced between a parent and its child.\label{step:intermediates}
-\end{enumerate}
-Note that, since connected components correspond to connected subtrees in the tree decomposition, then adding those variables preserves the properties of a tree decomposition.
-Also, each node in the former cluster tree introduces at most one new variable compared to its parent, so the elected node is uniquely defined by the rule stated in Step~\ref{step:election}.
-Finally, we note that the parity of a variable $v\in cc_i$ is entirely characterized by the parity $\Parity{i}$ of $v_i^\star$, and the parity of the distance between $v$ and $v_i^\star$ in the compatibility graph.
-Namely, the parity of $v$ is $\Parity{i}$ if the distance between $v$ and $v^{\star}_i$ is even, and opposite to $\Parity{i}$ if odd.
-It follows that, thanks to the new variables introduced in Step~\ref{step:addparities}, the domain of each of the original variables can be reduced to two alternatives instead of the original four.
-
-The computation of Alg.~\ref{alg:pf} on the new cluster tree, considering reduced assignments for position variables and additional odd/even assignments for parity variables, has complexity in $\mathcal{O}(n\,k\,2^{w+1}\,2^{c})$, where $n$ is the RNA length, $k$ is the number of structures, $w$ is the tree-width of the unmodified cluster tree, and $c$ is the maximum number of connected components represented in a node.
-It should be noted that $c\le \min(|CC|,w+1)$, so that the new complexity cannot be larger than the one stated in Th.~\ref{th:complexities}.
-Also, note that the number of nodes introduced by Step~\ref{step:intermediates} is bounded by $|CC|$, so that the introduction of these intermediate nodes does not impact the worst-case complexity.
-For the same reason, the complexity of the stochastic backtrack remains unchanged.
-
-\section{Parameters for the base pair and the stacking energy model}
-\label{appsec:modelparameters}
-
-We trained parameters for two RNA energy models to approximate the
-Turner energy model, as implemented in the \Software{ViennaRNA}
-package.  In the \Def{base pair model}, the total energy of a sequence $S$
-for an RNA structure $R$ is given as sum of base pair energies, where
-we consider six types of base pairs distinguishing by the bases A-U,
-C-G or G-U (symmetrically) and between stacked and non-stacked base
-pairs; here, we consider base pairs $(i,j)\in R$ \emph{stacked} iff
-$(i+1,j-1)\in R$, otherwise \emph{non-stacked}. In the \Def{stacking model},
-our features are defined by the stacks, i.e. pairs $(i,j)$ and
-$(i+1,j-1)$ which both occur in $R$; we distinguish 18 types based on
-$S_i,S_j,S_{i+1},S_{j-1}$ (i.e. all combinations that allow canonical
-base pairs; the configurations $S_i,S_j,S_{i+1},S_{j-1}$ and
-$S_{i+1},S_{j-1},S_j,S_i$ are symmetric).
-
-Both models describe the energy assigned to a pair of sequence and
-structure in linear dependency of the number of features and their
-weights. We can thus train weights for linear predictors of the Turner
-energy in both models.
-
-For this purpose, we generated 5000 uniform random RNA sequences of
-random lengths between 100 and 200. For each sample, we predict the
-minimum free energy and corresponding structure using the ViennaRNA
-package; then, we count the distinguished features (i.e., base pair or
-stack types). The parameters are estimated fitting linear models
-without intercept (\texttt{R} function \texttt{lm}). For both models,
-\texttt{R} reports an adjusted R-squared value of 0.99. The resulting
-parameters are reported in Tables~\ref{tab:basepairmodel} and
-\ref{tab:stackingmodel}.
-
-For validating the trained parameters, we generate a second
-independent test set of random RNA sequences in the same
-way. Fig.~\ref{fig:training-cor} shows correlation plots for the
-trained parameters in the base pair and stacking models for predicting
-the Turner energies in the test set with respective correlations of
-$0.95$ an $0.94$.
-
-\begin{figure}
-  \centering
-  A)\includegraphicstop[width=0.4\textwidth,trim=0 0 0 50,clip]{Figs/basepaircor}
-  B)\includegraphicstop[width=0.4\textwidth,trim=0 0 0 50,clip]{Figs/stackingcor}
-  \caption{Validation of the trained parameters for the A) base pair
-    model B) stacking model for predicting energies of the
-    independently sampled sequences in the test set. We show
-    correlation plots against the Turner energies reported by the
-    ViennaRNA package.}
-  \label{fig:training-cor}
-\end{figure}
-
-\begin{table}[b]
-  \centering
-  \caption{Trained weights for the base pair energy model.}
-  \label{tab:basepairmodel}
-  \begin{tabular}{c@{\quad}c@{\quad}c@{\quad}|@{\quad}c@{\quad}c@{\quad}c}
-    \multicolumn{3}{c}{non-stacked} & \multicolumn{3}{c}{stacked}\\
-    AU      & CG       & GU      & AU       & GC       & GU \\\hline
-    1.26630 & -0.09070 & 0.78566 & -0.52309 & -2.10208 & -0.88474
-  \end{tabular}
-\end{table}
-
-% GC_IN = -0.09070;
-% AU_IN = 1.26630;
-% GU_IN = 0.78566;
-% GC_TERM = -2.10208;
-% AU_TERM = -0.52309;
-% GU_TERM = -0.88474;
-
-\begin{table}[b]
-  \centering
-  \caption{Trained weights for the stacking energy model (the rows specify $S_i,S_j$; the columns, $S_{i+1},S_{j-1}$; we do not show the symmetric weights).}
-  \label{tab:stackingmodel}
-  \begin{tabular}{c@{\quad}|@{\quad}c@{\quad}c@{\quad}c@{\quad}c@{\quad}c@{\quad}c@{\quad}c@{\quad}c}
-       & AU & CG & GC & GU & UA & UG \\\hline
-    AU &
--0.18826 &
--1.13291 &
--1.09787 &
--0.38606 &
--0.26510 &
--0.62086
-    \\
-    CG &
--1.11752 &
--2.23740 &
--1.89434 &
--1.22942 &
--1.10548 &
--1.44085
-    \\
-    GU &
--0.55066 &
--1.26209 &
--1.58478 &
--0.72185 &
--0.49625 &
--0.68876
-    \\
-  \end{tabular}
-\end{table}
-
-\section{Monotonicity of the partial derivatives within weight calibration}
-\label{sec:weight_derivatives}
-%
-In weighted distributions, one witnesses a fairly predictable impact of the variation of any weight $\pi_\ell$ over the expected value $\mathbb{E}(E_\ell\mid \pmb{\pi})$, $\pmb{\pi}:= (\pi_1\cdots\pi_k)$, of the free energy $E_\ell$ of structure $\ell$.
-Let us  first remind the probability of a sequence $S$ in the weighted distribution
-\begin{align*}
-  \mathcal{B}_{\pmb{\pi}}(S) &= \prod_{i=1}^{k} \pi_i^{-E_i(S)}, &
-  \partfun{\pmb{\pi}}&=\sum_{S'}\mathcal{B}_{\pmb{\pi}}(S') &
-    \text{and}& &
-  \mathbb{P}(S\mid \pmb{\pi}) &= \frac{\mathcal{B}_{\pmb{\pi}}(S)}{\partfun{\pmb{\pi}}}.
-  \end{align*}
-
-Remind also the definition of the expectation of $E_\ell(S)$, for $S$ a Boltzmann-distributed sequence:
-$$\mathbb{E}(E_\ell\mid \pmb{\pi}) = \sum_S E_\ell(S)\cdot \mathbb{P}(S\mid \pmb{\pi}).$$
-
-We first remark that the partial derivative of $\mathcal{B}$ yields
-\begin{align*}
-  \frac{\partial \mathcal{B}_{\pmb{\pi}}(S)}{\partial \pi_\ell} = -E_\ell(S)\cdot \pi_\ell^{-E_\ell(S)-1}\prod_{\substack{i=1\\i\neq \ell}}^{k} \pi_i^{-E_i(S)} = \frac{-E_\ell(S)}{\pi_\ell}\cdot\mathcal{B}_{\pmb{\pi}}(S) = -E_\ell(S)\cdot \mathbb{P}(S\mid \pmb{\pi})\cdot \frac{\partfun{\pmb{\pi}}}{\pi_\ell }
-\end{align*}
-while the partial derivative of $\mathcal{Z}$ gives
-\begin{align*}
-  \frac{\partial \partfun{\pmb{\pi}}(S)}{\partial \pi_\ell} = \sum_{S'} -E_\ell(S')\cdot \mathbb{P}(S'\mid \pmb{\pi})\cdot \frac{\partfun{\pmb{\pi}}}{\pi_\ell }= -\mathbb{E}(E_\ell\mid \pmb{\pi})\cdot \frac{\partfun{\pmb{\pi}}}{\pi_\ell}.
-\end{align*}
-From these expressions, we conclude that
-\begin{align*}
-  \frac{\partial \mathbb{E}(E_\ell\mid \pmb{\pi})}{\partial \pi_\ell} &= \sum_S E_\ell(S)\cdot\frac{\partial \mathbb{P}(S\mid \pmb{\pi})}{\partial \pi_\ell}\\
-  %
-  & = \sum_S E_\ell(S)\left(\frac{\frac{\partial \mathcal{B}_{\pmb{\pi}}(S)}{\partial \pi_\ell}}{\mathcal{Z}_{\pmb{\pi}}} - \frac{\frac{\partial \mathcal{Z}_{\pmb{\pi}}}{\partial \pi_\ell}\cdot\mathcal{B}_{\pmb{\pi}}(S)}{\mathcal{Z}_{\pmb{\pi}}^2}\right)\\
-  %
-  & = \sum_S E_\ell(S)\left(\frac{-E_\ell(S)\cdot \mathbb{P}(S\mid \pmb{\pi})\cdot \frac{\mathcal{Z}_{\pmb{\pi}}}{\pi_\ell }}{\mathcal{Z}_{\pmb{\pi}}} + \frac{\mathbb{E}(E_\ell\mid \pmb{\pi})\cdot \frac{\mathcal{Z}_{\pmb{\pi}}}{\pi_\ell}\cdot\mathcal{B}_{\pmb{\pi}}(S)}{\mathcal{Z}_{\pmb{\pi}}^2}\right)\\
-  %
-  & = \sum_S - \frac{E_\ell(S)^2\cdot\mathbb{P}(S\mid \pmb{\pi})}{\pi_\ell} + \frac{\mathbb{E}(E_\ell\mid \pmb{\pi})}{\pi_\ell}\sum_S E_\ell(S)\cdot \mathbb{P}(S\mid \pmb{\pi})\\
-  & = -\frac{\mathbb{E}(E_\ell^2\mid \pmb{\pi}) - \mathbb{E}(E_\ell\mid \pmb{\pi})^2}{\pi_\ell}
-\end{align*}
-In the numerator of the above expression, one recognizes the variance of the distribution of $E_\ell$.
-Remark that a variance is always non-negative and, in our case, also strictly positive for $\pmb{\pi}$, $\pi_\ell>0$, as soon as there exist at least two distinct sequences $S$ and $S'$ such that $E_\ell(S)\neq E_\ell(S')$.
-Moreover, weights are always positive so the partial derivative in $\pi_\ell$ is always positive.
-Ultimately, it is always possible to increase (resp.
-decrease) the expected free energy of a structure by simply decreasing (resp.
-increasing) its weight, supporting our weight optimization procedure.
-\newpage
-\section{Detailed result of quality analysis}\label{sec:validity}
-N/A values correspond to data that could not be obtained by the time of this submission for two main reasons: In the case of the initial sampling (left columns), they correspond to instances which, in conjunction with an expressive energy model, resulted in very high tree-width, leading to unreasonable memory requirements incompatible with our available computing resources; The case of missing values after optimization  (right columns), they indicate situations where the initial optimization took too long ($\approx$ 1 day) and was killed.
-\input{table}
-
 \end{document}

### 2018-02-28 supplement.tex
Split off supplement and cross reference

\PassOptionsToPackage{utf8}{inputenc}
\documentclass{bioinfo}
\copyrightyear{2015} \pubyear{2015}

\access{Advance Access Publication Date: Day Month Year}
\appnotes{Manuscript Category}

%%% PACKAGES

\usepackage{xr}
\externaldocument{RNARedPrint-Bioinfo2018}

\usepackage{booktabs} % for much better looking tables
\usepackage{array} % for better arrays (eg matrices) in maths
% \usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
%   \let\itemize\compactitem
%   \let\enditemize\endcompactitem
%   \let\enumerate\compactenum
%   \let\endenumerate\endcompactenum
%   \let\description\compactdesc
%   \let\enddescription\endcompactdesc
%   \pltopsep=1pt
%   \plitemsep=1pt
%   \plparsep=1pt  
\usepackage{hyperref} 
\usepackage{xspace} 
%\usepackage{geometry} 
\usepackage{amsmath,amssymb} 
\usepackage{bm}
\usepackage{verbatim}
\usepackage{longtable}
\usepackage[vlined]{algorithm2e}

\usepackage{xifthen}
\usepackage{stmaryrd}

\usepackage{xcolor}

\hypersetup{
    bookmarks=false,         % show bookmarks bar?
    unicode=false,          % non-Latin characters in Acrobat’s bookmarks
    pdftoolbar=true,        % show Acrobat’s toolbar?
    pdfmenubar=true,        % show Acrobat’s menu?
    pdffitwindow=true,     % window fit to page when opened
    pdfstartview={FitH},    % fits the width of the page to the window
    colorlinks=true,       % false: boxed links; true: colored links
    linkcolor=red!70!black,          % color of internal links (change box color with linkbordercolor)
    citecolor=blue!50!black,        % color of links to bibliography
    filecolor=magenta,      % color of file links
    urlcolor=green!50!black           % color of external links
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Rolf's includegraphicstop
\makeatletter
\newsavebox{\@alignepsbox}
\newlength{\@aligneps}
\newcommand{\includegraphicstop}[2][]{%
\sbox{\@alignepsbox}{\includegraphics[#1]{#2}}%
\setlength{\@aligneps}{-\ht\@alignepsbox}%
\addtolength{\@aligneps}{2ex}%
\raisebox{\@aligneps}{\usebox{\@alignepsbox}}}
\makeatother

%\makeatletter
%\let\oldlt\longtable
%\let\endoldlt\endlongtable
%\def\longtable{\@ifnextchar[\longtable@i \longtable@ii}
%\def\longtable@i[#1]{\begin{figure}[t]
%\onecolumn
%\begin{minipage}{0.5\textwidth}
%\oldlt[#1]
%}
%\def\longtable@ii{\begin{figure}[t]
%\onecolumn
%\begin{minipage}{0.5\textwidth}
%\oldlt
%}
%\def\endlongtable{\endoldlt
%\end{minipage}
%\twocolumn
%\end{figure}}
%\makeatother

%%%%%%%%%%%%%%%%% Theorems %%%%%%%%%%%%%%%%%%%%%%%%%

 \newtheorem{theorem}{Theorem}
 \newtheorem{definition}[theorem]{Definition}
 \newtheorem{remark}[theorem]{Remark}
 \newtheorem{corollary}[theorem]{Corollary}
 \newtheorem{lemma}[theorem]{Lemma}
 \newtheorem{proposition}[theorem]{Proposition}

\newtheorem{observation}[theorem]{Observation}

%\newtheorem{algorithm}{Algorithm}
\newtheorem{axiom}{Axiom}
\newtheorem{hypothesis}{Working Hypothesis}
\renewenvironment{proof}[1][]{\noindent \em Proof\ifthenelse{\equal{#1}{}}{}{ (#1)}:~}{}

%%% macros for notation in DP framework
\newcommand{\network}{\mathcal{N}}
\newcommand{\val}{\bar S} % valuation aka assignment
\newcommand{\dep}{\operatorname{dep}}
\newcommand{\energy}[1]{\operatorname{e}_{#1}}
\newcommand{\numberof}{\operatorname{\#}}
\newcommand{\partfun}[1]{\mathcal{Z}_{#1}}
\newcommand{\separator}[2]{\operatorname{sep}(#1,#2)}
\newcommand{\difference}[2]{\operatorname{diff}(#1 \rightarrow #2)}
\newcommand{\real}{\mathbb{R}}
\newcommand{\genmarg}[1]{(\!|\!#1\!|\!)}
\newcommand{\gencomb}[1]{\langle\!|#1|\!\rangle}
\newcommand{\Message}[2]{m_{#1\rightarrow #2}}

\newcommand{\energyModel}{{\cal M}}
\newcommand{\structureElements}{{\cal SE}}
\newcommand{\powerSet}[1]{2^{#1}}
\newcommand{\underConstruction}[1]{{\LARGE$\triangle$\Large\!\!\!\!!}$\quad$\textcolor{red}{#1}}
\newcommand{\argmin}{\operatorname*{arg\,min}}
\newcommand{\objective}{{\mathbb{F}}}

\newcommand{\partseqs}{\mathcal{P\!S}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\R}{\mathcal{R}}
\renewcommand{\S}{\mathcal{S}}
\newcommand{\X}{\mathcal{X}}
\newcommand{\Y}{\mathcal{Y}}

\newcommand{\width}{w}

\newcommand{\sample}{\texttt{Sample}}
\newcommand{\elim}[2]{\operatorname{elim}(#1,#2)}
\newcommand{\edgesToR}{E^r_T}

\newcommand{\phitotal}{\phi_{\operatorname{m}}}

\newcommand{\Ebp}[2]{E^{\textrm{bp}}_{#1}(#2)}
\newcommand{\Ehp}[1]{E^{\textrm{hp}}(#1)}
\newcommand{\Eint}[1]{E^{\textrm{int}}(#1)}

\newcommand{\Def}[1]{{\bfseries #1}}

\newcommand{\TargetE}{E^{\star}}

\newcommand{\TODO}[1]{\textcolor{red!70!black}{\textbf{TODO: #1}}}

\newcommand{\parHead}[1]{\Final{\paragraph{#1}}}

\newcommand{\Final}[1]{#1}
%% Uncomment the line below for ``Final'' version
\renewcommand{\Final}[1]{}

\newcommand{\Design}[1]{{\sf Designs}^{\star}(#1)}
\newcommand{\NumDesign}{\ensuremath{\#}{\sf Designs}\xspace}
\newcommand{\IS}[1]{{\sf IndSets}(#1)}
\newcommand{\Nuc}[1]{{\sf #1}}
\newcommand{\Ab}{\Nuc{A}}
\newcommand{\Cb}{\Nuc{C}}
\newcommand{\Gb}{\Nuc{G}}
\newcommand{\Ub}{\Nuc{U}}

\newcommand{\GCb}{\Gb\Cb}

\newcommand{\Software}[1]{{\ttfamily #1}}

\newcommand{\ourprog}{\Software{RNARedPrint}}

\newcommand{\evalfor}[2]{#1\llbracket{}#2\rrbracket{}}
\newcommand{\substitute}[2]{#1\!\oplus\!#2}

\renewcommand{\gets}{:=}

\setlength{\parskip}{.2em}

%%% end macro defs

\begin{document}
%\newgeometry{margin=2cm}
\onecolumn

\appendix
{\centering \relsize{+4}Supplementary material\\%
%{\relsize{-1}\BackupTitle}\\
}
\relsize{+1}
\section{Approximate counting and random generation}
In fact, not only is $\#{\sf BIS}$ a reference problem in counting complexity, but it is also a landmark problem with respect to the complexity of approximate counting problems. In this context, it is the representative for a class of $\#{\sf BIS}$-hard problems~\citep{Bulatov2013} that are easier to approximate than $\# {\sf SAT}$, yet are widely believed not to admit any Fully Polynomial-time Randomized Approximation Scheme. Recent results reveal a surprising dichotomy in the behavior of $\#{\sf BIS}$: it admits a Fully Polynomial-Time Approximation Scheme (FPTAS) for graphs of max degree $\le 5$~\citep{Weitz2006}, but is as hard to approximate as the general $\#{\sf BIS}$ problem on graphs of degree $\ge 6$~\citep{Cai2016}. In other words, there is a clear threshold, in term of the max degree, separating (relatively) easy instances from really hard ones.

Additionally, let us note that, from the classic Vizing Theorem, any bipartite graph $G$ having maximum degree $\Delta$ can be decomposed in polynomial time in exactly $\Delta$ matchings. Any such matching can be reinterpreted as a secondary structure, possibly with crossing interactions (\Def{pseudoknots}). These results have two immediate consequences for the pseudoknotted version of the multiple design counting problem.
\begin{corollary}[as follows from~\citep{Weitz2006}]The number of designs compatible with $m\le 5$ pseudoknotted RNA structures can be approximated within any fixed ratio by a deterministic polynomial-time algorithm.
\end{corollary}
\begin{corollary}[as follows from~\citep{Cai2016}]
  As soon as the number of pseudoknotted RNA structures strictly exceeds $5$, \NumDesign is as hard to approximate as {\#{\sf BIS}}.
\end{corollary}

It is worth noting that the $\#{\sf P}$ hardness of \NumDesign does not immediately imply the hardness of generating a valid design uniformly at random, as demonstrated constructively by Jerrum, Valiant and Vazirani~\citep{Jerrum1986}. However, in the same work, the authors establish a strong connection between the complexity of approximate counting and the uniform random generation. Namely, they showed that, for problems associated with self-reducible relations, approximate counting is equally hard as (almost) uniform random generation. We conjecture that the (almost) uniform sampling of sequences from multiple structures with pseudoknots is in fact \#{\sf BIS}-hard as soon as the number of input structures strictly exceeds $5$, as indicated by~\citet{Goldberg2004}, motivating even further our parameterized approach.

\section{Tree decomposition for RNA design instances in practice}
\label{appsec:treedecomp}

For studying the typically expected treewidths and tree decomposition
run times in multi-target design instances, we consider five sets of
multi-target RNA design instances of different complexity. Our first
set consists of the Modena benchmark instances.

In addition, we generated four sets of instances of increasing
complexity. The instances of the sets RF3, RF4, RF5, and RF6, each
respectively specify 3,4,5, and 6 target structures for sequence
length 100.  For each instance (100 instances per set), we generated a
set of $k$ ($k=3,\dots,6$) compatible structures as follows
\begin{itemize}
\item Generate a random sequence of length 100;
\item Compute its minimum free energy structure (ViennaRNA package);
\item Add the new structure to the instances if the resulting base pair dependency graph is bipartite;
\item Repeat until $k$ structures are collected.
\end{itemize}
For each instance, we generated the dependency graphs in the base pair
model and in the stacking model. Then, we performed tree decomposition
(using strategy ``GreedyFillIn'' of LibTW~\citep{Dijk2006}) on each dependency
graph. The obtained treewidths are reported in
Fig.~\ref{fig:td-widths}, while Fig.~\ref{fig:td-times} shows the
corresponding run-times of the tree decomposition.
Fig.~\ref{fig:td-example} shows the tree decompositions for an example
instance from set RF3.

\begin{figure}
  %\hspace*{2cm}\textbf{Base pair model}\hspace{4.75cm}\textbf{Stacking model}\\[-28pt]
  \centering\includegraphics[width=0.9\textwidth]{Figs/td-widths}
  \caption{Treewidths for multi-target RNA design instances of
    different complexity. Distributions of treewidths shown as boxplots for the base pair (left) and stacking model (right).}
  \label{fig:td-widths}
\end{figure}

\begin{figure}
  %\hspace*{2cm}\textbf{Base pair model}\hspace{4.75cm}\textbf{Stacking model}\\[-28pt]
  \centering
  \includegraphics[width=0.9\textwidth]{Figs/td-times}
  \caption{Computation time of tree decompositions for
    multi-target RNA design instances of different complexity.
    Distributions of times (in ms/instance) shown as boxplots for the base pair (left) and stacking model (right).}
  \label{fig:td-times}
\end{figure}

\begin{figure}[t!]
  %% the figure shows benchmark_inputs/RNAfold/3str/f3.100.0.inp
  \centering
  \verbatiminput{Figs/f3_100_0.txt}
  \includegraphicstop[width=0.6\textwidth]{Figs/td-example-basepair}%
  \includegraphicstop[width=0.35\textwidth]{Figs/td-example-stacking}
  \caption{Example instance from RF3 (top) with its tree decompositions in the base pair (left) and stacking model (right). The respective treewidths are 2 and 12.}
  \label{fig:td-example}
\end{figure}

\section{Expressing higher arity functions as cliques in the dependency graph}
\label{appsec:dependency-cliques}

Our implementation relies on external tools for the tree decomposition. Therefore it is not trivial that the weight functions (i.e. terms of the energy function) can be captured within the tree decompositions returned by a specific tool. In other words, it is not immediately clear how one can construct a cluster tree from an arbitrary tree decomposition for a given network.

Crucially, one can show that, as long as the network the arguments $\dep(f)$ of a function $f$ are materialized by a clique within the network $\network$, then there exists at least one node in the tree decomposition that contains $\dep(f)$, thus allowing the evaluation of $f$.
\begin{lemma}\label{lem:cliques}
Let $G=(V,E)$ be an undirected graph, and  $T,\chi$ be a tree decomposition for $G$. For each clique $C\subset V$, there exists a node $n\in T$ such that $C\subset\chi(v)$.
\end{lemma}
\begin{proof}
  \newcommand{\TransVars}[1]{{\rm tverts}(#1)}
  \newcommand{\Children}[1]{{\rm Children}(#1)}
  For the sake of simplicity, let us consider $T$ as rooted on an arbitrary node, inducing an orientation, and denote by $\TransVars{n}\in V$ the set of vertices found in a node $n$ or its descendants, namely
    $$\TransVars{n} = \chi(n) \bigcup_{n'\in\Children{n}} \TransVars{n'}.$$

  Now consider (one of) the deepest node(s) $n\in T$, such that $C\subset\TransVars{v}$. We are going to prove that $C\subset\chi(n)$, using contradiction, by showing that $C\not\subset\chi(n)$ implies that $T$ is not a tree decomposition.

  Indeed, let us assume that $C\not\subset\chi(n)$, then there exists a node $v'\notin\chi(n)$ whose presence in $\TransVars{n}$ stems from its presence in some descendant of $n$. Let us denote by $n'\neq n$ the closest descendant of $n$ such that $v'\in\chi(n')$. Now, consider a node $v\in C$ such that $v\in \TransVars{v'}$ and $v\notin \TransVars{v'}$. Such a node always exists, otherwise $n'$, and not $n$, would be the deepest node such that $C\subset \TransVars{v'}$. From the definition of the tree decomposition, we know that neither $n'$ nor its descendants may contain $v$. On the other hand, none of the parents or siblings of $n'$ may contain $v'$. It follows that is no node $n''\in T$ whose vertex set $\chi(v'')$ includes, at the same time, $v$ and $v'$. Since both $v$ and $v'$ belong to a clique, one has $\{v,v'\}\in E$. The absence of a node in $T$ capturing an edge in $E$ contradicts our initial assumption that $T$ is tree decomposition for $G$.

  We conclude that $n$ is such that $C\not\subset\chi(n)$.
\end{proof}

This result implies that classic tree decomposition algorithms and, more importantly, implementations can be directly re-used to produce decompositions that capture energy models of arbitrary complexity, functions of arbitrary arity. Indeed, it suffices to add a clique, involving the parameters of the function, and Lemma~\ref{lem:cliques} guarantees that the tree decomposition will feature one node to which the function can be associated.

\section{Correctness of the FPT partition function algorithm}
\label{appsec:correctness}

\begin{theorem}[Correctness of Alg.~\ref{alg:pf}]
  \label{the:pfalgo-correctness}
  As computed by Alg.~\ref{alg:pf} for cluster tree $(T,\chi,\phi)$,
  the messages $\Message{u}{v}$, for all edges $u\to{}v\in T$, yield
  the partition functions of subtree of $u$ for the partial sequences
  $\val\in\partseqs(\separator{u}{v})$, i.e. the messages satisfy
  \begin{equation}
   \evalfor{\Message{u}{v}}{\val} = \sum_{\val'\in\partseqs(\chi(T_r(u))-\separator{u}{v})} \quad
   \prod_{f\in\phi(T_r(u))} \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}}),\label{eq:pfalgo-correct}
 \end{equation}
 where $\chi(T_r(u))$ denotes all $\chi$-assigned positions of nodes in $T_r(u)$;
 respectively $\phi(T_r(u))$; all $\phi$-assigned functions.
%
\end{theorem}

\begin{proof}
  Note that in more concise notation, Alg.~\ref{alg:pf} computes
  messages such that
\begin{equation}
  \evalfor{\Message{u}{v}}{\val} := \sum_{\val'\in\partseqs(\difference{u}{v})}\quad
  \prod_{f\in \phi(u) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}}) \prod_{(w\to{}u) \in T} \evalfor{\Message{w}{u}}{\substitute{\val'}{\val}}.\label{eq:messages}
\end{equation}

Proof by induction on $T$. If $u$ is a leaf, $\chi(r) = \chi(T_r(u))$,
there are no messages sent to $u$, and $\phi(u) = \phi(T_r(u));$ implying
Eq.~$(\ref{eq:pfalgo-correct}).$
%
Otherwise, since the algorithm traverses edges in postorder, $u$ received from its children
$w_1,\dots,w_q$ the messages $\Message{w_1}{u}, \dots, \Message{w_q}{u}$, which satisfy Eq.~$(\ref{eq:pfalgo-correct})$ (induction hypothesis). Let $\val\in\partseqs(\separator{u}{v})$; then, $\evalfor{\Message{u}{v}}{\val}$ is computed by the algorithm according to Eq.~$(\ref{eq:messages})$. We rewrite as follows
\begin{align*}
  & \sum_{\val'\in\partseqs(\difference{u}{v})}\quad
    \prod_{f\in \phi(u) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}})
    \prod_{(w\to{}u) \in T} \evalfor{\Message{w}{u}}{\substitute{\val'}{\val}}\\
  & = \sum_{\val'\in\partseqs(\chi(u)-\separator{u}{v})}\quad
    \prod_{f\in \phi(u) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}})
    \prod_{i=1}^q \evalfor{\Message{w_i}{u}}{\substitute{\val'}{\val}}\\
  & =_{IH}
    \sum_{\val'\in\partseqs(\chi(u)-\separator{u}{v})}\quad
    \prod_{f\in \phi(u) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}})
    \prod_{i=1}^q \sum_{\val''\in\partseqs(\chi(T_r(w_i))-\separator{w_i}{u})} \quad
    \prod_{f\in\phi(T_r(w_i))} \exp(-\beta \evalfor{f}{\substitute{\val''}{\substitute{\val'}{\val}}}) \\
& =_{*}
    \sum_{\val'\in\partseqs(\chi(u)-\separator{u}{v})}\quad
  \sum_{\val''\in\partseqs(\bigcup_{i=1}^q\chi(T_r(w_i))-\separator{w_i}{u})} \quad
  \prod_{f\in \phi(u) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}})
  \prod_{f\in\phi(T_r(w_i))} \exp(-\beta \evalfor{f}{\substitute{\val''}{\val'} | \val}) \\
  & =_{(**)} \sum_{\val'\in\partseqs(\chi(T_r(u))-\separator{u}{v})}\quad
    \prod_{f\in \phi(T_r(u)) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}})\\
\end{align*}
To see (*) and (**), we observe:
\begin{itemize}
\item The sets $\chi(u)-\separator{u}{v}$ and
  $\chi(T_r(w_i))-\separator{w_i}{u}$ are all disjoint due to
  Def.~\ref{def:treedecomp}, property 2. First, this property implies
  that any shared position between the subtrees of $w_i$ and $w_j$
  must be in $\chi(w_i)$, $\chi(w_j)$ and $\chi(u)$, thus the
  positions of $\chi(T_r(w_i))-\separator{w_i}{u}$ are
  disjoint. Second, if a position $\chi(T_r(w_i))$ occurs in
  $\chi(u)$, it must occur in $\chi(w_i)$ and consequently in $\separator{u}{v}$.
\item The union of the sets $\chi(u)-\separator{u}{v}$ and
  $\chi(T_r(w_i))-\separator{w_i}{u}$ is $\chi(T_r(u))-\separator{u}{v}).$
\end{itemize}

\end{proof}

\section{General complexity of the partition function computation by cluster tree elimination and generation of samples}
\label{appsec:algcomplexity}

\begin{proposition}
  \label{prop:general-complexity}
Given a cluster tree $(T,\chi,\phi)$, $T=(V,E)$ of the RNA design network $\network=(\X,\F)$ with treewidth $w$ and maximum separator size $s$, computing the partition function by Algorithm~\ref{alg:pf} takes $O((|F|+|V|)\cdot 4^{w+1})$ time and $O(|V| 4^s)$ space (for storing all messages). The sampling step has time complexity of $O((|F|+|V|)\cdot 4^D)$ per sample.
\end{proposition}

\begin{proof}[Proposition~\ref{prop:general-complexity}]
Let $d_u$ denote the degree of node $u$ in $T$, $s_u$ is the size of the separator between
$u$ and its parent in $T$ rooted at $r$. For each node in the cluster tree, Alg.~\ref{alg:pf} computes one message by combining $(|\phi(u)|+d_u-1)$ functions, each time enumerating $4^{|\chi(u)|}$ combinations; Alg.~\ref{alg:sampling} computes $4^{|\chi{u}|-s_u}$ partition functions each time combining $(|\phi(u)|+d_u-1)$ functions.  $4^{|\chi(u)|}$ is bound by $4^{w+1}$ and $4^{|\chi{u}|-s_u}$ by $4^D$; moreover $\sum_{u\in V} (|\phi(u)|+d_u-1) = |F|+|V|-1$.
\end{proof}

\section{Exploiting the bipartite nature of compatibility graphs to reduce the complexity}\label{sec:improvedComplexity}

As already noted by \citet{Flamm2001}, the
requirement of canonical base pairing induces a bi-partition of each connected
component within the compatibility graph, such that the nucleotides assigned to the two set of nodes in the partition are 
restricted to values $\{\Ab,\Gb\}$ and $\{\Cb,\Ub\}$ respectively. It follows that, within each connected component of the compatibility graph, the assignment of any position drastically restricts the domain of the remaining positions, \emph{i.e.} they may now be assigned to only 2 out of the original 4 nucleotides.
\newcommand{\Parity}[1]{p_{#1}}
To exploit this property, and further reduce the exponential factor of the complexity, we introduce the notion of \Def{parity} of a vertex/position, whereby vertices of \Def{even} (resp. \Def{odd}) polarities may only be assigned values in $\{\Ab,\Gb\}$ (resp. $\{\Cb,\Ub\}$). Neighbors of an odd vertex must be even and vice versa, so the choice of a parity for any single vertex caracterizes the parity of all vertices in a bipartite connected subgraph.

\textcolor{red}{\textbf{SW: Introducing variables now conflicts with
    the specialization of constraint networks to RNA design
    networks. Luckily, there should be an even easier way to describe
    the complexity reduction without introducing new variables; ideally let's discuss}}

Let $CC=\{cc_1,cc_2,\ldots\}$ denote the connected components of the compatibility graph, we consider a modified version of the cluster tree, obtained by:
\begin{enumerate}
\item Electing, in each connected component $cc_i$, the first position $v_i^\star$ from the component encountered from the root in a pre-order traversal of the cluster tree;\label{step:election}
\item Introducing new variables $\mathcal{P}=\{\Parity{1},\ldots,\Parity{|CC|}\}$, representing the parity of elected vertices;
\item Adding, to each node $x$ in the tree decomposition, the parity variables of connected components that are represented in the node, \emph{i.e.} such that at least one of the positions in the connected component is in $\chi(x)$; \label{step:addparities}
\item Introducing, in the tree decomposition, a chain of intermediate nodes to fix cases where more than a single (parity) variable is introduced between a parent and its child.\label{step:intermediates}
\end{enumerate}
Note that, since connected components correspond to connected subtrees in the tree decomposition, then adding those variables preserves the properties of a tree decomposition. Also, each node in the former cluster tree introduces at most one new variable compared to its parent, so the elected node is uniquely defined by the rule stated in Step~\ref{step:election}. Finally, we note that the parity of a variable $v\in cc_i$ is entirely characterized by the parity $\Parity{i}$ of $v_i^\star$, and the parity of the distance between $v$ and $v_i^\star$ in the compatibility graph. Namely, the parity of $v$ is $\Parity{i}$ if the distance between $v$ and $v^{\star}_i$ is even, and opposite to $\Parity{i}$ if odd. It follows that, thanks to the new variables introduced in Step~\ref{step:addparities}, the domain of each of the original variables can be reduced to two alternatives instead of the original four.

The computation of Alg.~\ref{alg:pf} on the new cluster tree, considering reduced assignments for position variables and additional odd/even assignments for parity variables, has complexity in $\mathcal{O}(n\,k\,2^{w+1}\,2^{c})$, where $n$ is the RNA length, $k$ is the number of structures, $w$ is the tree-width of the unmodified cluster tree, and $c$ is the maximum number of connected components represented in a node. It should be noted that $c\le \min(|CC|,w+1)$, so that the new complexity cannot be larger than the one stated in Th.~\ref{th:complexities}. Also, note that the number of nodes introduced by Step~\ref{step:intermediates} is bounded by $|CC|$, so that the introduction of these intermediate nodes does not impact the worst-case complexity. For the same reason, the complexity of the stochastic backtrack remains unchanged.

\section{Parameters for the base pair and the stacking energy model}
\label{appsec:modelparameters}

We trained parameters for two RNA energy models to approximate the
Turner energy model, as implemented in the \Software{ViennaRNA}
package.  In the \Def{base pair model}, the total energy of a sequence $S$
for an RNA structure $R$ is given as sum of base pair energies, where
we consider six types of base pairs distinguishing by the bases A-U,
C-G or G-U (symmetrically) and between stacked and non-stacked base
pairs; here, we consider base pairs $(i,j)\in R$ \emph{stacked} iff
$(i+1,j-1)\in R$, otherwise \emph{non-stacked}. In the \Def{stacking model},
our features are defined by the stacks, i.e. pairs $(i,j)$ and
$(i+1,j-1)$ which both occur in $R$; we distinguish 18 types based on
$S_i,S_j,S_{i+1},S_{j-1}$ (i.e. all combinations that allow canonical
base pairs; the configurations $S_i,S_j,S_{i+1},S_{j-1}$ and
$S_{i+1},S_{j-1},S_j,S_i$ are symmetric).

Both models describe the energy assigned to a pair of sequence and
structure in linear dependency of the number of features and their
weights. We can thus train weights for linear predictors of the Turner
energy in both models.

For this purpose, we generated 5000 uniform random RNA sequences of
random lengths between 100 and 200. For each sample, we predict the
minimum free energy and corresponding structure using the ViennaRNA
package; then, we count the distinguished features (i.e., base pair or
stack types). The parameters are estimated fitting linear models
without intercept (\texttt{R} function \texttt{lm}). For both models,
\texttt{R} reports an adjusted R-squared value of 0.99. The resulting
parameters are reported in Tables~\ref{tab:basepairmodel} and
\ref{tab:stackingmodel}.

For validating the trained parameters, we generate a second
independent test set of random RNA sequences in the same
way. Fig.~\ref{fig:training-cor} shows correlation plots for the
trained parameters in the base pair and stacking models for predicting
the Turner energies in the test set with respective correlations of
$0.95$ an $0.94$.

\begin{figure}
  \centering
  A)\includegraphicstop[width=0.4\textwidth,trim=0 0 0 50,clip]{Figs/basepaircor}
  B)\includegraphicstop[width=0.4\textwidth,trim=0 0 0 50,clip]{Figs/stackingcor}
  \caption{Validation of the trained parameters for the A) base pair
    model B) stacking model for predicting energies of the
    independently sampled sequences in the test set. We show
    correlation plots against the Turner energies reported by the
    ViennaRNA package.}
  \label{fig:training-cor}
\end{figure}

\begin{table}[b]
  \centering
  \caption{Trained weights for the base pair energy model.}
  \label{tab:basepairmodel}
  \begin{tabular}{c@{\quad}c@{\quad}c@{\quad}|@{\quad}c@{\quad}c@{\quad}c}
    \multicolumn{3}{c}{non-stacked} & \multicolumn{3}{c}{stacked}\\
    AU      & CG       & GU      & AU       & GC       & GU \\\hline
    1.26630 & -0.09070 & 0.78566 & -0.52309 & -2.10208 & -0.88474
  \end{tabular}
\end{table}

% GC_IN = -0.09070;
% AU_IN = 1.26630;
% GU_IN = 0.78566;
% GC_TERM = -2.10208;
% AU_TERM = -0.52309;
% GU_TERM = -0.88474;

\begin{table}[b]
  \centering
  \caption{Trained weights for the stacking energy model (the rows specify $S_i,S_j$; the columns, $S_{i+1},S_{j-1}$; we do not show the symmetric weights).}
  \label{tab:stackingmodel}
  \begin{tabular}{c@{\quad}|@{\quad}c@{\quad}c@{\quad}c@{\quad}c@{\quad}c@{\quad}c@{\quad}c@{\quad}c}
       & AU & CG & GC & GU & UA & UG \\\hline
    AU &
-0.18826 &
-1.13291 &
-1.09787 &
-0.38606 &
-0.26510 &
-0.62086
    \\
    CG &
-1.11752 &
-2.23740 &
-1.89434 &
-1.22942 &
-1.10548 &
-1.44085
    \\
    GU &
-0.55066 &
-1.26209 &
-1.58478 &
-0.72185 &
-0.49625 &
-0.68876
    \\
  \end{tabular}
\end{table}

\section{Monotonicity of the partial derivatives within weight calibration}
\label{sec:weight_derivatives}
%
In weighted distributions, one witnesses a fairly predictable impact of the variation of any weight $\pi_\ell$ over the expected value $\mathbb{E}(E_\ell\mid \pmb{\pi})$, $\pmb{\pi}:= (\pi_1\cdots\pi_k)$, of the free energy $E_\ell$ of structure $\ell$.  Let us  first remind the probability of a sequence $S$ in the weighted distribution
\begin{align*}
  \mathcal{B}_{\pmb{\pi}}(S) &= \prod_{i=1}^{k} \pi_i^{-E_i(S)}, &
  \partfun{\pmb{\pi}}&=\sum_{S'}\mathcal{B}_{\pmb{\pi}}(S') &
    \text{and}& &
  \mathbb{P}(S\mid \pmb{\pi}) &= \frac{\mathcal{B}_{\pmb{\pi}}(S)}{\partfun{\pmb{\pi}}}.
  \end{align*}

Remind also the definition of the expectation of $E_\ell(S)$, for $S$ a Boltzmann-distributed sequence:
$$\mathbb{E}(E_\ell\mid \pmb{\pi}) = \sum_S E_\ell(S)\cdot \mathbb{P}(S\mid \pmb{\pi}).$$

We first remark that the partial derivative of $\mathcal{B}$ yields
\begin{align*}
  \frac{\partial \mathcal{B}_{\pmb{\pi}}(S)}{\partial \pi_\ell} = -E_\ell(S)\cdot \pi_\ell^{-E_\ell(S)-1}\prod_{\substack{i=1\\i\neq \ell}}^{k} \pi_i^{-E_i(S)} = \frac{-E_\ell(S)}{\pi_\ell}\cdot\mathcal{B}_{\pmb{\pi}}(S) = -E_\ell(S)\cdot \mathbb{P}(S\mid \pmb{\pi})\cdot \frac{\partfun{\pmb{\pi}}}{\pi_\ell }
\end{align*}
while the partial derivative of $\mathcal{Z}$ gives
\begin{align*}
  \frac{\partial \partfun{\pmb{\pi}}(S)}{\partial \pi_\ell} = \sum_{S'} -E_\ell(S')\cdot \mathbb{P}(S'\mid \pmb{\pi})\cdot \frac{\partfun{\pmb{\pi}}}{\pi_\ell }= -\mathbb{E}(E_\ell\mid \pmb{\pi})\cdot \frac{\partfun{\pmb{\pi}}}{\pi_\ell}.
\end{align*}
From these expressions, we conclude that
\begin{align*}
  \frac{\partial \mathbb{E}(E_\ell\mid \pmb{\pi})}{\partial \pi_\ell} &= \sum_S E_\ell(S)\cdot\frac{\partial \mathbb{P}(S\mid \pmb{\pi})}{\partial \pi_\ell}\\
  %
  & = \sum_S E_\ell(S)\left(\frac{\frac{\partial \mathcal{B}_{\pmb{\pi}}(S)}{\partial \pi_\ell}}{\mathcal{Z}_{\pmb{\pi}}} - \frac{\frac{\partial \mathcal{Z}_{\pmb{\pi}}}{\partial \pi_\ell}\cdot\mathcal{B}_{\pmb{\pi}}(S)}{\mathcal{Z}_{\pmb{\pi}}^2}\right)\\
  %
  & = \sum_S E_\ell(S)\left(\frac{-E_\ell(S)\cdot \mathbb{P}(S\mid \pmb{\pi})\cdot \frac{\mathcal{Z}_{\pmb{\pi}}}{\pi_\ell }}{\mathcal{Z}_{\pmb{\pi}}} + \frac{\mathbb{E}(E_\ell\mid \pmb{\pi})\cdot \frac{\mathcal{Z}_{\pmb{\pi}}}{\pi_\ell}\cdot\mathcal{B}_{\pmb{\pi}}(S)}{\mathcal{Z}_{\pmb{\pi}}^2}\right)\\
  %
  & = \sum_S - \frac{E_\ell(S)^2\cdot\mathbb{P}(S\mid \pmb{\pi})}{\pi_\ell} + \frac{\mathbb{E}(E_\ell\mid \pmb{\pi})}{\pi_\ell}\sum_S E_\ell(S)\cdot \mathbb{P}(S\mid \pmb{\pi})\\
  & = -\frac{\mathbb{E}(E_\ell^2\mid \pmb{\pi}) - \mathbb{E}(E_\ell\mid \pmb{\pi})^2}{\pi_\ell}
\end{align*}
In the numerator of the above expression, one recognizes the variance of the distribution of $E_\ell$.
Remark that a variance is always non-negative and, in our case, also strictly positive for $\pmb{\pi}$, $\pi_\ell>0$, as soon as there exist at least two distinct sequences $S$ and $S'$ such that $E_\ell(S)\neq E_\ell(S')$.
Moreover, weights are always positive so the partial derivative in $\pi_\ell$ is always positive. Ultimately, it is always possible to increase (resp. decrease) the expected free energy of a structure by simply decreasing (resp. increasing) its weight, supporting our weight optimization procedure.
\newpage
\section{Detailed result of quality analysis}\label{sec:validity}
N/A values correspond to data that could not be obtained by the time of this submission for two main reasons: In the case of the initial sampling (left columns), they correspond to instances which, in conjunction with an expressive energy model, resulted in very high tree-width, leading to unreasonable memory requirements incompatible with our available computing resources; The case of missing values after optimization  (right columns), they indicate situations where the initial optimization took too long ($\approx$ 1 day) and was killed.
\input{table}

\bibliographystyle{natbib}
%\bibliographystyle{achemnat}
%\bibliographystyle{plainnat}
%\bibliographystyle{abbrv}
%\bibliographystyle{bioinformatics}
%
%\bibliographystyle{plain}
%
\bibliography{biblio}

\end{document}

### 2018-02-28 RNARedPrint-Bioinfo2018.tex
Typos

@@
 \newcommand{\numberof}{\operatorname{\#}}
-\newcommand{\partfun}[1]{\mathcal{Z}_{#1}}
+\newcommand{\partfun}[1]{Z_{#1}}
 \newcommand{\separator}[2]{\operatorname{sep}(#1,#2)}
@@
 $f\in\F$ evaluated for $S$, i.e.
-$\energy{\network}(S) := \sum_{f\in F} \evalfor{f}{S}.$
+$\energy{\network}(S) := \sum_{f\in\F} \evalfor{f}{S}.$
 
@@
 Eq.~$(\ref{eq:mainproblem})$ is modeled as network partition function
-$\partfun{\network} := \sum_{S}\exp(-\beta\energy{\network}(S)) = \sum_{S}\prod_{f\in F} \exp( -\beta\cdot
+$\partfun{\network} := \sum_{S}\exp(-\beta\energy{\network}(S)) = \sum_{S}\prod_{f\in\F} \exp( -\beta\cdot
 \evalfor{f}{S} ).$
@@
   \For{$\val\in\partseqs(\separator{u}{v})$}{
-    $\evalfor{\Message{u}{v}}{\val} \gets 0$\;
+    $x\gets 0$\;
     \For{$\val'\in\partseqs(\difference{u}{v})$}{
-     $p \gets$ product( $exp(-\beta \evalfor{f}{\substitute{\val}{\val'}})$ for $f\in \phi(u)$ )\;
-     $p \gets p\ \cdot\ $product( $\evalfor{\Message{w}{u}}{\substitute{\val}{\val'}}$ for $(w\to{}u)\in T$ )\;
-     $\evalfor{\Message{u}{v}}{\val} \gets \evalfor{\Message{u}{v}}{\val} + p$\;
+     $p \gets$ product( $exp(-\beta \evalfor{f}{\substitute{\val}{\val'}})$ for $f\in \phi(u)$ )\\
+     ${}\quad\qquad \cdot\ $product( $\evalfor{\Message{w}{u}}{\substitute{\val}{\val'}}$ for $(w\to{}u)\in T$ )\;
+     $x \gets x + p$\;
    }
+   $\evalfor{\Message{u}{v}}{\val} \gets x$\;
   }
@@
 
-     $p \gets$ product( $exp(-\beta \evalfor{f}{\substitute{\val}{\val'}})$ for $f\in \phi(u)$ )\;
-     $p \gets p\ \cdot\ $product( $\evalfor{\Message{w}{u}}{\substitute{\val}{\val'}}$ for $(w\to{}u)\in T$ )\;
+     $p \gets$ product( $exp(-\beta \evalfor{f}{\substitute{\val}{\val'}})$ for $f\in \phi(u)$\\
+     ${}\quad\qquad \cdot\ $product( $\evalfor{\Message{w}{u}}{\substitute{\val}{\val'}}$ for $(w\to{}u)\in T$ )\;
 %
@@
   	  \If{$r<0$}{
-  	    $a_{\rm res} \gets \substitute{\val}{\val'}$\;
+  	    $\val_{\rm res} \gets \substitute{\val}{\val'}$\;
   	  \For{$(v\to{}w)\in T$}{
-  	      $a_{\rm res} \gets \substitute{a_{\rm res}}{\Sample(v,\substitute{\val}{\val'};T,\chi,\phi,m)}$\;
+  	      $\val_{\rm res} \gets \substitute{\val_{\rm res}}{\Sample(v,\substitute{\val}{\val'};T,\chi,\phi,m)}$\;
      }
-  	  \Return {$a_{\rm res}$}
+  	  \Return {$\val_{\rm res}$}
   	  \;}

### 2018-02-28 supplement.tex
Add alternative strategy and description for complexity reduction due to bipartiteness (and constraints in more generality)

@@
 
-\section{Exploiting the bipartite nature of compatibility graphs to reduce the complexity}\label{sec:improvedComplexity}
+\section{Exploiting constraint consistency to reduce the complexity}\label{sec:improvedComplexity}
 
-As already noted by \citet{Flamm2001}, the
-requirement of canonical base pairing induces a bi-partition of each connected
-component within the compatibility graph, such that the nucleotides assigned to the two set of nodes in the partition are 
-restricted to values $\{\Ab,\Gb\}$ and $\{\Cb,\Ub\}$ respectively.
-It follows that, within each connected component of the compatibility graph, the assignment of any position drastically restricts the domain of the remaining positions, \emph{i.e.} they may now be assigned to only 2 out of the original 4 nucleotides.
-\newcommand{\Parity}[1]{p_{#1}}
-To exploit this property, and further reduce the exponential factor of the complexity, we introduce the notion of \Def{parity} of a vertex/position, whereby vertices of \Def{even} (resp.
-\Def{odd}) polarities may only be assigned values in $\{\Ab,\Gb\}$ (resp.
-$\{\Cb,\Ub\}$).
-Neighbors of an odd vertex must be even and vice versa, so the choice of a parity for any single vertex caracterizes the parity of all vertices in a bipartite connected subgraph.
+\textcolor{red}{\textbf{SW: Simpler alternative to
+    Sec.~\ref{sec:first-improvedComplexity}. Note also that the use of
+    variables in Sec.~\ref{sec:first-improvedComplexity} conflicts
+    with the (recent) specialization of constraint networks to RNA design
+    networks.}}
 
-\textcolor{red}{\textbf{SW: Introducing variables now conflicts with
-    the specialization of constraint networks to RNA design
-    networks. Luckily, there should be an even easier way to describe
-    the complexity reduction without introducing new variables; ideally let's discuss}}
+While Alg.~\ref{alg:pf} computes messages values for
+\emph{all} possible combinations of nucleotides for the positions in
+the separator set, we observe here that this is \emph{not} required for computing all relevant
+partition functions. In particular, the algorithm can be restricted to
+consider only \emph{valid} combinations, which satisfy the (hard)
+constraints on the nucleotides.
 
-Let $CC=\{cc_1,cc_2,\ldots\}$ denote the connected components of the compatibility graph, we consider a modified version of the cluster tree, obtained by:
+RNA design introduces constraints due to the requirement of canonical
+(aka complemenentary) base pairing, which can be exploited in a
+particularly simple and effective way.  As already noted by
+\citet{Flamm2001}, the base pair complementarity induces a
+bi-partition of each connected component within the base pair
+dependency graph, such that the nucleotides assigned to the two set of
+nodes in the partition are restricted to values $\{\Ab,\Gb\}$ and
+$\{\Cb,\Ub\}$ respectively; we call a partitial sequence
+\emph{cc-valid}, iff its determined positions satisfy this separation
+for all determined positions of the same connected component.
+
+% Let $CC=\{cc_1$, $cc_2, \dots\}$ denote the vertex sets of the
+% connected components of the base pair dependency graph. In this graph,
+% let each component have the bi-partition $cc_i=cc^0_i\cup cc^1_i$.
+
+One can now modify Alg.~\ref{alg:pf}, on tree decomposition
+$(T,\chi)$, such that the message values
+$\evalfor{\Message{u}{v}}{\val}$ are computed only for cc-valid
+partial sequences $\val\in\partseqs(\separator{u}{v})$. Moreover, the
+loop over $\val'\in\partseqs(\difference{u}{v})$ is restricted, such
+that $\substitute{\val}{\val'}$ are cc-valid. (The analogous change is
+made to the sampling algorithm Alg.~\ref{alg:sampling}.)
+
+The correctness of the modified algorithm still follows from an
+induction argument: in this way, the message computation of one node
+evaluates messages from its children only at cc-valid partial
+sequences. The result of this computation is a message, which is
+defined at all cc-valid partial sequences.
+
+Remarkably, this strongly improves the algorithm's time
+complexity. Where the original algorithm enumerates all partial
+sequences $\substitute{\val}{\val'}$, where
+$\val\in\partseqs(\separator{u}{v})$ and
+$\val'\in\partseqs(\difference{u}{v})$, the modified algorithm
+enumerates in two stages:
+\begin{enumerate}
+\item for each connected component of the base pair dependency graph
+  that involves any position in
+  $\separator{u}{v}\cup\difference{u}{v} = \chi(v)$, it first
+  enumerates all possibilities how the positions from the partition
+  sets are assigned to either $\{\Ab,\Gb\}$ or $\{\Cb,\Ub\}$
+\item it enumerates the possible partial sequences from the (according
+  to the decisions of the first stage) restricted domains of the
+  position in $\chi(v)$.
+\end{enumerate}
+Let $c(v)$ be the number of connected components involving positions
+in $\chi(v)$; then, the first stage enumerates $2^{c(v)}$
+combinations. For each such combination, the second stage enumerates
+only $2^{|\chi(v)|}$ partial sequences, since there remain only two
+possible nucleotide choices for each position. All message values for
+node $v$ are thus computed in time
+$\Theta(2^{c(v)+|\chi(v)|})$. Consequently, the total time complexity
+of the modified algorithm is $$O(nk2^{\hat w}),$$ where $n$ is the RNA
+length, $k$ is the number of structures, and
+$\hat w := \max_{v\in T} c(v)+|\chi(v)|$. Notably, $\hat w$ can not
+exceed $2(w+1)$ such that the modified algorithm does not increase the
+original complexity.
+%
+As useful corollary, one obtains the (looser) complexity
+bound $$O(nk2^{w+1}2^{cc})$$ in terms of treewidth $w$ and connected
+components $cc$.
+
+Finally, we remark that even stronger time savings could be possible
+in practice, since cc-valid partial sequences can still violate
+complementarity constraints, e.g. by assigning C and A to positions in
+different sets of a partition, thus satisfying the bi-partition
+constraints, where the positions base pair directly, rendering the
+partial sequence invalid. Moreover, applications of the sampling
+framework, can introduce additional constraints that further reduce
+the number of valid partial subsequences. However, exploiting all such
+constraints, in a complete and general way, likely causes significant
+overhead in an implementation and does not straightforwardly improve
+the theoretical bound.
+
+\section{Exploiting the bipartite nature of base pair dependency
+  graphs to reduce the complexity}\label{sec:first-improvedComplexity}
+\textcolor{red}{\textbf{TODO: discuss and merge with/replace by previous Section}}
+
+As already noted by \citet{Flamm2001}, the requirement of canonical
+base pairing induces a bi-partition of each connected component within
+the base pair dependency graph, such that the nucleotides assigned to
+the two set of nodes in the partition are restricted to values
+$\{\Ab,\Gb\}$ and $\{\Cb,\Ub\}$ respectively.
+%
+It follows that, within each connected component of the base pair
+dependency graph, the assignment of any position drastically restricts
+the domain of the remaining positions, \emph{i.e.} they may now be
+assigned to only 2 out of the original 4 nucleotides.
+\newcommand{\Parity}[1]{p_{#1}} To exploit this property, and further
+reduce the exponential factor of the complexity, we introduce the
+notion of \Def{parity} of a vertex/position, whereby vertices of
+\Def{even} (resp. \Def{odd}) polarities may only be assigned values in
+$\{\Ab,\Gb\}$ (resp. $\{\Cb,\Ub\}$). Neighbors of an odd vertex must
+be even and vice versa, so the choice of a parity for any single
+vertex caracterizes the parity of all vertices in a bipartite
+connected subgraph.
+
+Let $CC=\{cc_1,cc_2,\ldots\}$ denote the connected components of the base pair dependency graph, we consider a modified version of the cluster tree, obtained by:
 \begin{enumerate}
@@
 Also, each node in the former cluster tree introduces at most one new variable compared to its parent, so the elected node is uniquely defined by the rule stated in Step~\ref{step:election}.
-Finally, we note that the parity of a variable $v\in cc_i$ is entirely characterized by the parity $\Parity{i}$ of $v_i^\star$, and the parity of the distance between $v$ and $v_i^\star$ in the compatibility graph.
+Finally, we note that the parity of a variable $v\in cc_i$ is entirely characterized by the parity $\Parity{i}$ of $v_i^\star$, and the parity of the distance between $v$ and $v_i^\star$ in the base pair dependency graph.
 Namely, the parity of $v$ is $\Parity{i}$ if the distance between $v$ and $v^{\star}_i$ is even, and opposite to $\Parity{i}$ if odd.

### 2018-03-02 supplement.tex
Fix remarks to optimized complexity

@@
 length, $k$ is the number of structures, and
-$\hat w := \max_{v\in T} c(v)+|\chi(v)|$. Notably, $\hat w$ can not
-exceed $2(w+1)$ such that the modified algorithm does not increase the
-original complexity.
+$\hat w := \max_{v\in T} c(v)+|\chi(v)|$.
 %
-As useful corollary, one obtains the (looser) complexity
-bound $$O(nk2^{w+1}2^{cc})$$ in terms of treewidth $w$ and connected
-components $cc$.
+From this directly follows the slightly looser complexity
+bound $$O(nk2^{w+1}2^{c})$$ in terms of treewidth $w$ and the maximum
+number $c$ of connected components represented in a node of the tree
+decomposition.  Notably, $\hat w$, as well as $w+1+c$, can not exceed
+$2(w+1)$, since $c\leq w+1$. Consequently the modified algorithm does
+never increase the original complexity.
 

### 2018-03-27 RNARedPrint-Bioinfo2018.tex
Clarify potential mix-up
Minor changes

@@
     linkcolor=red!70!black,          % color of internal links (change box color with linkbordercolor)
-    citecolor=blue!50!black,        % color of links to bibliography
+    citecolor=red!70!black,        % color of links to bibliography
     filecolor=magenta,      % color of file links
-    urlcolor=green!50!black           % color of external links
+    urlcolor=red!70!black           % color of external links
 }
@@
 While heuristic in nature, this basic iteration was elected in our initial version of \ourprog{} because of its reasonable empirical behavior (choosing $\gamma=1.2$).
-
-\textcolor{red}{\bf RED ALERT! CHECK: IS THERE SOMETHING MISSING?}
 

### 2018-03-27 supplement.tex
Minor changes

--- 2018-02-28 supplement.tex
@@
 
-\section{Exploiting the bipartite nature of compatibility graphs to reduce the complexity}\label{sec:improvedComplexity}
+\section{Exploiting constraint consistency to reduce the complexity}\label{sec:improvedComplexity}
 
-As already noted by \citet{Flamm2001}, the
-requirement of canonical base pairing induces a bi-partition of each connected
-component within the compatibility graph, such that the nucleotides assigned to the two set of nodes in the partition are 
-restricted to values $\{\Ab,\Gb\}$ and $\{\Cb,\Ub\}$ respectively.
-It follows that, within each connected component of the compatibility graph, the assignment of any position drastically restricts the domain of the remaining positions, \emph{i.e.} they may now be assigned to only 2 out of the original 4 nucleotides.
-\newcommand{\Parity}[1]{p_{#1}}
-To exploit this property, and further reduce the exponential factor of the complexity, we introduce the notion of \Def{parity} of a vertex/position, whereby vertices of \Def{even} (resp.
-\Def{odd}) polarities may only be assigned values in $\{\Ab,\Gb\}$ (resp.
-$\{\Cb,\Ub\}$).
-Neighbors of an odd vertex must be even and vice versa, so the choice of a parity for any single vertex caracterizes the parity of all vertices in a bipartite connected subgraph.
+\textcolor{red}{\textbf{SW: Simpler alternative to
+    Sec.~\ref{sec:first-improvedComplexity}. Note also that the use of
+    variables in Sec.~\ref{sec:first-improvedComplexity} conflicts
+    with the (recent) specialization of constraint networks to RNA design
+    networks.}}
 
-\textcolor{red}{\textbf{SW: Introducing variables now conflicts with
-    the specialization of constraint networks to RNA design
-    networks. Luckily, there should be an even easier way to describe
-    the complexity reduction without introducing new variables; ideally let's discuss}}
+While Alg.~\ref{alg:pf} computes messages values for
+\emph{all} possible combinations of nucleotides for the positions 
+% in the separator set, 
+% yp: for all the positions in a node, right?
+in a node, 
+we observe here that many such combinations are \emph{not} required for computing all relevant
+partition functions. In particular, the algorithm can be restricted to
+consider only \Def{valid} combinations, satisfying the (hard)
+constraints induced by valid base pairs.
 
-Let $CC=\{cc_1,cc_2,\ldots\}$ denote the connected components of the compatibility graph, we consider a modified version of the cluster tree, obtained by:
-\begin{enumerate}
-\item Electing, in each connected component $cc_i$, the first position $v_i^\star$ from the component encountered from the root in a pre-order traversal of the cluster tree;\label{step:election}
-\item Introducing new variables $\mathcal{P}=\{\Parity{1},\ldots,\Parity{|CC|}\}$, representing the parity of elected vertices;
-\item Adding, to each node $x$ in the tree decomposition, the parity variables of connected components that are represented in the node, \emph{i.e.} such that at least one of the positions in the connected component is in $\chi(x)$; \label{step:addparities}
-\item Introducing, in the tree decomposition, a chain of intermediate nodes to fix cases where more than a single (parity) variable is introduced between a parent and its child.\label{step:intermediates}
-\end{enumerate}
-Note that, since connected components correspond to connected subtrees in the tree decomposition, then adding those variables preserves the properties of a tree decomposition.
+RNA design introduces constraints due to the requirement of canonical
+(aka complementary) base pairing, which can be exploited in a
+particularly simple and effective way to reduce the complexity.  
+As previously noted by \citet{Flamm2001}, the base pair complementarity induces a
+bi-partition of each connected component within the base pair
+dependency graph, such that the nucleotides assigned to the two set of
+nodes in the partition are restricted to values in $\{\Ab,\Gb\}$ and
+$\{\Cb,\Ub\}$ respectively. We call a partial sequence
+\Def{cc-valid}, iff its determined positions are consistent with such a separation
+for all determined positions of the same connected component.
+
+% Let $CC=\{cc_1$, $cc_2, \dots\}$ denote the vertex sets of the
+% connected components of the base pair dependency graph. In this graph,
+% let each component have the bi-partition $cc_i=cc^0_i\cup cc^1_i$.
+
+One can now modify Alg.~\ref{alg:pf}, on a tree decomposition
+$(T,\chi)$, such that the message values
+$\evalfor{\Message{u}{v}}{\val}$ are only computed for cc-valid
+partial sequences $\val\in\partseqs(\separator{u}{v})$. Moreover, the
+loop over $\val'\in\partseqs(\difference{u}{v})$ is restricted, such
+that $\substitute{\val}{\val'}$ are cc-valid. Analogous restrictions are 
+then implemented in the sampling algorithm Alg.~\ref{alg:sampling}.
+
+The correctness of the modified algorithm still follows from an
+induction argument: in this way, the message computation over one node
+evaluates messages from its children only at cc-valid partial
+sequences. The result of this computation is a message, which corresponds 
+to the partition function, restricted to cc-valid partial sequences. 
+Since invalid sequences have infinite energy, they do not contribute to 
+the partition function, and the restricted partition function coincides 
+with the initial one.
+
+Remarkably, this restriction drastically improves the time
+complexity. While the original algorithm enumerates all partial
+sequences $\chi(v):=\substitute{\val}{\val'}$ such that
+$\val\in\partseqs(\separator{u}{v})$ and
+$\val'\in\partseqs(\difference{u}{v})$, the modified algorithm
+only considers cc-valid sequences for $\chi(v)$ the nodes of $v$.
+Namely,
+
+For each connected component having at least a position $i$ in $\chi(v)$, 
+the algorithm considers assigning any possible nucleotide $x\in \Sigma$ to $i$ and restricts
+ the assignment of  for 
+any remaining position $j$ of the same connected component, it restricts the domain of $j$ to base-pairs that are consistent with the parity .
+
+Let $c(v)$ be the number of connected components involving positions
+in $\chi(v)$; then, the first stage enumerates $2^{c(v)}$
+combinations. For each such combination, the second stage enumerates
+only $2^{|\chi(v)|}$ partial sequences, since there remain only two
+possible nucleotide choices for each position. All message values for
+node $v$ are thus computed in time
+$\Theta(2^{c(v)+|\chi(v)|})$. Consequently, the total time complexity
+of the modified algorithm is $$O(nk2^{\hat w}),$$ where $n$ is the RNA
+length, $k$ is the number of structures, and
+$\hat w := \max_{v\in T} c(v)+|\chi(v)|$.
+%
+From this directly follows the slightly looser complexity
+bound $$O(nk2^{w+1}2^{c})$$ in terms of treewidth $w$ and the maximum
+number $c$ of connected components represented in a node of the tree
+decomposition.  Notably, $\hat w$, as well as $w+1+c$, can not exceed
+$2(w+1)$, since $c\leq w+1$. Consequently the modified algorithm does
+never increase the original complexity.
+
+Finally, we remark that even stronger time savings could be possible
+in practice, since cc-valid partial sequences can still violate
+complementarity constraints, e.g. by assigning C and A to positions in
+different sets of a partition, thus satisfying the bi-partition
+constraints, where the positions base pair directly, rendering the
+partial sequence invalid. Moreover, applications of the sampling
+framework, can introduce additional constraints that further reduce
+the number of valid partial subsequences. However, exploiting all such
+constraints, in a complete and general way, likely causes significant
+overhead in an implementation and does not straightforwardly improve
+the theoretical bound.
+
+%\section{Exploiting the bipartite nature of base pair dependency
+%  graphs to reduce the complexity}\label{sec:first-improvedComplexity}
+%\textcolor{red}{\textbf{TODO: discuss and merge with/replace by previous Section}}
+%
+%As already noted by \citet{Flamm2001}, the requirement of canonical
+%base pairing induces a bi-partition of each connected component within
+%the base pair dependency graph, such that the nucleotides assigned to
+%the two set of nodes in the partition are restricted to values
+%$\{\Ab,\Gb\}$ and $\{\Cb,\Ub\}$ respectively.
+%%
+%It follows that, within each connected component of the base pair
+%dependency graph, the assignment of any position drastically restricts
+%the domain of the remaining positions, \emph{i.e.} they may now be
+%assigned to only 2 out of the original 4 nucleotides.
+%\newcommand{\Parity}[1]{p_{#1}} To exploit this property, and further
+%reduce the exponential factor of the complexity, we introduce the
+%notion of \Def{parity} of a vertex/position, whereby vertices of
+%\Def{even} (resp. \Def{odd}) polarities may only be assigned values in
+%$\{\Ab,\Gb\}$ (resp. $\{\Cb,\Ub\}$). Neighbors of an odd vertex must
+%be even and vice versa, so the choice of a parity for any single
+%vertex caracterizes the parity of all vertices in a bipartite
+%connected subgraph.
+%
+%Let $CC=\{cc_1,cc_2,\ldots\}$ denote the connected components of the base pair dependency graph, we consider a modified version of the cluster tree, obtained by:
+%\begin{enumerate}
+%\item Electing, in each connected component $cc_i$, the first position $v_i^\star$ from the component encountered from the root in a pre-order traversal of the cluster tree;\label{step:election}
+%\item Introducing new variables $\mathcal{P}=\{\Parity{1},\ldots,\Parity{|CC|}\}$, representing the parity of elected vertices;
+%\item Adding, to each node $x$ in the tree decomposition, the parity variables of connected components that are represented in the node, \emph{i.e.} such that at least one of the positions in the connected component is in $\chi(x)$; \label{step:addparities}
+%\item Introducing, in the tree decomposition, a chain of intermediate nodes to fix cases where more than a single (parity) variable is introduced between a parent and its child.\label{step:intermediates}
+%\end{enumerate}
+%Note that, since connected components correspond to connected subtrees in the tree decomposition, then adding those variables preserves the properties of a tree decomposition.
 Also, each node in the former cluster tree introduces at most one new variable compared to its parent, so the elected node is uniquely defined by the rule stated in Step~\ref{step:election}.
-Finally, we note that the parity of a variable $v\in cc_i$ is entirely characterized by the parity $\Parity{i}$ of $v_i^\star$, and the parity of the distance between $v$ and $v_i^\star$ in the compatibility graph.
+Finally, we note that the parity of a variable $v\in cc_i$ is entirely characterized by the parity $\Parity{i}$ of $v_i^\star$, and the parity of the distance between $v$ and $v_i^\star$ in the base pair dependency graph.
 Namely, the parity of $v$ is $\Parity{i}$ if the distance between $v$ and $v^{\star}_i$ is even, and opposite to $\Parity{i}$ if odd.
 It follows that, thanks to the new variables introduced in Step~\ref{step:addparities}, the domain of each of the original variables can be reduced to two alternatives instead of the original four.
-
-The computation of Alg.~\ref{alg:pf} on the new cluster tree, considering reduced assignments for position variables and additional odd/even assignments for parity variables, has complexity in $\mathcal{O}(n\,k\,2^{w+1}\,2^{c})$, where $n$ is the RNA length, $k$ is the number of structures, $w$ is the tree-width of the unmodified cluster tree, and $c$ is the maximum number of connected components represented in a node.
+%
+%The computation of Alg.~\ref{alg:pf} on the new cluster tree, considering reduced assignments for position variables and additional odd/even assignments for parity variables, has complexity in $\mathcal{O}(n\,k\,2^{w+1}\,2^{c})$, where $n$ is the RNA length, $k$ is the number of structures, $w$ is the tree-width of the unmodified cluster tree, and $c$ is the maximum number of connected components represented in a node.
 It should be noted that $c\le \min(|CC|,w+1)$, so that the new complexity cannot be larger than the one stated in Th.~\ref{th:complexities}.

### 2018-03-29 RNARedPrint-Bioinfo2018.tex
Restated optimization

@@
 
-More complex \Def{Loop-based}
+More complex \Def{loop-based}
  energy models ---e.g.~the Turner model,
-which among others adds energy terms for special loops and dangling
-ends---are encoded as straightforward extensions. An interesting
+which among others includes energy terms for special loops and dangling
+ends---can also be encoded as straightforward extensions. An interesting
 stripped-down variant of the nearest neighbor model is the

### 2018-03-29 supplement.tex
Restated optimization

@@
 
-\textcolor{red}{\textbf{SW: Simpler alternative to
-    Sec.~\ref{sec:first-improvedComplexity}. Note also that the use of
-    variables in Sec.~\ref{sec:first-improvedComplexity} conflicts
-    with the (recent) specialization of constraint networks to RNA design
-    networks.}}
+%\textcolor{red}{\textbf{SW: Simpler alternative to
+%    Sec.~\ref{sec:first-improvedComplexity}. Note also that the use of
+%    variables in Sec.~\ref{sec:first-improvedComplexity} conflicts
+%    with the (recent) specialization of constraint networks to RNA design
+%    networks.}}
 
@@
 
-The correctness of the modified algorithm still follows from an
-induction argument: in this way, the message computation over one node
+The correctness of the modified algorithm follows from the same
+induction argument, where the message computation over one node
 evaluates messages from its children only at cc-valid partial
@@
 Since invalid sequences have infinite energy, they do not contribute to 
-the partition function, and the restricted partition function coincides 
+the partition function, and the partition function restricted to cc-valid sequences coincides 
 with the initial one.
 
-Remarkably, this restriction drastically improves the time
-complexity. While the original algorithm enumerates all partial
-sequences $\chi(v):=\substitute{\val}{\val'}$ such that
+This restriction drastically improves the time
+complexity. Indeed, for any given node $v$, the original algorithm sends message for 
+ $\chi(v):=\substitute{\val}{\val'}$ such that
 $\val\in\partseqs(\separator{u}{v})$ and
-$\val'\in\partseqs(\difference{u}{v})$, the modified algorithm
-only considers cc-valid sequences for $\chi(v)$ the nodes of $v$.
-Namely,
+$\val'\in\partseqs(\difference{u}{v})$, while the modified algorithm
+only considers cc-valid assignments for $\chi(v)$. Remark that, in any connected component $cc$, 
+assigning some nucleotide to a position reduces cc-valid assignments to (at most) 
+two alternatives for each of the $|cc|-1$ remaining positions.
+It follows that, for a node $v$ featuring positions from $\#cc(v)$ distinct connected components $\{cc_1,cc_2\ldots\}$ in the base pair dependency graph, the number of cc-valid assignments to positions in $\chi(v)$ is exactly
+$$4^{\#cc(v)}\prod_{i=1}^{\#cc(v)} 2^{|cc_i|-1} = 2^{\#cc(v)}\, 2^{|\chi(v)|} \in \Theta(2^{\#cc}\, 2^{w+1}),$$
+for a single node, where $\#cc$ is the total number of connected components in the base pair dependency graph, and $w$ is the tree-width of the tree decomposition $T$.
+Since the number of nodes in $T$ is in $\Theta(n)$, and the number of atomic energy contributions associated with $k$ structures is in $\mathcal{O}(k\,n)$, then the overall complexity grows like $\Theta(n\, k\, 2^{\#cc}\, 2^{w+1})$.
 
-For each connected component having at least a position $i$ in $\chi(v)$, 
-the algorithm considers assigning any possible nucleotide $x\in \Sigma$ to $i$ and restricts
- the assignment of  for 
-any remaining position $j$ of the same connected component, it restricts the domain of $j$ to base-pairs that are consistent with the parity .
-
-Let $c(v)$ be the number of connected components involving positions
-in $\chi(v)$; then, the first stage enumerates $2^{c(v)}$
-combinations. For each such combination, the second stage enumerates
-only $2^{|\chi(v)|}$ partial sequences, since there remain only two
-possible nucleotide choices for each position. All message values for
-node $v$ are thus computed in time
-$\Theta(2^{c(v)+|\chi(v)|})$. Consequently, the total time complexity
-of the modified algorithm is $$O(nk2^{\hat w}),$$ where $n$ is the RNA
-length, $k$ is the number of structures, and
-$\hat w := \max_{v\in T} c(v)+|\chi(v)|$.
+%For each connected component having at least a position $i$ in $\chi(v)$, 
+%the algorithm considers assigning any possible nucleotide $x\in \Sigma$ to $i$ and restricts
+% the assignment of  for 
+%any remaining position $j$ of the same connected component, it restricts the domain of $j$ to base-pairs that are consistent with the parity . 
 %
-From this directly follows the slightly looser complexity
-bound $$O(nk2^{w+1}2^{c})$$ in terms of treewidth $w$ and the maximum
-number $c$ of connected components represented in a node of the tree
-decomposition.  Notably, $\hat w$, as well as $w+1+c$, can not exceed
-$2(w+1)$, since $c\leq w+1$. Consequently the modified algorithm does
-never increase the original complexity.
+%
+%Let $c(v)$ be the number of connected components involving positions
+%in $\chi(v)$; then, the first stage enumerates $2^{c(v)}$
+%combinations. For each such combination, the second stage enumerates
+%only $2^{|\chi(v)|}$ partial sequences, since there remain only two
+%possible nucleotide choices for each position. All message values for
+%node $v$ are thus computed in time
+%$\Theta(2^{c(v)+|\chi(v)|})$. Consequently, the total time complexity
+%of the modified algorithm is $$O(nk2^{\hat w}),$$ where $n$ is the RNA
+%length, $k$ is the number of structures, and
+%$\hat w := \max_{v\in T} c(v)+|\chi(v)|$.
+%%
+%From this directly follows the slightly looser complexity
+%bound $$O(nk2^{w+1}2^{c})$$ in terms of treewidth $w$ and the maximum
+%number $c$ of connected components represented in a node of the tree
+%decomposition.  Notably, $\hat w$, as well as $w+1+c$, can not exceed
+%$2(w+1)$, since $c\leq w+1$. Consequently the modified algorithm does
+%never increase the original complexity.
 
@@
 the number of valid partial subsequences. However, exploiting all such
-constraints, in a complete and general way, likely causes significant
-overhead in an implementation and does not straightforwardly improve
-the theoretical bound.
+constraints, in a complete and general way, would likely cause significant
+implementation overhead, while not significantly improving
+the asymptotic complexity.
 
@@
 for an RNA structure $R$ is given as sum of base pair energies, where
-we consider six types of base pairs distinguishing by the bases A-U,
-C-G or G-U (symmetrically) and between stacked and non-stacked base
-pairs; here, we consider base pairs $(i,j)\in R$ \emph{stacked} iff
-$(i+1,j-1)\in R$, otherwise \emph{non-stacked}. In the \Def{stacking model},
+we consider six types of base pairs distinguishing by the bases $\Ab$-$\Ub$,
+$\Cb$-$\Gb$ or $\Gb$-$\Ub$ (symmetrically) and between stacked and non-stacked base
+pairs; here, we consider base pairs $(i,j)\in R$ \Def{stacked} iff
+$(i+1,j-1)\in R$, otherwise \Def{non-stacked}. In the \Def{stacking model},
 our features are defined by the stacks, i.e. pairs $(i,j)$ and

### 2018-04-25 RNARedPrint-Bioinfo2018.tex
what?
Revamped first paragraph of the results section

@@
 
+\usepackage[scale=.95]{inconsolata}
+
 \usepackage{xifthen}
@@
 %% Uncomment the line below for ``Final'' version
-\renewcommand{\Final}[1]{}
+%\renewcommand{\Final}[1]{}
 
@@
 \setlength{\parskip}{.2em}
+
+\newcommand{\RNAblueprint}{{\tt \bfseries\color{blue}RNABlueprint}}
 
@@
   models; this enables the generation of RNA sequences meeting
-  specific goals like expected free energies or \GCb-content. Finally,
+  specific goals like expected free energies or \GCb\%. Finally,
   we empirically study general properties of the approach and generate
@@
 additionally aiming at other features, such as specific
-\GCb-content~\citep{Reinharz2013} or the presence/absence of
+\GCb-content (\GCb\%) as previously done by~\citet{Reinharz2013} or the presence/absence of
 functionally relevant motifs (either anywhere or at specific
@@
 strategy was later shown to result in unrealistically high
-\GCb-contents in designed sequences. To address this issue,
-\Software{IncaRNAtion}~\citep{Reinharz2013} controls the \GCb-content
+\GCb\% in designed sequences. To address this issue,
+\Software{IncaRNAtion}~\citep{Reinharz2013} controls the \GCb\%
 through an adaptive sampling strategy.
@@
 approach~\citep{HoenerzuSiederdissen2013}, coupled with powerful local
-search within \Software{RNAblueprint}~\citep{Hammer2017}, solved the
+search within \RNAblueprint{}~\citep{Hammer2017}, solved the
 problem of sampling seeds from the uniform distribution for multiple
@@
 In particular, we wish to extend the flexibility of multi-structure design, leading to the following questions: \emph{How to sample, generally, from a Boltzmann distribution based on expressive energy models?
-How to enforce additional constraints, such as the \GCb-content, complex sequence constraints, or the energy of individual structures?}
+How to enforce additional constraints, such as the \GCb\%, complex sequence constraints, or the energy of individual structures?}
 
@@
 The tree is then used to compute the partition function, followed by a Boltzmann sampling of valid sequences (v).
-An adaptive scheme learns weights to achieve targeted energies and \GCb-content, leading to the production of suitable designs (vi).}
+An adaptive scheme learns weights to achieve targeted energies and \GCb\%, leading to the production of suitable designs (vi).}
 \label{fig:workflow}
@@
   sampling} strategy to control the free energies of the individual
-target structures and the \GCb-content%
+target structures and the \GCb\%%
 % (SubSection~\ref{sec:multiBoltzmann})
@@
 the energies of the single RNA structures in an RNA energy model, as
-well as sequence dependent features like \GCb-content.  Furthermore note
+well as sequence-dependent features like \GCb\%.  Furthermore note
 that $E_\R(S)$ is finite iff $S$ is valid for each structure
@@
 The flexibility of our framework supports an advanced sampling technique, named multidimensional Boltzmann sampling~\citep{Bodini2010} to (probabilistically) enforce additional constraints.
-This technique was previously used to control the \GCb-content~\citep{Waldispuehl2011,Reinharz2013} and dinucleotide content~\citep{Zhang2013} of sampled RNA sequences; it enables explicit control of the free energies $(\TargetE_1,\ldots,\TargetE_k)$ of the single targets.
+This technique was previously used to control the \GCb\%~\citep{Waldispuehl2011,Reinharz2013} and dinucleotide content~\citep{Zhang2013} of sampled RNA sequences; it enables explicit control of the free energies $(\TargetE_1,\ldots,\TargetE_k)$ of the single targets.
 %We use multidimensional Boltzmann sampling to generate sequences having predetermined targeted energies $(\TargetE_1,\ldots,\TargetE_k)$ for the $k$ structures.
 
-Multidimensional Boltzmann sampling requires the ability to \Def{sample from a weighted distribution} over the set of compatible sequences, where the probability of a sequence $S$ with free energies $(E_1,\ldots,E_k)$ for its target structures is
+Multidimensional Boltzmann sampling requires the ability to \Def{sample from a weighted distribution} over the set of valid sequences, where the probability of a sequence $S$ with free energies $(E_1,\ldots,E_k)$ for its target structures is
 $\mathbb{P}(S\mid \pmb{\pi}) = \frac{\prod_{\ell=1}^{k} \pi_i^{-E_i}}{\partfun{\pmb{\pi}}},$
@@
 $\varepsilon\ge 1/\sqrt{n}$, and $\Theta(n^{k/2})$ when
-$\varepsilon=0$. The \GCb-content of designs can also be controlled,
+$\varepsilon=0$. The \GCb\% of designs can also be controlled,
 jointly and in a similar way, as done in
@@
 
-\subsection{Targeting Turner energies and GC-content}
-We implemented the Boltzmann sampling approach (Algorithms
-\ref{alg:pf} and \ref{alg:sampling}), performing sampling for given
-target structures and weights $\pi_1,\dots,\pi_n$; moreover on top,
-multi-dimensional Boltzmann sampling (see
-Section~\ref{sec:multiBoltzmann}) to target specific energies and
-\GCb-content.  Our tool \ourprog{} evaluates energies according to the base
-pair energy model or the stacking energy model, using parameters which
-were trained to approximate Turner energies
-(Supp. Mat.~\ref{appsec:modelparameters}).
+\subsection{Targeting Turner energies and \GCb\%}
+We implemented our Boltzmann sampling strategy (Algorithms
+\ref{alg:pf} and \ref{alg:sampling}), to sample valid sequences for given
+target structures and weights $\pi_1,\dots,\pi_n$.
+Moreover, we used multi-dimensional Boltzmann sampling, as described in Section~\ref{sec:multiBoltzmann}, to target specific energies and \GCb\%.
+Our tool \ourprog{} evaluates energies according to the base
+pair energy model or a stacking energy model $E_{\mathcal{S}}$, whose parameters 
+were fitted to best approximate Turner energies (Supp. Mat.~\ref{appsec:modelparameters}).
 %
 %\paragraph{Implementation of multi-dimensional Boltzmann sampling}
-%As suggested in Section~\ref{sec:multiBoltzmann}, we heuristically optimize weights to target specific energies and GC content in an iterative procedure (based on the core sampling algorithm).
+%As suggested in Section~\ref{sec:multiBoltzmann}, we heuristically optimize weights to target specific energies and \GCb\% in an iterative procedure (based on the core sampling algorithm).
 %Specifically, energy weights are initialized as $w_\ell=e^{\frac{1}{RT}}$; the gc weight as $w_{gc}=1$.
@@
 
-To capture the realistic Turner model $E_{\mathcal{T}}$, we exploit its very good correlation (supp.
-Fig~\ref{fig:training-cor}) with our simple stacking-based model $E_{\mathcal{S}}$.
-Namely, we observe a structure-specific affine dependency between these Turner and stacking energy models, so that $E_{\mathcal{T}}(S_\ell;R_\ell) \approx \gamma_\ell\, E_{\mathcal{S};R_\ell}(S_\ell) + \delta_\ell$ for each structure $R_\ell$.
-We learn the $(\gamma_\ell,\delta_\ell)$ parameters from a set of sequences generated with homogenous weights $w=e^{\beta}$, tuning only the \GCb-content to a predetermined target frequency.
-Finally, we adjust the targeted energy of our stacking model to $E_{\mathcal{S}}^{\star} = (E_{\mathcal{T}}^{\star}- \delta_\ell)/\gamma_\ell$.
-
-To illustrate our above strategy, we sampled $n=1\,000$ sequences
-targeting ${\sf GC}\%=0.5$ and different Turner energies for the three
-structure targets of an example instance. Fig.~\ref{fig:energydist}A
-illustrates how the Turner energy distributions of the shown target
-structures can be accurately shifted to prescribed target energies;
-for comparison, we plot the respective energy distribution of
-uniformly and Boltzmann sampled sequences. Fig.~\ref{fig:energydist}B
-summarizes the targeting accuracy for the single structure energies
-over a larger set of instances from the \texttt{Modena} benchmark. We
-emphasize that only the simple stacking energies are directly targeted
-(at $\pm$10\% tolerance, expect for a few hard instances). Due to our
-linear adjustment, we finally achieve well-controlled energies in the
-realistic Turner energy model.
+To capture the realistic Turner model $E_{\mathcal{T}}$ while keeping the computational cost in check, we exploited the excellent linear fit  of $E_{\mathcal{T}}$  with the fitted stacking model $E_{\mathcal{S}}$ (supp.
+Fig~\ref{fig:training-cor}).
+Namely, we observed a structure-specific affine dependency between the Turner and stacking energy models, so that $E_{\mathcal{T}}(S;R) \approx \gamma\cdot E_{\mathcal{S}}(S;R) + \delta$ for any structure $R$ and sequence $S$.
+We learned the $(\gamma,\delta)$ parameters from a set of sequences generated with homogenous weights $w=e^{\beta}$, tuning only the \GCb\% to a predetermined value.
+Finally, we adjusted the targeted energies within our stacking model to $E_{\mathcal{S}}^{\star} = (E_{\mathcal{T}}^{\star}- \delta)/\gamma$ in order to reach, on average, the right targeted energy $E_{\mathcal{T}}^{\star}$ in the Turner model.
+
+Fig.~\ref{fig:energydist} illustrates our above strategy. Fig.~\ref{fig:energydist}A
+shows the relative distance between $E_{\mathcal{T}}$  and $E_{\mathcal{S}}$, observed for instances of the \texttt{Modena} benchmark.
+For each structure, $n=1\,000$ sequences were sampled, while targeting a ${\sf GC}\%$ of $50\%$, and reported the difference between the expected and sampled energies for both the Turner model and its approximation using a stacking model.
+Fig.~\ref{fig:energydist}B shows the Turner energy, for a single instance of the 3-structures \texttt{Modena}  benchmark, using different weights benchmark.
+These results demonstrate that our multidimensionnal Boltzmann sampling strategy can accurately and, to a large extent, independently shift the Turner energies of sampled sequences to prescribed targets.
+
+%To reduce the cost of the precomputation, we used a stacking energy model 
+%as a proxy for the Turner model. Indeed, we observed a strong linear 
+%dependency between the Turner energy and the 
+%emphasize that only the simple stacking energies are directly targeted
+%(at $\pm$10\% tolerance, except for a few hard instances). Due to our
+%linear adjustment, we finally achieve well-controlled energies in the
+%realistic Turner energy model.
 \begin{figure*}[t]
   \begin{center}
-    {\sf \bfseries A}\includegraphicstop[width=0.57\textwidth]{Figs/energy_distribution}
-    {\sf \bfseries B}\includegraphicstop[width=0.32\textwidth]{Figs/offset}
+    {\sf \bfseries A}\includegraphicstop[width=0.32\textwidth]{Figs/offset}\hspace{.5cm}
+    {\sf \bfseries B}\includegraphicstop[width=0.57\textwidth]{Figs/energy_distribution}
   \end{center}
@@
   	Targeting specific energies using multi-dimensional Boltzmann sampling. 
-(A) Turner energy distributions for three target structures (right) while targeting $(-40,-40,-20)$ and $(-20,-20,-20)$ free energy vectors.
-For comparison, we show the distribution of uniform and Boltzmann samples; respectively associated with homogenous weights $1$ and $e^\beta$.
-(B) Accuracy of targeting over all target structures of the \Software{Modena} benchmark instances \texttt{RNAtabupath} (2str), \texttt{3str} and \texttt{4str}.
+  	(A) Accuracy of targeting over all target structures of the \Software{Modena} benchmark instances \texttt{RNAtabupath} (2str), \texttt{3str} and \texttt{4str}.
 Shown is the relative deviation of targeted and achieved energies in the simple stacking model and the Turner model.
+  	(B) Turner energy distributions for three target structures (right) while targeting $(-40,-40,-20)$ and $(-20,-20,-20)$ free energy vectors.
+For comparison, we show the distribution of uniform and Boltzmann samples; respectively associated with homogenous weights $1$ (uniform) and $e^\beta$ (Boltzmann).
 %    ($\delta=0.05, \delta'=0.1, \gamma=1.1, \psi=20, n=1000$)
@@
 \subsection{Generating high-quality seeds for further optimization}
-We empirically study the capacity of \ourprog{} to improve the generation of seed sequences for subsequent local optimizations, showing an important application in biologically relevant scenario.
+We empirically study the capacity of \ourprog{} to improve the generation of seed sequences for subsequent local optimizations, showing an important application in biologically relevant scenarii.
 
@@
 energies are estimated from averaging the energy over samples at
-relatively high weights $e^{\beta}$, setting the targeted \GCb-content
+relatively high weights $e^{\beta}$, setting the targeted \GCb\%
 to 50\%. For each instance of the \Software{Modena}
@@
 energy that favor the stability of the target structures having
-moderate \GCb-content; all these properties are desirable objectives
+moderate \GCb\%; all these properties are desirable objectives
 for RNA design.  All sampled sequences are evaluated under the
@@
 function of Eq.~\eqref{eq:blueprintobjective}, to seed sequences
-uniformly sampled using \Software{RNAblueprint}~\citep{Hammer2017}
+uniformly sampled using \RNAblueprint{}~\citep{Hammer2017}
 (Table~\ref{tab:benchmark-results}, seeds). Our first analysis reveals
@@
 Moreover, for each generated seed sequence, we optimized the objective function $(\ref{eq:blueprintobjective})$ through an adaptive  greedy walk consisting of 500 steps.
-At each step, we resampled (uniformly random) the positions of a randomly selected component in the base pair dependency graph, and accepted the modification only if it resulted in a gain, as described in \Software{RNAblueprint}~\citep{Hammer2017}.
+At each step, we resampled (uniformly random) the positions of a randomly selected component in the base pair dependency graph, and accepted the modification only if it resulted in a gain, as described in \RNAblueprint{}.
 %
@@
 Table~\ref{tab:benchmark-results}, optimized). Notably, our subsequent
-optimization runs, which are directly inherited from the uniformly
-sampling RNABlueprint software, partially level the advantages of
+optimization runs, which are directly inherited from the uniform
+sampling \RNAblueprint{} software, partially level the advantages of
 Boltzmann sampling. In future work, we hope to improve this aspect by
@@
 
-We turn to the surjectivity of $\Psi$, i.e. the existence of a preimage for each element $S\in \IS{G}$. Let us consider the function $f$ defined as
-\begin{align}
- &\forall v_1\in V_1,v_2\in V_2:\notag\\& f(v_1) = \begin{cases} \Ab & \text{if }v_1\in S\\ \Gb & \text{if }v_1\notin S\end{cases} \text{\quad and\quad} 
- f(v_2) = \begin{cases} \Cb & \text{if }v_2\in S\\ \Ub & \text{if }v_2\notin S\end{cases}
-\end{align}
+We turn to the surjectivity of $\Psi$, i.e. the existence of a preimage for each element $S\in \IS{G}$. Let us consider the function $f$ defined for any $(v_1,v_2)\in V_1\times V_2$ as
+\begin{align*}
+f(v_1) &= \begin{cases} \Ab & \text{if }v_1\in S\\ \Gb & \text{if }v_1\notin S\end{cases} &
+\text{\quad and\quad}&& 
+ f(v_2) &= \begin{cases} \Cb & \text{if }v_2\in S\\ \Ub & \text{if }v_2\notin S.\end{cases}
+\end{align*}
 One easily verifies that $\Psi(f) = S$.
@@
 in the field of RNA sequence design by enforcing additional
-constraints, like the \GCb-content, while controlling the energy of
+constraints, like the \GCb\%, while controlling the energy of
 multiple target structures. Thus it presents a major advance over
@@
 %
-We thank Leonid Chindelevitch for suggesting using the stable sets analogy to optimize our FPT algorithm, and Arie Koster for practical recommendations
-for computing tree decompositions.
+We thank Leonid Chindelevitch for suggesting a drastic optimization of our FPT algorithm based on stable sets combinatorics, and Arie Koster for practical recommendations on tree decompositions.
 

### 2018-04-25 RNARedPrint-Bioinfo2018.tex
Mess with targeting figure and energy notation

@@
 
-\newcommand{\ourprog}{\Software{RNARedPrint}}
-
 \newcommand{\evalfor}[2]{#1\llbracket{}#2\rrbracket{}}
@@
 
-\newcommand{\RNAblueprint}{{\tt \bfseries\color{blue}RNABlueprint}}
+\newcommand{\RNAblueprint}{{\tt \bfseries{} RNA\textcolor{blue!70!black}{Blue}print}}
+\newcommand{\ourprog}{{\tt \bfseries{}RNA\textcolor{red!70!black}{Red}Print}}
+
+\newcommand{\Energy}[2]{\textcolor{green!70!black}{E(#2;#1)}}
+
+\newcommand{\EnergyTurner}{E_{\text{T}}}
+\newcommand{\EnergyStacking}{E_{\text{st}}}
 
@@
 %
-We define the energy of a sequence $S$ based on the set of structures
-$\R$ as $E_\R(S) \in \mathbb{R}\cup\{\infty\}$. In
-our setting, the energy $E_\R(S)$ is additively composed of
-the energies of the single RNA structures in an RNA energy model, as
-well as sequence-dependent features like \GCb\%.  Furthermore note
-that $E_\R(S)$ is finite iff $S$ is valid for each structure
-$R_1,\dots,R_k$.
+The \Def{energy} of a sequence $S$ is defined by an \Def{energy
+  function} $E$---defined for set of structures $\R$ and mapping
+sequences to $\mathbb{R}\cup\{\infty\}$---as
+$\Energy{\R}{S} \in \mathbb{R}\cup\{\infty\}$. In our setting, the
+energy $\Energy{\R}{S}$ is typically chosen as a linear combination of
+the energies of the single RNA structures (in some RNA energy model
+that defines $E(S,R) \in \mathbb{R}\cup\{\infty\}$ for each
+sequence $S$ and structure
+$R$), as well as sequence-dependent features like \GCb\%. Formally, we
+will require that
+$\Energy{\R}{S}$ can be expressed as sum over functions values, where
+each function depends on the nucleotides at certain positions of
+$S$ (see below). Furthermore note that
+$\Energy{\R}{S}$ is finite iff
+$S$ is valid for each structure $R_1,\dots,R_k$.
 
@@
     \label{eq:mainproblem}
-    \partfun{E_\R} = \sum_{S\in\S_n} \exp(-\beta E_\R(S)),
+    \partfun{E,\R} = \sum_{S\in\S_n} \exp(-\beta \Energy{\R}{S}),
   \end{equation}
@@
   As we elaborate in subsequent sections, our approach relies on
-  breaking down the energy function $E_\R(S)$ into additive
+  breaking down the energy function $\Energy{\R}{S}$ into additive
   components, each depending on only few sequence positions. Given
-  $\R$, we express $E_\R(S)$ as the sum of energy contributions $f(S)$
+  $\R$, we express $\Energy{\R}{S}$ as the sum of energy contributions $f(S)$
   over a set $\F$ (of functions $f:\S_n\to\real$),
-  s.t.~$E_\R(S)=\sum_{f\in\F} f(S)$. This captures realistic RNA
+  s.t.~$\Energy{\R}{S}=\sum_{f\in\F} f(S)$. This captures realistic RNA
   energy models---including nearest neighbor models,
@@
 Our tool \ourprog{} evaluates energies according to the base
-pair energy model or a stacking energy model $E_{\mathcal{S}}$, whose parameters 
+pair energy model or a stacking energy model $\EnergyStacking$, whose parameters 
 were fitted to best approximate Turner energies (Supp. Mat.~\ref{appsec:modelparameters}).
@@
 
-To capture the realistic Turner model $E_{\mathcal{T}}$ while keeping the computational cost in check, we exploited the excellent linear fit  of $E_{\mathcal{T}}$  with the fitted stacking model $E_{\mathcal{S}}$ (supp.
+To capture the realistic Turner model $\EnergyTurner$ while keeping the computational cost in check, we exploited the excellent linear fit  of $\EnergyTurner$  with the fitted stacking model $\EnergyStacking$ (supp.
 Fig~\ref{fig:training-cor}).
-Namely, we observed a structure-specific affine dependency between the Turner and stacking energy models, so that $E_{\mathcal{T}}(S;R) \approx \gamma\cdot E_{\mathcal{S}}(S;R) + \delta$ for any structure $R$ and sequence $S$.
+Namely, we observed a structure-specific affine dependency between the Turner and stacking energy models, so that $\EnergyTurner(S;R) \approx \gamma\cdot \EnergyStacking(S;R) + \delta$ for any structure $R$ and sequence $S$.
 We learned the $(\gamma,\delta)$ parameters from a set of sequences generated with homogenous weights $w=e^{\beta}$, tuning only the \GCb\% to a predetermined value.
-Finally, we adjusted the targeted energies within our stacking model to $E_{\mathcal{S}}^{\star} = (E_{\mathcal{T}}^{\star}- \delta)/\gamma$ in order to reach, on average, the right targeted energy $E_{\mathcal{T}}^{\star}$ in the Turner model.
+Finally, we adjusted the targeted energies within our stacking model to $\EnergyStacking^{\star} = (\EnergyTurner^{\star}- \delta)/\gamma$ in order to reach, on average, the right targeted energy $\EnergyTurner^{\star}$ in the Turner model.
 
 Fig.~\ref{fig:energydist} illustrates our above strategy. Fig.~\ref{fig:energydist}A
-shows the relative distance between $E_{\mathcal{T}}$  and $E_{\mathcal{S}}$, observed for instances of the \texttt{Modena} benchmark.
+shows the relative distance between $\EnergyTurner$  and $\EnergyStacking$, observed for instances of the \texttt{Modena} benchmark.
 For each structure, $n=1\,000$ sequences were sampled, while targeting a ${\sf GC}\%$ of $50\%$, and reported the difference between the expected and sampled energies for both the Turner model and its approximation using a stacking model.
@@
   \item More general energy  models $\to$ Turner. (Pseudoknots)
-  For this description, we assume that all input structures are non-crossing, while the framework would otherwise support crossing input structures as well as according energy models (e.g.
-the Hotknots model~\citep{Ren2005}) by considering special pseudoknot loops.
+  For this description, we assume that all input structures are non-crossing, while the framework would otherwise support crossing input structures as well as according energy models, e.g.{} the Hotknots model of~\citet{Ren2005}, by considering special pseudoknot loops.
 \end{enumerate}\label{lbl:theend}

### 2018-04-30 RNARedPrint-Bioinfo2018.tex
Edit Fig 2 caption

@@
   \caption{%
-  	Targeting specific energies using multi-dimensional Boltzmann sampling. 
-  	(A) Accuracy of targeting over all target structures of the \Software{Modena} benchmark instances \texttt{RNAtabupath} (2str), \texttt{3str} and \texttt{4str}.
-Shown is the relative deviation of targeted and achieved energies in the simple stacking model and the Turner model.
-  	(B) Turner energy distributions for three target structures (right) while targeting $(-40,-40,-20)$ and $(-20,-20,-20)$ free energy vectors.
-For comparison, we show the distribution of uniform and Boltzmann samples; respectively associated with homogenous weights $1$ (uniform) and $e^\beta$ (Boltzmann).
+    Targeting specific energies using multi-dimensional Boltzmann
+    sampling.  (A) Accuracy of targeting over all target structures of
+    the \Software{Modena} benchmark instances \texttt{RNAtabupath}
+    (2str), \texttt{3str} and \texttt{4str}. Shown is the relative
+    deviation of targeted and achieved energies in the simple stacking
+    model and the Turner model.  (B) Turner energy distributions for
+    three target structures S1, S2 and S3. We show the results of
+    targeting the respective free energies $(-40,-40,-20)$ and
+    $(-20,-20,-20)$ for the target structures (annotated in red and
+    blue, resepectively)---this demonstrates the
+    effectivity of our adaptive multi-dimensional Boltzmann sampling
+    procedure; moreover, for comparison, the distributions for uniform
+    (U) and Boltzmann (B) samples---respectively associated with
+    homogenous weights $1$ and $e^\beta$.
 %    ($\delta=0.05, \delta'=0.1, \gamma=1.1, \psi=20, n=1000$)

### 2018-04-30 RNARedPrint-Bioinfo2018.tex
Minor edits in Sec. 4; add todo note

@@
 \usepackage{xcolor}
+
+\usepackage{todonotes}
 
@@
 We implemented our Boltzmann sampling strategy (Algorithms
-\ref{alg:pf} and \ref{alg:sampling}), to sample valid sequences for given
-target structures and weights $\pi_1,\dots,\pi_n$.
-Moreover, we used multi-dimensional Boltzmann sampling, as described in Section~\ref{sec:multiBoltzmann}, to target specific energies and \GCb\%.
-Our tool \ourprog{} evaluates energies according to the base
-pair energy model or a stacking energy model $\EnergyStacking$, whose parameters 
-were fitted to best approximate Turner energies (Supp. Mat.~\ref{appsec:modelparameters}).
+\ref{alg:pf} and \ref{alg:sampling}), to sample valid sequences for
+given target structures and weights $\pi_1,\dots,\pi_n$.  Moreover, we
+used multi-dimensional Boltzmann sampling, as described in
+Section~\ref{sec:multiBoltzmann}, to target specific energies and
+\GCb\%.  Our tool \ourprog{} evaluates energies according to the
+stacking energy model $\EnergyStacking$, whose parameters were fitted
+to best approximate Turner energies. As well, we implemented and
+fitted a base pair energy model for \ourprog{}, which was not studied
+for its targeting performance (both models:
+Supp. Mat.~\ref{appsec:modelparameters}).
 %
@@
 
-To capture the realistic Turner model $\EnergyTurner$ while keeping the computational cost in check, we exploited the excellent linear fit  of $\EnergyTurner$  with the fitted stacking model $\EnergyStacking$ (supp.
-Fig~\ref{fig:training-cor}).
-Namely, we observed a structure-specific affine dependency between the Turner and stacking energy models, so that $\EnergyTurner(S;R) \approx \gamma\cdot \EnergyStacking(S;R) + \delta$ for any structure $R$ and sequence $S$.
-We learned the $(\gamma,\delta)$ parameters from a set of sequences generated with homogenous weights $w=e^{\beta}$, tuning only the \GCb\% to a predetermined value.
-Finally, we adjusted the targeted energies within our stacking model to $\EnergyStacking^{\star} = (\EnergyTurner^{\star}- \delta)/\gamma$ in order to reach, on average, the right targeted energy $\EnergyTurner^{\star}$ in the Turner model.
-
-Fig.~\ref{fig:energydist} illustrates our above strategy. Fig.~\ref{fig:energydist}A
-shows the relative distance between $\EnergyTurner$  and $\EnergyStacking$, observed for instances of the \texttt{Modena} benchmark.
-For each structure, $n=1\,000$ sequences were sampled, while targeting a ${\sf GC}\%$ of $50\%$, and reported the difference between the expected and sampled energies for both the Turner model and its approximation using a stacking model.
-Fig.~\ref{fig:energydist}B shows the Turner energy, for a single instance of the 3-structures \texttt{Modena}  benchmark, using different weights benchmark.
-These results demonstrate that our multidimensionnal Boltzmann sampling strategy can accurately and, to a large extent, independently shift the Turner energies of sampled sequences to prescribed targets.
+To capture the realistic Turner model $\EnergyTurner$ while keeping
+the computational cost in check, we exploited the close correlation of
+$\EnergyTurner$ and the fitted stacking model $\EnergyStacking$
+(Supp. Fig~\ref{fig:training-cor}). More precisely, we observed a
+structure-specific affine dependency between the Turner and stacking
+energy models, so that
+$\EnergyTurner(S;R) \approx \gamma\cdot \EnergyStacking(S;R) + \delta$
+for any structure $R$ and sequence $S$. We learned the
+$(\gamma,\delta)$ parameters from a set of sequences generated with
+homogenous weights $w=e^{\beta}$, tuning only the \GCb\% to a
+predetermined value.  Finally, we adjusted the targeted energies
+within our stacking model to
+$\EnergyStacking^{\star} = (\EnergyTurner^{\star}- \delta)/\gamma$ in
+order to reach, on average, the targeted energy
+$\EnergyTurner^{\star}$ in the Turner model.
+
+Fig.~\ref{fig:energydist} illustrates the described
+strategy. \todo{Update Figure 2A}Fig.~\ref{fig:energydist}A
+illustrates our capability to target Turner energies based on the
+stacking energy model. It shows the relative distance between
+$\EnergyTurner$ and $\EnergyStacking$, observed for instances of the
+\texttt{Modena} benchmark. For each structure, $n=1\,000$ sequences
+were sampled, while targeting a ${\sf GC}\%$ of $50\%$, and reported
+the difference between the expected and sampled energies for both the
+Turner model and its approximation using a stacking model.  For an
+example instance of the \texttt{Modena} benchmark \texttt{3str},
+Fig.~\ref{fig:energydist}B shows the Turner energy distributions of
+the single structures as they result from sampling with different weight
+parameters. The figure illustrates how our multidimensionnal Boltzmann
+sampling strategy can accurately and, to a large extent, independently
+shift the Turner energies of sampled sequences to prescribed targets.
 

### 2018-04-30 RNARedPrint-Bioinfo2018.tex
New Figure 2A and corresponding updates to the caption and text

@@
 Fig.~\ref{fig:energydist} illustrates the described
-strategy. \todo{Update Figure 2A}Fig.~\ref{fig:energydist}A
+strategy. Fig.~\ref{fig:energydist}A
 illustrates our capability to target Turner energies based on the
-stacking energy model. It shows the relative distance between
+stacking energy model. It shows the deviation between
 $\EnergyTurner$ and $\EnergyStacking$, observed for instances of the
@@
   \begin{center}
-    {\sf \bfseries A}\includegraphicstop[width=0.32\textwidth]{Figs/offset}\hspace{.5cm}
+    {\sf \bfseries A}\includegraphicstop[width=0.32\textwidth]{Figs/offset2}\hspace{.5cm}
     {\sf \bfseries B}\includegraphicstop[width=0.57\textwidth]{Figs/energy_distribution}
@@
     the \Software{Modena} benchmark instances \texttt{RNAtabupath}
-    (2str), \texttt{3str} and \texttt{4str}. Shown is the relative
-    deviation of targeted and achieved energies in the simple stacking
-    model and the Turner model.  (B) Turner energy distributions for
+    (2str), \texttt{3str} and \texttt{4str}. 
+    For the stacking and the Turner energy model, the relative offsets between the targeted and 
+    achieved energies were determined. We show the deviations of these offsets between both models.
+    (B) Turner energy distributions for
     three target structures S1, S2 and S3. We show the results of
@@
 Using our strategy, we generated at least 1\,000 seeds per instance of
-the subsets \texttt{RNAtabupath} (2str), \texttt{3str}, and \texttt{4str}, of
-the \Software{Modena} benchmark set with 2,3, and 4 structures. The
-results are compared, in terms of their value for the objective
-function of Eq.~\eqref{eq:blueprintobjective}, to seed sequences
-uniformly sampled using \RNAblueprint{}~\citep{Hammer2017}
+the subsets \texttt{RNAtabupath} (2str), \texttt{3str}, and
+\texttt{4str}, of the \Software{Modena} benchmark set with 2,3, and 4
+structures. The results are compared, in terms of their value for the
+objective function of Eq.~\eqref{eq:blueprintobjective}, to seed
+sequences uniformly sampled using \RNAblueprint{}~\citep{Hammer2017}
 (Table~\ref{tab:benchmark-results}, seeds). Our first analysis reveals
-that Boltzmann-sampled sequences constitute better seeds, associated
-with better values ($\sim\! 69\%$ improvement on average), compared to
-uniform seeds for \emph{all instances} of our benchmark
-(Supp. Mat.~\ref{sec:validity}).
+that Boltzmann-sampled sequences constitute better seeds
+($\sim\! 69\%$ improvement on average) compared to uniform seeds for
+\emph{all instances} of our benchmark (Supp. Mat.~\ref{sec:validity}).
 

### 2018-05-04 RNARedPrint-Bioinfo2018.tex
Add PK targeting figure; replace deviation by linear fit figure; move pk-free figure to supplemental material

@@
 i) \emph{Enacting an exact control over the energies would  be technically hard and costly.} Indeed, controlling the energies through dynamic programming would require explicit convolution products, generalizing~\citet{Cupal1996}, inducing additional $\Theta(n^{2k})$ time and $\Theta(n^k)$ space overheads;
-ii) \emph{Induced distributions are typically concentrated.} Intuitively, unless sequences are fully constrained individual energy terms are independent enough so that their sum is concentrated around its mean -- the targeted energy (cf Fig.~\ref{fig:energydist}).
+ii) \emph{Induced distributions are typically concentrated.} Intuitively, unless sequences are fully constrained individual energy terms are independent enough so that their sum is concentrated around its mean -- the targeted energy (cf Fig.~\ref{fig:energydist-pk}).
 %Indeed, as shown in Fig.~\ref{fig:shifting_mean}, empirical joint distributions in the energies have good fits towards Normal multidimensional distributions with linear (co-)variances.
@@
 
-Fig.~\ref{fig:energydist} illustrates the described
-strategy. Fig.~\ref{fig:energydist}A
-illustrates our capability to target Turner energies based on the
-stacking energy model. It shows the deviation between
-$\EnergyTurner$ and $\EnergyStacking$, observed for instances of the
-\texttt{Modena} benchmark. For each structure, $n=1\,000$ sequences
-were sampled, while targeting a ${\sf GC}\%$ of $50\%$, and reported
-the difference between the expected and sampled energies for both the
-Turner model and its approximation using a stacking model.  For an
-example instance of the \texttt{Modena} benchmark \texttt{3str},
-Fig.~\ref{fig:energydist}B shows the Turner energy distributions of
-the single structures as they result from sampling with different weight
-parameters. The figure illustrates how our multidimensionnal Boltzmann
-sampling strategy can accurately and, to a large extent, independently
-shift the Turner energies of sampled sequences to prescribed targets.
+Fig.~\ref{fig:energydist-pk} illustrates the described strategy. For
+the two target structures of Fig.~\ref{fig:energydist-pk}B,
+Fig.~\ref{fig:energydist-pk}A shows the good fit between Turner
+energies based and energies in the stacking energy model. For the
+shown fits we sampled $n=1\,000$ sequences, targeting a ${\sf GC}\%$
+of $50\%$.
+%
+For an example instance of the \texttt{Modena} benchmark with two
+pseudoknotted target structures, Fig.~\ref{fig:energydist-pk}B shows
+the Turner energy distributions of the single structures as they
+result from sampling with different weight parameters. The figure
+illustrates how our multidimensionnal Boltzmann sampling strategy can,
+to a large extent, independently shift the Turner energies of sampled
+sequences towards prescribed targets. See Supp. Mat.,
+Fig.~\ref{fig:energydist} for a further example with three
+pseudoknot-free target structures.
 
@@
   \begin{center}
-    {\sf \bfseries A}\includegraphicstop[width=0.32\textwidth]{Figs/offset2}\hspace{.5cm}
-    {\sf \bfseries B}\includegraphicstop[width=0.57\textwidth]{Figs/energy_distribution}
+    {\sf \bfseries A}\ \includegraphicstop[width=0.38\textwidth]{Figs/regression-pk}\hfill
+    {\sf \bfseries B}\ \includegraphicstop[width=0.55\textwidth]{Figs/energy_distribution-pk}
   \end{center}
   \caption{%
-    Targeting specific energies using multi-dimensional Boltzmann
-    sampling.  (A) Accuracy of targeting over all target structures of
-    the \Software{Modena} benchmark instances \texttt{RNAtabupath}
-    (2str), \texttt{3str} and \texttt{4str}. 
-    For the stacking and the Turner energy model, the relative offsets between the targeted and 
-    achieved energies were determined. We show the deviations of these offsets between both models.
+    Targeting specific energies for pseudoknotted structures using multi-dimensional Boltzmann
+    sampling.  (A) Linear fits between the energies in the stacking model to the Turner model for
+    sampled sequences; the good match enables more efficient targeting of Turner energies based on
+    targeting stacking model energies.
+    We show the fits for both target structures S1 and S2 of subfigure B.
     (B) Turner energy distributions for
-    three target structures S1, S2 and S3. We show the results of
-    targeting the respective free energies $(-40,-40,-20)$ and
-    $(-20,-20,-20)$ for the target structures (annotated in red and
-    blue, resepectively)---this demonstrates the
+    two target structures S1 and S2. We show the results of
+    targeting the respective free energies $(-20,-10)$, $(-10,-20)$, and $(-15,-15)$
+    for the target structures (annotated in red,
+    yellow, and blue resepectively)---this demonstrates the
     effectivity of our adaptive multi-dimensional Boltzmann sampling
@@
   }
-  \label{fig:energydist}
+  \label{fig:energydist-pk}
 \end{figure*}

### 2018-05-04 supplement.tex
Add PK targeting figure; replace deviation by linear fit figure; move pk-free figure to supplemental material

@@
 \appendix
-{\centering \relsize{+4}Supplementary material\\%
+{\centering \relsize{+4}\textbf{Supplementary Material}\\ Fixed-Parameter Tractable Sampling\\ for RNA Design with Multiple Target Structures\\%
 %{\relsize{-1}\BackupTitle}\\
@@
 \relsize{+1}
+
+\medskip
+\begin{center}
+\large Stefan Hammer, Yann Ponty, Wei Wang and Sebastian Will
+\end{center}
 \section{Approximate counting and random generation}
@@
 
+\section{Illustrating multi-dimensional Boltzmann sampling for three pseudoknot-free target structures}
+
+In analogy to Fig.~\ref{fig:energydist-pk} of the main text,
+Fig.~\ref{fig:energydist} illustrates our multi-dimensional sampling
+strategy for three pseudoknot-free target structures (example instance
+from the \texttt{Modena} benchmark). The figure shows Turner energy
+distributions of single structures from sampling with different weight
+parameters, where weights are trained to target specific energies. For
+comparison, the figure shows distributions from uniform and Boltzmann
+sampling at respective homogenous weights $1$ and $e^\beta$.
+
+\begin{figure*}[t]
+  \begin{center}
+    %{\sf \bfseries A}\includegraphicstop[width=0.32\textwidth]{Figs/offset2}\hspace{.5cm}
+    %{\sf \bfseries B}
+    \includegraphicstop[width=0.8\textwidth]{Figs/energy_distribution}
+  \end{center}
+  \caption{%
+    Targeting specific energies using multi-dimensional Boltzmann
+    sampling. We show the Turner energy distributions for
+    three target structures S1, S2 and S3. By
+    targeting the respective free energies $(-40,-40,-20)$ and
+    $(-20,-20,-20)$ for the target structures (annotated in red and
+    blue, resepectively), we demonstrate the
+    effectivity of our adaptive multi-dimensional Boltzmann sampling
+    procedure; moreover, for comparison, the distributions for uniform
+    (U) and Boltzmann (B) samples---respectively associated with
+    homogenous weights $1$ and $e^\beta$.
+%    ($\delta=0.05, \delta'=0.1, \gamma=1.1, \psi=20, n=1000$)
+  }
+  \label{fig:energydist}
+\end{figure*}
+
 \section{Parameters for the base pair and the stacking energy model}

### 2018-05-04 RNARedPrint-Bioinfo2018.tex
Dump green energy
Check/edit use of GC%; omit article like in '_the_ GC%'

@@
 
-\newcommand{\Energy}[2]{\textcolor{green!70!black}{E(#2;#1)}}
+\newcommand{\Energy}[2]{E(#2;#1)}
 
@@
   models; this enables the generation of RNA sequences meeting
-  specific goals like expected free energies or \GCb\%. Finally,
+  specific goals like expected free energies or \GCb-content. Finally,
   we empirically study general properties of the approach and generate
@@
 \GCb\% in designed sequences. To address this issue,
-\Software{IncaRNAtion}~\citep{Reinharz2013} controls the \GCb\%
+\Software{IncaRNAtion}~\citep{Reinharz2013} controls \GCb\%
 through an adaptive sampling strategy.
@@
 In particular, we wish to extend the flexibility of multi-structure design, leading to the following questions: \emph{How to sample, generally, from a Boltzmann distribution based on expressive energy models?
-How to enforce additional constraints, such as the \GCb\%, complex sequence constraints, or the energy of individual structures?}
+How to enforce additional constraints, such as \GCb\%, complex sequence constraints, or the energy of individual structures?}
 
@@
   sampling} strategy to control the free energies of the individual
-target structures and the \GCb\%%
+target structures and \GCb\%%
 % (SubSection~\ref{sec:multiBoltzmann})
@@
 The flexibility of our framework supports an advanced sampling technique, named multidimensional Boltzmann sampling~\citep{Bodini2010} to (probabilistically) enforce additional constraints.
-This technique was previously used to control the \GCb\%~\citep{Waldispuehl2011,Reinharz2013} and dinucleotide content~\citep{Zhang2013} of sampled RNA sequences; it enables explicit control of the free energies $(\TargetE_1,\ldots,\TargetE_k)$ of the single targets.
+This technique was previously used to control \GCb\%~\citep{Waldispuehl2011,Reinharz2013} and dinucleotide content~\citep{Zhang2013} of sampled RNA sequences; it enables explicit control of the free energies $(\TargetE_1,\ldots,\TargetE_k)$ of the single targets.
 %We use multidimensional Boltzmann sampling to generate sequences having predetermined targeted energies $(\TargetE_1,\ldots,\TargetE_k)$ for the $k$ structures.
@@
 $(\gamma,\delta)$ parameters from a set of sequences generated with
-homogenous weights $w=e^{\beta}$, tuning only the \GCb\% to a
+homogenous weights $w=e^{\beta}$, tuning only \GCb\% to a
 predetermined value.  Finally, we adjusted the targeted energies
@@
 in the field of RNA sequence design by enforcing additional
-constraints, like the \GCb\%, while controlling the energy of
+constraints, like \GCb\%, while controlling the energy of
 multiple target structures. Thus it presents a major advance over

### 2018-05-15 RNARedPrint-Bioinfo2018.tex
Highlight notes by macro Final (in red)
Update on Overleaf.
fix centering figure

@@
 
-\usepackage[scale=.95]{inconsolata}
+\usepackage{inconsolata}
 
@@
 
-\newcommand{\Final}[1]{#1}
+\newcommand{\Final}[1]{\begingroup\color{red!70!black}#1\endgroup}
 %% Uncomment the line below for ``Final'' version
@@
 \begin{figure*}[t]
-{\centering\includegraphics[width=.8\textwidth]{Figs/Workflow}\\}
+\begin{center}
+    \includegraphics[width=.8\textwidth]{Workflow.pdf}
+\end{center}
 \caption{General outline of \ourprog{} for base pair-based energy models.
@@
              & 3str & 18.09 ($\pm$3.98) & 30.49 ($\pm$5.41) & 71\%\\
-             & 4str & 19.94 ($\pm$3.84) & 32.29 ($\pm$5.24) & 63\%\\\midrule
+             & 4str & 19.94 ($\pm$3.84) & 32.29 ($\pm$5.24) & 63\%\\
+             & PK60 & 19.94 ($\pm$3.84) & 32.29 ($\pm$5.24) & 63\%\\
+             & PK80 & 19.94 ($\pm$3.84) & 32.29 ($\pm$5.24) & 63\%\\
+             & 4str & 19.94 ($\pm$3.84) & 32.29 ($\pm$5.24) & 63\%\\
+             \midrule
 
@@
              & 3str & 5.08 ($\pm$1.10) & 7.04 ($\pm$1.52) & 31\%\\
-             & 4str & 8.77($\pm$1.48) & 13.13 ($\pm$2.13) & 37\% \\ \bottomrule
+             & 4str & 8.77($\pm$1.48) & 13.13 ($\pm$2.13) & 37\% \\
+             \bottomrule
 \end{tabular}\\[1em]

### 2018-05-15 RNARedPrint-Bioinfo2018.tex
Update on Overleaf.

@@
              & PK80 & 19.94 ($\pm$3.84) & 32.29 ($\pm$5.24) & 63\%\\
-             & 4str & 19.94 ($\pm$3.84) & 32.29 ($\pm$5.24) & 63\%\\
+             & LE80 & 19.94 ($\pm$3.84) & 32.29 ($\pm$5.24) & 63\%\\
              \midrule

### 2018-05-15 RNARedPrint-Bioinfo2018.tex
Not sure what I did...

--- 2018-05-04 RNARedPrint-Bioinfo2018.tex
@@
     colorlinks=true,       % false: boxed links; true: colored links
-    linkcolor=red!70!black,          % color of internal links (change box color with linkbordercolor)
-    citecolor=red!70!black,        % color of links to bibliography
+%    linkcolor=red!70!black,          % color of internal links (change box color with linkbordercolor)
+%    citecolor=red!70!black,        % color of links to bibliography
+    linkcolor=black,          % color of internal links (change box color with linkbordercolor)
+    citecolor=black,        % color of links to bibliography
     filecolor=magenta,      % color of file links
@@
 %% Uncomment the line below for ``Final'' version
-%\renewcommand{\Final}[1]{}
+\renewcommand{\Final}[1]{}
 
@@
 
-\newcommand{\RNAblueprint}{{\tt \bfseries{} RNA\textcolor{blue!70!black}{Blue}print}}
-\newcommand{\ourprog}{{\tt \bfseries{}RNA\textcolor{red!70!black}{Red}Print}}
-
-\newcommand{\Energy}[2]{E(#2;#1)}
+\newcommand{\RNAblueprint}{{\tt \bfseries{}\color{black!75} RNA\textcolor{blue!70!black}{Blue}Print}}
+\newcommand{\ourprog}{{\tt \bfseries{}\color{black!75}RNA\textcolor{red!70!black}{Red}Print}}
+
+\newcommand{\Energy}[2]{\textcolor{green!70!black}{E(#2;#1)}}
 
@@
 \GCb\% in designed sequences. To address this issue,
-\Software{IncaRNAtion}~\citep{Reinharz2013} controls \GCb\%
+\Software{IncaRNAtion}~\citep{Reinharz2013} controls the \GCb\%
 through an adaptive sampling strategy.
@@
 In particular, we wish to extend the flexibility of multi-structure design, leading to the following questions: \emph{How to sample, generally, from a Boltzmann distribution based on expressive energy models?
-How to enforce additional constraints, such as \GCb\%, complex sequence constraints, or the energy of individual structures?}
+How to enforce additional constraints, such as the \GCb\%, complex sequence constraints, or the energy of individual structures?}
 
@@
   sampling} strategy to control the free energies of the individual
-target structures and \GCb\%%
+target structures and the \GCb\%%
 % (SubSection~\ref{sec:multiBoltzmann})
@@
 The flexibility of our framework supports an advanced sampling technique, named multidimensional Boltzmann sampling~\citep{Bodini2010} to (probabilistically) enforce additional constraints.
-This technique was previously used to control \GCb\%~\citep{Waldispuehl2011,Reinharz2013} and dinucleotide content~\citep{Zhang2013} of sampled RNA sequences; it enables explicit control of the free energies $(\TargetE_1,\ldots,\TargetE_k)$ of the single targets.
+This technique was previously used to control the \GCb\%~\citep{Waldispuehl2011,Reinharz2013} and dinucleotide content~\citep{Zhang2013} of sampled RNA sequences; it enables explicit control of the free energies $(\TargetE_1,\ldots,\TargetE_k)$ of the single targets.
 %We use multidimensional Boltzmann sampling to generate sequences having predetermined targeted energies $(\TargetE_1,\ldots,\TargetE_k)$ for the $k$ structures.
@@
 i) \emph{Enacting an exact control over the energies would  be technically hard and costly.} Indeed, controlling the energies through dynamic programming would require explicit convolution products, generalizing~\citet{Cupal1996}, inducing additional $\Theta(n^{2k})$ time and $\Theta(n^k)$ space overheads;
-ii) \emph{Induced distributions are typically concentrated.} Intuitively, unless sequences are fully constrained individual energy terms are independent enough so that their sum is concentrated around its mean -- the targeted energy (cf Fig.~\ref{fig:energydist-pk}).
+ii) \emph{Induced distributions are typically concentrated.} Intuitively, unless sequences are fully constrained individual energy terms are independent enough so that their sum is concentrated around its mean -- the targeted energy (cf Fig.~\ref{fig:energydist}).
 %Indeed, as shown in Fig.~\ref{fig:shifting_mean}, empirical joint distributions in the energies have good fits towards Normal multidimensional distributions with linear (co-)variances.
@@
 To capture the realistic Turner model $\EnergyTurner$ while keeping
-the computational cost in check, we exploited the close correlation of
+a reasonable computational cost, we exploit a tight correlation between
 $\EnergyTurner$ and the fitted stacking model $\EnergyStacking$
@@
 $\EnergyTurner(S;R) \approx \gamma\cdot \EnergyStacking(S;R) + \delta$
-for any structure $R$ and sequence $S$. We learned the
+for any structure $R$ and sequence $S$. We inferred the
 $(\gamma,\delta)$ parameters from a set of sequences generated with
-homogenous weights $w=e^{\beta}$, tuning only \GCb\% to a
+homogenous weights $w=e^{\beta}$, tuning only the \GCb\% to a
 predetermined value.  Finally, we adjusted the targeted energies
@@
 
-Fig.~\ref{fig:energydist-pk} illustrates the described strategy. For
-the two target structures of Fig.~\ref{fig:energydist-pk}B,
-Fig.~\ref{fig:energydist-pk}A shows the good fit between Turner
-energies based and energies in the stacking energy model. For the
-shown fits we sampled $n=1\,000$ sequences, targeting a ${\sf GC}\%$
-of $50\%$.
-%
-For an example instance of the \texttt{Modena} benchmark with two
-pseudoknotted target structures, Fig.~\ref{fig:energydist-pk}B shows
-the Turner energy distributions of the single structures as they
-result from sampling with different weight parameters. The figure
-illustrates how our multidimensionnal Boltzmann sampling strategy can,
-to a large extent, independently shift the Turner energies of sampled
-sequences towards prescribed targets. See Supp. Mat.,
-Fig.~\ref{fig:energydist} for a further example with three
-pseudoknot-free target structures.
+Fig.~\ref{fig:energydist} illustrates the described
+strategy. Fig.~\ref{fig:energydist}A
+illustrates our capability to target Turner energies based on the
+stacking energy model. It shows the deviation between
+$\EnergyTurner$ and $\EnergyStacking$, observed for instances of the
+\texttt{Modena} benchmark. For each structure, $n=1\,000$ sequences
+were sampled, while targeting a ${\sf GC}\%$ of $50\%$, and reported
+the difference between the expected and sampled energies for both the
+Turner model and its approximation using a stacking model.  For an
+example instance of the \texttt{Modena} benchmark \texttt{3str},
+Fig.~\ref{fig:energydist}B shows the Turner energy distributions of
+the single structures as they result from sampling with different weight
+parameters. The figure illustrates how our multidimensionnal Boltzmann
+sampling strategy can accurately and, to a large extent, independently
+shift the Turner energies of sampled sequences to prescribed targets.
 
@@
   \begin{center}
-    {\sf \bfseries A}\ \includegraphicstop[width=0.38\textwidth]{Figs/regression-pk}\hfill
-    {\sf \bfseries B}\ \includegraphicstop[width=0.55\textwidth]{Figs/energy_distribution-pk}
+    {\sf \bfseries A}\includegraphicstop[width=0.32\textwidth]{Figs/offset2}\hspace{.5cm}
+    {\sf \bfseries B}\includegraphicstop[width=0.57\textwidth]{Figs/energy_distribution}
   \end{center}
   \caption{%
-    Targeting specific energies for pseudoknotted structures using multi-dimensional Boltzmann
-    sampling.  (A) Linear fits between the energies in the stacking model to the Turner model for
-    sampled sequences; the good match enables more efficient targeting of Turner energies based on
-    targeting stacking model energies.
-    We show the fits for both target structures S1 and S2 of subfigure B.
+    Targeting specific energies using multi-dimensional Boltzmann
+    sampling.  (A) Accuracy of targeting over all target structures of
+    the \Software{Modena} benchmark instances \texttt{RNAtabupath}
+    (2str), \texttt{3str} and \texttt{4str}. 
+    For the stacking and the Turner energy model, the relative offsets between the targeted and 
+    achieved energies were determined. We show the deviations of these offsets between both models.
     (B) Turner energy distributions for
-    two target structures S1 and S2. We show the results of
-    targeting the respective free energies $(-20,-10)$, $(-10,-20)$, and $(-15,-15)$
-    for the target structures (annotated in red,
-    yellow, and blue resepectively)---this demonstrates the
+    three target structures S1, S2 and S3. We show the results of
+    targeting the respective free energies $(-40,-40,-20)$ and
+    $(-20,-20,-20)$ for the target structures (annotated in red and
+    blue, resepectively)---this demonstrates the
     effectivity of our adaptive multi-dimensional Boltzmann sampling
@@
   }
-  \label{fig:energydist-pk}
+  \label{fig:energydist}
 \end{figure*}
@@
 %We define the following notations: $\real^\infty:=\real\cup\{\infty\}$, $[i,j]:=\{i,\ldots,j\},$ \ldots
-Finally, we turn to the complexity of $\NumDesign(G)$, the problem of computing the number of valid sequences for a given compatibility graph $G=(V,E)$.
-Note that this problem corresponds to the partition function problem in a simple $(0/\infty)$-valued base pair model, with $\beta\neq 1$.
+We now establish the complexity of $\NumDesign(G)$, the problem of computing the number of valid sequences for a given compatibility graph $G=(V,E)$.
+Note that $\NumDesign(G)$ is equivalent to computing the partition function problem in a simple $(0/\infty)$-valued base pair model, with $\beta\neq 1$.
 As previously noted~\citep{Flamm2001}, a set of target structures admits a valid design iff its compatibility graph is bipartite, which can be tested in linear time.
@@
 
-The introduction of $\{\Gb, \Ub\}$ base pairs radically changes this picture, and we show that valid designs for a set of structures cannot be counted in polynomial time, unless ${\sf\# P}={\sf FP}$.
-The latter equality would, in particular, imply the classic ${\sf P}={\sf NP}$, and solving $\NumDesign(G)$ is polynomial time is therefore probably difficult.
+The problem becomes much harder when $\{\Gb, \Ub\}$ base pairs are allowed, and we show that valid designs for a set of structures cannot be counted in polynomial time, unless ${\sf\# P}={\sf FP}$.
+This equality would, in particular, imply the classic ${\sf P}={\sf NP}$.
 
@@
 Moreover remark that, as observed in Subsec.~\ref{sec:complexity}, assigning a nucleotide to a position $u\in V$ constrains the parity ($\{\Ab,\Gb\}$ or $\{\Cb,\Ub\}$) of all positions in the connected component of $u$.
-For this reason, we restrict our attention to the counting of valid designs \emph{up to trivial} $(\Ab\leftrightarrow \Cb/\Gb\leftrightarrow \Ub)$ symmetry, by constraining the positions in $V_{1}$ (resp.
-$V_2$) to only $\Ab$ and $\Gb$ (resp.
-$\Cb$ and $\Ub$).
+For this reason, we restrict our attention to the counting of valid designs \emph{up to trivial} $(\Ab\leftrightarrow \Cb/\Gb\leftrightarrow \Ub)$ symmetry, by constraining the positions in $V_{1}$ to $\Ab$ and $\Gb$.
 Let $\Design{G}$ denote the subset of all designs for $G$ under this constraint, noting that $\NumDesign(G) = 2\cdot|\Design{G}|$.
@@
 in the field of RNA sequence design by enforcing additional
-constraints, like \GCb\%, while controlling the energy of
+constraints, like the \GCb\%, while controlling the energy of
 multiple target structures. Thus it presents a major advance over

### 2018-05-15 RNARedPrint-Bioinfo2018.tex

--- 2018-05-15 RNARedPrint-Bioinfo2018.tex
@@
 
-\newcommand{\RNAblueprint}{{\tt \bfseries{} RNA\textcolor{blue!70!black}{Blue}print}}
-\newcommand{\ourprog}{{\tt \bfseries{}RNA\textcolor{red!70!black}{Red}Print}}
+\newcommand{\RNAblueprint}{{\tt \bfseries{}\color{black!75} RNA\textcolor{blue!70!black}{Blue}Print}}
+\newcommand{\ourprog}{{\tt \bfseries{}\color{black!75}RNA\textcolor{red!70!black}{Red}Print}}
 
@@
 $\EnergyTurner(S;R) \approx \gamma\cdot \EnergyStacking(S;R) + \delta$
-for any structure $R$ and sequence $S$. We learned the
+for any structure $R$ and sequence $S$. We inferred the
 $(\gamma,\delta)$ parameters from a set of sequences generated with

### 2018-05-15 RNARedPrint-Bioinfo2018.tex
Update on Overleaf.

--- 2018-05-15 RNARedPrint-Bioinfo2018.tex
@@
              &   \textbf{\textrm{Dataset}}   & {\bfseries\ourprog{}} & \textbf{Uniform} & \textbf{Improvement} \\\toprule
-  Seeds      & 2str & 21.67 ($\pm$4.38) & 37.74 ($\pm$6.45) & 73\%\\
-             & 3str & 18.09 ($\pm$3.98) & 30.49 ($\pm$5.41) & 71\%\\
-             & 4str & 19.94 ($\pm$3.84) & 32.29 ($\pm$5.24) & 63\%\\
-             & PK60 & 19.94 ($\pm$3.84) & 32.29 ($\pm$5.24) & 63\%\\
-             & PK80 & 19.94 ($\pm$3.84) & 32.29 ($\pm$5.24) & 63\%\\
-             & LE80 & 19.94 ($\pm$3.84) & 32.29 ($\pm$5.24) & 63\%\\
+  Seeds      & 2str & 21.69 ($\pm$4.37) & 37.74 ($\pm$6.45) & 74\%\\
+             & 3str & 18.10 ($\pm$3.98) & 30.49 ($\pm$5.41) & 71\%\\
+             & 4str & 20.15 ($\pm$3.83) & 32.32 ($\pm$5.23) & 62\%\\
+             & PK60 & 12.12 ($\pm$3.04) & 19.81 ($\pm$4.24) & 68\%\\
+             & PK80 & 13.81 ($\pm$3.63) & 24.05 ($\pm$4.93) & 82\%\\
+             & LE80 & 10.54 ($\pm$2.77) & 17.80 ($\pm$4.17) & 78\%\\
              \midrule
 
-  Optimized  & 2str & 5.84 ($\pm$1.31) & 7.95 ($\pm$1.76) & 28\%\\
-             & 3str & 5.08 ($\pm$1.10) & 7.04 ($\pm$1.52) & 31\%\\
-             & 4str & 8.77($\pm$1.48) & 13.13 ($\pm$2.13) & 37\% \\
+  Optimized  & 2str & 5.42 ($\pm$1.27) & 7.95 ($\pm$1.75) & 35\%\\
+             & 3str & 5.10 ($\pm$1.10) & 7.04 ($\pm$1.52) & 31\%\\
+             & 4str & 9.17($\pm$1.51) & 13.25 ($\pm$2.13) & 38\% \\
+             & PK60 & 3.75 ($\pm$1.19) & 6.45 ($\pm$1.84) & 125\%\\
+             & PK80 & 3.52 ($\pm$1.58) & 7.99 ($\pm$2.35) & 107\%\\
+             & LE80 & 3.59 ($\pm$1.11) & 5.88 ($\pm$1.59) & 215\%\\
              \bottomrule

### 2018-05-15 RNARedPrint-Bioinfo2018.tex
Update on Overleaf.

--- 2018-05-15 RNARedPrint-Bioinfo2018.tex
@@
   standard deviations. Smaller scores are better; the final column
-  reports our improvements over uniform sampling.}
+  reports the mean of our improvements over uniform sampling. Pseudoknotted}
 \label{tab:benchmark-results}

### 2018-05-15 RNARedPrint-Bioinfo2018.tex
Update on Overleaf.

--- 2018-05-15 RNARedPrint-Bioinfo2018.tex
@@
              & 4str & 20.15 ($\pm$3.83) & 32.32 ($\pm$5.23) & 62\%\\
+             \midrule
              & PK60 & 12.12 ($\pm$3.04) & 19.81 ($\pm$4.24) & 68\%\\
@@
              & 4str & 9.17($\pm$1.51) & 13.25 ($\pm$2.13) & 38\% \\
-             & PK60 & 3.75 ($\pm$1.19) & 6.45 ($\pm$1.84) & 125\%\\
-             & PK80 & 3.52 ($\pm$1.58) & 7.99 ($\pm$2.35) & 107\%\\
-             & LE80 & 3.59 ($\pm$1.11) & 5.88 ($\pm$1.59) & 215\%\\
+             \midrule
+             & PK60* & 3.75 ($\pm$1.19) & 6.45 ($\pm$1.84) & 125\%\\
+             & PK80* & 3.52 ($\pm$1.58) & 7.99 ($\pm$2.35) & 107\%\\
+             & LE80* & 3.59 ($\pm$1.11) & 5.88 ($\pm$1.59) & 215\%\\
              \bottomrule
@@
   standard deviations. Smaller scores are better; the final column
-  reports our improvements over uniform sampling.}
+  reports the mean of our improvements over uniform sampling.
+\TODO{PK datasets ensemble free energy was not available thus approximated with mfe energy.
+Mention negative scores somewhere!
+Asterisk benchmarks are not complete yet.}}
 \label{tab:benchmark-results}

### 2018-05-17 RNARedPrint-Bioinfo2018.tex
New targeting figure

--- 2018-05-15 RNARedPrint-Bioinfo2018.tex
@@
 the two target structures of Fig.~\ref{fig:energydist-pk}B,
-Fig.~\ref{fig:energydist-pk}A shows the good fit between Turner
-energies based and energies in the stacking energy model. For the
-shown fits we sampled $n=1\,000$ sequences, targeting a ${\sf GC}\%$
-of $50\%$.
+Fig.~\ref{fig:energydist-pk}A shows the good fit between realistic
+energies in the full-fledged Dirks and Pierce energy model for
+pseudoknots (D\&P model) and energies in the stacking energy
+model. For the shown fits we sampled $n=1\,000$ sequences, targeting a
+${\sf GC}\%$ of $60\%$.
 %
@@
   \begin{center}
-    {\sf \bfseries A}\ \includegraphicstop[width=0.38\textwidth]{Figs/regression-pk}\hfill
-    {\sf \bfseries B}\ \includegraphicstop[width=0.55\textwidth]{Figs/energy_distribution-pk}
+    {\sf \bfseries A}\ \includegraphicstop[width=0.38\textwidth]{Figs/PKB00211_PKB00239_0_regression}\hfill
+    {\sf \bfseries B}\ \includegraphicstop[width=0.55\textwidth]{Figs/PKB00211_PKB00239_0_energy_distribution}
   \end{center}
   \caption{%
-    Targeting specific energies for pseudoknotted structures using multi-dimensional Boltzmann
-    sampling.  (A) Linear fits between the energies in the stacking model to the Turner model for
-    sampled sequences; the good match enables more efficient targeting of Turner energies based on
-    targeting stacking model energies.
-    We show the fits for both target structures S1 and S2 of subfigure B.
-    (B) Turner energy distributions for
-    two target structures S1 and S2. We show the results of
-    targeting the respective free energies $(-20,-10)$, $(-10,-20)$, and $(-15,-15)$
-    for the target structures (annotated in red,
-    yellow, and blue resepectively)---this demonstrates the
-    effectivity of our adaptive multi-dimensional Boltzmann sampling
-    procedure; moreover, for comparison, the distributions for uniform
-    (U) and Boltzmann (B) samples---respectively associated with
-    homogenous weights $1$ and $e^\beta$.
+    Targeting specific energies for pseudoknotted structures using
+    multi-dimensional Boltzmann sampling.  (A) Linear fits between the
+    energies in the stacking model to the realistic pseudoknot energy
+    model by Dirks and Pierce (D\&P) for sampled sequences; the good
+    match enables more efficient targeting of Turner energies based on
+    targeting stacking model energies.  We show the fits for both
+    target structures S1 and S2 of subfigure B.  (B) D\&P energy
+    distributions for two target structures S1 and S2. We show the
+    results of targeting the respective free energies $(-30,-20)$,
+    $(-30,-30)$, $(-25,-25)$, and $(-20,-30)$ for the target
+    structures (annotated in red, yellow, and blue
+    resepectively)---this demonstrates the effectivity of our adaptive
+    multi-dimensional Boltzmann sampling procedure; moreover, for
+    comparison, the distributions for uniform and Boltzmann
+    samples---respectively associated with homogenous weights $1$ and
+    $e^\beta$.
 %    ($\delta=0.05, \delta'=0.1, \gamma=1.1, \psi=20, n=1000$)
@@
 relatively high weights $e^{\beta}$, setting the targeted \GCb\%
-to 50\%. For each instance of the \Software{Modena}
+to 60\%. For each instance of the \Software{Modena}
 benchmark~\citep{Taneda2015}, we estimated such target energies and
@@
 
-Moreover, for each generated seed sequence, we optimized the objective function $(\ref{eq:blueprintobjective})$ through an adaptive  greedy walk consisting of 500 steps.
+Moreover, for each generated seed sequence, we optimized the objective function $(\ref{eq:blueprintobjective})$ through an adaptive  greedy walk consisting of 500 steps \todo{(200 for PK benchmarksets)}.
 At each step, we resampled (uniformly random) the positions of a randomly selected component in the base pair dependency graph, and accepted the modification only if it resulted in a gain, as described in \RNAblueprint{}.
@@
 the quality of Boltzmann designs over uniform ones, suggesting a
-superior potential for optimization ($\sim 32\%$ avg improvement,
+superior potential for optimization ($\sim 32\%$ avg improvement, \todo{improvement not in percent}
 Table~\ref{tab:benchmark-results}, optimized). Notably, our subsequent
@@
   standard deviations. Smaller scores are better; the final column
-  reports our improvements over uniform sampling.}
+  reports the mean of our improvements over uniform sampling. \TODO{PK datasets ensemble free energy was}}
 \label{tab:benchmark-results}

### 2018-05-17 RNARedPrint-Bioinfo2018.tex

@@
   standard deviations. Smaller scores are better; the final column
-  reports the mean of our improvements over uniform sampling. \TODO{PK datasets ensemble free energy was}}
+  reports the mean of our improvements over uniform sampling.
+\TODO{PK datasets ensemble free energy was not available thus approximated with mfe energy.
+Mention negative scores somewhere!
+Asterisk benchmarks are not complete yet.}}
 \label{tab:benchmark-results}

### 2018-05-22 RNARedPrint-Bioinfo2018.tex
Minor changes

--- 2018-05-15 RNARedPrint-Bioinfo2018.tex
@@
 \begin{tabular}{@{}>{\bf}l@{\quad}>{\tt}l@{\quad}@{\quad}c@{\quad}c@{\quad}c@{}}
-             &   \textbf{\textrm{Dataset}}   & {\bfseries\ourprog{}} & \textbf{Uniform} & \textbf{Improvement} \\\toprule
+             &   \textbf{\textrm{Dataset}}   & {\bfseries\ourprog{}} & \textbf{Uniform} & \textbf{$\Delta_f$}\\\toprule
   Seeds      & 2str & 21.69 ($\pm$4.37) & 37.74 ($\pm$6.45) & 74\%\\
              & 3str & 18.10 ($\pm$3.98) & 30.49 ($\pm$5.41) & 71\%\\
-             & 4str & 20.15 ($\pm$3.83) & 32.32 ($\pm$5.23) & 62\%\\
-             \midrule
+             & 4str & 20.15 ($\pm$3.83) & 32.32 ($\pm$5.23) & 62\%\\ %\cline{2-5}
              & PK60 & 12.12 ($\pm$3.04) & 19.81 ($\pm$4.24) & 68\%\\
-             & PK80 & 13.81 ($\pm$3.63) & 24.05 ($\pm$4.93) & 82\%\\
+             & PK80 & 13.81 ($\pm$3.63) & 24.05 ($\pm$4.93) & 82\%\\ %\cline{2-5}
              & LE80 & 10.54 ($\pm$2.77) & 17.80 ($\pm$4.17) & 78\%\\
@@
              & 3str & 5.10 ($\pm$1.10) & 7.04 ($\pm$1.52) & 31\%\\
-             & 4str & 9.17($\pm$1.51) & 13.25 ($\pm$2.13) & 38\% \\
-             \midrule
-             & PK60* & 3.75 ($\pm$1.19) & 6.45 ($\pm$1.84) & 125\%\\
-             & PK80* & 3.52 ($\pm$1.58) & 7.99 ($\pm$2.35) & 107\%\\
-             & LE80* & 3.59 ($\pm$1.11) & 5.88 ($\pm$1.59) & 215\%\\
+             & 4str & 9.17($\pm$1.51) & 13.25 ($\pm$2.13) & 38\%\\ %\cline{2-5}
+             & PK60 & 3.75 ($\pm$1.19) & 6.45 ($\pm$1.84) & 125\%\\
+             & PK80 & 3.52 ($\pm$1.58) & 7.99 ($\pm$2.35) & 107\%\\ %\cline{2-5}
+             & LE80 & 3.59 ($\pm$1.11) & 5.88 ($\pm$1.59) & 215\%\\
              \bottomrule
@@
   standard deviations. Smaller scores are better; the final column
-  reports the mean of our improvements over uniform sampling.
-\TODO{PK datasets ensemble free energy was not available thus approximated with mfe energy.
-Mention negative scores somewhere!
-Asterisk benchmarks are not complete yet.}}
+  reports the mean of our improvements over uniform sampling. \TODO{PK datasets ensemble free energy was not available thus approximated with mfe energy.}}
 \label{tab:benchmark-results}

### 2018-05-22 RNARedPrint-Bioinfo2018.tex

--- 2018-05-17 RNARedPrint-Bioinfo2018.tex
@@
 \begin{tabular}{@{}>{\bf}l@{\quad}>{\tt}l@{\quad}@{\quad}c@{\quad}c@{\quad}c@{}}
-             &   \textbf{\textrm{Dataset}}   & {\bfseries\ourprog{}} & \textbf{Uniform} & \textbf{Improvement} \\\toprule
+             &   \textbf{\textrm{Dataset}}   & {\bfseries\ourprog{}} & \textbf{Uniform} & \textbf{$\Delta_f$}\\\toprule
   Seeds      & 2str & 21.69 ($\pm$4.37) & 37.74 ($\pm$6.45) & 74\%\\
              & 3str & 18.10 ($\pm$3.98) & 30.49 ($\pm$5.41) & 71\%\\
-             & 4str & 20.15 ($\pm$3.83) & 32.32 ($\pm$5.23) & 62\%\\
+             & 4str & 20.15 ($\pm$3.83) & 32.32 ($\pm$5.23) & 62\%\\ %\cline{2-5}
              \midrule
              & PK60 & 12.12 ($\pm$3.04) & 19.81 ($\pm$4.24) & 68\%\\
-             & PK80 & 13.81 ($\pm$3.63) & 24.05 ($\pm$4.93) & 82\%\\
+             & PK80 & 13.81 ($\pm$3.63) & 24.05 ($\pm$4.93) & 82\%\\ %\cline{2-5}
              & LE80 & 10.54 ($\pm$2.77) & 17.80 ($\pm$4.17) & 78\%\\
@@
              & 3str & 5.10 ($\pm$1.10) & 7.04 ($\pm$1.52) & 31\%\\
-             & 4str & 9.17($\pm$1.51) & 13.25 ($\pm$2.13) & 38\% \\
+             & 4str & 9.17($\pm$1.51) & 13.25 ($\pm$2.13) & 38\%\\ %\cline{2-5}
              \midrule

### 2018-05-22 RNARedPrint-Bioinfo2018.tex
Minor edits 'Sharp-P' section

--- 2018-05-17 RNARedPrint-Bioinfo2018.tex
@@
 %We define the following notations: $\real^\infty:=\real\cup\{\infty\}$, $[i,j]:=\{i,\ldots,j\},$ \ldots
-Finally, we turn to the complexity of $\NumDesign(G)$, the problem of computing the number of valid sequences for a given compatibility graph $G=(V,E)$.
-Note that this problem corresponds to the partition function problem in a simple $(0/\infty)$-valued base pair model, with $\beta\neq 1$.
-As previously noted~\citep{Flamm2001}, a set of target structures admits a valid design iff its compatibility graph is bipartite, which can be tested in linear time.
-Moreover, without $\{\Gb, \Ub\}$ base pairs, any connected component  $C\in {\rm CC}(G)$ is entirely determined by the assignment of a single of its nucleotides.
-The number of valid designs is thus simply $4^{\#{\rm CC}(G)}$, where $\#{\rm CC}(G)$ is the \Def{number of connected components}.
-
-The introduction of $\{\Gb, \Ub\}$ base pairs radically changes this picture, and we show that valid designs for a set of structures cannot be counted in polynomial time, unless ${\sf\# P}={\sf FP}$.
+Finally, we turn to the complexity of $\NumDesign(G)$, the problem of
+computing the number of valid sequences for a given compatibility
+graph $G=(V,E)$. Note that this problem corresponds to the partition
+function problem in a simple base pair model, assigning $0$ to
+canonical and $\infty$ to non-canonical base pairs, with
+$\beta\neq 0$. As previously noted~\citep{Flamm2001}, a set of target
+structures admits a valid design iff its compatibility graph is
+bipartite, which can be tested in linear time.  Moreover, without
+$\{\Gb, \Ub\}$ base pairs, the design of any connected component $C$
+of $G$ would be entirely determined by the assignment of a single of
+its nucleotides. The number of valid designs is thus simply
+$4^{\#{\rm CC}(G)}$, where $\#{\rm CC}(G)$ is the \Def{number of
+  connected components}.
+
+The potential to form $\{\Gb, \Ub\}$ base pairs radically changes this picture.
+As we show, the valid designs for a set of structures cannot be counted in polynomial time, unless ${\sf\# P}={\sf FP}$.
 The latter equality would, in particular, imply the classic ${\sf P}={\sf NP}$, and solving $\NumDesign(G)$ is polynomial time is therefore probably difficult.

### 2018-05-22 RNARedPrint-Bioinfo2018.tex
minor change

--- 2018-05-22 RNARedPrint-Bioinfo2018.tex
@@
 We define the \emph{maximum separator size} $s$ as
-$\max_{u,v\in V} | \separator{u}{v} |$ and denote the maximum size of
-$\difference{u}{v}$ over $(u,v)\in\edgesToR{}$ as $D$.  In the absence
+$\max_{u,v\in V} | \separator{u}{v} |$ and denote by  $D$ the maximum size of
+$\difference{u}{v}$ over $(u,v)\in\edgesToR{}$.  In the absence
 of specific optimizations, running Alg.~\ref{alg:pf} requires

### 2018-05-22 RNARedPrint-Bioinfo2018.tex
Update Benchmark numbers

--- 2018-05-22 RNARedPrint-Bioinfo2018.tex
@@
              &   \textbf{\textrm{Dataset}}   & {\bfseries\ourprog{}} & \textbf{Uniform} & \textbf{Improvement} \\\toprule
-  Seeds      & 2str & 21.69 ($\pm$4.37) & 37.74 ($\pm$6.45) & 74\%\\
-             & 3str & 18.10 ($\pm$3.98) & 30.49 ($\pm$5.41) & 71\%\\
-             & 4str & 20.15 ($\pm$3.83) & 32.32 ($\pm$5.23) & 62\%\\
+  Seeds      & 2str & 21.69 ($\pm$4.37) & 37.74 ($\pm$6.45) & 16.05\\
+             & 3str & 18.10 ($\pm$3.98) & 30.49 ($\pm$5.41) & 12.39\\
+             & 4str & 20.15 ($\pm$3.83) & 32.32 ($\pm$5.23) & 12.17\\
              \midrule
-             & PK60 & 12.12 ($\pm$3.04) & 19.81 ($\pm$4.24) & 68\%\\
-             & PK80 & 13.81 ($\pm$3.63) & 24.05 ($\pm$4.93) & 82\%\\
-             & LE80 & 10.54 ($\pm$2.77) & 17.80 ($\pm$4.17) & 78\%\\
+             & PK60 & 12.12 ($\pm$3.04) & 19.81 ($\pm$4.24) & 7.69\\
+             & PK80 & 13.81 ($\pm$3.63) & 24.05 ($\pm$4.93) & 10.24\\
+             & LE80 & 10.54 ($\pm$2.77) & 17.80 ($\pm$4.17) & 7.26\\
              \midrule
 
-  Optimized  & 2str & 5.42 ($\pm$1.27) & 7.95 ($\pm$1.75) & 35\%\\
-             & 3str & 5.10 ($\pm$1.10) & 7.04 ($\pm$1.52) & 31\%\\
-             & 4str & 9.17($\pm$1.51) & 13.25 ($\pm$2.13) & 38\% \\
+  Optimized  & 2str & 5.42 ($\pm$1.27) & 7.95 ($\pm$1.75) & 2.52\\
+             & 3str & 5.10 ($\pm$1.10) & 7.04 ($\pm$1.52) & 1.94\\
+             & 4str & 9.17($\pm$1.51) & 13.25 ($\pm$2.13) & 4.17\\
              \midrule
-             & PK60* & 3.75 ($\pm$1.19) & 6.45 ($\pm$1.84) & 125\%\\
-             & PK80* & 3.52 ($\pm$1.58) & 7.99 ($\pm$2.35) & 107\%\\
-             & LE80* & 3.59 ($\pm$1.11) & 5.88 ($\pm$1.59) & 215\%\\
+             & PK60* & 3.75 ($\pm$1.19) & 6.45 ($\pm$1.84) & 2.58\\
+             & PK80* & 3.52 ($\pm$1.58) & 7.99 ($\pm$2.35) & 4.34\\
+             & LE80  & 3.59 ($\pm$1.11) & 5.88 ($\pm$1.59) & 2.28\\
              \bottomrule

### 2018-05-22 RNARedPrint-Bioinfo2018.tex

--- 2018-05-22 RNARedPrint-Bioinfo2018.tex
@@
 %We define the following notations: $\real^\infty:=\real\cup\{\infty\}$, $[i,j]:=\{i,\ldots,j\},$ \ldots
-We now establish the complexity of $\NumDesign(G)$, the problem of computing the number of valid sequences for a given compatibility graph $G=(V,E)$.
-Note that $\NumDesign(G)$ is equivalent to computing the partition function problem in a simple $(0/\infty)$-valued base pair model, with $\beta\neq 1$.
-As previously noted~\citep{Flamm2001}, a set of target structures admits a valid design iff its compatibility graph is bipartite, which can be tested in linear time.
-Moreover, without $\{\Gb, \Ub\}$ base pairs, any connected component  $C\in {\rm CC}(G)$ is entirely determined by the assignment of a single of its nucleotides.
-The number of valid designs is thus simply $4^{\#{\rm CC}(G)}$, where $\#{\rm CC}(G)$ is the \Def{number of connected components}.
+Finally, we turn to the complexity of $\NumDesign(G)$, the problem of
+computing the number of valid sequences for a given compatibility
+graph $G=(V,E)$. Note that this problem corresponds to the partition
+function problem in a simple base pair model, assigning $0$ to
+canonical and $\infty$ to non-canonical base pairs and setting 
+$\beta<0$, so that any valid structure counts as $1$ in the partition 
+function. As previously noted~\citep{Flamm2001}, a set of target
+structures admits a valid design iff its compatibility graph is
+bipartite, which can be tested in linear time.  Moreover, without
+$\{\Gb, \Ub\}$ base pairs, the design of any connected component $C$
+of $G$ would be entirely determined by the assignment of a single of
+its nucleotides. The number of valid designs is thus simply
+$4^{\#{\rm CC}(G)}$, where $\#{\rm CC}(G)$ is the \Def{number of
+  connected components}.
 

### 2018-05-22 RNARedPrint-Bioinfo2018.tex

@@
 \begin{tabular}{@{}>{\bf}l@{\quad}>{\tt}l@{\quad}@{\quad}c@{\quad}c@{\quad}c@{}}
-             &   \textbf{\textrm{Dataset}}   & {\bfseries\ourprog{}} & \textbf{Uniform} & \textbf{$\Delta_f$}\\\toprule
-  Seeds      & 2str & 21.69 ($\pm$4.37) & 37.74 ($\pm$6.45) & 74\%\\
-             & 3str & 18.10 ($\pm$3.98) & 30.49 ($\pm$5.41) & 71\%\\
-             & 4str & 20.15 ($\pm$3.83) & 32.32 ($\pm$5.23) & 62\%\\ %\cline{2-5}
+  Seeds      & 2str & 21.69 ($\pm$4.37) & 37.74 ($\pm$6.45) & 16.05\\
+             & 3str & 18.10 ($\pm$3.98) & 30.49 ($\pm$5.41) & 12.39\\
+             & 4str & 20.15 ($\pm$3.83) & 32.32 ($\pm$5.23) & 12.17\\
              \midrule
-             & PK60 & 12.12 ($\pm$3.04) & 19.81 ($\pm$4.24) & 68\%\\
-             & PK80 & 13.81 ($\pm$3.63) & 24.05 ($\pm$4.93) & 82\%\\ %\cline{2-5}
-             & LE80 & 10.54 ($\pm$2.77) & 17.80 ($\pm$4.17) & 78\%\\
+             & PK60 & 12.12 ($\pm$3.04) & 19.81 ($\pm$4.24) & 7.69\\
+             & PK80 & 13.81 ($\pm$3.63) & 24.05 ($\pm$4.93) & 10.24\\
+             & LE80 & 10.54 ($\pm$2.77) & 17.80 ($\pm$4.17) & 7.26\\
              \midrule
 
-  Optimized  & 2str & 5.42 ($\pm$1.27) & 7.95 ($\pm$1.75) & 35\%\\
-             & 3str & 5.10 ($\pm$1.10) & 7.04 ($\pm$1.52) & 31\%\\
-             & 4str & 9.17($\pm$1.51) & 13.25 ($\pm$2.13) & 38\%\\ %\cline{2-5}
+  Optimized  & 2str & 5.42 ($\pm$1.27) & 7.95 ($\pm$1.75) & 2.52\\
+             & 3str & 5.10 ($\pm$1.10) & 7.04 ($\pm$1.52) & 1.94\\
+             & 4str & 9.17($\pm$1.51) & 13.25 ($\pm$2.13) & 4.17\\
              \midrule
-             & PK60* & 3.75 ($\pm$1.19) & 6.45 ($\pm$1.84) & 125\%\\
-             & PK80* & 3.52 ($\pm$1.58) & 7.99 ($\pm$2.35) & 107\%\\
-             & LE80* & 3.59 ($\pm$1.11) & 5.88 ($\pm$1.59) & 215\%\\
+             & PK60* & 3.75 ($\pm$1.19) & 6.45 ($\pm$1.84) & 2.58\\
+             & PK80* & 3.52 ($\pm$1.58) & 7.99 ($\pm$2.35) & 4.34\\
+             & LE80  & 3.59 ($\pm$1.11) & 5.88 ($\pm$1.59) & 2.28\\
              \bottomrule
@@
 For this reason, we restrict our attention to the counting of valid designs \emph{up to trivial} $(\Ab\leftrightarrow \Cb/\Gb\leftrightarrow \Ub)$ symmetry, by constraining the positions in $V_{1}$ to $\Ab$ and $\Gb$.
+Let $\Design{G}$ denote the subset of all designs for $G$ under this constraint, noting that $\NumDesign(G) = 2\cdot|\Design{G}|$.
+To establish that claim, we consider instances $G=(V_1\cup V_2, E)$ that are connected and bipartite ($(V_1\times V_2) \cap E = \varnothing$), noting that hardness over restricted instances implies hardness of the general problem.
+Moreover remark that, as observed in Subsec.~\ref{sec:complexity}, assigning a nucleotide to a position $u\in V$ constrains the parity ($\{\Ab,\Gb\}$ or $\{\Cb,\Ub\}$) of all positions in the connected component of $u$.
+For this reason, we restrict our attention to the counting of valid designs \emph{up to trivial} $(\Ab\leftrightarrow \Cb/\Gb\leftrightarrow \Ub)$ symmetry, by constraining the positions in $V_{1}$ (resp.
+$V_2$) to only $\Ab$ and $\Gb$ (resp.
+$\Cb$ and $\Ub$).
 Let $\Design{G}$ denote the subset of all designs for $G$ under this constraint, noting that $\NumDesign(G) = 2\cdot|\Design{G}|$.

### 2018-05-22 RNARedPrint-Bioinfo2018.tex
Added clines in Table
Table header and caption

@@
 \begin{tabular}{@{}>{\bf}l@{\quad}>{\tt}l@{\quad}@{\quad}c@{\quad}c@{\quad}c@{}}
+  &   \textbf{\textrm{Dataset}}   & {\bfseries\ourprog{}} & \textbf{Uniform} & \textbf{$\Delta$-Improvement}\\
   Seeds      & 2str & 21.69 ($\pm$4.37) & 37.74 ($\pm$6.45) & 16.05\\
@@
              & 4str & 20.15 ($\pm$3.83) & 32.32 ($\pm$5.23) & 12.17\\
-             \midrule
+             \cline{2-5}
              & PK60 & 12.12 ($\pm$3.04) & 19.81 ($\pm$4.24) & 7.69\\
@@
              & 4str & 9.17($\pm$1.51) & 13.25 ($\pm$2.13) & 4.17\\
-             \midrule
+             \cline{2-5}
              & PK60* & 3.75 ($\pm$1.19) & 6.45 ($\pm$1.84) & 2.58\\
@@
   standard deviations. Smaller scores are better; the final column
-  reports the mean of our improvements over uniform sampling.
-\TODO{PK datasets ensemble free energy was not available thus approximated with mfe energy.
-Mention negative scores somewhere!
-Asterisk benchmarks are not complete yet.}}
+  reports our improvements due to Boltzmann sampling over uniform
+  sampling (as difference of mean scores).}
 \label{tab:benchmark-results}

### 2018-05-22 RNARedPrint-Bioinfo2018.tex
new figure
update figure
Minor
edits Table 1 caption & results

@@
   \begin{center}
-    {\sf \bfseries A}\ \includegraphicstop[width=0.38\textwidth]{Figs/PKB00211_PKB00239_0_regression}\hfill
-    {\sf \bfseries B}\ \includegraphicstop[width=0.55\textwidth]{Figs/PKB00211_PKB00239_0_energy_distribution}
+    %{\sf \bfseries A}\
+    \includegraphicstop[width=0.99\textwidth]{Figs/energy_shift.pdf}\hfill
+    %{\sf \bfseries B}\ \includegraphicstop[width=0.55\textwidth]{Figs/PKB00211_PKB00239_0_energy_distribution}
   \end{center}
@@
 that Boltzmann-sampled sequences constitute better seeds
-($\sim\! 69\%$ improvement on average) compared to uniform seeds for
+(10.97 improvement of mean scores) compared to uniform seeds for
 \emph{all instances} of our benchmark (Supp. Mat.~\ref{sec:validity}).
 
-Moreover, for each generated seed sequence, we optimized the objective function $(\ref{eq:blueprintobjective})$ through an adaptive  greedy walk consisting of 500 steps \todo{(200 for PK benchmarksets)}.
-At each step, we resampled (uniformly random) the positions of a randomly selected component in the base pair dependency graph, and accepted the modification only if it resulted in a gain, as described in \RNAblueprint{}.
+Moreover, for each generated seed sequence, we optimized the objective
+function $(\ref{eq:blueprintobjective})$ through an adaptive greedy
+walk (500 steps for the pseudoknot free benchmark sets 2str, 3str, and
+4str; 200 for the pseudoknot benchmarksets PK60, PK80, and LE80). At
+each step, we resampled (uniformly random) the positions of a randomly
+selected component in the base pair dependency graph, and accepted the
+modification only if it resulted in a gain, as described in
+\RNAblueprint{}.
 %
@@
 the quality of Boltzmann designs over uniform ones, suggesting a
-superior potential for optimization ($\sim 32\%$ avg improvement, \todo{improvement not in percent}
-Table~\ref{tab:benchmark-results}, optimized). Notably, our subsequent
-optimization runs, which are directly inherited from the uniform
-sampling \RNAblueprint{} software, partially level the advantages of
-Boltzmann sampling. In future work, we hope to improve this aspect by
-exploiting Boltzmann sampling even during the optimization run.
+superior potential for optimization (2.97 improvement of mean
+scores, Table~\ref{tab:benchmark-results}, ``Optimized''). Notably,
+our subsequent optimization runs, which are directly inherited from
+the uniform sampling \RNAblueprint{} software, partially level the
+advantages of Boltzmann sampling. In future work, we hope to improve
+this aspect by exploiting Boltzmann sampling even during the
+optimization run.
 
@@
 
-\caption{Comparison of the multi-stable design objective function (
-  Eq.~(\ref{eq:blueprintobjective}) of sequences sampled by \ourprog{}
-  vs.\ uniform samples for three benchmark sets; before (seeds) and
-  after optimization (optimized). We report the average scores with
-  standard deviations. Smaller scores are better; the final column
-  reports our improvements due to Boltzmann sampling over uniform
-  sampling (as difference of mean scores).}
+\caption{Comparison of the multi-stable design objective function
+  (Eq.~(\ref{eq:blueprintobjective}) of sequences sampled by
+  \ourprog{} vs.\ uniform samples for three benchmark sets; before
+  (seeds) and after optimization (optimized). We report the average
+  scores with standard deviations; smaller scores are better. The
+  final column reports our improvements due to Boltzmann sampling over
+  uniform sampling (as difference of mean scores).}
 \label{tab:benchmark-results}

### 2018-05-22 RNARedPrint-Bioinfo2018.tex
New table for supplement

--- 2018-05-22 RNARedPrint-Bioinfo2018.tex
@@
 \newtheorem{hypothesis}{Working Hypothesis}
-\renewenvironment{proof}[1][]{\noindent \em Proof\ifthenelse{\equal{#1}{}}{}{ (#1)}:~}{}
+%\renewenvironment{proof}[1][]{\noindent \em Proof\ifthenelse{\equal{#1}{}}{}{ (#1)}:~}{}
 
@@
 \newcommand{\Ub}{\Nuc{U}}
+
+\newcommand{\Objectiv}{\Nuc{U}}
 
@@
 canonical and $\infty$ to non-canonical base pairs and setting 
-$\beta<0$, so that any valid structure counts as $1$ in the partition 
+$\beta>0$, so that any valid structure counts as $1$ in the partition 
 function. As previously noted~\citep{Flamm2001}, a set of target
@@
 Moreover remark that, as observed in Subsec.~\ref{sec:complexity}, assigning a nucleotide to a position $u\in V$ constrains the parity ($\{\Ab,\Gb\}$ or $\{\Cb,\Ub\}$) of all positions in the connected component of $u$.
-For this reason, we restrict our attention to the counting of valid designs \emph{up to trivial} $(\Ab\leftrightarrow \Cb/\Gb\leftrightarrow \Ub)$ symmetry, by constraining the positions in $V_{1}$ to $\Ab$ and $\Gb$.
+For this reason, we restrict our attention to the counting of valid designs \emph{up to trivial  symmetry} $(\Ab\leftrightarrow \Cb/\Gb\leftrightarrow \Ub)$, by constraining the positions in $V_{1}$ to $\Ab$ and $\Gb$.
 Let $\Design{G}$ denote the subset of all designs for $G$ under this constraint, noting that $\NumDesign(G) = 2\cdot|\Design{G}|$.
@@
 
-We turn to the surjectivity of $\Psi$, i.e. the existence of a preimage for each element $S\in \IS{G}$. Let us consider the function $f$ defined for any $(v_1,v_2)\in V_1\times V_2$ as
-\begin{align*}
-f(v_1) &= \begin{cases} \Ab & \text{if }v_1\in S\\ \Gb & \text{if }v_1\notin S\end{cases} &
-\text{\quad and\quad}&&
- f(v_2) &= \begin{cases} \Cb & \text{if }v_2\in S\\ \Ub & \text{if }v_2\notin S.\end{cases}
-\end{align*}
+We turn to the surjectivity of $\Psi$, i.e.
+the existence of a preimage for each element $I\in \IS{G}$.
+Let us consider the function $f$ defined as $f(v) = \Ab$ (resp.
+$\Gb$, $\Cb$ or $\Ub$) if $v\in V_1$ and $v\in S$ (resp.
+$v\in V_1$ and $v\notin S$, $v\in V_2$ and $v\in S$, or $v\in V_2$ and $v\notin S$).
+%
+%for any $(v_1,v_2)\in V_1\times V_2$ by 
+%\begin{align*}
+%f(v_1) &= \begin{cases} \Ab & \text{if }v_1\in S\\ \Gb & \text{if }v_1\notin S\end{cases} &
+%\text{\quad and\quad}&&
+% f(v_2) &= \begin{cases} \Cb & \text{if }v_2\in S\\ \Ub & \text{if }v_2\notin S.\end{cases}
+%\end{align*}
 One easily verifies that $\Psi(f) = S$.

### 2018-05-22 RNARedPrint-Bioinfo2018.tex

--- 2018-05-22 RNARedPrint-Bioinfo2018.tex
@@
 canonical and $\infty$ to non-canonical base pairs and setting 
-$\beta<0$, so that any valid structure counts as $1$ in the partition
+$\beta>0$, so that any valid structure counts as $1$ in the partition 
 function. As previously noted~\citep{Flamm2001}, a set of target
@@
 Moreover remark that, as observed in Subsec.~\ref{sec:complexity}, assigning a nucleotide to a position $u\in V$ constrains the parity ($\{\Ab,\Gb\}$ or $\{\Cb,\Ub\}$) of all positions in the connected component of $u$.
-For this reason, we restrict our attention to the counting of valid designs \emph{up to trivial} $(\Ab\leftrightarrow \Cb/\Gb\leftrightarrow \Ub)$ symmetry, by constraining the positions in $V_{1}$ to $\Ab$ and $\Gb$.
-Let $\Design{G}$ denote the subset of all designs for $G$ under this constraint, noting that $\NumDesign(G) = 2\cdot|\Design{G}|$.
-To establish that claim, we consider instances $G=(V_1\cup V_2, E)$ that are connected and bipartite ($(V_1\times V_2) \cap E = \varnothing$), noting that hardness over restricted instances implies hardness of the general problem.
-Moreover remark that, as observed in Subsec.~\ref{sec:complexity}, assigning a nucleotide to a position $u\in V$ constrains the parity ($\{\Ab,\Gb\}$ or $\{\Cb,\Ub\}$) of all positions in the connected component of $u$.
-For this reason, we restrict our attention to the counting of valid designs \emph{up to trivial} $(\Ab\leftrightarrow \Cb/\Gb\leftrightarrow \Ub)$ symmetry, by constraining the positions in $V_{1}$ (resp.
-$V_2$) to only $\Ab$ and $\Gb$ (resp.
-$\Cb$ and $\Ub$).
+For this reason, we restrict our attention to the counting of valid designs \emph{up to trivial  symmetry} $(\Ab\leftrightarrow \Cb/\Gb\leftrightarrow \Ub)$, by constraining the positions in $V_{1}$ to $\Ab$ and $\Gb$.
 Let $\Design{G}$ denote the subset of all designs for $G$ under this constraint, noting that $\NumDesign(G) = 2\cdot|\Design{G}|$.

### 2018-05-23 supplement.tex

@@
 \appendix
-{\centering \relsize{+4}\textbf{Supplementary Material}\\ Fixed-Parameter Tractable Sampling\\ for RNA Design with Multiple Target Structures\\%
+{\centering \relsize{+4}\textbf{Supplementary Material}\\[1em] Fixed-Parameter Tractable Sampling\\ for RNA Design with Multiple Target Structures\\%
 %{\relsize{-1}\BackupTitle}\\

### 2018-05-23 RNARedPrint-Bioinfo2018.tex
update finished benchmarks
Added figure

@@
 \newcommand{\TargetE}{E^{\star}}
+
+\newcommand{\Obj}{\text{\sf \relsize{-1}MultiDefect}}
 
@@
 
+To that purpose, we rely on a multi-stable design objective function introduced by \citet{Hammer2017}, named \Obj in the following:
+% \begin{align}
+%   f(S)= \quad & \frac{1}{k} \sum_{\ell=1}^{k} (E(S, R_\ell) - G(S))\notag\\
+%    & +\ \xi \frac{2}{k(k-1)} \sum\limits_{1\leq\ell<j\leq k}|E(S,R_\ell) - E(S,R_j)|.
+% \end{align}
+\begin{align}
+  \label{eq:blueprintobjective}
+    \Obj(S)= \quad & \frac{1}{k} \sum_{\ell=1}^{k} (E(S, R_\ell) - G(S))\notag\\
+   & +\ 0.5\frac{1}{\binom{k}{2}} \sum\limits_{1\leq\ell<j\leq k}|E(S,R_\ell) - E(S,R_j)|.
+\end{align}
+
 For a multi-target design instance, individual reasonable target
@@
 moderate \GCb\%; all these properties are desirable objectives
-for RNA design.  All sampled sequences are evaluated under the
-multi-stable design objective function of \citet{Hammer2017}:
-% \begin{align}
-%   f(S)= \quad & \frac{1}{k} \sum_{\ell=1}^{k} (E(S, R_\ell) - G(S))\notag\\
-%    & +\ \xi \frac{2}{k(k-1)} \sum\limits_{1\leq\ell<j\leq k}|E(S,R_\ell) - E(S,R_j)|.
-% \end{align}
-\begin{align}
-  \label{eq:blueprintobjective}
-    f(S)= \quad & \frac{1}{k} \sum_{\ell=1}^{k} (E(S, R_\ell) - G(S))\notag\\
-   & +\ 0.5\frac{1}{\binom{k}{2}} \sum\limits_{1\leq\ell<j\leq k}|E(S,R_\ell) - E(S,R_j)|.
-\end{align}
+for RNA design.
 
@@
 \begin{tabular}{@{}>{\bf}l@{\quad}>{\tt}l@{\quad}@{\quad}c@{\quad}c@{\quad}c@{}}
-  &   \textbf{\textrm{Dataset}}   & {\bfseries\ourprog{}} & \textbf{Uniform} & \textbf{$\Delta$-Improvement}\\
+  &   \textbf{\textrm{Dataset}}   & {\bfseries\ourprog{}} & \textbf{\RNAblueprint} & \textbf{$\Delta$-Improvement}\\ \toprule
   Seeds      & 2str & 21.69 ($\pm$4.37) & 37.74 ($\pm$6.45) & 16.05\\

### 2018-05-23 RNARedPrint-Bioinfo2018.tex
Finished revamping results section

--- 2018-05-22 RNARedPrint-Bioinfo2018.tex
@@
 \newcommand{\TargetE}{E^{\star}}
+\newcommand{\MFE}{\text{\rm MFE}}
+
+\newcommand{\Obj}{\text{\rm MultiDefect}}
 
@@
   biologically relevant multi-target Boltzmann-weighted designs for a
-  common design benchmark. Generating seed sequences with \ourprog{}, we demonstrate significant improvements over the previously best multi-target sampling strategy (uniform sampling).\\
+  common design benchmark, including pseudoknots. Generating seed sequences with \ourprog{}, we demonstrate significant improvements over the previously best multi-target sampling strategy.\\
   \textbf{Availability:} Our software is freely available at: \url{https://github.com/yannponty/RNARedPrint}\\
@@
 
+\enlargethispage{35.1pt}
 \section{Introduction}
@@
 macromolecules with highly specific properties and programmable functions.
-In particular, RNAs present themselves as well-suited tools for
+RNA present itself as a well-suited tool for a
 rational design targeting specific functions~\citep{Kushwaha2016}. RNA function is tightly
 coupled to the formation of secondary structure, as well as changes in
-base pairing propensities and the accessibility of regions, e.g. by
+base pairing propensities and the accessibility of regions, \emph{eg} by
 burying or exposing interaction sites~\citep{Rodrigo2014}. At the same time, the
@@
 computationally tractable~\citep{McCaskill1990}. Thus,  structure can serve as effective
-proxy within rational design approaches, ultimately targeting catalytic~\citep{Zhang2013} or regulatory~\citep{Rodrigo2014} functions.
+proxy for rational design approaches, ultimately targeting catalytic~\citep{Zhang2013} or regulatory functions.
 
@@
 \subsection{Generating high-quality seeds for further optimization}
-We empirically study the capacity of \ourprog{} to improve the generation of seed sequences for subsequent local optimizations, showing an important application in biologically relevant scenarii.
-
-For a multi-target design instance, individual reasonable target
-energies are estimated from averaging the energy over samples at
-relatively high weights $e^{\beta}$, setting the targeted \GCb\%
-to 60\%. For each instance of the \Software{Modena}
-benchmark~\citep{Taneda2015}, we estimated such target energies and
-subsequently generated $1\,000$ seed sequences at these energies.
-Our procedure is tailored to produce sequences with similar Turner
-energy that favor the stability of the target structures having
-moderate \GCb\%; all these properties are desirable objectives
-for RNA design.  All sampled sequences are evaluated under the
-multi-stable design objective function of \citet{Hammer2017}:
+Next, we empirically evaluated the capacity of \ourprog{} to generate seed sequences for multiple (pseudoknotted) target structures, possibly followed by subsequent local optimizations.
+As a baseline for comparison, we considered \RNAblueprint~\citep{Hammer2017}, the current leading tool for multiple design.
+As a quality measure, we used the multi-stable design objective function introduced by~\citet{Hammer2017}, defined as:
 % \begin{align}
@@
   \label{eq:blueprintobjective}
-    f(S)= \quad & \frac{1}{k} \sum_{\ell=1}^{k} (E(S, R_\ell) - G(S))\notag\\
-   & +\ 0.5\frac{1}{\binom{k}{2}} \sum\limits_{1\leq\ell<j\leq k}|E(S,R_\ell) - E(S,R_j)|.
+    \Obj(S)= \quad & \frac{1}{k} \sum_{\ell=1}^{k} (E(S, R_\ell) - \MFE(S))\notag\\
+   & +\frac{1}{2\binom{k}{2}} \sum\limits_{1\leq\ell<j\leq k}|E(S,R_\ell) - E(S,R_j)|.
 \end{align}
-
-Using our strategy, we generated at least 1\,000 seeds per instance of
-the subsets \texttt{RNAtabupath} (2str), \texttt{3str}, and
-\texttt{4str}, of the \Software{Modena} benchmark set with 2,3, and 4
-structures. The results are compared, in terms of their value for the
-objective function of Eq.~\eqref{eq:blueprintobjective}, to seed
-sequences uniformly sampled using \RNAblueprint{}~\citep{Hammer2017}
-(Table~\ref{tab:benchmark-results}, seeds). Our first analysis reveals
-that Boltzmann-sampled sequences constitute better seeds
-(10.97 improvement of mean scores) compared to uniform seeds for
-\emph{all instances} of our benchmark (Supp. Mat.~\ref{sec:validity}).
-
-Moreover, for each generated seed sequence, we optimized the objective
-function $(\ref{eq:blueprintobjective})$ through an adaptive greedy
-walk (500 steps for the pseudoknot free benchmark sets 2str, 3str, and
-4str; 200 for the pseudoknot benchmarksets PK60, PK80, and LE80). At
-each step, we resampled (uniformly random) the positions of a randomly
-selected component in the base pair dependency graph, and accepted the
-modification only if it resulted in a gain, as described in
-\RNAblueprint{}.
-%
-Once again, for all instances, we observe a positive improvement of
-the quality of Boltzmann designs over uniform ones, suggesting a
-superior potential for optimization (2.97 improvement of mean
-scores, Table~\ref{tab:benchmark-results}, ``Optimized''). Notably,
-our subsequent optimization runs, which are directly inherited from
-the uniform sampling \RNAblueprint{} software, partially level the
+where $\MFE(S)$ is the Minimum Free-Energy of $S$, computed by {\tt RNAfold}~\citep{Hofacker1994} or approximated by {\tt HotKnots}~\citep{Ren2005} for pseudoknotted targets.
+Intuitively, the first term of \Obj{} captures how distant from the $\MFE$ the targets are, while the second term penalizes the dispersion of targets.
+\Obj{} is best (minimized) when all targets simultaneously achieve the $\MFE$ of the sequence.
+
+We considered a benchmark of six sets of target structures described in \citet{Taneda2015}: {\sf 2str}, {\sf 3str}, and {\sf 4str} consist of non-pseudoknotted structures, while {\sf PK60}, {\sf PK80}, and {\sf LE80} contain pseudoknotted structures.
+%%For each instance of the \Software{Modena}
+%benchmark~\citep{Taneda2015}, we generated $1\,000$ seed sequences at these energies.
+%Our procedure is tailored to produce sequences with similar Turner
+%energy that favor the stability of the target structures having
+%moderate \GCb\%; all these properties are desirable objectives
+%for RNA design. 
+Using \ourprog{}, we generated at least 1\,000 seeds per instance for each instance 
+of the benchmark. We used target energies that were computed by averaging the energy over samples at
+higher weights $e^{\beta}$, and set the targeted \GCb\%
+to 60\%.
+ We compared the \Obj{} value of 
+these sequences against that of seed sequences, uniformly sampled 
+using \RNAblueprint{}.
+Moreover, for both sets, we used an adaptive greedy
+walk~\citep{Hammer2017} to \Def{optimize} the \Obj{} function. At
+each step, the local search resamples (uniformly at random) the positions 
+of a randomly selected component in the base pair dependency graph, 
+accepting the modification only if it results in a gain. The greedy descent
+was run for 500 steps in the case of pseudoknot-free datasets {\sf 2str}, {\sf 3str}, and
+{\sf 4str}; and 200 steps for the pseudoknotted ones {\sf PK60}, {\sf PK80}, and {\sf LE80}.
+
+The results, shown in Fig.~\ref{fig:benchmark-results}, confirm 
+that Boltzmann-sampled sequences outperform uniform seeds on every dataset, 
+leading to average improvements in \Obj{} values ranging from $7.26$ 
+({\sf LE80}) to $16.05$  ({\sf 2str}) units. Remarkably, this improvement is
+be observed for both of the terms in \Obj{} (see Supp. Mat.~\ref{sec:improvements}). 
+This means that \ourprog{} produces sequences whose targets are both substantially closer to their MFE, and more 
+similarly stable across targets. 
+In fact, for every sequence in our benchmark, consisting of  332 sets of 
+target structures in total, we observed better \Obj{} for Boltzmann-weighted sampling 
+than for uniform sampling (see Supp. Mat.~\ref{sec:validity}). 
+Notably, \ourprog{} remains 
+unfazed by the presence of pseudoknots, the main problem being the computation 
+of the \Obj{} function, which requires a costly free-energy minimization 
+in the presence of pseudoknots~\citep{Sheikh2012}.
+
+%
+Moreover, for all instances as well, the superiority of Boltzmann designs 
+remains unchallenged even after local optimizations, as shown by 
+Fig.~\ref{fig:benchmark-results} and Supp. Mat.~\ref{sec:improvements} 
+and \ref{sec:validity}. This observation is consistent with a 
+superior quality of the starting point for the greedy walk, probably leading 
+to better local minima of the \Obj{} function. However, it should be noted
+that the greedy walk is based on the uniform sampling of
+\RNAblueprint{}, and thus can be expected to partially level the 
 advantages of Boltzmann sampling. In future work, we hope to improve
-this aspect by exploiting Boltzmann sampling even during the
+this aspect by exploiting Boltzmann sampling during the
 optimization run.
 
-\begin{table}[t]
-\centering
-\medskip
-\begin{tabular}{@{}>{\bf}l@{\quad}>{\tt}l@{\quad}@{\quad}c@{\quad}c@{\quad}c@{}}
-  &   \textbf{\textrm{Dataset}}   & {\bfseries\ourprog{}} & \textbf{Uniform} & \textbf{$\Delta$-Improvement}\\
-  Seeds      & 2str & 21.69 ($\pm$4.37) & 37.74 ($\pm$6.45) & 16.05\\
-             & 3str & 18.10 ($\pm$3.98) & 30.49 ($\pm$5.41) & 12.39\\
-             & 4str & 20.15 ($\pm$3.83) & 32.32 ($\pm$5.23) & 12.17\\
-             \cline{2-5}
-             & PK60 & 12.12 ($\pm$3.04) & 19.81 ($\pm$4.24) & 7.69\\
-             & PK80 & 13.81 ($\pm$3.63) & 24.05 ($\pm$4.93) & 10.24\\
-             & LE80 & 10.54 ($\pm$2.77) & 17.80 ($\pm$4.17) & 7.26\\
-             \midrule
-
-  Optimized  & 2str & 5.42 ($\pm$1.27) & 7.95 ($\pm$1.75) & 2.52\\
-             & 3str & 5.10 ($\pm$1.10) & 7.04 ($\pm$1.52) & 1.94\\
-             & 4str & 9.17($\pm$1.51) & 13.25 ($\pm$2.13) & 4.17\\
-             \cline{2-5}
-             & PK60* & 3.75 ($\pm$1.19) & 6.45 ($\pm$1.84) & 2.58\\
-             & PK80* & 3.52 ($\pm$1.58) & 7.99 ($\pm$2.35) & 4.34\\
-             & LE80  & 3.59 ($\pm$1.11) & 5.88 ($\pm$1.59) & 2.28\\
-             \bottomrule
-\end{tabular}\\[1em]
-
-\caption{Comparison of the multi-stable design objective function
-  (Eq.~(\ref{eq:blueprintobjective}) of sequences sampled by
-  \ourprog{} vs.\ uniform samples for three benchmark sets; before
-  (seeds) and after optimization (optimized). We report the average
-  scores with standard deviations; smaller scores are better. The
-  final column reports our improvements due to Boltzmann sampling over
-  uniform sampling (as difference of mean scores).}
-\label{tab:benchmark-results}
-\end{table}
+%\begin{table}[t]
+%\centering
+%\medskip
+%\begin{tabular}{@{}>{\bf}l@{\quad}>{\tt}l@{\quad}@{\quad}c@{\quad}c@{\quad}c@{}}
+%  &   \textbf{\textrm{Dataset}}   & {\bfseries\ourprog{}} & \textbf{\RNAblueprint} & \textbf{$\Delta$-Improvement}\\ \toprule
+%  Seeds      & 2str & 21.69 ($\pm$4.37) & 37.74 ($\pm$6.45) & 16.05\\
+%             & 3str & 18.10 ($\pm$3.98) & 30.49 ($\pm$5.41) & 12.39\\
+%             & 4str & 20.15 ($\pm$3.83) & 32.32 ($\pm$5.23) & 12.17\\
+%             \cline{2-5}
+%             & PK60 & 12.12 ($\pm$3.04) & 19.81 ($\pm$4.24) & 7.69\\
+%             & PK80 & 13.81 ($\pm$3.63) & 24.05 ($\pm$4.93) & 10.24\\
+%             & LE80 & 10.54 ($\pm$2.77) & 17.80 ($\pm$4.17) & 7.26\\
+%             \midrule
+%
+%  Optimized  & 2str & 5.42 ($\pm$1.27) & 7.95 ($\pm$1.75) & 2.52\\
+%             & 3str & 5.10 ($\pm$1.10) & 7.04 ($\pm$1.52) & 1.94\\
+%             & 4str & 9.17($\pm$1.51) & 13.25 ($\pm$2.13) & 4.17\\
+%             \cline{2-5}
+%             & PK60  & 3.84 ($\pm$1.22) & 6.45 ($\pm$1.84) & 2.60\\
+%             & PK80* & 3.41 ($\pm$1.59) & 7.99 ($\pm$2.35) & 4.42\\
+%             & LE80  & 3.59 ($\pm$1.11) & 5.88 ($\pm$1.59) & 2.28\\
+%             \bottomrule
+%\end{tabular}\\[1em]
+%
+%\caption{Comparison of the multi-stable design objective function
+%  (Eq.~(\ref{eq:blueprintobjective}) of sequences sampled by
+%  \ourprog{} vs.\ uniform samples for three benchmark sets; before
+%  (seeds) and after optimization (optimized). We report the average
+%  scores with standard deviations; smaller scores are better. The
+%  final column reports our improvements due to Boltzmann sampling over
+%  uniform sampling (as difference of mean scores).}
+%\label{tab:benchmark-results}
+%\end{table}
+
+\begin{figure}
+  {\includegraphics[width=\linewidth,trim={.2cm .3cm .2cm 5cm},clip]{Figs/statistics-overall}}
+  \caption{Comparison of the \Obj{}
+  (see Eq.~\eqref{eq:blueprintobjective}; smaller values $\to$ better designs) for sequences sampled by
+  \ourprog{} and uniform sampling (\RNAblueprint) for six benchmark sets.
+For both sampling scheme, resulting sequences are either directly reported, or futher optimized using a local search heuristics.}
+\label{fig:benchmark-results}
+\end{figure}
 
@@
 Motivated by the---here established---hardness of the problem, we
-introduced a general framework and efficient algorithms for design of
-RNA sequences to multiple target structures with fine-tuned
+introduced a general framework, and efficient algorithms to design RNA 
+sequences that may adopt multiple target structures with fine-tuned
 properties. Our method combines an FPT stochastic sampling algorithm
@@
 approach generated significantly better seed sequences for instances
-of a common multi-target design benchmark.
-
-The presented method enables new possibilities for sequence generation
-in the field of RNA sequence design by enforcing additional
+of an extensive multi-target design benchmark, including pseudoknots.
+
+The presented method enables new possibilities for the sequence generation
+in the field of RNA sequence design, allowing to enforce additional
 constraints, like \GCb\%, while controlling the energy of
@@
 strategies. We have shown the practicality of such controlled sequence
-generation and studied its use for multi-target RNA design. Moreover,
-our framework is equipped to include more complex sequence
+generation, including the case of pseudoknotted structures, and studied its use for multi-target RNA design. Moreover,
+our framework is amenable to extensions, to include more complex sequence
 constraints, including mandatory/forbidden motifs at specific
-positions or anywhere in the designed sequences. Furthermore, the
-method can support even negative design principles, for instance by
-penalizing a set of alternative helices/structures. Based on this
-generality, we envision our framework---in extensions, which still
-require delicate elaboration---to support various complex RNA design
-scenarios far beyond the sketched applications.
-
-\Final{
-Extensions:
-\begin{enumerate}
-  \item What if the input graph is not bipartite? \\
-  Extract a minimal subset of edges such that new graph is bipartite!\\
-  $\to{}$ Graph bipartization (cf implementation by Hüffner, JGAA 2009)
-  \item Is the optimization version in the \emph{weighted base-pair model} already {\sf NP}-hard?
-  \item Does it help to know that there are cliques of size $k$ during the tree decomposition?\\
-  If anything, it provides a lower bound on tree decomposition (helps branch and bound algorithms)
-  \item What if some degree of freedom is allowed in the structure definition?\\ Can we still use tree decomposition?\\
-  This looks like a very reasonable first step towards the definition of an {\em optimal} DP decomposition for RNA folding (with Pseudoknots) or RNA-RNA interactions\ldots
-  \item Structures can also be penalized, but they must be specified explicitly.\\
-  However, any competing structure must consist of helices, and it is possible to penalize each of the $\Theta(n^2)$ helices of any given length $L$. What would be the impact on treewidth?
-  \item More general energy  models $\to$ Turner. (Pseudoknots)
-  For this description, we assume that all input structures are non-crossing, while the framework would otherwise support crossing input structures as well as according energy models, e.g.{} the Hotknots model of~\citet{Ren2005}, by considering special pseudoknot loops.
-\end{enumerate}\label{lbl:theend}
-}
+positions, or anywhere in the designed sequences, by adapting formal language constructs 
+introduced by \citet{Zhou2013}. Negative design principles could also 
+be explicitly supported at the generation stage, for instance by
+penalizing a set of alternative helices/structures.
+
+%\Final{
+%Extensions:
+%\begin{enumerate}
+%  \item What if the input graph is not bipartite? \\
+%  Extract a minimal subset of edges such that new graph is bipartite!\\
+%  $\to{}$ Graph bipartization (cf implementation by Hüffner, JGAA 2009)
+%  \item Is the optimization version in the \emph{weighted base-pair model} already {\sf NP}-hard?
+%  \item Does it help to know that there are cliques of size $k$ during the tree decomposition?\\
+%  If anything, it provides a lower bound on tree decomposition (helps branch and bound algorithms)
+%  \item What if some degree of freedom is allowed in the structure definition?\\ Can we still use tree decomposition?\\
+%  This looks like a very reasonable first step towards the definition of an {\em optimal} DP decomposition for RNA folding (with Pseudoknots) or RNA-RNA interactions\ldots
+%  \item Structures can also be penalized, but they must be specified explicitly.\\
+%  However, any competing structure must consist of helices, and it is possible to penalize each of the $\Theta(n^2)$ helices of any given length $L$. What would be the impact on treewidth?
+%  \item More general energy  models $\to$ Turner. (Pseudoknots)
+%  For this description, we assume that all input structures are non-crossing, while the framework would otherwise support crossing input structures as well as according energy models, e.g.{} the Hotknots model of~\citet{Ren2005}, by considering special pseudoknot loops.
+%\end{enumerate}\label{lbl:theend}
+%}
 

### 2018-05-23 supplement.tex
Finished revamping results section

@@
 \PassOptionsToPackage{utf8}{inputenc}
-\documentclass{bioinfo}
+\documentclass[12pt]{bioinfo}
 \copyrightyear{2015} \pubyear{2015}
@@
 \newpage
-\section{Detailed result of quality analysis}\label{sec:validity}
+\section{Detailed result of MultiDefect analysis}\label{sec:validity}
 N/A values correspond to data that could not be obtained by the time of this submission for two main reasons: In the case of the initial sampling (left columns), they correspond to instances which, in conjunction with an expressive energy model, resulted in very high tree-width, leading to unreasonable memory requirements incompatible with our available computing resources; The case of missing values after optimization  (right columns), they indicate situations where the initial optimization took too long ($\approx$ 1 day) and was killed.
 \input{table}
+
+\section{Contribution of MultiDefect terms to the improvement}
+\label{sec:improvements}
+\begin{figure}[h!]
+\includegraphics[width=.48\textwidth]{Figs/statistics-Term1}
+\includegraphics[width=.48\textwidth]{Figs/statistics-Term2}
+\caption{Break down of average MultiDefect values into their two components.}
+\end{figure}
 

### 2018-05-24 RNARedPrint-Bioinfo2018.tex
Minor edits of 4.2 and suggested fix of MFE vs. G in MultiDefect function

@@
 \documentclass{bioinfo}
-\copyrightyear{2015} \pubyear{2015}
+\copyrightyear{2018} \pubyear{2018}
 
@@
 
-\usepackage{todonotes}
+%\usepackage{todonotes}
 
@@
 \newcommand{\TargetE}{E^{\star}}
-\newcommand{\MFE}{\text{\rm MFE}}
+%\newcommand{\MFE}{\text{\rm MFE}}
+\newcommand{\EnsE}{\text{\rm G}} %ensemble energy
 
@@
 %%% end macro defs
+
+%% fix some weirdness of the bioinfo class (footer overlapping text on page 1)
+\makeatletter
+\setlength\footskip{30\p@}          % space above footer line
+\makeatother
 
@@
     Targeting specific energies for pseudoknotted structures using
-    multi-dimensional Boltzmann sampling.  (A) Linear fits between the
+    multi-dimensional Boltzmann sampling.  A) Linear fits between the
     energies in the stacking model to the realistic pseudoknot energy
@@
     targeting stacking model energies.  We show the fits for both
-    target structures S1 and S2 of subfigure B.  (B) D\&P energy
+    target structures S1 and S2 of subfigure B.  B) D\&P energy
     distributions for two target structures S1 and S2. We show the
@@
   \label{eq:blueprintobjective}
-    \Obj(S)= \quad & \frac{1}{k} \sum_{\ell=1}^{k} (E(S, R_\ell) - \MFE(S))\notag\\
+    \Obj(S)= \quad & \frac{1}{k} \sum_{\ell=1}^{k} (E(S, R_\ell) - \EnsE(S))\notag\\
    & +\frac{1}{2\binom{k}{2}} \sum\limits_{1\leq\ell<j\leq k}|E(S,R_\ell) - E(S,R_j)|.
 \end{align}
-where $\MFE(S)$ is the Minimum Free-Energy of $S$, computed by {\tt RNAfold}~\citep{Hofacker1994} or approximated by {\tt HotKnots}~\citep{Ren2005} for pseudoknotted targets.
-Intuitively, the first term of \Obj{} captures how distant from the $\MFE$ the targets are, while the second term penalizes the dispersion of targets.
-\Obj{} is best (minimized) when all targets simultaneously achieve the $\MFE$ of the sequence.
+where $\EnsE(S)$, the \emph{ensemble free energy} of $S$, is computed
+by {\tt RNAfold}~\citep{Hofacker1994} in the pseudoknot-free case; for
+pseudoknotted targets, $\EnsE(S)$ is approximated by the \emph{minimum
+  free energy} of S as estimated by {\tt HotKnots} in the energy model
+of~\citet{dirks-pierce-03}.  Intuitively, the first term of \Obj{}
+captures the distance of the targets from the ensemble free energy,
+while the second term penalizes the dispersion of targets; \Obj{} is
+best (minimal) when all targets simultaneously achieve the minimum
+free energy of the sequence.
 
@@
 %for RNA design. 
-Using \ourprog{}, we generated at least 1\,000 seeds per instance for each instance 
-of the benchmark. We used target energies that were computed by averaging the energy over samples at
-higher weights $e^{\beta}$, and set the targeted \GCb\%
-to 60\%.
- We compared the \Obj{} value of 
-these sequences against that of seed sequences, uniformly sampled 
-using \RNAblueprint{}.
-Moreover, for both sets, we used an adaptive greedy
-walk~\citep{Hammer2017} to \Def{optimize} the \Obj{} function. At
-each step, the local search resamples (uniformly at random) the positions 
-of a randomly selected component in the base pair dependency graph, 
-accepting the modification only if it results in a gain. The greedy descent
-was run for 500 steps in the case of pseudoknot-free datasets {\sf 2str}, {\sf 3str}, and
-{\sf 4str}; and 200 steps for the pseudoknotted ones {\sf PK60}, {\sf PK80}, and {\sf LE80}.
-
-The results, shown in Fig.~\ref{fig:benchmark-results}, confirm 
-that Boltzmann-sampled sequences outperform uniform seeds on every dataset, 
+Using \ourprog{}, we generated at least 1\,000 seeds per instance for
+each instance of the benchmark. We used target energies that were
+computed by averaging the energy over samples at higher weights
+$e^{\beta}$, and set the targeted \GCb\% to 60\%.  We compared the
+\Obj{} value of these sequences against that of seed sequences,
+uniformly sampled using \RNAblueprint{}.  Moreover, for both sets, we
+used an adaptive greedy walk~\citep{Hammer2017} to \Def{minimize} the
+\Obj{} function. At each step, the local search resamples (uniformly
+at random) the positions of a randomly selected component in the base
+pair dependency graph, accepting the modification only if it results
+in a gain. We performed 500 greedy descent steps in the case of
+pseudoknot-free datasets {\sf 2str}, {\sf 3str}, and {\sf 4str}; and
+200 steps for the pseudoknotted ones {\sf PK60}, {\sf PK80}, and {\sf
+  LE80}.
+
+The results, shown in Fig.~\ref{fig:benchmark-results}, confirm that
+Boltzmann-sampled sequences outperform uniform seeds on every dataset,
 leading to average improvements in \Obj{} values ranging from $7.26$
-({\sf LE80}) to $16.05$  ({\sf 2str}) units. Remarkably, this improvement is
-be observed for both of the terms in \Obj{} (see Supp. Mat.~\ref{sec:improvements}). 
-This means that \ourprog{} produces sequences whose targets are both substantially closer to their MFE, and more 
-similarly stable across targets. 
-In fact, for every sequence in our benchmark, consisting of  332 sets of 
-target structures in total, we observed better \Obj{} for Boltzmann-weighted sampling 
-than for uniform sampling (see Supp. Mat.~\ref{sec:validity}). 
-Notably, \ourprog{} remains 
-unfazed by the presence of pseudoknots, the main problem being the computation 
-of the \Obj{} function, which requires a costly free-energy minimization 
-in the presence of pseudoknots~\citep{Sheikh2012}.
+({\sf LE80}) to $16.05$ ({\sf 2str}) units. Remarkably, this
+improvement is be observed for both of the terms in \Obj{} (see
+Supp. Mat.~\ref{sec:improvements}).  This means that \ourprog{}
+produces sequences whose targets are both substantially closer to
+their MFE, and more similarly stable across targets.  In fact, for
+every sequence in our benchmark, consisting of 332 sets of target
+structures, we observed better \Obj{} for Boltzmann sampling than for
+uniform sampling (see Supp. Mat.~\ref{sec:validity}).  Notably,
+\ourprog{} performs equally well in the presence of pseudoknots;
+the difficulty rather lies in the computation of the \Obj{} function, since
+free-energy minimization is costly in the presence of
+pseudoknots~\citep{Sheikh2012} and good implementations are scarce.
 

### 2018-05-24 RNARedPrint-Bioinfo2018.tex
Edit caption of new results figure

@@
   {\includegraphics[width=\linewidth,trim={.2cm .3cm .2cm 5cm},clip]{Figs/statistics-overall}}
-  \caption{Comparison of the \Obj{}
-  (see Eq.~\eqref{eq:blueprintobjective}; smaller values $\to$ better designs) for sequences sampled by
-  \ourprog{} and uniform sampling (\RNAblueprint) for six benchmark sets.
-For both sampling scheme, resulting sequences are either directly reported, or futher optimized using a local search heuristics.}
+  \caption{Comparison of the \Obj{} (see
+    Eq.~\eqref{eq:blueprintobjective}; smaller values $\to$ better
+    designs) for sequences sampled by \ourprog{} and uniform sampling
+    (\RNAblueprint) for six benchmark sets. For both sampling schemes,
+    we show the MultiDefects of the sampled sequences, as well as
+    results after further optimization by local search (``+ Opt.'').}
 \label{fig:benchmark-results}

### 2018-05-24 RNARedPrint-Bioinfo2018.tex
Minor edit

@@
 
-\newcommand{\RNAblueprint}{{\tt \bfseries{}\color{black!75} RNA\textcolor{blue!70!black}{Blue}Print}}
-\newcommand{\ourprog}{{\tt \bfseries{}\color{black!75}RNA\textcolor{red!70!black}{Red}Print}}
+\newcommand{\RNAblueprint}{{\tt \bfseries{}\color{black!85} RNA\textcolor{blue!70!black}{Blue}Print}}
+\newcommand{\ourprog}{{\tt \bfseries{}\color{black!85}RNA\textcolor{red!70!black}{Red}Print}}
 
@@
 The expected value of an energy $E_\ell$ is estimated as $\hat\mu_\ell(\mathcal{S}) = \sum_{S\in\mathcal{S}}E_\ell(S)/|\mathcal{S}|$, and the weights are updated at the $t$-th iteration by %  $\pi_\ell^{[t+1]} = \pi_\ell^{[t]}\cdot\TargetE_\ell/\hat\mu_\ell(\mathcal{S})$.
-$\pi_\ell^{[t+1]} = \pi_\ell^{[t]}\cdot \gamma^{\hat\mu_\ell(\mathcal{S})-\TargetE_\ell}$. In practice, the constant $\gamma>1$ is chosen empirically to achieve effective optimization.
-While heuristic in nature, this basic iteration was elected in our initial version of \ourprog{} because of its reasonable empirical behavior (choosing $\gamma=1.2$).
+$\pi_\ell^{[t+1]} = \pi_\ell^{[t]}\cdot \gamma^{\hat\mu_\ell(\mathcal{S})-\TargetE_\ell}$. In practice, the constant $\gamma>1$ is chosen empirically  ($\gamma=1.2$) to achieve effective optimization.
+While heuristic in nature, this basic iteration was elected in our initial version of \ourprog{} because of its good empirical behavior.
 
@@
   \label{eq:blueprintobjective}
-    \Obj(S)= \quad & \frac{1}{k} \sum_{\ell=1}^{k} (E(S, R_\ell) - \EnsE(S))\notag\\
+    \Obj(S) &= \frac{1}{k} \sum_{\ell=1}^{k} (E(S, R_\ell) - \EnsE(S))\notag\\
    & +\frac{1}{2\binom{k}{2}} \sum\limits_{1\leq\ell<j\leq k}|E(S,R_\ell) - E(S,R_j)|.
 \end{align}
-where $\EnsE(S)$, the \emph{ensemble free energy} of $S$, is computed
+where $\EnsE(S)$, the \Def{ensemble free energy} of $S$, is computed
 by {\tt RNAfold}~\citep{Hofacker1994} in the pseudoknot-free case; for
-pseudoknotted targets, $\EnsE(S)$ is approximated by the \emph{minimum
-  free energy} of S as estimated by {\tt HotKnots} in the energy model
+pseudoknotted targets, $\EnsE(S)$ is approximated by the \Def{minimum
+  free energy} (MFE) of S as estimated by {\tt HotKnots} in the energy model
 of~\citet{dirks-pierce-03}.  Intuitively, the first term of \Obj{}
@@
 
-The results, shown in Fig.~\ref{fig:benchmark-results}, confirm that
+The results, shown in Fig.~\ref{fig:benchmark-results}, reveal that
 Boltzmann-sampled sequences outperform uniform seeds on every dataset,
@@
 produces sequences whose targets are both substantially closer to
-their MFE, and more similarly stable across targets.  In fact, for
+the sequence ensemble/minimal free energy, and more similarly stable across targets.  In fact, for
 every sequence in our benchmark, consisting of 332 sets of target
@@
 \begin{figure}
-  {\includegraphics[width=\linewidth,trim={.2cm .3cm .2cm 5cm},clip]{Figs/statistics-overall}}
+  {\includegraphics[width=.9\linewidth,trim={.2cm .3cm .2cm 5cm},clip]{Figs/statistics-overall}}
   \caption{Comparison of the \Obj{} (see
@@
 
-\section{Counting valid designs is \#{\sf P}-hard}\label{sec:counting}
-
+\section{\#{\sf P}-hardness of counting valid designs}\label{sec:counting}
 %%% reformulate later

### 2018-05-24 RNARedPrint-Bioinfo2018.tex
changed citation

@@
 \end{align}
+
 where $\EnsE(S)$, the \Def{ensemble free energy} of $S$, is computed
-by {\tt RNAfold}~\citep{Hofacker1994} in the pseudoknot-free case; for
+by {\tt RNAfold}~\citep{Lorenz2011} in the pseudoknot-free case; for
 pseudoknotted targets, $\EnsE(S)$ is approximated by the \Def{minimum

### 2018-08-28 RNARedPrint-Bioinfo2018.tex
Work on comments of reviewer R1---leave some open todo notes

@@
 
-%\usepackage{todonotes}
+\usepackage{todonotes}
 
@@
 
-\newcommand{\Def}[1]{{\bfseries #1}}
+\newcommand{\Def}[1]{\emph{#1}}
 
@@
 coupled to the formation of secondary structure, as well as changes in
-base pairing propensities and the accessibility of regions, \emph{eg} by
+base pairing propensities and the accessibility of regions, e.g.\ by
 burying or exposing interaction sites~\citep{Rodrigo2014}. At the same time, the
@@
 \GCb\% in designed sequences. To address this issue,
-\Software{IncaRNAtion}~\citep{Reinharz2013} controls \GCb\%
-through an adaptive sampling strategy.
+tools like \Software{antaRNA}~\citep{Kleinkauf2015} and \Software{IncaRNAtion}~\citep{Reinharz2013} control \GCb\%; the latter based on adaptive sampling.
 
 \parHead{Specificities and similarities of multi-target design.}
-Specifically, for multi-target design, virtually all available methods~\citep{Lyngsoe2012,HoenerzuSiederdissen2013,Taneda2015,Hammer2017} follow the same overall generation/optimization scheme.
-%
-Facing the complex sequence constraints induced by multiple targets, early methods such as \Software{Frnakenstein}~\citep{Lyngsoe2012} and \Software{Modena}~\citep{Taneda2015} did not attempt to solve sequence generation systematically, but rely on \emph{ad-hoc} sampling strategies.
+Specifically, for multi-target design, virtually all available methods~\citep{Lyngsoe2012,HoenerzuSiederdissen2013,Taneda2015,Hammer2017} follow the same overall generate-and-optimize strategy.
+%
+Facing the complex sequence constraints induced by multiple targets, early methods such as \Software{Frnakenstein}~\citep{Lyngsoe2012} and \Software{Modena}~\citep{Taneda2015} did not attempt to solve sequence generation systematically, but rely on ad-hoc sampling strategies.
 %
@@
 To answer these questions, we introduce a generic framework (illustrated in~Fig.~\ref{fig:workflow}) enabling efficient Boltzmann-weighted sampling over RNA sequences with multiple target structures (Section~\ref{sec:FPT}).
-Guided by a \Def{tree decomposition} of the network, we devise dynamic programming to compute partition functions and sample sequences from the Boltzmann distribution%
+Guided by a \Def{tree decomposition}, we devise dynamic programming to compute partition functions and sample sequences from the Boltzmann distribution%
 % (Subsection~\ref{sec:PF})
@@
 to a large set of multi-target RNA design instances from a
-representative benchmark Section~\ref{sec:results}).
+representative benchmark (Section~\ref{sec:results}).
 
@@
   energy models---including nearest neighbor models,
-  e.g.~\citet{Turner2009}, and even pseudoknot models,
-  e.g. \citet{Andronescu2010}, while bounding the dependencies to
+  e.g.~\citep{Turner2009}, and even pseudoknot models,
+  e.g. \citep{Andronescu2010}, while bounding the dependencies to
   sequence positions introduced by each single $f$. Formally, define
@@
   positions $\I\subseteq\{1,\dots,n\}$, where $f(S)=f(S')$ for all
-  sequences $S$ and $S'$ that agree at the positions in $\I$.
+  sequences $S$ and $S'$ that have equal nucleotides at the positions in $\I$.
 
@@
   each $v\in T$) $\chi(v)\subseteq X$ is a set of vertices assigned to
-  the node tree $v\in T$, such that
+  the node $v\in T$, such that
 \begin{enumerate}
@@
 
-\section{An FPT algorithm for the partition function and sampling of Boltzmann-weighted designs}
+\section{An fixed-parameter tractable (FPT) algorithm for the
+  partition function and sampling of Boltzmann-weighted designs}
 \label{sec:FPT}
@@
 The minimum energy, counting, and partition function
-over RNA design network can be computed by dynamic programming based
+over an RNA design network can be computed by dynamic programming based
 on a tree decomposition of the network's dependency graph
@@
 in $\Y$; furthermore, given partial sequences $\val$ and
-$\val'$, we define the \Def{combined partial sequence $\substitute{\val'}{\val''}$} such that
+$\val'$, we define the \Def{combined partial sequence $\substitute{\val}{\val'}$} such that
 $$
@@
 
-Let us now consider the \Def{computation of the partition function}.
+Let us now consider the computation of the partition function.
 Given is the RNA design network $\network=(\X,\F)$ and its cluster
@@
 messages along these directed edges $u\to v$ (i.e. always from some
-child $u$ to its parent $v$). Each message is a function that depends
+child $u$ to its parent $v$). Each message $m$ is a function that depends
 on the positions $\dep(m)\subseteq \X$ and yields a partition function
@@
         $p \gets$ product( $exp(-\beta \evalfor{f}{\substitute{\val}{\val'}})$ for $f\in \phi(u)$ )\\
-     ${}\quad\qquad \cdot\ $product( $\evalfor{\Message{w}{u}}{\substitute{\val}{\val'}}$ for $(w\to{}u)\in T$ )\;
+        ${}\quad\qquad \cdot\ $product( $\evalfor{\Message{u'}{u}}{\substitute{\val}{\val'}}$ for $(u'\to{}u)\in T$ )\;
         $x \gets x + p$\;
@@
   }
- \caption{FPT computation of the partition function using
-   dynamic programming (CTE). }\label{alg:pf}
+  \caption{FPT computation of the partition function using dynamic
+    programming, i.e.~cluster tree elimination
+    \citep[CTE]{Dechter2013}. The postorder traversal guarantees that
+    when processing edge $u\to{}v$, all messages
+    $\Message{u'}{u}$, corresponding to DP matrices, have been computed before. }\label{alg:pf}
 \end{algorithm}
 
-\SetKwProg{Fn}{Function}{}{}
- \SetKwFunction{Sample}{$\sample$}
- \SetKwFunction{Random}{UnifRand}
-
-The partition functions can then direct a \Def{stochastic backtrack} to achieve \Def{Boltzmann sampling of sequences}, such that one samples from the Boltzmann distribution of a given design network $\network$.
-The sampling algorithm assumes that the cluster tree was expanded and the messages $\Message{u}{v}$ for the edges in $\edgesToR{}$ are already generated by Algorithm \ref{alg:pf} for the expanded cluster tree.
-%
-Algorithm~\ref{alg:sampling} defines the recursive procedure
-$\sample(u,\val)$, which returns---randomly drawn from the
-Boltzmann distribution---a partial sequence that determines all
-sequence positions in the subtree rooted at $u$.  Called on $r$ and
-the empty partial sequence, which does not determine any positions,
-the procedure samples a random sequence from the Boltzmann
-distribution.
+The partition functions can then direct a stochastic backtracking procedure to sample sequences from the Boltzmann distribution based on the given design network $\network$.
+%
+For an expanded cluster tree, after the messages $\Message{u}{v}$ for
+the edges in $\edgesToR{}$ are generated by Algorithm \ref{alg:pf},
+one can repeatedly call Algorithm~\ref{alg:sampling}, each time
+randomly drawing another sequence from the Boltzmann distribution.
 
@@
 tree decompositions
-(Supp. Mat.~\ref{appsec:algcomplexity}). W.l.o.g. we assume that
+(Supp. Mat.~\ref{appsec:algcomplexity}). W.l.o.g.\ we assume that
 $D=1$; note that tree decompositions can generally be transformed,
@@
 
+% \begin{algorithm}
+%  \KwData{Node $u$, partial sequence $\val\in\partseqs(\separator{u}{v})$;\newline
+%  Cluster tree $(T,\chi,\phi)$ and partition functions $\Message{u'}{v'}[\val']$, $\forall (u'\to{}v')\in T$ and $\val'\in\partseqs(\separator{u'}{v'})$.}
+%  \KwResult{Boltzmann-distributed random partial sequence for the subtree rooted at $u$, specializing a partial sequence $\val$.}
+%  \Fn{\Sample$(u,\val)$}{
+%    $r \gets \Random(\evalfor{\Message{u}{v}}{\val})$\;
+%    \For{$\val'\in\partseqs(\difference{u}{v})$}{
+
+%      $p \gets$ product( $exp(-\beta \evalfor{f}{\substitute{\val}{\val'}})$ for $f\in \phi(u)$ )\\
+%      ${}\quad\qquad \cdot\ $product( $\evalfor{\Message{u'}{u}}{\substitute{\val}{\val'}}$ for $(u'\to{}u)\in T$ )\;
+%   	  $r \gets r - p$\;
+%   	  \If{$r<0$}{
+%   	    $\val_{\rm res} \gets \substitute{\val}{\val'}$\;
+%   	  \For{$(u'\to{}u)\in T$}{
+%   	      $\val_{\rm res} \gets \substitute{\val_{\rm res}}{\Sample(u',\substitute{\val}{\val'})}$\;
+%      }
+%   	  \Return {$\val_{\rm res}$}
+%   	  \;}
+%  }
+%  }
+%  \caption{Stochastic backtrack algorithm for partial sequences in the Boltzmann distribution.}\label{alg:sampling}
+% \end{algorithm}
+
 \begin{algorithm}
- \KwData{Node $u$, partial sequence $\val\in\partseqs(\separator{u}{v})$;\newline
- Cluster tree $(T,\chi,\phi)$ and partition functions $\Message{u'}{v'}[\val']$, $\forall (u'\to{}v')\in T$ and $\val'\in\partseqs(\separator{u'}{v'})$.}
- \KwResult{Boltzmann-distributed random partial sequence for the subtree rooted at $u$, specializing a partial sequence $\val$.}
- \Fn{\Sample$(u,\val;T,\chi,\phi,m)$}{
-   $r \gets \Random(\evalfor{\Message{u}{v}}{\val})$\;
+ \KwData{Cluster tree $(T,\chi,\phi)$ and partition functions $\Message{u'}{v'}$ for all $(u'\to{}v')\in T$.}
+ \KwResult{Boltzmann-distributed random sequence $\val$.}
+ $\val$ \gets $\varnothing$\;
+ \For{$u\to{}v\in T$ in preorder}{
+   $r \gets \text{uniform random number between $0$ and $\evalfor{\Message{u}{v}}{\val}$}$\;
    \For{$\val'\in\partseqs(\difference{u}{v})$}{
 
-     $p \gets$ product( $exp(-\beta \evalfor{f}{\substitute{\val}{\val'}})$ for $f\in \phi(u)$\\
-     ${}\quad\qquad \cdot\ $product( $\evalfor{\Message{w}{u}}{\substitute{\val}{\val'}}$ for $(w\to{}u)\in T$ )\;
-%
-%     $p \gets 1$\;
-%     \For{$f\in \phi(v)$}{
-%  	      $p \gets p \cdot \exp(-\beta \evalfor{f}{\substitute{\val}{\val'}})$\;
-%     }
-%     \For{$v\to{}w\in T$}{
-%  	      $p \gets p \cdot \evalfor{\Message{v}{w}}{\substitute{\val}{\val'}}$\;
-%     }
+     $p \gets$ product( $exp(-\beta \evalfor{f}{\substitute{\val}{\val'}})$ for $f\in \phi(u)$ )\\
+     ${}\quad\qquad \cdot\ $product( $\evalfor{\Message{u'}{u}}{\substitute{\val}{\val'}}$ for $(u'\to{}u)\in T$ )\;
      $r \gets r - p$\;
      \If{$r<0$}{
-  	    $\val_{\rm res} \gets \substitute{\val}{\val'}$\;
-  	  \For{$(v\to{}w)\in T$}{
-  	      $\val_{\rm res} \gets \substitute{\val_{\rm res}}{\Sample(v,\substitute{\val}{\val'};T,\chi,\phi,m)}$\;
-     }
-  	  \Return {$\val_{\rm res}$}
-  	  \;}
+       $\val \gets \substitute{\val}{\val'}$\;
      }
    }
- \caption{Stochastic backtrack algorithm for partial sequences in the Boltzmann distribution.}\label{alg:sampling}
+ }
+ \Return {$\val$}\;
+
+ \caption{Stochastic backtrack algorithm for partial sequences in the
+   Boltzmann distribution. Processing the edges $u\to{}v\in T$ in
+   preorder ensures that $\val$ invariantly determines all positions
+   of $v$ outside the subtree of $u$.}\label{alg:sampling}
 \end{algorithm}
+
 %For example, for design to a single non-crossing target in the nearest neighbor energy model, the complexity of Algs.~\ref{alg:pf} and \ref{alg:sampling} degrades to linear time (as reported for IncaRNAtion), since there are only $O(n)$ many functions and nodes (in the sequence length $n$).
@@
 \begin{theorem}[Complexities]\label{th:complexities}
-  For sequence length $n$, $k$ target structures, treewidth $w$ and
-  a base pair dependency graph having $c$ connected components, $t$ sequences
+  For sequence length $n$, $k$ target structures and treewidth $w$
+  % and a base pair dependency graph having $c$ connected components%
+  , $t$ sequences
   are generated from the Boltzmann distribution  in
@@
 \end{theorem}
+
+By this theorem, Boltzmann sampling in our setting is fixed parameter tractable (FPT) in the tree width $w$ (since the complexity is polynomial for fixed $w$).
+%
 As shown in Supp. Mat.~\ref{sec:improvedComplexity}, the complexity of the precomputation can be further improved to
@@
   models.}\label{sec:energy_models}
-
+\todo[inline]{Better motivate why we want different energy functions etc}
+%
 The functions in $\F$ allow expressing complex features of the
-sequences alone, e.g. rewarding or penalizing specific sequence
+sequences alone, e.g.\ rewarding or penalizing specific sequence
 motifs, as well as features depending on the target structures.
@@
 
-% To illustrate the expressivity of the framework, we outline the construction of function sets that express two different energy models for the energies $E(S;R_\ell)$ for encoding
-% \begin{equation}
-%   \label{eq:nussinov-network-energy}
-%   \energy{\network}(S) = \sum_{\ell\in[1,k]} w_\ell E(S;R_\ell)
-% \end{equation}
 In simple \Def{base pair-based energy models}, energy is defined as
 the sum of base pair (pseudo-)energies. If base pair energies
-$\Ebp{k}{i,j,x,y}$ (where $i$ and $j$ are sequence positions, $x$ and
-$y$ are bases in $\Sigma$) are given for each target structure $\ell$,
-s.t.  $ E(S;R_\ell) := \sum_{(i,j)\in R_\ell} \Ebp{k}{i,j,S_i,S_j}$,
+$\Ebp{\ell}{i,j,x,y}$ (where $i$ and $j$ are sequence positions, $x$ and
+$y$ are bases in $\Sigma$) are given for each target structure $R_\ell$,
+s.t.  $ E(S;R_\ell) := \sum_{(i,j)\in R_\ell} \Ebp{\ell}{i,j,S_i,S_j}$,
 we encode the network energy by the set of functions $f$ for each base
 pair $(i,j)\in R_\ell$ of each input structure $R_\ell$ that evaluate
-to $\ln(\pi_\ell) \Ebp{k}{i,j,\val(x_i),\val(x_j)}$ under partial sequence $\val$;
+to $\ln(\pi_\ell) \Ebp{\ell}{i,j,\val_i,\val_j}$ under partial sequence $\val$;
 here, $\pi_\ell>0$ is a weight that controls the
-influence of structure $\ell$ on the sampling (as elaborated later).
+influence of the $\ell$-th structure on the sampling (as elaborated later).
 
@@
 More complex \Def{loop-based}
- energy models ---e.g.~the Turner model,
+ energy models---e.g.~the Turner model,
 which among others includes energy terms for special loops and dangling
@@
 where $\pmb{\pi}:=(\pi_1\cdots\pi_k)$ is a vector of positive real-valued \Def{weights}, and $\partfun{\pmb{\pi}}$ is the weighted partition function.
-Such a distribution can be induced by a simple modification of the functions described in Sec.~\ref{sec:energy_models}, where any energy function $E(\val)$ for a structure $\ell$ is replaced by $E'(\val):= \ln(\pi_\ell)\, E(\val)/\beta$.
+Such a distribution can be induced by a simple modification of the functions described in Sec.~\ref{sec:energy_models}, where any energy function $E(\val)$ for a structure $\ell$ is replaced by $E'(\val):= \ln(\pi_\ell)\, E(\val)/\beta$\todo{is $/\beta$ necessary?}.
 The probability of a sequence $S$ is thus proportional to
@@
 
-One then needs to \Def{learn a weights vector} $\pmb{\pi}$ such that, on average, the targeted energies are achieved by a random sequences in the weighted distribution, \emph{i.e.} such that  $\mathbb{E}(E_\ell(S)\mid \pmb{\pi})=\TargetE_\ell$,  $\forall\ell\in[1,k]$.
+One then needs to \Def{learn a weights vector} $\pmb{\pi}$ such that, on average, the targeted energies are achieved by a random sequences in the weighted distribution, i.e.\ such that  $\mathbb{E}(E_\ell(S)\mid \pmb{\pi})=\TargetE_\ell$,  $\forall\ell\in[1,k]$.
 The expected value of $E_\ell$ is always decreasing for increasing weights $\pi_\ell$ (see Supp.
@@
 
-A further \Def{rejection step} is applied to the generated structures to retain only sequences whose energy for each structure $R_\ell$ belongs to $[\TargetE_\ell\cdot(1-\varepsilon),\TargetE_\ell\cdot(1+\varepsilon)]$, for $\varepsilon\ge 0$ some predefined \Def{tolerance}.
+A further \Def{rejection step} is applied to retain only those sequences whose energy for each structure $R_\ell$ belongs to $[\TargetE_\ell\cdot(1-\varepsilon),\TargetE_\ell\cdot(1+\varepsilon)]$, for $\varepsilon\ge 0$ some predefined \Def{tolerance}.
 The rejection approach is justified by the following considerations:
-i) \emph{Enacting an exact control over the energies would  be technically hard and costly.} Indeed, controlling the energies through dynamic programming would require explicit convolution products, generalizing~\citet{Cupal1996}, inducing additional $\Theta(n^{2k})$ time and $\Theta(n^k)$ space overheads;
+i) \emph{Enacting an exact control over the energies would  be technically hard and costly.} Indeed, controlling the energies through dynamic programming would require explicit convolution products, generalizing~\citep{Cupal1996}, inducing additional $\Theta(n^{2k})$ time and $\Theta(n^k)$ space overheads;
 ii) \emph{Induced distributions are typically concentrated.} Intuitively, unless sequences are fully constrained individual energy terms are independent enough so that their sum is concentrated around its mean -- the targeted energy (cf Fig.~\ref{fig:energydist-pk}).
 %Indeed, as shown in Fig.~\ref{fig:shifting_mean}, empirical joint distributions in the energies have good fits towards Normal multidimensional distributions with linear (co-)variances.
-For base pair-based energy models and special base pair dependency graph
+For base pair-based energy models and special base pair dependency graphs
 (paths, cycles\ldots) this property rigorously follows from analytic
-combinatorics, see \citet{Bender1983} and
-\citet{Drmota1997}. In such cases, the expected number of
+combinatorics, see \citep{Bender1983} and
+\citep{Drmota1997}. In such cases, the expected number of
 rejections before reaching the targeted energies remains constant when
@@
 \ref{alg:pf} and \ref{alg:sampling}), to sample valid sequences for
-given target structures and weights $\pi_1,\dots,\pi_n$.  Moreover, we
+given target structures and weights $\pi_1,\dots,\pi_k$.  Moreover, we
 used multi-dimensional Boltzmann sampling, as described in
@@
 model. For the shown fits we sampled $n=1\,000$ sequences, targeting a
-${\sf GC}\%$ of $60\%$.
+\GCb\% of $60\%$.
 %
@@
     Targeting specific energies for pseudoknotted structures using
-    multi-dimensional Boltzmann sampling.  A) Linear fits between the
+    multi-dimensional Boltzmann sampling. \textbf{(A)} Linear fits between the
     energies in the stacking model to the realistic pseudoknot energy
@@
     targeting stacking model energies.  We show the fits for both
-    target structures S1 and S2 of subfigure B.  B) D\&P energy
+    target structures S1 and S2 of subfigure B.  \textbf{(B)} D\&P energy
     distributions for two target structures S1 and S2. We show the
@@
     $(-30,-30)$, $(-25,-25)$, and $(-20,-30)$ for the target
-    structures (annotated in red, yellow, and blue
-    resepectively)---this demonstrates the effectivity of our adaptive
+    structures---this demonstrates the effectivity of our adaptive
     multi-dimensional Boltzmann sampling procedure; moreover, for
@@
   \label{fig:energydist-pk}
+  \todo[inline]{Rename S1,S2 to R1 R2. Discuss R2 values. Add legend to subfig A or explain colors}
 \end{figure*}
+
 \subsection{Generating high-quality seeds for further optimization}
@@
 pseudoknotted targets, $\EnsE(S)$ is approximated by the \Def{minimum
-  free energy} (MFE) of S as estimated by {\tt HotKnots} in the energy model
+  free energy} (MFE) of $S$ as estimated by {\tt HotKnots}~\citep{Ren2005} in the energy model
 of~\citet{dirks-pierce-03}.  Intuitively, the first term of \Obj{}
@@
 
-We considered a benchmark of six sets of target structures described in \citet{Taneda2015}: {\sf 2str}, {\sf 3str}, and {\sf 4str} consist of non-pseudoknotted structures, while {\sf PK60}, {\sf PK80}, and {\sf LE80} contain pseudoknotted structures.
+We considered a benchmark of six sets of target structures described in \citep{Taneda2015}: {\sf 2str}, {\sf 3str}, and {\sf 4str} consist of non-pseudoknotted structures, while {\sf PK60}, {\sf PK80}, and {\sf LE80} contain pseudoknotted structures.
 %%For each instance of the \Software{Modena}
@@
 %for RNA design.
-Using \ourprog{}, we generated at least 1\,000 seeds per instance for
+Using \ourprog{}, we generated at least 1\,000 seeds for
 each instance of the benchmark. We used target energies that were
@@
 
+\todo[inline]{R1: I am sure your "bipartite" definition is wrong (or your proof is broken), ie.
+is the bipartition not about V1 and V2, ie there is no edge WITHIN each vertex set but only BETWEEN the sets?
+Right now, your definition says that the intersection of all possible inter-set-edges (V1xV2) with the present edges E is empty..
+But the proof says in line 7-1-35 that any edge "in E involves two nodes $v1 \in V1$ and $v2 \in V2$".
+So you might correct the bipartite definition to something like "$(V1xV1)\cap E = \emptyset = (V2xV2)\cap E$"  }
+
 To establish that claim, we consider instances $G=(V_1\cup V_2, E)$ that are connected and bipartite ($(V_1\times V_2) \cap E = \varnothing$), noting that hardness over restricted instances implies hardness of the general problem.
-Moreover remark that, as observed in Subsec.~\ref{sec:complexity}, assigning a nucleotide to a position $u\in V$ constrains the parity ($\{\Ab,\Gb\}$ or $\{\Cb,\Ub\}$) of all positions in the connected component of $u$.
+Moreover note that, as observed in Subsec.~\ref{sec:complexity}, assigning a nucleotide to a position $u\in V$ constrains the parity ($\{\Ab,\Gb\}$ or $\{\Cb,\Ub\}$) of all positions in the connected component of $u$.
 For this reason, we restrict our attention to the counting of valid designs \emph{up to trivial  symmetry} $(\Ab\leftrightarrow \Cb/\Gb\leftrightarrow \Ub)$, by constraining the positions in $V_{1}$ to $\Ab$ and $\Gb$.
 Let $\Design{G}$ denote the subset of all designs for $G$ under this constraint, noting that $\NumDesign(G) = 2\cdot|\Design{G}|$.
+
+\todo[inline]{
+R1:
+7-1-16-18: do you identify the (maximal) independent set for each connected component or do you consider also subsets of the overall independent set to get the "set of all indepedent sets in G"?
+Please clearify!
+%
+7-1-29-39: You are using "S" instead of the defined independent set "I". Please correct!
+%
+7-1-34: The use of "B" comes quite surprisingly since its last occurrence was several pages back. Please add a helping statement like "one has a valid base pair (...) $\in$ B".
+}
 

### 2018-08-28 supplement.tex
Last modifs
Work on comments of reviewer R1---leave some open todo notes

@@
 
-Crucially, one can show that, as long as the network the arguments $\dep(f)$ of a function $f$ are materialized by a clique within the network $\network$, then there exists at least one node in the tree decomposition that contains $\dep(f)$, thus allowing the evaluation of $f$.
+Crucially, one can show that, as long as the arguments $\dep(f)$ of a function $f$ form a clique within the network $\network$, then there exists at least one node in the tree decomposition that contains $\dep(f)$, thus allowing the evaluation of $f$.
 \begin{lemma}\label{lem:cliques}
-Let $G=(V,E)$ be an undirected graph, and  $T,\chi$ be a tree decomposition for $G$. For each clique $C\subset V$, there exists a node $n\in T$ such that $C\subset\chi(v)$.
+\todo[inline]{R1: guess it should read "node v" instead of "n"!? please update the proof accordingly!}
+Let $G=(V,E)$ be an undirected graph, and  $(T,\chi)$ be a tree decomposition for $G$. For each clique $C\subset V$, there exists a node $n\in T$ such that $C\subset\chi(v)$.
 \end{lemma}
@@
 
-RNA design introduces constraints due to the requirement of canonical
-(aka complementary) base pairing, which can be exploited in a
-particularly simple and effective way to reduce the complexity.  
-As previously noted by \citet{Flamm2001}, the base pair complementarity induces a
+The given structures in the RNA design problem impose constraints on
+the sequences for ensuring canonical (aka complementary)
+base pairing. These constraints can be exploited in a particularly
+simple and effective way to reduce the complexity.  As previously
+noted by \citet{Flamm2001}, the base pair complementarity induces a
 bi-partition of each connected component within the base pair
@@
 nodes in the partition are restricted to values in $\{\Ab,\Gb\}$ and
-$\{\Cb,\Ub\}$ respectively. We call a partial sequence
-\Def{cc-valid}, iff its determined positions are consistent with such a separation
-for all determined positions of the same connected component.
+$\{\Cb,\Ub\}$ respectively. We call a partial sequence \Def{cc-valid},
+iff its determined positions are consistent with such a separation for
+all determined positions of the same connected component.
 
@@
 One can now modify Alg.~\ref{alg:pf}, on a tree decomposition
-$(T,\chi)$, such that the message values
-$\evalfor{\Message{u}{v}}{\val}$ are only computed for cc-valid
+$(T,\chi)$, such that the values of messages
+$\Message{u}{v}}$ are computed only for cc-valid
 partial sequences $\val\in\partseqs(\separator{u}{v})$. Moreover, the
@@
 partial sequence invalid. Moreover, applications of the sampling
-framework, can introduce additional constraints that further reduce
+framework can introduce additional constraints that further reduce
 the number of valid partial subsequences. However, exploiting all such
@@
 
-\section{Contribution of MultiDefect terms to the improvement}
+\section{Contribution of terms to the MultiDefect score}
 \label{sec:improvements}
@@
 \includegraphics[width=.48\textwidth]{Figs/statistics-Term2}
-\caption{Break down of average MultiDefect values into their two components.}
+\caption{Break-down of average MultiDefect values into their two components.}
 \end{figure}

### 2018-08-29 RNARedPrint-Bioinfo2018.tex
Rewrite proof of bijection---adress corresponding R1 comments; update Fig. 2 caption

@@
     targeting stacking model energies.  We show the fits for both
-    target structures S1 and S2 of subfigure B.  \textbf{(B)} D\&P energy
-    distributions for two target structures S1 and S2. We show the
+    target structures R1 and R2 of subfigure B.  \textbf{(B)} D\&P energy
+    distributions for two target structures R1 and R2. We show the
     results of targeting the respective free energies $(-30,-20)$,
@@
   \label{fig:energydist-pk}
-  \todo[inline]{Rename S1,S2 to R1 R2. Discuss R2 values. Add legend to subfig A or explain colors}
 \end{figure*}
@@
     \Obj(S) &= \frac{1}{k} \sum_{\ell=1}^{k} (E(S, R_\ell) - \EnsE(S))\notag\\
-   & +\frac{1}{2\binom{k}{2}} \sum\limits_{1\leq\ell<j\leq k}|E(S,R_\ell) - E(S,R_j)|.
+   & +\frac{1}{2\binom{k}{2}} \sum\limits_{1\leq\ell<j\leq k}|E(S,R_\ell) - E(S,R_j)|,
 \end{align}
-
+%
 where $\EnsE(S)$, the \Def{ensemble free energy} of $S$, is computed
@@
 
-\todo[inline]{R1: I am sure your "bipartite" definition is wrong (or your proof is broken), ie.
-is the bipartition not about V1 and V2, ie there is no edge WITHIN each vertex set but only BETWEEN the sets?
-Right now, your definition says that the intersection of all possible inter-set-edges (V1xV2) with the present edges E is empty..
-But the proof says in line 7-1-35 that any edge "in E involves two nodes $v1 \in V1$ and $v2 \in V2$".
-So you might correct the bipartite definition to something like "$(V1xV1)\cap E = \emptyset = (V2xV2)\cap E$"  }
-
-To establish that claim, we consider instances $G=(V_1\cup V_2, E)$ that are connected and bipartite ($(V_1\times V_2) \cap E = \varnothing$), noting that hardness over restricted instances implies hardness of the general problem.
+To establish that claim, we consider instances $G=(V_1\cup V_2, E)$ that are connected and bipartite ($E \cap (V_1\times V_2) = E$), noting that hardness over restricted instances implies hardness of the general problem.
 Moreover note that, as observed in Subsec.~\ref{sec:complexity}, assigning a nucleotide to a position $u\in V$ constrains the parity ($\{\Ab,\Gb\}$ or $\{\Cb,\Ub\}$) of all positions in the connected component of $u$.
@@
 
-\todo[inline]{
-R1:
-7-1-16-18: do you identify the (maximal) independent set for each connected component or do you consider also subsets of the overall independent set to get the "set of all indepedent sets in G"?
-Please clearify!
-%
-7-1-29-39: You are using "S" instead of the defined independent set "I". Please correct!
-%
-7-1-34: The use of "B" comes quite surprisingly since its last occurrence was several pages back. Please add a helping statement like "one has a valid base pair (...) $\in$ B".
-}
-
-We remind that an \Def{independent set} of $G=(V,E)$ is a subset $V'\subseteq V$ of nodes that are not connected by any edge in $E$. Let $\IS{G}$ denote the set of all independent sets in $G$.
+Let $\IS{G}$ denote the set of all independent sets in the connected graph $G$; recall that an \Def{independent set} of $G=(V,E)$ is a subset $V'\subseteq V$ of nodes that are not connected by any edge in $E$.
 
 \begin{proposition}
- $\Design{G}$ is in bijection with $\IS{G}$.
+  $\Design{G}$ and $\IS{G}$ have equal size.
 \end{proposition}
+%
 \begin{proof}
-Consider the function $\Psi: \Design{G} \to \IS{G}$ defined by $ \Psi(f) := \left\{v\in V\mid f(v)\in\{\Ab,\Cb\}\right\}.$
-
-Let us establish the injectivity of  $\Psi$, i.e. that $\Psi(f)\neq\Psi(f')$ for all $f\neq f'$.
+Consider $\Psi: \Design{G} \to \IS{G}$ defined by $ \Psi(f) := \left\{v\in V\mid f(v)\in\{\Ab,\Cb\}\right\}.$
+We show that $\Psi$ is bijective:%
+\begin{itemize}
+\item $\Psi$ is injective, i.e. $\Psi(f)\neq\Psi(f')$ for all $f\neq f'$.
 If $f\neq f'$, then there exists a node $v\in V$ such that $f(v)\neq f'(v)$.
-Assume that $v\in V_1$, and remind that the only nucleotides allowed in $V_1$ are $\Ab$ and $\Gb$. Since $f(v)\neq f'(v)$, then we have $\{f(v),f'(v)\}=\{\Ab,\Gb\}$
-and we conclude that $\Psi(f)$ differs from $\Psi(f')$ at least with respect to its inclusion/exclusion of $v$.
-
-We turn to the surjectivity of $\Psi$, i.e.
-the existence of a preimage for each element $I\in \IS{G}$.
-Let us consider the function $f$ defined as $f(v) = \Ab$ (resp.
-$\Gb$, $\Cb$ or $\Ub$) if $v\in V_1$ and $v\in S$ (resp.
-$v\in V_1$ and $v\notin S$, $v\in V_2$ and $v\in S$, or $v\in V_2$ and $v\notin S$).
+We discuss only the case $v\in V_1$, where we restricted the nucleotides to $\Ab$ and $\Gb$. Then, $\{f(v),f'(v)\}$ must equal $\{\Ab,\Gb\}$, such that either $v\in\Psi(f)$ or $v\in\Psi(f')$.
+%
+\item $\Psi$ is surjective, i.e. there is a preimage for each element $I\in \IS{G}$. Define $f\in\Design{G}$ as\begin{displaymath}
+f(v) = \left\{\begin{array}{ll@{\qquad}ll}
+\Ab{} & \text{if $v\in V_1$ and $v\in I$} &
+\Cb{} & \text{if $v\in V_2$ and $v\in I$}\\
+\Gb{} & \text{if $v\in V_1$ and $v\not\in I$} &
+\Ub{} & \text{if $v\in V_2$ and $v\not\in I$}
+\end{array}\right.
+\end{displaymath}
+One easily verifies that $\Psi(f) = I$.
 %
@@
 %\end{align*}
-One easily verifies that $\Psi(f) = S$.
-
-We are then left to determine if $f$ is a valid design for $G$, i.e.
-if for each $(v, v') \in E$ one has  $\{f(v),f(v')\}\in \B.$ Since $G$ is bipartite, any edge in $E$ involves two nodes $v_1\in V_1$ and $v_2\in V_2$.
-Remark that, among all the possible assignments $f(v_1)$ and $f(v_2)$, the only invalid combination of nucleotides is $(f(v_1),f(v_2)) = (\Ab,\Cb)$.
-However, such nucleotides are assigned to positions that are in the independent set $S$, and therefore cannot be adjacent.
-We conclude that $\Psi$ is surjective, and thus bijective.
+%
+It remains to show that $f$ is a valid design for $G$, i.e. if for each $(v, v') \in E$, $\{f(v),f(v')\}\in \B.$; please recall that we defined $\B$ as the set of all valid nucleotide pairs.
+Assume there is an edge $E=(v_1,v_2)$, violating $\{f(v_1),f(v_2)\} \in \B$. Since $G$ is bipartite,
+$v_1\in V_1$ and $v_2\in V_2$, such that $f(v_1)\in\{\Ab,\Gb\}$ and $f(v_2)\in\{\Cb,\Ub\}$.
+This implies that among all possible $\{f(v_1), f(v_2)\}$ only $\{\Ab,\Cb\}$ is not in $\B$, which in turn requires $v_1\in I$ and $v_2\in I$.
+Therefore, since $I$ is an independent set, $E$ cannot exist.
+\end{itemize}
 \end{proof}

### 2018-08-29 RNARedPrint-Bioinfo2018.tex
Fix detail in proof
Rewrite section 2

@@
 
-\newcommand{\phitotal}{\phi_{\operatorname{m}}}
-
-\newcommand{\Ebp}[2]{E^{\textrm{bp}}_{#1}(#2)}
-\newcommand{\Ehp}[1]{E^{\textrm{hp}}(#1)}
-\newcommand{\Eint}[1]{E^{\textrm{int}}(#1)}
+%\newcommand{\Ehp}[1]{E^{\textrm{hp}}(#1)}
+%\newcommand{\Eint}[1]{E^{\textrm{int}}(#1)}
+
+\newcommand{\EbpSym}{E^{\textrm{bp}}}
+\newcommand{\Ebp}[2]{\EbpSym_{#1}(#2)}
 
@@
 
-%In this work, we present a generic approach to multi-target sequence sampling that enables Boltzmann-weighted sampling (subsuming uniform sampling) for a wide class of objective functions that can even make use of the commonly used ``full-featured'' RNA energy models (like the Turner model~\citep{turner}).
-Due to its high expressivity the same framework allows combining objective functions with various hard and soft constraints, which can accommodate various desirable multi-target design scenarios.
-
-%To provide a simple concrete example of multi-target design, consider designing a single sequence where all target structures are as stable as possible.
-%That is, one asks for the sequence $S$ that minimizes the energy differences of the target structures to the minimum energy for this sequence among all possible structures; one aims to minimize
-%   \begin{equation}
-%\label{eq:design-objective}
-%\sum_{\ell\in[1,k]} E(S;R_\ell) - E_0(S),
-%\end{equation}
-%over all sequences $S$, where $E_0(S):=\min_R E(S;R)$.
-%
-%The proposed Boltzmann-weighted sampling approach can support this design procedure by sampling start structures from the Boltzmann distribution over
-%\begin{equation}
-%\label{eq:sampling-objective} \sum_{\ell\in[1,k]} w_\ell E(S;R_\ell)
-%\end{equation}
-%or extensions thereof, e.g.~for controlling base frequencies.
-Suitable weights $w_\ell$ to accommodate the design objective can be approximated automatically after the machinery for Boltzmann sampling over such functions is set up.
-%
-%\TODO{Rewrite (from here to "Contributions"): sounds too negative}
-%Notably, by Eq.~(\ref{eq:sampling-objective}) we do not suggest to sample over the energy difference of Eq.~$(\ref{eq:design-objective})$.
-The term $\min_R E(S;R)$, which---for each sequence---minimizes over the entire structure space, precludes efficient sampling even by the proposed approach.
-%
-%Already sampling over Eq.~(\ref{eq:sampling-objective}) strictly extends the only previously existing approach to controlled sampling from single to multi-target sampling (in complete analogy to IncaRNAtion only after GC content is controlled, which we present later) and subsumes uniform sampling for multi-target design.
-%
-%Naturally, also more complex objective functions are desirable for RNA design and have been discussed in the literature; e.g.
-the ensemble defect~\citep{Zadeh:2010}.
-Typically, more complex objective functions make design computationally much more challenging.
-Similarily, also the proposed sampling framework supports only a limited class of objective functions directly and efficiently---namely ``only'' all additive energy functions.
-Nevertheless, one can safely assume that the design under various more realistic objective functions profits from controlled sampling as long as the sampling is directed towards the final objective function.
-
-%\TODO{Recycled material: To be merged with intro later}
-%{\it
-%This graph coloring perspective was the basis for the uniform sampling method of \Software{RNADesign} / \Software{RNAblueprint}, which decomposes the graph hierarchically, counts the solutions in subgraphs, and finally samples solutions.
-Essentially, an \Def{ear decomposition} is used as a decision tree, where decisions are made according to the subsolution counts in order to guarantee uniform sampling.
-%
-%However, highly desirable objectives of RNA design, such as a control over the energy of individual structures, require one to depart from the uniform generation, and therefore generalize this setting.
-First, instead of subsolution counts, one needs partition functions for guiding the sampling procedure.
-Second, in the case of realistic energy models %like Eq.~(\ref{eq:sampling-objective}) are based on realistic RNA energy models
-%(e.g.~the Turner model), the atomic energy contributions depend on bases at more than two sequence positions---i.e. they cannot be captured by a strategy that only accounts for base-pairs.
-%%
-%
-%
-%These considerations motivate our sampling strategy, where we:
-%\begin{itemize}
-%\item Model RNA sequences as assignments over a constraint network.
-The network consists of one position per sequence position; constraints and energy contributions are modeled as functions of arbitrary arity over the positions, such that the total energy is the sum of all function values;
-%\item Translate the dependency hyper-graph into a binary dependency graph by inserting cliques (Appendix), such that a tree decomposition with small treewidth can be computed by a generic tree decomposition algorithm;
-%\item Compute the partition functions by specializing the generic ``Cluster Tree Elimination'' (CTE) dynamic programming algorithm, based on the cluster tree decomposition induced by the constraint network;
-%
-%\item Sample sequences are from the Boltzmann-distribution guided by the decomposition tree and the sub-partition functions.
-%\end{itemize}
-%}
-
 \section{Definitions and problem statement}
@@
 An \Def{RNA sequence $S$} is a word over the \Def{nucleotides
-  alphabet} $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}$; let $\S_n$ denote the set of
-sequences of length $n$. An \Def{RNA (secondary) structure $R$ of
+  alphabet} $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}.$
+An \Def{RNA (secondary) structure $R$ of
   length $n$} is a set of \Def{base pairs} $(i,j)$, where
@@
 
-We consider a fixed set of target RNA structures
-$\R:=\{R_1, \dots, R_k\}$ for sequences of length $n$. $\R$ induces a
+RNA energy models like the Turner energy model~\citep{Turner2009} define an energy function that maps a pair of sequence $S$ and structure $R$ to a real value, the \Def{energy}. 
+Typically, such energy functions are additively composed from energy contributions of structure elements (like single base pairs or loops).
+It is instructive, and immediately useful in this context, to first consider the simple \Def{base pair energy model}.
+This model defines energy contributions $\EbpSym(x,y)$ for base pairs $(x,y)\in\Sigma^2$ that specify the energy function as
+$$
+\EbpSym(S,R) = \sum_{(i,j)\in R} \EbpSym(S_i,S_j)
+$$
+for each sequence $S\in\Sigma^n$ and structure $R$ for length $n$.
+
+For multi-target RNA design, we fix the set of target RNA structures
+$\R:=\{R_1, \dots, R_k\}$ for sequences of length $n$.
+In this work, we are going to generate random sequences that satisfy a number of simple constraints like base pair complementarity for all base pairs in the target structures.
+Moreover, the generated samples shall (exactly or approximately) satisfy complex properties that we express in terms of features.
+An example of such a feature is the \GCb\%, where the sequence property would be to have a specific \GCb\%.
+The further properties we are going to enforce are specific 'target' energies for the single structures $R_1, \dots, R_k$; for this purpose, each such energy is considered as a feature.
+
+Formally, we define a feature $F$ as function on sequences, which is expressed by summing over an associated set of \emph{feature functions} on sequences, such that
+$$
+\F(S) = \sum_{\text{feature function $f$ of $F$}} f(S).
+$$
+We emphasize that our approach profits from simple feature functions that depend only on few sequence positions.
+For example, the $\GCb\%$ feature is expressed by feature functions $f^{GC}_i$ where $f^{GC}_i(S)$ equals 1 if $S_i='G'$ or $S_i='C'$ and 0 otherwise.
+Similarly the base pair energy for a target structure $R$ is expressed as feature with feature functions $f^R_{ij}(S)=\EbpSym(S_i,S_j),$ such that $F(S)=\EbpSym(S,R)$.
+In both cases, the feature functions are simple in the way that they depend on the nucleotides at a small (constantly bounded) number of sequence positions.
+
+\paragraph{Central problem.}
+Let us now consider the $k$ energy features $F_1,\dots F_k$ together with the negative $\GCb\%$ as feature $F_0$ to target all of them simultaneously.
+We base the targeting of the $k+1$ features $F_0,\dots,F_k$ on generating samples from a Boltzmann distribution parametrized by weights $\pi_0,\dots,\pi_k$, one for each feature.
+Sampling from this distribution requires to compute corresponding partition functions, such that we can draw sequences proportional to their Boltzmann weight:
+\begin{equation}
+\label{eq:sample-distribution}
+\Pr[S;  \pi_0,\dots,\pi_k] \propto \prod_{0\leq \ell\leq k} \pi_\ell^{-F_\ell(S)}.
+\end{equation}
+
+At the core of this work, is the efficient (fixed-parameter tractable; FPT) computation of feature-dependent partition functions over sequences.
+Given are the features $F_0,\dots,F_k$, which encode the energies functions for the fixed structures and the function for the \GCb\% all in terms of simple feature functions.
+For weights $\pi_0,\dots,\pi_k$  compute the partition function
+  \begin{equation}
+    \label{eq:mainproblem}
+    \partfun{\pi_0,\dots,\pi_k} = \sum_{S\in\Sigma^n} \prod_{0\leq \ell\leq k} \pi_\ell^{-F_\ell(S)}.
+  \end{equation}
+
+To stress the analogy to, in the RNA context more familiar, Boltzmann distributions over the structure ensemble of a sequence $S$, we emphasize the equivalence
+$$
+\prod_{0\leq \ell \leq k} \pi_\ell^{-F_\ell(S)} = exp(-\beta E(S)),
+$$
+where $\beta$ denotes inverse temperature and $E(S)$ is defined as linear combination of energies $E(S)=\sum_{1\leq \ell \leq k} \ln(\pi_\ell)/\beta E(S,R_\ell)$,
+where the energy $E(S,R_\ell)$ is treated as feature $F_\ell(S) = \EbpSym(S,R_\ell);$ as elaborated later, one could even Turner energies or pseudoknot energies can be expressed as features based on their additive nature.
+
+\paragraph{Foundations on the complexity of the central problem}
+
+The complexity of the central problem depends fundamentally on the dependency structure due to the entire set of feature functions.
+
+For the base pair energy model, the dependencies are immediately derived from the set of target structures 
+$\R$. It induces the
 \Def{base pair dependency graph} $G_{\R}$ with nodes $\{1,\dots,n\}$
@@
 
-One can interpret the valid sequences for $\R$ as colorings of
-$G_{\R}$ in a slightly modified graph coloring variant, where the
-colors (from $\Sigma$) assigned to adjacent vertices of $G_{\R}$ must
-constitute valid base pairs.
-%
-The \Def{energy} of a sequence $S$ is defined by an \Def{energy
-  function} $E$---defined for set of structures $\R$ and mapping
-sequences to $\mathbb{R}\cup\{\infty\}$---as
-$\Energy{\R}{S} \in \mathbb{R}\cup\{\infty\}$. In our setting, the
-energy $\Energy{\R}{S}$ is typically chosen as a linear combination of
-the energies of the single RNA structures (in some RNA energy model
-that defines $E(S,R) \in \mathbb{R}\cup\{\infty\}$ for each
-sequence $S$ and structure
-$R$), as well as sequence-dependent features like \GCb\%. Formally, we
-will require that
-$\Energy{\R}{S}$ can be expressed as sum over functions values, where
-each function depends on the nucleotides at certain positions of
-$S$ (see below). Furthermore note that
-$\Energy{\R}{S}$ is finite iff
-$S$ is valid for each structure $R_1,\dots,R_k$.
-
-At the core of this work, we study the computation of partition
-functions over sequences.\smallskip\\
-\textbf{Central problem (Partition function over sequences).}
-  Given an energy function $E$ and a set $\R$ of structures of length $n$, compute the
-  partition function
-  \begin{equation}
-    \label{eq:mainproblem}
-    \partfun{E,\R} = \sum_{S\in\S_n} \exp(-\beta \Energy{\R}{S}),
-  \end{equation}
-  where $\beta$ denotes the inverse pseudo-temperature, and $\S_n$ the set of sequences of length $n$.
-
-%We solve the problem efficiently for any fixed parameter $w$, whose value reflects the intrinsic complexity of the structures and energy function $E_\R$.
-%
-  As we elaborate in subsequent sections, our approach relies on
-  breaking down the energy function $\Energy{\R}{S}$ into additive
-  components, each depending on only few sequence positions. Given
-  $\R$, we express $\Energy{\R}{S}$ as the sum of energy contributions $f(S)$
-  over a set $\F$ (of functions $f:\S_n\to\real$),
-  s.t.~$\Energy{\R}{S}=\sum_{f\in\F} f(S)$. This captures realistic RNA
-  energy models---including nearest neighbor models,
-  e.g.~\citep{Turner2009}, and even pseudoknot models,
-  e.g. \citep{Andronescu2010}, while bounding the dependencies to
-  sequence positions introduced by each single $f$. Formally, define
+In more generality, we look at the dependencies due to the set of feature functions $\F$.
+First let us define, for single feature functions $f$,
   the \Def{dependencies} $\dep(f)$ of $f$ as the minimum set of sequence
@@
 
-  Each set $\F$ of functions (on sequences of length $n$) induces a
+Then, the set $\F$ of feature functions (on sequences of length $n$) induces a
   \Def{dependency graph} on sequence positions, namely the hypergraph
   $G_\F=(\{1,\dots,n\} ,\{\dep(f)\mid f\in \F\})$.  Our algorithms
-  will critically rely on a \Def{tree decomposition} of the dependency
-  graph, which we define below.
+  critically relies on a \Def{tree decomposition} of the dependency
+  graph.
 \begin{definition}[Tree decomposition and width]
@@
 
-%\begin{algorithm}[Boltzmann weighted sampling]
-%\label{alg:sampling}
-% \ \\
-%  Define $\sample(u,\val)$:
-%  \begin{enumerate}[1)]
-%  \item Let $\Delta:=\chi(u)-\dom(\val)$ be the set of free positions for $u$.
-%  \item For all assignments $\val'\in\assignments(\Delta)$, compute
-%  $$\partfun{\val'} := \prod_{f\in \phitotal(u)} \exp(-\beta \evalfor{f}{\substitute{val'}{\val}}).$$
-%  \item Choose one of the assignments $\val'\in\assignments(\Delta)$ with probability $\partfun{\val'}/\partfun{u}$, where $\partfun{u}$ is the sum of the $\partfun{\val'}$.
-%  \item For all edges $(v',u)$ in $\edgesToR{}$:\quad update $\val':=\sample(v',\substitute{\val'}{\val})$.
-%  \item Return $\val'$.
-%  \end{enumerate}
-%\end{algorithm}
-
 \subsection{Computational complexity of the multiple target sampling algorithm}\label{sec:complexity}
@@
 \subsection{Extension to multidimensional Boltzmann sampling}\label{sec:multiBoltzmann}
-The flexibility of our framework supports an advanced sampling technique, named multidimensional Boltzmann sampling~\citep{Bodini2010} to (probabilistically) enforce additional constraints.
-This technique was previously used to control \GCb\%~\citep{Waldispuehl2011,Reinharz2013} and dinucleotide content~\citep{Zhang2013} of sampled RNA sequences; it enables explicit control of the free energies $(\TargetE_1,\ldots,\TargetE_k)$ of the single targets.
+The flexibility of our framework supports the advanced sampling technique "multidimensional Boltzmann sampling"~\citep{Bodini2010}, which allows to (probabilistically) enforce additional, complex properties of the samples.
+This technique was previously used to control \GCb\%~\citep{Waldispuehl2011,Reinharz2013} and dinucleotide content~\citep{Zhang2013} of sampled RNA sequences; here, we additionally use it to control the free energies $(\TargetE_1,\ldots,\TargetE_k)$ of the single targets.
 %We use multidimensional Boltzmann sampling to generate sequences having predetermined targeted energies $(\TargetE_1,\ldots,\TargetE_k)$ for the $k$ structures.
@@
 %
-It remains to show that $f$ is a valid design for $G$, i.e. if for each $(v, v') \in E$, $\{f(v),f(v')\}\in \B.$; please recall that we defined $\B$ as the set of all valid nucleotide pairs.
-Assume there is an edge $E=(v_1,v_2)$, violating $\{f(v_1),f(v_2)\} \in \B$. Since $G$ is bipartite,
+It remains to show that $f$ is a valid design for $G$, i.e. for each $(v, v') \in E$, $\{f(v),f(v')\}\in \B$; please recall that we defined $\B$ as the set of all valid nucleotide pairs.
+Assume there is an edge $(v_1,v_2)\in E$, violating $\{f(v_1),f(v_2)\} \in \B$. Since $G$ is bipartite,
 $v_1\in V_1$ and $v_2\in V_2$, such that $f(v_1)\in\{\Ab,\Gb\}$ and $f(v_2)\in\{\Cb,\Ub\}$.
 This implies that among all possible $\{f(v_1), f(v_2)\}$ only $\{\Ab,\Cb\}$ is not in $\B$, which in turn requires $v_1\in I$ and $v_2\in I$.
-Therefore, since $I$ is an independent set, $E$ cannot exist.
+Therefore, since $I$ is an independent set, the edge $(v_1,v_2)\in E$ cannot exist.
 \end{itemize}

### 2018-08-29 RNARedPrint-Bioinfo2018.tex
More rewrite --- mostly 3.4 still needs adaption

@@
 \newcommand{\elim}[2]{\operatorname{elim}(#1,#2)}
-\newcommand{\edgesToR}{E^r_T}
 
@@
 To stress the analogy to, in the RNA context more familiar, Boltzmann distributions over the structure ensemble of a sequence $S$, we emphasize the equivalence
-$$
+\begin{equation}
+\label{eq:feature-energy-transformation}
 \prod_{0\leq \ell \leq k} \pi_\ell^{-F_\ell(S)} = exp(-\beta E(S)),
-$$
+\end{equation}
 where $\beta$ denotes inverse temperature and $E(S)$ is defined as linear combination of energies $E(S)=\sum_{1\leq \ell \leq k} \ln(\pi_\ell)/\beta E(S,R_\ell)$,
@@
 %
+
 In the RNA design network, (partially determined) RNA sequences
@@
 constraint networks. Partially determined RNA sequences, for short
-\Def{partial sequences}, are words $\val$ over the alphabet
-$\Sigma\cup\{?\}$ equivalently representing the set ${\mathcal S}(\val)$
-of RNA sequences, where for positions $1\leq i\leq n$,
-$\val_i\in\Sigma$ implies $S_i=\val_i$ for all $S\in{\mathcal
-  S}(\val)$. The positions $1\leq i\leq n$, where $\val_i\in\Sigma$,
+\Def{partial sequences}, are words $\val$ in
+$(\Sigma\cup\{?\})^n$.
+
+The positions $1\leq i\leq n$, where $\val_i\in\Sigma$,
 are called \Def{determined} by $\val$ and form its \Def{domain}.
-%
-Since the functions $f\in\F$ of Section~\ref{sec:problem-statement}
-depend on only the subset $\dep(f)$ of sequence positions, one can
+Note that
+$\val$ equivalently represents the set 
+of all RNA sequences in $\Sigma^n$ that are identical to $\val$ at the determined positions $i$.
+%
+Since each feature function $f\in\F$ (cf.~Section~\ref{sec:problem-statement})
+depends on only the subset $\dep(f)$ of sequence positions, one can
 evaluate them for partial sequences $\val$ that determine (at least)
@@
 \item $\X$ is the set of sequence positions $1,\dots,n$
-\item $\F$ is a set of \Def{functions} $f:\S_n\to\real$
+\item $\F$ is a set of \Def{functions} $f:\Sigma^n\to\real$
 \end{itemize}
@@
 %
-The \Def{energy $\energy{\network}(S)$ of a sequence} $S$ in a network
-$\network$ is defined as sum of the values of all functions in
+The network has the natural interpretation of assigning energy to sequences. 
+The \Def{network energy of a sequence} $S$ for the network
+$\network$ is the sum of the values of all functions in
 $f\in\F$ evaluated for $S$, i.e.
-$\energy{\network}(S) := \sum_{f\in\F} \evalfor{f}{S}.$
-
-The network energy $\energy{\network}(S)$ corresponds to the energy in
-Eq.~$(\ref{eq:mainproblem}),$ where this energy is modeled as sum of
-the functions in $\F$. Consequently, $\partfun{\R}$ of
-Eq.~$(\ref{eq:mainproblem})$ is modeled as network partition function
-$\partfun{\network} := \sum_{S}\exp(-\beta\energy{\network}(S)) = \sum_{S}\prod_{f\in\F} \exp( -\beta\cdot
-\evalfor{f}{S} ).$
-
-% For a given RNA design network, one is typically interested in the \Def{minimum free energy sequence} $\energy{\network}:=\min_{S}\energy{\network}(S)$, the \Def{enumeration of valid sequences} $\numberof{\network}$ and the \Def{partition function} of the network $\network$ as \begin{equation}
-%  \partfun{\network;\beta} := \sum_{S}\prod_{f\in F} \exp( -\beta\cdot \evalfor{f}{\val_S} ).
-%  \end{equation}
+$\sum_{f\in\F} \evalfor{f}{S}.$
+
+A further trick lets us handle the feature functions uniformly by `moving' the parameter $\pi_\ell$ of feature $F_\ell$ into each feature function $f$ of this feature, defining a new feature function $f'(S) = 1/\ln(\pi)f(S).$ Then, $exp(-f'(S))=\pi_\ell^{f(S)}.$ While all feature functions belong to a specific feature and therefore need to be weighted accordingly, by assuming that all feature functions are transformed in this way, they can be handled uniformly.
+
+Then, the network energy of $S$ in $\network$ corresponds to the energy in
+Eq.~$(\ref{eq:mainproblem}),$ considering the transformation of Eq.~(\ref{eq:feature-energy-transformation}). 
+Consequently, the network encodes the partition function $\partfun{\pi_0,\dots,\pi_k}$ of Eq.~(\ref{eq:mainproblem}) and can be used to (now efficiently, as explained in the next section) compute it.
 
@@
 over an RNA design network can be computed by dynamic programming based
-on a tree decomposition of the network's dependency graph
-(i.e. cluster tree elimination).
+on a tree decomposition of the network's dependency graph,
+i.e.~essentially cluster tree eliminationi (CTE) \citep{Dechter2013}.
 %This yields fixed parameter tractable (FPT) algorithms, based on the treewidth.
@@
 function.
-%Observe that counting is a special case of the partition function and can be obtained by setting $\beta =0$.%; moreover, efficient energy minimization can be obtained by a simple $(+,\times)\to (\min,+)$ algebraic substitution and dropping the transformation of energy contributions to Boltzmann weights.
 
@@
 connected and contains a dedicated node $r$, with
-$\chi(r)=\varnothing$ and $\phi(r)=\varnothing$, added as a virtual
-root $r$ connected to a node in each connected component of $T$.  Now,
-we consider the set of directed edges $\edgesToR{}$ of $T$ oriented to
-$r$; define $T_r(u)$ as the induced subtree of
-$u$. Algorithm~\ref{alg:pf} computes the partition function by passing
+$\chi(r)=\varnothing$ and $\phi(r)=\varnothing$. If such a root does not exist, it can be easily added to the tree decomposition and connected to one node in each connected component of $T$.
+
+Moreover, we assume that all edges in the tree decompostion are oriented towards this root.
+%and 
+%define $T_r(u)$ as the induced subtree of
+%$u$ (following this orientation). 
+Algorithm~\ref{alg:pf} computes the partition function by passing
 messages along these directed edges $u\to v$ (i.e. always from some
@@
 Alg.~\ref{alg:pf}, multiplying the 0-ary messages sent to the root $r$
-yields the total partition function:
+yields the total partition function of the network (i.e.~due to proper encoding the partition function of our design problem)
 \begin{math}
-  \partfun{\network} = \prod_{(u\to{}r)\in T} \evalfor{\Message{u}{r}}{\varnothing}.
+  \prod_{(u\to{}r)\in T} \evalfor{\Message{u}{r}}{\varnothing}.
 \end{math}
@@
       \For{$\val'\in\partseqs(\difference{u}{v})$}{
-        $p \gets$ product( $exp(-\beta \evalfor{f}{\substitute{\val}{\val'}})$ for $f\in \phi(u)$ )\\
+        $p \gets$ product( $exp(-\evalfor{f}{\substitute{\val}{\val'}})$ for $f\in \phi(u)$ )\\
         ${}\quad\qquad \cdot\ $product( $\evalfor{\Message{u'}{u}}{\substitute{\val}{\val'}}$ for $(u'\to{}u)\in T$ )\;
@@
   \caption{FPT computation of the partition function using dynamic
-    programming, i.e.~cluster tree elimination
-    \citep[CTE]{Dechter2013}. The postorder traversal guarantees that
+    programming, i.e.~cluster tree elimination (CTE). The postorder traversal guarantees that
     when processing edge $u\to{}v$, all messages
@@
 For an expanded cluster tree, after the messages $\Message{u}{v}$ for
-the edges in $\edgesToR{}$ are generated by Algorithm \ref{alg:pf},
+the edges in the tree decomposition are generated by Algorithm \ref{alg:pf},
 one can repeatedly call Algorithm~\ref{alg:sampling}, each time
@@
 
-%      $p \gets$ product( $exp(-\beta \evalfor{f}{\substitute{\val}{\val'}})$ for $f\in \phi(u)$ )\\
+%      $p \gets$ product( $exp(-\evalfor{f}{\substitute{\val}{\val'}})$ for $f\in \phi(u)$ )\\
 %      ${}\quad\qquad \cdot\ $product( $\evalfor{\Message{u'}{u}}{\substitute{\val}{\val'}}$ for $(u'\to{}u)\in T$ )\;
@@
 
-     $p \gets$ product( $exp(-\beta \evalfor{f}{\substitute{\val}{\val'}})$ for $f\in \phi(u)$ )\\
+     $p \gets$ product( $exp(-\evalfor{f}{\substitute{\val}{\val'}})$ for $f\in \phi(u)$ )\\
      ${}\quad\qquad \cdot\ $product( $\evalfor{\Message{u'}{u}}{\substitute{\val}{\val'}}$ for $(u'\to{}u)\in T$ )\;
@@
   models.}\label{sec:energy_models}
-\todo[inline]{Better motivate why we want different energy functions etc}
-%
-The functions in $\F$ allow expressing complex features of the
+
+Going beyond our encoding of the base pair model energy features and the \GCb\% feature, the feature functions provide more generality and allow expressing complex features of the
 sequences alone, e.g.\ rewarding or penalizing specific sequence
@@
 naturally expressed by assigning infinite penalties to invalid
-(partial) sequences. Finally, but less obviously, the framework
-captures various RNA energy models with bounded dependencies, which we
-describe briefly.
-
-In simple \Def{base pair-based energy models}, energy is defined as
-the sum of base pair (pseudo-)energies. If base pair energies
-$\Ebp{\ell}{i,j,x,y}$ (where $i$ and $j$ are sequence positions, $x$ and
-$y$ are bases in $\Sigma$) are given for each target structure $R_\ell$,
-s.t.  $ E(S;R_\ell) := \sum_{(i,j)\in R_\ell} \Ebp{\ell}{i,j,S_i,S_j}$,
-we encode the network energy by the set of functions $f$ for each base
-pair $(i,j)\in R_\ell$ of each input structure $R_\ell$ that evaluate
-to $\ln(\pi_\ell) \Ebp{\ell}{i,j,\val_i,\val_j}$ under partial sequence $\val$;
-here, $\pi_\ell>0$ is a weight that controls the
-influence of the $\ell$-th structure on the sampling (as elaborated later).
+(partial) sequences (even if in implementations, explicit handling of constraints is advantageous). In particular, the framework
+captures various RNA energy models, they can be handled efficiently (in the sense of FPT), as long as the dependencies of their additive energy contributions are bounded.
 
@@
 
-More complex \Def{loop-based}
+Complex \Def{loop-based}
  energy models---e.g.~the Turner model,
 which among others includes energy terms for special loops and dangling
-ends---can also be encoded as straightforward extensions. An interesting
-stripped-down variant of the nearest neighbor model is the
+ends---can be encoded as straightforward extension of our outline of the base pair energy model.
+
+In practice, the energy model has a strong influence on the computational complexity of our approach. Therefore, it is interesting to consider, in addition to the base pair model, other
+stripped-down variants of the nearest neighbor model, which could offer a compromise between the low-complexity base pair energy model and the high-accuracy Turner model. 
+Such a model, which is particularly promising for this purpose, is the
 \Def{stacking energy model}. This model assigns non-zero energy
@@
 \subsection{Extension to multidimensional Boltzmann sampling}\label{sec:multiBoltzmann}
+\todo[inline]{This is now better prepared due to the formulation of the main problem in terms of features with weights.
+Consequently, the text needs adaptions!
+Note that the energies E_1,\dots,E_k$ are still features $F_1,\dots,F_k$; there is also a feature $F_0$ for the GC content.
+}
 The flexibility of our framework supports the advanced sampling technique "multidimensional Boltzmann sampling"~\citep{Bodini2010}, which allows to (probabilistically) enforce additional, complex properties of the samples.
@@
 where $\pmb{\pi}:=(\pi_1\cdots\pi_k)$ is a vector of positive real-valued \Def{weights}, and $\partfun{\pmb{\pi}}$ is the weighted partition function.
-Such a distribution can be induced by a simple modification of the functions described in Sec.~\ref{sec:energy_models}, where any energy function $E(\val)$ for a structure $\ell$ is replaced by $E'(\val):= \ln(\pi_\ell)\, E(\val)/\beta$\todo{is $/\beta$ necessary?}.
+%Such a distribution can be induced by a simple modification of the functions described in Sec.~\ref{sec:energy_models}, where any energy function $E(\val)$ for a structure $\ell$ is replaced by $E'(\val):= \ln(\pi_\ell)\, E(\val)/\beta.$
 The probability of a sequence $S$ is thus proportional to

### 2018-08-29 RNARedPrint-Bioinfo2018.tex
WIP: iterate over text, more simplifications

@@
 
-\newcommand{\evalfor}[2]{#1\llbracket{}#2\rrbracket{}}
 \newcommand{\substitute}[2]{#1\!\oplus\!#2}
+%\newcommand{\evalfor}[2]{#1\llbracket{}#2\rrbracket{}}
+\newcommand{\evalfor}[2]{#1(#2)}
 
@@
 %
-Recently, the \Software{RNAdesign}
-approach~\citep{HoenerzuSiederdissen2013}, coupled with powerful local
+Recently, the approach \Software{RNAdesign}~\citep{HoenerzuSiederdissen2013}, coupled with local
 search within \RNAblueprint{}~\citep{Hammer2017}, solved the
-problem of sampling seeds from the uniform distribution for multiple
-targets. These methods adopt a graph coloring perspective, initially
-decomposing the graph hierarchically using various decomposition
-algorithms, and \Def{precomputing} the number of valid sequences
+problem of sampling seeds from the \emph{uniform} distribution for multiple
+target structures. \Software{RNAdesign} adopts a graph coloring perspective,
+coloring the sequence positions with nucleotides, such that the ends of each base pair receive compatible colors.
+Initially the method decomposes the graph hierarchically using ear decomposition, and then \Def{precomputes} the number of valid sequences
 within each subgraph. The decomposition is then reinterpreted as a
@@
 \Software{RNAdesign} grows like $\Theta(4^{\gamma})$, where the
-parameter $\gamma$ is bounded by the length of the designed RNA;
-typically, the decomposition strategy achieves much lower $\gamma$.
-
-\parHead{Motivation.} The exponential time and space requirements of the \Software{RNAdesign} method already raise the question of the \Def{complexity of (uniform) sampling for multi-target design}.
-Since stochastic backtrack can be performed in linear time per sample, the method is dominated by the precomputation step, which requires counting valid designs.
-Thus, we focus on the question: \emph{Is there a polynomial-time algorithm to count valid multi-target designs?} In Section~\ref{sec:counting}, we answer in the negative, showing that there exists no such algorithm unless ${\sf P}={\sf NP}$.
+parameter $\gamma$, while bounded only by the length of the designed RNA, is 
+typically much lower, due to the decomposition strategy.
+
+\parHead{Motivation.} 
+While uniform sampling in \Software{RNAdesign}/\RNAblueprint{} turned out to be very successful,
+the work on these tools gave no definitive answer on the \emph{complexity of uniform sampling for multi-target design}.
+On the one hand, time and space requirements grow exponential in the general case,
+the decomposition strategy seems to allow practical applications.
+Better understanding of the true complexity bounds is still highly relevant for the application of such methods, in particular for applications in more expressive settings (e.g.
+Boltzmann sampling as discussed in this work or sampling satisfying multiple additional constraints and properties).
+All such extensions are still bounded by the complexity of the multi-target uniform sampling problem, where counting of valid designs is the dominating step (since stochastic backtrack can be performed in linear time per sample).
+
+Consequently, for elucidating the theoretical fundament of RNA design sampling, we first need to clarify, whether efficient approaches for counting could exist.
+In Section~\ref{sec:counting}, we show that there can not exists any efficient, polynomial time algorithm unless ${\sf P}={\sf NP}$.
 Our result relies on a surprising bijection (up to a trivial symmetry) between valid sequences and independent sets of a bipartite graph, being the object of recent breakthroughs in approximate counting complexity~\citep{Bulatov2013,Cai2016}.
-The hardness of counting (and conjectured hardness of sampling) does not preclude, however, practically applicable algorithms for counting and sampling.
-In particular, we wish to extend the flexibility of multi-structure design, leading to the following questions: \emph{How to sample, generally, from a Boltzmann distribution based on expressive energy models?
+
+As already indicated by \Software{RNAdesign}, the hardness of counting (and conjectured hardness of sampling), does still not rule out practically applicable algorithms for counting and sampling.
+Addressing the following questions in more detail and generality led to the practical advances in this work, which extend the expressivity, flexibility, and performance of multi-structure design: \emph{How to sample, in a general way, from a Boltzmann distribution based on expressive energy models?
 How to enforce additional constraints, such as \GCb\%, complex sequence constraints, or the energy of individual structures?}
@@
 To answer these questions, we introduce a generic framework (illustrated in~Fig.~\ref{fig:workflow}) enabling efficient Boltzmann-weighted sampling over RNA sequences with multiple target structures (Section~\ref{sec:FPT}).
-Guided by a \Def{tree decomposition}, we devise dynamic programming to compute partition functions and sample sequences from the Boltzmann distribution%
+As main algorithmic contribution of this work, we devise a dynamic programming algorithm, based on a tree decomposition, to compute partition functions and sample sequences from the Boltzmann distribution%
 % (Subsection~\ref{sec:PF})
-. We show that these algorithms are \Def{fixed-parameter tractable} for the \Def{treewidth}%
+. We show that these algorithms are \Def{fixed-parameter tractable (FPT)} for the \Def{treewidth},
+in other words: they have polynomial time and space complexity for a fixed treewidth%
 % Subsection~\ref{sec:complexity}
 %
-; in practice, we limit this parameter by using state-of-the-art tree decomposition algorithms.
+. Due to the generality of our method, we can moreover strongly limit this parameter in practice by using state-of-the-art tree decomposition algorithms.
 %Uniform sampling is handled as a special case of Boltzmann sampling, where each valid sequence receives energy zero and---consequently---computing partition functions specializes to counting.
@@
 . %
-We observe that sampling based on less complex RNA energy models
-(taking only the most important energy contributions into account)
-still allows targeting realistic RNA energies in the well-accepted
-Turner RNA energy model. The resulting combination of efficiency and
-high accuracy finally enables generating biologically relevant
-multi-target designs in our final application of our overall strategy
+
+We observe that
+targeting realistic RNA energies in the realistic
+Turner RNA energy model
+works well by performing sampling based on less complex RNA energy models, which cause low treewidth.
+This result is highly relevant in practice, since it allows to combine high efficiency with high accuracy and moreover leaves room for extensions of the method.
+
+Eventually this strategy allows generating biologically relevant
+multi-target designs, as we demonstrate in our application 
 to a large set of multi-target RNA design instances from a
@@
 
-RNA energy models like the Turner energy model~\citep{Turner2009} define an energy function that maps a pair of sequence $S$ and structure $R$ to a real value, the \Def{energy}. 
+RNA energy models like the Turner energy model~\citep{Turner2009} define an energy function that maps a pair of a sequence $S$ and a structure $R$ to a real value, the \Def{energy}. 
 Typically, such energy functions are additively composed from energy contributions of structure elements (like single base pairs or loops).
-It is instructive, and immediately useful in this context, to first consider the simple \Def{base pair energy model}.
+It is most instructive, and in this context already practically useful, to develop pour approach for the simple \Def{base pair energy model}.
 This model defines energy contributions $\EbpSym(x,y)$ for base pairs $(x,y)\in\Sigma^2$ that specify the energy function as
+$
+\EbpSym(S,R) = \sum_{(i,j)\in R} \EbpSym(S_i,S_j)
+$
+for each sequence $S\in\Sigma^n$ and structure $R$ for length $n$.
+
+As the structural targets of our design, we fix the set of RNA structures
+$\R:=\{R_1, \dots, R_k\}$ for sequences of length $n$.
+Recall that we aim to generate random sequences that satisfy a number of constraints like the compatibility of paired bases.
+Moreover, we want to enforce that generated samples satisfy complex properties, which we express in terms of \emph{features}.
+An example of such a feature is the \GCb\% (the corresponding property of a sequence is to have a specific \GCb\%).
+The additional properties that we are going to enforce are specific 'target' energies for the single structures $R_1, \dots, R_k$; for this purpose, each such energy is considered as a feature.
+
+Formally, we define a \Def{feature $F$} as function on sequences, which is expressed by summing over an associated set of \Def{feature functions} on sequences, such that
 $$
-\EbpSym(S,R) = \sum_{(i,j)\in R} \EbpSym(S_i,S_j)
+F(S) = \sum_{\text{feature function $f$ associcated with $F$}} f(S).
 $$
-for each sequence $S\in\Sigma^n$ and structure $R$ for length $n$.
-
-For multi-target RNA design, we fix the set of target RNA structures
-$\R:=\{R_1, \dots, R_k\}$ for sequences of length $n$.
-In this work, we are going to generate random sequences that satisfy a number of simple constraints like base pair complementarity for all base pairs in the target structures.
-Moreover, the generated samples shall (exactly or approximately) satisfy complex properties that we express in terms of features.
-An example of such a feature is the \GCb\%, where the sequence property would be to have a specific \GCb\%.
-The further properties we are going to enforce are specific 'target' energies for the single structures $R_1, \dots, R_k$; for this purpose, each such energy is considered as a feature.
-
-Formally, we define a feature $F$ as function on sequences, which is expressed by summing over an associated set of \emph{feature functions} on sequences, such that
-$$
-\F(S) = \sum_{\text{feature function $f$ of $F$}} f(S).
-$$
-We emphasize that our approach profits from simple feature functions that depend only on few sequence positions.
+We emphasize that our approach is designed to profit (in terms of computational complexity) from feature functions that each depend only on few sequence positions.
 For example, the $\GCb\%$ feature is expressed by feature functions $f^{GC}_i$ where $f^{GC}_i(S)$ equals 1 if $S_i='G'$ or $S_i='C'$ and 0 otherwise.
 Similarly the base pair energy for a target structure $R$ is expressed as feature with feature functions $f^R_{ij}(S)=\EbpSym(S_i,S_j),$ such that $F(S)=\EbpSym(S,R)$.
-In both cases, the feature functions are simple in the way that they depend on the nucleotides at a small (constantly bounded) number of sequence positions.
+In both cases, the feature functions depend on the nucleotides at only a small (constantly bounded) number of sequence positions (one and two positions, respectively).
 
 \paragraph{Central problem.}
-Let us now consider the $k$ energy features $F_1,\dots F_k$ together with the negative $\GCb\%$ as feature $F_0$ to target all of them simultaneously.
-We base the targeting of the $k+1$ features $F_0,\dots,F_k$ on generating samples from a Boltzmann distribution parametrized by weights $\pi_0,\dots,\pi_k$, one for each feature.
-Sampling from this distribution requires to compute corresponding partition functions, such that we can draw sequences proportional to their Boltzmann weight:
+Let us consider the $k$ energy features $F_1,\dots F_k$ together with the negative $\GCb\%$ as feature $F_0$ to target all of them simultaneously.
+We base our strategy for targeting the $k+1$ features $F_0,\dots,F_k$ on generating samples from a Boltzmann distribution parametrized by weights $\pi_0,\dots,\pi_k$, one for each feature.
+These weights will essentially allow us to shift the sample distributions towards specific targets.
+Sampling from this distribution requires to compute corresponding partition functions, such that we can draw sequences with probabilities proportional to their Boltzmann weight:
 \begin{equation}
@@
 
-At the core of this work, is the efficient (fixed-parameter tractable; FPT) computation of feature-dependent partition functions over sequences.
+The workhorse of our approach is the efficient (i.e.~fixed-parameter tractable) computation of feature-dependent partition functions over sequences.
 Given are the features $F_0,\dots,F_k$, which encode the energies functions for the fixed structures and the function for the \GCb\% all in terms of simple feature functions.
@@
 
-To stress the analogy to, in the RNA context more familiar, Boltzmann distributions over the structure ensemble of a sequence $S$, we emphasize the equivalence
+To stress the analogy to, in the RNA context more familiar (e.g.~\citep{McCaskill1990}), Boltzmann distributions over the structure ensemble of a sequence $S$, we emphasize the equivalence
 \begin{equation}
@@
 where $\beta$ denotes inverse temperature and $E(S)$ is defined as linear combination of energies $E(S)=\sum_{1\leq \ell \leq k} \ln(\pi_\ell)/\beta E(S,R_\ell)$,
-where the energy $E(S,R_\ell)$ is treated as feature $F_\ell(S) = \EbpSym(S,R_\ell);$ as elaborated later, one could even Turner energies or pseudoknot energies can be expressed as features based on their additive nature.
-
-\paragraph{Foundations on the complexity of the central problem}
+where the energy $E(S,R_\ell)$ is treated as feature $F_\ell(S) = \EbpSym(S,R_\ell);$ as elaborated later, even Turner energies or pseudoknot energies can be expressed as features based on their additive nature.
+
+\paragraph{Preliminaries on the complexity}
 
@@
 
-For the base pair energy model, the dependencies are immediately derived from the set of target structures 
+For design in the base pair energy model or the related counting problem, the dependencies are immediately derived from the set of target structures 
 $\R$. It induces the
 \Def{base pair dependency graph} $G_{\R}$ with nodes $\{1,\dots,n\}$
-and edges $\bigcup_{\ell\in[1,k]} R_\ell$, which describe the minimal
-dependencies present in all relevant settings due to the requirement
+and edges $\bigcup_{\ell\in[1,k]} R_\ell$. As well, this graph describes the minimal
+dependency structure of multi-target RNA design, which is present in all relevant settings due to the requirement
 of canonical base pairing.
@@
 In more generality, we look at the dependencies due to the set of feature functions $\F$.
-First let us define, for single feature functions $f$,
-  the \Def{dependencies} $\dep(f)$ of $f$ as the minimum set of sequence
+The \Def{dependencies $\dep(f)$} of a single feature function $f$ are precisely described as the minimum set of sequence
   positions $\I\subseteq\{1,\dots,n\}$, where $f(S)=f(S')$ for all
@@
   \Def{dependency graph} on sequence positions, namely the hypergraph
-  $G_\F=(\{1,\dots,n\} ,\{\dep(f)\mid f\in \F\})$.  Our algorithms
-  critically relies on a \Def{tree decomposition} of the dependency
-  graph.
+  $G_\F=(\{1,\dots,n\} ,\{\dep(f)\mid f\in \F\})$. The \Def{tree decomposition} of this dependency
+  graph is fundamental for our algorithms.
+
 \begin{definition}[Tree decomposition and width]
@@
 \end{definition}
-
-\section{An fixed-parameter tractable (FPT) algorithm for the
+%Recall the essential role of the treewidth as fixed parameter in our FPT appraoch.
+
+\section{A fixed-parameter tractable (FPT) algorithm for the
   partition function and sampling of Boltzmann-weighted designs}
 \label{sec:FPT}
+
+\todo[inline]{Plan: make the RNA design network more specific, i.e. define the set of feature functions to containt the GC and energy features (in the specific form with internalized weights).}
 
@@
 \Def{partial sequences}, are words $\val$ in
-$(\Sigma\cup\{?\})^n$.
-
-The positions $1\leq i\leq n$, where $\val_i\in\Sigma$,
-are called \Def{determined} by $\val$ and form its \Def{domain}.
-Note that
-$\val$ equivalently represents the set 
-of all RNA sequences in $\Sigma^n$ that are identical to $\val$ at the determined positions $i$.
-%
+$(\Sigma\cup\{?\})^n$, where '?' is the dummy-value of undetermined positions, which allows representing only partially determined sequences.
+Consequently, the positions $1\leq i\leq n$, where $\val_i\in\Sigma$,
+are called \Def{determined} by $\val$.
+
 Since each feature function $f\in\F$ (cf.~Section~\ref{sec:problem-statement})
@@
 evaluate them for partial sequences $\val$ that determine (at least)
-the nucleotides at all positions in $\dep(f)$. Thus for functions $f$
-and partial sequences $\val$ that determine $\dep(f)$, we write
-$\evalfor{f}{\val}$ to \Def{evaluate $f$ for $\val$}; i.e.{}
-$\evalfor{f}{\val} := f(S),$ for any sequence
-$S\in{\mathcal S}(\val)$.
+the nucleotides at all positions in $\dep(f)$. Thus for any such function $f$
+and partial sequence $\val$, we know that $f(S)$ has the same value for any sequence $S\in \Sigma^n$ that equals $\val$ at its determined positions; thus, we define $\evalfor{f}{\val}$ as this common value.
 
@@
 \item $\X$ is the set of sequence positions $1,\dots,n$
-\item $\F$ is a set of \Def{functions} $f:\Sigma^n\to\real$
+\item $\F$ is a set of \Def{feature functions} $f:\Sigma^n\to\real$
 \end{itemize}
@@
 $f\in\F$ evaluated for $S$, i.e.
-$\sum_{f\in\F} \evalfor{f}{S}.$
+$\sum_{f\in\F} f(S).$
 

### 2018-08-30 RNARedPrint-Bioinfo2018.tex
Getting much closer...

@@
 \label{eq:sample-distribution}
-\Pr[S;  \pi_0,\dots,\pi_k] \propto \prod_{0\leq \ell\leq k} \pi_\ell^{-F_\ell(S)}.
+\Pr(S|\pi_0,\dots,\pi_k) \propto \prod_{0\leq \ell\leq k} \pi_\ell^{-F_\ell(S)}.
 \end{equation}
@@
 
-\todo[inline]{Plan: make the RNA design network more specific, i.e. define the set of feature functions to containt the GC and energy features (in the specific form with internalized weights).}
-
 For our algorithmic description, we translate the concepts of
 Section~\ref{sec:problem-statement} to the formalism of constraint networks, here
-specialized as RNA design network. This allows us to base our
+specialized as RNA design network. This abstraction allows us to base our
 algorithm on the cluster tree elimination (CTE) of~\citet{Dechter2013}.
-%
-
-In the RNA design network, (partially determined) RNA sequences
-replace the more general concept of (partial) assignments in
-constraint networks. Partially determined RNA sequences, for short
-\Def{partial sequences}, are words $\val$ in
-$(\Sigma\cup\{?\})^n$, where '?' is the dummy-value of undetermined positions, which allows representing only partially determined sequences.
-Consequently, the positions $1\leq i\leq n$, where $\val_i\in\Sigma$,
-are called \Def{determined} by $\val$.
+
+The RNA design network describes how to evaluate RNA sequences and, in particular, partial RNA sequences, which is required by our algorithms that construct results for the full sequence by combining partial information.
+%
+\Def{Partial RNA sequences} are words $\val$ in
+$(\Sigma\cup\{?\})^n$, where the symbol '?' represents not yet determined nucleotides; consequently, the positions, where $\val_i\in\Sigma$,
+are called \Def{determined} by $\val$ ($1\leq i\leq n$).
+%
+We remark that (partial) sequences correspond to the more general concept of (partial) assignments in
+constraint networks.
 
@@
 \end{definition}
-
-%
-The network has the natural interpretation of assigning energy to sequences. 
-The \Def{network energy of a sequence} $S$ for the network
-$\network$ is the sum of the values of all functions in
-$f\in\F$ evaluated for $S$, i.e.
-$\sum_{f\in\F} f(S).$
-
-A further trick lets us handle the feature functions uniformly by `moving' the parameter $\pi_\ell$ of feature $F_\ell$ into each feature function $f$ of this feature, defining a new feature function $f'(S) = 1/\ln(\pi)f(S).$ Then, $exp(-f'(S))=\pi_\ell^{f(S)}.$ While all feature functions belong to a specific feature and therefore need to be weighted accordingly, by assuming that all feature functions are transformed in this way, they can be handled uniformly.
-
-Then, the network energy of $S$ in $\network$ corresponds to the energy in
-Eq.~$(\ref{eq:mainproblem}),$ considering the transformation of Eq.~(\ref{eq:feature-energy-transformation}). 
-Consequently, the network encodes the partition function $\partfun{\pi_0,\dots,\pi_k}$ of Eq.~(\ref{eq:mainproblem}) and can be used to (now efficiently, as explained in the next section) compute it.
+%
+Note that we call this object a `network', since $\X$ could be interpreted as nodes, where $\F$ defines the (hyper-)edges between those nodes.
+In this way, it encodes a dependency structure (i.e.
+a graph or `network').
+%The network has the natural interpretation of assigning an energy to sequences. 
+%The \Def{network energy of a sequence} $S$ for the network
+%$\network$ is the sum of the values of all functions in
+%$f\in\F$ evaluated for $S$, i.e.
+%$\sum_{f\in\F} f(S).$
+
+\newcommand{\weighttransform}[2]{{#1}_{||{#2}}}
+
+Note that one can encode the weight $\pi$ of a features in its feature functions, such that all feature functions can be handled uniformly by our algorithms.
+For this purpose, we define the \Def{transformed feature function}
+$$\weighttransform{f}{\pi}(S) := ln(\pi)f(S),$$
+such that $exp(-\weighttransform{f}{\pi}(S))=\pi^{-f(S)}.$
+
+Concretely, for the design in the base pair model targeting \GCb\% and structures $\R$ with weights $\pi_0,\dots,\pi_k$, we construct the set $\F$ of the network as the collection of
+\begin{itemize}
+\item the transformed feature functions $\weighttransform{f^{GC}_i}{\pi_0}$ for our feature $F_0$, i.e.~$GC\%$ ($i=1,\dots,n$)
+\item the transformed feature functions $\weighttransform{f^{R_\ell}_{ij}}{\pi_\ell}$for each structure $R_\ell\in\R$ and $(i,j)\in R_\ell.$
+\end{itemize}
+
+By this choice of $\F$, the sum $\sum_f\in\F f(S)$ (for any $S\in\Sigma^n$) corresponds to the energy in
+Eq.~$(\ref{eq:mainproblem}),$ using the equivalence of Eq.~(\ref{eq:feature-energy-transformation}). 
+Consequently, the network encodes the partition function $\partfun{\pi_0,\dots,\pi_k}$ of Eq.~(\ref{eq:mainproblem}).
+Subsequently, we show how to efficiently compute this partition function based on the network.
 
 \subsection{Partition function and Boltzmann sampling through stochastic backtrack}\label{sec:PF}
-The minimum energy, counting, and partition function
+
+The partition function
 over an RNA design network can be computed by dynamic programming based
 on a tree decomposition of the network's dependency graph,
-i.e.~essentially cluster tree eliminationi (CTE) \citep{Dechter2013}.
+i.e.~essentially cluster tree elimination (CTE) \citep{Dechter2013}. Note that analogous algorithms could be constructed to count valid sequences or minimize energy.
 %This yields fixed parameter tractable (FPT) algorithms, based on the treewidth.
-We focus on the efficient computation of the partition
-function.
 
@@
 
-We require additional definitions: A \Def{cluster tree} for the
-network $\network=(\X,\F)$ is a tuple $(T,\chi,\phi)$, where
+Our algorithms are formulated to process a \Def{cluster tree} of the
+RNA design network, which is a tuple $(T,\chi,\phi)$, where
 $(T,\chi)$ is a tree decomposition of $G_\F$, and $\phi(v)$ represents
@@
 $\dep(f)\subseteq\chi(v)$ and $\phi(v)\cap \phi(v')=\varnothing$ for
-all $v\neq v'$.  For two nodes $v$ and $u$ of the cluster tree, define
+all $v\neq v'$.
+
+Two further notions are essential for our algorithms: for two nodes $v$ and $u$ of the cluster tree, define
 their \Def{separator} as $\separator{u}{v} := \chi(u)\cap\chi(v)$;
@@
 
-For a set $\Y$ of sequence positions, write $\partseqs(\Y)$ to
-denote the set of all partial sequences that determine exactly the positions
-in $\Y$; furthermore, given partial sequences $\val$ and
+Our algorithms need to iterate over sets of partial sequences that determine specific sequence positions
+from a set $\Y$; we denote the set of all partial sequences that exactly determine the positions in $\Y$ by $\partseqs(\Y)$.
+Furthermore, given partial sequences $\val$ and
 $\val'$, we define the \Def{combined partial sequence $\substitute{\val}{\val'}$} such that
@@
 
-Finally we assume, w.l.o.g., that all position difference sets
-$\difference{u}{v}$ are singleton: for any given
+\paragraph{FPT Computation of the partition function}
+Given is the RNA design network $\network=(\X,\F)$ and its cluster
+tree $(T,\chi,\phi)$.  
+%
+We assume several properties of this cluster tree, which otherwise could be easily obtained:
+\begin{itemize}
+\item
+$T$ is
+connected and contains a dedicated node $r$, with
+$\chi(r)=\varnothing$ and $\phi(r)=\varnothing$. If such a root does not exist, it can be added to the tree decomposition and connected to one node in each connected component of $T$.
+\item
+all edges in the tree decompostion are oriented towards this root.
+\item
+all sets $\difference{u}{v}$ are singleton: for any given
 cluster tree, an equivalent (in term of treewidth) cluster tree can
@@
 clusters.
-
-Let us now consider the computation of the partition function.
-Given is the RNA design network $\network=(\X,\F)$ and its cluster
-tree decomposition $(T,\chi,\phi)$.  W.l.o.g., we assume that $T$ is
-connected and contains a dedicated node $r$, with
-$\chi(r)=\varnothing$ and $\phi(r)=\varnothing$. If such a root does not exist, it can be easily added to the tree decomposition and connected to one node in each connected component of $T$.
-
-Moreover, we assume that all edges in the tree decompostion are oriented towards this root.
-%and 
-%define $T_r(u)$ as the induced subtree of
-%$u$ (following this orientation). 
+\end{itemize}
+
 Algorithm~\ref{alg:pf} computes the partition function by passing
-messages along these directed edges $u\to v$ (i.e. always from some
+messages along the directed edges $u\to v$ (which point from 
 child $u$ to its parent $v$). Each message $m$ is a function that depends
@@
 $\max_{u,v\in V} | \separator{u}{v} |$ and denote by  $D$ the maximum size of
-$\difference{u}{v}$ over $(u,v)\in\edgesToR{}$.  In the absence
+$\difference{u}{v}$ over $(u,v)\in E$.  In the absence
 of specific optimizations, running Alg.~\ref{alg:pf} requires
@@
 \subsection{Extension to multidimensional Boltzmann sampling}\label{sec:multiBoltzmann}
-\todo[inline]{This is now better prepared due to the formulation of the main problem in terms of features with weights.
-Consequently, the text needs adaptions!
-Note that the energies E_1,\dots,E_k$ are still features $F_1,\dots,F_k$; there is also a feature $F_0$ for the GC content.
-}
 The flexibility of our framework supports the advanced sampling technique "multidimensional Boltzmann sampling"~\citep{Bodini2010}, which allows to (probabilistically) enforce additional, complex properties of the samples.
-This technique was previously used to control \GCb\%~\citep{Waldispuehl2011,Reinharz2013} and dinucleotide content~\citep{Zhang2013} of sampled RNA sequences; here, we additionally use it to control the free energies $(\TargetE_1,\ldots,\TargetE_k)$ of the single targets.
-%We use multidimensional Boltzmann sampling to generate sequences having predetermined targeted energies $(\TargetE_1,\ldots,\TargetE_k)$ for the $k$ structures.
-
-Multidimensional Boltzmann sampling requires the ability to \Def{sample from a weighted distribution} over the set of valid sequences, where the probability of a sequence $S$ with free energies $(E_1,\ldots,E_k)$ for its target structures is
-$\mathbb{P}(S\mid \pmb{\pi}) = \frac{\prod_{\ell=1}^{k} \pi_i^{-E_i}}{\partfun{\pmb{\pi}}},$
-where $\pmb{\pi}:=(\pi_1\cdots\pi_k)$ is a vector of positive real-valued \Def{weights}, and $\partfun{\pmb{\pi}}$ is the weighted partition function.
+This technique was previously used to control \GCb\%~\citep{Waldispuehl2011,Reinharz2013} and dinucleotide content~\citep{Zhang2013} of sampled RNA sequences; here, in addition to controlling \GCb\% (our feature $F_0$) we use it to target the free energies $(\TargetE_1,\ldots,\TargetE_k)$ of the single targets (features $F_1,\dots,F_k$).
+
+For the multidimensional Boltzmann sampling, we require the already established ability to \Def{sample from a weighted distribution} over the set of valid sequences, where the probability of a sequence $S$ is
+$\mathbb{P}(S\mid \pmb{\pi}) = \frac{\prod_{\ell=0}^{k} \pi_i^{-F_i(S)}}{\partfun{\pmb{\pi}}},$
+where $\pmb{\pi}:=(\pi_0\cdots\pi_k)$ is the vector of the positive real-valued \Def{weights}, and $\partfun{\pmb{\pi}}$ is the weighted partition function.
+
 %Such a distribution can be induced by a simple modification of the functions described in Sec.~\ref{sec:energy_models}, where any energy function $E(\val)$ for a structure $\ell$ is replaced by $E'(\val):= \ln(\pi_\ell)\, E(\val)/\beta.$
-The probability of a sequence $S$ is thus proportional to
-$ \prod_{\ell=1}^{k} e^{-\ln(\pi_\ell)\, E_i} = \prod_{\ell=1}^{k} \pi_i^{-E_i}. $
-
-One then needs to \Def{learn a weights vector} $\pmb{\pi}$ such that, on average, the targeted energies are achieved by a random sequences in the weighted distribution, i.e.\ such that  $\mathbb{E}(E_\ell(S)\mid \pmb{\pi})=\TargetE_\ell$,  $\forall\ell\in[1,k]$.
-The expected value of $E_\ell$ is always decreasing for increasing weights $\pi_\ell$ (see Supp.
+Due to the construction of $\F$ of our RNA design network,
+the probability of a sequence $S$ is thus proportional to
+$\prod_{\ell=0}^{k} \pi_i^{-F_i(S)} = \prod_{f\in\F} f(S).$
+
+One then needs to \Def{learn a weights vector} $\pmb{\pi}$ such that, on average, the targeted energies are achieved by a random sequences in the weighted distribution, i.e.\ such that  $\mathbb{E}(F_\ell(S)\mid \pmb{\pi})=\TargetE_\ell$,  $\forall\ell\in[1,k]$ and, analogously, the expectation of $F_0(S)$ is the targeted GC content.
+The expected value of $F_\ell$ is always decreasing for increasing weights $\pi_\ell$ (see Supp.
 Mat.~\ref{sec:weight_derivatives}).
 More generally, computing a suitable $\pmb{\pi}$ can be restated as a convex optimization problem, and be efficiently solved using a wide array of methods~\citep{Denise2010,Bendkowski2017}.
-In practice, we use a simple heuristics which starts from an initial weight vector $\pmb{\pi}^{[0]}:=(e^\beta,\dots,e^\beta)$ and, at each iteration, generates a sample $\mathcal{S}$ of sequences.
-The expected value of an energy $E_\ell$ is estimated as $\hat\mu_\ell(\mathcal{S}) = \sum_{S\in\mathcal{S}}E_\ell(S)/|\mathcal{S}|$, and the weights are updated at the $t$-th iteration by %  $\pi_\ell^{[t+1]} = \pi_\ell^{[t]}\cdot\TargetE_\ell/\hat\mu_\ell(\mathcal{S})$.
+In practice, we use a simple heuristics which starts from an initial weight vector $\pmb{\pi}^{[0]}:=(e^\beta,\dots,e^\beta)$ for $\beta=1/(RT)$, T=$37^\circ$, and gas constant $R$.
+Then, at each iteration, it generates samples $\mathcal{S}$ of sequences.
+The expected value of an energy $F_\ell$ is estimated as $\hat\mu_\ell(\mathcal{S}) = \sum_{S\in\mathcal{S}}F_\ell(S)/|\mathcal{S}|$, and the weights are updated at the $t$-th iteration by %  $\pi_\ell^{[t+1]} = \pi_\ell^{[t]}\cdot\TargetE_\ell/\hat\mu_\ell(\mathcal{S})$.
 $\pi_\ell^{[t+1]} = \pi_\ell^{[t]}\cdot \gamma^{\hat\mu_\ell(\mathcal{S})-\TargetE_\ell}$. In practice, the constant $\gamma>1$ is chosen empirically  ($\gamma=1.2$) to achieve effective optimization.
@@
 $\varepsilon\ge 1/\sqrt{n}$, and $\Theta(n^{k/2})$ when
-$\varepsilon=0$. The \GCb\% of designs can also be controlled,
-jointly and in a similar way, as done in
-\Software{IncaRNAtion}~\citep{Reinharz2013}.
+$\varepsilon=0$. 
+%The \GCb\% of designs can also be controlled,
+%jointly and in a similar way, as done in
+%\Software{IncaRNAtion}~\citep{Reinharz2013}.
 

### 2018-08-30 supplement.tex
Getting much closer...

@@
 $(T,\chi)$, such that the values of messages
-$\Message{u}{v}}$ are computed only for cc-valid
+$\Message{u}{v}$ are computed only for cc-valid
 partial sequences $\val\in\partseqs(\separator{u}{v})$. Moreover, the

### 2018-08-30 supplement.tex
Reformat for BMC

@@
-\PassOptionsToPackage{utf8}{inputenc}
-\documentclass[12pt]{bioinfo}
-\copyrightyear{2015} \pubyear{2015}
-
-\access{Advance Access Publication Date: Day Month Year}
-\appnotes{Manuscript Category}
+\documentclass[10pt]{article}
 
@@
 \usepackage{xspace}
-%\usepackage{geometry} 
+
+\usepackage[margin=1in]{geometry}
+
 \usepackage{amsmath,amssymb} 
@@
 \newtheorem{hypothesis}{Working Hypothesis}
-\renewenvironment{proof}[1][]{\noindent \em Proof\ifthenelse{\equal{#1}{}}{}{ (#1)}:~}{}
+\newenvironment{proof}[1][]{\noindent \em Proof\ifthenelse{\equal{#1}{}}{}{ (#1)}:~}{}
 
@@
 
+\newcommand{\citep}[1]{\cite{#1}}
+\newcommand{\citet}[1]{\cite{#1}}
+
 %%% end macro defs
@@
 \begin{document}
-%\newgeometry{margin=2cm}
 \onecolumn
@@
 \appendix
-{\centering \relsize{+4}\textbf{Supplementary Material}\\[1em] Fixed-Parameter Tractable Sampling\\ for RNA Design with Multiple Target Structures\\%
-%{\relsize{-1}\BackupTitle}\\
+{\centering \Large \textbf{Supplementary Material}\\[1em] Fixed-Parameter Tractable Sampling\\ for RNA Design with Multiple Target Structures\\%
 }
-\relsize{+1}
 
@@
 
-\bibliographystyle{natbib}
+\bibliographystyle{plain}
 %\bibliographystyle{achemnat}

### 2018-08-30 RNARedPrint.tex
Rename main file

--- 2018-08-30 RNARedPrint-Bioinfo2018.tex
@@
-\PassOptionsToPackage{utf8}{inputenc}
-\documentclass{bioinfo}
-\copyrightyear{2018} \pubyear{2018}
-
-\access{Advance Access Publication Date: Day Month Year}
-\appnotes{Manuscript Category}
+%% BioMed_Central_Tex_Template_v1.06
+%%                                      %
+%  bmc_article.tex            ver: 1.06 %
+%                                       %
+
+%%IMPORTANT: do not delete the first line of this template
+%%It must be present to enable the BMC Submission system to
+%%recognise this template!!
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                     %%
+%%  LaTeX template for BioMed Central  %%
+%%     journal article submissions     %%
+%%                                     %%
+%%          <8 June 2012>              %%
+%%                                     %%
+%%                                     %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                                                 %%
+%% For instructions on how to fill out this Tex template           %%
+%% document please refer to Readme.html and the instructions for   %%
+%% authors page on the biomed central website                      %%
+%% http://www.biomedcentral.com/info/authors/                      %%
+%%                                                                 %%
+%% Please do not use \input{...} to include other tex files.       %%
+%% Submit your LaTeX manuscript as one .tex document.              %%
+%%                                                                 %%
+%% All additional figures and files should be attached             %%
+%% separately and not embedded in the \TeX\ document itself.       %%
+%%                                                                 %%
+%% BioMed Central currently use the MikTex distribution of         %%
+%% TeX for Windows) of TeX and LaTeX.  This is available from      %%
+%% http://www.miktex.org                                           %%
+%%                                                                 %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%% additional documentclass options:
+%  [doublespacing]
+%  [linenumbers]   - put the line numbers on margins
+
+%%% loading packages, author definitions
+
+%\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
+\documentclass{bmcart}
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                             %%
+%%  If you wish to display your graphics for   %%
+%%  your own use using includegraphic or       %%
+%%  includegraphics, then comment out the      %%
+%%  following two lines of code.               %%
+%%  NB: These line *must* be included when     %%
+%%  submitting to BMC.                         %%
+%%  All figure files must be submitted as      %%
+%%  separate graphics through the BMC          %%
+%%  submission process, not included in the    %%
+%%  submitted article.                         %%
+%%                                             %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%TODO uncomment
+%\def\includegraphic{}
+%\def\includegraphics{}
+
+%%% Put your definitions there:
+\startlocaldefs
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \newtheorem{hypothesis}{Working Hypothesis}
-%\renewenvironment{proof}[1][]{\noindent \em Proof\ifthenelse{\equal{#1}{}}{}{ (#1)}:~}{}
+\newenvironment{proof}[1][]{\noindent \em Proof\ifthenelse{\equal{#1}{}}{}{ (#1)}:~}{}
 
@@
 
+\endlocaldefs
+
+%%% Begin ...
 \begin{document}
-\firstpage{1}
-
-\subtitle{Research Article}
-
-\title[Efficient Sampling for Multi-Target RNA Design]{Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures}
-\author[Hammer, Ponty, Wang and Will]{Stefan Hammer\,$^{\text{\sfb 1,2,3}}$, Yann Ponty\,$^{\text{\sfb 4,5,}\star}$, Wei Wang\,$^{\text{\sfb 4}}$ and Sebastian Will\,$^{\text{\sfb 2}}$}
-\address{$^{\text{\sf 1}}$University Leipzig, Department of Computer Science and Interdisciplinary Center for Bioinformatics, 04107 Leipzig, Germany;
-$^{\text{\sf 2}}$University of Vienna, Faculty of Chemistry, Department of Theoretical Chemistry, 1090 Vienna, Austria;
-$^{\text{\sf 3}}$University of Vienna, Faculty of Computer Science, Research Group Bioinformatics and
-Computational Biology, 1090 Vienna, Austria;
-$^{\text{\sf 4}}$CNRS UMR 7161 LIX, Ecole Polytechnique, Bat. Turing, 91120 Palaiseau, France;
-and $^{\text{\sf 5}}$ AMIBio team, Inria Saclay, Bat Alan Turing, 91120 Palaiseau, France}
-
-\corresp{$^\ast$To whom correspondence should be addressed.}
-
-\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
-
-\editor{Associate Editor: XXXXXXX}
-
-\abstract{\textbf{Motivation:} The design of multi-stable RNA molecules has important applications in biology, medicine, and biotechnology.
+
+%%% Start of article front matter
+\begin{frontmatter}
+
+\begin{fmbox}
+\dochead{Research}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Enter the title of your article here     %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%[Efficient Sampling for Multi-Target RNA Design]
+\title{Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Enter the authors here                   %%
+%%                                          %%
+%% Specify information, if available,       %%
+%% in the form:                             %%
+%%   <key>={<id1>,<id2>}                    %%
+%%   <key>=                                 %%
+%% Comment or delete the keys which are     %%
+%% not used. Repeat \author command as much %%
+%% as required.                             %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% \author[Hammer, Ponty, Wang and Will]{Stefan Hammer\,$^{\text{\sfb 1,2,3}}$, Yann Ponty\,$^{\text{\sfb 4,5,}\star}$, Wei Wang\,$^{\text{\sfb 4}}$ and Sebastian Will\,$^{\text{\sfb 2}}$}
+
+\author[
+    addressref={aff1,aff2,aff3},
+    email={stefan.hammer@uni-leipzig.de}
+]{\inits{F}\fnm{Stefan} \snm{Hammer}}
+%
+\author[
+    addressref={aff4,aff5},
+    corref={aff4},                       % id of corresponding address, if any
+    email={yann.ponty@lix.polytechnique.fr}
+]{\inits{F}\fnm{Yann} \snm{Ponty}}
+%
+\author[
+    addressref={aff4},
+    email={wei.wang@lix.polytechnique.fr}
+]{\inits{F}\fnm{Wei} \snm{Wang}}
+%
+\author[
+    addressref={aff2,aff3},
+    %corref={aff2},                       % id of corresponding address, if any
+    email={will@tbi.univie.ac.at}
+]{\inits{S}\fnm{Sebastian} \snm{Will}}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Enter the authors' addresses here        %%
+%%                                          %%
+%% Repeat \address commands as much as      %%
+%% required.                                %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% \address{$^{\text{\sf 1}}$University Leipzig, Department of Computer Science and Interdisciplinary Center for Bioinformatics, 04107 Leipzig, Germany;
+\address[id=aff1]{%                           % unique id
+  \orgname{Dept.\ Computer Science, and Interdisciplinary Center
+      for Bioinformatics, Univ.\ Leipzig}, 
+  \street{H{\"a}rtelstr.\ 16-18},                     
+  \postcode{D-04107},   
+  \city{Leipzig}, 
+  \cny{Germany}  
+}
+
+% $^{\text{\sf 2}}$University of Vienna, Faculty of Chemistry, Department of Theoretical Chemistry, 1090 Vienna, Austria;
+\address[id=aff2]{%
+    \orgname{Dept.\ Theoretical Chemistry, Univ.\ Vienna},
+    \street{W{\"a}hringerstr.\ 17},
+    \postcode{A-1090}
+    \city{Wien},
+    \cny{Austria}
+}
+
+% $^{\text{\sf 3}}$University of Vienna, Faculty of Computer Science, Research Group Bioinformatics and
+% Computational Biology, 1090 Vienna, Austria;
+\address[id=aff3]{%
+    \orgname{Bioinformatics and Computational Biology Research Group, Univ.\ Vienna},
+    \street{W{\"a}hringerstr.\ 17},
+    \postcode{A-1090}
+    \city{Wien},
+    \cny{Austria}
+}
+
+% $^{\text{\sf 4}}$CNRS UMR 7161 LIX, Ecole Polytechnique, Bat. Turing, 91120 Palaiseau, France;
+\address[id=aff4]{%
+    \orgname{CNRS UMR 7161 LIX, Ecole Polytechnique},
+    \street{Bat. Alan Turing},
+    \postcode{91120}
+    \city{Palaiseau},
+    \cny{France}
+}
+
+% $^{\text{\sf 5}}$ AMIBio team, Inria Saclay, Bat Alan Turing, 91120 Palaiseau, France}
+\address[id=aff5]{%
+    \orgname{AMIBio team, Inria Saclay},
+    \street{Bat. Alan Turing},
+    \postcode{91120}
+    \city{Palaiseau},
+    \cny{France}
+}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Enter short notes here                   %%
+%%                                          %%
+%% Short notes will be after addresses      %%
+%% on first page.                           %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{artnotes}
+%\note{Sample of title note}     % note to the article
+%\note[id=n1]{Equal contributor} % note, connected to author
+\end{artnotes}
+
+\end{fmbox}% comment this for two column layout
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% The Abstract begins here                 %%
+%%                                          %%
+%% Please refer to the Instructions for     %%
+%% authors on http://www.biomedcentral.com  %%
+%% and include the section headings         %%
+%% accordingly for your article type.       %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{abstractbox}
+\begin{abstract}
+  \parttitle{Background}
+  The design of multi-stable RNA molecules has important applications in biology, medicine, and biotechnology.
 Synthetic design approaches profit strongly from effective in-silico methods, which can tremendously impact their cost and feasibility.
 \\
-  \textbf{Results:} We revisit a central ingredient of most in-silico
+  \parttitle{Results} We revisit a central ingredient of most in-silico
   design methods: the sampling of sequences for the design of
@@
   common design benchmark, including pseudoknots. Generating seed sequences with \ourprog{}, we demonstrate significant improvements over the previously best multi-target sampling strategy.\\
-  \textbf{Availability:} Our software is freely available at: \url{https://github.com/yannponty/RNARedPrint}\\
-  \textbf{Contact:} \href{yann.ponty@lix.polytechnique.fr}{yann.ponty@lix.polytechnique.fr}\\
-  \textbf{Supplementary information:} Supplementary data are available
-  at \textit{Bioinformatics} online.}
-
-\maketitle
-
-\enlargethispage{35.1pt}
-\section{Introduction}
+
+  \smallskip
+  \noindent
+  Our free software is available at: \url{https://github.com/yannponty/RNARedPrint}\\
+  Supplementary data are available online.
+\end{abstract}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% The keywords begin here                  %%
+%%                                          %%
+%% Put each keyword in separate \kwd{}.     %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{keyword}
+  \kwd{RNA multi-target design}
+  \kwd{RNA secondary structure}
+  \kwd{Multi-dimensional Boltzmann sampling}
+  \kwd{{$\#$}{\sf P}-hardness of RNA design}
+\end{keyword}
+
+% MSC classifications codes, if any
+%\begin{keyword}[class=AMS]
+%\kwd[Primary ]{}
+%\kwd{}
+%\kwd[; secondary ]{}
+%\end{keyword}
+
+\end{abstractbox}
+%
+%\end{fmbox}% uncomment this for twcolumn layout
+
+\end{frontmatter}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% The Main Body begins here                %%
+%%                                          %%
+%% Please refer to the instructions for     %%
+%% authors on:                              %%
+%% http://www.biomedcentral.com/info/authors%%
+%% and include the section headings         %%
+%% accordingly for your article type.       %%
+%%                                          %%
+%% See the Results and Discussion section   %%
+%% for details on how to create sub-sections%%
+%%                                          %%
+%% use \cite{...} to cite references        %%
+%%  \cite{koon} and                         %%
+%%  \cite{oreg,khar,zvai,xjon,schn,pond}    %%
+%%  \nocite{smith,marg,hunn,advi,koha,mouse}%%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%% start of article main body
+% <put your article body there>
+
+%%%%%%%%%%%%%%%%
+%% Background %%
+%%
+\section*{Background}
 \parHead{Design, applications and motivation for multiple design.}Synthetic biology endeavors the engineering of artificial biological
@@
 
-%\Final{
-%Extensions:
-%\begin{enumerate}
-%  \item What if the input graph is not bipartite? \\
-%  Extract a minimal subset of edges such that new graph is bipartite!\\
-%  $\to{}$ Graph bipartization (cf implementation by Hüffner, JGAA 2009)
-%  \item Is the optimization version in the \emph{weighted base-pair model} already {\sf NP}-hard?
-%  \item Does it help to know that there are cliques of size $k$ during the tree decomposition?\\
-%  If anything, it provides a lower bound on tree decomposition (helps branch and bound algorithms)
-%  \item What if some degree of freedom is allowed in the structure definition?\\ Can we still use tree decomposition?\\
-%  This looks like a very reasonable first step towards the definition of an {\em optimal} DP decomposition for RNA folding (with Pseudoknots) or RNA-RNA interactions\ldots
-%  \item Structures can also be penalized, but they must be specified explicitly.\\
-%  However, any competing structure must consist of helices, and it is possible to penalize each of the $\Theta(n^2)$ helices of any given length $L$. What would be the impact on treewidth?
-%  \item More general energy  models $\to$ Turner. (Pseudoknots)
-%  For this description, we assume that all input structures are non-crossing, while the framework would otherwise support crossing input structures as well as according energy models, e.g.{} the Hotknots model of~\citet{Ren2005}, by considering special pseudoknot loops.
-%\end{enumerate}\label{lbl:theend}
-%}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Backmatter begins here                   %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{backmatter}
+
+\section*{Competing interests}
+  The authors declare that they have no competing interests.
+
+\section*{Author's contributions}
+  All authors wrote the manuscript.
+    % Text for this section \ldots
+    % http://www.biomedcentral.com/submissions/editorial-policies\#authorship
 
@@
 
-\bibliographystyle{natbib}
-%\bibliographystyle{achemnat}
-%\bibliographystyle{plainnat}
-%\bibliographystyle{abbrv}
-%\bibliographystyle{bioinformatics}
-%
-%\bibliographystyle{plain}
-%
-\bibliography{biblio}
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                  The Bibliography                       %%
+%%                                                         %%
+%%  Bmc_mathpys.bst  will be used to                       %%
+%%  create a .BBL file for submission.                     %%
+%%  After submission of the .TEX file,                     %%
+%%  you will be prompted to submit your .BBL file.         %%
+%%                                                         %%
+%%                                                         %%
+%%  Note that the displayed Bibliography will not          %%
+%%  necessarily be rendered by Latex exactly as specified  %%
+%%  in the online Instructions for Authors.                %%
+%%                                                         %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% if your bibliography is in bibtex format, use those commands:
+\bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
+\bibliography{biblio}      % Bibliography file (usually '*.bib' )
+% for author-year bibliography (bmc-mathphys or spbasic)
+% a) write to bib file (bmc-mathphys only)
+% @settings{label, options="nameyear"}
+% b) uncomment next line
+%\nocite{label}
+
+% or include bibliography directly:
+% \begin{thebibliography}
+% \bibitem{b1}
+% \end{thebibliography}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                               %%
+%% Figures                       %%
+%%                               %%
+%% NB: this is for captions and  %%
+%% Titles. All graphics must be  %%
+%% submitted separately and NOT  %%
+%% included in the Tex document  %%
+%%                               %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%
+%% Do not use \listoffigures as most will included as separate files
+
+ \section*{Figures}
+%   \begin{figure}[h!]
+%   \caption{\csentence{Sample figure title.}
+%       A short description of the figure content
+%       should go here.}
+%       \end{figure}
+% 
+% \begin{figure}[h!]
+%   \caption{\csentence{Sample figure title.}
+%       Figure legend text.}
+%       \end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                               %%
+%% Tables                        %%
+%%                               %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%% Use of \listoftables is discouraged.
+%%
+% \section*{Tables}
+% \begin{table}[h!]
+% \caption{Sample table title. This is where the description of the table should go.}
+%       \begin{tabular}{cccc}
+%         \hline
+%            & B1  &B2   & B3\\ \hline
+%         A1 & 0.1 & 0.2 & 0.3\\
+%         A2 & ... & ..  & .\\
+%         A3 & ..  & .   & .\\ \hline
+%       \end{tabular}
+% \end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                               %%
+%% Additional Files              %%
+%%                               %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% \section*{Additional Files}
+%   \subsection*{Additional file 1 --- Sample additional file title}
+%     Additional file descriptions text (including details of how to
+%     view the file, if it is in a non-standard format or the file extension).  This might
+%     refer to a multi-page table or a figure.
+%
+%  \subsection*{Additional file 2 --- Sample additional file title}
+%    Additional file descriptions text.
+
+\end{backmatter}
 \end{document}

### 2018-08-31 supplement.tex
Write clean proof of hyper-graph tree decomposition

@@
 \usepackage{xr}
-\externaldocument{RNARedPrint-Bioinfo2018}
+\externaldocument{RNARedPrint}
 
@@
 \newtheorem{hypothesis}{Working Hypothesis}
-\newenvironment{proof}[1][]{\noindent \em Proof\ifthenelse{\equal{#1}{}}{}{ (#1)}:~}{}
+\newenvironment{proof}[1][]{\noindent \emph{Proof}\ifthenelse{\equal{#1}{}}{}{ (#1)}.~}{\hfill$\Box$}
 
@@
 Crucially, one can show that, as long as the arguments $\dep(f)$ of a function $f$ form a clique within the network $\network$, then there exists at least one node in the tree decomposition that contains $\dep(f)$, thus allowing the evaluation of $f$.
+%
+While this result is certainly part of folklore, we emphasize the importance for the generality of this work and present a brief direct proof for convenience.
+
 \begin{lemma}\label{lem:cliques}
-\todo[inline]{R1: guess it should read "node v" instead of "n"!? please update the proof accordingly!}
-Let $G=(V,E)$ be an undirected graph, and  $(T,\chi)$ be a tree decomposition for $G$. For each clique $C\subset V$, there exists a node $n\in T$ such that $C\subset\chi(v)$.
+Let $G=(V,E)$ be an undirected graph, and  $(T,\chi)$ be a tree decomposition for $G$. For each clique $C\subseteq V$, there exists a node $u \in T$ such that $C\subseteq\chi(u)$.
 \end{lemma}
-\begin{proof}
-  \newcommand{\TransVars}[1]{{\rm tverts}(#1)}
-  \newcommand{\Children}[1]{{\rm Children}(#1)}
-  For the sake of simplicity, let us consider $T$ as rooted on an arbitrary node, inducing an orientation, and denote by $\TransVars{n}\in V$ the set of vertices found in a node $n$ or its descendants, namely
-    $$\TransVars{n} = \chi(n) \bigcup_{n'\in\Children{n}} \TransVars{n'}.$$
 
-  Now consider (one of) the deepest node(s) $n\in T$, such that $C\subset\TransVars{v}$.
-We are going to prove that $C\subset\chi(n)$, using contradiction, by showing that $C\not\subset\chi(n)$ implies that $T$ is not a tree decomposition.
-
-  Indeed, let us assume that $C\not\subset\chi(n)$, then there exists a node $v'\notin\chi(n)$ whose presence in $\TransVars{n}$ stems from its presence in some descendant of $n$.
-Let us denote by $n'\neq n$ the closest descendant of $n$ such that $v'\in\chi(n')$.
-Now, consider a node $v\in C$ such that $v\in \TransVars{v'}$ and $v\notin \TransVars{v'}$.
-Such a node always exists, otherwise $n'$, and not $n$, would be the deepest node such that $C\subset \TransVars{v'}$.
-From the definition of the tree decomposition, we know that neither $n'$ nor its descendants may contain $v$.
-On the other hand, none of the parents or siblings of $n'$ may contain $v'$.
-It follows that is no node $n''\in T$ whose vertex set $\chi(v'')$ includes, at the same time, $v$ and $v'$.
+%\begin{proof}
+% \newcommand{\tn}{v_{\rm T}} %% node in tree
+% \newcommand{\TransVars}[1]{{\rm tverts}(#1)}
+% \newcommand{\Children}[1]{{\rm Children}(#1)}
+%  For the sake of simplicity, let us consider $T$ as rooted on an arbitrary node, inducing an orientation, and denote by $\TransVars{\tn}\in V$ the set of vertices found in the node $\tn$ or its descendants, namely
+%    $$\TransVars{\tn} = \chi(\tn) \cup \bigcup_{\tn'\in\Children{\tn}} \TransVars{\tn'}.$$
+%
+%  Now consider (one of) the deepest node(s) $\tn\in T$, such that $C\subseteq\TransVars{\tn}$.
+We are going to prove that $C\subseteq\chi(\tn)$, using contradiction, by showing that $C\not\subseteq\chi(\tn)$ implies that $T$ is not a tree decomposition.
+%
+%  Assume that $C\not\subseteq\chi(\tn)$, then there exists a node $v'\notin\chi(\tn)$ whose presence in $\TransVars{\tn}$ stems from its presence in some descendant of $\tn$.
+Let us denote by $\tn'\neq \tn$ the closest descendant of $\tn$ such that $v'\in\chi(\tn')$.
+Now, consider a node $v\in C$ such that $v\in \TransVars{\tn'}$ and $v\notin \TransVars{\tn'}$.
+Such a node always exists, otherwise $\tn'$, and not $\tn$, would be the deepest node such that $C\subseteq \TransVars{\tn'}$.
+From the definition of the tree decomposition, we know that neither $\tn'$ nor its descendants may contain $v$.
+On the other hand, none of the parents or siblings of $\tn'$ may contain $v'$.
+It follows that is no node $\tn''\in T$ whose vertex set $\chi(\tn'')$ includes, at the same time, $v$ and $v'$.
 Since both $v$ and $v'$ belong to a clique, one has $\{v,v'\}\in E$.
 The absence of a node in $T$ capturing an edge in $E$ contradicts our initial assumption that $T$ is tree decomposition for $G$.
+%
+%  We conclude that $\tn$ is such that $C\not\subseteq\chi(\tn)$.
+%\end{proof}
 
-  We conclude that $n$ is such that $C\not\subset\chi(n)$.
+\begin{proof}
+  \newcommand{\TransVars}[1]{{\rm tverts}(#1)}%
+  Root $T$ at an arbitrary tree node $r\in T$, which induces a parent/child relation between tree nodes.
+  We say a tree node $u$ \emph{represents} the node $v\in V$ (respectively, the set of nodes $X\subseteq V$) iff $v\in\chi(u)$ ($X\subseteq\chi(u)$).
+  Moreover, denote by $\TransVars{u}\subseteq V$ the vertices
+  represented by the nodes of the subtree of $u$, i.e.
+  $$\TransVars{u} := \chi(u) \cup \bigcup_{\text{children $u'$ of u}} \TransVars{u'}.$$
+
+  Let $u$ denote the node in $T$ with maximal distance to the root $r$,
+  such that still $C\subseteq\TransVars{u}$.
+  Let $C'$ be $C-\chi(u)$.
+  If $C'$ is empty than $u$ represents $C$---nothing remains to be shown;
+  thus assume, $C' \neq \emptyset$.
+  For exactly one of the children $u'$ of $u$,
+  $C'\subseteq\TransVars{u'}$, since otherwise for
+  some node $v'\in C'$, the subtree of the nodes representing $v'$ would be disconnected (property 2 of Def.~\ref{def:treedecomp}). Now, pick some $v'\in C'$ and some
+  $v''\in C-C'$, which must be represented by $u$ or some child $u''$ of $u$. In both cases, there cannot
+  be any node in T that represents the edge $\{v',v''\}\subseteq C$, since $v'$ can only be represented 
+  by nodes in the subtree of $u'$.
 \end{proof}
@@
 
-\bibliographystyle{plain}
+\bibliographystyle{vancouver}
 %\bibliographystyle{achemnat}

### 2018-08-31 RNARedPrint.tex
two column layout
Checkin ps figures
Update on Overleaf.

@@
 
-%\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
-\documentclass{bmcart}
+\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
+%\documentclass{bmcart}
 
@@
 \begin{center}
-    \includegraphics[width=.8\textwidth]{Workflow.pdf}
+    \includegraphics[width=.8\textwidth]{Figs/Workflow}
 \end{center}
@@
 
-\section{Definitions and problem statement}
+\section{Methods}
+\subsection*{Definitions and problem statement}
 \label{sec:problem-statement}
@@
 
-\section{A fixed-parameter tractable (FPT) algorithm for the
+\subsection*{A fixed-parameter tractable (FPT) algorithm for the
   partition function and sampling of Boltzmann-weighted designs}
@@
 
-\section{Results}\label{sec:results}
+\section*{Results}\label{sec:results}
 
@@
     %{\sf \bfseries A}\
-    \includegraphicstop[width=0.99\textwidth]{Figs/energy_shift.pdf}\hfill
+    \includegraphicstop[width=0.99\textwidth]{Figs/energy_shift}\hfill
     %{\sf \bfseries B}\ \includegraphicstop[width=0.55\textwidth]{Figs/PKB00211_PKB00239_0_energy_distribution}
@@
 
-\section{\#{\sf P}-hardness of counting valid designs}\label{sec:counting}
+\section*{\#{\sf P}-hardness of counting valid designs}\label{sec:counting}
 %%% reformulate later
@@
 
-\section{Conclusion}
+\section*{Conclusion}
 Motivated by the---here established---hardness of the problem, we

### 2018-08-31 RNARedPrint.tex
Remove sections numbers and adapt titles as dictated by BMC. Drop/reformulate some sec refs.

@@
 Consequently, for elucidating the theoretical fundament of RNA design sampling, we first need to clarify, whether efficient approaches for counting could exist.
-In Section~\ref{sec:counting}, we show that there can not exists any efficient, polynomial time algorithm unless ${\sf P}={\sf NP}$.
+In this work, we show that there can not exists any efficient, polynomial time algorithm unless ${\sf P}={\sf NP}$.
 Our result relies on a surprising bijection (up to a trivial symmetry) between valid sequences and independent sets of a bipartite graph, being the object of recent breakthroughs in approximate counting complexity~\citep{Bulatov2013,Cai2016}.
@@
 
-To answer these questions, we introduce a generic framework (illustrated in~Fig.~\ref{fig:workflow}) enabling efficient Boltzmann-weighted sampling over RNA sequences with multiple target structures (Section~\ref{sec:FPT}).
+To answer these questions, we introduce a generic framework (illustrated in~Fig.~\ref{fig:workflow}) enabling efficient Boltzmann-weighted sampling over RNA sequences with multiple target structures.
 As main algorithmic contribution of this work, we devise a dynamic programming algorithm, based on a tree decomposition, to compute partition functions and sample sequences from the Boltzmann distribution%
@@
 to a large set of multi-target RNA design instances from a
-representative benchmark (Section~\ref{sec:results}).
-
-\section{Methods}
+representative benchmark.
+
+\section*{Methods}
 \subsection*{Definitions and problem statement}
@@
 
-For our algorithmic description, we translate the concepts of
-Section~\ref{sec:problem-statement} to the formalism of constraint networks, here
+We are going to express our algorithms in the formalism of constraint networks, here
 specialized as RNA design network. This abstraction allows us to base our
@@
 
-The RNA design network describes how to evaluate RNA sequences and, in particular, partial RNA sequences, which is required by our algorithms that construct results for the full sequence by combining partial information.
+The RNA design network describes how to evaluate RNA sequences and, in
+particular, partial RNA sequences, which is required by our algorithms
+that construct results for the full sequence by combining partial
+information.
 %
@@
 
-Since each feature function $f\in\F$ (cf.~Section~\ref{sec:problem-statement})
+Since each feature function $f\in\F$ (as previously introduced)
 depends on only the subset $\dep(f)$ of sequence positions, one can
@@
 
-\subsection{Partition function and Boltzmann sampling through stochastic backtrack}\label{sec:PF}
+\subsection*{Partition function and Boltzmann sampling through stochastic backtrack}\label{sec:PF}
 
@@
 
-\subsection{Computational complexity of the multiple target sampling algorithm}\label{sec:complexity}
+\subsection*{Computational complexity of the multiple target sampling algorithm}\label{sec:complexity}
 
@@
 
-We define the \emph{maximum separator size} $s$ as
-$\max_{u,v\in V} | \separator{u}{v} |$ and denote by  $D$ the maximum size of
+Let $s$ denote the maximum size of any separator set $\separator{u}{v}$ and $D$ denote the maximum size of
 $\difference{u}{v}$ over $(u,v)\in E$.  In the absence
@@
 
-\subsection{Sequence features, constraints, and energy
+\subsection*{Sequence features, constraints, and energy
   models.}\label{sec:energy_models}
@@
 
-\subsection{Extension to multidimensional Boltzmann sampling}\label{sec:multiBoltzmann}
+\subsection*{Extension to multidimensional Boltzmann sampling}\label{sec:multiBoltzmann}
 The flexibility of our framework supports the advanced sampling technique "multidimensional Boltzmann sampling"~\citep{Bodini2010}, which allows to (probabilistically) enforce additional, complex properties of the samples.
@@
 
-\subsection{Targeting Turner energies and \GCb\%}
+\subsection*{Targeting Turner energies and \GCb\%}
 We implemented our Boltzmann sampling strategy (Algorithms
@@
 given target structures and weights $\pi_1,\dots,\pi_k$.  Moreover, we
-used multi-dimensional Boltzmann sampling, as described in
-Section~\ref{sec:multiBoltzmann}, to target specific energies and
+used multi-dimensional Boltzmann sampling to target specific energies and
 \GCb\%.  Our tool \ourprog{} evaluates energies according to the
@@
 
-\subsection{Generating high-quality seeds for further optimization}
+\subsection*{Generating high-quality seeds for further optimization}
 Next, we empirically evaluated the capacity of \ourprog{} to generate seed sequences for multiple (pseudoknotted) target structures, possibly followed by subsequent local optimizations.
@@
 
-\section*{\#{\sf P}-hardness of counting valid designs}\label{sec:counting}
+\subsection*{\#{\sf P}-hardness of counting valid designs}\label{sec:counting}
 %%% reformulate later
@@
 To establish that claim, we consider instances $G=(V_1\cup V_2, E)$ that are connected and bipartite ($E \cap (V_1\times V_2) = E$), noting that hardness over restricted instances implies hardness of the general problem.
-Moreover note that, as observed in Subsec.~\ref{sec:complexity}, assigning a nucleotide to a position $u\in V$ constrains the parity ($\{\Ab,\Gb\}$ or $\{\Cb,\Ub\}$) of all positions in the connected component of $u$.
+Moreover note that, as we explained before, assigning a nucleotide to a position $u\in V$ constrains the parity ($\{\Ab,\Gb\}$ or $\{\Cb,\Ub\}$) of all positions in the connected component of $u$.
 For this reason, we restrict our attention to the counting of valid designs \emph{up to trivial  symmetry} $(\Ab\leftrightarrow \Cb/\Gb\leftrightarrow \Ub)$, by constraining the positions in $V_{1}$ to $\Ab$ and $\Gb$.
@@
 \item $\Psi$ is surjective, i.e. there is a preimage for each element $I\in \IS{G}$. Define $f\in\Design{G}$ as\begin{displaymath}
-f(v) = \left\{\begin{array}{ll@{\qquad}ll}
-\Ab{} & \text{if $v\in V_1$ and $v\in I$} &
-\Cb{} & \text{if $v\in V_2$ and $v\in I$}\\
-\Gb{} & \text{if $v\in V_1$ and $v\not\in I$} &
+f(v) = \left\{\begin{array}{ll} % @{\qquad}ll
+\Ab{} & \text{if $v\in V_1$ and $v\in I$} \\ %&
+\Cb{} & \text{if $v\in V_2$ and $v\in I$} \\
+\Gb{} & \text{if $v\in V_1$ and $v\not\in I$} \\ %&
 \Ub{} & \text{if $v\in V_2$ and $v\not\in I$}

### 2018-08-31 supplement.tex
Fix of hyperedge proof

@@
 
+  Assume there is no node in T that represents $C$.
+
   Let $u$ denote the node in $T$ with maximal distance to the root $r$,
   such that still $C\subseteq\TransVars{u}$.
+  Since there is no single node $u$ that represents $C$, this maximal node is unique, since
+  if a node $u$, $C\subseteq\TransVars{u}$, has two subtrees that cover $C$, then $u$ must represent $C$ due to property 2 of Def.~\ref{def:treedecomp}.
+
   Let $C'$ be $C-\chi(u)$.
-  If $C'$ is empty than $u$ represents $C$---nothing remains to be shown;
-  thus assume, $C' \neq \emptyset$.
+  $C'$ cannot be empty, since otherwise $u$ represents $C$i.
   For exactly one of the children $u'$ of $u$,
@@
   be any node in T that represents the edge $\{v',v''\}\subseteq C$, since $v'$ can only be represented 
-  by nodes in the subtree of $u'$.
+  by nodes in the subtree of $u'$. Contradiction.
 \end{proof}

### 2018-09-03 RNARedPrint.tex
Update on Overleaf.

@@
 
-\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
-%\documentclass{bmcart}
+%\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
+\documentclass[linenumbers,doublespacing]{bmcart}
 
@@
 
-RNA energy models like the Turner energy model~\citep{Turner2009} define an energy function that maps a pair of a sequence $S$ and a structure $R$ to a real value, the \Def{energy}. 
-Typically, such energy functions are additively composed from energy contributions of structure elements (like single base pairs or loops).
+RNA energy models like the Turner energy model~\citep{Turner2009} define a function that maps a pair of a sequence $S$ and a structure $R$ to a real value, the \Def{energy}. 
+Typically, such \Def{energy functions} are additively composed from \Def{energy contributions} of structure elements (like single base pairs or loops).
 It is most instructive, and in this context already practically useful, to develop pour approach for the simple \Def{base pair energy model}.
-This model defines energy contributions $\EbpSym(x,y)$ for base pairs $(x,y)\in\Sigma^2$ that specify the energy function as
+This model defines energy contributions $\EbpSym(x,y)$ for pairs $(x,y)\in\Sigma^2$ that specify the energy function as
 $
@@
 where $\beta$ denotes inverse temperature and $E(S)$ is defined as linear combination of energies $E(S)=\sum_{1\leq \ell \leq k} \ln(\pi_\ell)/\beta E(S,R_\ell)$,
-where the energy $E(S,R_\ell)$ is treated as feature $F_\ell(S) = \EbpSym(S,R_\ell);$ as elaborated later, even Turner energies or pseudoknot energies can be expressed as features based on their additive nature.
+where the energy $E(S,R_\ell)$ is treated as feature $F_\ell(S) = \EbpSym(S,R_\ell);$ as elaborated later, even Turner energies or pseudoknot energies can be expressed as features due to their additive nature.
 
@@
 
-The complexity of the central problem depends fundamentally on the dependency structure due to the entire set of feature functions.
+The complexity of the central problem depends fundamentally on the dependency structure that is imposed by the entire set of feature functions.
 
@@
 and edges $\bigcup_{\ell\in[1,k]} R_\ell$. As well, this graph describes the minimal
-dependency structure of multi-target RNA design, which is present in all relevant settings due to the requirement
-of canonical base pairing.
-
-In more generality, we look at the dependencies due to the set of feature functions $\F$.
-The \Def{dependencies $\dep(f)$} of a single feature function $f$ are precisely described as the minimum set of sequence
-  positions $\I\subseteq\{1,\dots,n\}$, where $f(S)=f(S')$ for all
-  sequences $S$ and $S'$ that have equal nucleotides at the positions in $\I$.
-
-Then, the set $\F$ of feature functions (on sequences of length $n$) induces a
-  \Def{dependency graph} on sequence positions, namely the hypergraph
-  $G_\F=(\{1,\dots,n\} ,\{\dep(f)\mid f\in \F\})$. The \Def{tree decomposition} of this dependency
-  graph is fundamental for our algorithms.
+dependency structure of multi-target RNA design, which is present in all relevant settings, since base pairings must be canonical.
+
+In more generality, we consider dependencies due to the set of feature functions $\F$.
+Recall that each feature function, even if evaluated for entire sequences, depends only on a subset of all sequence positions, which we call the \Def{dependencies of $f$} and write as $\dep(f)$;
+thus, $\dep(f)$ is the minimal subset of $\{1,\dots,n\}$ s.t.~$\forall S,S'\in\Sigma^n: f(S)=f(S') \text{ only if } (\forall i\in\dep(f): S_i=S_i')$.
+
+Then, the set $\F$ of feature functions induces a
+  \Def{dependency graph} on sequence positions by interpreting the dependencies as hyperedges, namely the hypergraph 
+  $G_\F=(\{1,\dots,n\} ,\{\dep(f)\mid f\in \F\})$. The \Def{tree decomposition} of $G_\F$ is fundamental for our algorithms.
 
@@
 
-\subsection*{A fixed-parameter tractable (FPT) algorithm for the
-  partition function and sampling of Boltzmann-weighted designs}
+\subsection*{Partition function and Boltzmann sampling of RNA designs: A fixed-parameter tractable (FPT) algorithm}
 \label{sec:FPT}
@@
 
-Since each feature function $f\in\F$ (as previously introduced)
+Since each feature function $f\in\F$
 depends on only the subset $\dep(f)$ of sequence positions, one can
@@
 %
-Note that we call this object a `network', since $\X$ could be interpreted as nodes, where $\F$ defines the (hyper-)edges between those nodes.
-In this way, it encodes a dependency structure (i.e.
-a graph or `network').
-%The network has the natural interpretation of assigning an energy to sequences. 
-%The \Def{network energy of a sequence} $S$ for the network
-%$\network$ is the sum of the values of all functions in
-%$f\in\F$ evaluated for $S$, i.e.
-%$\sum_{f\in\F} f(S).$
+Notably the term `network' refers to the implied graph structure, namely the dependency graph.
 
@@
 
-Note that one can encode the weight $\pi$ of a features in its feature functions, such that all feature functions can be handled uniformly by our algorithms.
+For the algorithms, it is convenient to encode the weight $\pi$ of each feature in its feature functions, such that all feature functions can be handled uniformly.
 For this purpose, we define the \Def{transformed feature function}
-$$\weighttransform{f}{\pi}(S) := ln(\pi)f(S),$$
+$\weighttransform{f}{\pi}(S) := \ln(\pi)f(S),$
 such that $exp(-\weighttransform{f}{\pi}(S))=\pi^{-f(S)}.$ 
-
-Concretely, for the design in the base pair model targeting \GCb\% and structures $\R$ with weights $\pi_0,\dots,\pi_k$, we construct the set $\F$ of the network as the collection of
+%
+
+Let us now specify the concrete set $\F$ of feature functions that we use
+for the design in the base pair model targeting \GCb\% and structures $\R$ with weights $\pi_0,\dots,\pi_k$. $\F$ consists of
 \begin{itemize}
@@
 
-By this choice of $\F$, the sum $\sum_f\in\F f(S)$ (for any $S\in\Sigma^n$) corresponds to the energy in
+By this choice of $\F$, the sum $\sum_{f\in\F} f(S)$ (for any $S\in\Sigma^n$) corresponds to the energy in
 Eq.~$(\ref{eq:mainproblem}),$ using the equivalence of Eq.~(\ref{eq:feature-energy-transformation}). 
@@
 \subsection*{Generating high-quality seeds for further optimization}
-Next, we empirically evaluated the capacity of \ourprog{} to generate seed sequences for multiple (pseudoknotted) target structures, possibly followed by subsequent local optimizations.
+We empirically evaluated the capacity of \ourprog{} to generate seed sequences for multiple (pseudoknotted) target structures, possibly followed by subsequent local optimizations.
 As a baseline for comparison, we considered \RNAblueprint~\citep{Hammer2017}, the current leading tool for multiple design.
-As a quality measure, we used the multi-stable design objective function introduced by~\citet{Hammer2017}, defined as:
+As a quality measure, we applied the objective function that was introduced by~\citet{Hammer2017} for multi-stable design, defined as:
 % \begin{align}
@@
 % \end{align}
-\begin{align}
+\begin{multline}
   \label{eq:blueprintobjective}
-    \Obj(S) &= \frac{1}{k} \sum_{\ell=1}^{k} (E(S, R_\ell) - \EnsE(S))\notag\\
-   & +\frac{1}{2\binom{k}{2}} \sum\limits_{1\leq\ell<j\leq k}|E(S,R_\ell) - E(S,R_j)|,
-\end{align}
-%
-where $\EnsE(S)$, the \Def{ensemble free energy} of $S$, is computed
+    \Obj(S) = \frac{1}{k} \sum_{\ell=1}^{k} (E(S, R_\ell) - \EnsE(S))\\
+    +\frac{1}{2\binom{k}{2}} \sum\limits_{1\leq\ell<j\leq k}|E(S,R_\ell) - E(S,R_j)|,
+\end{multline}
+%
+where the free energies $E(S, R)$ as well as the \Def{ensemble free energy} $\EnsE(S)$ of $S$ are computed
 by {\tt RNAfold}~\citep{Lorenz2011} in the pseudoknot-free case; for
@@
 %
-\item $\Psi$ is surjective, i.e. there is a preimage for each element $I\in \IS{G}$. Define $f\in\Design{G}$ as\begin{displaymath}
+b\item $\Psi$ is surjective, i.e. there is a preimage for each element $I\in \IS{G}$. Define $f\in\Design{G}$ as\begin{displaymath}
 f(v) = \left\{\begin{array}{ll} % @{\qquad}ll
@@
 
-  Let us now hypothesize the existence of a polynomial-time algorithm $\mathcal{A}$ for \NumDesign over strongly-connected graphs $G$.
-Consider the (polynomial-time) algorithm $\mathcal{A}'$ that first executes $\mathcal{A}$ on $G$ to produce $\NumDesign(G)$, and returns $\NumDesign(G)/2=|\Design{G}| = |\IS{G}|$.
-Clearly $\mathcal{A}'$ solves $\#{\sf BIS}$ in polynomial-time.
-This means that $\NumDesign$  is at least as hard as $\#{\sf BIS}$, thus does not admit a polynomial time exact algorithm unless $\#{\sf P}={\sf FP}$.
+  Given any algorithm $\mathcal{A}$ that solves \NumDesign over strongly-connected graphs $G$, one can  construct an algorithm $\mathcal{A}'$ that solves $\#{\sf BIS}$: 1) calculate $\NumDesign(G)$ by calling $\mathcal{A}$ 2) return $\NumDesign(G)/2=|\Design{G}| = |\IS{G}|$.
+    Obviously, if $\mathcal{A}$ is a polynomial-time algorithm, so is $\mathcal{A'}$.
+  This implies that $\NumDesign$  is at least as hard as $\#{\sf BIS}$, thus does not admit a polynomial time exact algorithm unless $\#{\sf P}={\sf FP}$.
 \end{proof}
@@
 \section*{Author's contributions}
-  All authors wrote the manuscript.
+  YP and SW initiated the initiated the project and worked out the theoretical framework.
+YP and WW wrote the core software, while SW and SH contributed additional scripts.
+SH devised and conducted the computational experiments.
+All authors wrote, read, and approved the manuscript.
     % Text for this section \ldots
-    % http://www.biomedcentral.com/submissions/editorial-policies\#authorship
+    % https://www.biomedcentral.com/getpublished/editorial-policies#authorship
 

### 2018-09-03 RNARedPrint.tex
Update on Overleaf.

@@
 \begin{proof}
-Consider $\Psi: \Design{G} \to \IS{G}$ defined by $ \Psi(f) := \left\{v\in V\mid f(v)\in\{\Ab,\Cb\}\right\}.$
+Consider the mapping $\Psi: \Design{G} \to \IS{G}$ defined by $ \Psi(f) := \left\{v\in V\mid f(v)\in\{\Ab,\Cb\}\right\}.$
 We show that $\Psi$ is bijective:%
@@
 \section*{Author's contributions}
-  YP and SW initiated the initiated the project and worked out the theoretical framework.
-YP and WW wrote the core software, while SW and SH contributed additional scripts.
-SH devised and conducted the computational experiments.
-All authors wrote, read, and approved the manuscript.
+  YP and SW initiated the project and worked out the algorithms.
+  % SW---I guess we could be less detailed...
+  YP contributed the multi-dimensional Boltzmann sampling background and the \#P-hardness proof.
+  SW contributed the integration with the constraint network setting and its specialization to RNA design.
+  YP and WW wrote the core software, while SW and SH contributed additional scripts. SH devised and conducted the computational experiments. All authors wrote, read, and approved the manuscript.
     % Text for this section \ldots

### 2018-09-04 RNARedPrint.tex
Author order; add proper initials

--- 2018-09-03 RNARedPrint.tex
@@
 
-% \author[Hammer, Ponty, Wang and Will]{Stefan Hammer\,$^{\text{\sfb 1,2,3}}$, Yann Ponty\,$^{\text{\sfb 4,5,}\star}$, Wei Wang\,$^{\text{\sfb 4}}$ and Sebastian Will\,$^{\text{\sfb 2}}$}
-
 \author[
@@
     email={stefan.hammer@uni-leipzig.de}
-]{\inits{F}\fnm{Stefan} \snm{Hammer}}
+]{\inits{SH} \fnm{Stefan} \snm{Hammer}}
+%
+\author[
+    addressref={aff4},
+    email={wei.wang@lix.polytechnique.fr}
+]{\inits{WW} \fnm{Wei} \snm{Wang}}
+%
+\author[
+    addressref={aff2,aff3},
+    email={will@tbi.univie.ac.at}
+]{\inits{SW} \fnm{Sebastian} \snm{Will}}
 %
@@
     email={yann.ponty@lix.polytechnique.fr}
-]{\inits{F}\fnm{Yann} \snm{Ponty}}
-%
-\author[
-    addressref={aff4},
-    email={wei.wang@lix.polytechnique.fr}
-]{\inits{F}\fnm{Wei} \snm{Wang}}
-%
-\author[
-    addressref={aff2,aff3},
-    %corref={aff2},                       % id of corresponding address, if any
-    email={will@tbi.univie.ac.at}
-]{\inits{S}\fnm{Sebastian} \snm{Will}}
+]{\inits{YP} \fnm{Yann} \snm{Ponty}}
 

### 2018-09-05 supplement.tex
Add section/plots "benefits of neg for pos design" to supplement

@@
 
+\section{Immediate benefits of negative design for positive design}
+
+Already previous work on RNA Design, e.g.
+INFO-RNA and Incarnation, observed that---briefly put---positive design benefits from negative design.
+Here, more concretely, we only \emph{demonstrate} this effect (for small examples), where we consider \Def{positive design} as
+designing RNA sequences that fold with high probability $Pr(R|S)$ into the given target structure $R$; and \Def{negative design}, as generating one or many sequences that have a low energy for $R$.
+
+For a double-stem structure, we produce a uniform and a weighted Boltzmann sampling, each of 1000 sequenes. For each sampled sequence $S$, we determine the Turner energy $E(S,R)$ as well as the
+%difference $E_\Delta(S,R)$ of $E(S,R)$ to the ensemble energy $E_{\text{ens}}(S)$ of $S$ 
+probability $Pr(R|S)$ of the structure $R$ in the ensemble of sequence $S$ 
+(where the energies are calculated in the Turner model using RNAfold).
+
+Note that low energy $E(S|R)$ is (non-linearly) correlated with high probability $Pr(S|R)$, the probability to see $S$ in the Boltzmann distribution of sequences folding into $R$.
+Recall that the probability $Pr(R|S)$ can be calculated as $\exp(-E(S,R)/RT)/Z(S)$,
+where $Z(S)=\sum_{R'} \exp(-E(S,R')/RT)$. 
+%
+Consequently, Figure~\ref{fig:negforpos} ultimately illustrates the dependency between $Pr(S|R)$ and $Pr(R|S)$ for uniform samples and low energy samples (where the latter can be efficiently generated due to our Boltzmann sampling strategy).
+
+% and high energy difference $E_\Delta(S,R)=E_{\text{ens}}(S)-E(S,R)$, with high probability $\Pr(R|S)$.
+% 
+% To see the latter, recall that the \Def{ensemble energy $E_{\text{Ens}}(S)$} is defined as
+% $$
+% E_{\text{ens}}(S) = -RT \ln( Z(S) ),
+% $$ 
+% based on the partition function $Z(S)=\sum_{R'} \exp(-E(S,R')/RT)$
+% such that 
+% $$
+% E(S,R) - E_{\text{ens}}(S) 
+% =
+% -RT \ln Pr(R|S),
+% $$  
+% which can be derived by term transformation 
+% \begin{align*}
+% -RT \ln P(R|S) &= -RT \ln [ exp(-E(S,R)/RT) / Z(S) ] \\&= -RT \ln (exp(-E(S,R)/RT)) - (-RT\ln Z(S))
+% \\&=  E(S,R) - E_{\text{Ens}}(S). 
+% \end{align*}
+
+\begin{figure}
+\textbf{A}~\includegraphicstop[width=0.48\textwidth]{Figs/Plots/doublestem_weight1}%
+\textbf{B}~\includegraphicstop[width=0.48\textwidth]{Figs/Plots/doublestem_weight500}
+\caption{Comparision of (A) Uniform sampling to (B) Boltzmann sampling with weight $\pi=500$.
+For given structure $R$, we plot the energies $E(S,R)$ of sampled sequences against the probabilities $Pr(R|S)$; the former is controlled by Boltzmann sampling, while the latter is a common positive design objective.}
+\label{fig:negforpos}
+\end{figure}
+
 \bibliographystyle{vancouver}

### 2018-09-06 supplement.tex
Fix mixup positive vs negative

@@
 
-\section{Immediate benefits of negative design for positive design}
+\section{Immediate benefits of positive design for negative design}
 
-Already previous work on RNA Design, e.g.
-INFO-RNA and Incarnation, observed that---briefly put---positive design benefits from negative design.
-Here, more concretely, we only \emph{demonstrate} this effect (for small examples), where we consider \Def{positive design} as
-designing RNA sequences that fold with high probability $Pr(R|S)$ into the given target structure $R$; and \Def{negative design}, as generating one or many sequences that have a low energy for $R$.
+Already previous work on RNA Design, e.g. INFO-RNA and Incarnation, observed
+that---briefly put---negative design benefits from positive design. Here, more
+concretely, we only \emph{demonstrate} this effect (for small examples), where
+we consider \Def{negative design} as  
+designing RNA sequences that fold with high probability $Pr(R|S)$ into the
+given target structure $R$; and \Def{positive design}, as generating one or
+many sequences that have a low energy for $R$.
 
-For a double-stem structure, we produce a uniform and a weighted Boltzmann sampling, each of 1000 sequenes. For each sampled sequence $S$, we determine the Turner energy $E(S,R)$ as well as the
+For a double-stem structure, we produce a uniform and a weighted Boltzmann
+sampling, each of 10000 sequenes. For each sampled sequence $S$, we determine
+the Turner energy $E(S,R)$ as well as the
 %difference $E_\Delta(S,R)$ of $E(S,R)$ to the ensemble energy $E_{\text{ens}}(S)$ of $S$ 
@@
 
-Note that low energy $E(S|R)$ is (non-linearly) correlated with high probability $Pr(S|R)$, the probability to see $S$ in the Boltzmann distribution of sequences folding into $R$.
-Recall that the probability $Pr(R|S)$ can be calculated as $\exp(-E(S,R)/RT)/Z(S)$,
+Note that low energy $E(S|R)$ is (non-linearly) correlated with high
+probability $Pr(S|R)$, the probability to see $S$ in the Boltzmann distribution
+of sequences folding into $R$. Recall that the probability $Pr(R|S)$ can be
+calculated as $\exp(-E(S,R)/RT)/Z(S)$,
 where $Z(S)=\sum_{R'} \exp(-E(S,R')/RT)$. 
 %
-Consequently, Figure~\ref{fig:negforpos} ultimately illustrates the dependency between $Pr(S|R)$ and $Pr(R|S)$ for uniform samples and low energy samples (where the latter can be efficiently generated due to our Boltzmann sampling strategy).
+Consequently, Figure~\ref{fig:negforpos} ultimately illustrates the dependency
+between $Pr(S|R)$ and $Pr(R|S)$ for uniform samples and low energy samples
+(where the latter can be efficiently generated due to our Boltzmann sampling
+strategy).
 
@@
 \textbf{B}~\includegraphicstop[width=0.48\textwidth]{Figs/Plots/doublestem_weight500}
-\caption{Comparision of (A) Uniform sampling to (B) Boltzmann sampling with weight $\pi=500$.
-For given structure $R$, we plot the energies $E(S,R)$ of sampled sequences against the probabilities $Pr(R|S)$; the former is controlled by Boltzmann sampling, while the latter is a common positive design objective.}
+\caption{Comparision of (A) Uniform sampling to (B) Boltzmann sampling with
+weight $\pi=500$. For given structure $R$, we plot the energies $E(S,R)$ of
+sampled sequences against the probabilities $Pr(R|S)$; the former is controlled
+by Boltzmann sampling, while the latter is a common negative design objective.}
 \label{fig:negforpos}

### 2018-09-10 RNARedPrint.tex
Minor edits

@@
 %\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
-\documentclass[linenumbers,doublespacing]{bmcart}
+\documentclass{bmcart}
 
@@
 Typically, such \Def{energy functions} are additively composed from \Def{energy contributions} of structure elements (like single base pairs or loops).
-It is most instructive, and in this context already practically useful, to develop pour approach for the simple \Def{base pair energy model}.
+It is most instructive, and in this context already practically useful, to develop our approach for the simple \Def{base pair energy model}.
 This model defines energy contributions $\EbpSym(x,y)$ for pairs $(x,y)\in\Sigma^2$ that specify the energy function as
@@
 Recall that we aim to generate random sequences that satisfy a number of constraints like the compatibility of paired bases.
-Moreover, we want to enforce that generated samples satisfy complex properties, which we express in terms of \emph{features}.
+Moreover, we want to enforce complex properties, which we express in terms of \emph{features}.
 An example of such a feature is the \GCb\% (the corresponding property of a sequence is to have a specific \GCb\%).

### 2018-09-12 RNARedPrint.tex

@@
     email={will@tbi.univie.ac.at}
-]{\inits{SW} \fnm{Sebastian} \snm{Will}}
+]{\inits{SW} \fnm{Sebastian} \snm{Will$^*$}}
 %
@@
 
-An \Def{RNA sequence $S$} is a word over the \Def{nucleotides
+An \Def{RNA sequence $S$} is a word over the \Def{nucleotide
   alphabet} $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}.$
@@
   length $n$} is a set of \Def{base pairs} $(i,j)$, where
-$1\leq i<j\leq n$, where for all different $(i,j), (i',j')\in R$:
+$1\leq i<j\leq n$ and for all different $(i,j), (i',j')\in R$:
 $\{i,j\}\cap\{i',j'\}=\emptyset$ (``degree $\leq$ 1'').
@@
 \Def{Valid base pair} must pair bases from
-$\B:=\left\{\{\Ab,\Ub\},\{\Gb,\Cb\},\{\Gb,\Ub\}\right\}.$
-Consequently, $S$ is \Def{valid} for $R$, iff $\{S_i,S_j\}\in \B$ for
+$\B:=\left\{\{\Ab,\Ub\},\{\Gb,\Cb\},\{\Gb,\Ub\}\right\};$
+we call $S$ \Def{valid} for $R$, iff $\{S_i,S_j\}\in \B$ for
 all $(i,j)\in R$.
 
-RNA energy models like the Turner energy model~\citep{Turner2009} define a function that maps a pair of a sequence $S$ and a structure $R$ to a real value, the \Def{energy}. 
+RNA energy models like the Turner energy model~\citep{Turner2009} define a function $E$ that maps a pair of a sequence $S$ and a structure $R$ to a real value, the \Def{energy} $E(S,R)$. 
 Typically, such \Def{energy functions} are additively composed from \Def{energy contributions} of structure elements (like single base pairs or loops).
@@
 
-As the structural targets of our design, we fix the set of RNA structures
-$\R:=\{R_1, \dots, R_k\}$ for sequences of length $n$.
+As the structural targets of our design, we fix the set $\R:=\{R_1, \dots, R_k\}$ of $k$ RNA structures for sequences of length $n$.
 Recall that we aim to generate random sequences that satisfy a number of constraints like the compatibility of paired bases.
@@
 We emphasize that our approach is designed to profit (in terms of computational complexity) from feature functions that each depend only on few sequence positions.
-For example, the $\GCb\%$ feature is expressed by feature functions $f^{GC}_i$ where $f^{GC}_i(S)$ equals 1 if $S_i='G'$ or $S_i='C'$ and 0 otherwise.
-Similarly the base pair energy for a target structure $R$ is expressed as feature with feature functions $f^R_{ij}(S)=\EbpSym(S_i,S_j),$ such that $F(S)=\EbpSym(S,R)$.
-In both cases, the feature functions depend on the nucleotides at only a small (constantly bounded) number of sequence positions (one and two positions, respectively).
+For example, the $\GCb\%$ feature is expressed by $n$ feature functions $f^{GC}_i$ where $f^{GC}_i(S)$ equals 1 if $S_i='G'$ or $S_i='C'$ and 0 otherwise.
+Similarly the base pair energy for a target structure $R$ is expressed as feature with $|R|$-many feature functions $f^R_{ij}(S)=\EbpSym(S_i,S_j)$ for each $(i,j)\in R,$ such that $F^R(S)=\EbpSym(S,R)$.
+In both cases, the feature functions depend on the nucleotides at only a small, constantly bounded number of sequence positions (one and two positions, respectively).
 
@@
 
-The workhorse of our approach is the efficient (i.e.~fixed-parameter tractable) computation of feature-dependent partition functions over sequences.
+The workhorse of our approach is the efficient (more precisely, fixed-parameter tractable) computation of feature-dependent partition functions over sequences.
 Given are the features $F_0,\dots,F_k$, which encode the energies functions for the fixed structures and the function for the \GCb\% all in terms of simple feature functions.
-For weights $\pi_0,\dots,\pi_k$  compute the partition function
+For weights $\pi_0,\dots,\pi_k$, one wants to compute the partition function
   \begin{equation}
@@
 \label{eq:feature-energy-transformation}
-\prod_{0\leq \ell \leq k} \pi_\ell^{-F_\ell(S)} = exp(-\beta E(S)),
+\prod_{1\leq \ell \leq k} \pi_\ell^{-F_\ell(S)} = exp(-\beta E(S)),
 \end{equation}
@@
 
-\paragraph{Preliminaries on the complexity}
+\paragraph{Further preliminaries for complexity and algorithmic considerations}
 
@@
 In more generality, we consider dependencies due to the set of feature functions $\F$.
-Recall that each feature function, even if evaluated for entire sequences, depends only on a subset of all sequence positions, which we call the \Def{dependencies of $f$} and write as $\dep(f)$;
+Recall that each feature function, even if `formally' evaluated for entire sequences, depends only on a subset of all sequence positions, which we call the \Def{dependencies of $f$} and write as $\dep(f)$;
 thus, $\dep(f)$ is the minimal subset of $\{1,\dots,n\}$ s.t.~$\forall S,S'\in\Sigma^n: f(S)=f(S') \text{ only if } (\forall i\in\dep(f): S_i=S_i')$.
@@
 specialized as RNA design network. This abstraction allows us to base our
-algorithm on the cluster tree elimination (CTE) of~\citet{Dechter2013}.
+algorithm on cluster tree elimination (CTE)~\citet{Dechter2013}.
 

### 2018-09-12 RNARedPrint.tex

@@
 
-By this choice of $\F$, the sum $\sum_{f\in\F} f(S)$ (for any $S\in\Sigma^n$) corresponds to the energy in
-Eq.~$(\ref{eq:mainproblem}),$ using the equivalence of Eq.~(\ref{eq:feature-energy-transformation}). 
-Consequently, the network encodes the partition function $\partfun{\pi_0,\dots,\pi_k}$ of Eq.~(\ref{eq:mainproblem}).
+By these definitions, the network encodes the partition function $\partfun{\pi_0,\dots,\pi_k}$ of Eq.~(\ref{eq:mainproblem}); also confer Eq.~(\ref{eq:feature-energy-transformation}).
 Subsequently, we show how to efficiently compute this partition function based on the network.

### 2018-09-12 RNARedPrint.tex
Edits in long Jitsi session with Yann

@@
 \section*{Methods}
-\subsection*{Definitions and problem statement}
+\subsection*{Problem statement}
 \label{sec:problem-statement}
-
-An \Def{RNA sequence $S$} is a word over the \Def{nucleotide
-  alphabet} $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}.$
-An \Def{RNA (secondary) structure $R$ of
-  length $n$} is a set of \Def{base pairs} $(i,j)$, where
-$1\leq i<j\leq n$ and for all different $(i,j), (i',j')\in R$:
-$\{i,j\}\cap\{i',j'\}=\emptyset$ (``degree $\leq$ 1'').
-%
+The main computational problem addressed in this work is to generate sequences over the alphabet $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}$ that achieving prescribed free-energies and $\GCb\%$.
+Despite their apparent diversity, such criteria can be homogeneously abstracted as constraints on the value of \Def{features}, each defined as sums over atomic feature contributions depending on subsets of sequence positions.
+
+\paragraph{Central computational problem.}
+Let us consider $k$ features $F_1,\dots F_k$ to target all of them simultaneously.
+We base our strategy for targeting the $k$ features $F_1,\dots,F_k$ on generating samples from a Boltzmann distribution parameterized by weights $\pi_1,\dots,\pi_k$, one for each feature.
+These weights essentially allows us to shift the sample distributions towards specific targets.
+Sampling from this distribution requires to compute corresponding partition functions, such that we can draw sequences with probabilities proportional to their Boltzmann weight:
+\begin{equation}
+\label{eq:sample-distribution}
+\Pr(S|\pi_1,\dots,\pi_k) \propto \prod_{1\leq \ell\leq k} \pi_\ell^{-F_\ell(S)}.
+\end{equation}
+
+The workhorse of our approach is the efficient (more precisely, fixed-parameter tractable) computation of feature-dependent partition functions over sequences.
+Given are the features $F_1,\dots,F_k$, which encode the energies functions for the fixed structures and the function for the \GCb\% all in terms of simple atomic energy contributions.
+For weights $\pi_1,\dots,\pi_k$, one wants to compute the partition function
+  \begin{equation}
+    \label{eq:mainproblem}
+    \partfun{\pi_1,\dots,\pi_k} = \sum_{S\in\Sigma^n} \prod_{1\leq \ell\leq k} \pi_\ell^{-F_\ell(S)}.
+  \end{equation}
+
+\paragraph{Representing \GCb\%-content and energies as features.}
+Formally, we define a \Def{feature $F$} as function on sequences, whose value is obtained by summing over an associated set of \Def{atomic feature contributions} $f_1, f_2\ldots$, briefly called \emph{contributions}, that each are associated with a restricted set of positions, such that
+$$
+F(S) =\!\!\!\!\sum_{\substack{\text{contribution $f_i$ of $F$}\\ \text{on positions $x_1,\ldots,x_p$}}}\!\!\!\! f_i(\{x_1\mapsto S_{x_1},\ldots,x_p\mapsto S_{x_p}\}).
+$$
+
+As simple example, the $\GCb\%$ feature is expressed by $n$ contributions $f^{\GC}_i$ where $f^{\GC}_i(\{i\mapsto S_i\})$ equals $-1$ if $S_i=\Gb$ or $\Cb$, and $0$ otherwise.
+
+For the definition of energy features, we abstract an \Def{RNA (secondary) structure $R$ of  length $n$} as a set of \Def{base pairs} $(i,j)$, where
+$1\leq i<j\leq n$ and for all different $(i,j), (i',j')\in R$: $\{i,j\}\cap\{i',j'\}=\emptyset$ (``degree $\leq$ 1'')---without precluding pseudoknots.
 % We call an RNA structure $R$ \Def{non-crossing}, iff it does not
@@
 %
-\Def{Valid base pair} must pair bases from
-$\B:=\left\{\{\Ab,\Ub\},\{\Gb,\Cb\},\{\Gb,\Ub\}\right\};$
-we call $S$ \Def{valid} for $R$, iff $\{S_i,S_j\}\in \B$ for
-all $(i,j)\in R$.
-
-RNA energy models like the Turner energy model~\citep{Turner2009} define a function $E$ that maps a pair of a sequence $S$ and a structure $R$ to a real value, the \Def{energy} $E(S,R)$. 
-Typically, such \Def{energy functions} are additively composed from \Def{energy contributions} of structure elements (like single base pairs or loops).
+We call $S$ \Def{valid} for $R$, iff $\{S_i,S_j\}\in \B$ for
+all $(i,j)\in R$, where $\B:=\left\{\{\Ab,\Ub\},\{\Gb,\Cb\},\{\Gb,\Ub\}\right\}$ describes the valid base pairs.
+
+In a simple \Def{base pair energy model}, the free-energy of a target structure $R$ is expressed as feature with contributions $f^R_{ij}(\{i\mapsto x, j\mapsto y\})=\EbpSym(x,y)$ for each $(i,j)\in R,$ such that $F^R(S)=\EbpSym(S,R)$.
+
+Note that for \GCb\% as well as for the base pair energy features, the contributions depend on the nucleotides at only a small, constantly bounded number of sequence positions (one and two positions, respectively).
+This is important for the feasibility of our approach, which is designed to profit from contributions that each depend on only few sequence positions.
+We emphasize that nevertheless, the same framework can capture more expressive RNA energy models like the Turner energy model~\citep{Turner2009}, since such models generally define a function $E$ that maps a pair of a sequence $S$ and a structure $R$ to a real value, the \Def{energy} $E(S,R)$.
+Here, our approach automatically adapts to the more complex energy functions, leading to a controlled increase of the computational complexity.
+
+%Typically, such \Def{energy functions} are additively composed from \Def{energy contributions} of structure elements (like single base pairs or loops).
 It is most instructive, and in this context already practically useful, to develop our approach for the simple \Def{base pair energy model}.
 This model defines energy contributions $\EbpSym(x,y)$ for pairs $(x,y)\in\Sigma^2$ that specify the energy function as
-$
-\EbpSym(S,R) = \sum_{(i,j)\in R} \EbpSym(S_i,S_j)
-$
-for each sequence $S\in\Sigma^n$ and structure $R$ for length $n$.
-
-As the structural targets of our design, we fix the set $\R:=\{R_1, \dots, R_k\}$ of $k$ RNA structures for sequences of length $n$.
-Recall that we aim to generate random sequences that satisfy a number of constraints like the compatibility of paired bases.
+%$
+%\EbpSym(S,R) = \sum_{(i,j)\in R} \EbpSym(S_i,S_j)
+%$
+%for each sequence $S\in\Sigma^n$ and structure $R$ for length $n$.
+
+%As the structural targets of our design, we fix the set $\R:=\{R_1, \dots, R_k\}$ of $k$ RNA structures for sequences of length $n$.
+%Recall that we aim to generate random sequences that satisfy a number of constraints like the compatibility of paired bases.
 Moreover, we want to enforce complex properties, which we express in terms of \emph{features}.
@@
 
-Formally, we define a \Def{feature $F$} as function on sequences, which is expressed by summing over an associated set of \Def{feature functions} on sequences, such that
-$$
-F(S) = \sum_{\text{feature function $f$ associcated with $F$}} f(S).
-$$
-We emphasize that our approach is designed to profit (in terms of computational complexity) from feature functions that each depend only on few sequence positions.
-For example, the $\GCb\%$ feature is expressed by $n$ feature functions $f^{GC}_i$ where $f^{GC}_i(S)$ equals 1 if $S_i='G'$ or $S_i='C'$ and 0 otherwise.
-Similarly the base pair energy for a target structure $R$ is expressed as feature with $|R|$-many feature functions $f^R_{ij}(S)=\EbpSym(S_i,S_j)$ for each $(i,j)\in R,$ such that $F^R(S)=\EbpSym(S,R)$.
-In both cases, the feature functions depend on the nucleotides at only a small, constantly bounded number of sequence positions (one and two positions, respectively).
-
-\paragraph{Central problem.}
-Let us consider the $k$ energy features $F_1,\dots F_k$ together with the negative $\GCb\%$ as feature $F_0$ to target all of them simultaneously.
-We base our strategy for targeting the $k+1$ features $F_0,\dots,F_k$ on generating samples from a Boltzmann distribution parametrized by weights $\pi_0,\dots,\pi_k$, one for each feature.
-These weights will essentially allow us to shift the sample distributions towards specific targets.
-Sampling from this distribution requires to compute corresponding partition functions, such that we can draw sequences with probabilities proportional to their Boltzmann weight:
-\begin{equation}
-\label{eq:sample-distribution}
-\Pr(S|\pi_0,\dots,\pi_k) \propto \prod_{0\leq \ell\leq k} \pi_\ell^{-F_\ell(S)}.
-\end{equation}
-
-The workhorse of our approach is the efficient (more precisely, fixed-parameter tractable) computation of feature-dependent partition functions over sequences.
-Given are the features $F_0,\dots,F_k$, which encode the energies functions for the fixed structures and the function for the \GCb\% all in terms of simple feature functions.
-For weights $\pi_0,\dots,\pi_k$, one wants to compute the partition function
-  \begin{equation}
-    \label{eq:mainproblem}
-    \partfun{\pi_0,\dots,\pi_k} = \sum_{S\in\Sigma^n} \prod_{0\leq \ell\leq k} \pi_\ell^{-F_\ell(S)}.
-  \end{equation}
-
-To stress the analogy to, in the RNA context more familiar (e.g.~\citep{McCaskill1990}), Boltzmann distributions over the structure ensemble of a sequence $S$, we emphasize the equivalence
+Finally, to stress the analogy to, in the RNA context more familiar (e.g.~\citep{McCaskill1990}), Boltzmann distributions over the structure ensemble of a sequence $S$, we emphasize the equivalence
 \begin{equation}
@@
  \paragraph{Further preliminaries for complexity and algorithmic considerations} 
-
-The complexity of the central problem depends fundamentally on the dependency structure that is imposed by the entire set of feature functions.
-
-For design in the base pair energy model or the related counting problem, the dependencies are immediately derived from the set of target structures 
-$\R$. It induces the
-\Def{base pair dependency graph} $G_{\R}$ with nodes $\{1,\dots,n\}$
-and edges $\bigcup_{\ell\in[1,k]} R_\ell$. As well, this graph describes the minimal
-dependency structure of multi-target RNA design, which is present in all relevant settings, since base pairings must be canonical.
-
-In more generality, we consider dependencies due to the set of feature functions $\F$.
-Recall that each feature function, even if `formally' evaluated for entire sequences, depends only on a subset of all sequence positions, which we call the \Def{dependencies of $f$} and write as $\dep(f)$;
+The complexity of the central problem depends fundamentally on the dependency structure that is imposed by the entire set of contributions.
+
+%For design in the base pair energy model or the related counting problem, the dependencies are immediately derived from the set of target structures $\R$.
+
+First, for capturing the minimal
+dependency structure of multi-target RNA design, which is induced by the requirement of canonical base pairing,
+we define the \Def{base pair dependency graph} $G_{\R}$ as the graph with nodes $\{1,\dots,n\}$
+and edges $\bigcup_{\ell\in[1,k]} R_\ell$.
+
+More generally, we consider dependencies induced by the complete set of contributions $\F := \bigcup_{F}\{f\mid f\text{ contribution of } F\}$.
+Recall that each contribution, even if `formally' evaluated for entire sequences, depends only on a subset of all sequence positions, which we call the \Def{dependencies of $f$} and write as $\dep(f)$;
 thus, $\dep(f)$ is the minimal subset of $\{1,\dots,n\}$ s.t.~$\forall S,S'\in\Sigma^n: f(S)=f(S') \text{ only if } (\forall i\in\dep(f): S_i=S_i')$.
 
-Then, the set $\F$ of feature functions induces a
-  \Def{dependency graph} on sequence positions by interpreting the dependencies as hyperedges, namely the hypergraph 
+Then, the set $\F$ of contributions induces a
+  \Def{dependency graph} $G_\F$ on sequence positions by interpreting the dependencies as hyperedges, namely the hypergraph 
   $G_\F=(\{1,\dots,n\} ,\{\dep(f)\mid f\in \F\})$. The \Def{tree decomposition} of $G_\F$ is fundamental for our algorithms.
@@
 
-\subsection*{Partition function and Boltzmann sampling of RNA designs: A fixed-parameter tractable (FPT) algorithm}
+\subsection*{Fixed-parameter tractable (FPT) algorithm}
 \label{sec:FPT}
 
-We are going to express our algorithms in the formalism of constraint networks, here
-specialized as RNA design network. This abstraction allows us to base our
-algorithm on cluster tree elimination (CTE)~\citet{Dechter2013}.
-
-The RNA design network describes how to evaluate RNA sequences and, in
-particular, partial RNA sequences, which is required by our algorithms
-that construct results for the full sequence by combining partial
-information.
-%
-\Def{Partial RNA sequences} are words $\val$ in
-$(\Sigma\cup\{?\})^n$, where the symbol '?' represents not yet determined nucleotides; consequently, the positions, where $\val_i\in\Sigma$,
+%We express our algorithms in the formalism of constraint networks, here
+%specialized as RNA design network. This abstraction allows us to base our
+%algorithm on cluster tree elimination (CTE)~\citet{Dechter2013}.
+
+%The RNA design network describes how to evaluate RNA sequences and, in
+%particular, partial RNA sequences, which is
+
+Our algorithms iteratively merge partial solutions. To capture this we introduce 
+\Def{Partial RNA sequences} as words $\val$ in
+$(\Sigma\cup\{?\})^n$, where the symbol '?' indicate unassigned positions; consequently, positions such that  $\val_i\in\Sigma$,
 are called \Def{determined} by $\val$ ($1\leq i\leq n$).
@@
 
-Since each feature function $f\in\F$
+Since each contribution $f\in\F$
 depends on only the subset $\dep(f)$ of sequence positions, one can
@@
 
-\begin{definition}
-An \Def{RNA design network} (for sequences of length $n$) is a tuple $\network=(\X,\F)$, where\vspace{-6pt}
-\begin{itemize}
-\item $\X$ is the set of sequence positions $1,\dots,n$
-\item $\F$ is a set of \Def{feature functions} $f:\Sigma^n\to\real$
-\end{itemize}
-\end{definition}
-%
-Notably the term `network' refers to the implied graph structure, namely the dependency graph.
+%\begin{definition}
+%An \Def{RNA design network} (for sequences of length $n$) is a tuple $\network=(\X,\F)$, where
+%\begin{itemize}
+%\item $\X$ is the set of sequence positions $1,\dots,n$
+%\item $\F$ is a set of \Def{contributions} $f:\Sigma^n\to\real$
+%\end{itemize}
+%\end{definition}
+%
+%Notably the term `network' refers to the implied graph structure, namely the dependency graph.
 
@@
 
-For the algorithms, it is convenient to encode the weight $\pi$ of each feature in its feature functions, such that all feature functions can be handled uniformly.
-For this purpose, we define the \Def{transformed feature function}
+For the algorithms, it is convenient to encode the weight $\pi$ of each feature in its contributions, such that all contributions can be handled uniformly.
+For this purpose, we define the \Def{transformed contribution}
 $\weighttransform{f}{\pi}(S) := \ln(\pi)f(S),$
@@
 
-Let us now specify the concrete set $\F$ of feature functions that we use
-for the design in the base pair model targeting \GCb\% and structures $\R$ with weights $\pi_0,\dots,\pi_k$. $\F$ consists of
+Let us now specify the concrete set $\F$ of contributions that we use
+for the design in the base pair model targeting \GCb\% and structures $\R$ with weights $\pi_0,\dots,\pi_k$. The set $\F$ thus consists of
 \begin{itemize}
-\item the transformed feature functions $\weighttransform{f^{GC}_i}{\pi_0}$ for our feature $F_0$, i.e.~$GC\%$ ($i=1,\dots,n$)
-\item the transformed feature functions $\weighttransform{f^{R_\ell}_{ij}}{\pi_\ell}$for each structure $R_\ell\in\R$ and $(i,j)\in R_\ell.$
+\item the transformed contributions $\weighttransform{f^{\GCb}_i}{\pi_0}$ for the ~$\GCb\%$ feature ($i=1,\dots,n$)
+\item the transformed contributions $\weighttransform{f^{R_\ell}_{ij}}{\pi_\ell}$for each structure $R_\ell\in\R$ and $(i,j)\in R_\ell.$
 \end{itemize}
@@
 
-\subsection*{Partition function and Boltzmann sampling through stochastic backtrack}\label{sec:PF}
+\subsubsection*{Partition function and  stochastic backtrack algorithms}\label{sec:PF}
 
@@
 \begin{theorem}[Complexities]\label{th:complexities}
-  For sequence length $n$, $k$ target structures and treewidth $w$
+  For sequence length $n$, $k$ target structures and treewidth $w$,
   % and a base pair dependency graph having $c$ connected components%
-  , $t$ sequences
+  $t$ sequences
   are generated from the Boltzmann distribution  in
@@
 
-\subsection*{Sequence features, constraints, and energy
-  models.}\label{sec:energy_models}
-
-Going beyond our encoding of the base pair model energy features and the \GCb\% feature, the feature functions provide more generality and allow expressing complex features of the
-sequences alone, e.g.\ rewarding or penalizing specific sequence
-motifs, as well as features depending on the target structures.
-Furthermore, constraints, which enforce or forbid features, are
-naturally expressed by assigning infinite penalties to invalid
-(partial) sequences (even if in implementations, explicit handling of constraints is advantageous). In particular, the framework
-captures various RNA energy models, they can be handled efficiently (in the sense of FPT), as long as the dependencies of their additive energy contributions are bounded.
+\subsection*{Design for expressive energy models.}\label{sec:energy_models}
+
+\textbf{Move the approximation of more complex models from "Results" to this subsection}
+
+%%!!!!!!!!!!!!!!! consider to merge into disciussion
+%Going beyond our encoding of the base pair model energy features and the \GCb\% feature, the contributions provide more generality and allow expressing complex features of the
+%sequences alone, e.g.\ rewarding or penalizing specific sequence
+%motifs, as well as features depending on the target structures.
+%Furthermore, constraints, which enforce or forbid features, are
+%naturally expressed by assigning infinite penalties to invalid
+%(partial) sequences (even if in implementations, explicit handling of constraints is advantageous). In particular, the %framework
+%aptures various RNA energy models, they can be handled efficiently (in the sense of FPT), as long as the dependencies of their additive energy contributions are bounded.
 
@@
 \section*{Results}\label{sec:results}
-
-\subsection*{Targeting Turner energies and \GCb\%}
+Our Boltzmann sampling strategy (Algorithms
+\ref{alg:pf} and \ref{alg:sampling}) were implemented, using a combination of {\tt C++} and {\tt Python}, in a software called \ourprog{}, available at:
+
+{\centering \url{https://github.com/yannponty/RNARedPrint}\\[.3em]}
+
+The driver script of \ourprog{} takes as input a set of (secondary) structures, ... and outputs ....
+
+{\bfseries Stefan: Please list available options/features from the Python interface + output format}
+
+\subsection*{Effectively targeting Turner energies}
 We implemented our Boltzmann sampling strategy (Algorithms

### 2018-09-12 RNARedPrint.tex
Make consistent again, edits, ...

@@
 
-\newcommand{\substitute}[2]{#1\!\oplus\!#2}
+\newcommand{\substitute}[2]{#1\cup#2}
 %\newcommand{\evalfor}[2]{#1\llbracket{}#2\rrbracket{}}
@@
   establishing the $\#${\sf P}-hardness of uniform sampling. Modeling
-  the problem as a constraint network, \ourprog{} supports generic
+  the problem by specializing constraint networks, \ourprog{} supports generic
   Boltzmann-weighted sampling for arbitrary additive RNA energy
@@
 %Uniform sampling is handled as a special case of Boltzmann sampling, where each valid sequence receives energy zero and---consequently---computing partition functions specializes to counting.
-By evaluating (partial) sequences in a weighted constraint
+By evaluating (partial) sequences in a specialized weighted constraint
 network, we support arbitrary multi-ary constraints and thus
@@
 \paragraph{Representing \GCb\%-content and energies as features.}
-Formally, we define a \Def{feature $F$} as function on sequences, whose value is obtained by summing over an associated set of \Def{atomic feature contributions} $f_1, f_2\ldots$, briefly called \emph{contributions}, that each are associated with a restricted set of positions, such that
+Formally, we define a \Def{feature $F$} as function on sequences, whose value is obtained by summing over an associated set of \Def{atomic feature contributions} $f_1, f_2\ldots$, briefly called \emph{contributions}, that each are associated with a restricted set of positions, its \emph{dependencies $\dep(f)$}, such that
 $$
-F(S) =\!\!\!\!\sum_{\substack{\text{contribution $f_i$ of $F$}\\ \text{on positions $x_1,\ldots,x_p$}}}\!\!\!\! f_i(\{x_1\mapsto S_{x_1},\ldots,x_p\mapsto S_{x_p}\}).
+F(S) =\!\!\!\!\!\!\sum_{\substack{\text{contribution $f_i$ of $F$}\\\text{with $\dep(f)=\{x_1,\ldots,x_p\}$}}}\!\!\!\!\!\! f_i(\{x_1\mapsto S_{x_1},\ldots,x_p\mapsto S_{x_p}\}).
 $$
 
-As simple example, the $\GCb\%$ feature is expressed by $n$ contributions $f^{\GC}_i$ where $f^{\GC}_i(\{i\mapsto S_i\})$ equals $-1$ if $S_i=\Gb$ or $\Cb$, and $0$ otherwise.
+As simple example, the $\GCb\%$ feature is expressed by $n$ contributions $f^{\GCb}_i$ where $f^{\GCb}_i(\{i\mapsto S_i\})$ equals $-1$ if $S_i=\Gb$ or $\Cb$, and $0$ otherwise.
 
@@
 % $i\leq i'\leq j \leq j'.$
-%
-We call $S$ \Def{valid} for $R$, iff $\{S_i,S_j\}\in \B$ for
-all $(i,j)\in R$, where $\B:=\left\{\{\Ab,\Ub\},\{\Gb,\Cb\},\{\Gb,\Ub\}\right\}$ describes the valid base pairs.
 
@@
 Here, our approach automatically adapts to the more complex energy functions, leading to a controlled increase of the computational complexity.
+
+Typically, such RNA energy models consider a sequence $S$ \Def{valid} for $R$, iff $\{S_i,S_j\}\in \B$ for
+all $(i,j)\in R$, where $\B:=\left\{\{\Ab,\Ub\},\{\Gb,\Cb\},\{\Gb,\Ub\}\right\}$ describes the valid base pairs.
 
@@
 
- \paragraph{Further preliminaries for complexity and algorithmic considerations} 
-The complexity of the central problem depends fundamentally on the dependency structure that is imposed by the entire set of contributions.
+\paragraph{Dependency structure and complexity}
+%The complexity of the central problem depends fundamentally on the dependency structure that is imposed by the entire set of contributions.
 
@@
 
-More generally, we consider dependencies induced by the complete set of contributions $\F := \bigcup_{F}\{f\mid f\text{ contribution of } F\}$.
-Recall that each contribution, even if `formally' evaluated for entire sequences, depends only on a subset of all sequence positions, which we call the \Def{dependencies of $f$} and write as $\dep(f)$;
-thus, $\dep(f)$ is the minimal subset of $\{1,\dots,n\}$ s.t.~$\forall S,S'\in\Sigma^n: f(S)=f(S') \text{ only if } (\forall i\in\dep(f): S_i=S_i')$.
-
-Then, the set $\F$ of contributions induces a
+More generally, we consider dependencies induced by the complete set of contributions $\F := \bigcup_{\ell}\{f\mid f\text{ contribution of } F_\ell\}$.
+%
+%Let us denote the \emph{dependencies of a contribution $f$} as \dep{f}.
+%
+The set $\F$ of contributions induces a
   \Def{dependency graph} $G_\F$ on sequence positions by interpreting the dependencies as hyperedges, namely the hypergraph 
@@
 
-Our algorithms iteratively merge partial solutions. To capture this we introduce 
-\Def{Partial RNA sequences} as words $\val$ in
-$(\Sigma\cup\{?\})^n$, where the symbol '?' indicate unassigned positions; consequently, positions such that  $\val_i\in\Sigma$,
-are called \Def{determined} by $\val$ ($1\leq i\leq n$).
-%
-We remark that (partial) sequences correspond to the more general concept of (partial) assignments in
-constraint networks.
-
-Since each contribution $f\in\F$
-depends on only the subset $\dep(f)$ of sequence positions, one can
-evaluate them for partial sequences $\val$ that determine (at least)
-the nucleotides at all positions in $\dep(f)$. Thus for any such function $f$
-and partial sequence $\val$, we know that $f(S)$ has the same value for any sequence $S\in \Sigma^n$ that equals $\val$ at its determined positions; thus, we define $\evalfor{f}{\val}$ as this common value.
+To formalize our algorithms, which iteratively merges evaluations of partial solutions, we extend the idea of atomic feature contributions, which are evaluated at sets of the form $\{x_1\mapsto v_1, \dots, x_p\mapsto v_d\}$.
+Let us call the latter object \Def{partial sequence}; they are handy to specify partial knowledge on the entire sequence at some point of the algorithm.
+Easily, we can extend the definition of contributions $f$ to sets $\{x_1\mapsto v_1, \dots, x_p\mapsto v_d\}$, where $\{x_1\dots x_d\}$ is any super-set of $\dep(f)$ by ignoring the superfluous assignments $x\mapsto v$, where $x\not\in\dep(f)$.
+
+Our algorithms specialize the idea of cluster tree elimination (CTE)~\citep{Dechter2013}, which operates on constraint networks.
+In this correspondence, (partial) sequences specialize (partial) assignments and the constraint network would be given by variables for each sequence position, constraints due to valid base pairing, and the set of atomic feature contributions $\cal F$.
 
@@
 
-For the algorithms, it is convenient to encode the weight $\pi$ of each feature in its contributions, such that all contributions can be handled uniformly.
+For algorithmic treatment, it is convenient to encode the weight $\pi$ of each feature in its contributions, such that all contributions can be handled uniformly.
 For this purpose, we define the \Def{transformed contribution}
@@
 
-By these definitions, the network encodes the partition function $\partfun{\pi_0,\dots,\pi_k}$ of Eq.~(\ref{eq:mainproblem}); also confer Eq.~(\ref{eq:feature-energy-transformation}).
-Subsequently, we show how to efficiently compute this partition function based on the network.
-
-\subsubsection*{Partition function and  stochastic backtrack algorithms}\label{sec:PF}
-
-The partition function
-over an RNA design network can be computed by dynamic programming based
-on a tree decomposition of the network's dependency graph,
-i.e.~essentially cluster tree elimination (CTE) \citep{Dechter2013}. Note that analogous algorithms could be constructed to count valid sequences or minimize energy.
+By these definitions, the set $\F$ encodes the partition function $\partfun{\pi_0,\dots,\pi_k}$ of Eq.~(\ref{eq:mainproblem}); also confer Eq.~(\ref{eq:feature-energy-transformation}). 
+%Subsequently, we show how to efficiently compute this partition function based on the 
+%corresponding dependency graph and its tree decomposition.
+
+\subsubsection*{Partition function and stochastic backtracking}\label{sec:PF}
+
+We compute the partition function (as specified by $\F$) by dynamic programming based
+on a tree decomposition of the $\F$'s dependency graph.
+Note that analogous algorithms could be constructed to count valid sequences or minimize energy.
 %This yields fixed parameter tractable (FPT) algorithms, based on the treewidth.
@@
 
-Our algorithms are formulated to process a \Def{cluster tree} of the
-RNA design network, which is a tuple $(T,\chi,\phi)$, where
+Our algorithms are formulated to process a \Def{cluster tree}, which is a tuple $(T,\chi,\phi)$, where
 $(T,\chi)$ is a tree decomposition of $G_\F$, and $\phi(v)$ represents
@@
 
-Our algorithms need to iterate over sets of partial sequences that determine specific sequence positions
-from a set $\Y$; we denote the set of all partial sequences that exactly determine the positions in $\Y$ by $\partseqs(\Y)$.
-Furthermore, given partial sequences $\val$ and
-$\val'$, we define the \Def{combined partial sequence $\substitute{\val}{\val'}$} such that
+Since our algorithms iterate over specific sets of sequence positions, we moreover define
+the \Def{set of partial sequences $\partseqs(\Y)$ determining $\Y\subseteq\{1,\dots,n\}$} as
 $$
-(\substitute{\val}{\val'})_i :=
-\begin{cases}
-  \val'_i & \text{if } \val'_i\in \Sigma\\
-  \val_i & \text{otherwise}
-\end{cases}
+\partseqs(\Y)=\{ x\mapsto v \mid \text{$x\in\Y$ and $v\in \{\Ab,\Cb,\Gb,\Ub\}$}\}.
 $$
-
-\paragraph{FPT Computation of the partition function}
-Given is the RNA design network $\network=(\X,\F)$ and its cluster
-tree $(T,\chi,\phi)$.  
-%
-We assume several properties of this cluster tree, which otherwise could be easily obtained:
+%
+We assume several properties of the given cluster tree (reflecting $\F$), which otherwise could be easily obtained:
 \begin{itemize}
@@
 on the positions $\dep(m)\subseteq \X$ and yields a partition function
-in $\real\cup\{\infty\}$. The message from $u$ to $v$ represents the
+in $\real$. The message from $u$ to $v$ represents the
 partition functions of the subtree of $u$ for all possible partial
@@
 Alg.~\ref{alg:pf}, multiplying the 0-ary messages sent to the root $r$
-yields the total partition function of the network (i.e.~due to proper encoding the partition function of our design problem)
+yields the total partition function (i.e.~due to proper encoding the partition function of our design problem)
 \begin{math}
@@
 
-The partition functions can then direct a stochastic backtracking procedure to sample sequences from the Boltzmann distribution based on the given design network $\network$.
+The partition functions can then direct a stochastic backtracking procedure to sample sequences from the Boltzmann distribution (according to $\F$).
 %
@@
 
-\textbf{Move the approximation of more complex models from "Results" to this subsection}
-
-%%!!!!!!!!!!!!!!! consider to merge into disciussion
-%Going beyond our encoding of the base pair model energy features and the \GCb\% feature, the contributions provide more generality and allow expressing complex features of the
-%sequences alone, e.g.\ rewarding or penalizing specific sequence
-%motifs, as well as features depending on the target structures.
-%Furthermore, constraints, which enforce or forbid features, are
-%naturally expressed by assigning infinite penalties to invalid
-%(partial) sequences (even if in implementations, explicit handling of constraints is advantageous). In particular, the %framework
-%aptures various RNA energy models, they can be handled efficiently (in the sense of FPT), as long as the dependencies of their additive energy contributions are bounded.
-
 % \paragraph{Nearest-neighbor model.}
@@
 
+So far, we discussed sampling only in base pair energy models, although one is finally interested in targeting energies in realistic energy models like the Turner model or pseudoknot models like {\tt HotKnots}~\citep{Ren2005}.
+Notably, for this purpose, the discussed sampling strategy can be extended in two ways: 1) one can directly sample based on more expressive energy models, 2) sampling in a simple energy model can be used to approximate sampling in more complex models.
+
+\paragraph{Sampling based on more expressive energy features}
 Complex \Def{loop-based}
@@
 
-The arity of the introduced functions provides an important bound on the
-treewidth of the network (and therefore computational
-complexity). Thus, it is noteworthy that the base pair energy model
+The arity of the introduced functions bounds the
+treewidth of the dependency graph. %(and therefore computational complexity).
+Thus, it is noteworthy that the base pair energy model
 requires only binary functions; the stacking model, only quarternary
@@
 
+\paragraph{Approximating Turner Energy using Simpler Energy Models}
+To capture the realistic Turner model $\EnergyTurner$ while keeping
+a reasonable computational cost, we exploit a tight correlation between
+$\EnergyTurner$ and the fitted stacking model $\EnergyStacking$
+(Supp. Fig~\ref{fig:training-cor}). More precisely, we observed a
+structure-specific affine dependency between the Turner and stacking
+energy models, so that
+$\EnergyTurner(S;R) \approx \gamma\cdot \EnergyStacking(S;R) + \delta$
+for any structure $R$ and sequence $S$. We inferred the
+$(\gamma,\delta)$ parameters from a set of sequences generated with
+homogenous weights $w=e^{\beta}$, tuning only \GCb\% to a
+predetermined value.  Finally, we adjusted the targeted energies
+within our stacking model to
+$\EnergyStacking^{\star} = (\EnergyTurner^{\star}- \delta)/\gamma$ in
+order to reach, on average, the targeted energy
+$\EnergyTurner^{\star}$ in the Turner model.
+
 \subsection*{Extension to multidimensional Boltzmann sampling}\label{sec:multiBoltzmann}
@@
 %Such a distribution can be induced by a simple modification of the functions described in Sec.~\ref{sec:energy_models}, where any energy function $E(\val)$ for a structure $\ell$ is replaced by $E'(\val):= \ln(\pi_\ell)\, E(\val)/\beta.$
-Due to the construction of $\F$ of our RNA design network,
+Due to the construction of $\F$,
 the probability of a sequence $S$ is thus proportional to
@@
 
-To capture the realistic Turner model $\EnergyTurner$ while keeping
-a reasonable computational cost, we exploit a tight correlation between
-$\EnergyTurner$ and the fitted stacking model $\EnergyStacking$
-(Supp. Fig~\ref{fig:training-cor}). More precisely, we observed a
-structure-specific affine dependency between the Turner and stacking
-energy models, so that
-$\EnergyTurner(S;R) \approx \gamma\cdot \EnergyStacking(S;R) + \delta$
-for any structure $R$ and sequence $S$. We inferred the
-$(\gamma,\delta)$ parameters from a set of sequences generated with
-homogenous weights $w=e^{\beta}$, tuning only \GCb\% to a
-predetermined value.  Finally, we adjusted the targeted energies
-within our stacking model to
-$\EnergyStacking^{\star} = (\EnergyTurner^{\star}- \delta)/\gamma$ in
-order to reach, on average, the targeted energy
-$\EnergyTurner^{\star}$ in the Turner model.
-
-Fig.~\ref{fig:energydist-pk} illustrates the described strategy. For
+Fig.~\ref{fig:energydist-pk} illustrates how well complex realistic energy models can be approximated based on simpler, but better tractable ones. For
 the two target structures of Fig.~\ref{fig:energydist-pk}B,
@@
 \section*{Conclusion}
+%%!!!!!!!!!!!!!!! consider to merge into disciussion
+%Going beyond our encoding of the base pair model energy features and the \GCb\% feature, the contributions provide more generality and allow expressing complex features of the
+%sequences alone, e.g.\ rewarding or penalizing specific sequence
+%motifs, as well as features depending on the target structures.
+%Furthermore, constraints, which enforce or forbid features, are
+%naturally expressed by assigning infinite penalties to invalid
+%(partial) sequences (even if in implementations, explicit handling of constraints is advantageous). In particular, the %framework
+%aptures various RNA energy models, they can be handled efficiently (in the sense of FPT), as long as the dependencies of their additive energy contributions are bounded.
+
 Motivated by the---here established---hardness of the problem, we
@@
   YP and SW initiated the project and worked out the algorithms.
-  % SW---I guess we could be less detailed...
-  YP contributed the multi-dimensional Boltzmann sampling background and the \#P-hardness proof.
-  SW contributed the integration with the constraint network setting and its specialization to RNA design.
+  % SW---I guess we don't need to be super-detailed, otherwise something like...
+  %YP contributed the multi-dimensional Boltzmann sampling background and the \#P-hardness proof.
+  %SW contributed the integration with the constraint network setting and its specialization to RNA design.
   YP and WW wrote the core software, while SW and SH contributed additional scripts. SH devised and conducted the computational experiments. All authors wrote, read, and approved the manuscript.

### 2018-09-13 RNARedPrint.tex
Fix PS(Y)

@@
 Since our algorithms iterate over specific sets of sequence positions, we moreover define
-the \Def{set of partial sequences $\partseqs(\Y)$ determining $\Y\subseteq\{1,\dots,n\}$} as
-$$
-\partseqs(\Y)=\{ x\mapsto v \mid \text{$x\in\Y$ and $v\in \{\Ab,\Cb,\Gb,\Ub\}$}\}.
-$$
+the \Def{set $\partseqs(\Y)$ of all partial sequences determining the positions of $\Y\subseteq\{1,\dots,n\}$} in all combinations of nucleotides $\{\Ab,\Cb,\Gb,\Ub\}$.
 %

### 2018-09-13 RNARedPrint.tex
Edits; move #P to end of Methods section

@@
 \begin{fmbox}
-\dochead{Research}
+\dochead{Methodology}
+%\dochead{Research}
 
@@
   The design of multi-stable RNA molecules has important applications in biology, medicine, and biotechnology.
-Synthetic design approaches profit strongly from effective in-silico methods, which can tremendously impact their cost and feasibility.
+Synthetic design approaches profit strongly from effective in-silico methods, which impact their cost and feasibility.
 \\
-  \parttitle{Results} We revisit a central ingredient of most in-silico
-  design methods: the sampling of sequences for the design of
-  multi-target structures, possibly including pseudoknots. For this
-  task, we present the efficient, tree decomposition-based algorithm
-  \ourprog{}. Our fixed parameter tractable approach is underpinned by
-  establishing the $\#${\sf P}-hardness of uniform sampling. Modeling
+%
+  \parttitle{Results} We devise a novel approach to a central ingredient of most in-silico
+  design methods: the sampling of sequences for multi-target RNA design. For this task, we present the efficient (fixed-parameter tractable) approach \ourprog{}.
+  % 
+  Modeling
   the problem by specializing constraint networks, \ourprog{} supports generic
-  Boltzmann-weighted sampling for arbitrary additive RNA energy
-  models; this enables the generation of RNA sequences meeting
-  specific goals like expected free energies or \GCb-content. Finally,
-  we empirically study general properties of the approach and generate
+  Boltzmann-weighted sampling, which enables the positive design of RNA sequences
+  with specific free energies (for each of multiple, possibly pseudoknotted, target structures) and \GCb-content.
+%
+  Securing the theoretical foundations of our method, we establish the $\#${\sf P}-hardness of uniform sampling. 
+%
+  Finally, we empirically study general properties of the approach and generate
   biologically relevant multi-target Boltzmann-weighted designs for a
-  common design benchmark, including pseudoknots. Generating seed sequences with \ourprog{}, we demonstrate significant improvements over the previously best multi-target sampling strategy.\\
+  common design benchmark, including pseudoknots.
+  Generating seed sequences with \ourprog{}, we demonstrate significant improvements over the previously best multi-target sampling strategy.\\
+%
+  \parttitle{Conclusion} \ourprog{} introduces a novel, flexible and---as we demonstrate--effective approach to multi-target RNA design, which promises broad applicability and extensibility.
 
@@
   Our free software is available at: \url{https://github.com/yannponty/RNARedPrint}\\
+
+  \noindent
   Supplementary data are available online.
@@
 \section*{Background}
-\parHead{Design, applications and motivation for multiple design.}Synthetic biology endeavors the engineering of artificial biological
+\parHead{Design, applications and motivation for multiple design.}Synthetic biology strives for the engineering of artificial biological
 systems, promising broad applications in biology, biotechnology and
@@
 macromolecules with highly specific properties and programmable functions.
-RNA present itself as a well-suited tool for a
-rational design targeting specific functions~\citep{Kushwaha2016}. RNA function is tightly
+RNAs are particularly well-suited tools for
+rational design targeting specific functions~\citep{Kushwaha2016}: on the one hand, RNA function is tightly
 coupled to the formation of secondary structure, as well as changes in
 base pairing propensities and the accessibility of regions, e.g.\ by
-burying or exposing interaction sites~\citep{Rodrigo2014}. At the same time, the
+burying or exposing interaction sites~\citep{Rodrigo2014}; on the other hand, the
 thermodynamics of RNA secondary structure is well understood and its prediction is
 computationally tractable~\citep{McCaskill1990}. Thus,  structure can serve as effective
-proxy for rational design approaches, ultimately targeting catalytic~\citep{Zhang2013} or regulatory functions.
+proxy for rational design approaches, ultimately targeting catalytic or regulatory functions~\citep{Zhang2013}.
 
@@
 conformations. Classic examples include riboswitches, which
-notoriously adopt different stable structures upon binding a specific
+adopt different stable structures upon binding a specific
 ligand. Riboswitches have been a popular application of rational
 design~\citep{Wachsmuth2013,Domin2017}, partly motivated by their
-capacity to act as biosensors~\citep{Findeiss2017}. At the
-co-transcriptional level, certain RNA families feature alternative,
+capacity to act as biosensors~\citep{Findeiss2017}.
+In particular due to the kinetic coupling of RNA folding with RNA transcription, RNA families can feature alternative,
 evolutionarily conserved, transient structures~\citep{Zhu2013}, which
-facilitate the ultimate adoption of functional structures at full
-elongation.  More generally, simultaneous compatibility to multiple
+are essential for the formation of their functional structures.  More generally, simultaneous compatibility to multiple
 structures is a relevant design objective for engineering kinetically
 controlled RNAs, finally targeting prescribed folding pathways. Thus,
-modern applications of RNA design often target multiple structures,
+advanced applications of RNA design often target multiple structures,
 additionally aiming at other features, such as specific
@@
 \parHead{On the importance of sampling for design.}
-Many computational methods for RNA design rely on similar overall
-strategies: initially generating one or several \Def{seed} sequences
-and optimizing them subsequently. In many cases, the seed quality was
+Many computational methods for RNA design follow the ``generate-and-optimize'' strategy: they randomly generate one or several \Def{seed} sequences, which are then optimized.
+In many cases, the seed quality was
 found to be critical for the empirical performance of RNA design
-methods~\citep{Levin2012}. For instance, random seed generation
+methods~\citep{Levin2012}. Generally, random seed generation
 improves the prospect of subsequent optimizations, helping to overcome
@@
 across designs~\citep{Reinharz2013}.  For single-target approaches,
-\Software{INFO-RNA}~\citep{Busch2006} made significant improvements
-mainly by starting its local search from the minimum energy sequence
-for the target structure instead of (uniform) random sequences for the
-early \Software{RNAinverse} algorithm~\citep{Hofacker1994}. This
-strategy was later shown to result in unrealistically high
-\GCb\% in designed sequences. To address this issue,
-tools like \Software{antaRNA}~\citep{Kleinkauf2015} and \Software{IncaRNAtion}~\citep{Reinharz2013} control \GCb\%; the latter based on adaptive sampling.
+\Software{INFO-RNA}~\citep{Busch2006} could significantly improve the success rate over \Software{RNAinverse}~\citep{Hofacker1994}, by starting its local search from the minimum energy sequence
+for the target structure. Since this
+strategy typically designs sequences with unrealistically high
+\GCb\%,
+more recent approaches like \Software{antaRNA}~\citep{Kleinkauf2015} and \Software{IncaRNAtion}~\citep{Reinharz2013} explicitly control \GCb\%; the applying adaptive sampling.
 
 \parHead{Specificities and similarities of multi-target design.}
-Specifically, for multi-target design, virtually all available methods~\citep{Lyngsoe2012,HoenerzuSiederdissen2013,Taneda2015,Hammer2017} follow the same overall generate-and-optimize strategy.
+Specifically for multi-target design, virtually all available methods~\citep{Lyngsoe2012,HoenerzuSiederdissen2013,Taneda2015,Hammer2017} follow the same overall generate-and-optimize strategy.
 %
@@
 coloring the sequence positions with nucleotides, such that the ends of each base pair receive compatible colors.
-Initially the method decomposes the graph hierarchically using ear decomposition, and then \Def{precomputes} the number of valid sequences
+Initially the method decomposes the graph hierarchically, using ear decomposition, and then \Def{precomputes} the number of valid sequences
 within each subgraph. The decomposition is then reinterpreted as a
-decision tree to perform a \Def{stochastic backtrack}, inspired by
+decision tree to perform \Def{stochastic backtracking}, inspired by
 Ding and Lawrence~\citep{Ding2003}. Uniform sampling is achieved by
@@
 While uniform sampling in \Software{RNAdesign}/\RNAblueprint{} turned out to be very successful,
-the work on these tools gave no definitive answer on the \emph{complexity of uniform sampling for multi-target design}.
-On the one hand, time and space requirements grow exponential in the general case,
+the work on these tools gave no definitive answer on the \emph{complexity of uniform sampling for multi-target design}. While time and space requirements grow exponential in the general case,
 the decomposition strategy seems to allow practical applications.
@@
 Boltzmann sampling as discussed in this work or sampling satisfying multiple additional constraints and properties).
-All such extensions are still bounded by the complexity of the multi-target uniform sampling problem, where counting of valid designs is the dominating step (since stochastic backtrack can be performed in linear time per sample).
-
-Consequently, for elucidating the theoretical fundament of RNA design sampling, we first need to clarify, whether efficient approaches for counting could exist.
+All such extensions are still bounded by the complexity of the multi-target uniform sampling problem, where counting of valid designs is the dominating step (note that stochastic backtracking can be performed in linear time per sample).
+
+Consequently, for elucidating the theoretical fundament of RNA design sampling, we first need to clarify, whether efficient counting is possible.
 In this work, we show that there can not exists any efficient, polynomial time algorithm unless ${\sf P}={\sf NP}$.
-Our result relies on a surprising bijection (up to a trivial symmetry) between valid sequences and independent sets of a bipartite graph, being the object of recent breakthroughs in approximate counting complexity~\citep{Bulatov2013,Cai2016}.
-
-As already indicated by \Software{RNAdesign}, the hardness of counting (and conjectured hardness of sampling), does still not rule out practically applicable algorithms for counting and sampling.
+Our result relies on a surprising bijection between valid sequences and independent sets of a bipartite graph; the latter being the object of recent breakthroughs in approximate counting complexity~\citep{Bulatov2013,Cai2016}.
+
+As already demonstrated by \Software{RNAdesign}, the hardness of counting (and conjectured hardness of sampling), does still not rule out practically applicable algorithms for counting and sampling.
 Addressing the following questions in more detail and generality led to the practical advances in this work, which extend the expressivity, flexibility, and performance of multi-structure design: \emph{How to sample, in a general way, from a Boltzmann distribution based on expressive energy models?
@@
 % (Subsection~\ref{sec:PF})
-. We show that these algorithms are \Def{fixed-parameter tractable (FPT)} for the \Def{treewidth},
-in other words: they have polynomial time and space complexity for a fixed treewidth%
+.
+We show that these algorithms are time and space-efficient depending on a specific property of the tree decomposition, namely the treewidth; in short, our algorithms are fixed-parameter tractable (FPT).%
 % Subsection~\ref{sec:complexity}
 %
-. Due to the generality of our method, we can moreover strongly limit this parameter in practice by using state-of-the-art tree decomposition algorithms.
+Due to the generality of our method, we can moreover strongly limit this parameter in practice by using state-of-the-art tree decomposition algorithms.
 %Uniform sampling is handled as a special case of Boltzmann sampling, where each valid sequence receives energy zero and---consequently---computing partition functions specializes to counting.
@@
 Since our algorithms iterate over specific sets of sequence positions, we moreover define
-the \Def{set $\partseqs(\Y)$ of all partial sequences determining the positions of $\Y\subseteq\{1,\dots,n\}$} in all combinations of nucleotides $\{\Ab,\Cb,\Gb,\Ub\}$.
+the \Def{set $\partseqs(\Y)$ of all partial sequences determining the positions of $\Y=\subseteq\{1,\dots,n\}$} in all combinations of nucleotides $\{\Ab,\Cb,\Gb,\Ub\}$,
+i.e.~for $\Y=\{y_1,\dots,y_{q}\}$,
+\begin{align*}
+\partseqs(\Y) = \{ & \{ y_i\mapsto v_i \mid i=1,\dots,q \} \\& | (v_1,\dots,v_{q}) \in \{\Ab,\Cb,\Gb,\Ub\}^{q} \}.
+\end{align*}
 %
@@
 
-\subsection*{Design for expressive energy models.}\label{sec:energy_models}
+\subsection*{Design for expressive energy models}\label{sec:energy_models}
 
@@
 %\Software{IncaRNAtion}~\citep{Reinharz2013}.
+
+\subsection*{\#{\sf P}-hardness of counting valid designs}\label{sec:counting}
+%%% reformulate later
+%We define the following notations: $\real^\infty:=\real\cup\{\infty\}$, $[i,j]:=\{i,\ldots,j\},$ \ldots
+Finally, we turn to the complexity of $\NumDesign(G)$, the problem of
+computing the number of valid sequences for a given compatibility
+graph $G=(V,E)$. Note that this problem corresponds to the partition
+function problem in a simple base pair model, assigning $0$ to
+canonical and $\infty$ to non-canonical base pairs and setting
+$\beta>0$, so that any valid structure counts as $1$ in the partition
+function. As previously noted~\citep{Flamm2001}, a set of target
+structures admits a valid design iff its compatibility graph is
+bipartite, which can be tested in linear time.  Moreover, without
+$\{\Gb, \Ub\}$ base pairs, the design of any connected component $C$
+of $G$ would be entirely determined by the assignment of a single of
+its nucleotides. The number of valid designs is thus simply
+$4^{\#{\rm CC}(G)}$, where $\#{\rm CC}(G)$ is the \Def{number of
+  connected components}.
+
+The problem becomes much harder when $\{\Gb, \Ub\}$ base pairs are allowed, and we show that valid designs for a set of structures cannot be counted in polynomial time, unless ${\sf\# P}={\sf FP}$.
+This equality would, in particular, imply the classic ${\sf P}={\sf NP}$.
+
+To establish that claim, we consider instances $G=(V_1\cup V_2, E)$ that are connected and bipartite ($E \cap (V_1\times V_2) = E$), noting that hardness over restricted instances implies hardness of the general problem.
+Moreover note that, as we explained before, assigning a nucleotide to a position $u\in V$ constrains the parity ($\{\Ab,\Gb\}$ or $\{\Cb,\Ub\}$) of all positions in the connected component of $u$.
+For this reason, we restrict our attention to the counting of valid designs \emph{up to trivial  symmetry} $(\Ab\leftrightarrow \Cb/\Gb\leftrightarrow \Ub)$, by constraining the positions in $V_{1}$ to $\Ab$ and $\Gb$.
+Let $\Design{G}$ denote the subset of all designs for $G$ under this constraint, noting that $\NumDesign(G) = 2\cdot|\Design{G}|$.
+
+Let $\IS{G}$ denote the set of all independent sets in the connected graph $G$; recall that an \Def{independent set} of $G=(V,E)$ is a subset $V'\subseteq V$ of nodes that are not connected by any edge in $E$.
+
+\begin{proposition}
+  $\Design{G}$ and $\IS{G}$ have equal size.
+\end{proposition}
+%
+\begin{proof}
+Consider the mapping $\Psi: \Design{G} \to \IS{G}$ defined by $ \Psi(f) := \left\{v\in V\mid f(v)\in\{\Ab,\Cb\}\right\}.$
+We show that $\Psi$ is bijective:%
+\begin{itemize}
+\item $\Psi$ is injective, i.e. $\Psi(f)\neq\Psi(f')$ for all $f\neq f'$.
+If $f\neq f'$, then there exists a node $v\in V$ such that $f(v)\neq f'(v)$.
+We discuss only the case $v\in V_1$, where we restricted the nucleotides to $\Ab$ and $\Gb$. Then, $\{f(v),f'(v)\}$ must equal $\{\Ab,\Gb\}$, such that either $v\in\Psi(f)$ or $v\in\Psi(f')$.
+%
+b\item $\Psi$ is surjective, i.e. there is a preimage for each element $I\in \IS{G}$. Define $f\in\Design{G}$ as\begin{displaymath}
+f(v) = \left\{\begin{array}{ll} % @{\qquad}ll
+\Ab{} & \text{if $v\in V_1$ and $v\in I$} \\ %&
+\Cb{} & \text{if $v\in V_2$ and $v\in I$} \\
+\Gb{} & \text{if $v\in V_1$ and $v\not\in I$} \\ %&
+\Ub{} & \text{if $v\in V_2$ and $v\not\in I$}
+\end{array}\right.
+\end{displaymath}
+One easily verifies that $\Psi(f) = I$.
+%
+%for any $(v_1,v_2)\in V_1\times V_2$ by
+%\begin{align*}
+%f(v_1) &= \begin{cases} \Ab & \text{if }v_1\in S\\ \Gb & \text{if }v_1\notin S\end{cases} &
+%\text{\quad and\quad}&&
+% f(v_2) &= \begin{cases} \Cb & \text{if }v_2\in S\\ \Ub & \text{if }v_2\notin S.\end{cases}
+%\end{align*}
+%
+It remains to show that $f$ is a valid design for $G$, i.e. for each $(v, v') \in E$, $\{f(v),f(v')\}\in \B$; please recall that we defined $\B$ as the set of all valid nucleotide pairs.
+Assume there is an edge $(v_1,v_2)\in E$, violating $\{f(v_1),f(v_2)\} \in \B$. Since $G$ is bipartite,
+$v_1\in V_1$ and $v_2\in V_2$, such that $f(v_1)\in\{\Ab,\Gb\}$ and $f(v_2)\in\{\Cb,\Ub\}$.
+This implies that among all possible $\{f(v_1), f(v_2)\}$ only $\{\Ab,\Cb\}$ is not in $\B$, which in turn requires $v_1\in I$ and $v_2\in I$.
+Therefore, since $I$ is an independent set, the edge $(v_1,v_2)\in E$ cannot exist.
+\end{itemize}
+\end{proof}
+
+Now we can build on the connection between the two problems to obtain complexity results for \NumDesign.
+Counting independent sets in bipartite graphs ($\#{\sf BIS}$) is indeed a well-studied \#{\sf P}-hard problem~\citep{Ge2012}, from which we immediately conclude:
+\begin{corollary}
+  \NumDesign is $\#{\sf P}$-hard
+\end{corollary}
+\begin{proof}
+  Note that $\#{\sf BIS}$ is also \#{\sf P}-hard on connected graphs, as the number of independent sets for a disconnected graph $G$ is given by $|\IS{G}|=\prod_{cc\in CC(G)} |\IS{cc}|$.
+Thus any efficient algorithm for $\#{\sf BIS}$ on connected instances provides an efficient algorithm for general graphs.
+
+  Given any algorithm $\mathcal{A}$ that solves \NumDesign over strongly-connected graphs $G$, one can  construct an algorithm $\mathcal{A}'$ that solves $\#{\sf BIS}$: 1) calculate $\NumDesign(G)$ by calling $\mathcal{A}$ 2) return $\NumDesign(G)/2=|\Design{G}| = |\IS{G}|$.
+    Obviously, if $\mathcal{A}$ is a polynomial-time algorithm, so is $\mathcal{A'}$.
+  This implies that $\NumDesign$  is at least as hard as $\#{\sf BIS}$, thus does not admit a polynomial time exact algorithm unless $\#{\sf P}={\sf FP}$.
+\end{proof}
 
@@
 
-\subsection*{\#{\sf P}-hardness of counting valid designs}\label{sec:counting}
-%%% reformulate later
-%We define the following notations: $\real^\infty:=\real\cup\{\infty\}$, $[i,j]:=\{i,\ldots,j\},$ \ldots
-Finally, we turn to the complexity of $\NumDesign(G)$, the problem of
-computing the number of valid sequences for a given compatibility
-graph $G=(V,E)$. Note that this problem corresponds to the partition
-function problem in a simple base pair model, assigning $0$ to
-canonical and $\infty$ to non-canonical base pairs and setting
-$\beta>0$, so that any valid structure counts as $1$ in the partition
-function. As previously noted~\citep{Flamm2001}, a set of target
-structures admits a valid design iff its compatibility graph is
-bipartite, which can be tested in linear time.  Moreover, without
-$\{\Gb, \Ub\}$ base pairs, the design of any connected component $C$
-of $G$ would be entirely determined by the assignment of a single of
-its nucleotides. The number of valid designs is thus simply
-$4^{\#{\rm CC}(G)}$, where $\#{\rm CC}(G)$ is the \Def{number of
-  connected components}.
-
-The problem becomes much harder when $\{\Gb, \Ub\}$ base pairs are allowed, and we show that valid designs for a set of structures cannot be counted in polynomial time, unless ${\sf\# P}={\sf FP}$.
-This equality would, in particular, imply the classic ${\sf P}={\sf NP}$.
-
-To establish that claim, we consider instances $G=(V_1\cup V_2, E)$ that are connected and bipartite ($E \cap (V_1\times V_2) = E$), noting that hardness over restricted instances implies hardness of the general problem.
-Moreover note that, as we explained before, assigning a nucleotide to a position $u\in V$ constrains the parity ($\{\Ab,\Gb\}$ or $\{\Cb,\Ub\}$) of all positions in the connected component of $u$.
-For this reason, we restrict our attention to the counting of valid designs \emph{up to trivial  symmetry} $(\Ab\leftrightarrow \Cb/\Gb\leftrightarrow \Ub)$, by constraining the positions in $V_{1}$ to $\Ab$ and $\Gb$.
-Let $\Design{G}$ denote the subset of all designs for $G$ under this constraint, noting that $\NumDesign(G) = 2\cdot|\Design{G}|$.
-
-Let $\IS{G}$ denote the set of all independent sets in the connected graph $G$; recall that an \Def{independent set} of $G=(V,E)$ is a subset $V'\subseteq V$ of nodes that are not connected by any edge in $E$.
-
-\begin{proposition}
-  $\Design{G}$ and $\IS{G}$ have equal size.
-\end{proposition}
-%
-\begin{proof}
-Consider the mapping $\Psi: \Design{G} \to \IS{G}$ defined by $ \Psi(f) := \left\{v\in V\mid f(v)\in\{\Ab,\Cb\}\right\}.$
-We show that $\Psi$ is bijective:%
-\begin{itemize}
-\item $\Psi$ is injective, i.e. $\Psi(f)\neq\Psi(f')$ for all $f\neq f'$.
-If $f\neq f'$, then there exists a node $v\in V$ such that $f(v)\neq f'(v)$.
-We discuss only the case $v\in V_1$, where we restricted the nucleotides to $\Ab$ and $\Gb$. Then, $\{f(v),f'(v)\}$ must equal $\{\Ab,\Gb\}$, such that either $v\in\Psi(f)$ or $v\in\Psi(f')$.
-%
-b\item $\Psi$ is surjective, i.e. there is a preimage for each element $I\in \IS{G}$. Define $f\in\Design{G}$ as\begin{displaymath}
-f(v) = \left\{\begin{array}{ll} % @{\qquad}ll
-\Ab{} & \text{if $v\in V_1$ and $v\in I$} \\ %&
-\Cb{} & \text{if $v\in V_2$ and $v\in I$} \\
-\Gb{} & \text{if $v\in V_1$ and $v\not\in I$} \\ %&
-\Ub{} & \text{if $v\in V_2$ and $v\not\in I$}
-\end{array}\right.
-\end{displaymath}
-One easily verifies that $\Psi(f) = I$.
-%
-%for any $(v_1,v_2)\in V_1\times V_2$ by
-%\begin{align*}
-%f(v_1) &= \begin{cases} \Ab & \text{if }v_1\in S\\ \Gb & \text{if }v_1\notin S\end{cases} &
-%\text{\quad and\quad}&&
-% f(v_2) &= \begin{cases} \Cb & \text{if }v_2\in S\\ \Ub & \text{if }v_2\notin S.\end{cases}
-%\end{align*}
-%
-It remains to show that $f$ is a valid design for $G$, i.e. for each $(v, v') \in E$, $\{f(v),f(v')\}\in \B$; please recall that we defined $\B$ as the set of all valid nucleotide pairs.
-Assume there is an edge $(v_1,v_2)\in E$, violating $\{f(v_1),f(v_2)\} \in \B$. Since $G$ is bipartite,
-$v_1\in V_1$ and $v_2\in V_2$, such that $f(v_1)\in\{\Ab,\Gb\}$ and $f(v_2)\in\{\Cb,\Ub\}$.
-This implies that among all possible $\{f(v_1), f(v_2)\}$ only $\{\Ab,\Cb\}$ is not in $\B$, which in turn requires $v_1\in I$ and $v_2\in I$.
-Therefore, since $I$ is an independent set, the edge $(v_1,v_2)\in E$ cannot exist.
-\end{itemize}
-\end{proof}
-
-Now we can build on the connection between the two problems to obtain complexity results for \NumDesign.
-Counting independent sets in bipartite graphs ($\#{\sf BIS}$) is indeed a well-studied \#{\sf P}-hard problem~\citep{Ge2012}, from which we immediately conclude:
-\begin{corollary}
-  \NumDesign is $\#{\sf P}$-hard
-\end{corollary}
-\begin{proof}
-  Note that $\#{\sf BIS}$ is also \#{\sf P}-hard on connected graphs, as the number of independent sets for a disconnected graph $G$ is given by $|\IS{G}|=\prod_{cc\in CC(G)} |\IS{cc}|$.
-Thus any efficient algorithm for $\#{\sf BIS}$ on connected instances provides an efficient algorithm for general graphs.
-
-  Given any algorithm $\mathcal{A}$ that solves \NumDesign over strongly-connected graphs $G$, one can  construct an algorithm $\mathcal{A}'$ that solves $\#{\sf BIS}$: 1) calculate $\NumDesign(G)$ by calling $\mathcal{A}$ 2) return $\NumDesign(G)/2=|\Design{G}| = |\IS{G}|$.
-    Obviously, if $\mathcal{A}$ is a polynomial-time algorithm, so is $\mathcal{A'}$.
-  This implies that $\NumDesign$  is at least as hard as $\#{\sf BIS}$, thus does not admit a polynomial time exact algorithm unless $\#{\sf P}={\sf FP}$.
-\end{proof}
-
 \section*{Conclusion}
@@
 
-Motivated by the---here established---hardness of the problem, we
-introduced a general framework, and efficient algorithms to design RNA
-sequences that may adopt multiple target structures with fine-tuned
-properties. Our method combines an FPT stochastic sampling algorithm
+We introduced a novel approach, based on a general framework and efficient algorithms, to design RNA
+sequences, while targeting very specific complex properties; in particular, we describe the targeting of the free energies of multiple target strucutures and the \GCb-content.
+Our method combines an FPT stochastic sampling algorithm
 with multi-dimensional Boltzmann sampling over distributions
@@
 of an extensive multi-target design benchmark, including pseudoknots.
+%
+To substantiate our work additionally, we establish the $\#${\sf P}-hardness of uniform sampling;
+from a complexity-theoretic point of view, this motivates the FPT, tree decomposition-based nature of our method.
 
@@
 generation, including the case of pseudoknotted structures, and studied its use for multi-target RNA design. Moreover,
-our framework is amenable to extensions, to include more complex sequence
+our framework is extensible to include various more complex sequence
 constraints, including mandatory/forbidden motifs at specific
 positions, or anywhere in the designed sequences, by adapting formal language constructs
-introduced by \citet{Zhou2013}. Negative design principles could also
+introduced by \citet{Zhou2013}. In future work, negative design principles could 
 be explicitly supported at the generation stage, for instance by

### 2018-09-13 RNARedPrint.tex
Update P(S|R)~P(R|S) corelation plots in supplement

@@
 Turner RNA energy model
-works well by performing sampling based on less complex RNA energy models, which cause low treewidth.
+works well by performing sampling based on less complex RNA energy models, which keeps the treewidth low.
 This result is highly relevant in practice, since it allows to combine high efficiency with high accuracy and moreover leaves room for extensions of the method.
@@
 to a large set of multi-target RNA design instances from a
-representative benchmark.
+meaningful, representative benchmark.
 

### 2018-09-13 supplement.tex
Update P(S|R)~P(R|S) corelation plots in supplement

@@
 Already previous work on RNA Design, e.g. INFO-RNA and Incarnation, observed
-that---briefly put---negative design benefits from positive design. Here, more
-concretely, we only \emph{demonstrate} this effect (for small examples), where
+that---put briefly---negative design benefits from positive design. Here, more
+concretely, we simply \emph{demonstrate} this effect (for small examples), where
 we consider \Def{negative design} as  
@@
 
-For a double-stem structure, we produce a uniform and a weighted Boltzmann
+For each of two small examples, a single-stem and a double-stem structure, we produce a uniform and a weighted Boltzmann
 sampling, each of 10000 sequenes. For each sampled sequence $S$, we determine
@@
 %
-Consequently, Figure~\ref{fig:negforpos} ultimately illustrates the dependency
+Consequently, Figure~\ref{fig:negforpos} ultimately illustrates the dependencies
 between $Pr(S|R)$ and $Pr(R|S)$ for uniform samples and low energy samples
-(where the latter can be efficiently generated due to our Boltzmann sampling
-strategy).
+(where those samples can be efficiently generated due to our Boltzmann sampling
+strategy). For our examples, we observe (Fig.~\ref{fig:negforpos}) that Boltzmann sampling with weight 500 (green), generates good negative designs 
+with substantially higher probability than uniform sampling (blue).
 
@@
 \begin{figure}
-\textbf{A}~\includegraphicstop[width=0.48\textwidth]{Figs/Plots/doublestem_weight1}%
-\textbf{B}~\includegraphicstop[width=0.48\textwidth]{Figs/Plots/doublestem_weight500}
-\caption{Comparision of (A) Uniform sampling to (B) Boltzmann sampling with
-weight $\pi=500$. For given structure $R$, we plot the energies $E(S,R)$ of
+\textbf{A}~\includegraphicstop[width=0.48\textwidth]{Figs/Plots/stem_weight1-500}%
+\textbf{B}~\includegraphicstop[width=0.48\textwidth]{Figs/Plots/doublestem_weight1-500}%
+\caption{Comparision of uniform sampling and Boltzmann sampling with
+weight $\pi=500$. We show results for two different small structures: (A) single stem structure (B) double stem structure. For each respective structure $R$, we plot the energies $E(S,R)$ of
 sampled sequences against the probabilities $Pr(R|S)$; the former is controlled

### 2018-09-13 supplement.tex
Add GC distribution plots

@@
 
+Fig.~\ref{fig:gc-content} shows the distributions of the GC content of the samples, while targeting different combinations of energies in this experiment.
+For completeness, we also report corresponding results for the analogous experiment from the main text.
+We observe that in addition to the target energies, also the feature $\GCb\%$ is narrowly defined by the multi-dimensional Boltzmann sampling strategy.
+In comparison, for uniform sampling and Boltzmann sampling at high weight $e^\beta$, where the \GCb\% is not controlled, we obtain much broader distributions.
+
 \begin{figure*}[t]
@@
 \end{figure*}
+
+\begin{figure}
+  \begin{center}
+    {\sf \bfseries A}\includegraphicstop[width=0.45\textwidth]{Figs/Plots/PKB00211_PKB00239_0_GC_content}
+    ~~~~~{\sf \bfseries B}
+    \includegraphicstop[width=0.45\textwidth]{Figs/Plots/supp_3str_GC_content}
+  \end{center}
+  \caption{Distribution of the \GCb-content \GCb\% in the experiments of (A) Fig.~\ref{fig:energydist-pk} of the main text and (B) Suppl.~Fig.~\ref{fig:energydist}.}
+  \label{fig:gc-content}
+\end{figure}
 

### 2018-09-14 RNARedPrint.tex
Updates from ShareLaTeX

@@
 \newtheorem{proposition}[theorem]{Proposition}
-
 \newtheorem{observation}[theorem]{Observation}
@@
   \parttitle{Results} We devise a novel approach to a central ingredient of most in-silico
-  design methods: the sampling of sequences for multi-target RNA design. For this task, we present the efficient (fixed-parameter tractable) approach \ourprog{}.
+  design methods: the generation of sequences with multiple structural targets. For this task, we present the efficient (fixed-parameter tractable) approach \ourprog{}.
   % 
@@
 thermodynamics of RNA secondary structure is well understood and its prediction is
-computationally tractable~\citep{McCaskill1990}. Thus,  structure can serve as effective
-proxy for rational design approaches, ultimately targeting catalytic or regulatory functions~\citep{Zhang2013}.
+computationally tractable~\citep{McCaskill1990}. Thus, in rational design approaches, structure can serve as effective
+proxy for, the ultimately targeted, catalytic or regulatory functions~\citep{Zhang2013}.
 
@@
 \parHead{Motivation.}
-While uniform sampling in \Software{RNAdesign}/\RNAblueprint{} turned out to be very successful,
+While uniform sampling in \Software{RNAdesign}/\RNAblueprint{} turned out to be highly beneficial for general RNA design,
 the work on these tools gave no definitive answer on the \emph{complexity of uniform sampling for multi-target design}. While time and space requirements grow exponential in the general case,
@@
 .
-We show that these algorithms are time and space-efficient depending on a specific property of the tree decomposition, namely the treewidth; in short, our algorithms are fixed-parameter tractable (FPT).%
+We show that these algorithms are time and space-efficient depending on a specific property of the tree decomposition, namely the treewidth; in short, our algorithms are fixed-parameter tractable (FPT).
 % Subsection~\ref{sec:complexity}
@@
 Eventually this strategy allows generating biologically relevant
-multi-target designs, as we demonstrate in our application 
-to a large set of multi-target RNA design instances from a
-meaningful, representative benchmark.
+multi-target designs, as we demonstrate by performing negative multi-target RNA design for a comprehensive, 
+meaningful benchmark set~\citep{Taneda2015}.
 
@@
 \label{sec:problem-statement}
-The main computational problem addressed in this work is to generate sequences over the alphabet $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}$ that achieving prescribed free-energies and $\GCb\%$.
-Despite their apparent diversity, such criteria can be homogeneously abstracted as constraints on the value of \Def{features}, each defined as sums over atomic feature contributions depending on subsets of sequence positions.
+The main computational problem addressed in this work is to generate sequences over the alphabet $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}$ with prescribed energies, for each single target structure, and $\GCb\%$.
+Remarkably, such criteria can be homogeneously abstracted as constraints on the value of \Def{features}, each defined as sums over atomic feature contributions depending on limited subsets of sequence positions.
 
@@
 We base our strategy for targeting the $k$ features $F_1,\dots,F_k$ on generating samples from a Boltzmann distribution parameterized by weights $\pi_1,\dots,\pi_k$, one for each feature.
-These weights essentially allows us to shift the sample distributions towards specific targets.
+These weights essentially allow us to shift the sample distributions towards specific targets.
 Sampling from this distribution requires to compute corresponding partition functions, such that we can draw sequences with probabilities proportional to their Boltzmann weight:
@@
 The workhorse of our approach is the efficient (more precisely, fixed-parameter tractable) computation of feature-dependent partition functions over sequences.
-Given are the features $F_1,\dots,F_k$, which encode the energies functions for the fixed structures and the function for the \GCb\% all in terms of simple atomic energy contributions.
+Given are the features $F_1,\dots,F_k$, which encode the energies functions for the fixed structures and the function for the \GCb\%, all in terms of simple atomic energy contributions.
 For weights $\pi_1,\dots,\pi_k$, one wants to compute the partition function
@@
 \paragraph{Representing \GCb\%-content and energies as features.}
-Formally, we define a \Def{feature $F$} as function on sequences, whose value is obtained by summing over an associated set of \Def{atomic feature contributions} $f_1, f_2\ldots$, briefly called \emph{contributions}, that each are associated with a restricted set of positions, its \emph{dependencies $\dep(f)$}, such that
+Formally, we define a \Def{feature $F$} as function on sequences, whose value is obtained by summing over an associated set of \Def{atomic feature contributions}, briefly called \emph{contributions}.
+Each feature $f$ is associated with a restricted set of positions, its \emph{dependencies $\dep(f)$}, such that
 $$
-F(S) =\!\!\!\!\!\!\sum_{\substack{\text{contribution $f_i$ of $F$}\\\text{with $\dep(f)=\{x_1,\ldots,x_p\}$}}}\!\!\!\!\!\! f_i(\{x_1\mapsto S_{x_1},\ldots,x_p\mapsto S_{x_p}\}).
+F(S) =\!\!\!\!\!\!\sum_{\substack{\text{$f$: $f$ is contribution of $F$,}\\\text{ $\dep(f)=\{x_1,\ldots,x_p\}$}}}\!\!\!\!\!\! f(\{x_1\mapsto S_{x_1},\ldots,x_p\mapsto S_{x_p}\}).
 $$
 
-As simple example, the $\GCb\%$ feature is expressed by $n$ contributions $f^{\GCb}_i$ where $f^{\GCb}_i(\{i\mapsto S_i\})$ equals $-1$ if $S_i=\Gb$ or $\Cb$, and $0$ otherwise.
+Thus, the $\GCb\%$ feature is expressed by $n$ contributions $f^{\GCb}_i$ where $f^{\GCb}_i(\{i\mapsto S_i\})$ equals $-1$ if $S_i=\Gb$ or $\Cb$, and $0$ otherwise $(i=1,\dots,n)$.
 
@@
 
-In a simple \Def{base pair energy model}, the free-energy of a target structure $R$ is expressed as feature with contributions $f^R_{ij}(\{i\mapsto x, j\mapsto y\})=\EbpSym(x,y)$ for each $(i,j)\in R,$ such that $F^R(S)=\EbpSym(S,R)$.
+In a simple \Def{base pair energy model}, the energy of a target structure $R$ is expressed as feature with contributions $f^R_{ij}(\{i\mapsto x, j\mapsto y\})=\EbpSym(x,y)$ for each $(i,j)\in R,$ such that $F^R(S)=\EbpSym(S,R)$.
 
 Note that for \GCb\% as well as for the base pair energy features, the contributions depend on the nucleotides at only a small, constantly bounded number of sequence positions (one and two positions, respectively).
-This is important for the feasibility of our approach, which is designed to profit from contributions that each depend on only few sequence positions.
+This is important for the efficiency of our approach, which is designed to profit from contributions that each depend on only few sequence positions.
 We emphasize that nevertheless, the same framework can capture more expressive RNA energy models like the Turner energy model~\citep{Turner2009}, since such models generally define a function $E$ that maps a pair of a sequence $S$ and a structure $R$ to a real value, the \Def{energy} $E(S,R)$.
@@
 
-To formalize our algorithms, which iteratively merges evaluations of partial solutions, we extend the idea of atomic feature contributions, which are evaluated at sets of the form $\{x_1\mapsto v_1, \dots, x_p\mapsto v_d\}$.
+To formalize our algorithms, which iteratively merge evaluations of partial solutions, we extend the idea of atomic feature contributions, which are evaluated at sets of the form $\{x_1\mapsto v_1, \dots, x_d\mapsto v_p\}$.
 Let us call the latter object \Def{partial sequence}; they are handy to specify partial knowledge on the entire sequence at some point of the algorithm.
-Easily, we can extend the definition of contributions $f$ to sets $\{x_1\mapsto v_1, \dots, x_p\mapsto v_d\}$, where $\{x_1\dots x_d\}$ is any super-set of $\dep(f)$ by ignoring the superfluous assignments $x\mapsto v$, where $x\not\in\dep(f)$.
+Easily, we can extend the definition of contributions $f$ to sets $\{x_1\mapsto v_1, \dots, x_p\mapsto v_p\}$, where $\{x_1\dots x_p\}$ is any super-set of $\dep(f)$ by ignoring the superfluous assignments $x\mapsto v$, where $x\not\in\dep(f)$.
 
@@
 
-\newcommand{\weighttransform}[2]{{#1}_{||{#2}}}
+%\newcommand{\weighttransform}[2]{{#1}_{||{#2}}}
 
 For algorithmic treatment, it is convenient to encode the weight $\pi$ of each feature in its contributions, such that all contributions can be handled uniformly.
-For this purpose, we define the \Def{transformed contribution}
-$\weighttransform{f}{\pi}(S) := \ln(\pi)f(S),$
-such that $exp(-\weighttransform{f}{\pi}(S))=\pi^{-f(S)}.$ 
-%
-
+This transformation works by multiplying all contributions with $ln(\pi)$, where $\pi$ is the weight of the corresponding feature, since then
+$exp(-\ln(\pi)f(S))=\pi^{-f(S)}.$
+
+%For this purpose, we define the \Def{transformed contribution} 
+%$\weighttransform{f}{\pi}(S) := \ln(\pi)f(S),$
+%such that $exp(-\weighttransform{f}{\pi}(S))=\pi^{-f(S)}.$ 
+%
 Let us now specify the concrete set $\F$ of contributions that we use
@@
 \begin{itemize}
-\item the transformed contributions $\weighttransform{f^{\GCb}_i}{\pi_0}$ for the ~$\GCb\%$ feature ($i=1,\dots,n$)
-\item the transformed contributions $\weighttransform{f^{R_\ell}_{ij}}{\pi_\ell}$for each structure $R_\ell\in\R$ and $(i,j)\in R_\ell.$
+\item the transformed contributions $\ln(\pi_0)f^{\GCb}_i$ for the ~$\GCb\%$ feature ($i=1,\dots,n$)
+\item the transformed contributions $\ln(\pi_\ell)f^{R_\ell}_{ij}$for each structure $R_\ell\in\R$ and $(i,j)\in R_\ell.$
 \end{itemize}
@@
 We compute the partition function (as specified by $\F$) by dynamic programming based
-on a tree decomposition of the $\F$'s dependency graph.
+on a tree decomposition of $\F$'s dependency graph.
 Note that analogous algorithms could be constructed to count valid sequences or minimize energy.

### 2018-09-14 RNARedPrint.tex
Changes from overleaf v2

@@
 
-Our algorithms are formulated to process a \Def{cluster tree}, which is a tuple $(T,\chi,\phi)$, where
+Our algorithms are formulated to process a \Def{cluster tree} of $\F$, which is a tuple $(T,\chi,\phi)$, where
 $(T,\chi)$ is a tree decomposition of $G_\F$, and $\phi(v)$ represents
@@
 messages along the directed edges $u\to v$ (which point from 
-child $u$ to its parent $v$). Each message $m$ is a function that depends
+child $u$ to its parent $v$). Each message $m$ is  that depends
 on the positions $\dep(m)\subseteq \X$ and yields a partition function

### 2018-09-14 RNARedPrint.tex
edits on overleaf

@@
 messages along the directed edges $u\to v$ (which point from 
-child $u$ to its parent $v$). Each message $m$ is  that depends
-on the positions $\dep(m)\subseteq \X$ and yields a partition function
+child $u$ to its parent $v$). Each message $m$ has the form of a contribution, i.e. it takes a partial sequence, depends
+on the positions $\dep(m)\subseteq \X$, and yields a partition function
 in $\real$. The message from $u$ to $v$ represents the
@@
 % The treewidth $w$ of a cluster tree $(T,\chi,\phi)$, $T=(V,E)$, is $\max_{v\in V} |\chi(v)| - 1$, i.e.~the maximum number of positions in any of its clusters minus 1.
-
-Note that in the following complexity analysis, we omit time and space for computing the
-tree decomposition itself, since we observed that the computation time
-of tree decomposition (\Software{GreedyFillIn}, implemented in
-\Software{LibTW}~by \citet{Dijk2006}) for multi-target sampling is
-negligible compared to Alg.~\ref{alg:pf}
-(Supp. Mat.~\ref{appsec:treedecomp} and
-\ref{appsec:dependency-cliques}).
-
 Let $s$ denote the maximum size of any separator set $\separator{u}{v}$ and $D$ denote the maximum size of
@@
 
+Note that in this complexity analysis, we do not include time and space for computing the
+tree decomposition itself, since we observed that the computation time
+of tree decomposition (\Software{GreedyFillIn}, implemented in
+\Software{LibTW}~by \citet{Dijk2006}) for multi-target sampling is
+negligible compared to Alg.~\ref{alg:pf}
+(Supp. Mat.~\ref{appsec:treedecomp} and
+\ref{appsec:dependency-cliques}).
+
 \subsection*{Design for expressive energy models}\label{sec:energy_models}
@@
 \paragraph{Approximating Turner Energy using Simpler Energy Models}
-To capture the realistic Turner model $\EnergyTurner$ while keeping
-a reasonable computational cost, we exploit a tight correlation between
+To capture the realistic Turner model $\EnergyTurner$ more efficiently, we exploit the tight correlation between
 $\EnergyTurner$ and the fitted stacking model $\EnergyStacking$
@@
 \subsection*{Extension to multidimensional Boltzmann sampling}\label{sec:multiBoltzmann}
-The flexibility of our framework supports the advanced sampling technique "multidimensional Boltzmann sampling"~\citep{Bodini2010}, which allows to (probabilistically) enforce additional, complex properties of the samples.
+The flexibility of our framework supports the advanced sampling technique "multidimensional Boltzmann sampling"~\citep{Bodini2010}, which allows (probabilistically) enforcing additional, complex properties of the samples.
 This technique was previously used to control \GCb\%~\citep{Waldispuehl2011,Reinharz2013} and dinucleotide content~\citep{Zhang2013} of sampled RNA sequences; here, in addition to controlling \GCb\% (our feature $F_0$) we use it to target the free energies $(\TargetE_1,\ldots,\TargetE_k)$ of the single targets (features $F_1,\dots,F_k$).
@@
 Mat.~\ref{sec:weight_derivatives}).
-More generally, computing a suitable $\pmb{\pi}$ can be restated as a convex optimization problem, and be efficiently solved using a wide array of methods~\citep{Denise2010,Bendkowski2017}.
+More generally, computing a suitable parameter vector $\pmb{\pi}$ can be restated as a convex optimization problem, and be efficiently solved using a wide array of methods~\citep{Denise2010,Bendkowski2017}.
 In practice, we use a simple heuristics which starts from an initial weight vector $\pmb{\pi}^{[0]}:=(e^\beta,\dots,e^\beta)$ for $\beta=1/(RT)$, T=$37^\circ$, and gas constant $R$.
@@
 i) \emph{Enacting an exact control over the energies would  be technically hard and costly.} Indeed, controlling the energies through dynamic programming would require explicit convolution products, generalizing~\citep{Cupal1996}, inducing additional $\Theta(n^{2k})$ time and $\Theta(n^k)$ space overheads;
-ii) \emph{Induced distributions are typically concentrated.} Intuitively, unless sequences are fully constrained individual energy terms are independent enough so that their sum is concentrated around its mean -- the targeted energy (cf Fig.~\ref{fig:energydist-pk}).
+ii) \emph{Induced distributions are typically concentrated.} Intuitively, unless sequences are fully constrained individual energy terms are independent enough so that their sum is concentrated around its mean -- the targeted energy (cf.{} Fig.~\ref{fig:energydist-pk}).
 %Indeed, as shown in Fig.~\ref{fig:shifting_mean}, empirical joint distributions in the energies have good fits towards Normal multidimensional distributions with linear (co-)variances.
@@
 graph $G=(V,E)$. Note that this problem corresponds to the partition
-function problem in a simple base pair model, assigning $0$ to
-canonical and $\infty$ to non-canonical base pairs and setting
-$\beta>0$, so that any valid structure counts as $1$ in the partition
-function. As previously noted~\citep{Flamm2001}, a set of target
+function problem with weights $\pi_\ell=1$ for each target structure, so that each valid structure contributes $1$ to the partition
+function; each invalid structure, 0.
+
+As previously noted~\citep{Flamm2001}, a set of target
 structures admits a valid design iff its compatibility graph is
@@
 \subsection*{Generating high-quality seeds for further optimization}
-We empirically evaluated the capacity of \ourprog{} to generate seed sequences for multiple (pseudoknotted) target structures, possibly followed by subsequent local optimizations.
+We empirically evaluated \ourprog{} for generating seed sequences targeting multiple (pseudoknotted) structures, possibly followed by subsequent local optimizations.
 As a baseline for comparison, we considered \RNAblueprint~\citep{Hammer2017}, the current leading tool for multiple design.
@@
 previously applied ad-hoc sampling and even efficient uniform sampling
-strategies. We have shown the practicality of such controlled sequence
+strategies. We demonstrated the practical feasibility of such controlled sequence
 generation, including the case of pseudoknotted structures, and studied its use for multi-target RNA design. Moreover,

### 2018-09-14 RNARedPrint.tex
supplement: edits for consistency etc, move tables to end

--- 2018-09-14 RNARedPrint.tex
@@
 \newtheorem{proposition}[theorem]{Proposition}
+
 \newtheorem{observation}[theorem]{Observation}
@@
   \parttitle{Results} We devise a novel approach to a central ingredient of most in-silico
-  design methods: the generation of sequences with multiple structural targets. For this task, we present the efficient (fixed-parameter tractable) approach \ourprog{}.
+  design methods: the sampling of sequences for multi-target RNA design. For this task, we present the efficient (fixed-parameter tractable) approach \ourprog{}.
   % 
@@
 thermodynamics of RNA secondary structure is well understood and its prediction is
-computationally tractable~\citep{McCaskill1990}. Thus, in rational design approaches, structure can serve as effective
-proxy for, the ultimately targeted, catalytic or regulatory functions~\citep{Zhang2013}.
+computationally tractable~\citep{McCaskill1990}. Thus,  structure can serve as effective
+proxy for rational design approaches, ultimately targeting catalytic or regulatory functions~\citep{Zhang2013}.
 
@@
 \GCb\%,
-more recent approaches like \Software{antaRNA}~\citep{Kleinkauf2015} and \Software{IncaRNAtion}~\citep{Reinharz2013} explicitly control \GCb\%; the applying adaptive sampling.
+more recent approaches like \Software{antaRNA}~\citep{Kleinkauf2015} and \Software{IncaRNAtion}~\citep{Reinharz2013} explicitly control \GCb\%; the latter applying adaptive sampling.
 
@@
 \parHead{Motivation.}
-While uniform sampling in \Software{RNAdesign}/\RNAblueprint{} turned out to be highly beneficial for general RNA design,
+While uniform sampling in \Software{RNAdesign}/\RNAblueprint{} turned out to be very successful,
 the work on these tools gave no definitive answer on the \emph{complexity of uniform sampling for multi-target design}. While time and space requirements grow exponential in the general case,
@@
 .
-We show that these algorithms are time and space-efficient depending on a specific property of the tree decomposition, namely the treewidth; in short, our algorithms are fixed-parameter tractable (FPT).
+We show that these algorithms are time and space-efficient depending on a specific property of the tree decomposition, namely the treewidth; in short, our algorithms are fixed-parameter tractable (FPT).%
 % Subsection~\ref{sec:complexity}
@@
 Eventually this strategy allows generating biologically relevant
-multi-target designs, as we demonstrate by performing negative multi-target RNA design for a comprehensive, 
-meaningful benchmark set~\citep{Taneda2015}.
+multi-target designs, as we demonstrate in our application 
+to a large set of multi-target RNA design instances from a
+meaningful, representative benchmark.
 
@@
 \label{sec:problem-statement}
-The main computational problem addressed in this work is to generate sequences over the alphabet $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}$ with prescribed energies, for each single target structure, and $\GCb\%$.
-Remarkably, such criteria can be homogeneously abstracted as constraints on the value of \Def{features}, each defined as sums over atomic feature contributions depending on limited subsets of sequence positions.
+The main computational problem addressed in this work is to generate sequences over the alphabet $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}$ that achieving prescribed free-energies and $\GCb\%$.
+Despite their apparent diversity, such criteria can be homogeneously abstracted as constraints on the value of \Def{features}, each defined as sums over atomic feature contributions depending on subsets of sequence positions.
 
@@
 We base our strategy for targeting the $k$ features $F_1,\dots,F_k$ on generating samples from a Boltzmann distribution parameterized by weights $\pi_1,\dots,\pi_k$, one for each feature.
-These weights essentially allow us to shift the sample distributions towards specific targets.
+These weights essentially allows us to shift the sample distributions towards specific targets.
 Sampling from this distribution requires to compute corresponding partition functions, such that we can draw sequences with probabilities proportional to their Boltzmann weight:
@@
 The workhorse of our approach is the efficient (more precisely, fixed-parameter tractable) computation of feature-dependent partition functions over sequences.
-Given are the features $F_1,\dots,F_k$, which encode the energies functions for the fixed structures and the function for the \GCb\%, all in terms of simple atomic energy contributions.
+Given are the features $F_1,\dots,F_k$, which encode the energies functions for the fixed structures and the function for the \GCb\% all in terms of simple atomic energy contributions.
 For weights $\pi_1,\dots,\pi_k$, one wants to compute the partition function
@@
 \paragraph{Representing \GCb\%-content and energies as features.}
-Formally, we define a \Def{feature $F$} as function on sequences, whose value is obtained by summing over an associated set of \Def{atomic feature contributions}, briefly called \emph{contributions}.
-Each feature $f$ is associated with a restricted set of positions, its \emph{dependencies $\dep(f)$}, such that
+Formally, we define a \Def{feature $F$} as function on sequences, whose value is obtained by summing over an associated set of \Def{atomic feature contributions} $f_1, f_2\ldots$, briefly called \emph{contributions}, that each are associated with a restricted set of positions, its \emph{dependencies $\dep(f)$}, such that
 $$
-F(S) =\!\!\!\!\!\!\sum_{\substack{\text{$f$: $f$ is contribution of $F$,}\\\text{ $\dep(f)=\{x_1,\ldots,x_p\}$}}}\!\!\!\!\!\! f(\{x_1\mapsto S_{x_1},\ldots,x_p\mapsto S_{x_p}\}).
+F(S) =\!\!\!\!\!\!\sum_{\substack{\text{contribution $f_i$ of $F$}\\\text{with $\dep(f)=\{x_1,\ldots,x_p\}$}}}\!\!\!\!\!\! f_i(\{x_1\mapsto S_{x_1},\ldots,x_p\mapsto S_{x_p}\}).
 $$
 
-Thus, the $\GCb\%$ feature is expressed by $n$ contributions $f^{\GCb}_i$ where $f^{\GCb}_i(\{i\mapsto S_i\})$ equals $-1$ if $S_i=\Gb$ or $\Cb$, and $0$ otherwise $(i=1,\dots,n)$.
+As simple example, the $\GCb\%$ feature is expressed by $n$ contributions $f^{\GCb}_i$ where $f^{\GCb}_i(\{i\mapsto S_i\})$ equals $-1$ if $S_i=\Gb$ or $\Cb$, and $0$ otherwise.
 
@@
 
-In a simple \Def{base pair energy model}, the energy of a target structure $R$ is expressed as feature with contributions $f^R_{ij}(\{i\mapsto x, j\mapsto y\})=\EbpSym(x,y)$ for each $(i,j)\in R,$ such that $F^R(S)=\EbpSym(S,R)$.
+In a simple \Def{base pair energy model}, the free-energy of a target structure $R$ is expressed as feature with contributions $f^R_{ij}(\{i\mapsto x, j\mapsto y\})=\EbpSym(x,y)$ for each $(i,j)\in R,$ such that $F^R(S)=\EbpSym(S,R)$.
 
 Note that for \GCb\% as well as for the base pair energy features, the contributions depend on the nucleotides at only a small, constantly bounded number of sequence positions (one and two positions, respectively).
-This is important for the efficiency of our approach, which is designed to profit from contributions that each depend on only few sequence positions.
+This is important for the feasibility of our approach, which is designed to profit from contributions that each depend on only few sequence positions.
 We emphasize that nevertheless, the same framework can capture more expressive RNA energy models like the Turner energy model~\citep{Turner2009}, since such models generally define a function $E$ that maps a pair of a sequence $S$ and a structure $R$ to a real value, the \Def{energy} $E(S,R)$.
@@
 
-To formalize our algorithms, which iteratively merge evaluations of partial solutions, we extend the idea of atomic feature contributions, which are evaluated at sets of the form $\{x_1\mapsto v_1, \dots, x_d\mapsto v_p\}$.
+To formalize our algorithms, which iteratively merges evaluations of partial solutions, we extend the idea of atomic feature contributions, which are evaluated at sets of the form $\{x_1\mapsto v_1, \dots, x_p\mapsto v_d\}$.
 Let us call the latter object \Def{partial sequence}; they are handy to specify partial knowledge on the entire sequence at some point of the algorithm.
-Easily, we can extend the definition of contributions $f$ to sets $\{x_1\mapsto v_1, \dots, x_p\mapsto v_p\}$, where $\{x_1\dots x_p\}$ is any super-set of $\dep(f)$ by ignoring the superfluous assignments $x\mapsto v$, where $x\not\in\dep(f)$.
+Easily, we can extend the definition of contributions $f$ to sets $\{x_1\mapsto v_1, \dots, x_p\mapsto v_d\}$, where $\{x_1\dots x_d\}$ is any super-set of $\dep(f)$ by ignoring the superfluous assignments $x\mapsto v$, where $x\not\in\dep(f)$.
 
@@
 
-%\newcommand{\weighttransform}[2]{{#1}_{||{#2}}}
+\newcommand{\weighttransform}[2]{{#1}_{||{#2}}}
 
 For algorithmic treatment, it is convenient to encode the weight $\pi$ of each feature in its contributions, such that all contributions can be handled uniformly.
-This transformation works by multiplying all contributions with $ln(\pi)$, where $\pi$ is the weight of the corresponding feature, since then
-$exp(-\ln(\pi)f(S))=\pi^{-f(S)}.$
-
-%For this purpose, we define the \Def{transformed contribution} 
-%$\weighttransform{f}{\pi}(S) := \ln(\pi)f(S),$
-%such that $exp(-\weighttransform{f}{\pi}(S))=\pi^{-f(S)}.$ 
-%
+For this purpose, we define the \Def{transformed contribution}
+$\weighttransform{f}{\pi}(S) := \ln(\pi)f(S),$
+such that $exp(-\weighttransform{f}{\pi}(S))=\pi^{-f(S)}.$ 
+%
+
 Let us now specify the concrete set $\F$ of contributions that we use
@@
 \begin{itemize}
-\item the transformed contributions $\ln(\pi_0)f^{\GCb}_i$ for the ~$\GCb\%$ feature ($i=1,\dots,n$)
-\item the transformed contributions $\ln(\pi_\ell)f^{R_\ell}_{ij}$for each structure $R_\ell\in\R$ and $(i,j)\in R_\ell.$
+\item the transformed contributions $\weighttransform{f^{\GCb}_i}{\pi_0}$ for the ~$\GCb\%$ feature ($i=1,\dots,n$)
+\item the transformed contributions $\weighttransform{f^{R_\ell}_{ij}}{\pi_\ell}$for each structure $R_\ell\in\R$ and $(i,j)\in R_\ell.$
 \end{itemize}
@@
 We compute the partition function (as specified by $\F$) by dynamic programming based
-on a tree decomposition of $\F$'s dependency graph.
+on a tree decomposition of the $\F$'s dependency graph.
 Note that analogous algorithms could be constructed to count valid sequences or minimize energy.

### 2018-09-14 supplement.tex
Update Suppl Fig 4
supplement: edits for consistency etc, move tables to end

@@
 %%% macros for notation in DP framework
-\newcommand{\network}{\mathcal{N}}
 \newcommand{\val}{\bar S} % valuation aka assignment
@@
 
-\newcommand{\evalfor}[2]{#1\llbracket{}#2\rrbracket{}}
-\newcommand{\substitute}[2]{#1\!\oplus\!#2}
+\newcommand{\evalfor}[2]{#1(#2)}
+\newcommand{\substitute}[2]{#1\cup#2}
 
@@
 
-\section{Expressing higher arity functions as cliques in the dependency graph}
+\section{Tree decomposition with hyper-dependencies}
 \label{appsec:dependency-cliques}
 
-Our implementation relies on external tools for the tree decomposition.
-Therefore it is not trivial that the weight functions (i.e.
-terms of the energy function) can be captured within the tree decompositions returned by a specific tool.
-In other words, it is not immediately clear how one can construct a cluster tree from an arbitrary tree decomposition for a given network.
+Our implementation relies on external tools for the tree decomposition, which do not directly support tree decomposition of hyper-graphs. Therefore it is not immediately obvious how 
+to obtain a suitable tree decomposition, such that even multi-ary contributions are respected and a cluster tree can be generated.
 
-Crucially, one can show that, as long as the arguments $\dep(f)$ of a function $f$ form a clique within the network $\network$, then there exists at least one node in the tree decomposition that contains $\dep(f)$, thus allowing the evaluation of $f$.
+Crucially, one can show that, as long as certain vertices of a graph $G$ from a clique, there exists at least one node in the tree decomposition of $G$ that contains all such vertices.
 %
-While this result is certainly part of folklore, we emphasize the importance for the generality of this work and present a brief direct proof for convenience.
+This allows us to generate suitable tree decompositions, which respect the hyper-edges of our dependency graph, by expanding the hyper-edges to cliques, and then calling the tree-decomposer on the resulting (non-hyper-)graph.
+
+%
+While the result is certainly part of folklore, we emphasize the importance for the generality of this work and present a brief direct proof for convenience.
 
@@
        \evalfor{\Message{u}{v}}{\val} = \sum_{\val'\in\partseqs(\chi(T_r(u))-\separator{u}{v})} \quad
-   \prod_{f\in\phi(T_r(u))} \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}}),\label{eq:pfalgo-correct}
+       \prod_{f\in\phi(T_r(u))} \exp(-\evalfor{f}{\substitute{\val'}{\val}}),\label{eq:pfalgo-correct}
      \end{equation}
@@
       \evalfor{\Message{u}{v}}{\val} := \sum_{\val'\in\partseqs(\difference{u}{v})}\quad
-  \prod_{f\in \phi(u) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}}) \prod_{(w\to{}u) \in T} \evalfor{\Message{w}{u}}{\substitute{\val'}{\val}}.\label{eq:messages}
+      \prod_{f\in \phi(u) } \exp(-\evalfor{f}{\substitute{\val'}{\val}}) \prod_{(w\to{}u) \in T} \evalfor{\Message{w}{u}}{\substitute{\val'}{\val}}.\label{eq:messages}
     \end{equation}
@@
       & \sum_{\val'\in\partseqs(\difference{u}{v})}\quad
-    \prod_{f\in \phi(u) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}})
+        \prod_{f\in \phi(u) } \exp(-\evalfor{f}{\substitute{\val'}{\val}})
         \prod_{(w\to{}u) \in T} \evalfor{\Message{w}{u}}{\substitute{\val'}{\val}}\\
       & = \sum_{\val'\in\partseqs(\chi(u)-\separator{u}{v})}\quad
-    \prod_{f\in \phi(u) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}})
+        \prod_{f\in \phi(u) } \exp(-\evalfor{f}{\substitute{\val'}{\val}})
         \prod_{i=1}^q \evalfor{\Message{w_i}{u}}{\substitute{\val'}{\val}}\\
@@
         \sum_{\val'\in\partseqs(\chi(u)-\separator{u}{v})}\quad
-    \prod_{f\in \phi(u) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}})
+        \prod_{f\in \phi(u) } \exp(-\evalfor{f}{\substitute{\val'}{\val}})
         \prod_{i=1}^q \sum_{\val''\in\partseqs(\chi(T_r(w_i))-\separator{w_i}{u})} \quad
-    \prod_{f\in\phi(T_r(w_i))} \exp(-\beta \evalfor{f}{\substitute{\val''}{\substitute{\val'}{\val}}}) \\
+        \prod_{f\in\phi(T_r(w_i))} \exp(-\evalfor{f}{\substitute{\val''}{\substitute{\val'}{\val}}}) \\
     & =_{*}
@@
       \sum_{\val''\in\partseqs(\bigcup_{i=1}^q\chi(T_r(w_i))-\separator{w_i}{u})} \quad
-  \prod_{f\in \phi(u) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}})
-  \prod_{f\in\phi(T_r(w_i))} \exp(-\beta \evalfor{f}{\substitute{\val''}{\val'} | \val}) \\
+      \prod_{f\in \phi(u) } \exp(-\evalfor{f}{\substitute{\val'}{\val}})
+      \prod_{f\in\phi(T_r(w_i))} \exp(-\evalfor{f}{\substitute{\val''}{\val'} \cup \val}) \\
       & =_{(**)} \sum_{\val'\in\partseqs(\chi(T_r(u))-\separator{u}{v})}\quad
-    \prod_{f\in \phi(T_r(u)) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}})\\
+        \prod_{f\in \phi(T_r(u)) } \exp(-\evalfor{f}{\substitute{\val'}{\val}})\\
     \end{align*}
@@
       \label{prop:general-complexity}
-Given a cluster tree $(T,\chi,\phi)$, $T=(V,E)$ of the RNA design network $\network=(\X,\F)$ with treewidth $w$ and maximum separator size $s$, computing the partition function by Algorithm~\ref{alg:pf} takes $O((|F|+|V|)\cdot 4^{w+1})$ time and $O(|V| 4^s)$ space (for storing all messages).
+    Given a cluster tree $(T,\chi,\phi)$, $T=(V,E)$ of the set of contributions $\F{}$ with treewidth $w$ and maximum separator size $s$, computing the partition function by Algorithm~\ref{alg:pf} takes $O((|F|+|V|)\cdot 4^{w+1})$ time and $O(|V| 4^s)$ space (for storing all messages).
 The sampling step has time complexity of $O((|F|+|V|)\cdot 4^D)$ per sample.
@@
     \section{Exploiting constraint consistency to reduce the complexity}\label{sec:improvedComplexity}
-
-%\textcolor{red}{\textbf{SW: Simpler alternative to
-%    Sec.~\ref{sec:first-improvedComplexity}. Note also that the use of
-%    variables in Sec.~\ref{sec:first-improvedComplexity} conflicts
-%    with the (recent) specialization of constraint networks to RNA design
-%    networks.}}
 
@@
         sampling. We show the Turner energy distributions for
-    three target structures S1, S2 and S3. By
+        three target structures R1, R2 and R3. By
         targeting the respective free energies $(-40,-40,-20)$ and
@@
 increasing) its weight, supporting our weight optimization procedure.
-\newpage
-\section{Detailed result of MultiDefect analysis}\label{sec:validity}
-N/A values correspond to data that could not be obtained by the time of this submission for two main reasons: In the case of the initial sampling (left columns), they correspond to instances which, in conjunction with an expressive energy model, resulted in very high tree-width, leading to unreasonable memory requirements incompatible with our available computing resources; The case of missing values after optimization  (right columns), they indicate situations where the initial optimization took too long ($\approx$ 1 day) and was killed.
-\input{table}
 
@@
 
+\section{Detailed result of MultiDefect analysis}\label{sec:validity}
+N/A values correspond to data that could not be obtained by the time of this submission for two main reasons: In the case of the initial sampling (left columns), they correspond to instances which, in conjunction with an expressive energy model, resulted in very high tree-width, leading to unreasonable memory requirements incompatible with our available computing resources; The case of missing values after optimization  (right columns), they indicate situations where the initial optimization took too long ($\approx$ 1 day) and was killed.
+\input{table}
+
 \bibliographystyle{vancouver}

### 2018-09-14 supplement.tex
cosmetics

@@
   \begin{center}
-    {\sf \bfseries A}\includegraphicstop[width=0.45\textwidth]{Figs/Plots/PKB00211_PKB00239_0_GC_content}
-    ~~~~~{\sf \bfseries B}
-    \includegraphicstop[width=0.45\textwidth]{Figs/Plots/supp_3str_GC_content}
+    {\sf \bfseries A}~\includegraphicstop[width=0.45\textwidth]{Figs/Plots/PKB00211_PKB00239_0_GC_content}
+    ~~~~~{\sf \bfseries B}~\includegraphicstop[width=0.45\textwidth]{Figs/Plots/supp_3str_GC_content}
   \end{center}

### 2018-09-19 RNARedPrint.tex

@@
 Let us consider $k$ features $F_1,\dots F_k$ to target all of them simultaneously.
-We base our strategy for targeting the $k$ features $F_1,\dots,F_k$ on generating samples from a Boltzmann distribution parameterized by weights $\pi_1,\dots,\pi_k$, one for each feature.
+We base our strategy for targeting the $k$ features $F_1,\dots,F_k$ on generating samples from a Boltzmann distribution parametrized by weights $\pi_1,\dots,\pi_k$, one for each feature.
 These weights essentially allow us to shift the sample distributions towards specific targets.
@@
 \section*{Conclusion}
-%%!!!!!!!!!!!!!!! consider to merge into disciussion
+%%!!!!!!!!!!!!!!! consider to merge into discussion
 %Going beyond our encoding of the base pair model energy features and the \GCb\% feature, the contributions provide more generality and allow expressing complex features of the
@@
 %(partial) sequences (even if in implementations, explicit handling of constraints is advantageous). In particular, the %framework
-%aptures various RNA energy models, they can be handled efficiently (in the sense of FPT), as long as the dependencies of their additive energy contributions are bounded.
+%captures various RNA energy models, they can be handled efficiently (in the sense of FPT), as long as the dependencies of their additive energy contributions are bounded.
 

### 2018-09-19 RNARedPrint.tex
* Added 'contribution' section to intro * Reshaped complexity paragraph

@@
 
-\hypersetup{
-    bookmarks=false,         % show bookmarks bar?
-    unicode=false,          % non-Latin characters in Acrobat’s bookmarks
-    pdftoolbar=true,        % show Acrobat’s toolbar?
-    pdfmenubar=true,        % show Acrobat’s menu?
-    pdffitwindow=true,     % window fit to page when opened
-    pdfstartview={FitH},    % fits the width of the page to the window
-    colorlinks=true,       % false: boxed links; true: colored links
+%\hypersetup{
+%    bookmarks=false,         % show bookmarks bar?
+%    unicode=false,          % non-Latin characters in Acrobat’s bookmarks
+%    pdftoolbar=true,        % show Acrobat’s toolbar?
+%    pdfmenubar=true,        % show Acrobat’s menu?
+%    pdffitwindow=true,     % window fit to page when opened
+%    pdfstartview={FitH},    % fits the width of the page to the window
+%    colorlinks=true,       % false: boxed links; true: colored links
 %    linkcolor=red!70!black,          % color of internal links (change box color with linkbordercolor)
 %    citecolor=red!70!black,        % color of links to bibliography
-    linkcolor=black,          % color of internal links (change box color with linkbordercolor)
-    citecolor=black,        % color of links to bibliography
-    filecolor=magenta,      % color of file links
-    urlcolor=red!70!black           % color of external links
-}
+%    linkcolor=black,          % color of internal links (change box color with linkbordercolor)
+%    citecolor=black,        % color of links to bibliography
+%    filecolor=magenta,      % color of file links
+%    urlcolor=red!70!black           % color of external links
+%}
 
@@
 \newtheorem{hypothesis}{Working Hypothesis}
-\newenvironment{proof}[1][]{\noindent \em Proof\ifthenelse{\equal{#1}{}}{}{ (#1)}:~}{}
+\newenvironment{proof}[1][]{\noindent Proof\ifthenelse{\equal{#1}{}}{}{ (#1)}:~}{}
 
@@
 \parHead{Specificities and similarities of multi-target design.}
-Specifically for multi-target design, virtually all available methods~\citep{Lyngsoe2012,HoenerzuSiederdissen2013,Taneda2015,Hammer2017} follow the same overall generate-and-optimize strategy.
+In the case of multi-target design, virtually all available methods~\citep{Lyngsoe2012,HoenerzuSiederdissen2013,Taneda2015,Hammer2017} follow the same overall generate-and-optimize strategy.
 %
@@
 the decomposition strategy seems to allow practical applications.
-Better understanding of the true complexity bounds is still highly relevant for the application of such methods, in particular for applications in more expressive settings (e.g.
+%
+Better understanding of the true complexity bounds is highly relevant for the application of such methods, in particular for applications in more expressive settings (e.g.
 Boltzmann sampling as discussed in this work or sampling satisfying multiple additional constraints and properties).
-All such extensions are still bounded by the complexity of the multi-target uniform sampling problem, where counting of valid designs is the dominating step (note that stochastic backtracking can be performed in linear time per sample).
-
-Consequently, for elucidating the theoretical fundament of RNA design sampling, we first need to clarify, whether efficient counting is possible.
-In this work, we show that there can not exists any efficient, polynomial time algorithm unless ${\sf P}={\sf NP}$.
-Our result relies on a surprising bijection between valid sequences and independent sets of a bipartite graph; the latter being the object of recent breakthroughs in approximate counting complexity~\citep{Bulatov2013,Cai2016}.
-
-As already demonstrated by \Software{RNAdesign}, the hardness of counting (and conjectured hardness of sampling), does still not rule out practically applicable algorithms for counting and sampling.
-Addressing the following questions in more detail and generality led to the practical advances in this work, which extend the expressivity, flexibility, and performance of multi-structure design: \emph{How to sample, in a general way, from a Boltzmann distribution based on expressive energy models?
+All such extensions are still limited by the complexity of the multi-target uniform sampling problem, where counting of valid designs is the dominating step.
+In other words key questions include: \emph{How to sample, in a general way, from a Boltzmann distribution based on expressive energy models?
 How to enforce additional constraints, such as \GCb\%, complex sequence constraints, or the energy of individual structures?}
+
+\subsection*{Contributions}
 
@@
 
-To answer these questions, we introduce a generic framework (illustrated in~Fig.~\ref{fig:workflow}) enabling efficient Boltzmann-weighted sampling over RNA sequences with multiple target structures.
+Addressing these questions in more detail and generality leads to the practical advances in this work, which extend the expressivity, flexibility, and performance of multi-structure design.
+Namely, we introduce a generic framework (illustrated in~Fig.~\ref{fig:workflow}) enabling efficient Boltzmann-weighted sampling over RNA sequences with multiple target structures.
 As main algorithmic contribution of this work, we devise a dynamic programming algorithm, based on a tree decomposition, to compute partition functions and sample sequences from the Boltzmann distribution%
@@
 .
-We show that these algorithms are time and space-efficient depending on a specific property of the tree decomposition, namely the treewidth; in short, our algorithms are fixed-parameter tractable (FPT).
+We show that these algorithms are time and space-efficient, while requiring exponential time based on a specific property of the tree decomposition, called treewidth, whose value is typically low for typical instances; in short, our algorithms are fixed-parameter tractable (FPT).
+
+This exponential dependency is arguably necessary, as the underlying counting problem is \#{\sf P}-hard, \emph{i.e.}  we show that there can not exists any efficient, polynomial time algorithm unless ${\sf P}={\sf NP}$.
+Our result relies on a surprising bijection between valid sequences and independent sets of a bipartite graph; the latter being the object of recent breakthroughs in approximate counting complexity~\citep{Bulatov2013,Cai2016}.
+
 % Subsection~\ref{sec:complexity}
@@
 We observe that
-targeting realistic RNA energies in the realistic
+targeting realistic RNA energies in the
 Turner RNA energy model
@@
 $$
-F(S) =\!\!\!\!\!\!\sum_{\substack{\text{$f$: $f$ is contribution of $F$,}\\\text{ $\dep(f)=\{x_1,\ldots,x_p\}$}}}\!\!\!\!\!\! f(\{x_1\mapsto S_{x_1},\ldots,x_p\mapsto S_{x_p}\}).
+F(S) =\!\!\!\!\!\!\!\sum_{\substack{\text{$f$: $f$ is contribution of $F$,}\\\text{ $\dep(f)=\{x_1,\ldots,x_p\}$}}}\!\!\!\!\!\!\! f(\{x_1\mapsto S_{x_1},\ldots,x_p\mapsto S_{x_p}\}).
 $$
@@
   connected components}.
-
 The problem becomes much harder when $\{\Gb, \Ub\}$ base pairs are allowed, and we show that valid designs for a set of structures cannot be counted in polynomial time, unless ${\sf\# P}={\sf FP}$.
@@
 
-\begin{proposition}
-  $\Design{G}$ and $\IS{G}$ have equal size.
+\begin{proposition}\label{prop:bijection}
+  $|\Design{G}| = |\IS{G}|$.
 \end{proposition}
@@
 \begin{proof}
-Consider the mapping $\Psi: \Design{G} \to \IS{G}$ defined by $ \Psi(f) := \left\{v\in V\mid f(v)\in\{\Ab,\Cb\}\right\}.$
+Consider the mapping $\Psi: \Design{G} \to \IS{G}, f \mapsto \left\{v\in V\mid f(v)\in\{\Ab,\Cb\}\right\}.$
 We show that $\Psi$ is bijective:%
@@
 Now we can build on the connection between the two problems to obtain complexity results for \NumDesign.
-Counting independent sets in bipartite graphs ($\#{\sf BIS}$) is indeed a well-studied \#{\sf P}-hard problem~\citep{Ge2012}, from which we immediately conclude:
+Counting independent sets in bipartite graphs ($\#{\sf BIS}$) is indeed a well-studied \#{\sf P}-hard problem~\citep{Ge2012}.
+Now assume the existence of an efficient (polynomial-time) algorithm $\mathcal{A}$ for computing $|\NumDesign(G)|$ on connected (bipartite) graphs.
+Then, running $\mathcal{A}$ and returning $|\NumDesign(G)|/2$ constitutes an efficient algorithm for $\#{\sf BIS}$ on connected graphs.
+However, $\#{\sf BIS}$ is also \#{\sf P}-hard on connected graphs, as the number of independent sets for a disconnected graph $G$ can be efficiently computed by $|\IS{G}|=\prod_{cc\in CC(G)} |\IS{cc}|$.
+In other words, any efficient algorithm for \NumDesign implies an efficient algorithm for $\#{\sf BIS}$, and we conclude with the hardness of \NumDesign.
+
 \begin{corollary}
@@
 \end{corollary}
-\begin{proof}
-  Note that $\#{\sf BIS}$ is also \#{\sf P}-hard on connected graphs, as the number of independent sets for a disconnected graph $G$ is given by $|\IS{G}|=\prod_{cc\in CC(G)} |\IS{cc}|$.
-Thus any efficient algorithm for $\#{\sf BIS}$ on connected instances provides an efficient algorithm for general graphs.
-
-  Given any algorithm $\mathcal{A}$ that solves \NumDesign over strongly-connected graphs $G$, one can  construct an algorithm $\mathcal{A}'$ that solves $\#{\sf BIS}$: 1) calculate $\NumDesign(G)$ by calling $\mathcal{A}$ 2) return $\NumDesign(G)/2=|\Design{G}| = |\IS{G}|$.
-    Obviously, if $\mathcal{A}$ is a polynomial-time algorithm, so is $\mathcal{A'}$.
-  This implies that $\NumDesign$  is at least as hard as $\#{\sf BIS}$, thus does not admit a polynomial time exact algorithm unless $\#{\sf P}={\sf FP}$.
-\end{proof}
+% \begin{proof}
+
+%   Given any algorithm $\mathcal{A}$ that solves \NumDesign over strongly-connected graphs $G$, one can  construct an algorithm $\mathcal{A}'$ that solves $\#{\sf BIS}$:
+%   \begin{enumerate}
+%     \item[1)] calculate $\NumDesign(G)$ by calling $\mathcal{A}$
+%     \item[2)] return\\$\NumDesign(G)/2=|\Design{G}| = |\IS{G}|$.
+%   \end{enumerate}
+%     Obviously, if $\mathcal{A}$ is a polynomial-time algorithm, so is $\mathcal{A'}$.
+%   This implies that $\NumDesign$  is at least as hard as $\#{\sf BIS}$, thus does not admit a polynomial time exact algorithm unless $\#{\sf P}={\sf FP}$.
+% \end{proof}
 
@@
 Our Boltzmann sampling strategy (Algorithms
-\ref{alg:pf} and \ref{alg:sampling}) were implemented, using a combination of {\tt C++} and {\tt Python}, in a software called \ourprog{}, available at:
+\ref{alg:pf} and \ref{alg:sampling}) was implemented, using a combination of {\tt C++} and {\tt Python}, in a software called \ourprog{}, available at:
 
@@
 
-{\bfseries Stefan: Please list available options/features from the Python interface + output format}
+We implemented the core algorithms in C++, resulting in the tool
+C++-tool \ourprog{}, which takes a set of target structures, as well as
+weights for each energy feature and \GCb\%; then,
+it generates a sample (of prescribed size) from the corresponding
+Boltzmann distribution. For \ourprog{}, we implemented the base pair
+energy model and the stacking energy model.
+Moreover, we provide a Python wrapper script for the C++-tool
+\ourprog{} to perform targeting, based on multi-dimensional Boltzmann
+sampling.
+For given secondary structures, together with prescribed target energies
+and target \GCb\%, this script generates
+a series of sequences that satisfy the target values for the energy and
+\GCb\% features within configurable tolerances.
+Notably, these target energies are actual free energies in the realistic
+Turner energy model, which are targeted by efficiently sampling in the
+base pair energy model.
 
@@
     %{\sf \bfseries A}\
-    \includegraphicstop[width=0.99\textwidth]{Figs/energy_shift}\hfill
+    \includegraphicstop[width=0.97\textwidth]{Figs/energy_shift}\hfill
     %{\sf \bfseries B}\ \includegraphicstop[width=0.55\textwidth]{Figs/PKB00211_PKB00239_0_energy_distribution}
@@
 penalizing a set of alternative helices/structures.
+{\em Add RNASKetch to conclusion for more complex design scenario + need for generic constratins-based framework for flexibility}
 

### 2018-09-21 supplement.tex

--- 2018-09-14 supplement.tex
@@
   \begin{center}
-    {\sf \bfseries A}\includegraphicstop[width=0.45\textwidth]{Figs/Plots/PKB00211_PKB00239_0_GC_content}
-    ~~~~~{\sf \bfseries B}
-    \includegraphicstop[width=0.45\textwidth]{Figs/Plots/supp_3str_GC_content}
+    {\sf \bfseries A}~\includegraphicstop[width=0.45\textwidth]{Figs/Plots/PKhttps://v2.overleaf.com/6339486874jnyvnycthtddB00211_PKB00239_0_GC_content}
+    ~~~~~{\sf \bfseries B}~\includegraphicstop[width=0.45\textwidth]{Figs/Plots/supp_3str_GC_content}
   \end{center}

### 2018-09-26 RNARedPrint.tex

@@
 We introduced a novel approach, based on a general framework and efficient algorithms, to design RNA
-sequences, while targeting very specific complex properties; in particular, we describe the targeting of the free energies of multiple target strucutures and the \GCb-content.
+sequences, while targeting very specific complex properties. In particular, we describe the targeting of the free energies of multiple target strucutures and the \GCb-content.
 Our method combines an FPT stochastic sampling algorithm
@@
 %
-To substantiate our work additionally, we establish the $\#${\sf P}-hardness of uniform sampling;
-from a complexity-theoretic point of view, this motivates the FPT, tree decomposition-based nature of our method.
+To substantiate our work additionally, we establish the $\#${\sf P}-hardness of uniform sampling which,
+from a complexity-theoretic point of view, motivates the FPT, tree decomposition-based nature of our method.
 
@@
 strategies. We demonstrated the practical feasibility of such controlled sequence
-generation, including the case of pseudoknotted structures, and studied its use for multi-target RNA design. Moreover,
-our framework is extensible to include various more complex sequence
-constraints, including mandatory/forbidden motifs at specific
+generation, including the case of pseudoknotted structures, and studied its use for multi-target RNA design.
+
+%perspectives:
+Our framework is extensible to include various more complex sequence constraints, including mandatory/forbidden motifs at specific
 positions, or anywhere in the designed sequences, by adapting formal language constructs
-introduced by \citet{Zhou2013}. In future work, negative design principles could 
+of Zhou \emph{et al}~\citet{Zhou2013}. In future work, negative design principles could 
 be explicitly supported at the generation stage, for instance by
 penalizing a set of alternative helices/structures.
-{\em Add RNASKetch to conclusion for more complex design scenario + need for generic constratins-based framework for flexibility}
+
+{Stefan: \em Mention RNAsketch to perspectives for more complex design scenario + need for generic constraints-based framework for flexibility ($\to$ generic constraint framework).}
+
+{Yann: \em Connection with statistics (design as perfect random sequences for null-model), mention Rivas/Eddy lncRNAs papers...}
 

### 2018-09-26 supplement.tex
Minor edits

--- 2018-09-14 supplement.tex
@@
 \begin{figure}
-\textbf{A}~\includegraphicstop[width=0.48\textwidth]{Figs/Plots/stem_weight1-500}%
-\textbf{B}~\includegraphicstop[width=0.48\textwidth]{Figs/Plots/doublestem_weight1-500}%
-\caption{Comparision of uniform sampling and Boltzmann sampling with
+\textbf{A}~\includegraphicstop[width=0.48\textwidth]{Figs/Plots/stem_weight1-500_mean}%
+\textbf{B}~\includegraphicstop[width=0.48\textwidth]{Figs/Plots/doublestem_weight1-500_mean}%
+\caption{Comparison of uniform sampling and Boltzmann sampling with
 weight $\pi=500$. We show results for two different small structures: (A) single stem structure (B) double stem structure. For each respective structure $R$, we plot the energies $E(S,R)$ of
-sampled sequences against the probabilities $Pr(R|S)$; the former is controlled
-by Boltzmann sampling, while the latter is a common negative design objective.}
+10000 sampled sequences against the probabilities $Pr(R|S)$ (in log scale).
+For illustrating the trend, we show moving averages over windows of $\pm 1$ kcal.
+Note that while only the energy $E(S,R)$ is directly controlled by Boltzmann sampling, this strongly impacts the common negative design objective $Pr(R|S)$.}
 \label{fig:negforpos}

### 2018-09-26 RNARedPrint.tex
Edit intro

@@
 \parHead{On the importance of sampling for design.}
-Many computational methods for RNA design follow the ``generate-and-optimize'' strategy: they randomly generate one or several \Def{seed} sequences, which are then optimized.
-In many cases, the seed quality was
-found to be critical for the empirical performance of RNA design
-methods~\citep{Levin2012}. Generally, random seed generation
-improves the prospect of subsequent optimizations, helping to overcome
-local optima of the objective function, and increases the diversity
+Many computational methods for RNA design follow the ``generate-and-optimize'' strategy: \Def{seed} sequences are randomly generated and then optimized.
+While the quality of the seeds was
+found to be performance critical for such RNA design
+methods~\citep{Levin2012}, random seed generation
+can improve the prospect of subsequent optimizations
+and increases the diversity
 across designs~\citep{Reinharz2013}.  For single-target approaches,
@@
 \parHead{Specificities and similarities of multi-target design.}
-In the case of multi-target design, virtually all available methods~\citep{Lyngsoe2012,HoenerzuSiederdissen2013,Taneda2015,Hammer2017} follow the same overall generate-and-optimize strategy.
-%
-Facing the complex sequence constraints induced by multiple targets, early methods such as \Software{Frnakenstein}~\citep{Lyngsoe2012} and \Software{Modena}~\citep{Taneda2015} did not attempt to solve sequence generation systematically, but rely on ad-hoc sampling strategies.
+The available methods for multi-target RNA design~\citep{Lyngsoe2012,HoenerzuSiederdissen2013,Taneda2015,Hammer2017} all follow the same overall generate-and-optimize strategy.
+%
+Faced with the complex constraints due to the multiple targets, early methods such as \Software{Frnakenstein}~\citep{Lyngsoe2012} and \Software{Modena}~\citep{Taneda2015} passed on any systematic generation of sequences, but rely on ad-hoc sampling strategies.
 %
 Recently, the approach \Software{RNAdesign}~\citep{HoenerzuSiederdissen2013}, coupled with local
-search within \RNAblueprint{}~\citep{Hammer2017}, solved the
+search in \RNAblueprint{}~\citep{Hammer2017}, solved the
 problem of sampling seeds from the \emph{uniform} distribution for multiple
@@
 choosing individual nucleotide assignments with probabilities derived
-from the subsolution counts. The overall complexity of
-\Software{RNAdesign} grows like $\Theta(4^{\gamma})$, where the
-parameter $\gamma$, while bounded only by the length of the designed RNA, is
-typically much lower, due to the decomposition strategy.
+from the subsolution counts. While, due to its decomposition strategy, 
+\Software{RNAdesign} performs much better than the theoretical bound of $O(4^n)$, no attempts were made to characterize or justify its---still exponential---complexity; leaving important theoretical questions of the complexity of counting and uniform sampling open.
+As well, the \Software{RNAdesign}/\RNAblueprint{} approach is very specialized to uniform sampling, which limits its direct extensibility.
+Both aspects call for a more systematic redesign of multi-target sampling, which builds on established concepts in computer science.
+In turn, this promises a fundamentally broader range of applications for extensions of the sampling method.
 
 \parHead{Motivation.}
-While uniform sampling in \Software{RNAdesign}/\RNAblueprint{} turned out to be highly beneficial for general RNA design,
-the work on these tools gave no definitive answer on the \emph{complexity of uniform sampling for multi-target design}. While time and space requirements grow exponential in the general case,
-the decomposition strategy seems to allow practical applications.
-%
-Better understanding of the true complexity bounds is highly relevant for the application of such methods, in particular for applications in more expressive settings (e.g.
+%Better understanding of the true complexity bounds is highly relevant for the application of such methods, in particular for applications in more expressive settings (e.g.
 Boltzmann sampling as discussed in this work or sampling satisfying multiple additional constraints and properties).
-All such extensions are still limited by the complexity of the multi-target uniform sampling problem, where counting of valid designs is the dominating step.
-In other words key questions include: \emph{How to sample, in a general way, from a Boltzmann distribution based on expressive energy models?
-How to enforce additional constraints, such as \GCb\%, complex sequence constraints, or the energy of individual structures?}
+
+%All such extensions are still limited by the complexity of the multi-target uniform sampling problem, where counting of valid designs is the dominating step.
 
 \subsection*{Contributions}
-
+%
 \begin{figure*}[t]
@@
 \end{figure*}
-
-Addressing these questions in more detail and generality leads to the practical advances in this work, which extend the expressivity, flexibility, and performance of multi-structure design.
-Namely, we introduce a generic framework (illustrated in~Fig.~\ref{fig:workflow}) enabling efficient Boltzmann-weighted sampling over RNA sequences with multiple target structures.
-As main algorithmic contribution of this work, we devise a dynamic programming algorithm, based on a tree decomposition, to compute partition functions and sample sequences from the Boltzmann distribution%
+%
+In particular, we ask in this paper: \emph{How to sample, in a general way, from a Boltzmann distribution (which is a natural generalization of uniform sampling) based on expressive energy models?
+How to enforce additional constraints, such as \GCb\%, complex sequence constraints, or the energy of individual structures?}
+
+For addressing these questions,
+we introduce a generic framework (illustrated in~Fig.~\ref{fig:workflow}) enabling efficient Boltzmann-weighted sampling over RNA sequences with multiple target structures.
+As main algorithmic contribution, we devise a dynamic programming algorithm, based on a tree decomposition, to compute partition functions.
+This serves as precomputation for a second algorithm, which samples sequences from the Boltzmann distribution%
 % (Subsection~\ref{sec:PF})
-.
-We show that these algorithms are time and space-efficient, while requiring exponential time based on a specific property of the tree decomposition, called treewidth, whose value is typically low for typical instances; in short, our algorithms are fixed-parameter tractable (FPT).
-
-This exponential dependency is arguably necessary, as the underlying counting problem is \#{\sf P}-hard, \emph{i.e.}  we show that there can not exists any efficient, polynomial time algorithm unless ${\sf P}={\sf NP}$.
-Our result relies on a surprising bijection between valid sequences and independent sets of a bipartite graph; the latter being the object of recent breakthroughs in approximate counting complexity~\citep{Bulatov2013,Cai2016}.
+. We show that both algorithms are time and space-efficient for a fixed 
+ property of the tree decomposition, called the treewidth.
+This parameter, has an exponential impact on the complexity, but is low for typical instances; in short, our algorithms are fixed-parameter tractable (FPT).
+
+To theoretically justify the exponential dependency on the treewidth, we show that the underlying counting problem is \#{\sf P}-hard; this implies that even the counting problem cannot be solved efficiently (i.e.~in polynomial time) unless ${\sf P}={\sf NP}$.
+This complexity result relies on a surprising bijection between valid sequences and independent sets of a bipartite graph; the latter being the object of recent breakthroughs in approximate counting complexity~\citep{Bulatov2013,Cai2016}.
 
@@
 %
-Due to the generality of our method, we can moreover strongly limit this parameter in practice by using state-of-the-art tree decomposition algorithms.
-%Uniform sampling is handled as a special case of Boltzmann sampling, where each valid sequence receives energy zero and---consequently---computing partition functions specializes to counting.
-By evaluating (partial) sequences in a specialized weighted constraint
-network, we support arbitrary multi-ary constraints and thus
+In practice, due to the systematic construction of our method, which is based on tree decomposition, we can strongly limit the treewidth by using state-of-the-art tree decomposition algorithms.
+%
+Moreover, we support arbitrary multi-ary constraints and thus
 arbitrarily complex energy models,
@@
 %(Subsection~\ref{sec:energy_models})
-.  Moreover, we describe an \Def{adaptive
+.
+
+Finally, we describe an \Def{adaptive
   sampling} strategy to control the free energies of the individual
@@
 . %
-
+%
 We observe that
@@
 works well by performing sampling based on less complex RNA energy models, which keeps the treewidth low.
-This result is highly relevant in practice, since it allows to combine high efficiency with high accuracy and moreover leaves room for extensions of the method.
+In practice, this allows to combine high efficiency with high accuracy and moreover leaves room for extensions of the method.
 

### 2018-09-27 RNARedPrint.tex
Revamped intro

@@
 additionally aiming at other features, such as specific
-\GCb-content (\GCb\%) as previously done by~\citet{Reinharz2013} or the presence/absence of
-functionally relevant motifs (either anywhere or at specific
-positions)~\citep{Zhou2013}; these objectives motivate flexible
+\GCb-content (\GCb\%)~\citet{Reinharz2013} or the presence/absence of
+functionally relevant motifs, either anywhere or at specific
+positions~\citep{Zhou2013}; these objectives motivate flexible
 computational design methods.
@@
 \parHead{Motivation.}
-%Better understanding of the true complexity bounds is highly relevant for the application of such methods, in particular for applications in more expressive settings (e.g.
+Better understanding of the true complexity bounds is highly relevant for the application of such methods, in particular for applications in more expressive settings (e.g.
 Boltzmann sampling as discussed in this work or sampling satisfying multiple additional constraints and properties).
+All such extensions are still limited by the complexity of the multi-target uniform sampling problem, where counting of valid designs is the dominating step.
+In other words key questions include: \emph{How to sample, in a general way, from a Boltzmann distribution based on expressive energy models?
+How to enforce additional constraints, such as \GCb\%, complex sequence constraints, or the energy of individual structures?}
 
@@
 \subsection*{Contributions}
-%
+
 \begin{figure*}[t]
@@
 \end{figure*}
-%
-In particular, we ask in this paper: \emph{How to sample, in a general way, from a Boltzmann distribution (which is a natural generalization of uniform sampling) based on expressive energy models?
-How to enforce additional constraints, such as \GCb\%, complex sequence constraints, or the energy of individual structures?}
-
-For addressing these questions,
-we introduce a generic framework (illustrated in~Fig.~\ref{fig:workflow}) enabling efficient Boltzmann-weighted sampling over RNA sequences with multiple target structures.
-As main algorithmic contribution, we devise a dynamic programming algorithm, based on a tree decomposition, to compute partition functions.
-This serves as precomputation for a second algorithm, which samples sequences from the Boltzmann distribution%
+
+Addressing these questions in more detail and generality leads to the practical advances in this work, which extend the expressivity, flexibility, and performance of multi-structure design.
+Namely, we introduce a generic framework, summarized in~Fig.~\ref{fig:workflow}, enabling efficient Boltzmann-weighted sampling over RNA sequences with multiple target structures.
+The main algorithmic contribution of this work consists in dynamic programming algorithms, based on the concept of \Def{tree decomposition}, to compute the dual partition function, and sample sequences from the Boltzmann distribution%
 % (Subsection~\ref{sec:PF})
-. We show that both algorithms are time and space-efficient for a fixed 
- property of the tree decomposition, called the treewidth.
-This parameter, has an exponential impact on the complexity, but is low for typical instances; in short, our algorithms are fixed-parameter tractable (FPT).
-
-To theoretically justify the exponential dependency on the treewidth, we show that the underlying counting problem is \#{\sf P}-hard; this implies that even the counting problem cannot be solved efficiently (i.e.~in polynomial time) unless ${\sf P}={\sf NP}$.
-This complexity result relies on a surprising bijection between valid sequences and independent sets of a bipartite graph; the latter being the object of recent breakthroughs in approximate counting complexity~\citep{Bulatov2013,Cai2016}.
+.
+The complexities of our algorithms depend exponentially on a specific property of the tree decomposition, called the \Def{treewidth}, whose value remains typically low for typical instances.
+However, for any given value of the treewidth, the complexity only scales linearly with the size of designed sequences and the number of targeted structures, \emph{i.e.} our algorithms are Fixed-Parameter Tractable (FPT).
+
+This exponential dependency is arguably necessary.
+Indeed, the underlying counting problem is \#{\sf P}-hard, meaning that there is no efficient, polynomial time algorithm for the problem unless ${\sf P}={\sf NP}$.
+Our result relies on a surprising bijection between valid sequences and independent sets of a bipartite graph, the latter being the object of recent breakthroughs in approximate counting complexity~\citep{Bulatov2013,Cai2016}.
 
@@
 %
-In practice, due to the systematic construction of our method, which is based on tree decomposition, we can strongly limit the treewidth by using state-of-the-art tree decomposition algorithms.
-%
-Moreover, we support arbitrary multi-ary constraints and thus
+Due to the generality of our method, we can moreover strongly limit this parameter in practice by using state-of-the-art tree decomposition algorithms.
+%Uniform sampling is handled as a special case of Boltzmann sampling, where each valid sequence receives energy zero and---consequently---computing partition functions specializes to counting.
+By evaluating (partial) sequences in a specialized weighted constraint
+network, we support arbitrary multi-ary constraints and thus
 arbitrarily complex energy models,
@@
 %(Subsection~\ref{sec:energy_models})
-.
-
-Finally, we describe an \Def{adaptive
+.  Moreover, we describe an \Def{adaptive
   sampling} strategy to control the free energies of the individual
@@
 . %
-%
+
 We observe that
@@
 works well by performing sampling based on less complex RNA energy models, which keeps the treewidth low.
-In practice, this allows to combine high efficiency with high accuracy and moreover leaves room for extensions of the method.
+This result is highly relevant in practice, since it allows to combine high efficiency with high accuracy and moreover leaves room for extensions of the method.
 
 Eventually this strategy allows generating biologically relevant
-multi-target designs, as we demonstrate by performing negative multi-target RNA design for a comprehensive, 
-meaningful benchmark set~\citep{Taneda2015}.
+multi-target designs, as demonstrated by its leading performance on a comprehensive benchmark set for multi-target RNA design~\citep{Taneda2015}.
 
@@
 \label{sec:problem-statement}
-The main computational problem addressed in this work is to generate sequences over the alphabet $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}$ with prescribed energies, for each single target structure, and $\GCb\%$.
-Remarkably, such criteria can be homogeneously abstracted as constraints on the value of \Def{features}, each defined as sums over atomic feature contributions depending on limited subsets of sequence positions.
-
-\paragraph{Central computational problem.}
+The main computational problem addressed in this work is the design of sequences over the alphabet $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}$, compatible with a set of secondary structures, associated with prescribed free-energies, and featuring a given $\GCb\%$.
+Such criteria can be homogeneously abstracted as constraints on the value of \Def{features}, each defined as the sum  of real-valued contributions depending on subsets of the sequence positions.
+
+To achieve maximal diversity within candidate designs, we abstract the problem as a random generation of sequences within a Boltzmann distribution.
+The probability of a sequence then depends on its associated features, and associated weights which
+ influence the expected feature values of generated sequences, allowing us to shift the sample distributions towards specific targets.
+Sampling from this distribution requires to compute corresponding partition functions, such that we can draw sequences with probabilities proportional to their Boltzmann weight.
+
+%\paragraph{Central computational problem.}
 Let us consider $k$ features $F_1,\dots F_k$ to target all of them simultaneously.
 We base our strategy for targeting the $k$ features $F_1,\dots,F_k$ on generating samples from a Boltzmann distribution parametrized by weights $\pi_1,\dots,\pi_k$, one for each feature.
-These weights essentially allow us to shift the sample distributions towards specific targets.
-Sampling from this distribution requires to compute corresponding partition functions, such that we can draw sequences with probabilities proportional to their Boltzmann weight:
 \begin{equation}
@@
 
-\paragraph{Representing \GCb\%-content and energies as features.}
+\subsection*{Representing \GCb\%-content and energies as features.}
 Formally, we define a \Def{feature $F$} as function on sequences, whose value is obtained by summing over an associated set of \Def{atomic feature contributions}, briefly called \emph{contributions}.
@@
 
-For the definition of energy features, we abstract an \Def{RNA (secondary) structure $R$ of length $n$} as a set of \Def{base pairs} $(i,j)$, where
+For the definition of energy features, we abstract an \Def{RNA (secondary) structure} $R$ of length $n$ as a set of \Def{base pairs} $(i,j)$, where
 $1\leq i<j\leq n$ and for all different $(i,j), (i',j')\in R$: $\{i,j\}\cap\{i',j'\}=\emptyset$ (``degree $\leq$ 1'')---without precluding pseudoknots.
@@
 Note that for \GCb\% as well as for the base pair energy features, the contributions depend on the nucleotides at only a small, constantly bounded number of sequence positions (one and two positions, respectively).
-This is important for the efficiency of our approach, which is designed to profit from contributions that each depend on only few sequence positions.
-We emphasize that nevertheless, the same framework can capture more expressive RNA energy models like the Turner energy model~\citep{Turner2009}, since such models generally define a function $E$ that maps a pair of a sequence $S$ and a structure $R$ to a real value, the \Def{energy} $E(S,R)$.
+
+%This is important for the efficiency of our approach, which is designed to profit from contributions that each depend on only few sequence positions. 
+%We emphasize that nevertheless, the same framework can capture more expressive RNA energy models like the Turner energy model~\citep{Turner2009}, since such models generally define a function $E$ that maps a pair of a sequence $S$ and a structure $R$ to a real value, the \Def{energy} $E(S,R)$.
 Here, our approach automatically adapts to the more complex energy functions, leading to a controlled increase of the computational complexity.
-
-Typically, such RNA energy models consider a sequence $S$ \Def{valid} for $R$, iff $\{S_i,S_j\}\in \B$ for
-all $(i,j)\in R$, where $\B:=\left\{\{\Ab,\Ub\},\{\Gb,\Cb\},\{\Gb,\Ub\}\right\}$ describes the valid base pairs.
 
@@
 
-Finally, to stress the analogy to, in the RNA context more familiar (e.g.~\citep{McCaskill1990}), Boltzmann distributions over the structure ensemble of a sequence $S$, we emphasize the equivalence
-\begin{equation}
-\label{eq:feature-energy-transformation}
-\prod_{1\leq \ell \leq k} \pi_\ell^{-F_\ell(S)} = exp(-\beta E(S)),
-\end{equation}
-where $\beta$ denotes inverse temperature and $E(S)$ is defined as linear combination of energies $E(S)=\sum_{1\leq \ell \leq k} \ln(\pi_\ell)/\beta E(S,R_\ell)$,
-where the energy $E(S,R_\ell)$ is treated as feature $F_\ell(S) = \EbpSym(S,R_\ell);$ as elaborated later, even Turner energies or pseudoknot energies can be expressed as features due to their additive nature.
-
-\paragraph{Dependency structure and complexity}
+%Finally, to stress the analogy to, in the RNA context more familiar (e.g.~\citep{McCaskill1990}), Boltzmann distributions over the structure ensemble of a sequence $S$, we emphasize the equivalence
+%\begin{equation}
+%\label{eq:feature-energy-transformation}
+%\prod_{1\leq \ell \leq k} \pi_\ell^{-F_\ell(S)} = exp(-\beta E(S)),
+%\end{equation}
+%where $\beta$ denotes inverse temperature and $E(S)$ is defined as linear combination of energies $E(S)=\sum_{1\leq \ell \leq k} \ln(\pi_\ell)/\beta E(S,R_\ell)$,
+%where the energy $E(S,R_\ell)$ is treated as feature $F_\ell(S) = \EbpSym(S,R_\ell);$ as elaborated later, even Turner energies or pseudoknot energies can be expressed as features due to their additive nature.
+
+\subsection*{Dependency structure, tree decomposition and width}
 %The complexity of the central problem depends fundamentally on the dependency structure that is imposed by the entire set of contributions.
@@
 %For design in the base pair energy model or the related counting problem, the dependencies are immediately derived from the set of target structures $\R$.
+
+Typically, such RNA energy models consider a sequence $S$ \Def{valid} for $R$, iff $\{S_i,S_j\}\in \B$ for
+all $(i,j)\in R$, where $\B:=\left\{\{\Ab,\Ub\},\{\Gb,\Cb\},\{\Gb,\Ub\}\right\}$ describes the valid base pairs.
 

### 2018-09-28 RNARedPrint.tex
Pass through all doc

@@
 
+\newcommand*{\QEDA}{\hfill\ensuremath{\blacksquare}}%
+
 %\newtheorem{algorithm}{Algorithm}
@@
 \newtheorem{hypothesis}{Working Hypothesis}
-\newenvironment{proof}[1][]{\noindent Proof\ifthenelse{\equal{#1}{}}{}{ (#1)}:~}{}
+\newenvironment{proof}[1][]{\noindent Proof\ifthenelse{\equal{#1}{}}{}{ (#1)}.~}{}
 
@@
 \newcommand{\citet}[1]{\cite{#1}}
+
+\renewcommand{\Pr}[1]{\mathbb{P}{#1}}
 
@@
 \parHead{Specificities and similarities of multi-target design.}
-The available methods for multi-target RNA design~\citep{Lyngsoe2012,HoenerzuSiederdissen2013,Taneda2015,Hammer2017} all follow the same overall generate-and-optimize strategy.
+Available methods for multi-target RNA design~\citep{Lyngsoe2012,HoenerzuSiederdissen2013,Taneda2015,Hammer2017} all follow the same overall generate-and-optimize strategy.
 %
@@
 \section*{Methods}
+The main computational problem addressed in this work is the design of sequences over the alphabet $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}$, featuring a given $\GCb\%$, and compatible with a set of secondary structures, associated with prescribed free-energies.
+Such criteria are here abstracted as constraints on the value of \Def{features}, functions of the sequence that are expressed as sums over real-valued contributions.
+Such contributions typically depend on the nucleotides assigned to a subset of sequence positions.
+
+To achieve maximal diversity within candidate designs, we abstract the problem as a random generation of sequences within a Boltzmann distribution.
+The probability of a sequence then depends on its associated features, and of weights that are associated features.
+Those influence the expected feature values of generated sequences, allowing us to shift the sample distributions towards specific targets.
+Sampling from this distribution requires to compute corresponding partition functions, such that we can draw sequences with probabilities proportional to their Boltzmann weight.
+A final rejection step, made efficient by the fine calibration of weights, allows to produce sequences satisfying the input set of constraints.
+
 \subsection*{Problem statement}
 \label{sec:problem-statement}
-The main computational problem addressed in this work is the design of sequences over the alphabet $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}$, compatible with a set of secondary structures, associated with prescribed free-energies, and featuring a given $\GCb\%$.
-Such criteria can be homogeneously abstracted as constraints on the value of \Def{features}, each defined as the sum  of real-valued contributions depending on subsets of the sequence positions.
-
-To achieve maximal diversity within candidate designs, we abstract the problem as a random generation of sequences within a Boltzmann distribution.
-The probability of a sequence then depends on its associated features, and associated weights which
- influence the expected feature values of generated sequences, allowing us to shift the sample distributions towards specific targets.
-Sampling from this distribution requires to compute corresponding partition functions, such that we can draw sequences with probabilities proportional to their Boltzmann weight.
-
 %\paragraph{Central computational problem.}
-Let us consider $k$ features $F_1,\dots F_k$ to target all of them simultaneously.
-We base our strategy for targeting the $k$ features $F_1,\dots,F_k$ on generating samples from a Boltzmann distribution parametrized by weights $\pi_1,\dots,\pi_k$, one for each feature.
+Let us consider $m$ (secondary) structures $R_1,\dots, R_m$, each abstracted as a set of base pairs, and $k$ features $F_1,\dots, F_k$, associated with weights $\pi_1,\dots,\pi_k$, our goal is to sample from the Boltzmann distribution, defined over the set of sequences that satisfy classic base-pairing rules for each structure as:
 \begin{equation}
 \label{eq:sample-distribution}
-\Pr(S|\pi_1,\dots,\pi_k) \propto \prod_{1\leq \ell\leq k} \pi_\ell^{-F_\ell(S)}.
-\end{equation}
-
-The workhorse of our approach is the efficient (more precisely, fixed-parameter tractable) computation of feature-dependent partition functions over sequences.
-Given are the features $F_1,\dots,F_k$, which encode the energies functions for the fixed structures and the function for the \GCb\%, all in terms of simple atomic energy contributions.
+\Pr(S\mid\pi_1,\dots,\pi_k) \propto \displaystyle\prod_{1\leq \ell\leq k}\!\! \pi_\ell^{-F_\ell(S)} \end{equation}
+
+The workhorse of our approach is the efficient (more precisely, fixed-parameter tractable) computation of feature-dependent partition functions over sequences:
+%Given are the features $F_1,\dots,F_k$, which encode the energies functions for the fixed structures and the function for the \GCb\%, all in terms of simple atomic energy contributions.
 For weights $\pi_1,\dots,\pi_k$, one wants to compute the partition function
@@
 \subsection*{Representing \GCb\%-content and energies as features.}
-Formally, we define a \Def{feature $F$} as function on sequences, whose value is obtained by summing over an associated set of \Def{atomic feature contributions}, briefly called \emph{contributions}.
-Each feature $f$ is associated with a restricted set of positions, its \emph{dependencies $\dep(f)$}, such that
+Formally, we define a \Def{feature $F$} as a function on sequences, whose value is obtained by summing over an associated set of \Def{contributions}.
+Each contribution $f$ depends on a restricted set of positions, its \emph{dependencies $\dep(f)$}, such that
 $$
@@
 
-Thus, the $\GCb\%$ feature is expressed by $n$ contributions $f^{\GCb}_i$ where $f^{\GCb}_i(\{i\mapsto S_i\})$ equals $-1$ if $S_i=\Gb$ or $\Cb$, and $0$ otherwise $(i=1,\dots,n)$.
-
-For the definition of energy features, we abstract an \Def{RNA (secondary) structure} $R$ of length $n$ as a set of \Def{base pairs} $(i,j)$, where
-$1\leq i<j\leq n$ and for all different $(i,j), (i',j')\in R$: $\{i,j\}\cap\{i',j'\}=\emptyset$ (``degree $\leq$ 1'')---without precluding pseudoknots.
+For instance, the $\GCb\%$ can be expressed using $n$ contributions $f^{\GCb}_i$, $i\in[1,n]$,  where $f^{\GCb}_i(\{i\mapsto S_i\})=-1$ if $S_i=\Gb$ or $\Cb$, and $0$ otherwise.
+
+%For the definition of energy features, we abstract an \Def{RNA (secondary) structure} $R$ of length $n$ as a set of \Def{base pairs} $(i,j)$, where
+%$1\leq i<j\leq n$ and for all different $(i,j), (i',j')\in R$: $\{i,j\}\cap\{i',j'\}=\emptyset$ (``degree $\leq$ 1'')---without precluding pseudoknots.
 % We call an RNA structure $R$ \Def{non-crossing}, iff it does not
@@
 
-In a simple \Def{base pair energy model}, the energy of a target structure $R$ is expressed as feature with contributions $f^R_{ij}(\{i\mapsto x, j\mapsto y\})=\EbpSym(x,y)$ for each $(i,j)\in R,$ such that $F^R(S)=\EbpSym(S,R)$.
+Moreover, in a simple \Def{base pair energy model}, the energy of a target structure $R$ is expressed as a feature associated with contributions $f^R_{ij}(\{i\mapsto x, j\mapsto y\})=\EbpSym(x,y)$ for each base pair $(i,j)$ of $R,$ such that $F^R(S)=\EbpSym(S,R)$.
 
@@
 
-Typically, such RNA energy models consider a sequence $S$ \Def{valid} for $R$, iff $\{S_i,S_j\}\in \B$ for
-all $(i,j)\in R$, where $\B:=\left\{\{\Ab,\Ub\},\{\Gb,\Cb\},\{\Gb,\Ub\}\right\}$ describes the valid base pairs.
-
-First, for capturing the minimal
-dependency structure of multi-target RNA design, which is induced by the requirement of canonical base pairing,
+%Typically, such RNA energy models consider a sequence $S$ \Def{valid} for $R$, iff $\{S_i,S_j\}\in \B$ for
+%all $(i,j)\in R$, where $\B:=\left\{\{\Ab,\Ub\},\{\Gb,\Cb\},\{\Gb,\Ub\}\right\}$ describes the valid base pairs.
+
+First, in order to capture the requirement of canonical base pairing for each structure,
 we define the \Def{base pair dependency graph} $G_{\R}$ as the graph with nodes $\{1,\dots,n\}$
 and edges $\bigcup_{\ell\in[1,k]} R_\ell$.
+Any edge between two nodes $i$ and $j$ materializes the existence the existence of a base pair between positions $i$ and $j$ in some structure, inducing a direct depency between the assignments of these two positions.
 
@@
   \label{def:treedecomp}
-  Let $G=(X, E)$ be a hypergraph. A \Def{tree decomposition} of $G$ is
-  a pair $(T,\chi)$, where $T$ is an unrooted tree/forest, and (for
-  each $v\in T$) $\chi(v)\subseteq X$ is a set of vertices assigned to
+  Let $G=(X, E)$ be a (hyper)graph with nodes in $X$ and (hyper)edges in $E$. A \Def{tree decomposition} of $G$ is
+  a pair $(T,\chi)$, where $T$ is an unrooted tree/forest and, for
+  each $v\in T$, $\chi(v)\subseteq X$ is a set of vertices assigned to
   the node $v\in T$, such that
@@
 \begin{itemize}
-\item the transformed contributions $\ln(\pi_0)f^{\GCb}_i$ for the ~$\GCb\%$ feature ($i=1,\dots,n$)
+\item the transformed contributions $\ln(\pi_0)f^{\GCb}_i$ for the ~$\GCb\%$ feature ($i=1,\dots,n$);
 \item the transformed contributions $\ln(\pi_\ell)f^{R_\ell}_{ij}$for each structure $R_\ell\in\R$ and $(i,j)\in R_\ell.$
@@
 We compute the partition function (as specified by $\F$) by dynamic programming based
-on a tree decomposition of $\F$'s dependency graph.
-Note that analogous algorithms could be constructed to count valid sequences or minimize energy.
+on a tree decomposition of $G_{\F}$, the dependency graph associated with $\F$.
+Note that analogous algorithms could be easily derived to count valid sequences, or produce the sequence with minimal free-energy.
 %This yields fixed parameter tractable (FPT) algorithms, based on the treewidth.
@@
 connected and contains a dedicated node $r$, with
-$\chi(r)=\varnothing$ and $\phi(r)=\varnothing$. If such a root does not exist, it can be added to the tree decomposition and connected to one node in each connected component of $T$.
+$\chi(r)=\varnothing$ and $\phi(r)=\varnothing$. If such a root does not exist, it can be added to the tree decomposition and connected to one node in each connected component of $T$;
 \item
-all edges in the tree decompostion are oriented towards this root.
+all edges in the tree decompostion are oriented towards this root;
 \item
@@
 Alg.~\ref{alg:pf}, multiplying the 0-ary messages sent to the root $r$
-yields the total partition function (i.e.~due to proper encoding the partition function of our design problem)
+yields the total partition function (i.e.~due to proper encoding the partition function of our design problem) through
 \begin{math}
@@
 
-By this theorem, Boltzmann sampling in our setting is fixed parameter tractable (FPT) in the tree width $w$ (since the complexity is polynomial for fixed $w$).
+By this theorem, the complexity is polynomial for fixed value of $w$, and Boltzmann sampling in our setting is thus fixed parameter tractable (FPT) in the tree width.
 %
@@
 
-\subsection*{Design for expressive energy models}\label{sec:energy_models}
+\subsection*{Design within expressive energy models}\label{sec:energy_models}
 
@@
 
-So far, we discussed sampling only in base pair energy models, although one is finally interested in targeting energies in realistic energy models like the Turner model or pseudoknot models like {\tt HotKnots}~\citep{Ren2005}.
-Notably, for this purpose, the discussed sampling strategy can be extended in two ways: 1) one can directly sample based on more expressive energy models, 2) sampling in a simple energy model can be used to approximate sampling in more complex models.
-
-\paragraph{Sampling based on more expressive energy features}
+In order to capture realistic energy models like the Turner model or pseudoknot models like {\tt HotKnots}~\citep{Ren2005}, our sampling strategy can be extended in two ways: 1) either by directly sampling based on more expressive energy models; 2) or by sampling in a simple energy model can be used to approximate sampling in more complex models.
+In practice, complex energy models has a strong influence on the tree width of the dependency graph and, in turn computational complexity of our approach.
+Therefore, it is interesting to consider, in addition to the base pair model, other
+stripped-down variants of the nearest neighbor model, which could offer a compromise between the low-complexity base pair energy model and the high-accuracy Turner model.
+
+\paragraph{Exact energy models.}
+A first model, which is particularly promising, is the
+\Def{stacking energy model}. This model assigns only assigns energy
+contributions $\Delta G(x_i,x_j,x_{i+1},x_{j-1})$ to stacks consisting of two nested base pairs $(i,j)$ and $(i+1,j-1)$.
+Within our framework, this energy model is captured by contributions $f_S(\{x_i\mapsto s_i, x_{i+1}\mapsto s_{i+1}, x_{j-1}\mapsto s_{j-1}, x_{j}\mapsto s_{j}\}):=\Delta G(x_i,x_j,x_{i+1},x_{j-1})$ associated with stacks occurring in at least one of the input structures.
+
 Complex \Def{loop-based}
  energy models---e.g.~the Turner model,
-which among others includes energy terms for special loops and dangling
-ends---can be encoded as straightforward extension of our outline of the base pair energy model.
-
-In practice, the energy model has a strong influence on the computational complexity of our approach. Therefore, it is interesting to consider, in addition to the base pair model, other
-stripped-down variants of the nearest neighbor model, which could offer a compromise between the low-complexity base pair energy model and the high-accuracy Turner model. 
-Such a model, which is particularly promising for this purpose, is the
-\Def{stacking energy model}. This model assigns non-zero energy
-contributions only to stacks, i.e. interior loops with closing base
-pair $(i,j)$ and inner base pair $(i+1,j-1)$.
-
-The arity of the introduced functions bounds the
-treewidth of the dependency graph. %(and therefore computational complexity).
-Thus, it is noteworthy that the base pair energy model
-requires only binary functions; the stacking model, only quarternary
-dependencies. This arity is increased in a few cases by the commonly
-used Turner 2004 model~\citep{Turner2009} for encoding tabulated
-special hairpin and interior loop contributions, which depend on up to
-nine bases for the interior loops with a total of 5 unpaired bases
-(``2x3'' interior loops)---all other energy contributions (like
-dangling ends) still depend on at most four bases of the sequence.
-
-\paragraph{Approximating Turner Energy using Simpler Energy Models}
+which, among others, includes energy terms for special loops and dangling
+ends---can also be encoded exactly as instances of our general framework.
+Namely, each loop $L$ involving positions $x_1$,\ldots,$x_p$ will be modeled by a contribution $f_L(\{x_1\mapsto s_1,\ldots, x_p\mapsto s_p\}):=\Delta G(s_1,\ldots,s_p)$, where $\Delta G(s_1,\ldots,s_p)$ is the energy assigned to the loop in the energy model for a given nucleotide content $s_1,\ldots,s_p$.
+Note that the maximum arity of contributions constitutes a lower bound on the treewidth, which may impact the practical complexity of our algorithms.
+For instance, loop contributions in the Turner 2004 model~\citep{Turner2009} may depend on up to
+nine bases for interior loops, with a total of 5 unpaired bases (``2x3'' interior loops)--- although all other energy contributions, including dangling ends, only depend on at most four nucleotides.
+
+%The arity of the introduced functions influences the
+%treewidth of the dependency graph.
+%Thus, it is noteworthy that the base pair energy model
+%requires only binary functions; the stacking model, only quarternary
+%dependencies. This arity is increased in a few cases by the commonly
+%used Turner 2004 model~\citep{Turner2009} for encoding tabulated
+%special hairpin and interior loop contributions, which depend on up to
+%nine bases for the interior loops with a total of 5 unpaired bases
+%(``2x3'' interior loops)---all other energy contributions (like
+%dangling ends) still depend on at most four bases of the sequence.
+
+\paragraph{Approximating Turner Energy using Simpler Energy Models.}
 To capture the realistic Turner model $\EnergyTurner$ more efficiently, we exploit the tight correlation between
@@
 \subsection*{Extension to multidimensional Boltzmann sampling}\label{sec:multiBoltzmann}
-The flexibility of our framework supports the advanced sampling technique "multidimensional Boltzmann sampling"~\citep{Bodini2010}, which allows (probabilistically) enforcing additional, complex properties of the samples.
-This technique was previously used to control \GCb\%~\citep{Waldispuehl2011,Reinharz2013} and dinucleotide content~\citep{Zhang2013} of sampled RNA sequences; here, in addition to controlling \GCb\% (our feature $F_0$) we use it to target the free energies $(\TargetE_1,\ldots,\TargetE_k)$ of the single targets (features $F_1,\dots,F_k$).
+The flexibility of our framework allows to support the advanced sampling technique called "multidimensional Boltzmann sampling"~\citep{Bodini2010}, which allows to enforce (probabilistically)  additional, complex properties of the samples through an additional rejection.
+This technique was previously used to control \GCb\%~\citep{Waldispuehl2011,Reinharz2013} and dinucleotide content~\citep{Zhang2013} of sampled RNA sequences; here, in addition to controlling \GCb\% (our feature $F_0$) we use it to target the free energies $(\TargetE_1,\ldots,\TargetE_k)$ of the individual target structures (features $F_1,\dots,F_m$).
 
@@
 %Such a distribution can be induced by a simple modification of the functions described in Sec.~\ref{sec:energy_models}, where any energy function $E(\val)$ for a structure $\ell$ is replaced by $E'(\val):= \ln(\pi_\ell)\, E(\val)/\beta.$
-Due to the construction of $\F$,
-the probability of a sequence $S$ is thus proportional to
-$\prod_{\ell=0}^{k} \pi_i^{-F_i(S)} = \prod_{f\in\F} f(S).$
-
-One then needs to \Def{learn a weights vector} $\pmb{\pi}$ such that, on average, the targeted energies are achieved by a random sequences in the weighted distribution, i.e.\ such that  $\mathbb{E}(F_\ell(S)\mid \pmb{\pi})=\TargetE_\ell$,  $\forall\ell\in[1,k]$ and, analogously, the expectation of $F_0(S)$ is the targeted GC content.
+%Due to the construction of $\F$,
+%the probability of a sequence $S$ is thus proportional to
+%$\prod_{\ell=0}^{k} \pi_i^{-F_i(S)} = \prod_{f\in\F} f(S).$
+
+One then needs to \Def{learn a weights vector} $\pmb{\pi}$ such that, on average, the targeted energies are achieved by a random sequences in the weighted distribution.
+In other words,  $\mathbb{E}(F_\ell(S)\mid \pmb{\pi})=\TargetE_\ell$,  $\forall\ell\in[1,k]$ and, analogously, the expectation of $F_0(S)$ is the targeted GC content.
 The expected value of $F_\ell$ is always decreasing for increasing weights $\pi_\ell$ (see Supp.
@@
 More generally, computing a suitable parameter vector $\pmb{\pi}$ can be restated as a convex optimization problem, and be efficiently solved using a wide array of methods~\citep{Denise2010,Bendkowski2017}.
+
 In practice, we use a simple heuristics which starts from an initial weight vector $\pmb{\pi}^{[0]}:=(e^\beta,\dots,e^\beta)$ for $\beta=1/(RT)$, T=$37^\circ$, and gas constant $R$.
@@
 Therefore, since $I$ is an independent set, the edge $(v_1,v_2)\in E$ cannot exist.
+\QEDA
 \end{itemize}
@@
 
-The driver script of \ourprog{} takes as input a set of (secondary) structures, ... and outputs ....
-
-We implemented the core algorithms in C++, resulting in the tool
-C++-tool \ourprog{}, which takes a set of target structures, as well as
-weights for each energy feature and \GCb\%; then,
-it generates a sample (of prescribed size) from the corresponding
-Boltzmann distribution. For \ourprog{}, we implemented the base pair
+We implemented the core algorithms in \texttt{C++}, resulting in the tool
+\ourprog{}, which takes a set of target structures, as well as
+weights for each energy feature and \GCb\%, and generates a sample set of sequences compatible with the structures in the corresponding
+Boltzmann distribution. \ourprog{} currently supports the base pair
 energy model and the stacking energy model.
-Moreover, we provide a Python wrapper script for the C++-tool
-\ourprog{} to perform targeting, based on multi-dimensional Boltzmann
-sampling.
-For given secondary structures, together with prescribed target energies
-and target \GCb\%, this script generates
-a series of sequences that satisfy the target values for the energy and
+Moreover, we provide a {\tt Python} wrapper script that targets prescribed energies using multi-dimensional Boltzmann sampling.
+For a given set of secondary structures, together with prescribed target energies
+and target \GCb\%, this script generates a series of sequences that satisfy the target values for the energy and
 \GCb\% features within configurable tolerances.
 Notably, these target energies are actual free energies in the realistic
-Turner energy model, which are targeted by efficiently sampling in the
-base pair energy model.
+Turner energy model, and are targeted by efficiently sampling in the
+base pair energy model, and filtering sequences based on the {\tt RNAeval} tool from the Vienna package~\cite{Lorenz2011}.
 
 \subsection*{Effectively targeting Turner energies}
-We implemented our Boltzmann sampling strategy (Algorithms
+We used our Boltzmann sampling strategy (Algorithms
 \ref{alg:pf} and \ref{alg:sampling}), to sample valid sequences for
@@
 
+One could also envision using positive design as a null model, for instance in order to assess the significance of observed properties.
+Indeed, the complex set of constraints required in order to achieve a desired function must be equally enforced in RNA design, and to produce random sequences that are known to be constrained at a phenotypic level.
+One could therefore use constrained design to assess whether P-values, and therefore our perception of the statistical significance, are affected by the overly simplistic simple null models (\emph{e.g.} dinucleotide shuffling) used to model random RNAs~\cite{Rivas2017}.
+
 {Stefan: \em Mention RNAsketch to perspectives for more complex design scenario + need for generic constraints-based framework for flexibility ($\to$ generic constraint framework).}
-
-{Yann: \em Connection with statistics (design as perfect random sequences for null-model), mention Rivas/Eddy lncRNAs papers...}
 
@@
 
-\section*{Figures}
+%\section*{Figures}
 %   \begin{figure}[h!]

### 2018-10-08 RNARedPrint.tex
sync

@@
 \section*{Results}\label{sec:results}
-Our Boltzmann sampling strategy (Algorithms
-\ref{alg:pf} and \ref{alg:sampling}) was implemented, using a combination of {\tt C++} and {\tt Python}, in a software called \ourprog{}, available at:
+
+We implemented the core algorithms in \texttt{C++}, resulting in the tool
+\ourprog{}, available at:
 
@@
 
-We implemented the core algorithms in \texttt{C++}, resulting in the tool
-\ourprog{}, which takes a set of target structures, as well as
+\ourprog{}  takes a set of target structures, as well as
 weights for each energy feature and \GCb\%, and generates a sample set of sequences compatible with the structures in the corresponding
@@
 energy model and the stacking energy model.
+
 Moreover, we provide a {\tt Python} wrapper script that targets prescribed energies using multi-dimensional Boltzmann sampling.
@@
 %for RNA design.
-Using \ourprog{}, we generated at least 1\,000 seeds for
-each instance of the benchmark. We used target energies that were
-computed by averaging the energy over samples at higher weights
-$e^{\beta}$, and set the targeted \GCb\% to 60\%.  We compared the
+Using \ourprog{}, for
+each instance of the benchmark, we generated at least 1\,000 seed sequences with similar
+energies for all target structures. For this purpose, we determine good common target energies that
+can be succesfully targeted for all single target structures simultaneously.
+Generally, we targeted 60\% \GCb\%.  We compared the
 \Obj{} value of these sequences against that of seed sequences,

### 2018-10-23 RNARedPrint.tex
sync from overleaf

@@
 \author[
-    addressref={aff4,aff5},
-    corref={aff4},                       % id of corresponding address, if any
-    email={yann.ponty@lix.polytechnique.fr}
+    addressref={aff4},
+    corref={aff2,aff4},                       % id of corresponding address, if any
+    email={will@tbi.univie.ac.at; yann.ponty@lix.polytechnique.fr}
 ]{\inits{YP} \fnm{Yann} \snm{Ponty}}
@@
 % $^{\text{\sf 5}}$ AMIBio team, Inria Saclay, Bat Alan Turing, 91120 Palaiseau, France}
-\address[id=aff5]{%
-    \orgname{AMIBio team, Inria Saclay},
-    \street{Bat. Alan Turing},
-    \postcode{91120}
-    \city{Palaiseau},
-    \cny{France}
-}
 
@@
 
-Moreover, we provide a {\tt Python} wrapper script that targets prescribed energies using multi-dimensional Boltzmann sampling.
+Moreover, we provide two {\tt Python} wrapper scripts. The first script targets prescribed energies using multi-dimensional Boltzmann sampling.
 For a given set of secondary structures, together with prescribed target energies
@@
 base pair energy model, and filtering sequences based on the {\tt RNAeval} tool from the Vienna package~\cite{Lorenz2011}.
+The second script generates high quality seed sequences suitable for negative RNA design.
+The details of this approach are described in the subsections below.
 
@@
 
-{Stefan: \em Mention RNAsketch to perspectives for more complex design scenario + need for generic constraints-based framework for flexibility ($\to$ generic constraint framework).}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \section*{Author's contributions}
-  YP and SW initiated the project and worked out the algorithms.
+  YP and SW initiated the project and designed the algorithms.
   % SW---I guess we don't need to be super-detailed, otherwise something like...
@@
   %SW contributed the integration with the constraint network setting and its specialization to RNA design.
-  YP and WW wrote the core software, while SW and SH contributed additional scripts. SH devised and conducted the computational experiments. All authors wrote, read, and approved the manuscript.
+  YP, SW and WW wrote the core software, while SW and SH contributed the weight optimization scripts.
+SH, SW and YP devised the computational experiments, which were conducted by SH.
+All authors wrote, read, and approved the manuscript.
     % Text for this section \ldots

### 2018-12-15 BMC-Response-Round1.tex
Added response

% !TEX TS-program = pdflatex
% !TEX encoding = UTF-8 Unicode

% This is a simple template for a LaTeX document using the "article" class.
% See "book", "report", "letter" for other types of document.

\documentclass[11pt,hyperref]{article} % use larger type; default would be 10pt

\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)

%%% Examples of Article customizations
% These packages are optional, depending whether you want the features they provide.
% See the LaTeX Companion or other references for full information.

%%% PAGE DIMENSIONS
\usepackage{geometry} % to change the page dimensions
\geometry{a4paper} % or letterpaper (US) or a5paper or....
% \geometry{margin=2in} % for example, change the margins to 2 inches all round
% \geometry{landscape} % set up the page for landscape
%   read geometry.pdf for detailed page layout information

\usepackage{graphicx} % support the \includegraphics command and options

% \usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent

%%% PACKAGES
\usepackage{amsmath} % for much better looking tables
\usepackage{xspace}
\usepackage{url}
\usepackage{soul}
\usepackage{xcolor}
\usepackage{amssymb} % for much better looking tables
\usepackage{booktabs} % for much better looking tables
\usepackage{array} % for better arrays (eg matrices) in maths
\usepackage{ntheorem} % for better arrays (eg matrices) in maths
\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim
\usepackage{subfig} % make it possible to include more than one captioned figure/table in a single float
\usepackage{framed}
% These packages are all incorporated in the memoir class to one degree or another...

%%% HEADERS & FOOTERS
\usepackage{fancyhdr} % This should be set AFTER setting up the page geometry
\pagestyle{fancy} % options: empty , plain , fancy
\renewcommand{\headrulewidth}{0pt} % customise the layout...
\lhead{}\chead{}\rhead{}
\lfoot{}\cfoot{\thepage}\rfoot{}

%%% SECTION TITLE APPEARANCE
\usepackage{sectsty}
\allsectionsfont{\sffamily\mdseries\upshape} % (See the fntguide.pdf for font help)
% (This matches ConTeXt defaults)

%%% ToC (table of contents) APPEARANCE
\usepackage[nottoc,notlof,notlot]{tocbibind} % Put the bibliography in the ToC
\usepackage[titles,subfigure]{tocloft} % Alter the style of the Table of Contents
\renewcommand{\cftsecfont}{\rmfamily\mdseries\upshape}
\renewcommand{\cftsecpagefont}{\rmfamily\mdseries\upshape} % No bold!

\newcommand{\Answer}[1]{\noindent{\color{gray!55!black}\textsf{\ul{Response:} }{\sf#1}}}
\newcommand{\Comment}[1]{\item{\textsf{\ul{Comment:} }{\it#1}}}
\newcommand{\AnswerTODO}[2][All]{\noindent{\color{blue!80!black}\textsf{\ul{TODO Respond (#1):} }{\sf#2}}}

%%% END Article customizations

%%% The "real" document content comes below...

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors here                   %%
%%                                          %%
%% Specify information, if available,       %%
%% in the form:                             %%
%%   <key>={<id1>,<id2>}                    %%
%%   <key>=                                 %%
%% Comment or delete the keys which are     %%
%% not used. Repeat \author command as much %%
%% as required.                             %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{Response to reviewers\\[.3em]Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures}
\author{
Stefan Hammer \and Wei Wang \and Sebastian Will \and Yann Ponty}

\date{} % Activate to display a given date or no date (if empty),
         % otherwise the current date is printed

\begin{document}
\maketitle

First of all, the authors wish to express their gratitude to the reviewers for their encouraging comments and constructive criticisms. We tried to address these points to the best of our abilities in this revised version of the manuscript.
\section{General response}

\section{Detailed responses}

\subsection{Reviewer \#1}
\begin{enumerate}
\Comment{While this manuscript exhaustively described the algorithms of RNARedPrint and its advancement from theoretical aspects, it would be hard for BMC bioinformatics readers to understand what are the sophisticated points applied in RNARedPrint for a practical use.
Related to that problem, it is also not clear what methods (or what options or approximations) are applied for each comparison in the results.
Thus, this manuscript is considered to be further improved in terms of intelligibility of the methods and results using fair evaluation criteria.\\[1em]}%
\AnswerTODO[Yann]{...}
\Comment{For evaluation of a method to design RNA molecules, the number of the solutions it produces and its running time are considered to be important.\\[1em]}%
\Answer{Indeed, this is a relevant parameter to ensure a fair comparison. We note, however, that one must distinguish the number of sequences produced as the final output of our method from the total number of sequences generated by our rejection sampling strategy, the large majority of which are discarded for not satisfying certain properties. While this may appear wasteful, this probabilistic approach saves very substantial amount of time (and memory) compared to natural alternatives (classified dynamic programming, or constraint-based systems), at the price of some rejections, expected to remain reasonable from general analytic combinatorics principles.}
\Comment{
In the scope of this study, the running time and memory comparison should be shown to produce the same number of sequences by some minor variants of RNARedPrint method, RNABluePrint, and MODENA.\\[1em]}%%
\AnswerTODO[Stefan]{
We did not fully understand this remark, and interpreted it to mean that a fair comparison should dedicate equal amount of runtime to each of the compared method. 
(Stefan, is that the case?)
In the revised version, we reported the runtime required by all those approaches in the setting of the benchmark.
}
\Comment{
It would be informative if the authors show the number of rejected sequences produced by RNARedPrint considering the specific criteria (GC\% or free energy). 
Since RNARedPrint keeps the sequences to be under the specific constraint by a simple rejection method, such information for various target conditions would be better to estimate the practical running time of RNARedPrint.\\[1em]}%
\AnswerTODO[Sebastian]{
We now report the frequency of rejections measured under various conditions. (Further comments...)
}
\Comment{
Another is the quality of their solutions in terms of the stability of each target structure.\\[1em]}%
\AnswerTODO[Seb+Stefan]{ Can we report that? If so, why not...
}
\Comment{
According to their results, adaptive sampling has a strong power to improve the results. Unfortunately, the authors only introduce the existence of weights first in the first section of methods.
It would be better that the authors introduce the concept of weights at the beginning (and whether they are completely same with those in RNAblueprint or not).\\[1em]}%
\AnswerTODO[Yann]{
Weights are used for two different purposes:
\begin{itemize}
    \item In order to induce a Boltzmann distribution on the set of sequences;
    \item As a technical artifact, to ensure the efficiency of a rejection-based strategy to control auxiliary features (free-energies for structures, GC\%\ldots)
\end{itemize}
}
\Comment{
Moreover, it is not well described how we can choose the initial weights and what values they are converged in the actual cases (e.g., Fig2 and S4) after adaptive sampling.\\[1em]}%
\AnswerTODO[Stefan]{
The calibration of weights relies on a numerical iteration aiming at computing weights such that expected properties of sampled sequences (which depend on weights) match their targeted values. Associated problems are usually convex, and we describe in supplementary material a numerical iteration to compute suitable weights.
We also added to Fig 2 and 4 the final weights.
}
\Comment{
The influence of long loops and stacking should be measured for the sequences having different GC contents and length. Moreover, the number of base pairs in target structures should be considered since such approximation might be critical when the long loops appear in the target structures.
\\[1em]}%
\AnswerTODO[Yann]{Our approximation only hold for a single structure. Long loops tend to be oblivious of their content so, while approximating the energy model for long loops would be a problem in the context of structure prediction, they are not in the context of design. (to be clarified/elaborated)

}
\Comment{
$R^2$ values and $p$-values are also necessary to evaluate the correlation between two energy models. (What are the lines in Fig2A representing?)
\\[1em]}%
\AnswerTODO[Stefan+Sebastian]{
Compute $R^2$ (detaching plot). Write paragraph about goodness of fit
}
\Comment{
Through the analyses, the names of energy models should be specified more clearly. The original model, simplified base paired model, and stacking model are confusing. Remove base-paired (simplified Turner).
\\[1em]}%
\AnswerTODO[Yann]{
To avoid any confusion, we moved the paragraph describing the integration of various energy models to the appendix. We now only describe our simplified stacking model, which we more explicitly show to be highly correlated with the Turner energy model.
}
\Comment{
Multi defect can be a good measure to understand the overall performance of the sequence designing method, but there is little explanation why Multidefect is enough to evaluate the quality of sequences for the multi (target-structure) objective problem in a practical use case such as Riboswitch design.
\\[1em]}%
\AnswerTODO[Yann]{Not really our job, but we can maybe remind some of the key arguments.
}
\Comment{
Although RNABluePrint was compared with MODENA in the previous study, RNABluePrint did not overwhelm MODENA in all of conditions (and it is unclear which version (variant) of RNABluePrint is applied in this study).
For that reason, RNARedPrint still should be compared with MODENA (in the same running time) using Multidefect and the criteria used in the previous studies such as $\delta$e1 and $\delta$e2. Also, RNARedprint and RNABlueprint are optimised to increase only a Multidefect value.\\[1em]}%
\AnswerTODO[Stefan]{First of all, it is perhaps worth reminding that RNAredprint is not intended as a tool for negative design, but rather a good seed generation algorithm, amenable to further optimization using other tools. For this reason, we focused our validation on this potential for better initialization, and compared the quality of sequences generated by RNAredprint (Boltzmann weighted) to RNAblueprint (uniform). We noticed that, while RNAredprint is not explicitly tailored to optimize MultiDefect, it nevertheless produces sequences that: 1) achieve better values; and 2) are amenable to better further optimizations than the uniform version implemented by RNABluePrint.
}

\Comment{
Hence, the distribution of Multidefect, previous criteria, and the first and second term of Multidefect should be shown in box plots or violin plots to know the robustness of RNARedPrint It is also interesting what is the characteristic of the sequence designed by each method in a specific example of functional RNA molecules. \\[1em]}%
\AnswerTODO[Yann]{Again, we do not adhere to such a benchmark-driven view of design tasks. The time of design is generally negligible compared to the time taken to experimentally assess the ultimate functionality of a design. What we believe to matter more to the community is the expressivity of the tools, allowing to address various use-cases.
}
\Comment{(Is RNARedPrint able to produce, for example, the combination of two long stem loops which are high and low GC-content, respectively?)\\[1em]}%
\AnswerTODO[Yann]{In the current implementation, it is only possible to constrain the global GC\% of the sequence. It is however possible to support this feature within our general framework, since one can associated individual objectives to structures (eg  structures are already associated with individual target energies). Different stems can be expressed as different structures by "peeling the structure", so this feature could be added to a future version, should the concrete need arise.
}
\Comment{
The section "problem statement" is not helpful enough for readers, both of algorithmic and biological people.\\[1em]}%
\AnswerTODO[Yann]{We attempted to clarify it.
}
\Comment{
In particular, it is difficult to comprehend the relationships between the variables due to insufficient descriptions and lack of definitions.
For example, they do not specify any relationship between R and F in this section while they should not be independent. (e.g., m should be equal or less than k?)
Moreover, I cannot find a definition of several variables. (e.g., What is $S$, $S_i$, $x$, $xp$, $xd$, $vp$, $bu{chi}$ (P9L24) ? Where "its dependence dep $f$" comes from?)
Because this manuscript contains many variables and abbreviations, it is recommended to have a table or image to explain the variables and their actual examples following the sampling example in Fig1. The description of the brief aim and summary of each step would be a good guide for readers.\\[1em]}%
\AnswerTODO[Yan]{
Add method overview
}
\Comment{
* Ensemble free energy: might be an ambiguous definition\\[1em]}%
\Answer{The concept of ensemble free-energy is a well-defined concept in statistical mechanics, and is used here in a way that, to the best of our knowledge, matches its usual definition.}
\Comment{
* While $\chi$ is a subset of nodes $X$, is a subset of $\chi$ an edge? (P7 L19)\\[1em]}%
\AnswerTODO{
??
}
\Comment[Sebastian :)]{
* Is the output of algorithm 2 only one sequence? It is unclear when two (or more?) sequences are merged by $S \cup S'$.\\[1em]}%
\AnswerTODO{???
}
\Comment{
Is it correct no break at P10L23?\\[1em]}%
\AnswerTODO{
??
}
\Comment{
* In Fig2, the notation like $(a, b)$ is confusing with the range (of free energies or something).\\[1em]}%
\AnswerTODO{
}
\Comment{
* Unfortunately, I could not run RNARedPrint (while it can be just compiled when I make bin directory). More detailed documentation is required such as OS dependency or the version information the authors confirmed.\\[1em]}%
\AnswerTODO[Yanns]{
Yann will check
}
\Comment{
* It seems not trivial that Boltzmann sampling works better to control GC contents, but the stability of structures. What's the reason the authors include GC contents for the criteria of Boltzmann-weighted sampling?\\[1em]}%
\Answer{GC\%, just like the free-energy of any given target structure, can be expressed uniformly in our framework in a way that is additive on base pairs. Control over GC\% has recurrently been identified as critical, since a good stability usually implies high GC\%, yet actual sequences usually combine good stability with reasonable GC\%. Modern computational objectives of design are to ensure good stability (+specificity) while producing sequences of prescribed GC\%\\[1em]}%
\AnswerTODO{
}
\Comment{
* The order of authors is different between the main manuscript and supplementary (not sure whether they used just an alphabetical order).\\[1em]}%
\AnswerTODO{
?
}
\Comment{
* Is the section "\# P-hardness of counting valid designs" necessary in the main script (not supplementary) to explain the results? The complexity of the problem is important to consider theoretical limitations, but the assumption in that section seems not to be necessary since the authors applied a greedy sampling strategy and not in the scope of this study.\\[1em]}%
\Answer{we respectfully disagree with that assessment.
The core of our approach consists in generating, uniformly at random (resp. in a Boltzmann distribution, which identifies with the uniform distribution when $T\to +\infty$). Such a generation algorithm usually starts by precomputing some numbers of solutions (here, sequences that satisfy the classic base pairing rules) later to be used to ensure the uniformity of the generation through stochastic choices. The precomputation step  requires an efficient algorithm for counting the number of solutions. In this work, we provide an {\bfseries exact} algorithm (not a heuristics) for the problem which, while being FPT, still requires exponential time when the tree-width increases with the sequence length. This raises the natural question: Can a truly polynomial algorithm exist for the problem? Our proof of \#P-hardness provides a negative answer to this problem under fairly reasonable assumptions. We tried to clarify those points in our introduction of the \#P hardness result.
}
\Comment{
* Move the first section in the result to the method.\\[1em]}%
\AnswerTODO[Yann]{
Ok (but include efficiency of multidim sampling, keeping correlation analysis)
}
\end{enumerate}
\subsection{Reviewer \#2}
\begin{enumerate}
\Comment{
1) The "Methods" section is very long, it should be shortened.\\[1em]}%
\AnswerTODO[Yann]{
Well, some methods require longer descriptions ;)
Move expression of various energy models as constraints Turner to appendix
Move numerical weight optim to appendix
Better motivations for \#P-hardness
New results : 
  1) Validation of stacking approximation: Correlation coefficient of Stacking and Turner (+ goodness of fit)
  2) Rejection analysis for multidim sampling
  3) Tree width and Time benchmark
  5) Proof-of-concept -Potential of better seeds
}
\Comment{
2) Although, it was shown previously, that RNABluePrint is a leading tool for multiple design, it would be valuable to include and compare the results generated by the other tools, e.g. Frnakenstein.\\[1em]}%
\AnswerTODO[Stefan+Sebastian]{
The three currently available tools are Frnakenstein, RNABluePrint and Modena. Shall we do that?
}
\Comment{
3) Please include computational times of the performed designs with the given benchmark.\\[1em]}%
\AnswerTODO[Stefan]{
Will do
}
\Comment{
4) The paper should be checked for English grammar and wording.\\[1em]}%
\AnswerTODO{
Will do
}
\end{enumerate}
\subsection{Reviewer \#3}
\begin{enumerate}
\Comment{
In Figure 3 of Supplementary Material: more detailed description for explaining the two examples of tree decomposition is appreciated, particularly for the readers who are not familiar with tree decomposition; e.g. why these two examples are output, and the other examples for RF3 exist or not?\\[1em]}%
\AnswerTODO[Sebastian]{
Discussion
}
\Comment{
In "L Detailed result of MultiDefect Analysis" of Supplementary Material, running time is missing. Since the Boltzmann sampling in RNARedPrint utilizes the rejection approach, I would like to know the difference between the running times of the Boltzmann sampling and the uniform sampling. \\[1em]}%
\AnswerTODO[Yann]{ Added runtimes.
None, but for multidim sampling, rejection and weight optimization stack up (we report them).
}
\Comment{
In addition, I think it is better that the N/A result is indicated not by a blank but by a "N/A" in the table.\\[1em]}%
\AnswerTODO[Stefan/Sebastian]{
Will add results times}
\end{enumerate}
%\bibliographystyle{amsplain}
%\bibliography{response}
\end{document}

### 2018-12-19 BMC-Response-Round1.tex
Update response letter

@@
 
-First of all, the authors wish to express their gratitude to the reviewers for their encouraging comments and constructive criticisms.
-We tried to address these points to the best of our abilities in this revised version of the manuscript.
-\section{General response}
+We thank the reviewers for their encouraging comments and constructive criticisms and thoroughly addressed their points in our revised version of the manuscript.
+
+\medskip
+In particular, the reviews helped us to state the goal of this work more clearly.
+While we demonstrate the applicability of our approach to negative RNA design, which (here by optimizing 'MultiDefect') aims at high and similar probability of the target structures, the focus of this work is the positive design of sequences that fold into multiple target structures with highly specific properties.
+We discuss this positive design, in particular, for specific folding energies (for each single target structure) and GC content.
+Moreover, as we like to emphasize, the method is even more general, which we see as most important strength of our approach; it is prepared to handle many additional constraints and sequence features based on the same mechanism.
+
+Naturally, we hope that this method turns out to be highly useful for single- and multi-target negative design (and various other applications, like generating background models for assessing significance) in future applications, but---while we present preliminary results for negative design (based on generating seed sequences for the MODENA benchmark)---a more in-depth evaluation of its use in negative design is beyond the scope of this paper.
+This seems even more acceptable, since the principal benefits of controlled sampling have been shown with RNABlueprint.
+We apologize for potential misunderstandings about the objectives of the manuscript and hope to better clarify these aspects in our revision.
+
+To emphasize our focus on positive design and the prototypical nature of our brief proof-of-concept study in the use of our method for negative design, we carefully rephrased the respective paragraph at the end of the introduction as well statements in the subsection about the optimization of the MultiDefect.
+(TODO: write!)
 
@@
 Thus, this manuscript is considered to be further improved in terms of intelligibility of the methods and results using fair evaluation criteria.\\[1em]}%
-\AnswerTODO[Yann]{...}
+\AnswerTODO[Yann]{We structured the results section more clearly and adapted the introduction accordingly.
+In Results, we introduce single subsections for the following sub-results to reflect our main results in a logical order:
+  \begin{itemize}
+    \item The full-featured RNA energy model (Turner) can be well approximated by the much more efficiently tractable stacking model.
+    \item Controlled Boltzmann sampling for typical multi-target design instances requires little time (corresponding to small tree-width).
+    \item Multi-dimensional Boltzmann sampling solves the positive design problem effectively, requiring only a limited number of rejections (even for multiple simultaneously considered features).
+    \item The positively designed sequences are suitable for improving negative design approaches.
+\end{itemize}}
+
 \Comment{For evaluation of a method to design RNA molecules, the number of the solutions it produces and its running time are considered to be important.\\[1em]}%
-\Answer{Indeed, this is a relevant parameter to ensure a fair comparison.
+\Answer{We agree that running times are important for assessing the method; consequently, we included details timing information in the revision.
+As we show now explicitly, the running times directly depend on the already reported tree-widths.
+The run-times are generally modest for the relevant design instances.}
+
+%\Answer{Indeed, this is a relevant parameter to ensure a fair comparison.
 We note, however, that one must distinguish the number of sequences produced as the final output of our method from the total number of sequences generated by our rejection sampling strategy, the large majority of which are discarded for not satisfying certain properties.
@@
 \AnswerTODO[Stefan]{
+  We do now report and discuss the run-times of such methods for generating samples.
+Since MODENA does not allow to measure separate times for sequence sampling, we estimated those times and report our estimates.
+It should be emphasized that MODENA produces a very different kind of sampling compared to RNARedPrint (and as well RNABluePrint), such one should be careful to compare only run-times.
+While RNARedPrint produces a very specific distribution of samples, solving a hard positive design problem, MODENA
+
   We did not fully understand this remark, and interpreted it to mean that a fair comparison should dedicate equal amount of runtime to each of the compared method. 

### 2018-12-19 RNARedPrint.tex
Update response letter

@@
 
-\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
-%\documentclass{bmcart}
+%\documentclass[doublespacing,linenumbers]{bmcart}% uncomment this for twocolumn layout and comment line below
+\documentclass{bmcart}
 
@@
 
+\newcommand{\revised}[1]{\textcolor{red}{#1}}
+
 %%% end macro defs
@@
 
-By these definitions, the set $\F$ encodes the partition function $\partfun{\pi_0,\dots,\pi_k}$ of Eq.~(\ref{eq:mainproblem}); also confer Eq.~(\ref{eq:feature-energy-transformation}). 
+By these definitions, the set $\F$ encodes the partition function $\partfun{\pi_0,\dots,\pi_k}$ of Eq.~(\ref{eq:mainproblem}).
 %Subsequently, we show how to efficiently compute this partition function based on the 
@@
 As a baseline for comparison, we considered \RNAblueprint~\citep{Hammer2017}, the current leading tool for multiple design.
-As a quality measure, we applied the objective function that was introduced by~\citet{Hammer2017} for multi-stable design, defined as:
+As a quality measure, we applied the objective function introduced by~\citet{Hammer2017} \revised{based on \cite{Flamm2001,HoenerzuSiederdissen2013}} for multi-stable design, defined as:
 % \begin{align}
@@
 
+\section*{Availability of data and materials}
+The software and data of this work are available in the Github repository \url{https://github.com/yannponty/RNARedPrint}.
+
 \section*{Author's contributions}

### 2018-12-19 BMC-Response-Round1.tex
two columns, as it ought to be :)

--- 2018-12-15 BMC-Response-Round1.tex
@@
 \newcommand{\Comment}[1]{\item{\textsf{\ul{Comment:} }{\it#1}}}
-\newcommand{\AnswerTODO}[2][All]{\noindent{\color{blue!80!black}\textsf{\ul{TODO Respond (#1):} }{\sf#2}}}
+
+\usepackage{todonotes}
+\newcommand{\AnswerTODO}[2][All]{{\todo[inline]{\ul{TODO (#1)}: #2}}}
 
@@
 
-First of all, the authors wish to express their gratitude to the reviewers for their encouraging comments and constructive criticisms.
-We tried to address these points to the best of our abilities in this revised version of the manuscript.
-\section{General response}
+We thank the reviewers for their encouraging comments and constructive criticisms and thoroughly addressed their points in our revised version of the manuscript.
+
+\section{General comments}
+
+In particular, the reviews helped us to state the goal of this work more clearly.
+While we demonstrate the applicability of our approach to negative RNA design, which (here by optimizing 'MultiDefect') aims at high and similar probability of the target structures, the focus of this work is the positive design of sequences that fold into multiple target structures with highly specific properties.
+We discuss this positive design, in particular, for specific folding energies (for each single target structure) and GC content.
+Moreover, as we like to emphasize, the method is even more general, which we see as most important strength of our approach; it is prepared to handle many additional constraints and sequence features based on the same mechanism.
+
+Naturally, we hope that this method turns out to be highly useful for single- and multi-target negative design (and various other applications, like generating background models for assessing significance) in future applications, but---while we present preliminary results for negative design (based on generating seed sequences for the MODENA benchmark)---a more in-depth evaluation of its use in negative design is beyond the scope of this paper.
+This seems even more acceptable, since the principal benefits of controlled sampling have been shown with RNABlueprint.
+We apologize for potential misunderstandings about the objectives of the manuscript and hope to better clarify these aspects in our revision.
+
+To emphasize our focus on positive design and the prototypical nature of our brief proof-of-concept study in the use of our method for negative design, we carefully rephrased the respective paragraph at the end of the introduction as well statements in the subsection about the optimization of the MultiDefect.
+(TODO: write!)
+
+\todo[inline]{Mention that our sampling could be used in combination with MODENA / not necessarily competition}
 
@@
 Thus, this manuscript is considered to be further improved in terms of intelligibility of the methods and results using fair evaluation criteria.\\[1em]}%
-\AnswerTODO[Yann]{...}
-\Comment{For evaluation of a method to design RNA molecules, the number of the solutions it produces and its running time are considered to be important.\\[1em]}%
-\Answer{Indeed, this is a relevant parameter to ensure a fair comparison.
-We note, however, that one must distinguish the number of sequences produced as the final output of our method from the total number of sequences generated by our rejection sampling strategy, the large majority of which are discarded for not satisfying certain properties.
-While this may appear wasteful, this probabilistic approach saves very substantial amount of time (and memory) compared to natural alternatives (classified dynamic programming, or constraint-based systems), at the price of some rejections, expected to remain reasonable from general analytic combinatorics principles.}
-\Comment{
-In the scope of this study, the running time and memory comparison should be shown to produce the same number of sequences by some minor variants of RNARedPrint method, RNABluePrint, and MODENA.\\[1em]}%%
-\AnswerTODO[Stefan]{
-We did not fully understand this remark, and interpreted it to mean that a fair comparison should dedicate equal amount of runtime to each of the compared method. 
-(Stefan, is that the case?)
-In the revised version, we reported the runtime required by all those approaches in the setting of the benchmark.
-}
+\Answer{We structured and rephrased the Results section more clearly and adapted the introduction accordingly.
+In Results, we introduce single subsections for the following sub-results to reflect our main results in a logical order:
+  \begin{itemize}
+    \item The full-featured RNA energy model (Turner) can be well approximated by the much more efficiently tractable stacking model;
+    \item Controlled Boltzmann sampling for typical multi-target design instances requires little time (corresponding to small tree-width); Multi-dimensional Boltzmann sampling solves the positive design problem effectively, requiring only a limited number of rejections (even for multiple simultaneously considered features);
+    \item The positively designed sequences are suitable for improving negative design approaches.
+\end{itemize}}
+\AnswerTODO{Restructure Results and Intro)}
+
+\Comment{For evaluation of a method to design RNA molecules, the number of the solutions it produces and its running time are considered to be important.\\[1em]
+In the scope of this study, the running time and memory comparison should be shown to produce the same number of sequences by some minor variants of RNARedPrint method, RNABluePrint, and MODENA.\\[1em]}%
+\Answer{We agree that running times are important for assessing the method; consequently, we included details timing information in the revision.
+As we show now explicitly, the running times directly depend on the already reported tree-widths.
+The run-time requirements are generally modest for the relevant design instances.
+%
+Concerning the comparison: since MODENA does not allow to measure separate times for sequence sampling, we could only estimate its run-times by running the MODENA binary with parameters that turn off the optimization as far as this is possible.
+Moreover, we want to re-emphasize that MODENA, while it implements a sophisticated optimization procedure, does not put any specific focus on the sampling procedure for generating its seed sequences.
+Consequently, it does not sample from a specific distribution and produces random sequences without targeting any specific properties (neither good stability nor specific GC-content, etc.).
+In contrast, RNARedPrint produces highly specific samples, thereby solving a fundamentally different problem (i.e.
+positive design).
+This crucial difference should kept in mind, when comparing the run-times of these tools.
+\AnswerTODO{...}}
 \Comment{
@@
 \AnswerTODO[Sebastian]{
-We now report the frequency of rejections measured under various conditions. (Further comments...)
+We now report the frequency of rejections measured under various conditions. +add plots (moving distributions over rounds...)
 }
@@
 Another is the quality of their solutions in terms of the stability of each target structure.\\[1em]}%
-\AnswerTODO[Seb+Stefan]{ Can we report that? If so, why not...
-}
+\Answer{We were unsure how to interpret this remark: RNARedPrint enforces an explicit control on target free-energies through multidimensional Boltzmann sampling, so it cannot be used as an external validation of the quality of our output.
+Moreover the final targeted free energies differ greatly across structures across our various benchmark, so one would have to report them on a per sequence basis, leading to a result that would be hard to interpret.
+Finally, note that the MODENA multidefect objective function already depends critically from achieving low, consistent, energies, and already arguably captures this notion of stability.}
 \Comment{
@@
 It would be better that the authors introduce the concept of weights at the beginning (and whether they are completely same with those in RNAblueprint or not).\\[1em]}%
-\AnswerTODO[Yann]{
+\Answer{
 Weights are used for two different purposes:
@@
 }
+\AnswerTODO[Yann]{
+  Introducing gently the concept of adaptive sampling and weights in intro + Diplomatic answer
+}
 \Comment{
 Moreover, it is not well described how we can choose the initial weights and what values they are converged in the actual cases (e.g., Fig2 and S4) after adaptive sampling.\\[1em]}%
-\AnswerTODO[Stefan]{
+\Answer{
 The calibration of weights relies on a numerical iteration aiming at computing weights such that expected properties of sampled sequences (which depend on weights) match their targeted values.
 Associated problems are usually convex, and we describe in supplementary material a numerical iteration to compute suitable weights.
-We also added to Fig 2 and 4 the final weights.
+We also added to Fig. 2 and 4 the final weights.
 }
@@
 (to be clarified/elaborated)
-
 }
@@
 \Comment{
-Through the analyses, the names of energy models should be specified more clearly.
-The original model, simplified base paired model, and stacking model are confusing.
-Remove base-paired (simplified Turner).
+Through the analyses, the names of energy models should be specified more clearly. The original model, simplified base paired model, and stacking model are confusing.
 \\[1em]}%
-\AnswerTODO[Yann]{
-To avoid any confusion, we moved the paragraph describing the integration of various energy models to the appendix.
-We now only describe our simplified stacking model, which we more explicitly show to be highly correlated with the Turner energy model.
-}
+\Answer{While we believe that the capacity of our framework to describe various energy models is important, and demonstrates its versatility and potential for future developments, it is true that only the stacking model is absolutely necessary to understand the method.
+To avoid any further confusion, we moved the paragraph describing the integration of various energy models to the appendix.
+We now only describe our simplified stacking model, which we more explicitly show to be highly correlated with the Turner energy model.}
+\AnswerTODO[Sebastian]{Check that BP model does not appear too recurrently in manuscript}
 \Comment{
@@
 \\[1em]}%
-\AnswerTODO[Yann]{Not really our job, but we can maybe remind some of the key arguments.
+\Answer{The MultiDefect objective function is based on the rationale that target structures should achieve high stability, and that the chosen RNA cannot adopt alternative, more stable, structures.
+It is a well accepted metrics introduced by other contributors, which we did not feel necessary to question or further improve.
 }
@@
 Also, RNARedprint and RNABlueprint are optimised to increase only a Multidefect value.\\[1em]}%
-\AnswerTODO[Stefan]{First of all, it is perhaps worth reminding that RNAredprint is not intended as a tool for negative design, but rather a good seed generation algorithm, amenable to further optimization using other tools.
-For this reason, we focused our validation on this potential for better initialization, and compared the quality of sequences generated by RNAredprint (Boltzmann weighted) to RNAblueprint (uniform).
-We noticed that, while RNAredprint is not explicitly tailored to optimize MultiDefect, it nevertheless produces sequences that: 1) achieve better values; and 2) are amenable to better further optimizations than the uniform version implemented by RNABluePrint.
-}
+\Answer{First of all, it is perhaps worth reminding that RNAredprint is not intended as a tool for negative design, but rather a generation algorithm, not only addressing a large array of design use-cases, but also able to generate good seed sequences, amenable to further optimization using other tools.
+For this reason, we focused our validation on this potential for better initialization, and compared the quality of sequences generated by RNARedPrint (Boltzmann weighted) to RNABlueprint (uniform).
+We noticed that, while RNAredprint is not explicitly tailored to optimize MultiDefect, it nevertheless produces sequences that: 1) achieve better values; and 2) are amenable to better further optimizations than the uniform version implemented by RNABluePrint.}
 
-\Comment{
-Hence, the distribution of Multidefect, previous criteria, and the first and second term of Multidefect should be shown in box plots or violin plots to know the robustness of RNARedPrint It is also interesting what is the characteristic of the sequence designed by each method in a specific example of functional RNA molecules.
-\\[1em]}%
-\AnswerTODO[Yann]{Again, we do not adhere to such a benchmark-driven view of design tasks.
-The time of design is generally negligible compared to the time taken to experimentally assess the ultimate functionality of a design.
-What we believe to matter more to the community is the expressivity of the tools, allowing to address various use-cases.
+\Comment{Hence, the distribution of Multidefect, previous criteria, and the first and second term of Multidefect should be shown in box plots or violin plots to know the robustness of RNARedPrint.
+It is also interesting what is the characteristic of the sequence designed by each method in a specific example of functional RNA molecules.\\[1em]}%
+\AnswerTODO[Yann]{Again, we do not feel the need to validate, or further analyze, the MultiDefect metrics, which has been a standard for multiple structure design.
+The time taken by the computational design is generally negligible compared to the time taken to experimentally validate the ultimate functionality of a design.
+What we believe to matter more to the community is the expressivity of the tool which, in this case, allows to address various use-cases, including some that could not be addressed before.
 }
 \Comment{(Is RNARedPrint able to produce, for example, the combination of two long stem loops which are high and low GC-content, respectively?)\\[1em]}%
-\AnswerTODO[Yann]{In the current implementation, it is only possible to constrain the global GC\% of the sequence.
+\Answer{In the current implementation, it is only possible to constrain the global GC\% of the sequence.
 It is however possible to support this feature within our general framework, since one can associated individual objectives to structures (eg  structures are already associated with individual target energies).
-Different stems can be expressed as different structures by "peeling the structure", so this feature could be added to a future version, should the concrete need arise.
-}
+Different stems can be expressed as different structures by "peeling the structure", so this feature could be added to a future version, should the concrete need arise.}
 \Comment{
-The section "problem statement" is not helpful enough for readers, both of algorithmic and biological people.\\[1em]}%
+The section "problem statement" is not helpful enough for readers, both of algorithmic and biological people.
+In particular, it is difficult to comprehend the relationships between the variables due to insufficient descriptions and lack of definitions.
+For example, they do not specify any relationship between R and F in this section while they should not be independent. (e.g., m should be equal or less than k?)
+\\[1em]}%
 \AnswerTODO[Yann]{We attempted to clarify it.
@@
 \Comment{
-In particular, it is difficult to comprehend the relationships between the variables due to insufficient descriptions and lack of definitions.
-For example, they do not specify any relationship between R and F in this section while they should not be independent. (e.g., m should be equal or less than k?)
 Moreover, I cannot find a definition of several variables. (e.g., What is $S$, $S_i$, $x$, $xp$, $xd$, $vp$, $bu{chi}$ (P9L24) ? Where "its dependence dep $f$" comes from?)
@@
 }
+
+\noindent
+{\bf Detailed comments:}
+
 \Comment{
@@
 \AnswerTODO{
-??
+ $\chi(v)$ is a subset of the positions in the RNA sequence.
+For an edge/base pair to be included in this set simply means that both of its edges are part of that set.
+It is crucial that all edges/base pairs are captured by at least one node $v$, so that the sequence compatibility can be ensured.
 }
@@
 * Is the output of algorithm 2 only one sequence? It is unclear when two (or more?) sequences are merged by $S \cup S'$.\\[1em]}%
-\AnswerTODO{???
+\AnswerTODO{Indeed. In order to simplify the notations, we use $S$ and $S'$ to mean partial assignments, which can be merged through an union.
 }
@@
 More detailed documentation is required such as OS dependency or the version information the authors confirmed.\\[1em]}%
-\AnswerTODO[Yanns]{
-Yann will check
+\AnswerTODO[Yann]{
+Yann will check; mention OS dependencies---not checked on MacOS. Merge pyscript branch into master branch!
 }
@@
 * The order of authors is different between the main manuscript and supplementary (not sure whether they used just an alphabetical order).\\[1em]}%
-\AnswerTODO{
-?
+\Answer{Thank you for spotting that issue, devoid of any deeper significance.
 }
@@
 The complexity of the problem is important to consider theoretical limitations, but the assumption in that section seems not to be necessary since the authors applied a greedy sampling strategy and not in the scope of this study.\\[1em]}%
-\Answer{we respectfully disagree with that assessment.
+\Answer{We respectfully disagree with that assessment.
 The core of our approach consists in generating, uniformly at random (resp.
@@
 In this work, we provide an {\bfseries exact} algorithm (not a heuristics) for the problem which, while being FPT, still requires exponential time when the tree-width increases with the sequence length.
-This raises the natural question: Can a truly polynomial algorithm exist for the problem?
+This motivates the natural question: Can a truly polynomial algorithm exist for the problem?
 Our proof of \#P-hardness provides a negative answer to this problem under fairly reasonable assumptions.
@@
 }
+\AnswerTODO{Revamp intro for \#P result + maybe overall intro}
 \Comment{
@@
 \AnswerTODO[Yann]{
-Ok (but include efficiency of multidim sampling, keeping correlation analysis)
+Ok (but include efficiency of multidim sampling, keeping correlation analysis) TODO:move parts of the Subsection: Efficiently targeting... to Methods.
 }
@@
 1) The "Methods" section is very long, it should be shortened.\\[1em]}%
+\Answer{
+As this reviewer could probably notice, the techniques used in this paper are quite divers, and stem from several subfields of computer science (graph algorithms, enumerative/analytic combinatorics...).
+For this reason, a more concise description of the method would be poorly self-contained, possibly to the point of irreproducibility.
+However, we share the concern of this reviewer regarding an imbalanced manuscript, and moved less original/central parts of the method to the appendix.
+This includes the expression of various energy models as constraints Turner, and the description of our naive numerical weight optimization.
+}
 \AnswerTODO[Yann]{
-Well, some methods require longer descriptions ;)
-Move expression of various energy models as constraints Turner to appendix
-Move numerical weight optim to appendix
+Move material to appendix
 Better motivations for \#P-hardness
@@
   1) Validation of stacking approximation: Correlation coefficient of Stacking and Turner (+ goodness of fit)
-  2) Rejection analysis for multidim sampling
-  3) Tree width and Time benchmark
-  5) Proof-of-concept -Potential of better seeds
+  2) Tree width and Time benchmark
+  3) Rejection analysis for multidim sampling
+  4) Proof-of-concept -Potential of better seeds
 }
@@
 Frnakenstein.\\[1em]}%
-\AnswerTODO[Stefan+Sebastian]{
-The three currently available tools are Frnakenstein, RNABluePrint and Modena. Shall we do that?
-}
+\AnswerTODO[Stefan+Sebastian]{Unfortunately, we could no longer locate the implementation of FRNAkenstein.
+Moreover, in the context of positive, the strategies of MODENA and FRNAkenstein are virtually identical.}
 \Comment{
@@
 why these two examples are output, and the other examples for RF3 exist or not?\\[1em]}%
+\Answer{This figure illustrates the impact of the energy model on the tree-decomposition, a simple energy model leading to a much simpler tree-decomposition than the stacking model.
+We now describe more explicitly
+}
 \AnswerTODO[Sebastian]{
-Discussion
+Add a description of tree decompositions and why they differ + New Illustration
 }
 \Comment{
-In "L Detailed result of MultiDefect Analysis" of Supplementary Material, running time is missing.
+In "Detailed result of MultiDefect Analysis" of Supplementary Material, running time is missing.
 Since the Boltzmann sampling in RNARedPrint utilizes the rejection approach, I would like to know the difference between the running times of the Boltzmann sampling and the uniform sampling.
 \\[1em]}%
-\AnswerTODO[Yann]{ Added runtimes.
-None, but for multidim sampling, rejection and weight optimization stack up (we report them).
-}
+\Answer{We were unsure about what was meant in this comment. Just to be precise, we distinguish the term 'Boltzmann sampling' from 'multi-dimensional Bolzmann sampling' in our paper. 
+On the one hand, Boltzmann sampling, i.e.
+drawing sequences from the Boltzmann distribution does not involve any rejection, and is actually exactly as costly as drawing them uniformly.
+In such a uniform mode, RNARedPrint performs equally as fast as RNABluePrint, and usually faster, since the graph decomposition used in RNABluePrint implies time and memory consumptions that are provably strictly less efficient than the tree decomposition underlying RNARedPrint.
+On the other hand, multi-dimensional Boltzmann sampling requires a rejection step to target specific properties that are typically difficult to obtain by chance in the uniform distribution.
+There is, of course, an added cost associated to this rejection, and the simultaneous weight optimization, which we discuss further in a new "Result" subsection, where we also report the exact time differences.}
 \Comment{
 In addition, I think it is better that the N/A result is indicated not by a blank but by a "N/A" in the table.\\[1em]}%
-\AnswerTODO[Stefan/Sebastian]{
-Will add results times}
+\Answer{We used N/A in the table.}
 \end{enumerate}

### 2018-12-20 RNARedPrint.tex
two columns, as it ought to be :)
Added figure and made some def more human-friendly

@@
 %\documentclass[doublespacing,linenumbers]{bmcart}% uncomment this for twocolumn layout and comment line below
-\documentclass{bmcart}
+\documentclass[twocolumn]{bmcart}
 
@@
 
-\begin{definition}[Tree decomposition and width]
+\begin{definition}[Tree decomposition]
   \label{def:treedecomp}
@@
 \end{enumerate}
+\end{definition}
+\revised{Intuitively, the tree decomposition of an (hypergraph) $G$ is a tree structure which features all the vertices of $G$, captures all the hyperedges of $G$, and helps ensure some notion of consistency.
+Figure~\ref{fig:treedec} illustrates one of the optimal tree decomposition for a pair of secondary structure under the stacking energy model.}
+\begin{definition}[Tree width]
 The \Def{width} of a tree decomposition $(T,\chi)$ is defined as
-$\width(T,\chi) = \min_{u\in T} |\chi(u)| - 1 $. The \Def{treewidth}
-of $G$ is the smallest width of any tree decomposition of $G$.
+$$\min_{u\in T} |\chi(u)| - 1.$$ The \Def{treewidth}
+of a graph $G$ is the smallest width of any tree decomposition of $G$.
 \end{definition}
 %Recall the essential role of the treewidth as fixed parameter in our FPT appraoch.
+
+\begin{figure}
+{\centering \includegraphics[width=.9\linewidth]{Figs/example-treedec-stacking}\\}
+\caption{\revised{Toy example of a tree decomposition associated with two target structures in the stacking energy model.
+Two target secondary structures ({\bfseries\sffamily A}) are merged into a joint hypergraph ({\bfseries\sffamily B}), whose hyperedges correspond to the quadruplets of positions involved in stacking pairs (colored).
+A valid tree decomposition ({\bfseries\sffamily C}) for the hypergraph ensures, among other properties, that each stacking pair/hyperedge is represented in at least one of its node, so that the energy features can be adequately evaluated.
+The width of this tree decomposition is 3, and is provably optimal for any tree decomposition of the input hypergraph.}\label{fig:treedec}}
+\end{figure}
 

### 2018-12-20 RNARedPrint.tex
Added low-res figure

@@
 $$
-F(S) =\!\!\!\!\!\!\!\sum_{\substack{\text{$f$: $f$ is contribution of $F$,}\\\text{ $\dep(f)=\{x_1,\ldots,x_p\}$}}}\!\!\!\!\!\!\! f(\{x_1\mapsto S_{x_1},\ldots,x_p\mapsto S_{x_p}\}).
+F(S) =\!\!\!\!\!\!\!\sum_{\substack{\text{$f$ contribution of $F$,}\\\text{ $\dep(f)=\{x_1,\ldots,x_p\}$}}}\!\!\!\!\!\!\! f(\{x_1\mapsto S_{x_1},\ldots,x_p\mapsto S_{x_p}\}).
 $$
@@
 
-Moreover, in a simple \Def{base pair energy model}, the energy of a target structure $R$ is expressed as a feature associated with contributions $f^R_{ij}(\{i\mapsto x, j\mapsto y\})=\EbpSym(x,y)$ for each base pair $(i,j)$ of $R,$ such that $F^R(S)=\EbpSym(S,R)$.
+% Moreover, in a simple \Def{base pair energy model}, the energy of a target structure $R$ is expressed as a feature associated with contributions $f^R_{ij}(\{i\mapsto x, j\mapsto y\})=\EbpSym(x,y)$ for each base pair $(i,j)$ of $R,$ such that $F^R(S)=\EbpSym(S,R)$.
 
@@
 and edges $\bigcup_{\ell\in[1,k]} R_\ell$.
-Any edge between two nodes $i$ and $j$ materializes the existence the existence of a base pair between positions $i$ and $j$ in some structure, inducing a direct depency between the assignments of these two positions.
+Any edge between two nodes $i$ and $j$ materializes the existence of a base pair between positions $i$ and $j$ in some structure, inducing a direct dependency between the chosen nucleotide assignments for the two positions.
 
@@
 Two target secondary structures ({\bfseries\sffamily A}) are merged into a joint hypergraph ({\bfseries\sffamily B}), whose hyperedges correspond to the quadruplets of positions involved in stacking pairs (colored).
-A valid tree decomposition ({\bfseries\sffamily C}) for the hypergraph ensures, among other properties, that each stacking pair/hyperedge is represented in at least one of its node, so that the energy features can be adequately evaluated.
-The width of this tree decomposition is 3, and is provably optimal for any tree decomposition of the input hypergraph.}\label{fig:treedec}}
+A valid tree decomposition ({\bfseries\sffamily C}) for the hypergraph ensures, among other properties, that each base pair and each stacking pair/hyperedge is represented in at least one of its node, so that features can be adequately evaluated.
+The width of this tree decomposition is 3, a provably optimal value for this input hypergraph.}\label{fig:treedec}}
 \end{figure}

### 2018-12-20 RNARedPrint.tex
minor edits

--- 2018-12-20 RNARedPrint.tex
@@
 $$
-F(S) =\!\!\!\!\!\!\!\sum_{\substack{\text{$f$: $f$ is contribution of $F$,}\\\text{ $\dep(f)=\{x_1,\ldots,x_p\}$}}}\!\!\!\!\!\!\! f(\{x_1\mapsto S_{x_1},\ldots,x_p\mapsto S_{x_p}\}).
+F(S) =\!\!\!\!\!\!\!\!\!\!\sum_{\substack{\text{$f$: $f$ is contribution of $F$,}\\\text{ $\dep(f)=\{x_1,\ldots,x_p\}$}}}\!\!\!\!\!\!\!\!\!\! f(\{x_1\mapsto S_{x_1},\ldots,x_p\mapsto S_{x_p}\}).
 $$
@@
 
-Moreover, in a simple \Def{base pair energy model}, the energy of a target structure $R$ is expressed as a feature associated with contributions $f^R_{ij}(\{i\mapsto x, j\mapsto y\})=\EbpSym(x,y)$ for each base pair $(i,j)$ of $R,$ such that $F^R(S)=\EbpSym(S,R)$.
-
-Note that for \GCb\% as well as for the base pair energy features, the contributions depend on the nucleotides at only a small, constantly bounded number of sequence positions (one and two positions, respectively).
+%Moreover, in a simple \Def{base pair energy model}, the energy of a target structure $R$ is expressed as a feature associated with contributions $f^R_{ij}(\{i\mapsto x, j\mapsto y\})=\EbpSym(x,y)$ for each base pair $(i,j)$ of $R,$ such that $F^R(S)=\EbpSym(S,R)$.
+
+%Note that the \GCb\% as well as for the base pair energy features, the contributions depend on the nucleotides at only a small, constantly bounded, number of sequence positions (one and two positions, respectively).
 
@@
 \end{definition}
-\revised{Intuitively, the tree decomposition of an (hypergraph) $G$ is a tree structure which features all the vertices of $G$, captures all the hyperedges of $G$, and helps ensure some notion of consistency.
-Figure~\ref{fig:treedec} illustrates one of the optimal tree decomposition for a pair of secondary structure under the stacking energy model.}
+\revised{Intuitively, a tree decomposition for an (hyper)graph $G$ is a tree structure which features all the vertices of $G$, captures all the (hyper)edges of $G$, and helps ensure some notion of consistency.
+Figure~\ref{fig:treedec} illustrates shows an optimal tree decomposition for a pair of structures under the stacking energy model.}
 \begin{definition}[Tree width]
@@
 
-To formalize our algorithms, which iteratively merge evaluations of partial solutions, we extend the idea of atomic feature contributions, which are evaluated at sets of the form $\{x_1\mapsto v_1, \dots, x_d\mapsto v_p\}$.
-Let us call the latter object \Def{partial sequence}; they are handy to specify partial knowledge on the entire sequence at some point of the algorithm.
-Easily, we can extend the definition of contributions $f$ to sets $\{x_1\mapsto v_1, \dots, x_p\mapsto v_p\}$, where $\{x_1\dots x_p\}$ is any super-set of $\dep(f)$ by ignoring the superfluous assignments $x\mapsto v$, where $x\not\in\dep(f)$.
-
 Our algorithms specialize the idea of cluster tree elimination (CTE)~\citep{Dechter2013}, which operates on constraint networks.
 In this correspondence, (partial) sequences specialize (partial) assignments and the constraint network would be given by variables for each sequence position, constraints due to valid base pairing, and the set of atomic feature contributions $\cal F$.
+
+To formalize our algorithms, which iteratively merge evaluations of partial solutions, we extend the idea of atomic feature contributions, which are evaluated at sets of the form $\{x_1\mapsto v_1, \dots, x_d\mapsto v_p\}$.
+Let us call the latter object a \Def{partial sequence}; such an object will help specify partial knowledge on the sequence at some point of the algorithm.
+Easily, we can extend the definition of contributions $f$ to sets $\{x_1\mapsto v_1, \dots, x_p\mapsto v_p\}$, where $\{x_1\dots x_p\}$ is any super-set of $\dep(f)$ by ignoring the superfluous assignments $x\mapsto v$, where $x\not\in\dep(f)$.
 
@@
 on a tree decomposition of $G_{\F}$, the dependency graph associated with $\F$.
-Note that analogous algorithms could be easily derived to count valid sequences, or produce the sequence with minimal free-energy.
+Note that analogous algorithms could be easily derived to count valid sequences, or list sequences having minimal free-energy.
 %This yields fixed parameter tractable (FPT) algorithms, based on the treewidth.
@@
 %
-We assume several properties of the given cluster tree (reflecting $\F$), which otherwise could be easily obtained:
+We assume the following properties of the given cluster tree (reflecting $\F$):
 \begin{itemize}
@@
 
-\subsection*{Computational complexity of the multiple target sampling algorithm}\label{sec:complexity}
+\subsection*{Complexity considerations}\label{sec:complexity}
 
@@
 %We define the following notations: $\real^\infty:=\real\cup\{\infty\}$, $[i,j]:=\{i,\ldots,j\},$ \ldots
+This 
 Finally, we turn to the complexity of $\NumDesign(G)$, the problem of

### 2018-12-20 RNARedPrint.tex
Shortened intro to #P hardness result Added explanatory examples of features

@@
 
-\newcommand{\revised}[1]{\textcolor{red}{#1}}
+\newcommand{\revised}[1]{{\color{red} #1}}
 
@@
 
-\subsection*{Representing \GCb\%-content and energies as features.}
+\subsection*{Expressing \GCb\%-content, sequence validity and energies as features}
 Formally, we define a \Def{feature $F$} as a function on sequences, whose value is obtained by summing over an associated set of \Def{contributions}.
-Each contribution $f$ depends on a restricted set of positions, its \emph{dependencies $\dep(f)$}, such that
+Each contribution $f$ takes values in $\mathbb{R}\cup \{+\infty\}$, and depends on the nucleotides assigned to a restricted set of positions, its \emph{dependencies $\dep(f)$}, such that
 $$
-F(S) =\!\!\!\!\!\!\!\!\!\!\sum_{\substack{\text{$f$: $f$ is contribution of $F$,}\\\text{ $\dep(f)=\{x_1,\ldots,x_p\}$}}}\!\!\!\!\!\!\!\!\!\! f(\{x_1\mapsto S_{x_1},\ldots,x_p\mapsto S_{x_p}\}).
+F(S) =\!\!\!\!\!\!\!\sum_{\substack{\text{$f$ contribution of $F$,}\\\text{ $\dep(f)=\{x_1,\ldots,x_p\}$}}}\!\!\!\!\!\!
+f\left(\left\{\substack{x_1\mapsto S_{x_1}\\\cdots\\x_p\mapsto S_{x_p}}\right\}\right).
 $$
 
-For instance, the $\GCb\%$ can be expressed using $n$ contributions $f^{\GCb}_i$, $i\in[1,n]$,  where $f^{\GCb}_i(\{i\mapsto S_i\})=-1$ if $S_i=\Gb$ or $\Cb$, and $0$ otherwise.
+The $\GCb\%$ can be trivially expressed using $n$ contributions $f^{\GCb}_i$, $i\in[1,n]$, whose dependencies are restricted to individual nucleotides,  and such that 
+$$f^{\GCb}_i(\{i\mapsto c\})=\begin{cases}-1&\text{ if }c=\Gb\text{ or }\Cb\\ 0&\text{ otherwise}.\end{cases}$$ 
+\revised{By summing $f^{\GCb}_i$ over the whole sequence, one simply counts the occurrences of $\Gb$ and $\Cb$.}
+
+\revised{
+	Such features can also be used to count the number of \Def{valid sequences}, \emph{i.e.} sequences inducing only base pairs in $\mathcal{B}:=\{\{\Ab,\Ub\}, \{\Gb,\Cb\}, \{\Gb,\Ub\}\}$.
+Indeed, consider a feature $F^{\tt BP}$ composed of  contributions $f^{\tt BP}_{i,j}$, for each base pair $(i,j)$ occurring in some structure, such that
+	$$f^{{\sf BP}}_{i,j}(\left\{\substack{i\mapsto a\\j\mapsto b}\right\})=\begin{cases}0&\text{ if }\{a,b\}\in\mathcal{B} \\+\infty&\text{ otherwise}.\end{cases}$$ 
+	The value of $F^{\tt BP}$ is $0$ for any valid sequence, and $+\infty$ as soon as some non canonical base pair is created.
+For any associated weight $\pi_{{\sf BP}}>1$, the contribution of a valid sequence is $\pi_{{\sf BP}}^0 = 1$, and the contribution of an invalid seqeunce is $\pi_{{\sf BP}}^{+\infty} = 0$, so that Equation~\ref{eq:mainproblem} simply counts the number of valid sequences when restricted to to $F^{\tt BP}$.
+
+	Energy models for structure prediction vary considerably, yet can always be expressed as a sum over contributions associated with local structural motifs (base pairs, stacking pairs, loops\ldots) under a certain nucleotide assignment.
+Energy models can thus be captured generically by introducing, to each motif $m$ occurring in a structure, a contribution $f_m$, taking a specific value for each assignment of nucleotides to its specific set of positions $\dep(f_m)$.
+For instance, the contribution of a stacking pair, consisting of two pairs $(i,j)$ and $(i+1,j-1)$, will be expressed through a
+
+}
 
@@
 
-\subsection*{Dependency structure, tree decomposition and width}
+\subsection*{Dependency graph, tree decomposition and width}
 %The complexity of the central problem depends fundamentally on the dependency structure that is imposed by the entire set of contributions.
@@
 
-First, in order to capture the requirement of canonical base pairing for each structure,
+In order to capture the requirement of canonical base pairing for each structure,
 we define the \Def{base pair dependency graph} $G_{\R}$ as the graph with nodes $\{1,\dots,n\}$
@@
 
-More generally, we consider dependencies induced by the complete set of contributions $\F := \bigcup_{\ell}\{f\mid f\text{ contribution of } F_\ell\}$.
+More generally, we consider dependencies induced by the complete set of contributions $$\F := \bigcup_{\ell}\{f\mid f\text{ contribution of } F_\ell\}.$$
 %
@@
 The set $\F$ of contributions induces a
-  \Def{dependency graph} $G_\F$ on sequence positions by interpreting the dependencies as hyperedges, namely the hypergraph 
-  $G_\F=(\{1,\dots,n\} ,\{\dep(f)\mid f\in \F\})$. The \Def{tree decomposition} of $G_\F$ is fundamental for our algorithms.
+  \Def{dependency graph} $G_\F$ on sequence positions by interpreting the dependencies as hyperedges, leading to the hypergraph 
+  $$G_\F=(\{1,\dots,n\} ,\{\dep(f)\mid f\in \F\}).$$ 
+  Let us now remind the \Def{tree decomposition} of a graph $G_\F$, a fundamental ingredient of our algorithms.
 
@@
 \end{definition}
-\revised{Intuitively, a tree decomposition for an (hyper)graph $G$ is a tree structure which features all the vertices of $G$, captures all the (hyper)edges of $G$, and helps ensure some notion of consistency.
+\revised{Intuitively, a tree decomposition for an (hyper)graph $G$ is a tree structure which features all the vertices of $G$, captures all the (hyper)edges of $G$, and helps ensure some notion of consistency by allowing the propagation of values.
 Figure~\ref{fig:treedec} illustrates shows an optimal tree decomposition for a pair of structures under the stacking energy model.}
@@
 
-For algorithmic treatment, it is convenient to encode the weight $\pi$ of each feature in its contributions, such that all contributions can be handled uniformly.
+Moreover, to ensure a uniform algorithmic treatment of contributions, it is convenient to encode the weight $\pi$ of each feature in its contributions.
 This transformation works by multiplying all contributions with $ln(\pi)$, where $\pi$ is the weight of the corresponding feature, since then
@@
 For the multidimensional Boltzmann sampling, we require the already established ability to \Def{sample from a weighted distribution} over the set of valid sequences, where the probability of a sequence $S$ is
-$\mathbb{P}(S\mid \pmb{\pi}) = \frac{\prod_{\ell=0}^{k} \pi_i^{-F_i(S)}}{\partfun{\pmb{\pi}}},$
+$$\mathbb{P}(S\mid \pmb{\pi}) = \frac{\prod_{\ell=0}^{k} \pi_i^{-F_i(S)}}{\partfun{\pmb{\pi}}},$$
 where $\pmb{\pi}:=(\pi_0\cdots\pi_k)$ is the vector of the positive real-valued \Def{weights}, and $\partfun{\pmb{\pi}}$ is the weighted partition function.
@@
 %We define the following notations: $\real^\infty:=\real\cup\{\infty\}$, $[i,j]:=\{i,\ldots,j\},$ \ldots
-This 
-Finally, we turn to the complexity of $\NumDesign(G)$, the problem of
-computing the number of valid sequences for a given compatibility
-graph $G=(V,E)$. Note that this problem corresponds to the partition
-function problem with weights $\pi_\ell=1$ for each target structure, so that each valid structure contributes $1$ to the partition
-function; each invalid structure, 0.
-
-As previously noted~\citep{Flamm2001}, a set of target
-structures admits a valid design iff its compatibility graph is
-bipartite, which can be tested in linear time.  Moreover, without
-$\{\Gb, \Ub\}$ base pairs, the design of any connected component $C$
-of $G$ would be entirely determined by the assignment of a single of
-its nucleotides. The number of valid designs is thus simply
-$4^{\#{\rm CC}(G)}$, where $\#{\rm CC}(G)$ is the \Def{number of
-  connected components}.
-The problem becomes much harder when $\{\Gb, \Ub\}$ base pairs are allowed, and we show that valid designs for a set of structures cannot be counted in polynomial time, unless ${\sf\# P}={\sf FP}$.
+\revised{While efficient, both in practice and in theory for graphs of bounded tree width, our algorithms remain exponential in the worst case scenario, since the tree-width of a dependency graph can then become arbitrarily large.
+This exponential complexity in the worst case seems intrinsic.
+Indeed, we show that a specialization of our core problem, namely the enumeration of designs that respect canonical base pairing rules ($\Ab \leftrightarrow \Ub$, $\Gb \leftrightarrow \Cb$, $\Gb \leftrightarrow \Ub$) is $\#{\sf P}$-hard, even when the dependency graph is bipartite and connected.
+The existence of a polynomial time algorithm for computing the partition function of Equation \eqref{eq:mainproblem} is thus unlikely, and it would imply that  ${\sf\# P}={\sf FP}$ and, in turn, that ${\sf P}={\sf NP}$.}
+
+%Our contributed algorithms, while 
+%Finally, we turn to the complexity of $\NumDesign(G)$, the problem of
+%computing the number of valid sequences for a given compatibility
+%graph $G=(V,E)$. Note that this problem corresponds to the partition
+%function problem with weights $\pi_\ell=1$ for each target structure, so that each valid structure contributes $1$ to the partition
+%function; each invalid structure, 0.
+%
+%As previously noted~\citep{Flamm2001}, a set of target
+%structures admits a valid design iff its compatibility graph is
+%bipartite, which can be tested in linear time.  Moreover, without
+%$\{\Gb, \Ub\}$ base pairs, the design of any connected component $C$
+%of $G$ would be entirely determined by the assignment of a single of
+%its nucleotides. The number of valid designs is thus simply
+%$4^{\#{\rm CC}(G)}$, where $\#{\rm CC}(G)$ is the \Def{number of
+%  connected components}.
+%The problem becomes much harder when $\{\Gb, \Ub\}$ base pairs are allowed, and we show that valid designs for a set of structures cannot be counted in polynomial time, unless ${\sf\# P}={\sf FP}$.
 This equality would, in particular, imply the classic ${\sf P}={\sf NP}$.
 
-To establish that claim, we consider instances $G=(V_1\cup V_2, E)$ that are connected and bipartite ($E \cap (V_1\times V_2) = E$), noting that hardness over restricted instances implies hardness of the general problem.
-Moreover note that, as we explained before, assigning a nucleotide to a position $u\in V$ constrains the parity ($\{\Ab,\Gb\}$ or $\{\Cb,\Ub\}$) of all positions in the connected component of $u$.
+To establish that claim, we consider a dependency graph $G=(V_1\cup V_2, E)$ that is connected and bipartite ($E \cap (V_1\times V_2) = E$).
+Note that, assigning a nucleotide to a position $u\in V$ constrains the parity ($\{\Ab,\Gb\}$ or $\{\Cb,\Ub\}$) of all positions in the connected component of $u$.
 For this reason, we restrict our attention to the counting of valid designs \emph{up to trivial  symmetry} $(\Ab\leftrightarrow \Cb/\Gb\leftrightarrow \Ub)$, by constraining the positions in $V_{1}$ to $\Ab$ and $\Gb$.
@@
 
-Let $\IS{G}$ denote the set of all independent sets in the connected graph $G$; recall that an \Def{independent set} of $G=(V,E)$ is a subset $V'\subseteq V$ of nodes that are not connected by any edge in $E$.
+Finally, let $\IS{G}$ denote the set of all independent sets in the connected graph $G$; recall that an \Def{independent set} of $G=(V,E)$ is a subset $V'\subseteq V$ of nodes that are not connected by any edge in $E$.
 
@@
 
-Now we can build on the connection between the two problems to obtain complexity results for \NumDesign.
-Counting independent sets in bipartite graphs ($\#{\sf BIS}$) is indeed a well-studied \#{\sf P}-hard problem~\citep{Ge2012}.
+Counting independent sets in bipartite graphs ($\#{\sf BIS}$) is a well-studied problem, shown to be \#{\sf P}-hard~\citep{Ge2012} even on connected graphs.
 Now assume the existence of an efficient (polynomial-time) algorithm $\mathcal{A}$ for computing $|\NumDesign(G)|$ on connected (bipartite) graphs.
 Then, running $\mathcal{A}$ and returning $|\NumDesign(G)|/2$ constitutes an efficient algorithm for $\#{\sf BIS}$ on connected graphs.
-However, $\#{\sf BIS}$ is also \#{\sf P}-hard on connected graphs, as the number of independent sets for a disconnected graph $G$ can be efficiently computed by $|\IS{G}|=\prod_{cc\in CC(G)} |\IS{cc}|$.
-In other words, any efficient algorithm for \NumDesign implies an efficient algorithm for $\#{\sf BIS}$, and we conclude with the hardness of \NumDesign.
+%However, $\#{\sf BIS}$ is also \#{\sf P}-hard on connected graphs, as the number of independent sets for a disconnected graph $G$ can be efficiently computed by $|\IS{G}|=\prod_{cc\in CC(G)} |\IS{cc}|$.
+In other words, any efficient algorithm for \NumDesign implies an efficient algorithm for $\#{\sf BIS}$, thus our conclusion on the complexity of \NumDesign.
 

### 2018-12-20 RNARedPrint.tex
Updates from ShareLaTeX

--- 2018-12-20 RNARedPrint.tex
@@
 \Software{RNAdesign} performs much better than the theoretical bound of $O(4^n)$, no attempts were made to characterize or justify its---still exponential---complexity; leaving important theoretical questions of the complexity of counting and uniform sampling open.
-As well, the \Software{RNAdesign}/\RNAblueprint{} approach is very specialized to uniform sampling, which limits its direct extensibility.
+As well, the \Software{RNAdesign}/\RNAblueprint{} approach is specialized to uniform sampling, which limits its direct extensibility.
 Both aspects call for a more systematic redesign of multi-target sampling, which builds on established concepts in computer science.
@@
 
-\parHead{Motivation.}
+\todo[inline]{SW: The following paragraph (together with the first sentence of Subsection Contributions) is rather confusing than helpful.
+We could drop this here and give brief explanations of the complexity issues in the contributions section.}
 Better understanding of the true complexity bounds is highly relevant for the application of such methods, in particular for applications in more expressive settings (e.g.
@@
 How to enforce additional constraints, such as \GCb\%, complex sequence constraints, or the energy of individual structures?}
-
-%All such extensions are still limited by the complexity of the multi-target uniform sampling problem, where counting of valid designs is the dominating step.
 

### 2019-01-22 supplement.tex
Updates from ShareLaTeX
Add run times section to supplement

@@
 
-\newcommand{\ourprog}{\Software{RNARedPrint}}
+\newcommand{\RNAblueprint}{{\tt \bfseries{}\color{black!85} RNA\textcolor{blue!70!black}{Blue}Print}}
+\newcommand{\ourprog}{{\tt \bfseries{}\color{black!85}RNA\textcolor{red!70!black}{Red}Print}}
 
@@
 \newcommand{\citet}[1]{\cite{#1}}
+
+%% highlight changes for revision in red
+\newcommand{\revised}[1]{{\color{red} #1}}
+\newenvironment{revision}{\color{red}}{\color{black}}
 
@@
 \begin{center}
-\large Stefan Hammer, Yann Ponty, Wei Wang and Sebastian Will
+\large Stefan Hammer, Wei Wang, Sebastian Will and Yann Ponty
 \end{center}
@@
 \end{figure}
+
+\begin{revision}
+  \section{Run times for generating samples}
+
+To illustrate the efficiency and practical feasibility of our Boltzmann sampling approach, we show the distributions of run times for generating 1000 samples (over the Modena benchmark instances and grouped by benchmark sub-set).
+Since producing a controlled Boltzmann sampling requires significant pre-computation, it is informative to separately show construction and sampling times (Figure~\ref{fig:run-times-redprint}).
+
+\todo[inline]{@Stefan: could you please include technical data about the machines used for the timings?
+Please also check the claims/statements in the text for correctness.
+If necessary---please edit.}
+
+A fair comparison to sampling times by other tools can hardly be given, since the other tools produce samplings of very different nature.
+In particular, Modena does not make any guarantees on properties of the sampling.
+Non-surprisingly, this allows it to sample significantly faster (several orders of magnitudes).
+
+The comparison to run times of \RNAblueprint{} is still interesting, since this tool generates---similarly to our tool---\emph{controlled} samples.
+Nevertheless, as we argue the Boltzmann sampling is fundamentally more flexible than uniform sampling as generated by \RNAblueprint.
+Moreover, in this concrete case, \ourprog{} samples based on the rather complex stacking energy model, which introduces much stronger dependencies than required for uniform sampling.
+Technically, these stronger dependencies increase the tree width; here, the increase is exactly the same as the one from the base pair to the stacking model as reported in Figure~\ref{}.
+In this light, \ourprog{} performs even particularly well, beating \RNAblueprint{} most of the time with only a few exceptions for particularly hard instances.
+The results for \RNAblueprint{} are shown in Figure~\ref{fig:run-times-rbp}.
+
+\begin{figure}[tbh]
+  \centering
+  \includegraphics[width=0.45\textwidth]{Figs/RunTimes/construction_time_sets}\quad%
+  \includegraphics[width=0.45\textwidth]{Figs/RunTimes/sample_time_sets}
+  \caption{\todo[inline]{caption Stefan?}}
+  \label{fig:run-times-redprint}
+\end{figure}
+\begin{figure}[tbh]
+  \centering
+  \includegraphics[width=0.45\textwidth]{Figs/RunTimes/construction_time_sets_rbp}\quad%
+  \includegraphics[width=0.45\textwidth]{Figs/RunTimes/sample_time_sets_rbp}
+  \caption{\todo[inline]{caption Stefan?}}
+  \label{fig:run-times-rbp}
+\end{figure}
+
+\end{revision}
 

### 2019-01-22 supplement.tex
Reorder supplement sections; add text to MultiDecect-contributions section

@@
 \end{center}
+
+\begin{revision}
+  \section{Run times for generating samples}
+
+To illustrate the efficiency and practical feasibility of our Boltzmann sampling approach, we show the distributions of run times for generating 1000 samples (over the Modena benchmark instances and grouped by benchmark sub-set).
+Since producing a controlled Boltzmann sampling requires significant pre-computation, it is informative to separately show construction and sampling times (Figure~\ref{fig:run-times-redprint}).
+
+\todo[inline]{@Stefan: could you please include technical data about the machines used for the timings?
+Please also check the claims/statements in the text for correctness.
+If necessary---please edit.}
+
+A fair comparison to sampling times by other tools can hardly be given, since the other tools produce samplings of very different nature.
+In particular, Modena does not make any guarantees on properties of the sampling.
+Non-surprisingly, this allows it to sample significantly faster (several orders of magnitudes).
+
+The comparison to run times of \RNAblueprint{} is still interesting, since this tool generates---similarly to our tool---\emph{controlled} samples.
+Nevertheless, as we argue the Boltzmann sampling is fundamentally more flexible than uniform sampling as generated by \RNAblueprint.
+Moreover, in this concrete case, \ourprog{} samples based on the rather complex stacking energy model, which introduces much stronger dependencies than required for uniform sampling.
+Technically, these stronger dependencies increase the tree width; here, the increase is exactly the same as the one from the base pair to the stacking model as reported in Figure~\ref{}.
+In this light, \ourprog{} performs even particularly well, beating \RNAblueprint{} most of the time with only a few exceptions for particularly hard instances.
+The results for \RNAblueprint{} are shown in Figure~\ref{fig:run-times-rbp}.
+
+\begin{figure}[tbh]
+  \centering
+  \includegraphics[width=0.45\textwidth]{Figs/RunTimes/construction_time_sets}\quad%
+  \includegraphics[width=0.45\textwidth]{Figs/RunTimes/sample_time_sets}
+  \caption{\todo[inline]{caption Stefan?}}
+  \label{fig:run-times-redprint}
+\end{figure}
+\begin{figure}[tbh]
+  \centering
+  \includegraphics[width=0.45\textwidth]{Figs/RunTimes/construction_time_sets_rbp}\quad%
+  \includegraphics[width=0.45\textwidth]{Figs/RunTimes/sample_time_sets_rbp}
+  \caption{\todo[inline]{caption Stefan?}}
+  \label{fig:run-times-rbp}
+\end{figure}
+
+\end{revision}
+
+\section{Illustrating multi-dimensional Boltzmann sampling for three pseudoknot-free target structures}
+
+    In analogy to Fig.~\ref{fig:energydist-pk} of the main text,
+    Fig.~\ref{fig:energydist} illustrates our multi-dimensional sampling
+    strategy for three pseudoknot-free target structures (example instance
+    from the \texttt{Modena} benchmark). The figure shows Turner energy
+    distributions of single structures from sampling with different weight
+    parameters, where weights are trained to target specific energies. For
+    comparison, the figure shows distributions from uniform and Boltzmann
+    sampling at respective homogenous weights $1$ and $e^\beta$.
+
+    Fig.~\ref{fig:gc-content} shows the distributions of the GC content of the samples, while targeting different combinations of energies in this experiment.
+For completeness, we also report corresponding results for the analogous experiment from the main text.
+We observe that in addition to the target energies, also the feature $\GCb\%$ is narrowly defined by the multi-dimensional Boltzmann sampling strategy.
+In comparison, for uniform sampling and Boltzmann sampling at high weight $e^\beta$, where the \GCb\% is not controlled, we obtain much broader distributions.
+
+\begin{figure}[tb]
+      \begin{center}
+        %{\sf \bfseries A}\includegraphicstop[width=0.32\textwidth]{Figs/offset2}\hspace{.5cm}
+        %{\sf \bfseries B}
+        \includegraphicstop[width=0.8\textwidth]{Figs/energy_distribution}
+      \end{center}
+      \caption{%
+        Targeting specific energies using multi-dimensional Boltzmann
+        sampling. We show the Turner energy distributions for
+        three target structures R1, R2 and R3. By
+        targeting the respective free energies $(-40,-40,-20)$ and
+        $(-20,-20,-20)$ for the target structures (annotated in red and
+        blue, resepectively), we demonstrate the
+        effectivity of our adaptive multi-dimensional Boltzmann sampling
+        procedure; moreover, for comparison, the distributions for uniform
+        (U) and Boltzmann (B) samples---respectively associated with
+        homogenous weights $1$ and $e^\beta$.
+%    ($\delta=0.05, \delta'=0.1, \gamma=1.1, \psi=20, n=1000$)
+  }
+  \label{fig:energydist}
+\end{figure}
+
+\begin{figure}
+  \begin{center}
+    {\sf \bfseries A}~\includegraphicstop[width=0.45\textwidth]{Figs/Plots/PKB00211_PKB00239_0_GC_content}
+    ~~~~~{\sf \bfseries B}~\includegraphicstop[width=0.45\textwidth]{Figs/Plots/supp_3str_GC_content}
+  \end{center}
+  \caption{Distribution of the \GCb-content \GCb\% in the experiments of (A) Fig.~\ref{fig:energydist-pk} of the main text and (B) Suppl.~Fig.~\ref{fig:energydist}.}
+  \label{fig:gc-content}
+\end{figure}
+
+\section{Immediate benefits of positive design for negative design}
+
+Already previous work on RNA Design, e.g. INFO-RNA and Incarnation, observed
+that---put briefly---negative design benefits from positive design. Here, more
+concretely, we simply \emph{demonstrate} this effect (for small examples), where
+we consider \Def{negative design} as  
+designing RNA sequences that fold with high probability $Pr(R|S)$ into the
+given target structure $R$; and \Def{positive design}, as generating one or
+many sequences that have a low energy for $R$.
+
+For each of two small examples, a single-stem and a double-stem structure, we produce a uniform and a weighted Boltzmann
+sampling, each of 10000 sequenes. For each sampled sequence $S$, we determine
+the Turner energy $E(S,R)$ as well as the
+%difference $E_\Delta(S,R)$ of $E(S,R)$ to the ensemble energy $E_{\text{ens}}(S)$ of $S$ 
+probability $Pr(R|S)$ of the structure $R$ in the ensemble of sequence $S$ 
+(where the energies are calculated in the Turner model using RNAfold).
+
+Note that low energy $E(S|R)$ is (non-linearly) correlated with high
+probability $Pr(S|R)$, the probability to see $S$ in the Boltzmann distribution
+of sequences folding into $R$. Recall that the probability $Pr(R|S)$ can be
+calculated as $\exp(-E(S,R)/RT)/Z(S)$,
+where $Z(S)=\sum_{R'} \exp(-E(S,R')/RT)$.
+%
+Consequently, Figure~\ref{fig:negforpos} ultimately illustrates the dependencies
+between $Pr(S|R)$ and $Pr(R|S)$ for uniform samples and low energy samples
+(where those samples can be efficiently generated due to our Boltzmann sampling
+strategy). For our examples, we observe (Fig.~\ref{fig:negforpos}) that Boltzmann sampling with weight 500 (green), generates good negative designs 
+with substantially higher probability than uniform sampling (blue).
+
+% and high energy difference $E_\Delta(S,R)=E_{\text{ens}}(S)-E(S,R)$, with high probability $\Pr(R|S)$.
+% 
+% To see the latter, recall that the \Def{ensemble energy $E_{\text{Ens}}(S)$} is defined as
+% $$
+% E_{\text{ens}}(S) = -RT \ln( Z(S) ),
+% $$ 
+% based on the partition function $Z(S)=\sum_{R'} \exp(-E(S,R')/RT)$
+% such that 
+% $$
+% E(S,R) - E_{\text{ens}}(S) 
+% =
+% -RT \ln Pr(R|S),
+% $$  
+% which can be derived by term transformation 
+% \begin{align*}
+% -RT \ln P(R|S) &= -RT \ln [ exp(-E(S,R)/RT) / Z(S) ] \\&= -RT \ln (exp(-E(S,R)/RT)) - (-RT\ln Z(S))
+% \\&=  E(S,R) - E_{\text{Ens}}(S). 
+% \end{align*}
+
+\begin{figure}
+\textbf{A}~\includegraphicstop[width=0.48\textwidth]{Figs/Plots/stem_weight1-500_mean}%
+\textbf{B}~\includegraphicstop[width=0.48\textwidth]{Figs/Plots/doublestem_weight1-500_mean}%
+\caption{Comparison of uniform sampling and Boltzmann sampling with
+weight $\pi=500$. We show results for two different small structures: (A) single stem structure (B) double stem structure. For each respective structure $R$, we plot the energies $E(S,R)$ of
+10000 sampled sequences against the probabilities $Pr(R|S)$ (in log scale).
+For illustrating the trend, we show moving averages over windows of $\pm 1$ kcal.
+Note that while only the energy $E(S,R)$ is directly controlled by Boltzmann sampling, this strongly impacts the common negative design objective $Pr(R|S)$.}
+\label{fig:negforpos}
+\end{figure}
+
+\section{Parameters for the base pair and the stacking energy model}
+\label{appsec:modelparameters}
+
+We trained parameters for two RNA energy models to approximate the
+Turner energy model, as implemented in the \Software{ViennaRNA}
+package.  In the \Def{base pair model}, the total energy of a sequence $S$
+for an RNA structure $R$ is given as sum of base pair energies, where
+we consider six types of base pairs distinguishing by the bases $\Ab$-$\Ub$,
+$\Cb$-$\Gb$ or $\Gb$-$\Ub$ (symmetrically) and between stacked and non-stacked base
+pairs; here, we consider base pairs $(i,j)\in R$ \Def{stacked} iff
+$(i+1,j-1)\in R$, otherwise \Def{non-stacked}. In the \Def{stacking model},
+our features are defined by the stacks, i.e. pairs $(i,j)$ and
+$(i+1,j-1)$ which both occur in $R$; we distinguish 18 types based on
+$S_i,S_j,S_{i+1},S_{j-1}$ (i.e. all combinations that allow canonical
+base pairs; the configurations $S_i,S_j,S_{i+1},S_{j-1}$ and
+$S_{i+1},S_{j-1},S_j,S_i$ are symmetric).
+
+Both models describe the energy assigned to a pair of sequence and
+structure in linear dependency of the number of features and their
+weights. We can thus train weights for linear predictors of the Turner
+energy in both models.
+
+For this purpose, we generated 5000 uniform random RNA sequences of
+random lengths between 100 and 200. For each sample, we predict the
+minimum free energy and corresponding structure using the ViennaRNA
+package; then, we count the distinguished features (i.e., base pair or
+stack types). The parameters are estimated fitting linear models
+without intercept (\texttt{R} function \texttt{lm}). For both models,
+\texttt{R} reports an adjusted R-squared value of 0.99. The resulting
+parameters are reported in Tables~\ref{tab:basepairmodel} and
+\ref{tab:stackingmodel}.
+
+For validating the trained parameters, we generate a second
+independent test set of random RNA sequences in the same
+way. Fig.~\ref{fig:training-cor} shows correlation plots for the
+trained parameters in the base pair and stacking models for predicting
+the Turner energies in the test set with respective correlations of
+$0.95$ an $0.94$.
+
+\begin{figure}
+  \centering
+  A)\includegraphicstop[width=0.4\textwidth,trim=0 0 0 50,clip]{Figs/basepaircor}
+  B)\includegraphicstop[width=0.4\textwidth,trim=0 0 0 50,clip]{Figs/stackingcor}
+  \caption{Validation of the trained parameters for the A) base pair
+    model B) stacking model for predicting energies of the
+    independently sampled sequences in the test set. We show
+    correlation plots against the Turner energies reported by the
+    ViennaRNA package.}
+  \label{fig:training-cor}
+\end{figure}
+
+\begin{table}[b]
+  \centering
+  \caption{Trained weights for the base pair energy model.}
+  \label{tab:basepairmodel}
+  \begin{tabular}{c@{\quad}c@{\quad}c@{\quad}|@{\quad}c@{\quad}c@{\quad}c}
+    \multicolumn{3}{c}{non-stacked} & \multicolumn{3}{c}{stacked}\\
+    AU      & CG       & GU      & AU       & GC       & GU \\\hline
+    1.26630 & -0.09070 & 0.78566 & -0.52309 & -2.10208 & -0.88474
+  \end{tabular}
+\end{table}
+
+% GC_IN = -0.09070;
+% AU_IN = 1.26630;
+% GU_IN = 0.78566;
+% GC_TERM = -2.10208;
+% AU_TERM = -0.52309;
+% GU_TERM = -0.88474;
+
+\begin{table}[b]
+  \centering
+  \caption{Trained weights for the stacking energy model (the rows specify $S_i,S_j$; the columns, $S_{i+1},S_{j-1}$; we do not show the symmetric weights).}
+  \label{tab:stackingmodel}
+  \begin{tabular}{c@{\quad}|@{\quad}c@{\quad}c@{\quad}c@{\quad}c@{\quad}c@{\quad}c@{\quad}c@{\quad}c}
+       & AU & CG & GC & GU & UA & UG \\\hline
+    AU &
+-0.18826 &
+-1.13291 &
+-1.09787 &
+-0.38606 &
+-0.26510 &
+-0.62086
+    \\
+    CG &
+-1.11752 &
+-2.23740 &
+-1.89434 &
+-1.22942 &
+-1.10548 &
+-1.44085
+    \\
+    GU &
+-0.55066 &
+-1.26209 &
+-1.58478 &
+-0.72185 &
+-0.49625 &
+-0.68876
+    \\
+  \end{tabular}
+\end{table}
+
+\section{Contribution of terms to the MultiDefect score}
+
+Figure~\ref{fig:contribtions-multidefect} breaks down the multi-defect
+score of Eq.~$(\ref{eq:blueprintobjective})$, which is used to
+evaluate the quality of multi-target designs. Recall that the total
+score is composed of the term $\frac{1}{k} \sum_{\ell=1}^{k} (E(S,
+R_\ell) - \EnsE(S))$, which evaluates the distances of the energies of
+target structures to the minimum free energy (Distance to MFE), and
+the term $\frac{1}{2\binom{k}{2}} \sum\limits_{1\leq\ell<j\leq
+  k}|E(S,R_\ell) - E(S,R_j)|$, which evaluates the similarity of the
+target structure energies of the target structures (Heterogeneity).
+
+\label{appsec:multidefect}
+\begin{figure}[h!]
+\includegraphics[width=.48\textwidth]{Figs/statistics-Term1}
+\includegraphics[width=.48\textwidth]{Figs/statistics-Term2}
+\caption{Break-down of average MultiDefect values into their two components.}
+\label{fig:contribtions-multidefect}
+\end{figure}
+
 \section{Approximate counting and random generation}
@@
 \end{figure}
-
-\begin{revision}
-  \section{Run times for generating samples}
-
-To illustrate the efficiency and practical feasibility of our Boltzmann sampling approach, we show the distributions of run times for generating 1000 samples (over the Modena benchmark instances and grouped by benchmark sub-set).
-Since producing a controlled Boltzmann sampling requires significant pre-computation, it is informative to separately show construction and sampling times (Figure~\ref{fig:run-times-redprint}).
-
-\todo[inline]{@Stefan: could you please include technical data about the machines used for the timings?
-Please also check the claims/statements in the text for correctness.
-If necessary---please edit.}
-
-A fair comparison to sampling times by other tools can hardly be given, since the other tools produce samplings of very different nature.
-In particular, Modena does not make any guarantees on properties of the sampling.
-Non-surprisingly, this allows it to sample significantly faster (several orders of magnitudes).
-
-The comparison to run times of \RNAblueprint{} is still interesting, since this tool generates---similarly to our tool---\emph{controlled} samples.
-Nevertheless, as we argue the Boltzmann sampling is fundamentally more flexible than uniform sampling as generated by \RNAblueprint.
-Moreover, in this concrete case, \ourprog{} samples based on the rather complex stacking energy model, which introduces much stronger dependencies than required for uniform sampling.
-Technically, these stronger dependencies increase the tree width; here, the increase is exactly the same as the one from the base pair to the stacking model as reported in Figure~\ref{}.
-In this light, \ourprog{} performs even particularly well, beating \RNAblueprint{} most of the time with only a few exceptions for particularly hard instances.
-The results for \RNAblueprint{} are shown in Figure~\ref{fig:run-times-rbp}.
-
-\begin{figure}[tbh]
-  \centering
-  \includegraphics[width=0.45\textwidth]{Figs/RunTimes/construction_time_sets}\quad%
-  \includegraphics[width=0.45\textwidth]{Figs/RunTimes/sample_time_sets}
-  \caption{\todo[inline]{caption Stefan?}}
-  \label{fig:run-times-redprint}
-\end{figure}
-\begin{figure}[tbh]
-  \centering
-  \includegraphics[width=0.45\textwidth]{Figs/RunTimes/construction_time_sets_rbp}\quad%
-  \includegraphics[width=0.45\textwidth]{Figs/RunTimes/sample_time_sets_rbp}
-  \caption{\todo[inline]{caption Stefan?}}
-  \label{fig:run-times-rbp}
-\end{figure}
-
-\end{revision}
 
@@
 
-    \section{Illustrating multi-dimensional Boltzmann sampling for three pseudoknot-free target structures}
-
-    In analogy to Fig.~\ref{fig:energydist-pk} of the main text,
-    Fig.~\ref{fig:energydist} illustrates our multi-dimensional sampling
-    strategy for three pseudoknot-free target structures (example instance
-    from the \texttt{Modena} benchmark). The figure shows Turner energy
-    distributions of single structures from sampling with different weight
-    parameters, where weights are trained to target specific energies. For
-    comparison, the figure shows distributions from uniform and Boltzmann
-    sampling at respective homogenous weights $1$ and $e^\beta$.
-
-    Fig.~\ref{fig:gc-content} shows the distributions of the GC content of the samples, while targeting different combinations of energies in this experiment.
-For completeness, we also report corresponding results for the analogous experiment from the main text.
-We observe that in addition to the target energies, also the feature $\GCb\%$ is narrowly defined by the multi-dimensional Boltzmann sampling strategy.
-In comparison, for uniform sampling and Boltzmann sampling at high weight $e^\beta$, where the \GCb\% is not controlled, we obtain much broader distributions.
-
-    \begin{figure*}[t]
-      \begin{center}
-        %{\sf \bfseries A}\includegraphicstop[width=0.32\textwidth]{Figs/offset2}\hspace{.5cm}
-        %{\sf \bfseries B}
-        \includegraphicstop[width=0.8\textwidth]{Figs/energy_distribution}
-      \end{center}
-      \caption{%
-        Targeting specific energies using multi-dimensional Boltzmann
-        sampling. We show the Turner energy distributions for
-        three target structures R1, R2 and R3. By
-        targeting the respective free energies $(-40,-40,-20)$ and
-        $(-20,-20,-20)$ for the target structures (annotated in red and
-        blue, resepectively), we demonstrate the
-        effectivity of our adaptive multi-dimensional Boltzmann sampling
-        procedure; moreover, for comparison, the distributions for uniform
-        (U) and Boltzmann (B) samples---respectively associated with
-        homogenous weights $1$ and $e^\beta$.
-%    ($\delta=0.05, \delta'=0.1, \gamma=1.1, \psi=20, n=1000$)
-  }
-  \label{fig:energydist}
-\end{figure*}
-
-\begin{figure}
-  \begin{center}
-    {\sf \bfseries A}~\includegraphicstop[width=0.45\textwidth]{Figs/Plots/PKB00211_PKB00239_0_GC_content}
-    ~~~~~{\sf \bfseries B}~\includegraphicstop[width=0.45\textwidth]{Figs/Plots/supp_3str_GC_content}
-  \end{center}
-  \caption{Distribution of the \GCb-content \GCb\% in the experiments of (A) Fig.~\ref{fig:energydist-pk} of the main text and (B) Suppl.~Fig.~\ref{fig:energydist}.}
-  \label{fig:gc-content}
-\end{figure}
-
-\section{Parameters for the base pair and the stacking energy model}
-\label{appsec:modelparameters}
-
-We trained parameters for two RNA energy models to approximate the
-Turner energy model, as implemented in the \Software{ViennaRNA}
-package.  In the \Def{base pair model}, the total energy of a sequence $S$
-for an RNA structure $R$ is given as sum of base pair energies, where
-we consider six types of base pairs distinguishing by the bases $\Ab$-$\Ub$,
-$\Cb$-$\Gb$ or $\Gb$-$\Ub$ (symmetrically) and between stacked and non-stacked base
-pairs; here, we consider base pairs $(i,j)\in R$ \Def{stacked} iff
-$(i+1,j-1)\in R$, otherwise \Def{non-stacked}. In the \Def{stacking model},
-our features are defined by the stacks, i.e. pairs $(i,j)$ and
-$(i+1,j-1)$ which both occur in $R$; we distinguish 18 types based on
-$S_i,S_j,S_{i+1},S_{j-1}$ (i.e. all combinations that allow canonical
-base pairs; the configurations $S_i,S_j,S_{i+1},S_{j-1}$ and
-$S_{i+1},S_{j-1},S_j,S_i$ are symmetric).
-
-Both models describe the energy assigned to a pair of sequence and
-structure in linear dependency of the number of features and their
-weights. We can thus train weights for linear predictors of the Turner
-energy in both models.
-
-For this purpose, we generated 5000 uniform random RNA sequences of
-random lengths between 100 and 200. For each sample, we predict the
-minimum free energy and corresponding structure using the ViennaRNA
-package; then, we count the distinguished features (i.e., base pair or
-stack types). The parameters are estimated fitting linear models
-without intercept (\texttt{R} function \texttt{lm}). For both models,
-\texttt{R} reports an adjusted R-squared value of 0.99. The resulting
-parameters are reported in Tables~\ref{tab:basepairmodel} and
-\ref{tab:stackingmodel}.
-
-For validating the trained parameters, we generate a second
-independent test set of random RNA sequences in the same
-way. Fig.~\ref{fig:training-cor} shows correlation plots for the
-trained parameters in the base pair and stacking models for predicting
-the Turner energies in the test set with respective correlations of
-$0.95$ an $0.94$.
-
-\begin{figure}
-  \centering
-  A)\includegraphicstop[width=0.4\textwidth,trim=0 0 0 50,clip]{Figs/basepaircor}
-  B)\includegraphicstop[width=0.4\textwidth,trim=0 0 0 50,clip]{Figs/stackingcor}
-  \caption{Validation of the trained parameters for the A) base pair
-    model B) stacking model for predicting energies of the
-    independently sampled sequences in the test set. We show
-    correlation plots against the Turner energies reported by the
-    ViennaRNA package.}
-  \label{fig:training-cor}
-\end{figure}
-
-\begin{table}[b]
-  \centering
-  \caption{Trained weights for the base pair energy model.}
-  \label{tab:basepairmodel}
-  \begin{tabular}{c@{\quad}c@{\quad}c@{\quad}|@{\quad}c@{\quad}c@{\quad}c}
-    \multicolumn{3}{c}{non-stacked} & \multicolumn{3}{c}{stacked}\\
-    AU      & CG       & GU      & AU       & GC       & GU \\\hline
-    1.26630 & -0.09070 & 0.78566 & -0.52309 & -2.10208 & -0.88474
-  \end{tabular}
-\end{table}
-
-% GC_IN = -0.09070;
-% AU_IN = 1.26630;
-% GU_IN = 0.78566;
-% GC_TERM = -2.10208;
-% AU_TERM = -0.52309;
-% GU_TERM = -0.88474;
-
-\begin{table}[b]
-  \centering
-  \caption{Trained weights for the stacking energy model (the rows specify $S_i,S_j$; the columns, $S_{i+1},S_{j-1}$; we do not show the symmetric weights).}
-  \label{tab:stackingmodel}
-  \begin{tabular}{c@{\quad}|@{\quad}c@{\quad}c@{\quad}c@{\quad}c@{\quad}c@{\quad}c@{\quad}c@{\quad}c}
-       & AU & CG & GC & GU & UA & UG \\\hline
-    AU &
--0.18826 &
--1.13291 &
--1.09787 &
--0.38606 &
--0.26510 &
--0.62086
-    \\
-    CG &
--1.11752 &
--2.23740 &
--1.89434 &
--1.22942 &
--1.10548 &
--1.44085
-    \\
-    GU &
--0.55066 &
--1.26209 &
--1.58478 &
--0.72185 &
--0.49625 &
--0.68876
-    \\
-  \end{tabular}
-\end{table}
-
 \section{Monotonicity of the partial derivatives within weight calibration}
@@
 
-\section{Contribution of terms to the MultiDefect score}
-\label{sec:improvements}
-\begin{figure}[h!]
-\includegraphics[width=.48\textwidth]{Figs/statistics-Term1}
-\includegraphics[width=.48\textwidth]{Figs/statistics-Term2}
-\caption{Break-down of average MultiDefect values into their two components.}
-\end{figure}
-
-\section{Immediate benefits of positive design for negative design}
-
-Already previous work on RNA Design, e.g. INFO-RNA and Incarnation, observed
-that---put briefly---negative design benefits from positive design. Here, more
-concretely, we simply \emph{demonstrate} this effect (for small examples), where
-we consider \Def{negative design} as  
-designing RNA sequences that fold with high probability $Pr(R|S)$ into the
-given target structure $R$; and \Def{positive design}, as generating one or
-many sequences that have a low energy for $R$.
-
-For each of two small examples, a single-stem and a double-stem structure, we produce a uniform and a weighted Boltzmann
-sampling, each of 10000 sequenes. For each sampled sequence $S$, we determine
-the Turner energy $E(S,R)$ as well as the
-%difference $E_\Delta(S,R)$ of $E(S,R)$ to the ensemble energy $E_{\text{ens}}(S)$ of $S$ 
-probability $Pr(R|S)$ of the structure $R$ in the ensemble of sequence $S$ 
-(where the energies are calculated in the Turner model using RNAfold).
-
-Note that low energy $E(S|R)$ is (non-linearly) correlated with high
-probability $Pr(S|R)$, the probability to see $S$ in the Boltzmann distribution
-of sequences folding into $R$. Recall that the probability $Pr(R|S)$ can be
-calculated as $\exp(-E(S,R)/RT)/Z(S)$,
-where $Z(S)=\sum_{R'} \exp(-E(S,R')/RT)$.
-%
-Consequently, Figure~\ref{fig:negforpos} ultimately illustrates the dependencies
-between $Pr(S|R)$ and $Pr(R|S)$ for uniform samples and low energy samples
-(where those samples can be efficiently generated due to our Boltzmann sampling
-strategy). For our examples, we observe (Fig.~\ref{fig:negforpos}) that Boltzmann sampling with weight 500 (green), generates good negative designs 
-with substantially higher probability than uniform sampling (blue).
-
-% and high energy difference $E_\Delta(S,R)=E_{\text{ens}}(S)-E(S,R)$, with high probability $\Pr(R|S)$.
-% 
-% To see the latter, recall that the \Def{ensemble energy $E_{\text{Ens}}(S)$} is defined as
-% $$
-% E_{\text{ens}}(S) = -RT \ln( Z(S) ),
-% $$ 
-% based on the partition function $Z(S)=\sum_{R'} \exp(-E(S,R')/RT)$
-% such that 
-% $$
-% E(S,R) - E_{\text{ens}}(S) 
-% =
-% -RT \ln Pr(R|S),
-% $$  
-% which can be derived by term transformation 
-% \begin{align*}
-% -RT \ln P(R|S) &= -RT \ln [ exp(-E(S,R)/RT) / Z(S) ] \\&= -RT \ln (exp(-E(S,R)/RT)) - (-RT\ln Z(S))
-% \\&=  E(S,R) - E_{\text{Ens}}(S). 
-% \end{align*}
-
-\begin{figure}
-\textbf{A}~\includegraphicstop[width=0.48\textwidth]{Figs/Plots/stem_weight1-500_mean}%
-\textbf{B}~\includegraphicstop[width=0.48\textwidth]{Figs/Plots/doublestem_weight1-500_mean}%
-\caption{Comparison of uniform sampling and Boltzmann sampling with
-weight $\pi=500$. We show results for two different small structures: (A) single stem structure (B) double stem structure. For each respective structure $R$, we plot the energies $E(S,R)$ of
-10000 sampled sequences against the probabilities $Pr(R|S)$ (in log scale).
-For illustrating the trend, we show moving averages over windows of $\pm 1$ kcal.
-Note that while only the energy $E(S,R)$ is directly controlled by Boltzmann sampling, this strongly impacts the common negative design objective $Pr(R|S)$.}
-\label{fig:negforpos}
-\end{figure}
-
 \section{Detailed result of MultiDefect analysis}\label{sec:validity}

### 2019-01-22 BMC-Response-Round1.tex
Address Reviewer 3, comment 1

@@
 
-In particular, the reviews helped us to state the goal of this work more clearly.
+In particular, the reviews helped us to state the goal of this work more clearly and improve its presentation.
 While we demonstrate the applicability of our approach to negative RNA design, which (here by optimizing 'MultiDefect') aims at high and similar probability of the target structures, the focus of this work is the positive design of sequences that fold into multiple target structures with highly specific properties.
@@
 why these two examples are output, and the other examples for RF3 exist or not?\\[1em]}%
-\Answer{This figure illustrates the impact of the energy model on the tree-decomposition, a simple energy model leading to a much simpler tree-decomposition than the stacking model.
-We now describe more explicitly
-}
-\AnswerTODO[Sebastian]{
-Add a description of tree decompositions and why they differ + New Illustration
+\Answer{
+  We addressed this comment in two ways: First, we added an intuitive explanation of tree decomposition i(after its formal definition) together with an illustration of a simple tree decomposition to the main text.
+  Second, we rewrote the corresponding text in the supplement to further clarify the purpose of the figure and provide additional explanations.
 }

### 2019-01-22 supplement.tex
Address Reviewer 3, comment 1

@@
 
+\newcommand{\EnsE}{\text{\rm G}} %ensemble energy
+
 Figure~\ref{fig:contribtions-multidefect} breaks down the multi-defect
@@
 corresponding run-times of the tree decomposition.
-Fig.~\ref{fig:td-example} shows the tree decompositions for an example
-instance from set RF3.
+
+\revised{
+  Finally, for further illustration, we show concrete tree decompositions for a selected instance in the benchmark set RF3 in Figure~\ref{fig:td-example}.
+Note that this figure is provided solely for illustrational purposes:
+  First, it depicts the tree decomposition for a typical, realistic sampling task. Second, it exemplarily illustrates the typical impact of the energy model on the tree decomposition.
+  Compared to the tree decomposition based on the stacking model, the tree decomposition for the base pair model is much simpler.
+This shows in many connected components, which correspond to independently solvable sub-problems, and fewer positions per bag (i.e.
+smaller tree width), which allows far more efficient evaluation per tree node.
+  Note that the complexity of the tree decomposition ultimately shows the complexity of the dependencies that need to be considered for simultaneous energy evaluation for all target structures.
+This complexity is substantially increased in the stacking model over the simpler base pair model.
+}
 
@@
 
-\section{Detailed result of MultiDefect analysis}\label{sec:validity}
+\section{Detailed results of MultiDefect analysis}\label{sec:validity}
 N/A values correspond to data that could not be obtained by the time of this submission for two main reasons: In the case of the initial sampling (left columns), they correspond to instances which, in conjunction with an expressive energy model, resulted in very high tree-width, leading to unreasonable memory requirements incompatible with our available computing resources; The case of missing values after optimization  (right columns), they indicate situations where the initial optimization took too long ($\approx$ 1 day) and was killed.

### 2019-01-22 RNARedPrint.tex
Merge
Edits in intro, contributions, Fig1 caption

@@
   \parttitle{Results} We devise a novel approach to a central ingredient of most in-silico
-  design methods: the generation of sequences with multiple structural targets. For this task, we present the efficient (fixed-parameter tractable) approach \ourprog{}.
-  % 
-  Modeling
-  the problem by specializing constraint networks, \ourprog{} supports generic
+  design methods: the generation of sequences \revised{that fold well into multiple target structures.} For this task, we present the fixed-parameter tractable (FPT) approach \ourprog{}.
+  % 
+  \revised{Based on} constraint networks, \ourprog{} supports generic
   Boltzmann-weighted sampling, which enables the positive design of RNA sequences
@@
 %
-  Securing the theoretical foundations of our method, we establish the $\#${\sf P}-hardness of uniform sampling. 
-%
-  Finally, we empirically study general properties of the approach and generate
+  %Securing the theoretical foundations of our method
+  Moreover, we establish the $\#${\sf P}-hardness of \revised{counting to theoretically justify the development of an FPT approach}. 
+%
+  Finally, we study general properties of our approach empirically and generate
   biologically relevant multi-target Boltzmann-weighted designs for a
-  common design benchmark, including pseudoknots.
-  Generating seed sequences with \ourprog{}, we demonstrate significant improvements over the previously best multi-target sampling strategy.\\
+  common design benchmark.
+  We demonstrate significant improvements over the previously best multi-target sampling strategy, due to generating seed sequences with \ourprog{}.\\
 %
@@
 While the quality of the seeds was
-found to be performance critical for such RNA design
+found to be performance-critical for such RNA design
 methods~\citep{Levin2012}, random seed generation
@@
 \parHead{Specificities and similarities of multi-target design.}
-Available methods for multi-target RNA design~\citep{Lyngsoe2012,HoenerzuSiederdissen2013,Taneda2015,Hammer2017} all follow the same overall generate-and-optimize strategy.
-%
-Faced with the complex constraints due to the multiple targets, early methods such as \Software{Frnakenstein}~\citep{Lyngsoe2012} and \Software{Modena}~\citep{Taneda2015} passed on any systematic generation of sequences, but rely on ad-hoc sampling strategies.
+The available methods for multi-target RNA design~\citep{Lyngsoe2012,HoenerzuSiederdissen2013,Taneda2015,Hammer2017} all follow the same overall generate-and-optimize strategy.
+%
+Faced with the complex constraints due to the multiple targets, early methods such as \Software{Frnakenstein}~\citep{Lyngsoe2012} and \Software{Modena}~\citep{Taneda2015} \revised{do not even attempt to sample sequences systematically from a controlled distribution}, but rely on ad-hoc generation strategies.
 %
@@
 
-\todo[inline]{SW: The following paragraph (together with the first sentence of Subsection Contributions) is rather confusing than helpful.
+%\todo[inline]{SW: The following paragraph (together with the first sentence of Subsection Contributions) is rather confusing than helpful.
 We could drop this here and give brief explanations of the complexity issues in the contributions section.}
-Better understanding of the true complexity bounds is highly relevant for the application of such methods, in particular for applications in more expressive settings (e.g.
+%
+%Better understanding of the true complexity bounds is highly relevant for the application of such methods, in particular for applications in more expressive settings (e.g.
 Boltzmann sampling as discussed in this work or sampling satisfying multiple additional constraints and properties).
@@
 \end{center}
-\caption{General outline of \ourprog{} for base pair-based energy models.
-From a set of target secondary structures (i), base-pairs are merged (ii) into a (base pair) dependency graph (iii) and transformed into a tree decomposition (iv).
+\caption{General outline of \ourprog{}.
+From a set of target secondary structures (i), base pairs are merged (ii) into a (base pair) dependency graph (iii) and transformed into a tree decomposition (iv).
 The tree is then used to compute the partition function, followed by a Boltzmann sampling of valid sequences (v).
-An adaptive scheme learns weights to achieve targeted energies and \GCb\%, leading to the production of suitable designs (vi).}
+An adaptive scheme learns weights to achieve targeted energies and \GCb\%, leading to the production of suitable designs (vi).
+\revised{For simplicity, we assume that only dependencies between the ends of base pairs have to be considered.
+Our computations based on a more complex energy model (stacking model) require additional dependencies.}
+  }
 \label{fig:workflow}
@@
 
-Addressing these questions in more detail and generality leads to the practical advances in this work, which extend the expressivity, flexibility, and performance of multi-structure design.
-Namely, we introduce a generic framework, summarized in~Fig.~\ref{fig:workflow}, enabling efficient Boltzmann-weighted sampling over RNA sequences with multiple target structures.
-The main algorithmic contribution of this work consists in dynamic programming algorithms, based on the concept of \Def{tree decomposition}, to compute the dual partition function, and sample sequences from the Boltzmann distribution%
+\revised{As central main contribution, we provide a systematic and flexibly extensible technique for sampling that targets multiple versatile features.
+For the sake of clarity, we introduce this method specialized to the sampling of sequences that target the energies of multiple structures and the \GCb\%.}
+
+\revised{Fig.~\ref{fig:workflow} summarizes our generic framework, which enables} efficient Boltzmann-weighted sampling over RNA sequences with multiple target structures.
+The main \emph{algorithmic} contribution of this work consists of dynamic programming (DP) algorithms, based on the concept of \Def{tree decomposition}, to compute partition functions, and sample sequences from the Boltzmann distribution%
 % (Subsection~\ref{sec:PF})
 .
-The complexities of our algorithms depend exponentially on a specific property of the tree decomposition, called the \Def{treewidth}, whose value remains typically low for typical instances.
-However, for any given value of the treewidth, the complexity only scales linearly with the size of designed sequences and the number of targeted structures, \emph{i.e.} our algorithms are Fixed-Parameter Tractable (FPT).
+\revised{Generally, tree decompositions are data structures that capture the specific dependencies of a problem instance (here, the dependencies induced by the target structures) to guide the efficient processing by DP algorithms.}
+%
+\revised{Building on this principle, }the complexities of our algorithms depend exponentially on a specific property of the tree decomposition, called the \Def{treewidth}, which remains low for typical instances.
+However, for any given value of the treewidth, the complexity scales only linearly with the size of designed sequences and the number of targeted structures, \emph{i.e.} our algorithms are fixed-parameter tractable (FPT).
 
@@
 Turner RNA energy model
-works well by performing sampling based on less complex RNA energy models, which keeps the treewidth low.
-This result is highly relevant in practice, since it allows to combine high efficiency with high accuracy and moreover leaves room for extensions of the method.
-
-Eventually this strategy allows generating biologically relevant
-multi-target designs, as demonstrated by its leading performance on a comprehensive benchmark set for multi-target RNA design~\citep{Taneda2015}.
+works well by performing sampling based on \revised{a simplified RNA energy model, which induces lower treewidth}.
+This result is highly relevant in practice, since it allows to combine high efficiency with high accuracy (and even leaves room for \revised{future extensions supporting additional constraints}).
+
+\revised{Eventually our proof-of-concept results 
+ on a comprehensive multi-target RNA design benchmark~\citep{Taneda2015} 
+  suggest that our sampling strategy supports designing biologically relevant
+RNAs for multiple targets.}
 
@@
 %\paragraph{Central computational problem.}
-Let us consider $m$ (secondary) structures $R_1,\dots, R_m$, each abstracted as a set of base pairs, and $k$ features $F_1,\dots, F_k$, associated with weights $\pi_1,\dots,\pi_k$, our goal is to sample from the Boltzmann distribution, defined over the set of sequences that satisfy classic base-pairing rules for each structure as:
+Let us consider $m$ (secondary) structures $R_1,\dots, R_m$, each abstracted as a set of base pairs, and $k$ features $F_1,\dots, F_k$, associated with weights $\pi_1,\dots,\pi_k$, our goal is to sample from the Boltzmann distribution, defined over the set of sequences that satisfy classic base pairing rules for each structure as:
 \begin{equation}
@@
 \revised{Intuitively, a tree decomposition for an (hyper)graph $G$ is a tree structure which features all the vertices of $G$, captures all the (hyper)edges of $G$, and helps ensure some notion of consistency by allowing the propagation of values.
-Figure~\ref{fig:treedec} illustrates shows an optimal tree decomposition for a pair of structures under the stacking energy model.}
+Figure~\ref{fig:treedec} shows an optimal tree decomposition for a pair of structures under the stacking energy model.}
 \begin{definition}[Tree width]
@@
  \KwData{Cluster tree $(T,\chi,\phi)$ and partition functions $\Message{u'}{v'}$ for all $(u'\to{}v')\in T$.}
- \KwResult{Boltzmann-distributed random sequence $\val$.}
+ \KwResult{One random sequence $\val$ sampled from the Boltzmann distribution}
  $\val$ \gets $\varnothing$\;

### 2019-01-23 supplement.tex
Updates from ShareLaTeX

@@
 
-To illustrate the efficiency and practical feasibility of our Boltzmann sampling approach, we show the distributions of run times for generating 1000 samples (over the Modena benchmark instances and grouped by benchmark sub-set).
+To illustrate the efficiency and practical feasibility of our Boltzmann sampling approach, we show the distributions of run times (Intel(R) Core(TM) i7-4770 CPU) for generating 20.000 samples (over the Modena benchmark instances and grouped by benchmark sub-set).
 Since producing a controlled Boltzmann sampling requires significant pre-computation, it is informative to separately show construction and sampling times (Figure~\ref{fig:run-times-redprint}).
-
-\todo[inline]{@Stefan: could you please include technical data about the machines used for the timings?
-Please also check the claims/statements in the text for correctness.
-If necessary---please edit.}
 
@@
   \includegraphics[width=0.45\textwidth]{Figs/RunTimes/construction_time_sets}\quad%
-  \includegraphics[width=0.45\textwidth]{Figs/RunTimes/sample_time_sets}
+  \includegraphics[width=0.45\textwidth]{Figs/RunTimes/sample_time_sets}\\
+  \includegraphics[width=0.45\textwidth]{Figs/RunTimes/construction_time_sets_bp}\quad%
+  \includegraphics[width=0.45\textwidth]{Figs/RunTimes/sample_time_sets_bp}
   \caption{\todo[inline]{caption Stefan?}}

### 2019-01-23 RNARedPrint.tex

@@
 %\documentclass[doublespacing,linenumbers]{bmcart}% uncomment this for twocolumn layout and comment line below
-\documentclass[twocolumn]{bmcart}
+\documentclass[]{bmcart}
+%\documentclass[twocolumn]{bmcart}
 
@@
 An adaptive scheme learns weights to achieve targeted energies and \GCb\%, leading to the production of suitable designs (vi).
-\revised{For simplicity, we assume that only dependencies between the ends of base pairs have to be considered.
+\revised{Note that for simplicity, we assume in this figure that only dependencies between the ends of base pairs are considered to evaluate the energies of structures.
 Our computations based on a more complex energy model (stacking model) require additional dependencies.}
@@
 
-\revised{As central main contribution, we provide a systematic and flexibly extensible technique for sampling that targets multiple versatile features.
+\revised{As our central contribution, we provide a systematic and flexibly extensible technique for sampling that targets multiple versatile features.
 For the sake of clarity, we introduce this method specialized to the sampling of sequences that target the energies of multiple structures and the \GCb\%.}
@@
 \revised{Fig.~\ref{fig:workflow} summarizes our generic framework, which enables} efficient Boltzmann-weighted sampling over RNA sequences with multiple target structures.
-The main \emph{algorithmic} contribution of this work consists of dynamic programming (DP) algorithms, based on the concept of \Def{tree decomposition}, to compute partition functions, and sample sequences from the Boltzmann distribution%
+\revised{\emph{Algorithmically}, we contribute} dynamic programming (DP) algorithms, based on the concept of \Def{tree decomposition}, to compute partition functions and sample sequences from the Boltzmann distribution%
 % (Subsection~\ref{sec:PF})
@@
 
-This exponential dependency is arguably necessary.
-Indeed, the underlying counting problem is \#{\sf P}-hard, meaning that there is no efficient, polynomial time algorithm for the problem unless ${\sf P}={\sf NP}$.
-Our result relies on a surprising bijection between valid sequences and independent sets of a bipartite graph, the latter being the object of recent breakthroughs in approximate counting complexity~\citep{Bulatov2013,Cai2016}.
+\revised{Remarkably, we show that it is not possible to find a better, efficient method for sampling (unless ${\sf P}={\sf NP}$), since the underlying counting problem is \#{\sf P}-hard.
+The practical relevance of this result is that it rules out substantially better sampling techniques, such that there will always remain---in practice---an upper limit on the tractable number of structures and their heterogeneity, the complexity of the directly treatable energy model, and the number and complexity of additional constraints that could be considered in future sampling-based applications.} Technically, this result relies on a surprising bijection between valid sequences and independent sets of a bipartite graph, the latter being the object of recent breakthroughs in approximate counting complexity~\citep{Bulatov2013,Cai2016}.
 

### 2019-01-23 supplement.tex
Updates from ShareLaTeX

@@
 Since producing a controlled Boltzmann sampling requires significant pre-computation, it is informative to separately show construction and sampling times (Figure~\ref{fig:run-times-redprint}).
+Moreover, \ourprog{} can be used with a simpler base pair model or a rather complex stacking energy model, which introduces much stronger dependencies than required for uniform sampling.
+Technically, these stronger dependencies increase the tree width and thus the runtime, as noticeable in Figure~\ref{fig:run-times-redprint}.
 
@@
 Non-surprisingly, this allows it to sample significantly faster (several orders of magnitudes).
-
-The comparison to run times of \RNAblueprint{} is still interesting, since this tool generates---similarly to our tool---\emph{controlled} samples.
-Nevertheless, as we argue the Boltzmann sampling is fundamentally more flexible than uniform sampling as generated by \RNAblueprint.
-Moreover, in this concrete case, \ourprog{} samples based on the rather complex stacking energy model, which introduces much stronger dependencies than required for uniform sampling.
-Technically, these stronger dependencies increase the tree width; here, the increase is exactly the same as the one from the base pair to the stacking model as reported in Figure~\ref{}.
-In this light, \ourprog{} performs even particularly well, beating \RNAblueprint{} most of the time with only a few exceptions for particularly hard instances.
+Nevertheless, the comparison to run times of \RNAblueprint{} is still interesting, since this tool generates---similarly to our tool---\emph{controlled} samples.
+As we argue that the Boltzmann sampling is fundamentally more flexible than uniform sampling as generated by \RNAblueprint, \ourprog{} performs even particularly well, beating \RNAblueprint{} most of the time with only a few exceptions for particularly hard instances.
 The results for \RNAblueprint{} are shown in Figure~\ref{fig:run-times-rbp}.
@@
   \includegraphics[width=0.45\textwidth]{Figs/RunTimes/sample_time_sets_bp}
-  \caption{\todo[inline]{caption Stefan?}}
+  \caption{Boxplots showing the run-time distribution of sequence sampling with \ourprog{} for the various benchmark sets.
+The times are split into construction time (left) and sampling time (right).
+Construction includes all the pre-computation, while the latter is the time needed to generate 20.000 sequences.
+Furthermore, we distinguish between two energy models, stacking (top) and basepair (bottom).
+}
   \label{fig:run-times-redprint}
@@
   \includegraphics[width=0.45\textwidth]{Figs/RunTimes/sample_time_sets_rbp}
-  \caption{\todo[inline]{caption Stefan?}}
+  \caption{Boxplots showing the run-time distribution of sequence sampling with \RNAblueprint{} for the various benchmark sets.
+The times are split into construction time (left) and sampling time (right).
+Construction includes all the pre-computation, while the latter is the time needed to generate 20.000 sequences.}
   \label{fig:run-times-rbp}
@@
   Finally, for further illustration, we show concrete tree decompositions for a selected instance in the benchmark set RF3 in Figure~\ref{fig:td-example}.
-Note that this figure is provided solely for illustrational purposes:
+This figure is provided solely for illustrational purposes:
   First, it depicts the tree decomposition for a typical, realistic sampling task. Second, it exemplarily illustrates the typical impact of the energy model on the tree decomposition.

### 2019-01-23 supplement.tex
fig change

@@
   \includegraphics[width=0.45\textwidth]{Figs/RunTimes/construction_time_sets}\quad%
-  \includegraphics[width=0.45\textwidth]{Figs/RunTimes/sample_time_sets}\\
-  \includegraphics[width=0.45\textwidth]{Figs/RunTimes/construction_time_sets_bp}\quad%
-  \includegraphics[width=0.45\textwidth]{Figs/RunTimes/sample_time_sets_bp}
+  \includegraphics[width=0.45\textwidth]{Figs/RunTimes/sample_time_sets}
   \caption{Boxplots showing the run-time distribution of sequence sampling with \ourprog{} for the various benchmark sets.
@@
 Construction includes all the pre-computation, while the latter is the time needed to generate 20.000 sequences.
-Furthermore, we distinguish between two energy models, stacking (top) and basepair (bottom).
 }

### 2019-01-23 supplement.tex
Suppl section about seed generation

@@
 
+\section{Immediate benefits of positive design for negative design}
+
+Already previous work on RNA Design, e.g. INFO-RNA and Incarnation, observed
+that---put briefly---negative design benefits from positive design. Here, more
+concretely, we simply \emph{demonstrate} this effect (for small examples), where
+we consider \Def{negative design} as  
+designing RNA sequences that fold with high probability $Pr(R|S)$ into the
+given target structure $R$; and \Def{positive design}, as generating one or
+many sequences that have a low energy for $R$.
+
+For each of two small examples, a single-stem and a double-stem structure, we produce a uniform and a weighted Boltzmann
+sampling, each of 10000 sequenes. For each sampled sequence $S$, we determine
+the Turner energy $E(S,R)$ as well as the
+%difference $E_\Delta(S,R)$ of $E(S,R)$ to the ensemble energy $E_{\text{ens}}(S)$ of $S$ 
+probability $Pr(R|S)$ of the structure $R$ in the ensemble of sequence $S$ 
+(where the energies are calculated in the Turner model using RNAfold).
+
+Note that low energy $E(S|R)$ is (non-linearly) correlated with high
+probability $Pr(S|R)$, the probability to see $S$ in the Boltzmann distribution
+of sequences folding into $R$. Recall that the probability $Pr(R|S)$ can be
+calculated as $\exp(-E(S,R)/RT)/Z(S)$,
+where $Z(S)=\sum_{R'} \exp(-E(S,R')/RT)$.
+%
+Consequently, Figure~\ref{fig:negforpos} ultimately illustrates the dependencies
+between $Pr(S|R)$ and $Pr(R|S)$ for uniform samples and low energy samples
+(where those samples can be efficiently generated due to our Boltzmann sampling
+strategy). For our examples, we observe (Fig.~\ref{fig:negforpos}) that Boltzmann sampling with weight 500 (green), generates good negative designs 
+with substantially higher probability than uniform sampling (blue).
+
+% and high energy difference $E_\Delta(S,R)=E_{\text{ens}}(S)-E(S,R)$, with high probability $\Pr(R|S)$.
+% 
+% To see the latter, recall that the \Def{ensemble energy $E_{\text{Ens}}(S)$} is defined as
+% $$
+% E_{\text{ens}}(S) = -RT \ln( Z(S) ),
+% $$ 
+% based on the partition function $Z(S)=\sum_{R'} \exp(-E(S,R')/RT)$
+% such that 
+% $$
+% E(S,R) - E_{\text{ens}}(S) 
+% =
+% -RT \ln Pr(R|S),
+% $$  
+% which can be derived by term transformation 
+% \begin{align*}
+% -RT \ln P(R|S) &= -RT \ln [ exp(-E(S,R)/RT) / Z(S) ] \\&= -RT \ln (exp(-E(S,R)/RT)) - (-RT\ln Z(S))
+% \\&=  E(S,R) - E_{\text{Ens}}(S). 
+% \end{align*}
+
+\begin{figure}[h!]
+\textbf{A}~\includegraphicstop[width=0.48\textwidth]{Figs/Plots/stem_weight1-500_mean}%
+\textbf{B}~\includegraphicstop[width=0.48\textwidth]{Figs/Plots/doublestem_weight1-500_mean}%
+\caption{Comparison of uniform sampling and Boltzmann sampling with
+weight $\pi=500$. We show results for two different small structures: (A) single stem structure (B) double stem structure. For each respective structure $R$, we plot the energies $E(S,R)$ of
+10000 sampled sequences against the probabilities $Pr(R|S)$ (in log scale).
+For illustrating the trend, we show moving averages over windows of $\pm 1$ kcal.
+Note that while only the energy $E(S,R)$ is directly controlled by Boltzmann sampling, this strongly impacts the common negative design objective $Pr(R|S)$.}
+\label{fig:negforpos}
+\end{figure}
+
+\section{Tree decomposition for RNA design instances in practice}
+\label{appsec:treedecomp}
+
+For studying the typically expected treewidths and tree decomposition
+run times in multi-target design instances, we consider five sets of
+multi-target RNA design instances of different complexity. Our first
+set consists of the Modena benchmark instances.
+
+In addition, we generated four sets of instances of increasing
+complexity. The instances of the sets RF3, RF4, RF5, and RF6, each
+respectively specify 3,4,5, and 6 target structures for sequence
+length 100.  For each instance (100 instances per set), we generated a
+set of $k$ ($k=3,\dots,6$) compatible structures as follows
+\begin{itemize}
+\item Generate a random sequence of length 100;
+\item Compute its minimum free energy structure (ViennaRNA package);
+\item Add the new structure to the instances if the resulting base pair dependency graph is bipartite;
+\item Repeat until $k$ structures are collected.
+\end{itemize}
+For each instance, we generated the dependency graphs in the base pair
+model and in the stacking model. Then, we performed tree decomposition
+(using strategy ``GreedyFillIn'' of LibTW~\citep{Dijk2006}) on each dependency
+graph. The obtained treewidths are reported in
+Fig.~\ref{fig:td-widths}, while Fig.~\ref{fig:td-times} shows the
+corresponding run-times of the tree decomposition.
+
+\revised{
+  Finally, for further illustration, we show concrete tree decompositions for a selected instance in the benchmark set RF3 in Figure~\ref{fig:td-example}.
+This figure is provided solely for illustrational purposes:
+  First, it depicts the tree decomposition for a typical, realistic sampling task. Second, it exemplarily illustrates the typical impact of the energy model on the tree decomposition.
+  Compared to the tree decomposition based on the stacking model, the tree decomposition for the base pair model is much simpler.
+This shows in many connected components, which correspond to independently solvable sub-problems, and fewer positions per bag (i.e.
+smaller tree width), which allows far more efficient evaluation per tree node.
+  Note that the complexity of the tree decomposition ultimately shows the complexity of the dependencies that need to be considered for simultaneous energy evaluation for all target structures.
+This complexity is substantially increased in the stacking model over the simpler base pair model.
+}
+
+\begin{figure}
+  %\hspace*{2cm}\textbf{Base pair model}\hspace{4.75cm}\textbf{Stacking model}\\[-28pt]
+  \centering\includegraphics[width=0.9\textwidth]{Figs/td-widths}
+  \caption{Treewidths for multi-target RNA design instances of
+    different complexity. Distributions of treewidths shown as boxplots for the base pair (left) and stacking model (right).}
+  \label{fig:td-widths}
+\end{figure}
+
+\begin{figure}
+  %\hspace*{2cm}\textbf{Base pair model}\hspace{4.75cm}\textbf{Stacking model}\\[-28pt]
+  \centering
+  \includegraphics[width=0.9\textwidth]{Figs/td-times}
+  \caption{Computation time of tree decompositions for
+    multi-target RNA design instances of different complexity.
+    Distributions of times (in ms/instance) shown as boxplots for the base pair (left) and stacking model (right).}
+  \label{fig:td-times}
+\end{figure}
+
+\begin{figure}[t!]
+  %% the figure shows benchmark_inputs/RNAfold/3str/f3.100.0.inp
+  \centering
+  \verbatiminput{Figs/f3_100_0.txt}
+  \includegraphicstop[width=0.6\textwidth]{Figs/td-example-basepair}%
+  \includegraphicstop[width=0.35\textwidth]{Figs/td-example-stacking}
+  \caption{Example instance from RF3 (top) with its tree decompositions in the base pair (left) and stacking model (right). The respective treewidths are 2 and 12.}
+  \label{fig:td-example}
+\end{figure}
+
 \begin{revision}
+
   \section{Run times for generating samples}
@@
 
-\section{Immediate benefits of positive design for negative design}
+\section{Effectivity of the seed sequence generation}
+\todo[inline]{TODO}
 
-Already previous work on RNA Design, e.g. INFO-RNA and Incarnation, observed
-that---put briefly---negative design benefits from positive design. Here, more
-concretely, we simply \emph{demonstrate} this effect (for small examples), where
-we consider \Def{negative design} as  
-designing RNA sequences that fold with high probability $Pr(R|S)$ into the
-given target structure $R$; and \Def{positive design}, as generating one or
-many sequences that have a low energy for $R$.
+Figure~\ref{fig:seed-generation} shows relative numbers of finished instances and admissible sequences at each iteration of the seed generation procedure for the benchmark set \texttt{3str}.
+We obtain qualitatively equivalent results for all other benchmark sets (not shown).
+The top plot illustrates the  precomputation step of our strategy, where weights are calibrated for \GCb\% of 0.6.
+In this step, we produce sequence of the targeted \GCb\% to determine the linear transformation from the sampling energy model (stacking model) to the Turner model.
+The bottom plot shows the actual generation of the seed sequences after precomputation.
+The plot shows that most of the instances are finished after the first iterations, with only a few harder instances that require 2--3 times more iterations.
+Recall that the seed sequences were designed to have similar energy for all target structures.
 
-For each of two small examples, a single-stem and a double-stem structure, we produce a uniform and a weighted Boltzmann
-sampling, each of 10000 sequenes. For each sampled sequence $S$, we determine
-the Turner energy $E(S,R)$ as well as the
-%difference $E_\Delta(S,R)$ of $E(S,R)$ to the ensemble energy $E_{\text{ens}}(S)$ of $S$ 
-probability $Pr(R|S)$ of the structure $R$ in the ensemble of sequence $S$ 
-(where the energies are calculated in the Turner model using RNAfold).
-
-Note that low energy $E(S|R)$ is (non-linearly) correlated with high
-probability $Pr(S|R)$, the probability to see $S$ in the Boltzmann distribution
-of sequences folding into $R$. Recall that the probability $Pr(R|S)$ can be
-calculated as $\exp(-E(S,R)/RT)/Z(S)$,
-where $Z(S)=\sum_{R'} \exp(-E(S,R')/RT)$.
-%
-Consequently, Figure~\ref{fig:negforpos} ultimately illustrates the dependencies
-between $Pr(S|R)$ and $Pr(R|S)$ for uniform samples and low energy samples
-(where those samples can be efficiently generated due to our Boltzmann sampling
-strategy). For our examples, we observe (Fig.~\ref{fig:negforpos}) that Boltzmann sampling with weight 500 (green), generates good negative designs 
-with substantially higher probability than uniform sampling (blue).
-
-% and high energy difference $E_\Delta(S,R)=E_{\text{ens}}(S)-E(S,R)$, with high probability $\Pr(R|S)$.
-% 
-% To see the latter, recall that the \Def{ensemble energy $E_{\text{Ens}}(S)$} is defined as
-% $$
-% E_{\text{ens}}(S) = -RT \ln( Z(S) ),
-% $$ 
-% based on the partition function $Z(S)=\sum_{R'} \exp(-E(S,R')/RT)$
-% such that 
-% $$
-% E(S,R) - E_{\text{ens}}(S) 
-% =
-% -RT \ln Pr(R|S),
-% $$  
-% which can be derived by term transformation 
-% \begin{align*}
-% -RT \ln P(R|S) &= -RT \ln [ exp(-E(S,R)/RT) / Z(S) ] \\&= -RT \ln (exp(-E(S,R)/RT)) - (-RT\ln Z(S))
-% \\&=  E(S,R) - E_{\text{Ens}}(S). 
-% \end{align*}
-
-\begin{figure}
-\textbf{A}~\includegraphicstop[width=0.48\textwidth]{Figs/Plots/stem_weight1-500_mean}%
-\textbf{B}~\includegraphicstop[width=0.48\textwidth]{Figs/Plots/doublestem_weight1-500_mean}%
-\caption{Comparison of uniform sampling and Boltzmann sampling with
-weight $\pi=500$. We show results for two different small structures: (A) single stem structure (B) double stem structure. For each respective structure $R$, we plot the energies $E(S,R)$ of
-10000 sampled sequences against the probabilities $Pr(R|S)$ (in log scale).
-For illustrating the trend, we show moving averages over windows of $\pm 1$ kcal.
-Note that while only the energy $E(S,R)$ is directly controlled by Boltzmann sampling, this strongly impacts the common negative design objective $Pr(R|S)$.}
-\label{fig:negforpos}
+\begin{figure}[tbh]
+  \centering
+  \includegraphics[width=0.7\textwidth]{Figs/SeedGeneration/iterations_gc_3str.pdf}\\
+  \includegraphics[width=0.7\textwidth]{Figs/SeedGeneration/iterations_en_3str.pdf}
+  \caption{Relative numbers of finished instances and admissible sequences at each iteration of the seed generation procedure for the benchmark set \texttt{3str}.}
+  \label{fig:seed-generation}
 \end{figure}
@@
 \end{table}
-
-\section{Contribution of terms to the MultiDefect score}
-
-\newcommand{\EnsE}{\text{\rm G}} %ensemble energy
-
-Figure~\ref{fig:contribtions-multidefect} breaks down the multi-defect
-score of Eq.~$(\ref{eq:blueprintobjective})$, which is used to
-evaluate the quality of multi-target designs. Recall that the total
-score is composed of the term $\frac{1}{k} \sum_{\ell=1}^{k} (E(S,
-R_\ell) - \EnsE(S))$, which evaluates the distances of the energies of
-target structures to the minimum free energy (Distance to MFE), and
-the term $\frac{1}{2\binom{k}{2}} \sum\limits_{1\leq\ell<j\leq
-  k}|E(S,R_\ell) - E(S,R_j)|$, which evaluates the similarity of the
-target structure energies of the target structures (Heterogeneity).
-
-\label{appsec:multidefect}
-\begin{figure}[h!]
-\includegraphics[width=.48\textwidth]{Figs/statistics-Term1}
-\includegraphics[width=.48\textwidth]{Figs/statistics-Term2}
-\caption{Break-down of average MultiDefect values into their two components.}
-\label{fig:contribtions-multidefect}
-\end{figure}
-
-\section{Approximate counting and random generation}
-In fact, not only is $\#{\sf BIS}$ a reference problem in counting complexity, but it is also a landmark problem with respect to the complexity of approximate counting problems.
-In this context, it is the representative for a class of $\#{\sf BIS}$-hard problems~\citep{Bulatov2013} that are easier to approximate than $\# {\sf SAT}$, yet are widely believed not to admit any Fully Polynomial-time Randomized Approximation Scheme.
-Recent results reveal a surprising dichotomy in the behavior of $\#{\sf BIS}$: it admits a Fully Polynomial-Time Approximation Scheme (FPTAS) for graphs of max degree $\le 5$~\citep{Weitz2006}, but is as hard to approximate as the general $\#{\sf BIS}$ problem on graphs of degree $\ge 6$~\citep{Cai2016}.
-In other words, there is a clear threshold, in term of the max degree, separating (relatively) easy instances from really hard ones.
-
-Additionally, let us note that, from the classic Vizing Theorem, any bipartite graph $G$ having maximum degree $\Delta$ can be decomposed in polynomial time in exactly $\Delta$ matchings.
-Any such matching can be reinterpreted as a secondary structure, possibly with crossing interactions (\Def{pseudoknots}).
-These results have two immediate consequences for the pseudoknotted version of the multiple design counting problem.
-\begin{corollary}[as follows from~\citep{Weitz2006}]The number of designs compatible with $m\le 5$ pseudoknotted RNA structures can be approximated within any fixed ratio by a deterministic polynomial-time algorithm.
-\end{corollary}
-\begin{corollary}[as follows from~\citep{Cai2016}]
-  As soon as the number of pseudoknotted RNA structures strictly exceeds $5$, \NumDesign is as hard to approximate as {\#{\sf BIS}}.
-\end{corollary}
-
-It is worth noting that the $\#{\sf P}$ hardness of \NumDesign does not immediately imply the hardness of generating a valid design uniformly at random, as demonstrated constructively by Jerrum, Valiant and Vazirani~\citep{Jerrum1986}.
-However, in the same work, the authors establish a strong connection between the complexity of approximate counting and the uniform random generation.
-Namely, they showed that, for problems associated with self-reducible relations, approximate counting is equally hard as (almost) uniform random generation.
-We conjecture that the (almost) uniform sampling of sequences from multiple structures with pseudoknots is in fact \#{\sf BIS}-hard as soon as the number of input structures strictly exceeds $5$, as indicated by~\citet{Goldberg2004}, motivating even further our parameterized approach.
-
-\section{Tree decomposition for RNA design instances in practice}
-\label{appsec:treedecomp}
-
-For studying the typically expected treewidths and tree decomposition
-run times in multi-target design instances, we consider five sets of
-multi-target RNA design instances of different complexity. Our first
-set consists of the Modena benchmark instances.
-
-In addition, we generated four sets of instances of increasing
-complexity. The instances of the sets RF3, RF4, RF5, and RF6, each
-respectively specify 3,4,5, and 6 target structures for sequence
-length 100.  For each instance (100 instances per set), we generated a
-set of $k$ ($k=3,\dots,6$) compatible structures as follows
-\begin{itemize}
-\item Generate a random sequence of length 100;
-\item Compute its minimum free energy structure (ViennaRNA package);
-\item Add the new structure to the instances if the resulting base pair dependency graph is bipartite;
-\item Repeat until $k$ structures are collected.
-\end{itemize}
-For each instance, we generated the dependency graphs in the base pair
-model and in the stacking model. Then, we performed tree decomposition
-(using strategy ``GreedyFillIn'' of LibTW~\citep{Dijk2006}) on each dependency
-graph. The obtained treewidths are reported in
-Fig.~\ref{fig:td-widths}, while Fig.~\ref{fig:td-times} shows the
-corresponding run-times of the tree decomposition.
-
-\revised{
-  Finally, for further illustration, we show concrete tree decompositions for a selected instance in the benchmark set RF3 in Figure~\ref{fig:td-example}.
-This figure is provided solely for illustrational purposes:
-  First, it depicts the tree decomposition for a typical, realistic sampling task. Second, it exemplarily illustrates the typical impact of the energy model on the tree decomposition.
-  Compared to the tree decomposition based on the stacking model, the tree decomposition for the base pair model is much simpler.
-This shows in many connected components, which correspond to independently solvable sub-problems, and fewer positions per bag (i.e.
-smaller tree width), which allows far more efficient evaluation per tree node.
-  Note that the complexity of the tree decomposition ultimately shows the complexity of the dependencies that need to be considered for simultaneous energy evaluation for all target structures.
-This complexity is substantially increased in the stacking model over the simpler base pair model.
-}
-
-\begin{figure}
-  %\hspace*{2cm}\textbf{Base pair model}\hspace{4.75cm}\textbf{Stacking model}\\[-28pt]
-  \centering\includegraphics[width=0.9\textwidth]{Figs/td-widths}
-  \caption{Treewidths for multi-target RNA design instances of
-    different complexity. Distributions of treewidths shown as boxplots for the base pair (left) and stacking model (right).}
-  \label{fig:td-widths}
-\end{figure}
-
-\begin{figure}
-  %\hspace*{2cm}\textbf{Base pair model}\hspace{4.75cm}\textbf{Stacking model}\\[-28pt]
-  \centering
-  \includegraphics[width=0.9\textwidth]{Figs/td-times}
-  \caption{Computation time of tree decompositions for
-    multi-target RNA design instances of different complexity.
-    Distributions of times (in ms/instance) shown as boxplots for the base pair (left) and stacking model (right).}
-  \label{fig:td-times}
-\end{figure}
-
-\begin{figure}[t!]
-  %% the figure shows benchmark_inputs/RNAfold/3str/f3.100.0.inp
-  \centering
-  \verbatiminput{Figs/f3_100_0.txt}
-  \includegraphicstop[width=0.6\textwidth]{Figs/td-example-basepair}%
-  \includegraphicstop[width=0.35\textwidth]{Figs/td-example-stacking}
-  \caption{Example instance from RF3 (top) with its tree decompositions in the base pair (left) and stacking model (right). The respective treewidths are 2 and 12.}
-  \label{fig:td-example}
-\end{figure}
 
@@
 
+\section{Approximate counting and random generation}
+In fact, not only is $\#{\sf BIS}$ a reference problem in counting complexity, but it is also a landmark problem with respect to the complexity of approximate counting problems.
+In this context, it is the representative for a class of $\#{\sf BIS}$-hard problems~\citep{Bulatov2013} that are easier to approximate than $\# {\sf SAT}$, yet are widely believed not to admit any Fully Polynomial-time Randomized Approximation Scheme.
+Recent results reveal a surprising dichotomy in the behavior of $\#{\sf BIS}$: it admits a Fully Polynomial-Time Approximation Scheme (FPTAS) for graphs of max degree $\le 5$~\citep{Weitz2006}, but is as hard to approximate as the general $\#{\sf BIS}$ problem on graphs of degree $\ge 6$~\citep{Cai2016}.
+In other words, there is a clear threshold, in term of the max degree, separating (relatively) easy instances from really hard ones.
+
+Additionally, let us note that, from the classic Vizing Theorem, any bipartite graph $G$ having maximum degree $\Delta$ can be decomposed in polynomial time in exactly $\Delta$ matchings.
+Any such matching can be reinterpreted as a secondary structure, possibly with crossing interactions (\Def{pseudoknots}).
+These results have two immediate consequences for the pseudoknotted version of the multiple design counting problem.
+\begin{corollary}[as follows from~\citep{Weitz2006}]The number of designs compatible with $m\le 5$ pseudoknotted RNA structures can be approximated within any fixed ratio by a deterministic polynomial-time algorithm.
+\end{corollary}
+\begin{corollary}[as follows from~\citep{Cai2016}]
+  As soon as the number of pseudoknotted RNA structures strictly exceeds $5$, \NumDesign is as hard to approximate as {\#{\sf BIS}}.
+\end{corollary}
+
+It is worth noting that the $\#{\sf P}$ hardness of \NumDesign does not immediately imply the hardness of generating a valid design uniformly at random, as demonstrated constructively by Jerrum, Valiant and Vazirani~\citep{Jerrum1986}.
+However, in the same work, the authors establish a strong connection between the complexity of approximate counting and the uniform random generation.
+Namely, they showed that, for problems associated with self-reducible relations, approximate counting is equally hard as (almost) uniform random generation.
+We conjecture that the (almost) uniform sampling of sequences from multiple structures with pseudoknots is in fact \#{\sf BIS}-hard as soon as the number of input structures strictly exceeds $5$, as indicated by~\citet{Goldberg2004}, motivating even further our parameterized approach.
+
+\section{Contribution of terms to the MultiDefect score}
+
+\newcommand{\EnsE}{\text{\rm G}} %ensemble energy
+
+Figure~\ref{fig:contribtions-multidefect} breaks down the multi-defect
+score of Eq.~$(\ref{eq:blueprintobjective})$, which is used to
+evaluate the quality of multi-target designs. Recall that the total
+score is composed of the term $\frac{1}{k} \sum_{\ell=1}^{k} (E(S,
+R_\ell) - \EnsE(S))$, which evaluates the distances of the energies of
+target structures to the minimum free energy (Distance to MFE), and
+the term $\frac{1}{2\binom{k}{2}} \sum\limits_{1\leq\ell<j\leq
+  k}|E(S,R_\ell) - E(S,R_j)|$, which evaluates the similarity of the
+target structure energies of the target structures (Heterogeneity).
+
+\label{appsec:multidefect}
+\begin{figure}[h!]
+\includegraphics[width=.48\textwidth]{Figs/statistics-Term1}
+\includegraphics[width=.48\textwidth]{Figs/statistics-Term2}
+\caption{Break-down of average MultiDefect values into their two components.}
+\label{fig:contribtions-multidefect}
+\end{figure}
+
 \section{Detailed results of MultiDefect analysis}\label{sec:validity}

### 2019-01-23 supplement.tex
mark new section

@@
 
+\begin{revision}
 \section{Effectivity of the seed sequence generation}
-\todo[inline]{TODO}
-
 Figure~\ref{fig:seed-generation} shows relative numbers of finished instances and admissible sequences at each iteration of the seed generation procedure for the benchmark set \texttt{3str}.
@@
 
-\begin{figure}[tbh]
+\begin{figure}[h!]
   \centering
@@
 \end{figure}
+\end{revision}
 

### 2019-01-23 supplement.tex

@@
 
+\begin{figure}[h!]
+  %\hspace*{2cm}\textbf{Base pair model}\hspace{4.75cm}\textbf{Stacking model}\\[-28pt]
+  \centering\includegraphics[width=0.9\textwidth]{Figs/td-widths}
+  \caption{Treewidths for multi-target RNA design instances of
+    different complexity. Distributions of treewidths shown as boxplots for the base pair (left) and stacking model (right).}
+  \label{fig:td-widths}
+\end{figure}
+
+\begin{figure}[h!]
+  %\hspace*{2cm}\textbf{Base pair model}\hspace{4.75cm}\textbf{Stacking model}\\[-28pt]
+  \centering
+  \includegraphics[width=0.9\textwidth]{Figs/td-times}
+  \caption{Computation time of tree decompositions for
+    multi-target RNA design instances of different complexity.
+    Distributions of times (in ms/instance) shown as boxplots for the base pair (left) and stacking model (right).}
+  \label{fig:td-times}
+\end{figure}
+
 \revised{
@@
 
-\begin{figure}
-  %\hspace*{2cm}\textbf{Base pair model}\hspace{4.75cm}\textbf{Stacking model}\\[-28pt]
-  \centering\includegraphics[width=0.9\textwidth]{Figs/td-widths}
-  \caption{Treewidths for multi-target RNA design instances of
-    different complexity. Distributions of treewidths shown as boxplots for the base pair (left) and stacking model (right).}
-  \label{fig:td-widths}
-\end{figure}
-
-\begin{figure}
-  %\hspace*{2cm}\textbf{Base pair model}\hspace{4.75cm}\textbf{Stacking model}\\[-28pt]
-  \centering
-  \includegraphics[width=0.9\textwidth]{Figs/td-times}
-  \caption{Computation time of tree decompositions for
-    multi-target RNA design instances of different complexity.
-    Distributions of times (in ms/instance) shown as boxplots for the base pair (left) and stacking model (right).}
-  \label{fig:td-times}
-\end{figure}
-
-\begin{figure}[t!]
+\begin{figure}[h!]
   %% the figure shows benchmark_inputs/RNAfold/3str/f3.100.0.inp
@@
 
-To illustrate the efficiency and practical feasibility of our Boltzmann sampling approach, we show the distributions of run times (Intel(R) Core(TM) i7-4770 CPU) for generating 20.000 samples (over the Modena benchmark instances and grouped by benchmark sub-set).
+To illustrate the efficiency and practical feasibility of our Boltzmann sampling approach, we show the distributions of run times (Intel(R) Core(TM) i7-4770 CPU) for generating 10000 samples (over the Modena benchmark instances and grouped by benchmark sub-set).
 Since producing a controlled Boltzmann sampling requires significant pre-computation, it is informative to separately show construction and sampling times (Figure~\ref{fig:run-times-redprint}).
@@
 The times are split into construction time (left) and sampling time (right).
-Construction includes all the pre-computation, while the latter is the time needed to generate 20.000 sequences.
+Construction includes all the pre-computation, while the latter is the time needed to generate 10000 sequences.
 }
@@
 \end{figure}
-\begin{figure}[tbh]
+\begin{figure}[h!]
   \centering
@@
 The times are split into construction time (left) and sampling time (right).
-Construction includes all the pre-computation, while the latter is the time needed to generate 20.000 sequences.}
+Construction includes all the pre-computation, while the latter is the time needed to generate 10000 sequences.}
   \label{fig:run-times-rbp}
@@
 
-\begin{figure}[tb]
+\begin{figure}[h!]
       \begin{center}
@@
 
-\begin{figure}
+\begin{figure}[h!]
   \begin{center}
@@
 
-\begin{figure}
+\begin{figure}[h!]
   \centering

### 2019-01-23 RNARedPrint.tex
Some changes

@@
 	The value of $F^{\tt BP}$ is $0$ for any valid sequence, and $+\infty$ as soon as some non canonical base pair is created.
-For any associated weight $\pi_{{\sf BP}}>1$, the contribution of a valid sequence is $\pi_{{\sf BP}}^0 = 1$, and the contribution of an invalid seqeunce is $\pi_{{\sf BP}}^{+\infty} = 0$, so that Equation~\ref{eq:mainproblem} simply counts the number of valid sequences when restricted to to $F^{\tt BP}$.
-
-	Energy models for structure prediction vary considerably, yet can always be expressed as a sum over contributions associated with local structural motifs (base pairs, stacking pairs, loops\ldots) under a certain nucleotide assignment.
-Energy models can thus be captured generically by introducing, to each motif $m$ occurring in a structure, a contribution $f_m$, taking a specific value for each assignment of nucleotides to its specific set of positions $\dep(f_m)$.
-For instance, the contribution of a stacking pair, consisting of two pairs $(i,j)$ and $(i+1,j-1)$, will be expressed through a
+For any associated weight $\pi_{{\sf BP}}>1$, the contribution of a valid sequence is $\pi_{{\sf BP}}^0 = 1$, and the contribution of an invalid sequence is $\pi_{{\sf BP}}^{+\infty} = 0$, so that Equation~\ref{eq:mainproblem} simply counts the number of valid sequences when restricted to $F^{\tt BP}$.
+
+	Energy models for structure prediction vary considerably, yet can always be expressed as sums over contributions associated with local structural motifs (base pairs, stacking pairs, loops\ldots) under a certain nucleotide assignment.
+Energy models can thus be captured generically by introducing, for each motif $m$ occurring in a target structure, a contribution $f_m$, taking a specific value for each assignment of nucleotides to its positions $\dep(f_m)$.
+For instance, the contribution of a stacking pair, consisting of two pairs $(i,j)$ and $(i+1,j-1)$, can be captured by the introduction of a function $f^{\textsf{Stack}}_{\substack{i,j}}$ such that $\dep(f^{\textsf{Stack}}_{\substack{i,j}}) = \{i,i+1,j-1,j\}$.
 
@@
 
-In order to capture the requirement of canonical base pairing for each structure,
+\revised{In order to compute the partition function of Eq.~\ref{eq:mainproblem}, and thus sample in a well-defined definition, one needs an abstraction to express dependencies beyond pairs of positions.
+Namely, we must consider dependencies induced by the complete set of contributions $$\F := \bigcup_{\ell}\{f\mid f\text{ contribution of } F_\ell\}.$$
+%
+%Let us denote the \emph{dependencies of a contribution $f$} as \dep{f}.
+%
+The set $\F$ of contributions induces a
+\Def{dependency graph} $G_\F$ on sequence positions by interpreting the dependencies as hyperedges, leading to the hypergraph 
+$$G_\F=(\{1,\dots,n\} ,\{\dep(f)\mid f\in \F\}).$$ 
+Let us now remind the \Def{tree decomposition} of a graph $G_\F$, a fundamental ingredient of our algorithms.}
+
+Moreover, to capture the requirement of canonical base pairing for each structure,
 we define the \Def{base pair dependency graph} $G_{\R}$ as the graph with nodes $\{1,\dots,n\}$
@@
 Any edge between two nodes $i$ and $j$ materializes the existence of a base pair between positions $i$ and $j$ in some structure, inducing a direct dependency between the chosen nucleotide assignments for the two positions.
-
-More generally, we consider dependencies induced by the complete set of contributions $$\F := \bigcup_{\ell}\{f\mid f\text{ contribution of } F_\ell\}.$$
-%
-%Let us denote the \emph{dependencies of a contribution $f$} as \dep{f}.
-%
-The set $\F$ of contributions induces a
-  \Def{dependency graph} $G_\F$ on sequence positions by interpreting the dependencies as hyperedges, leading to the hypergraph 
-  $$G_\F=(\{1,\dots,n\} ,\{\dep(f)\mid f\in \F\}).$$ 
-  Let us now remind the \Def{tree decomposition} of a graph $G_\F$, a fundamental ingredient of our algorithms.
 
@@
 %However, $\#{\sf BIS}$ is also \#{\sf P}-hard on connected graphs, as the number of independent sets for a disconnected graph $G$ can be efficiently computed by $|\IS{G}|=\prod_{cc\in CC(G)} |\IS{cc}|$.
-In other words, any efficient algorithm for \NumDesign implies an efficient algorithm for $\#{\sf BIS}$, thus our conclusion on the complexity of \NumDesign.
-
-\begin{corollary}
-  \NumDesign is $\#{\sf P}$-hard
-\end{corollary}
+In other words, any efficient algorithm for \NumDesign implies an efficient algorithm for $\#{\sf BIS}$, thus our conclusion that \NumDesign is $\#{\sf P}$-hard.
+
 % \begin{proof}
@@
 
-{\centering \url{https://github.com/yannponty/RNARedPrint}\\[.3em]}
+{\centering \url{https://github.com/yannponty/RNARedPrint}\\[.3em]}%
 

### 2019-01-23 RNARedPrint.tex

--- 2019-01-23 RNARedPrint.tex
@@
 \section*{Methods}
-The main computational problem addressed in this work is the design of sequences over the alphabet $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}$, featuring a given $\GCb\%$, and compatible with a set of secondary structures, associated with prescribed free-energies.
-Such criteria are here abstracted as constraints on the value of \Def{features}, functions of the sequence that are expressed as sums over real-valued contributions.
+The main computational problem addressed in this work is the generation of sequences over the alphabet $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}$, \revised{such that the sequences feature a given $\GCb\%$, and have prescribed energies for a set of target secondary structures.
+Here, such criteria are modeled} as constraints on the value of \Def{features}, functions of the sequence that are expressed as sums over real-valued contributions.
 Such contributions typically depend on the nucleotides assigned to a subset of sequence positions.

### 2019-01-23 RNARedPrint.tex
Meeting Seb

@@
 
-\subsection*{Effectively targeting Turner energies}
+\subsection*{A parsimonious stacking energy model as a proxy for the Turner energy model}
+Sampling for fixed weight
+Time benchmark
+
+Figure 10 - Correlation plot -- Right
+Figure 2 -- Distribution plot -- Right (Turner is even untractable to compute the tree width itself in benchmark instances)
+Figure 3 -- Right -- Distribution
+
+\subsection*{Effectively targeting Turner energies using multidimensionnal sampling}
+\revised{
+
 We used our Boltzmann sampling strategy (Algorithms
@@
 result from sampling with different weight parameters. The figure
-illustrates how our multidimensionnal Boltzmann sampling strategy can,
+illustrates how our multidimensional Boltzmann sampling strategy can,
 to a large extent, independently shift the Turner energies of sampled
@@
 Fig.~\ref{fig:energydist} for a further example with three
-pseudoknot-free target structures.
+pseudoknot-free target structures.}
 

### 2019-01-23 RNARedPrint.tex
Updates from ShareLaTeX
Hypergraph introduction

@@
 
-\subsection*{Dependency graph, tree decomposition and width}
+\subsection*{Dependency (hyper)graph, tree decomposition and width}
 %The complexity of the central problem depends fundamentally on the dependency structure that is imposed by the entire set of contributions.
@@
 
-\revised{In order to compute the partition function of Eq.~\ref{eq:mainproblem}, and thus sample in a well-defined definition, one needs an abstraction to express dependencies beyond pairs of positions.
-Namely, we must consider dependencies induced by the complete set of contributions $$\F := \bigcup_{\ell}\{f\mid f\text{ contribution of } F_\ell\}.$$
+\revised{In order to compute the partition function of Eq.~\ref{eq:mainproblem}, and thus sample in a well-defined distribution, one needs to abstract dependencies beyond pairs of positions.
+To that purpose, we consider the notion of \Def{hypergraph}, which consists of vertices (here, the sequence positions) connected by \Def{hyperedges} defined as sets of vertices.
+Undirected graphs can be modeled as hypergraphs whose hyperedges are pairs of vertices, but hypergraphs also allow to express complex dependencies.
+
+Here, we consider the complete set of contributions $$\F := \bigcup_{\ell}\{f\mid f\text{ contribution of } F_\ell\}.$$
 %
@@
 The set $\F$ of contributions induces a
-\Def{dependency graph} $G_\F$ on sequence positions by interpreting the dependencies as hyperedges, leading to the hypergraph 
-$$G_\F=(\{1,\dots,n\} ,\{\dep(f)\mid f\in \F\}).$$ 
-Let us now remind the \Def{tree decomposition} of a graph $G_\F$, a fundamental ingredient of our algorithms.}
-
-Moreover, to capture the requirement of canonical base pairing for each structure,
-we define the \Def{base pair dependency graph} $G_{\R}$ as the graph with nodes $\{1,\dots,n\}$
-and edges $\bigcup_{\ell\in[1,k]} R_\ell$.
-Any edge between two nodes $i$ and $j$ materializes the existence of a base pair between positions $i$ and $j$ in some structure, inducing a direct dependency between the chosen nucleotide assignments for the two positions.
+\Def{dependency (hyper)graph} $G_\F$ on sequence positions by interpreting the dependencies as hyperedges, leading to an hypergraph 
+$G_\F=(V, H)$ having vertices $V=\{1,\dots,n\}$ and hyperedges $H = \{\dep(f)\mid f\in \F\})$.
+
+Irrespectively of the chosen energy model, to capture the requirements of canonical base pairing for each structure,
+we also define the \Def{base pair dependency graph} $G_{\R} = (V,E)$ having vertices $V=\{1,\dots,n\}$ 
+and edges $E=\bigcup_{\ell\in[1,k]} R_\ell$.
+Any edge between two nodes $i$ and $j$ materializes the existence of a base pair between positions $i$ and $j$ in at least one of the input structures, and induce a direct dependency between the nucleotide assignments for those positions.
+}
+
+Let us now remind the \Def{tree decomposition} of an (hyper)graph $G_\F$, a fundamental ingredient of our algorithms.
 
@@
 \begin{definition}[Tree width]
-The \Def{width} of a tree decomposition $(T,\chi)$ is defined as
+The \Def{width} of a tree decomposition $(T,\chi)$ is 
 $$\min_{u\in T} |\chi(u)| - 1.$$ The \Def{treewidth}
-of a graph $G$ is the smallest width of any tree decomposition of $G$.
+of an (hyper)graph $G$ is the smallest width of a tree decomposition of $G$.
 \end{definition}
@@
 \revised{While efficient, both in practice and in theory for graphs of bounded tree width, our algorithms remain exponential in the worst case scenario, since the tree-width of a dependency graph can then become arbitrarily large.
-This exponential complexity in the worst case seems intrinsic.
+This exponential complexity in the worst case appears to be intrinsic.
 Indeed, we show that a specialization of our core problem, namely the enumeration of designs that respect canonical base pairing rules ($\Ab \leftrightarrow \Ub$, $\Gb \leftrightarrow \Cb$, $\Gb \leftrightarrow \Ub$) is $\#{\sf P}$-hard, even when the dependency graph is bipartite and connected.
-The existence of a polynomial time algorithm for computing the partition function of Equation \eqref{eq:mainproblem} is thus unlikely, and it would imply that  ${\sf\# P}={\sf FP}$ and, in turn, that ${\sf P}={\sf NP}$.}
+The existence of a polynomial time algorithm for computing the partition function of Equation \eqref{eq:mainproblem} is thus unlikely, as it would imply that  ${\sf\# P}={\sf FP}$ and, in turn, that ${\sf P}={\sf NP}$.}
 
@@
 In other words, any efficient algorithm for \NumDesign implies an efficient algorithm for $\#{\sf BIS}$, thus our conclusion that \NumDesign is $\#{\sf P}$-hard.
+
+\revised{Proposition~\ref{prop:bijection} also strongly impacts the complexity of computing the partition function.
+Indeed it implies that, among the $4^k$ possible assignments of nucleotides to $k$ connected positions (in the base-pair dependency graph), at most $2^k$ are compatible with base pairing rules.
+One can thus sharply reduce the complexity of Algorithm~\ref{alg:pf} by restricting the precomputations to compatible assignments.}
 

### 2019-01-23 RNARedPrint.tex
intermediary state of editing

--- 2019-01-23 RNARedPrint.tex
@@
 
-\revised{As our central contribution, we provide a systematic and flexibly extensible technique for sampling that targets multiple versatile features.
+\revised{As central contribution, we provide a systematic and flexibly extensible technique for sampling that targets multiple versatile features.
 For the sake of clarity, we introduce this method specialized to the sampling of sequences that target the energies of multiple structures and the \GCb\%.}
 
-\revised{Fig.~\ref{fig:workflow} summarizes our generic framework, which enables} efficient Boltzmann-weighted sampling over RNA sequences with multiple target structures.
-\revised{\emph{Algorithmically}, we contribute} dynamic programming (DP) algorithms, based on the concept of \Def{tree decomposition}, to compute partition functions and sample sequences from the Boltzmann distribution%
+\revised{Fig.~\ref{fig:workflow} summarizes our generic framework, which enables the generation of sequence samples with specific energies for multiple target structures (based on multi-dimensional Boltzmann sampling).
+\emph{Algorithmically}, we originally contribute} dynamic programming (DP) algorithms, based on the concept of \Def{tree decomposition}, to compute partition functions and sample sequences from the Boltzmann distribution%
 % (Subsection~\ref{sec:PF})
@@
 %
-\revised{Building on this principle, }the complexities of our algorithms depend exponentially on a specific property of the tree decomposition, called the \Def{treewidth}, which remains low for typical instances.
-However, for any given value of the treewidth, the complexity scales only linearly with the size of designed sequences and the number of targeted structures, \emph{i.e.} our algorithms are fixed-parameter tractable (FPT).
-
-\revised{Remarkably, we show that it is not possible to find a better, efficient method for sampling (unless ${\sf P}={\sf NP}$), since the underlying counting problem is \#{\sf P}-hard.
-The practical relevance of this result is that it rules out substantially better sampling techniques, such that there will always remain---in practice---an upper limit on the tractable number of structures and their heterogeneity, the complexity of the directly treatable energy model, and the number and complexity of additional constraints that could be considered in future sampling-based applications.} Technically, this result relies on a surprising bijection between valid sequences and independent sets of a bipartite graph, the latter being the object of recent breakthroughs in approximate counting complexity~\citep{Bulatov2013,Cai2016}.
+\revised{Building on this principle, the complexities of our algorithms depend exponentially on a specific property of the tree decomposition, called the \Def{treewidth}.
+Thus, it is essential for the applicability of our approach that---by appropriate design choices---we can keep this parameter low for typical instances.
+For} any fixed value of the treewidth, the complexity scales only linearly with the size of designed sequences and the number of targeted structures, \emph{i.e.} our algorithms are \emph{fixed-parameter tractable (FPT)}.
+
+\revised{Moreover, we could show that it is not possible to find a better, efficient method for sampling (unless ${\sf P}={\sf NP}$), since the underlying counting problem is \#{\sf P}-hard.
+The practical relevance of this theoretical result is that it rules out substantially better sampling techniques, such that there will always---even using improved sampling methods----remain an upper limit on the---in practice---tractable number and heterogeneity of structures, the complexity of the directly treatable energy model, and the number and complexity of additional constraints that could be considered in future sampling-based applications.} Technically, this result relies on a surprising bijection between valid sequences and independent sets of a bipartite graph, the latter being the object of recent breakthroughs in approximate counting complexity~\citep{Bulatov2013,Cai2016}.
 
@@
 %Uniform sampling is handled as a special case of Boltzmann sampling, where each valid sequence receives energy zero and---consequently---computing partition functions specializes to counting.
-By evaluating (partial) sequences in a specialized weighted constraint
-network, we support arbitrary multi-ary constraints and thus
+By evaluating sequences in a specialized weighted constraint
+network, we support---in principle---arbitrary multi-ary constraints and thus
 arbitrarily complex energy models,
@@
 Turner RNA energy model
-works well by performing sampling based on \revised{a simplified RNA energy model, which induces lower treewidth}.
-This result is highly relevant in practice, since it allows to combine high efficiency with high accuracy (and even leaves room for \revised{future extensions supporting additional constraints}).
+works well by performing sampling based on \revised{a simplified RNA energy model, which induces much lower treewidth than the Turner model}.
+This result is \revised{essential for the applicability of our method, since it allows to combine high efficiency (by keeping the tree width low) with sufficient accuracy to precisely target realistic Turner energies.}
 
@@
 \section*{Methods}
-The main computational problem addressed in this work is the generation of sequences over the alphabet $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}$, \revised{such that the sequences feature a given $\GCb\%$, and have prescribed energies for a set of target secondary structures.
-Here, such criteria are modeled} as constraints on the value of \Def{features}, functions of the sequence that are expressed as sums over real-valued contributions.
-Such contributions typically depend on the nucleotides assigned to a subset of sequence positions.
-
-To achieve maximal diversity within candidate designs, we abstract the problem as a random generation of sequences within a Boltzmann distribution.
-The probability of a sequence then depends on its associated features, and of weights that are associated features.
-Those influence the expected feature values of generated sequences, allowing us to shift the sample distributions towards specific targets.
-Sampling from this distribution requires to compute corresponding partition functions, such that we can draw sequences with probabilities proportional to their Boltzmann weight.
-A final rejection step, made efficient by the fine calibration of weights, allows to produce sequences satisfying the input set of constraints.
+\revised{The main computational problem addressed in this work is the positive design of RNA sequences for multiple target structures; more specifically, the generation of sequences over the alphabet $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}$, such that the sequences feature a given $\GCb\%$, and have prescribed energies for a set of target secondary structures.
+Here, such criteria are modeled} as constraints on the values of \Def{features}, functions of the sequence that are expressed as sums over real-valued contributions.
+Such contributions typically depend on the nucleotides at specific sequence positions.
+
+\revised{To generate diverse design candidates, we randomly generate sequences from a Boltzmann distribution.
+The probability of a sequence then depends on its features (e.g.
+the energies of the target structures), and the weight of each feature (which influences its distribution).
+Sampling from the (multi-feature) Boltzmann distribution requires to compute corresponding partition functions, such that we can draw sequences with probabilities proportional to their Boltzmann weight.
+On this basis, we can finely calibrate the weights, to increase the probability that sampled sequences meet prescribed target values for each feature.
+Together with a final rejection step this results in an effective procedure for generating highly specific sequences.}
 
@@
 \label{sec:problem-statement}
+
 %\paragraph{Central computational problem.}
-Let us consider $m$ (secondary) structures $R_1,\dots, R_m$, each abstracted as a set of base pairs, and $k$ features $F_1,\dots, F_k$, associated with weights $\pi_1,\dots,\pi_k$, our goal is to sample from the Boltzmann distribution, defined over the set of sequences that satisfy classic base pairing rules for each structure as:
+Let us consider $m$ (secondary) structures $R_1,\dots, R_m$, each abstracted as a set of base pairs, and \revised{$k\geq m$} features $F_1,\dots, F_k$, associated with weights $\pi_1,\dots,\pi_k$.
+Our goal is to sample from the Boltzmann distribution, defined over the set of sequences that satisfy classic base pairing rules for each structure as:
 \begin{equation}
@@
 
-The workhorse of our approach is the efficient (more precisely, fixed-parameter tractable) computation of feature-dependent partition functions over sequences:
+The workhorse of our approach is the \revised{fixed-parameter tractable} computation of feature-dependent partition functions over sequences:
 %Given are the features $F_1,\dots,F_k$, which encode the energies functions for the fixed structures and the function for the \GCb\%, all in terms of simple atomic energy contributions.
@@
 \revised{
-	Such features can also be used to count the number of \Def{valid sequences}, \emph{i.e.} sequences inducing only base pairs in $\mathcal{B}:=\{\{\Ab,\Ub\}, \{\Gb,\Cb\}, \{\Gb,\Ub\}\}$.
-Indeed, consider a feature $F^{\tt BP}$ composed of  contributions $f^{\tt BP}_{i,j}$, for each base pair $(i,j)$ occurring in some structure, such that
+  To start with a simple example of evaluating the energy of sequences by features, let us explain how they are used to count the number of \Def{valid sequences}, \emph{i.e.} sequences inducing only base pairs in $\mathcal{B}:=\{\{\Ab,\Ub\}, \{\Gb,\Cb\}, \{\Gb,\Ub\}\}$.
+Consider a feature $F^{\tt BP}$ composed of  contributions $f^{\tt BP}_{i,j}$, for each base pair $(i,j)$ occurring in some structure, such that
 	$$f^{{\sf BP}}_{i,j}(\left\{\substack{i\mapsto a\\j\mapsto b}\right\})=\begin{cases}0&\text{ if }\{a,b\}\in\mathcal{B} \\+\infty&\text{ otherwise}.\end{cases}$$ 
         The value of $F^{\tt BP}$ is $0$ for any valid sequence, and $+\infty$ as soon as some non canonical base pair is created.
-For any associated weight $\pi_{{\sf BP}}>1$, the contribution of a valid sequence is $\pi_{{\sf BP}}^0 = 1$, and the contribution of an invalid sequence is $\pi_{{\sf BP}}^{+\infty} = 0$, so that Equation~\ref{eq:mainproblem} simply counts the number of valid sequences when restricted to $F^{\tt BP}$.
+For any associated weight $\pi_{{\sf BP}}>1$, the contribution of a valid sequence is $\pi_{{\sf BP}}^0 = 1$, and the contribution of an invalid sequence is $\pi_{{\sf BP}}^{+\infty} = 0$, so that Eq.~\ref{eq:mainproblem} (when restricted to $F^{\tt BP}) $simply counts the number of valid sequences.
 
@@
 }
+
+\todo[inline]{Check use of the term stacking model; the term is not introduced here yet!}
 
@@
 
-\subsection*{Dependency graph, tree decomposition and width}
+\subsection*{Dependency graph, tree decomposition and treewidth}
 %The complexity of the central problem depends fundamentally on the dependency structure that is imposed by the entire set of contributions.
@@
 
-\revised{In order to compute the partition function of Eq.~\ref{eq:mainproblem}, and thus sample in a well-defined definition, one needs an abstraction to express dependencies beyond pairs of positions.
-Namely, we must consider dependencies induced by the complete set of contributions $$\F := \bigcup_{\ell}\{f\mid f\text{ contribution of } F_\ell\}.$$
+\revised{In order to compute the partition function of Eq.~\ref{eq:mainproblem}, and thus sample in a well-defined way, one must consider dependencies induced by the complete set of contributions $$\F := \bigcup_{\ell}\{f\mid f\text{ contribution of } F_\ell\}.$$
 %
@@
 %
+
+\revised{
+In the simplest case, this set captures the requirement of canonical base pairing for each structure.
+To express this, let us define the \Def{base pair dependency graph} $G_{\R}$ as the graph with nodes $\{1,\dots,n\}$
+and edges $\bigcup_{\ell\in[1,k]} R_\ell$.
+%Any edge between two nodes $i$ and $j$ materializes the existence of a base pair between positions $i$ and $j$ in some structure, inducing a direct dependency between the chosen nucleotide assignments for the two positions.
+}
+
 The set $\F$ of contributions induces a
@@
 $$G_\F=(\{1,\dots,n\} ,\{\dep(f)\mid f\in \F\}).$$ 
-Let us now remind the \Def{tree decomposition} of a graph $G_\F$, a fundamental ingredient of our algorithms.}
-
-Moreover, to capture the requirement of canonical base pairing for each structure,
-we define the \Def{base pair dependency graph} $G_{\R}$ as the graph with nodes $\{1,\dots,n\}$
-and edges $\bigcup_{\ell\in[1,k]} R_\ell$.
-Any edge between two nodes $i$ and $j$ materializes the existence of a base pair between positions $i$ and $j$ in some structure, inducing a direct dependency between the chosen nucleotide assignments for the two positions.
-
-\begin{definition}[Tree decomposition]
+Let us now define the \Def{tree decomposition} of the graph $G_\F$, a fundamental ingredient of our algorithms, which also determines their efficiency (most importantly, via its property called treewidth).
+}
+
+\begin{definition}[Tree decomposition and treewidth]
   \label{def:treedecomp}
@@
 \end{enumerate}
+
+\smallskip
+\noindent
+The \Def{treewidth} of a tree decomposition $(T,\chi)$ is defined as
+$$\min_{u\in T} |\chi(u)| - 1.$$ 
+The \Def{treewidth}
+of a graph $G$ is the smallest treewidth of any tree decomposition of $G$.
+\todo[inline]{Check: do we really need to define treewidth of a graph?}
 \end{definition}
+
 \revised{Intuitively, a tree decomposition for an (hyper)graph $G$ is a tree structure which features all the vertices of $G$, captures all the (hyper)edges of $G$, and helps ensure some notion of consistency by allowing the propagation of values.
 Figure~\ref{fig:treedec} shows an optimal tree decomposition for a pair of structures under the stacking energy model.}
-\begin{definition}[Tree width]
-The \Def{width} of a tree decomposition $(T,\chi)$ is defined as
-$$\min_{u\in T} |\chi(u)| - 1.$$ The \Def{treewidth}
-of a graph $G$ is the smallest width of any tree decomposition of $G$.
-\end{definition}
 %Recall the essential role of the treewidth as fixed parameter in our FPT appraoch.

### 2019-01-23 RNARedPrint.tex

@@
 
-\subsection*{Dependency graph, tree decomposition and treewidth}
+\subsection*{Dependency (hyper)graph, tree decomposition and treewidth}
+
 %The complexity of the central problem depends fundamentally on the dependency structure that is imposed by the entire set of contributions.
@@
 
-\revised{In order to compute the partition function of Eq.~\ref{eq:mainproblem}, and thus sample in a well-defined way, one must consider dependencies induced by the complete set of contributions $$\F := \bigcup_{\ell}\{f\mid f\text{ contribution of } F_\ell\}.$$
+\revised{In order to compute the partition function of Eq.~\ref{eq:mainproblem}, and thus sample in a well-defined way, one must consider dependencies induced by the complete set of contributions $$\F := \bigcup_{\ell}\{f\mid f\text{ contribution of } F_\ell\}.$$}
+
 %
@@
 %Any edge between two nodes $i$ and $j$ materializes the existence of a base pair between positions $i$ and $j$ in some structure, inducing a direct dependency between the chosen nucleotide assignments for the two positions.
-}
-
-The set $\F$ of contributions induces a
-\Def{dependency graph} $G_\F$ on sequence positions by interpreting the dependencies as hyperedges, leading to the hypergraph 
-$$G_\F=(\{1,\dots,n\} ,\{\dep(f)\mid f\in \F\}).$$ 
+
+Since $\F$ defines potentially more complex dependencies, which can
+relate more than two positions, its dependencies cannot be similarily represented by a
+graph. Instead, this requires a structure known as \Def{hypergraph}, which
+consists of vertices (here, the sequence positions) connected by
+\Def{hyperedges} are sets of vertices---in generalization of graphs, where edges are pairs of vertices.
+%Undirected graphs can be
+%modeled as hypergraphs whose hyperedges are pairs of vertices, but
+%hypergraphs also allow to express complex dependencies.
+
+The set $\F$ of contributions induces the \Def{dependency
+  (hyper)graph} $G_\F=(V, H)$ on sequence positions $V=\{1,\dots,n\}$
+by interpreting the dependencies as hyperedges, i.e. $H =
+\{\dep(f)\mid f\in \F\})$.
+
 Let us now define the \Def{tree decomposition} of the graph $G_\F$, a fundamental ingredient of our algorithms, which also determines their efficiency (most importantly, via its property called treewidth).
 }
-
 \begin{definition}[Tree decomposition and treewidth]
@@
 Figure~\ref{fig:treedec} shows an optimal tree decomposition for a pair of structures under the stacking energy model.}
-%Recall the essential role of the treewidth as fixed parameter in our FPT appraoch.
 

### 2019-01-23 RNARedPrint.tex
correction

@@
 Since $\F$ defines potentially more complex dependencies, which can
-relate more than two positions, its dependencies cannot be similarily represented by a
-graph. Instead, this requires a structure known as \Def{hypergraph}, which
-consists of vertices (here, the sequence positions) connected by
-\Def{hyperedges} are sets of vertices---in generalization of graphs, where edges are pairs of vertices.
+relate more than two positions, its dependencies cannot be similarily
+represented by a graph. Instead, this requires a structure known as
+\Def{hypergraph}, which consists of vertices (here, the sequence
+positions) connected by \Def{hyperedges}, which are sets of
+vertices. In this way, hypergraphs generalize undirected graphs, where
+edges are sets of two vertices.
 %Undirected graphs can be

### 2019-01-24 BMC-Response-Round1.tex
Meeting Seb
Minor edits; check treewidth, stacking energy model}

@@
   \begin{itemize}
-    \item The full-featured RNA energy model (Turner) can be well approximated by the much more efficiently tractable stacking model;
-    \item Controlled Boltzmann sampling for typical multi-target design instances requires little time (corresponding to small tree-width); Multi-dimensional Boltzmann sampling solves the positive design problem effectively, requiring only a limited number of rejections (even for multiple simultaneously considered features);
+    \item The full-featured RNA energy model (Turner) can be well approximated by the much more efficiently tractable stacking model; in this model, controlled Boltzmann sampling for typical multi-target design instances requires little time (corresponding to small tree-width);
+    \item Multi-dimensional Boltzmann sampling solves the positive design problem effectively, requiring only a limited number of rejections (even for multiple simultaneously considered features);
     \item The positively designed sequences are suitable for improving negative design approaches.
@@
 In the scope of this study, the running time and memory comparison should be shown to produce the same number of sequences by some minor variants of RNARedPrint method, RNABluePrint, and MODENA.\\[1em]}%
-\Answer{We agree that running times are important for assessing the method; consequently, we included details timing information in the revision.
-As we show now explicitly, the running times directly depend on the already reported tree-widths.
-The run-time requirements are generally modest for the relevant design instances.
-%
+\Answer{We agree that running times are important for assessing the method; consequently, we included detailed timing information in the revision.
+The running times directly depend on the already reported treewidths (which we actively keep low due to approximating the full energy model and using state-of-the-art decomposition methods).
+In consequence, the run-time requirements are generally modest for the relevant design instances; making our approach applicable in practice.
+
 Concerning the comparison: since MODENA does not allow to measure separate times for sequence sampling, we could only estimate its run-times by running the MODENA binary with parameters that turn off the optimization as far as this is possible.
@@
 positive design).
-This crucial difference should kept in mind, when comparing the run-times of these tools.
+This crucial difference should be kept in mind, when comparing the run-times of these tools.
 \AnswerTODO{...}}
+
 \Comment{

### 2019-01-24 RNARedPrint.tex
Minor edits; check treewidth, stacking energy model}

@@
 \revised{Note that for simplicity, we assume in this figure that only dependencies between the ends of base pairs are considered to evaluate the energies of structures.
-Our computations based on a more complex energy model (stacking model) require additional dependencies.}
+Our computations based on a more complex energy model, which considers energy contributions of base pair stacks, require additional dependencies.}
   }
@@
 works well by performing sampling based on \revised{a simplified RNA energy model, which induces much lower treewidth than the Turner model}.
-This result is \revised{essential for the applicability of our method, since it allows to combine high efficiency (by keeping the tree width low) with sufficient accuracy to precisely target realistic Turner energies.}
+This result is \revised{essential for the applicability of our method, since it allows to combine high efficiency (by keeping the treewidth low) with sufficient accuracy to precisely target realistic Turner energies.}
 
@@
 \revised{The main computational problem addressed in this work is the positive design of RNA sequences for multiple target structures; more specifically, the generation of sequences over the alphabet $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}$, such that the sequences feature a given $\GCb\%$, and have prescribed energies for a set of target secondary structures.
-Here, such criteria are modeled} as constraints on the values of \Def{features}, functions of the sequence that are expressed as sums over real-valued contributions.
-Such contributions typically depend on the nucleotides at specific sequence positions.
+Here, these desired sequence properties are modeled as constraints on the values of \Def{features}, which are functions of the sequence that are expressed as sums over real-valued \Def{contributions}.
+Each contribution depends on the nucleotides at---typically few---specific sequence positions.}
 
@@
 
-	Energy models for structure prediction vary considerably, yet can always be expressed as sums over contributions associated with local structural motifs (base pairs, stacking pairs, loops\ldots) under a certain nucleotide assignment.
+        Energy models for structure prediction vary considerably, yet can always be expressed as sums over contributions associated with local structural motifs (base pairs, base pair stacks, loops\ldots) under a certain nucleotide assignment.
 Energy models can thus be captured generically by introducing, for each motif $m$ occurring in a target structure, a contribution $f_m$, taking a specific value for each assignment of nucleotides to its positions $\dep(f_m)$.
-For instance, the contribution of a stacking pair, consisting of two pairs $(i,j)$ and $(i+1,j-1)$, can be captured by the introduction of a function $f^{\textsf{Stack}}_{\substack{i,j}}$ such that $\dep(f^{\textsf{Stack}}_{\substack{i,j}}) = \{i,i+1,j-1,j\}$.
-
+For instance, the contribution of a \Def{base pair stack}, consisting of two pairs $(i,j)$ and $(i+1,j-1)$, can be captured by the introduction of a function $f^{\textsf{Stack}}_{\substack{i,j}}$ such that $\dep(f^{\textsf{Stack}}_{\substack{i,j}}) = \{i,i+1,j-1,j\}$.
+        We refer to energy models that consider the contributions of all base pair stacks (and thus introduce the corresponding dependencies) collectively as the \Def{stacking energy model}.
 }
-
-\todo[inline]{Check use of the term stacking model; the term is not introduced here yet!}
 
@@
 The \Def{treewidth} of a tree decomposition $(T,\chi)$ is defined as
-$$\min_{u\in T} |\chi(u)| - 1.$$ 
-The \Def{treewidth}
-of a graph $G$ is the smallest treewidth of any tree decomposition of $G$.
-\todo[inline]{Check: do we really need to define treewidth of a graph?}
+$\min_{u\in T} |\chi(u)| - 1.$ 
+%The \Def{treewidth} of a graph $G$ is the smallest treewidth of any tree decomposition of $G$.
 \end{definition}
@@
 \caption{\revised{Toy example of a tree decomposition associated with two target structures in the stacking energy model.
-Two target secondary structures ({\bfseries\sffamily A}) are merged into a joint hypergraph ({\bfseries\sffamily B}), whose hyperedges correspond to the quadruplets of positions involved in stacking pairs (colored).
-A valid tree decomposition ({\bfseries\sffamily C}) for the hypergraph ensures, among other properties, that each base pair and each stacking pair/hyperedge is represented in at least one of its node, so that features can be adequately evaluated.
-The width of this tree decomposition is 3, a provably optimal value for this input hypergraph.}\label{fig:treedec}}
+Two target secondary structures ({\bfseries\sffamily A}) are merged into a joint hypergraph ({\bfseries\sffamily B}), whose hyperedges correspond to the quadruplets of positions involved in base pair stacks (colored).
+A valid tree decomposition ({\bfseries\sffamily C}) for the hypergraph ensures, among other properties, that each base pair and each base pair stack is represented in at least one of its node, so that features can be adequately evaluated.
+The treewidth of this tree decomposition is 3, a provably optimal value for this input hypergraph.}\label{fig:treedec}}
 \end{figure}
@@
 
-By this theorem, the complexity is polynomial for fixed value of $w$, and Boltzmann sampling in our setting is thus fixed parameter tractable (FPT) in the tree width.
+By this theorem, the complexity is polynomial for fixed value of $w$, and Boltzmann sampling in our setting is thus fixed parameter tractable (FPT) in the treewidth.
 %
@@
 In order to capture realistic energy models like the Turner model or pseudoknot models like {\tt HotKnots}~\citep{Ren2005}, our sampling strategy can be extended in two ways: 1) either by directly sampling based on more expressive energy models; 2) or by sampling in a simple energy model can be used to approximate sampling in more complex models.
-In practice, complex energy models has a strong influence on the tree width of the dependency graph and, in turn computational complexity of our approach.
+In practice, complex energy models has a strong influence on the treewidth (of optimal tree decompositions) of the dependency graph and, in turn the computational complexity of our approach.
 Therefore, it is interesting to consider, in addition to the base pair model, other
@@
 %We define the following notations: $\real^\infty:=\real\cup\{\infty\}$, $[i,j]:=\{i,\ldots,j\},$ \ldots
-\revised{While efficient, both in practice and in theory for graphs of bounded tree width, our algorithms remain exponential in the worst case scenario, since the tree-width of a dependency graph can then become arbitrarily large.
+\revised{While efficient, both in practice and in theory for graphs of bounded treewidth, our algorithms remain exponential in the worst case scenario, since the treewidth of a dependency graph can then become arbitrarily large.
 This exponential complexity in the worst case appears to be intrinsic.
@@
 Figure 10 - Correlation plot -- Right
-Figure 2 -- Distribution plot -- Right (Turner is even untractable to compute the tree width itself in benchmark instances)
+Figure 2 -- Distribution plot -- Right (Turner is even untractable to compute the treewidth itself in benchmark instances)
 Figure 3 -- Right -- Distribution

### 2019-01-24 supplement.tex
Updates from ShareLaTeX

@@
 
+\begin{figure}[h!]
+  %\hspace*{2cm}\textbf{Base pair model}\hspace{4.75cm}\textbf{Stacking model}\\[-28pt]
+  \centering\includegraphics[width=0.9\textwidth]{Figs/td-widths}
+  \caption{Treewidths for multi-target RNA design instances of
+    different complexity. Distributions of treewidths shown as boxplots for the base pair (left) and stacking model (right).}
+  \label{fig:td-widths}
+\end{figure}
+
+\begin{figure}[h!]
+  %\hspace*{2cm}\textbf{Base pair model}\hspace{4.75cm}\textbf{Stacking model}\\[-28pt]
+  \centering
+  \includegraphics[width=0.9\textwidth]{Figs/td-times}
+  \caption{Computation time of tree decompositions for
+    multi-target RNA design instances of different complexity.
+    Distributions of times (in ms/instance) shown as boxplots for the base pair (left) and stacking model (right).}
+  \label{fig:td-times}
+\end{figure}
+
 In addition, we generated four sets of instances of increasing
@@
 \begin{figure}[h!]
-  %\hspace*{2cm}\textbf{Base pair model}\hspace{4.75cm}\textbf{Stacking model}\\[-28pt]
-  \centering\includegraphics[width=0.9\textwidth]{Figs/td-widths}
-  \caption{Treewidths for multi-target RNA design instances of
-    different complexity. Distributions of treewidths shown as boxplots for the base pair (left) and stacking model (right).}
-  \label{fig:td-widths}
-\end{figure}
-
-\begin{figure}[h!]
-  %\hspace*{2cm}\textbf{Base pair model}\hspace{4.75cm}\textbf{Stacking model}\\[-28pt]
+  %% the figure shows benchmark_inputs/RNAfold/3str/f3.100.0.inp
   \centering
-  \includegraphics[width=0.9\textwidth]{Figs/td-times}
-  \caption{Computation time of tree decompositions for
-    multi-target RNA design instances of different complexity.
-    Distributions of times (in ms/instance) shown as boxplots for the base pair (left) and stacking model (right).}
-  \label{fig:td-times}
+  {\small \verbatiminput{Figs/f3_100_0.txt}}
+  \includegraphicstop[width=0.55\textwidth]{Figs/td-example-basepair}%
+  \includegraphicstop[width=0.30\textwidth]{Figs/td-example-stacking}
+  \caption{Example instance from RF3 (top) with its tree decompositions in the base pair (left) and stacking model (right). The respective treewidths are 2 and 12.}
+  \label{fig:td-example}
 \end{figure}
@@
 
-\begin{figure}[h!]
-  %% the figure shows benchmark_inputs/RNAfold/3str/f3.100.0.inp
-  \centering
-  \verbatiminput{Figs/f3_100_0.txt}
-  \includegraphicstop[width=0.6\textwidth]{Figs/td-example-basepair}%
-  \includegraphicstop[width=0.35\textwidth]{Figs/td-example-stacking}
-  \caption{Example instance from RF3 (top) with its tree decompositions in the base pair (left) and stacking model (right). The respective treewidths are 2 and 12.}
-  \label{fig:td-example}
-\end{figure}
-
 \begin{revision}
@@
 \section{Run times for generating samples}
-
-To illustrate the efficiency and practical feasibility of our Boltzmann sampling approach, we show the distributions of run times (Intel(R) Core(TM) i7-4770 CPU) for generating 10000 samples (over the Modena benchmark instances and grouped by benchmark sub-set).
-Since producing a controlled Boltzmann sampling requires significant pre-computation, it is informative to separately show construction and sampling times (Figure~\ref{fig:run-times-redprint}).
-Moreover, \ourprog{} can be used with a simpler base pair model or a rather complex stacking energy model, which introduces much stronger dependencies than required for uniform sampling.
-Technically, these stronger dependencies increase the tree width and thus the runtime, as noticeable in Figure~\ref{fig:run-times-redprint}.
-
-A fair comparison to sampling times by other tools can hardly be given, since the other tools produce samplings of very different nature.
-In particular, Modena does not make any guarantees on properties of the sampling.
-Non-surprisingly, this allows it to sample significantly faster (several orders of magnitudes).
-Nevertheless, the comparison to run times of \RNAblueprint{} is still interesting, since this tool generates---similarly to our tool---\emph{controlled} samples.
-As we argue that the Boltzmann sampling is fundamentally more flexible than uniform sampling as generated by \RNAblueprint, \ourprog{} performs even particularly well, beating \RNAblueprint{} most of the time with only a few exceptions for particularly hard instances.
-The results for \RNAblueprint{} are shown in Figure~\ref{fig:run-times-rbp}.
 
@@
 \end{figure}
+
+To illustrate the efficiency and practical feasibility of our Boltzmann sampling approach, we show the distributions of run times (Intel(R) Core(TM) i7-4770 CPU) for generating 10000 samples (over the Modena benchmark instances and grouped by benchmark sub-set).
+Since producing a controlled Boltzmann sampling requires significant pre-computation, it is informative to separately show construction and sampling times (Figure~\ref{fig:run-times-redprint}).
+Moreover, \ourprog{} can be used with a simpler base pair model or a rather complex stacking energy model, which introduces much stronger dependencies than required for uniform sampling.
+Technically, these stronger dependencies increase the tree width and thus the runtime, as noticeable in Figure~\ref{fig:run-times-redprint}.
+
+A fair comparison to sampling times by other tools can hardly be given, since the other tools produce samplings of very different nature.
+In particular, Modena does not make any guarantees on properties of the sampling.
+Non-surprisingly, this allows it to sample significantly faster (several orders of magnitudes).
+Nevertheless, the comparison to run times of \RNAblueprint{} is still interesting, since this tool generates---similarly to our tool---\emph{controlled} samples.
+As we argue that the Boltzmann sampling is fundamentally more flexible than uniform sampling as generated by \RNAblueprint, \ourprog{} performs even particularly well, beating \RNAblueprint{} most of the time with only a few exceptions for particularly hard instances.
+The results for \RNAblueprint{} are shown in Figure~\ref{fig:run-times-rbp}.
 
@@
 \section{Effectivity of the seed sequence generation}
+
 Figure~\ref{fig:seed-generation} shows relative numbers of finished instances and admissible sequences at each iteration of the seed generation procedure for the benchmark set \texttt{3str}.
@@
   \centering
-  \includegraphics[width=0.7\textwidth]{Figs/SeedGeneration/iterations_gc_3str.pdf}\\
-  \includegraphics[width=0.7\textwidth]{Figs/SeedGeneration/iterations_en_3str.pdf}
+  \includegraphics[width=0.48\textwidth]{Figs/SeedGeneration/iterations_gc_3str.pdf}
+  \includegraphics[width=0.48\textwidth]{Figs/SeedGeneration/iterations_en_3str.pdf}
   \caption{Relative numbers of finished instances and admissible sequences at each iteration of the seed generation procedure for the benchmark set \texttt{3str}.}
@@
 \end{figure}
+
 \end{revision}

### 2019-01-24 supplement.tex

@@
 
+In addition, we generated four sets of instances of increasing
+complexity. The instances of the sets RF3, RF4, RF5, and RF6, each
+respectively specify 3,4,5, and 6 target structures for sequence
+length 100.  For each instance (100 instances per set), we generated a
+set of $k$ ($k=3,\dots,6$) compatible structures as follows
+
+\begin{itemize}
+\item Generate a random sequence of length 100;
+\item Compute its minimum free energy structure (ViennaRNA package);
+\item Add the new structure to the instances if the resulting base pair dependency graph is bipartite;
+\item Repeat until $k$ structures are collected.
+\end{itemize}
+
 \begin{figure}[h!]
   %\hspace*{2cm}\textbf{Base pair model}\hspace{4.75cm}\textbf{Stacking model}\\[-28pt]
-  \centering\includegraphics[width=0.9\textwidth]{Figs/td-widths}
+  \centering\includegraphics[width=0.8\textwidth]{Figs/td-widths}
   \caption{Treewidths for multi-target RNA design instances of
@@
   \centering
-  \includegraphics[width=0.9\textwidth]{Figs/td-times}
+  \includegraphics[width=0.8\textwidth]{Figs/td-times}
   \caption{Computation time of tree decompositions for
@@
 
-In addition, we generated four sets of instances of increasing
-complexity. The instances of the sets RF3, RF4, RF5, and RF6, each
-respectively specify 3,4,5, and 6 target structures for sequence
-length 100.  For each instance (100 instances per set), we generated a
-set of $k$ ($k=3,\dots,6$) compatible structures as follows
-\begin{itemize}
-\item Generate a random sequence of length 100;
-\item Compute its minimum free energy structure (ViennaRNA package);
-\item Add the new structure to the instances if the resulting base pair dependency graph is bipartite;
-\item Repeat until $k$ structures are collected.
-\end{itemize}
 For each instance, we generated the dependency graphs in the base pair
@@
 
+To illustrate the efficiency and practical feasibility of our Boltzmann sampling approach, we show the distributions of run times (Intel(R) Core(TM) i7-4770 CPU) for generating 10000 samples (over the Modena benchmark instances and grouped by benchmark sub-set).
+Since producing a controlled Boltzmann sampling requires significant pre-computation, it is informative to separately show construction and sampling times (Figure~\ref{fig:run-times-redprint}).
+Moreover, \ourprog{} can be used with a simpler base pair model or a rather complex stacking energy model, which introduces much stronger dependencies than required for uniform sampling.
+Technically, these stronger dependencies increase the tree width and thus the runtime, as noticeable in Figure~\ref{fig:run-times-redprint}.
+
 \begin{figure}[h!]
   \centering
-  \includegraphics[width=0.45\textwidth]{Figs/RunTimes/construction_time_sets}\quad%
-  \includegraphics[width=0.45\textwidth]{Figs/RunTimes/sample_time_sets}
+  \includegraphics[width=0.43\textwidth]{Figs/RunTimes/construction_time_sets}\quad%
+  \includegraphics[width=0.43\textwidth]{Figs/RunTimes/sample_time_sets}
   \caption{Boxplots showing the run-time distribution of sequence sampling with \ourprog{} for the various benchmark sets.
@@
   \centering
-  \includegraphics[width=0.45\textwidth]{Figs/RunTimes/construction_time_sets_rbp}\quad%
-  \includegraphics[width=0.45\textwidth]{Figs/RunTimes/sample_time_sets_rbp}
+  \includegraphics[width=0.43\textwidth]{Figs/RunTimes/construction_time_sets_rbp}\quad%
+  \includegraphics[width=0.43\textwidth]{Figs/RunTimes/sample_time_sets_rbp}
   \caption{Boxplots showing the run-time distribution of sequence sampling with \RNAblueprint{} for the various benchmark sets.
@@
 \end{figure}
-
-To illustrate the efficiency and practical feasibility of our Boltzmann sampling approach, we show the distributions of run times (Intel(R) Core(TM) i7-4770 CPU) for generating 10000 samples (over the Modena benchmark instances and grouped by benchmark sub-set).
-Since producing a controlled Boltzmann sampling requires significant pre-computation, it is informative to separately show construction and sampling times (Figure~\ref{fig:run-times-redprint}).
-Moreover, \ourprog{} can be used with a simpler base pair model or a rather complex stacking energy model, which introduces much stronger dependencies than required for uniform sampling.
-Technically, these stronger dependencies increase the tree width and thus the runtime, as noticeable in Figure~\ref{fig:run-times-redprint}.
 

### 2019-01-24 RNARedPrint.tex
minor reformulation

@@
 Since $\F$ defines potentially more complex dependencies, which can
-relate more than two positions, its dependencies cannot be similarily
+relate more than two positions, in general its dependencies cannot be 
 represented by a graph. Instead, this requires a structure known as
 \Def{hypergraph}, which consists of vertices (here, the sequence
-positions) connected by \Def{hyperedges}, which are sets of
+positions) connected by \Def{hyperedges}, which are arbitrary sets of
 vertices. In this way, hypergraphs generalize undirected graphs, where
-edges are sets of two vertices.
-%Undirected graphs can be
-%modeled as hypergraphs whose hyperedges are pairs of vertices, but
-%hypergraphs also allow to express complex dependencies.
-
-The set $\F$ of contributions induces the \Def{dependency
-  (hyper)graph} $G_\F=(V, H)$ on sequence positions $V=\{1,\dots,n\}$
+each edge is a set of exactly two vertices.
+%
+The \Def{dependency (hyper)graph induced by $\F$} is then defined as the hypergraph $G_\F=(V, H)$ on sequence positions $V=\{1,\dots,n\}$
 by interpreting the dependencies as hyperedges, i.e. $H =

### 2019-01-24 RNARedPrint.tex
Minor edits; edit tree decomp intuition

@@
 
-\revised{Moreover, we could show that it is not possible to find a better, efficient method for sampling (unless ${\sf P}={\sf NP}$), since the underlying counting problem is \#{\sf P}-hard.
-The practical relevance of this theoretical result is that it rules out substantially better sampling techniques, such that there will always---even using improved sampling methods----remain an upper limit on the---in practice---tractable number and heterogeneity of structures, the complexity of the directly treatable energy model, and the number and complexity of additional constraints that could be considered in future sampling-based applications.} Technically, this result relies on a surprising bijection between valid sequences and independent sets of a bipartite graph, the latter being the object of recent breakthroughs in approximate counting complexity~\citep{Bulatov2013,Cai2016}.
+\revised{Remarkably, we could show that it is not possible to find a better, efficient method for sampling (unless ${\sf P}={\sf NP}$), since the underlying counting problem is \#{\sf P}-hard.
+The practical relevance of this theoretical result is that it rules out substantially better sampling techniques, such that there will always---even using improved sampling methods---remain an upper limit on the---in practice---tractable number and heterogeneity of structures, the complexity of the directly treatable energy model, and the number and complexity of additional constraints that could be considered in future sampling-based applications.} Technically, this result relies on a surprising bijection between valid sequences and independent sets of a bipartite graph, the latter being the object of recent breakthroughs in approximate counting complexity~\citep{Bulatov2013,Cai2016}.
 
@@
 %
-Due to the generality of our method, we can moreover strongly limit this parameter in practice by using state-of-the-art tree decomposition algorithms.
+Due to the generality of our method, we can moreover strongly limit \revised{the treewidth} in practice by using state-of-the-art tree decomposition algorithms.
 %Uniform sampling is handled as a special case of Boltzmann sampling, where each valid sequence receives energy zero and---consequently---computing partition functions specializes to counting.
@@
  on a comprehensive multi-target RNA design benchmark~\citep{Taneda2015} 
-  suggest that our sampling strategy supports designing biologically relevant
+  suggest that our sampling strategy well supports designing biologically relevant
 RNAs for multiple targets.}
@@
 Sampling from the (multi-feature) Boltzmann distribution requires to compute corresponding partition functions, such that we can draw sequences with probabilities proportional to their Boltzmann weight.
-On this basis, we can finely calibrate the weights, to increase the probability that sampled sequences meet prescribed target values for each feature.
+On this basis, we can finely calibrate the weights, to increase the probability that sampled sequences meet the desired target values for each feature.
 Together with a final rejection step this results in an effective procedure for generating highly specific sequences.}
@@
 
-The $\GCb\%$ can be trivially expressed using $n$ contributions $f^{\GCb}_i$, $i\in[1,n]$, whose dependencies are restricted to individual nucleotides,  and such that 
+The $\GCb\%$ can be simply expressed using $n$ contributions $f^{\GCb}_i$, $i\in[1,n]$, whose dependencies are restricted to individual nucleotides,  and such that 
 $$f^{\GCb}_i(\{i\mapsto c\})=\begin{cases}-1&\text{ if }c=\Gb\text{ or }\Cb\\ 0&\text{ otherwise}.\end{cases}$$ 
-\revised{By summing $f^{\GCb}_i$ over the whole sequence, one simply counts the occurrences of $\Gb$ and $\Cb$.}
+\revised{By summing $f^{\GCb}_i(\{i\mapsto S_i\})$ over the whole sequence, one simply counts the occurrences of $\Gb$ and $\Cb$.}
 
@@
 
-Let us now define the \Def{tree decomposition} of the graph $G_\F$, a fundamental ingredient of our algorithms, which also determines their efficiency (most importantly, via its property called treewidth).
+Let us finally define the \Def{tree decomposition} of the graph $G_\F$, a fundamental ingredient of our algorithms, which also determines their efficiency (most importantly, via its property called treewidth).
 }
@@
 \end{definition}
-
-\revised{Intuitively, a tree decomposition for an (hyper)graph $G$ is a tree structure which features all the vertices of $G$, captures all the (hyper)edges of $G$, and helps ensure some notion of consistency by allowing the propagation of values.
+%
+\revised{Intuitively, a tree decomposition of an (hyper)graph $G$ is a tree that captures all the vertices and (hyper)edges of $G$, and properly relates dependend subproblems to ensure consistency in a recursive computation.
 Figure~\ref{fig:treedec} shows an optimal tree decomposition for a pair of structures under the stacking energy model.}

### 2019-01-24 RNARedPrint.tex
edit caption of toy tree decomp figure

@@
 {\centering \includegraphics[width=.9\linewidth]{Figs/example-treedec-stacking}\\}
-\caption{\revised{Toy example of a tree decomposition associated with two target structures in the stacking energy model.
+\caption{\revised{Toy example of a tree decomposition associated with two target structures in the stacking energy model (where the four positions of each base pair stack depend on each other).
 Two target secondary structures ({\bfseries\sffamily A}) are merged into a joint hypergraph ({\bfseries\sffamily B}), whose hyperedges correspond to the quadruplets of positions involved in base pair stacks (colored).
-A valid tree decomposition ({\bfseries\sffamily C}) for the hypergraph ensures, among other properties, that each base pair and each base pair stack is represented in at least one of its node, so that features can be adequately evaluated.
+A valid tree decomposition ({\bfseries\sffamily C}) for the hypergraph ensures, among other properties, that each base pair and each base pair stack is represented in at least one of its node, so that features can be correctly evaluated.
 The treewidth of this tree decomposition is 3, a provably optimal value for this input hypergraph.}\label{fig:treedec}}

### 2019-01-24 RNARedPrint.tex
Partially rewrite abstract (thanks to Maria for comments)

@@
   The design of multi-stable RNA molecules has important applications in biology, medicine, and biotechnology.
-Synthetic design approaches profit strongly from effective in-silico methods, which impact their cost and feasibility.
-\\
+Synthetic design approaches profit strongly from effective in-silico methods, \revised{which substantially reduce the need for costly wet-lab experiments.} \\
 %
   \parttitle{Results} We devise a novel approach to a central ingredient of most in-silico
-  design methods: the generation of sequences \revised{that fold well into multiple target structures.} For this task, we present the fixed-parameter tractable (FPT) approach \ourprog{}.
-  % 
-  \revised{Based on} constraint networks, \ourprog{} supports generic
+  design methods: the generation of sequences \revised{that fold well into multiple target structures.} 
+  % 
+  \revised{Based on constraint networks, our approach \ourprog{}} supports generic
   Boltzmann-weighted sampling, which enables the positive design of RNA sequences
@@
 %
-  %Securing the theoretical foundations of our method
-  Moreover, we establish the $\#${\sf P}-hardness of \revised{counting to theoretically justify the development of an FPT approach}. 
-%
-  Finally, we study general properties of our approach empirically and generate
+  Moreover, we study general properties of our approach empirically and generate
   biologically relevant multi-target Boltzmann-weighted designs for a
-  common design benchmark.
-  We demonstrate significant improvements over the previously best multi-target sampling strategy, due to generating seed sequences with \ourprog{}.\\
-%
-  \parttitle{Conclusion} \ourprog{} introduces a novel, flexible and---as we demonstrate--effective approach to multi-target RNA design, which promises broad applicability and extensibility.
+  \revised{established} design benchmark. \revised{Our results demonstrate the effectivity and feasiblity of the method in practice
+  as well as the benefits of Boltzmann sampling over the previously best multi-target sampling strategy---even for the case of negative design of multi-stable RNAs.}\\
+%
+  \revised{Besides empirically studies, we finally justify the algorithmic details due to a fundamental theoretic result about multi-stable RNA design, namely the \#P-hardness of the counting of designs.}
+%
+  \parttitle{Conclusion} \ourprog{} introduces a novel, flexible and---as we demonstrate---effective approach to multi-target RNA design, which promises broad applicability and extensibility.
 

### 2019-01-24 RNARedPrint.tex

@@
 coloring the sequence positions with nucleotides, such that the ends of each base pair receive compatible colors.
-Initially the method decomposes the graph hierarchically, using ear decomposition, and then \Def{precomputes} the number of valid sequences
+Initially, the method decomposes the graph hierarchically and then \Def{precomputes} the number of valid sequences
 within each subgraph. The decomposition is then reinterpreted as a
@@
 \revised{As central contribution, we provide a systematic and flexibly extensible technique for sampling that targets multiple versatile features.
-For the sake of clarity, we introduce this method specialized to the sampling of sequences that target the energies of multiple structures and the \GCb\%.}
+For the sake of clarity, we introduce this method specialized to the sampling of RNA sequences that target the energies of multiple structures and the \GCb\%.}
 

### 2019-01-24 RNARedPrint.tex
minor reformulations

--- 2019-01-24 RNARedPrint.tex
@@
 
-\revised{Fig.~\ref{fig:workflow} summarizes our generic framework, which enables the generation of sequence samples with specific energies for multiple target structures (based on multi-dimensional Boltzmann sampling).
+\revised{Figure~\ref{fig:workflow} summarizes our generic framework, which enables this targeted sequence generation based on multi-dimensional Boltzmann sampling.
 \emph{Algorithmically}, we originally contribute} dynamic programming (DP) algorithms, based on the concept of \Def{tree decomposition}, to compute partition functions and sample sequences from the Boltzmann distribution%
@@
 Sampling from the (multi-feature) Boltzmann distribution requires to compute corresponding partition functions, such that we can draw sequences with probabilities proportional to their Boltzmann weight.
-On this basis, we can finely calibrate the weights, to increase the probability that sampled sequences meet the desired target values for each feature.
+On this basis, we can finely calibrate the weights, to maximize the probability that sampled sequences meet the desired target values for each feature.
 Together with a final rejection step this results in an effective procedure for generating highly specific sequences.}

### 2019-01-24 RNARedPrint.tex
Updates from ShareLaTeX

--- 2019-01-24 RNARedPrint.tex
@@
 
-\revised{Eventually our proof-of-concept results 
+\revised{Eventually, our proof-of-concept results 
  on a comprehensive multi-target RNA design benchmark~\citep{Taneda2015} 

### 2019-01-24 RNARedPrint.tex
Fix app refs (partially)

--- 2019-01-24 RNARedPrint.tex
@@
 the correctness of the algorithm
-(Supp. Mat.~\ref{appsec:correctness}).  After running
+(Supp.\ Mat.~\ref{appsec:correctness}).  After running
 Alg.~\ref{alg:pf}, multiplying the 0-ary messages sent to the root $r$
@@
 $\mathcal{O}((|\F|+|V|)\cdot 4^{w+1})$ time and
-$\mathcal{O}(|V|\cdot4^s)$ space
-(Supp. Mat.~\ref{appsec:algcomplexity}); Alg.~\ref{alg:sampling} would
+$\mathcal{O}(|V|\cdot4^s)$ space; Alg.~\ref{alg:sampling} would
 require $\mathcal{O}((|\F|+|V|)\cdot 4^D)$ per sample on arbitrary
 tree decompositions
-(Supp. Mat.~\ref{appsec:algcomplexity}). W.l.o.g.\ we assume that
+(Supp.\ Mat.~\ref{appsec:algcomplexity}). W.l.o.g.\ we assume that
 $D=1$; note that tree decompositions can generally be transformed,
@@
 %
-As shown in Supp. Mat.~\ref{sec:improvedComplexity}, the complexity of the precomputation can be further improved to
-$\mathcal{O}(n\,k\,2^{w+1}\,2^{c})$, where $c$ is the maximum number of connected components represented in a node of the tree decomposition ($c\le w+1$).
+The complexity of the precomputation can be further improved to
+$\mathcal{O}(n\,k\,2^{w+1}\,2^{c})$, where $c$ ($c\le w+1$) is the maximum number of connected components represented in a node of the tree decomposition (Supp.\ Mat.~\ref{appsec:improvedComplexity}).
 
@@
 negligible compared to Alg.~\ref{alg:pf}
-(Supp. Mat.~\ref{appsec:treedecomp} and
+(Supp.\ Mat.~\ref{appsec:treedecomp} and
 \ref{appsec:dependency-cliques}).
@@
 $\EnergyTurner$ and the fitted stacking model $\EnergyStacking$
-(Supp. Fig~\ref{fig:training-cor}). More precisely, we observed a
+(Supp.\ Mat.~\ref{appsec:modelparameters}). More precisely, we observed a
 structure-specific affine dependency between the Turner and stacking
@@
 In other words,  $\mathbb{E}(F_\ell(S)\mid \pmb{\pi})=\TargetE_\ell$,  $\forall\ell\in[1,k]$ and, analogously, the expectation of $F_0(S)$ is the targeted GC content.
-The expected value of $F_\ell$ is always decreasing for increasing weights $\pi_\ell$ (see Supp.
-Mat.~\ref{sec:weight_derivatives}).
+The expected value of $F_\ell$ is always decreasing for increasing weights $\pi_\ell$ (see Supp.\ Mat.~\ref{appsec:weight_derivatives}).
 More generally, computing a suitable parameter vector $\pmb{\pi}$ can be restated as a convex optimization problem, and be efficiently solved using a wide array of methods~\citep{Denise2010,Bendkowski2017}.
@@
 for its targeting performance (both models:
-Supp. Mat.~\ref{appsec:modelparameters}).
+Supp.\ Mat.~\ref{appsec:modelparameters}).
 %
@@
 to a large extent, independently shift the Turner energies of sampled
-sequences towards prescribed targets. See Supp. Mat.,
-Fig.~\ref{fig:energydist} for a further example with three
+sequences towards prescribed targets. See Supp.\ Mat.~\ref{appsec:illustrating-mdbs} for a further example with three
 pseudoknot-free target structures.}
@@
 improvement is be observed for both of the terms in \Obj{} (see
-Supp. Mat.~\ref{sec:improvements}).  This means that \ourprog{}
+Supp. Mat.~\ref{appsec:contributions-multidefect}).  This means that \ourprog{}
 produces sequences whose targets are both substantially closer to
@@
 structures, we observed better \Obj{} for Boltzmann sampling than for
-uniform sampling (see Supp. Mat.~\ref{sec:validity}).  Notably,
+uniform sampling (see Supp. Mat.~\ref{appsec:detailed-benchmark-results}).  Notably,
 \ourprog{} performs equally well in the presence of pseudoknots;
@@
 remains unchallenged even after local optimizations, as shown by
-Fig.~\ref{fig:benchmark-results} and Supp. Mat.~\ref{sec:improvements}
-and \ref{sec:validity}. This observation is consistent with a
+Fig.~\ref{fig:benchmark-results} and Supp. Mat.~\ref{appsec:contributions-multidefect}
+and \ref{appsec:detailed-benchmark-results}. This observation is consistent with a
 superior quality of the starting point for the greedy walk, probably leading

### 2019-01-24 supplement.tex
Fix app refs (partially)

@@
 \section{Immediate benefits of positive design for negative design}
-
+\label{appsec:immediate-benefits-for-negative-design}
 Already previous work on RNA Design, e.g. INFO-RNA and Incarnation, observed
@@
 %
-Consequently, Figure~\ref{fig:negforpos} ultimately illustrates the dependencies
+Consequently, Figure~\ref{appfig:negforpos} ultimately illustrates the dependencies
 between $Pr(S|R)$ and $Pr(R|S)$ for uniform samples and low energy samples
 (where those samples can be efficiently generated due to our Boltzmann sampling
-strategy). For our examples, we observe (Fig.~\ref{fig:negforpos}) that Boltzmann sampling with weight 500 (green), generates good negative designs 
+strategy). For our examples, we observe (Fig.~\ref{appfig:negforpos}) that Boltzmann sampling with weight 500 (green), generates good negative designs 
 with substantially higher probability than uniform sampling (blue).
@@
 
-\begin{figure}[h!]
+\begin{figure}[!h]
 \textbf{A}~\includegraphicstop[width=0.48\textwidth]{Figs/Plots/stem_weight1-500_mean}%
@@
 Note that while only the energy $E(S,R)$ is directly controlled by Boltzmann sampling, this strongly impacts the common negative design objective $Pr(R|S)$.}
-\label{fig:negforpos}
+\label{appfig:negforpos}
 \end{figure}
@@
     different complexity. Distributions of treewidths shown as boxplots for the base pair (left) and stacking model (right).}
-  \label{fig:td-widths}
+  \label{appfig:td-widths}
 \end{figure}
@@
     Distributions of times (in ms/instance) shown as boxplots for the base pair (left) and stacking model (right).}
-  \label{fig:td-times}
+  \label{appfig:td-times}
 \end{figure}
@@
 graph. The obtained treewidths are reported in
-Fig.~\ref{fig:td-widths}, while Fig.~\ref{fig:td-times} shows the
+Fig.~\ref{appfig:td-widths}, while Fig.~\ref{appfig:td-times} shows the
 corresponding run-times of the tree decomposition.
@@
   \caption{Example instance from RF3 (top) with its tree decompositions in the base pair (left) and stacking model (right). The respective treewidths are 2 and 12.}
-  \label{fig:td-example}
+  \label{appfig:td-example}
 \end{figure}
@@
 \revised{
-  Finally, for further illustration, we show concrete tree decompositions for a selected instance in the benchmark set RF3 in Figure~\ref{fig:td-example}.
+  Finally, for further illustration, we show concrete tree decompositions for a selected instance in the benchmark set RF3 in Figure~\ref{appfig:td-example}.
 This figure is provided solely for illustrational purposes:
@@
 
-\section{Run times for generating samples}
+\section{Run times for Boltzmann sampling (at fixed weights)}
+\label{appsec:run-times}
 
 To illustrate the efficiency and practical feasibility of our Boltzmann sampling approach, we show the distributions of run times (Intel(R) Core(TM) i7-4770 CPU) for generating 10000 samples (over the Modena benchmark instances and grouped by benchmark sub-set).
-Since producing a controlled Boltzmann sampling requires significant pre-computation, it is informative to separately show construction and sampling times (Figure~\ref{fig:run-times-redprint}).
+Since producing a controlled Boltzmann sampling requires significant pre-computation, it is informative to separately show construction and sampling times (Figure~\ref{appfig:run-times-redprint}).
 Moreover, \ourprog{} can be used with a simpler base pair model or a rather complex stacking energy model, which introduces much stronger dependencies than required for uniform sampling.
-Technically, these stronger dependencies increase the tree width and thus the runtime, as noticeable in Figure~\ref{fig:run-times-redprint}.
+Technically, these stronger dependencies increase the tree width and thus the runtime, as noticeable in Figure~\ref{appfig:run-times-redprint}.
 
@@
 }
-  \label{fig:run-times-redprint}
+  \label{appfig:run-times-redprint}
 \end{figure}
@@
 Construction includes all the pre-computation, while the latter is the time needed to generate 10000 sequences.}
-  \label{fig:run-times-rbp}
+  \label{appfig:run-times-rbp}
 \end{figure}
@@
 As we argue that the Boltzmann sampling is fundamentally more flexible than uniform sampling as generated by \RNAblueprint, \ourprog{} performs even particularly well, beating \RNAblueprint{} most of the time with only a few exceptions for particularly hard instances.
-The results for \RNAblueprint{} are shown in Figure~\ref{fig:run-times-rbp}.
+The results for \RNAblueprint{} are shown in Figure~\ref{appfig:run-times-rbp}.
 
@@
 \section{Illustrating multi-dimensional Boltzmann sampling for three pseudoknot-free target structures}
+\label{appsec:illustrating-mdbs}
 
 In analogy to Fig.~\ref{fig:energydist-pk} of the main text,
-    Fig.~\ref{fig:energydist} illustrates our multi-dimensional sampling
+    Fig.~\ref{appfig:energydist} illustrates our multi-dimensional sampling
     strategy for three pseudoknot-free target structures (example instance
@@
 
-    Fig.~\ref{fig:gc-content} shows the distributions of the GC content of the samples, while targeting different combinations of energies in this experiment.
+    Fig.~\ref{appfig:gc-content} shows the distributions of the GC content of the samples, while targeting different combinations of energies in this experiment.
 For completeness, we also report corresponding results for the analogous experiment from the main text.
@@
   }
-  \label{fig:energydist}
+  \label{appfig:energydist}
 \end{figure}
@@
   \end{center}
-  \caption{Distribution of the \GCb-content \GCb\% in the experiments of (A) Fig.~\ref{fig:energydist-pk} of the main text and (B) Suppl.~Fig.~\ref{fig:energydist}.}
-  \label{fig:gc-content}
+  \caption{Distribution of the \GCb-content \GCb\% in the experiments of (A) Fig.~\ref{fig:energydist-pk} of the main text and (B) Suppl.~Fig.~\ref{appfig:energydist}.}
+  \label{appfig:gc-content}
 \end{figure}
@@
 \section{Effectivity of the seed sequence generation}
+\label{appsec:effectivity-seed-generation}
 
-Figure~\ref{fig:seed-generation} shows relative numbers of finished instances and admissible sequences at each iteration of the seed generation procedure for the benchmark set \texttt{3str}.
+Figure~\ref{appfig:seed-generation} shows relative numbers of finished instances and admissible sequences at each iteration of the seed generation procedure for the benchmark set \texttt{3str}.
 We obtain qualitatively equivalent results for all other benchmark sets (not shown).
-The top plot illustrates the  precomputation step of our strategy, where weights are calibrated for \GCb\% of 0.6.
+The first plot illustrates the  precomputation step of our strategy, where weights are calibrated for \GCb\% of 0.6.
 In this step, we produce sequence of the targeted \GCb\% to determine the linear transformation from the sampling energy model (stacking model) to the Turner model.
-The bottom plot shows the actual generation of the seed sequences after precomputation.
+The second plot shows the actual generation of the seed sequences after precomputation.
 The plot shows that most of the instances are finished after the first iterations, with only a few harder instances that require 2--3 times more iterations.
@@
   \includegraphics[width=0.48\textwidth]{Figs/SeedGeneration/iterations_en_3str.pdf}
-  \caption{Relative numbers of finished instances and admissible sequences at each iteration of the seed generation procedure for the benchmark set \texttt{3str}.}
-  \label{fig:seed-generation}
+  \caption{Relative numbers of finished instances and admissible sequences at each iteration of the seed generation procedure for the benchmark set \texttt{3str}.
+The left illustrates the procedure for targeting just the \GCb\% in the precomputation step; this step is used to learn instance-specific transformations of approximate energies to the Turner model from the generated sample.
+The right plots illustrates generating samples targeting \GCb\% and the Turner energies.
+}
+  \label{appfig:seed-generation}
 \end{figure}
@@
 independent test set of random RNA sequences in the same
-way. Fig.~\ref{fig:training-cor} shows correlation plots for the
+way. Fig.~\ref{appfig:training-cor} shows correlation plots for the
 trained parameters in the base pair and stacking models for predicting
@@
     ViennaRNA package.}
-  \label{fig:training-cor}
+  \label{appfig:training-cor}
 \end{figure}
@@
 
-    \section{Exploiting constraint consistency to reduce the complexity}\label{sec:improvedComplexity}
+    \section{Exploiting constraint consistency to reduce the complexity}
+    \label{appsec:improvedComplexity}
 
@@
 
-    %\section{Exploiting the bipartite nature of base pair dependency
-    %  graphs to reduce the complexity}\label{sec:first-improvedComplexity}
+    %section{Exploiting the bipartite nature of base pair dependency
+    %  graphs to reduce the complexity}\label{appsec:first-improvedComplexity}
     %\textcolor{red}{\textbf{TODO: discuss and merge with/replace by previous Section}}
@@
 \section{Monotonicity of the partial derivatives within weight calibration}
-\label{sec:weight_derivatives}
+\label{appsec:weight_derivatives}
+
 %
@@
 \section{Approximate counting and random generation}
+\label{appsec:approximate-counting}
+
 In fact, not only is $\#{\sf BIS}$ a reference problem in counting complexity, but it is also a landmark problem with respect to the complexity of approximate counting problems.
@@
 \section{Contribution of terms to the MultiDefect score}
+\label{appsec:contributions-multidefect}
 
@@
 
-Figure~\ref{fig:contribtions-multidefect} breaks down the multi-defect
+Figure~\ref{appfig:contributions-multidefect} breaks down the multi-defect
 score of Eq.~$(\ref{eq:blueprintobjective})$, which is used to
@@
 
-\label{appsec:multidefect}
 \begin{figure}[h!]
@@
 \caption{Break-down of average MultiDefect values into their two components.}
-\label{fig:contribtions-multidefect}
+\label{appfig:contributions-multidefect}
 \end{figure}
 
-\section{Detailed results of MultiDefect analysis}\label{sec:validity}
+\section{Detailed results of the negative design benchmark}
+\label{appsec:detailed-benchmark-results}
+
 N/A values correspond to data that could not be obtained by the time of this submission for two main reasons: In the case of the initial sampling (left columns), they correspond to instances which, in conjunction with an expressive energy model, resulted in very high tree-width, leading to unreasonable memory requirements incompatible with our available computing resources; The case of missing values after optimization  (right columns), they indicate situations where the initial optimization took too long ($\approx$ 1 day) and was killed.
+
 \input{table}

### 2019-01-24 RNARedPrint.tex
add R-squared

@@
 pseudoknots (D\&P model) and energies in the stacking energy
-model. For the shown fits we sampled $n=1\,000$ sequences, targeting a
+model\revised{, which is obtained for each of the two target structures (with respective $R^2$ values of $0.846$ and $0.841$)}. For the shown fits we sampled $n=1\,000$ sequences, targeting a
 \GCb\% of $60\%$.
@@
     energies in the stacking model to the realistic pseudoknot energy
-    model by Dirks and Pierce (D\&P) for sampled sequences; the good
-    match enables more efficient targeting of Turner energies based on
+    model by Dirks and Pierce (D\&P) for the sampled sequences and both target structures R1 and R2. The good
+    match \revised{(respective $R^2$ values of $0.846$ and $0.841$)} enables more efficient targeting of Turner energies based on
     targeting stacking model energies.  We show the fits for both
@@
 ({\sf LE80}) to $16.05$ ({\sf 2str}) units. Remarkably, this
-improvement is be observed for both of the terms in \Obj{} (see
+improvement is observed for both of the terms in \Obj{} (see
 Supp. Mat.~\ref{appsec:contributions-multidefect}).  This means that \ourprog{}
@@
 %
-Moreover, for all instances as well, the superiority of Boltzmann designs
-remains unchallenged even after local optimizations, as shown by
+Moreover, for all instances as well, the Boltzmann designs
+remain superior even after local optimizations, as shown by
 Fig.~\ref{fig:benchmark-results} and Supp. Mat.~\ref{appsec:contributions-multidefect}
@@
 with multi-dimensional Boltzmann sampling over distributions
-controlled by expressive RNA energy models. Compared to the
-previously best available sampling method (uniform sampling), this
-approach generated significantly better seed sequences for instances
-of an extensive multi-target design benchmark, including pseudoknots.
+controlled by expressive RNA energy models.
+\revised{We demonstrate that the approach, despite of its theoretical hardness, performs well on typical multi-stable RNA design instances in practice, which is a direct consequence of the approximability of Turner energies as well as the systematic algorithmic framework.
+In a proof-of-concept study
+  on an established benchmark set for multi-target RNA design (including pseudoknotted instances), 
+  the approach generated significantly better seed sequences than
+the previously best available method (uniform sampling). These results strongly suggest that the presented technique for positive design can be highly beneficial in future negative design approaches.}
 %

### 2019-01-24 RNARedPrint.tex
SOme edits

--- 2019-01-24 RNARedPrint.tex
@@
   The design of multi-stable RNA molecules has important applications in biology, medicine, and biotechnology.
-Synthetic design approaches profit strongly from effective in-silico methods, \revised{which substantially reduce the need for costly wet-lab experiments.} \\
+Synthetic design approaches profit strongly from effective in-silico methods, which impact their cost and feasibility.
+\\
 %
   \parttitle{Results} We devise a novel approach to a central ingredient of most in-silico
-  design methods: the generation of sequences \revised{that fold well into multiple target structures.} 
-  % 
-  \revised{Based on constraint networks, our approach \ourprog{}} supports generic
+  design methods: the generation of sequences \revised{that fold well into multiple target structures.} For this task, we present the fixed-parameter tractable (FPT) approach \ourprog{}.
+  % 
+  \revised{Based on} constraint networks, \ourprog{} supports generic
   Boltzmann-weighted sampling, which enables the positive design of RNA sequences
@@
 %
-  Moreover, we study general properties of our approach empirically and generate
+  %Securing the theoretical foundations of our method
+  Moreover, we establish the $\#${\sf P}-hardness of \revised{counting to theoretically justify the development of an FPT approach}. 
+%
+  Finally, we study general properties of our approach empirically and generate
   biologically relevant multi-target Boltzmann-weighted designs for a
-  \revised{established} design benchmark. \revised{Our results demonstrate the effectivity and feasiblity of the method in practice
-  as well as the benefits of Boltzmann sampling over the previously best multi-target sampling strategy---even for the case of negative design of multi-stable RNAs.}\\
-%
-  \revised{Besides empirically studies, we finally justify the algorithmic details due to a fundamental theoretic result about multi-stable RNA design, namely the \#P-hardness of the counting of designs.}
-%
-  \parttitle{Conclusion} \ourprog{} introduces a novel, flexible and---as we demonstrate---effective approach to multi-target RNA design, which promises broad applicability and extensibility.
+  common design benchmark.
+  We demonstrate significant improvements over the previously best multi-target sampling strategy, due to generating seed sequences with \ourprog{}.\\
+%
+  \parttitle{Conclusion} \ourprog{} introduces a novel, flexible and---as we demonstrate--effective approach to multi-target RNA design, which promises broad applicability and extensibility.
 
@@
 coloring the sequence positions with nucleotides, such that the ends of each base pair receive compatible colors.
-Initially, the method decomposes the graph hierarchically and then \Def{precomputes} the number of valid sequences
+Initially the method decomposes the graph hierarchically, using ear decomposition, and then \Def{precomputes} the number of valid sequences
 within each subgraph. The decomposition is then reinterpreted as a
@@
 \revised{Note that for simplicity, we assume in this figure that only dependencies between the ends of base pairs are considered to evaluate the energies of structures.
-Our computations based on a more complex energy model, which considers energy contributions of base pair stacks, require additional dependencies.}
+Our computations based on a more complex energy model (stacking model) require additional dependencies.}
   }
@@
 
-\revised{As central contribution, we provide a systematic and flexibly extensible technique for sampling that targets multiple versatile features.
-For the sake of clarity, we introduce this method specialized to the sampling of RNA sequences that target the energies of multiple structures and the \GCb\%.}
-
-\revised{Figure~\ref{fig:workflow} summarizes our generic framework, which enables this targeted sequence generation based on multi-dimensional Boltzmann sampling.
-\emph{Algorithmically}, we originally contribute} dynamic programming (DP) algorithms, based on the concept of \Def{tree decomposition}, to compute partition functions and sample sequences from the Boltzmann distribution%
+\revised{As our central contribution, we provide a systematic and flexibly extensible technique for sampling that targets multiple versatile features.
+For the sake of clarity, we introduce this method specialized to the sampling of sequences that target the energies of multiple structures and the \GCb\%.}
+
+\revised{Fig.~\ref{fig:workflow} summarizes our generic framework, which enables} efficient Boltzmann-weighted sampling over RNA sequences with multiple target structures.
+\revised{\emph{Algorithmically}, we contribute} dynamic programming (DP) algorithms, based on the concept of \Def{tree decomposition}, to compute partition functions and sample sequences from the Boltzmann distribution%
 % (Subsection~\ref{sec:PF})
@@
 %
-\revised{Building on this principle, the complexities of our algorithms depend exponentially on a specific property of the tree decomposition, called the \Def{treewidth}.
-Thus, it is essential for the applicability of our approach that---by appropriate design choices---we can keep this parameter low for typical instances.
-For} any fixed value of the treewidth, the complexity scales only linearly with the size of designed sequences and the number of targeted structures, \emph{i.e.} our algorithms are \emph{fixed-parameter tractable (FPT)}.
-
-\revised{Remarkably, we could show that it is not possible to find a better, efficient method for sampling (unless ${\sf P}={\sf NP}$), since the underlying counting problem is \#{\sf P}-hard.
-The practical relevance of this theoretical result is that it rules out substantially better sampling techniques, such that there will always---even using improved sampling methods---remain an upper limit on the---in practice---tractable number and heterogeneity of structures, the complexity of the directly treatable energy model, and the number and complexity of additional constraints that could be considered in future sampling-based applications.} Technically, this result relies on a surprising bijection between valid sequences and independent sets of a bipartite graph, the latter being the object of recent breakthroughs in approximate counting complexity~\citep{Bulatov2013,Cai2016}.
+\revised{Building on this principle, }the complexities of our algorithms depend exponentially on a specific property of the tree decomposition, called the \Def{treewidth}, which remains low for typical instances.
+However, for any given value of the treewidth, the complexity scales only linearly with the size of designed sequences and the number of targeted structures, \emph{i.e.} our algorithms are fixed-parameter tractable (FPT).
+
+\revised{Remarkably, we show that it is not possible to find a better, efficient method for sampling (unless ${\sf P}={\sf NP}$), since the underlying counting problem is \#{\sf P}-hard.
+The practical relevance of this result is that it rules out substantially better sampling techniques, such that there will always remain---in practice---an upper limit on the tractable number of structures and their heterogeneity, the complexity of the directly treatable energy model, and the number and complexity of additional constraints that could be considered in future sampling-based applications.} Technically, this result relies on a surprising bijection between valid sequences and independent sets of a bipartite graph, the latter being the object of recent breakthroughs in approximate counting complexity~\citep{Bulatov2013,Cai2016}.
 
@@
 %
-Due to the generality of our method, we can moreover strongly limit \revised{the treewidth} in practice by using state-of-the-art tree decomposition algorithms.
+Due to the generality of our method, we can moreover strongly limit this parameter in practice by using state-of-the-art tree decomposition algorithms.
 %Uniform sampling is handled as a special case of Boltzmann sampling, where each valid sequence receives energy zero and---consequently---computing partition functions specializes to counting.
-By evaluating sequences in a specialized weighted constraint
-network, we support---in principle---arbitrary multi-ary constraints and thus
+By evaluating (partial) sequences in a specialized weighted constraint
+network, we support arbitrary multi-ary constraints and thus
 arbitrarily complex energy models,
@@
 Turner RNA energy model
-works well by performing sampling based on \revised{a simplified RNA energy model, which induces much lower treewidth than the Turner model}.
-This result is \revised{essential for the applicability of our method, since it allows to combine high efficiency (by keeping the treewidth low) with sufficient accuracy to precisely target realistic Turner energies.}
+works well by performing sampling based on \revised{a simplified RNA energy model, which induces lower treewidth}.
+This result is highly relevant in practice, since it allows to combine high efficiency with high accuracy (and even leaves room for \revised{future extensions supporting additional constraints}).
 
@@
  on a comprehensive multi-target RNA design benchmark~\citep{Taneda2015} 
-  suggest that our sampling strategy well supports designing biologically relevant
+  suggest that our sampling strategy supports designing biologically relevant
 RNAs for multiple targets.}
@@
 \section*{Methods}
-\revised{The main computational problem addressed in this work is the positive design of RNA sequences for multiple target structures; more specifically, the generation of sequences over the alphabet $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}$, such that the sequences feature a given $\GCb\%$, and have prescribed energies for a set of target secondary structures.
-Here, these desired sequence properties are modeled as constraints on the values of \Def{features}, which are functions of the sequence that are expressed as sums over real-valued \Def{contributions}.
-Each contribution depends on the nucleotides at---typically few---specific sequence positions.}
-
-\revised{To generate diverse design candidates, we randomly generate sequences from a Boltzmann distribution.
-The probability of a sequence then depends on its features (e.g.
-the energies of the target structures), and the weight of each feature (which influences its distribution).
-Sampling from the (multi-feature) Boltzmann distribution requires to compute corresponding partition functions, such that we can draw sequences with probabilities proportional to their Boltzmann weight.
-On this basis, we can finely calibrate the weights, to maximize the probability that sampled sequences meet the desired target values for each feature.
-Together with a final rejection step this results in an effective procedure for generating highly specific sequences.}
+The main computational problem addressed in this work is the generation of sequences over the alphabet $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}$, \revised{such that the sequences feature a given $\GCb\%$, and have prescribed energies for a set of target secondary structures.
+Here, such criteria are modeled} as constraints on the value of \Def{features}, functions of the sequence that are expressed as sums over real-valued contributions.
+Such contributions typically depend on the nucleotides assigned to a subset of sequence positions.
+
+To achieve maximal diversity within candidate designs, we abstract the problem as a random generation of sequences within a Boltzmann distribution.
+The probability of a sequence then depends on its associated features, and of weights that are associated features.
+Those influence the expected feature values of generated sequences, allowing us to shift the sample distributions towards specific targets.
+Sampling from this distribution requires to compute corresponding partition functions, such that we can draw sequences with probabilities proportional to their Boltzmann weight.
+A final rejection step, made efficient by the fine calibration of weights, allows to produce sequences satisfying the input set of constraints.
 
@@
 \label{sec:problem-statement}
-
 %\paragraph{Central computational problem.}
-Let us consider $m$ (secondary) structures $R_1,\dots, R_m$, each abstracted as a set of base pairs, and \revised{$k\geq m$} features $F_1,\dots, F_k$, associated with weights $\pi_1,\dots,\pi_k$.
-Our goal is to sample from the Boltzmann distribution, defined over the set of sequences that satisfy classic base pairing rules for each structure as:
+Let us consider $m$ (secondary) structures $R_1,\dots, R_m$, each abstracted as a set of base pairs, and $k$ features $F_1,\dots, F_k$, associated with weights $\pi_1,\dots,\pi_k$, our goal is to sample from the Boltzmann distribution, defined over the set of sequences that satisfy classic base pairing rules for each structure as:
 \begin{equation}
@@
 
-The workhorse of our approach is the \revised{fixed-parameter tractable} computation of feature-dependent partition functions over sequences:
+The workhorse of our approach is the efficient (more precisely, fixed-parameter tractable) computation of feature-dependent partition functions over sequences:
 %Given are the features $F_1,\dots,F_k$, which encode the energies functions for the fixed structures and the function for the \GCb\%, all in terms of simple atomic energy contributions.
@@
 
-The $\GCb\%$ can be simply expressed using $n$ contributions $f^{\GCb}_i$, $i\in[1,n]$, whose dependencies are restricted to individual nucleotides,  and such that 
+The $\GCb\%$ can be trivially expressed using $n$ contributions $f^{\GCb}_i$, $i\in[1,n]$, whose dependencies are restricted to individual nucleotides,  and such that 
 $$f^{\GCb}_i(\{i\mapsto c\})=\begin{cases}-1&\text{ if }c=\Gb\text{ or }\Cb\\ 0&\text{ otherwise}.\end{cases}$$ 
-\revised{By summing $f^{\GCb}_i(\{i\mapsto S_i\})$ over the whole sequence, one simply counts the occurrences of $\Gb$ and $\Cb$.}
+\revised{By summing $f^{\GCb}_i$ over the whole sequence, one simply counts the occurrences of $\Gb$ and $\Cb$.}
 
 \revised{
-  To start with a simple example of evaluating the energy of sequences by features, let us explain how they are used to count the number of \Def{valid sequences}, \emph{i.e.} sequences inducing only base pairs in $\mathcal{B}:=\{\{\Ab,\Ub\}, \{\Gb,\Cb\}, \{\Gb,\Ub\}\}$.
-Consider a feature $F^{\tt BP}$ composed of  contributions $f^{\tt BP}_{i,j}$, for each base pair $(i,j)$ occurring in some structure, such that
+	Such features can also be used to count the number of \Def{valid sequences}, \emph{i.e.} sequences inducing only base pairs in $\mathcal{B}:=\{\{\Ab,\Ub\}, \{\Gb,\Cb\}, \{\Gb,\Ub\}\}$.
+Indeed, consider a feature $F^{\tt BP}$ composed of  contributions $f^{\tt BP}_{i,j}$, for each base pair $(i,j)$ occurring in some structure, such that
 	$$f^{{\sf BP}}_{i,j}(\left\{\substack{i\mapsto a\\j\mapsto b}\right\})=\begin{cases}0&\text{ if }\{a,b\}\in\mathcal{B} \\+\infty&\text{ otherwise}.\end{cases}$$ 
 	The value of $F^{\tt BP}$ is $0$ for any valid sequence, and $+\infty$ as soon as some non canonical base pair is created.
-For any associated weight $\pi_{{\sf BP}}>1$, the contribution of a valid sequence is $\pi_{{\sf BP}}^0 = 1$, and the contribution of an invalid sequence is $\pi_{{\sf BP}}^{+\infty} = 0$, so that Eq.~\ref{eq:mainproblem} (when restricted to $F^{\tt BP}) $simply counts the number of valid sequences.
-
-        Energy models for structure prediction vary considerably, yet can always be expressed as sums over contributions associated with local structural motifs (base pairs, base pair stacks, loops\ldots) under a certain nucleotide assignment.
+For any associated weight $\pi_{{\sf BP}}>1$, the contribution of a valid sequence is $\pi_{{\sf BP}}^0 = 1$, and the contribution of an invalid sequence is $\pi_{{\sf BP}}^{+\infty} = 0$, so that Equation~\ref{eq:mainproblem} simply counts the number of valid sequences when restricted to $F^{\tt BP}$.
+
+	Energy models for structure prediction vary considerably, yet can always be expressed as sums over contributions associated with local structural motifs (base pairs, stacking pairs, loops\ldots) under a certain nucleotide assignment.
 Energy models can thus be captured generically by introducing, for each motif $m$ occurring in a target structure, a contribution $f_m$, taking a specific value for each assignment of nucleotides to its positions $\dep(f_m)$.
-For instance, the contribution of a \Def{base pair stack}, consisting of two pairs $(i,j)$ and $(i+1,j-1)$, can be captured by the introduction of a function $f^{\textsf{Stack}}_{\substack{i,j}}$ such that $\dep(f^{\textsf{Stack}}_{\substack{i,j}}) = \{i,i+1,j-1,j\}$.
-        We refer to energy models that consider the contributions of all base pair stacks (and thus introduce the corresponding dependencies) collectively as the \Def{stacking energy model}.
+For instance, the contribution of a stacking pair, consisting of two pairs $(i,j)$ and $(i+1,j-1)$, can be captured by the introduction of a function $f^{\textsf{Stack}}_{\substack{i,j}}$ such that $\dep(f^{\textsf{Stack}}_{\substack{i,j}}) = \{i,i+1,j-1,j\}$.
+
 }
@@
 
-\subsection*{Dependency (hyper)graph, tree decomposition and treewidth}
-
+\subsection*{Dependency (hyper)graph, tree decomposition and width}
 %The complexity of the central problem depends fundamentally on the dependency structure that is imposed by the entire set of contributions.
@@
 
-\revised{In order to compute the partition function of Eq.~\ref{eq:mainproblem}, and thus sample in a well-defined way, one must consider dependencies induced by the complete set of contributions $$\F := \bigcup_{\ell}\{f\mid f\text{ contribution of } F_\ell\}.$$}
-
+\revised{In order to compute the partition function of Eq.~\ref{eq:mainproblem}, and thus sample in a well-defined distribution, one needs to abstract dependencies beyond pairs of positions.
+To that purpose, we consider the notion of \Def{hypergraph}, which consists of vertices (here, the sequence positions) connected by \Def{hyperedges} defined as sets of vertices.
+Undirected graphs can be modeled as hypergraphs whose hyperedges are pairs of vertices, but hypergraphs also allow to express complex dependencies.
+
+Here, we consider the complete set of contributions $$\F := \bigcup_{\ell}\{f\mid f\text{ contribution of } F_\ell\}.$$
 %
@@
 %
-
-\revised{
-In the simplest case, this set captures the requirement of canonical base pairing for each structure.
-To express this, let us define the \Def{base pair dependency graph} $G_{\R}$ as the graph with nodes $\{1,\dots,n\}$
-and edges $\bigcup_{\ell\in[1,k]} R_\ell$.
-%Any edge between two nodes $i$ and $j$ materializes the existence of a base pair between positions $i$ and $j$ in some structure, inducing a direct dependency between the chosen nucleotide assignments for the two positions.
-
-Since $\F$ defines potentially more complex dependencies, which can
-relate more than two positions, in general its dependencies cannot be 
-represented by a graph. Instead, this requires a structure known as
-\Def{hypergraph}, which consists of vertices (here, the sequence
-positions) connected by \Def{hyperedges}, which are arbitrary sets of
-vertices. In this way, hypergraphs generalize undirected graphs, where
-each edge is a set of exactly two vertices.
-%
-The \Def{dependency (hyper)graph induced by $\F$} is then defined as the hypergraph $G_\F=(V, H)$ on sequence positions $V=\{1,\dots,n\}$
-by interpreting the dependencies as hyperedges, i.e. $H =
-\{\dep(f)\mid f\in \F\})$.
-
-Let us finally define the \Def{tree decomposition} of the graph $G_\F$, a fundamental ingredient of our algorithms, which also determines their efficiency (most importantly, via its property called treewidth).
+The set $\F$ of contributions induces a
+\Def{dependency (hyper)graph} $G_\F$ on sequence positions by interpreting the dependencies as hyperedges, leading to an hypergraph 
+$G_\F=(V, H)$ having vertices $V=\{1,\dots,n\}$ and hyperedges $H = \{\dep(f)\mid f\in \F\})$.
+
+Irrespectively of the chosen energy model, to capture the requirements of canonical base pairing for each structure,
+we also define the \Def{base pair dependency graph} $G_{\R} = (V,E)$ having vertices $V=\{1,\dots,n\}$ 
+and edges $E=\bigcup_{\ell\in[1,k]} R_\ell$.
+Any edge between two nodes $i$ and $j$ materializes the existence of a base pair between positions $i$ and $j$ in at least one of the input structures, and induce a direct dependency between the nucleotide assignments for those positions.
 }
-\begin{definition}[Tree decomposition and treewidth]
+
+Let us now remind the \Def{tree decomposition} of an (hyper)graph $G_\F$, a fundamental ingredient of our algorithms.
+
+\begin{definition}[Tree decomposition]
   \label{def:treedecomp}
@@
 \end{enumerate}
-
-\smallskip
-\noindent
-The \Def{treewidth} of a tree decomposition $(T,\chi)$ is defined as
-$\min_{u\in T} |\chi(u)| - 1.$ 
-%The \Def{treewidth} of a graph $G$ is the smallest treewidth of any tree decomposition of $G$.
 \end{definition}
-%
-\revised{Intuitively, a tree decomposition of an (hyper)graph $G$ is a tree that captures all the vertices and (hyper)edges of $G$, and properly relates dependend subproblems to ensure consistency in a recursive computation.
+\revised{Intuitively, a tree decomposition for an (hyper)graph $G$ is a tree structure which features all the vertices of $G$, captures all the (hyper)edges of $G$, and helps ensure some notion of consistency by allowing the propagation of values.
 Figure~\ref{fig:treedec} shows an optimal tree decomposition for a pair of structures under the stacking energy model.}
+\begin{definition}[Tree width]
+The \Def{width} of a tree decomposition $(T,\chi)$ is 
+$$\min_{u\in T} |\chi(u)| - 1.$$ The \Def{treewidth}
+of an (hyper)graph $G$ is the smallest width of a tree decomposition of $G$.
+\end{definition}
+%Recall the essential role of the treewidth as fixed parameter in our FPT appraoch.
 
@@
 {\centering \includegraphics[width=.9\linewidth]{Figs/example-treedec-stacking}\\}
-\caption{\revised{Toy example of a tree decomposition associated with two target structures in the stacking energy model (where the four positions of each base pair stack depend on each other).
-Two target secondary structures ({\bfseries\sffamily A}) are merged into a joint hypergraph ({\bfseries\sffamily B}), whose hyperedges correspond to the quadruplets of positions involved in base pair stacks (colored).
-A valid tree decomposition ({\bfseries\sffamily C}) for the hypergraph ensures, among other properties, that each base pair and each base pair stack is represented in at least one of its node, so that features can be correctly evaluated.
-The treewidth of this tree decomposition is 3, a provably optimal value for this input hypergraph.}\label{fig:treedec}}
+\caption{\revised{Toy example of a tree decomposition associated with two target structures in the stacking energy model.
+Two target secondary structures ({\bfseries\sffamily A}) are merged into a joint hypergraph ({\bfseries\sffamily B}), whose hyperedges correspond to the quadruplets of positions involved in stacking pairs (colored).
+A valid tree decomposition ({\bfseries\sffamily C}) for the hypergraph ensures, among other properties, that each base pair and each stacking pair/hyperedge is represented in at least one of its node, so that features can be adequately evaluated.
+The width of this tree decomposition is 3, a provably optimal value for this input hypergraph.}\label{fig:treedec}}
 \end{figure}
@@
 the correctness of the algorithm
-(Supp.\ Mat.~\ref{appsec:correctness}).  After running
+(Supp. Mat.~\ref{appsec:correctness}).  After running
 Alg.~\ref{alg:pf}, multiplying the 0-ary messages sent to the root $r$
@@
 $\mathcal{O}((|\F|+|V|)\cdot 4^{w+1})$ time and
-$\mathcal{O}(|V|\cdot4^s)$ space; Alg.~\ref{alg:sampling} would
+$\mathcal{O}(|V|\cdot4^s)$ space
+(Supp. Mat.~\ref{appsec:algcomplexity}); Alg.~\ref{alg:sampling} would
 require $\mathcal{O}((|\F|+|V|)\cdot 4^D)$ per sample on arbitrary
 tree decompositions
-(Supp.\ Mat.~\ref{appsec:algcomplexity}). W.l.o.g.\ we assume that
+(Supp. Mat.~\ref{appsec:algcomplexity}). W.l.o.g.\ we assume that
 $D=1$; note that tree decompositions can generally be transformed,
@@
 
-By this theorem, the complexity is polynomial for fixed value of $w$, and Boltzmann sampling in our setting is thus fixed parameter tractable (FPT) in the treewidth.
-%
-The complexity of the precomputation can be further improved to
-$\mathcal{O}(n\,k\,2^{w+1}\,2^{c})$, where $c$ ($c\le w+1$) is the maximum number of connected components represented in a node of the tree decomposition (Supp.\ Mat.~\ref{appsec:improvedComplexity}).
+By this theorem, the complexity is polynomial for fixed value of $w$, and Boltzmann sampling in our setting is thus fixed parameter tractable (FPT) in the tree width.
+%
+As shown in Supp. Mat.~\ref{sec:improvedComplexity}, the complexity of the precomputation can be further improved to
+$\mathcal{O}(n\,k\,2^{w+1}\,2^{c})$, where $c$ is the maximum number of connected components represented in a node of the tree decomposition ($c\le w+1$).
 
@@
 negligible compared to Alg.~\ref{alg:pf}
-(Supp.\ Mat.~\ref{appsec:treedecomp} and
+(Supp. Mat.~\ref{appsec:treedecomp} and
 \ref{appsec:dependency-cliques}).
@@
 In order to capture realistic energy models like the Turner model or pseudoknot models like {\tt HotKnots}~\citep{Ren2005}, our sampling strategy can be extended in two ways: 1) either by directly sampling based on more expressive energy models; 2) or by sampling in a simple energy model can be used to approximate sampling in more complex models.
-In practice, complex energy models has a strong influence on the treewidth (of optimal tree decompositions) of the dependency graph and, in turn the computational complexity of our approach.
+In practice, complex energy models has a strong influence on the tree width of the dependency graph and, in turn computational complexity of our approach.
 Therefore, it is interesting to consider, in addition to the base pair model, other
@@
 $\EnergyTurner$ and the fitted stacking model $\EnergyStacking$
-(Supp.\ Mat.~\ref{appsec:modelparameters}). More precisely, we observed a
+(Supp. Fig~\ref{fig:training-cor}). More precisely, we observed a
 structure-specific affine dependency between the Turner and stacking
@@
 In other words,  $\mathbb{E}(F_\ell(S)\mid \pmb{\pi})=\TargetE_\ell$,  $\forall\ell\in[1,k]$ and, analogously, the expectation of $F_0(S)$ is the targeted GC content.
-The expected value of $F_\ell$ is always decreasing for increasing weights $\pi_\ell$ (see Supp.\ Mat.~\ref{appsec:weight_derivatives}).
+The expected value of $F_\ell$ is always decreasing for increasing weights $\pi_\ell$ (see Supp.
+Mat.~\ref{sec:weight_derivatives}).
 More generally, computing a suitable parameter vector $\pmb{\pi}$ can be restated as a convex optimization problem, and be efficiently solved using a wide array of methods~\citep{Denise2010,Bendkowski2017}.
@@
 %We define the following notations: $\real^\infty:=\real\cup\{\infty\}$, $[i,j]:=\{i,\ldots,j\},$ \ldots
-\revised{While efficient, both in practice and in theory for graphs of bounded treewidth, our algorithms remain exponential in the worst case scenario, since the treewidth of a dependency graph can then become arbitrarily large.
+\revised{While efficient, both in practice and in theory for graphs of bounded tree width, our algorithms remain exponential in the worst case scenario, since the tree-width of a dependency graph can then become arbitrarily large.
 This exponential complexity in the worst case appears to be intrinsic.
@@
 \subsection*{A parsimonious stacking energy model as a proxy for the Turner energy model}
+
 Sampling for fixed weight
@@
 Figure 10 - Correlation plot -- Right
-Figure 2 -- Distribution plot -- Right (Turner is even untractable to compute the treewidth itself in benchmark instances)
+Figure 2 -- Distribution plot -- Right (Turner is even untractable to compute the tree width itself in benchmark instances)
 Figure 3 -- Right -- Distribution
@@
 for its targeting performance (both models:
-Supp.\ Mat.~\ref{appsec:modelparameters}).
+Supp. Mat.~\ref{appsec:modelparameters}).
 %
@@
 to a large extent, independently shift the Turner energies of sampled
-sequences towards prescribed targets. See Supp.\ Mat.~\ref{appsec:illustrating-mdbs} for a further example with three
+sequences towards prescribed targets. See Supp. Mat.,
+Fig.~\ref{fig:energydist} for a further example with three
 pseudoknot-free target structures.}
@@
 improvement is be observed for both of the terms in \Obj{} (see
-Supp. Mat.~\ref{appsec:contributions-multidefect}).  This means that \ourprog{}
+Supp. Mat.~\ref{sec:improvements}).  This means that \ourprog{}
 produces sequences whose targets are both substantially closer to
@@
 structures, we observed better \Obj{} for Boltzmann sampling than for
-uniform sampling (see Supp. Mat.~\ref{appsec:detailed-benchmark-results}).  Notably,
+uniform sampling (see Supp. Mat.~\ref{sec:validity}).  Notably,
 \ourprog{} performs equally well in the presence of pseudoknots;
@@
 remains unchallenged even after local optimizations, as shown by
-Fig.~\ref{fig:benchmark-results} and Supp. Mat.~\ref{appsec:contributions-multidefect}
-and \ref{appsec:detailed-benchmark-results}. This observation is consistent with a
+Fig.~\ref{fig:benchmark-results} and Supp. Mat.~\ref{sec:improvements}
+and \ref{sec:validity}. This observation is consistent with a
 superior quality of the starting point for the greedy walk, probably leading

### 2019-01-24 RNARedPrint.tex
fig3 caption clearer

--- 2019-01-24 RNARedPrint.tex
@@
     energies in the stacking model to the realistic pseudoknot energy
-    model by Dirks and Pierce (D\&P) for the sampled sequences and both target structures R1 and R2. The good
+    model by Dirks and Pierce (D\&P) for \revised{initially} sampled sequences and both target structures R1 and R2 (shown in subfigure B). The good
     match \revised{(respective $R^2$ values of $0.846$ and $0.841$)} enables more efficient targeting of Turner energies based on
-    targeting stacking model energies.  We show the fits for both
-    target structures R1 and R2 of subfigure B.  \textbf{(B)} D\&P energy
-    distributions for two target structures R1 and R2. We show the
-    results of targeting the respective free energies $(-30,-20)$,
-    $(-30,-30)$, $(-25,-25)$, and $(-20,-30)$ for the target
-    structures---this demonstrates the effectivity of our adaptive
-    multi-dimensional Boltzmann sampling procedure; moreover, for
-    comparison, the distributions for uniform and Boltzmann
-    samples---respectively associated with homogenous weights $1$ and
-    $e^\beta$.
+    targeting stacking model energies. \textbf{(B)} Resulting D\&P energy distributions for the two target structures R1 and R2 when aiming for the respective free energies $(-30,-20)$,
+    $(-30,-30)$, $(-25,-25)$, and $(-20,-30)$. These demonstrate the effectivity of our adaptive
+    multi-dimensional Boltzmann sampling procedure, especially by comparing the distributions to those of uniform and Boltzmann sampled sequences, with homogenous weights $1$ and
+    $e^\beta$, respectively.
 %    ($\delta=0.05, \delta'=0.1, \gamma=1.1, \psi=20, n=1000$)

### 2019-01-24 RNARedPrint.tex
Updates from ShareLaTeX
Questions for sebastian in todo

@@
 Our goal is to sample from the Boltzmann distribution, defined over the set of sequences that satisfy classic base pairing rules for each structure as:
+\todo{$S$ sequence not defined!}
 \begin{equation}
@@
 %Given are the features $F_1,\dots,F_k$, which encode the energies functions for the fixed structures and the function for the \GCb\%, all in terms of simple atomic energy contributions.
+
 For weights $\pi_1,\dots,\pi_k$, one wants to compute the partition function
@@
 \subsection*{Expressing \GCb\%-content, sequence validity and energies as features}
+\todo{$x_1$ position(?) not defined!\\$S_x_1$ nucleotide at position(?) not defined!\\$p$ number of dependent positions(?) from $x_p$ not defined!}
 Formally, we define a \Def{feature $F$} as a function on sequences, whose value is obtained by summing over an associated set of \Def{contributions}.
@@
 $$f^{\GCb}_i(\{i\mapsto c\})=\begin{cases}-1&\text{ if }c=\Gb\text{ or }\Cb\\ 0&\text{ otherwise}.\end{cases}$$ 
+\todo{Why here $i$ and not $x_i$??}
 \revised{By summing $f^{\GCb}_i(\{i\mapsto S_i\})$ over the whole sequence, one simply counts the occurrences of $\Gb$ and $\Cb$.}
@@
         The value of $F^{\tt BP}$ is $0$ for any valid sequence, and $+\infty$ as soon as some non canonical base pair is created.
-For any associated weight $\pi_{{\sf BP}}>1$, the contribution of a valid sequence is $\pi_{{\sf BP}}^0 = 1$, and the contribution of an invalid sequence is $\pi_{{\sf BP}}^{+\infty} = 0$, so that Eq.~\ref{eq:mainproblem} (when restricted to $F^{\tt BP}) $simply counts the number of valid sequences.
+For any associated weight $\pi_{{\sf BP}}>1$, the contribution of a valid sequence is $\pi_{{\sf BP}}^0 = 1$, and the contribution of an invalid sequence is $\pi_{{\sf BP}}^{+\infty} = 0$, so that Eq.~\ref{eq:mainproblem} (when restricted to $F^{\tt BP})$ simply counts the number of valid sequences.
 
@@
 pseudoknots (D\&P model) and energies in the stacking energy
-model\revised{, which is obtained for each of the two target structures (with respective $R^2$ values of $0.846$ and $0.841$)}. For the shown fits we sampled $n=1\,000$ sequences, targeting a
+model\revised{, which is obtained for each of the two target structures (with respective $R^2$ values of $0.846$ and $0.841$)}. For the shown fits we sampled $n=10\,000$ sequences, targeting a
 \GCb\% of $60\%$.

### 2019-01-24 RNARedPrint.tex
Changes due to Maria's comments. add 'use case' idea to conclusion

--- 2019-01-24 RNARedPrint.tex
@@
 design~\citep{Wachsmuth2013,Domin2017}, partly motivated by their
-capacity to act as biosensors~\citep{Findeiss2017}.
+capacity to act as biosensors~\citep{Findeiss2017}, \revised{which suggests them for biotechnological applications.}
 In particular due to the kinetic coupling of RNA folding with RNA transcription, RNA families can feature alternative,
@@
 target structures. \Software{RNAdesign} adopts a graph coloring perspective,
-coloring the sequence positions with nucleotides, such that the ends of each base pair receive compatible colors.
-Initially, the method decomposes the graph hierarchically and then \Def{precomputes} the number of valid sequences
+\revised{assigning nucleotide symbols (like ``colors'') to the sequence positions, such that compatible nucleotides are assigned to the ends of each base pair.} Initially, the method decomposes the graph hierarchically and then \Def{precomputes} the number of valid sequences
 within each subgraph. The decomposition is then reinterpreted as a
@@
 As well, the \Software{RNAdesign}/\RNAblueprint{} approach is specialized to uniform sampling, which limits its direct extensibility.
-Both aspects call for a more systematic redesign of multi-target sampling, which builds on established concepts in computer science.
-In turn, this promises a fundamentally broader range of applications for extensions of the sampling method.
+\revised{Substantial improvements of multi-target sampling thus require a systematically redesigned appraoch.
+To enable a fundamentally broader range of applications in extensions of the sampling method, we build our approach, from the start, on established concepts in computer science.}
 
@@
 \revised{As central contribution, we provide a systematic and flexibly extensible technique for sampling that targets multiple versatile features.
-For the sake of clarity, we introduce this method specialized to the sampling of RNA sequences that target the energies of multiple structures and the \GCb\%.}
+For the sake of clarity, we introduce this method specialized to the sampling of RNA sequences that have specific energies for multiple structures and specific \GCb\%.}
 
@@
 .
-\revised{Generally, tree decompositions are data structures that capture the specific dependencies of a problem instance (here, the dependencies induced by the target structures) to guide the efficient processing by DP algorithms.}
+\revised{Generally, tree decompositions are data structures that capture the specific dependencies of a problem instance (here, the dependencies between sequence positions induced by the target structures), such that they can guide the efficient processing by DP algorithms.}
 %
@@
 By evaluating sequences in a specialized weighted constraint
-network, we support---in principle---arbitrary multi-ary constraints and thus
-arbitrarily complex energy models,
-notably subsuming all commonly
+network, we support---in principle---arbitrary \revised{complex constraints and energy models},
+notably subsuming the commonly
 used RNA energy models%
@@
     energies in the stacking model to the realistic pseudoknot energy
-    model by Dirks and Pierce (D\&P) for \revised{initially} sampled sequences and both target structures R1 and R2 (shown in subfigure B). The good
+    model by Dirks and Pierce (D\&P) for the sampled sequences and both target structures R1 and R2. The good
     match \revised{(respective $R^2$ values of $0.846$ and $0.841$)} enables more efficient targeting of Turner energies based on
-    targeting stacking model energies. \textbf{(B)} Resulting D\&P energy distributions for the two target structures R1 and R2 when aiming for the respective free energies $(-30,-20)$,
-    $(-30,-30)$, $(-25,-25)$, and $(-20,-30)$. These demonstrate the effectivity of our adaptive
-    multi-dimensional Boltzmann sampling procedure, especially by comparing the distributions to those of uniform and Boltzmann sampled sequences, with homogenous weights $1$ and
-    $e^\beta$, respectively.
+    targeting stacking model energies.  We show the fits for both
+    target structures R1 and R2 of subfigure B.  \textbf{(B)} D\&P energy
+    distributions for two target structures R1 and R2. We show the
+    results of targeting the respective free energies $(-30,-20)$,
+    $(-30,-30)$, $(-25,-25)$, and $(-20,-30)$ for the target
+    structures---this demonstrates the effectivity of our adaptive
+    multi-dimensional Boltzmann sampling procedure; moreover, for
+    comparison, the distributions for uniform and Boltzmann
+    samples---respectively associated with homogenous weights $1$ and
+    $e^\beta$.
 %    ($\delta=0.05, \delta'=0.1, \gamma=1.1, \psi=20, n=1000$)
@@
 \revised{We demonstrate that the approach, despite of its theoretical hardness, performs well on typical multi-stable RNA design instances in practice, which is a direct consequence of the approximability of Turner energies as well as the systematic algorithmic framework.
-In a proof-of-concept study
-  on an established benchmark set for multi-target RNA design (including pseudoknotted instances), 
+By conducting a proof-of-concept study
+  on an established benchmark set for negative multi-target RNA design (including pseudoknotted instances), we showcase a typical application of our tool \ourprog{}. In this study, 
   the approach generated significantly better seed sequences than
-the previously best available method (uniform sampling). These results strongly suggest that the presented technique for positive design can be highly beneficial in future negative design approaches.}
+the previously best available method (uniform sampling). These results strongly suggest that the presented technique for positive design can be highly beneficial in future negative design approaches.
+}
 %

### 2019-01-24 RNARedPrint.tex
Addded paragraph

--- 2019-01-24 RNARedPrint.tex
@@
 Our goal is to sample from the Boltzmann distribution, defined over the set of sequences that satisfy classic base pairing rules for each structure as:
-\todo{$S$ sequence not defined!}
 \begin{equation}
@@
 %Given are the features $F_1,\dots,F_k$, which encode the energies functions for the fixed structures and the function for the \GCb\%, all in terms of simple atomic energy contributions.
-
 For weights $\pi_1,\dots,\pi_k$, one wants to compute the partition function
@@
 \subsection*{Expressing \GCb\%-content, sequence validity and energies as features}
-\todo{$x_1$ position(?) not defined!\\$S_x_1$ nucleotide at position(?) not defined!\\$p$ number of dependent positions(?) from $x_p$ not defined!}
 Formally, we define a \Def{feature $F$} as a function on sequences, whose value is obtained by summing over an associated set of \Def{contributions}.
@@
 $$f^{\GCb}_i(\{i\mapsto c\})=\begin{cases}-1&\text{ if }c=\Gb\text{ or }\Cb\\ 0&\text{ otherwise}.\end{cases}$$ 
-\todo{Why here $i$ and not $x_i$??}
 \revised{By summing $f^{\GCb}_i(\{i\mapsto S_i\})$ over the whole sequence, one simply counts the occurrences of $\Gb$ and $\Cb$.}
@@
         The value of $F^{\tt BP}$ is $0$ for any valid sequence, and $+\infty$ as soon as some non canonical base pair is created.
-For any associated weight $\pi_{{\sf BP}}>1$, the contribution of a valid sequence is $\pi_{{\sf BP}}^0 = 1$, and the contribution of an invalid sequence is $\pi_{{\sf BP}}^{+\infty} = 0$, so that Eq.~\ref{eq:mainproblem} (when restricted to $F^{\tt BP})$ simply counts the number of valid sequences.
+For any associated weight $\pi_{{\sf BP}}>1$, the contribution of a valid sequence is $\pi_{{\sf BP}}^0 = 1$, and the contribution of an invalid sequence is $\pi_{{\sf BP}}^{+\infty} = 0$, so that Eq.~\ref{eq:mainproblem} (when restricted to $F^{\tt BP}) $simply counts the number of valid sequences.
 
@@
 
-Moreover, we provide two {\tt Python} wrapper scripts. The first script targets prescribed energies using multi-dimensional Boltzmann sampling.
+Moreover, we provide two {\tt Python} wrapper scripts. 
+The first script targets prescribed energies using multi-dimensional Boltzmann sampling.
 For a given set of secondary structures, together with prescribed target energies
@@
 
-\subsection*{A parsimonious stacking energy model as a proxy for the Turner energy model}
-Sampling for fixed weight
-Time benchmark
-
-Figure 10 - Correlation plot -- Right
-Figure 2 -- Distribution plot -- Right (Turner is even untractable to compute the treewidth itself in benchmark instances)
-Figure 3 -- Right -- Distribution
+\subsection*{Practical efficacy of Boltzmann sampling for sequences}
+
+\begin{figure}
+	\includegraphicstop[width=0.8\textwidth]{Figs/Plots/doublestem_weight1-500_mean}
+	\caption{\revised{Comparison of free-energy and Boltzmann probability for 10~000 uniform ($\pi=1$; blue dots) and Boltzmann distributed ($\pi=500$; green dots) sequences, targeting a simple structure consisting of two adjacent helices.}}\label{fig:BoltzmannBenefits}
+\end{figure}
+
+\revised{First, we show how seed sequences can be generated in a Boltzmann distribution, leading to designs that are substantially more stable that those generated uniformly.
+As can be seen in Figure~\ref{fig:BoltzmannBenefits} and Supp.
+Mat.
+\ref{appsec:immediate-benefits-for-negative-design}, sequences generated in the Boltzmann distribution not only reach lower free-energies than those generated in a uniform setting, but also achieve better Boltzmann probabilities.
+While the former is expected since the Boltzmann distribution explicitly favors low-energy candidates, the latter is somewhat surprising, since the Boltzmann probability of a target structure could, in principle, decrease under Boltzmann sampling due to the partition function growing faster than the Boltzmann factor.
+The empirical superiority of Boltzmann sampling appears robust, and is at the foundation of other approaches~\cite{Reinharz2013}.
+
+\begin{figure}
+	  \begin{minipage}{.35\textwidth}\centering 
+{\sffamily {\bfseries A} -- 
+	Stacking model}\\
+	\resizebox{\textwidth}{!}{
+		\begin{tabular}{c|c@{\hspace{.5em}}c@{\hspace{.5em}}c@{\hspace{.5em}}c@{\hspace{.5em}}c@{\hspace{.5em}}c}
+		$\Delta G$ 	& \Ab\Ub & \Cb\Gb & \Gb\Cb & \Gb\Ub & \Ub\Ab & \Ub\Gb \\[.4em]\hline
+			\Ab\Ub &
+			-0.18&
+			-1.13&
+			-1.09&
+			-0.38&
+			-0.26&
+			-0.62
+			\\[.4em]
+			\Cb\Gb&
+			-1.11&
+			-2.23&
+			-1.89&
+			-1.22&
+			-1.10&
+			-1.44
+			\\[.4em]
+			\Gb\Ub&
+			-0.55&
+			-1.26&
+			-1.58&
+			-0.72&
+			-0.50&
+			-0.69
+			\\
+	\end{tabular}}\\[3em]
+{\sffamily {\bfseries C} -- Tree widths}\\
+\includegraphics[width=\textwidth,trim=350 0 0 50,clip]{Figs/td-widths}
+\end{minipage}
+	  \begin{minipage}{.6\textwidth}\centering
+	{\sffamily {\bfseries B} -- Correlation plot}\\
+	  	\includegraphicstop[width=\textwidth,trim=0 0 0 50,clip]{Figs/stackingcor} 
+  	\end{minipage}
+
+	\caption{\revised{A fitted energy model based on stacking pairs ({\sffamily\bfseries A}) leads to approximated free-energies that are highly correlated (R=0.99) with free-energies in the Turner energy model ({\sffamily\bfseries B}), yet induces tree widths that are amenable to practical sampling ({\sffamily\bfseries C})}.\label{fig:correlation}}
+\end{figure}
+
+However, while in principle feasible, sampling in a Boltzmann distribution directly using the Turner energy model may induce extreme computational demands, with tree-widths scaling at least as large as the number of nucleotides in the largest loop.
+Fortunately, we found that intricacy of the Turner energy model can be circumvented with minimal loss of precision by using a simpler stacking energy model.
+As can be shown in Figure~\ref{fig:correlation}, a simple stacking energy model, further described in Supp.
+Mat.~\ref{appsec:modelparameters} can be used to approximate the Turner energy model (correlation coefficient $R=0.99$), while greatly reducing the tree width, and thus the computational requirements of the whole method even for complex instances.
+
+}
 
@@
 pseudoknots (D\&P model) and energies in the stacking energy
-model\revised{, which is obtained for each of the two target structures (with respective $R^2$ values of $0.846$ and $0.841$)}. For the shown fits we sampled $n=10\,000$ sequences, targeting a
+model\revised{, which is obtained for each of the two target structures (with respective $R^2$ values of $0.846$ and $0.841$)}. For the shown fits we sampled $n=1\,000$ sequences, targeting a
 \GCb\% of $60\%$.

### 2019-01-24 RNARedPrint.tex
Responded to Stefan's comments

@@
 Let us consider $m$ (secondary) structures $R_1,\dots, R_m$, each abstracted as a set of base pairs, and \revised{$k\geq m$} features $F_1,\dots, F_k$, associated with weights $\pi_1,\dots,\pi_k$.
-Our goal is to sample from the Boltzmann distribution, defined over the set of sequences that satisfy classic base pairing rules for each structure as:
+Our goal is to sample from the Boltzmann distribution, such that the probability of a sequence $S\in \Sigma^\star$ of length $n$, within the set of sequences that satisfy classic base pairing rules, is defined as:
 \todo{$S$ sequence not defined!}
@@
 \subsection*{Expressing \GCb\%-content, sequence validity and energies as features}
-\todo{$x_1$ position(?) not defined!\\$S_x_1$ nucleotide at position(?) not defined!\\$p$ number of dependent positions(?) from $x_p$ not defined!}
+\todo{$x_1$ position(?) not defined!\\$S_{x_1}$ nucleotide at position(?) not defined!\\$p$ number of dependent positions(?) from $x_p$ not defined!\\
+Yann: We just defined the dependencies as positions in the sequence, so $x_i$ should be fine from the context.
+$p$ is purely declarative (it differs for each $f$), so it should be ok, but since it is unused, we can probably drop it...
+}
 Formally, we define a \Def{feature $F$} as a function on sequences, whose value is obtained by summing over an associated set of \Def{contributions}.
@@
 $$
-F(S) =\!\!\!\!\!\!\!\sum_{\substack{\text{$f$ contribution of $F$,}\\\text{ $\dep(f)=\{x_1,\ldots,x_p\}$}}}\!\!\!\!\!\!
-f\left(\left\{\substack{x_1\mapsto S_{x_1}\\\cdots\\x_p\mapsto S_{x_p}}\right\}\right).
+F(S) =\!\!\!\!\!\!\!\sum_{\substack{\text{$f$ contribution of $F$,}\\\text{ $\dep(f)=\{x_1,x_2,\ldots\}$}}}\!\!\!\!\!\!
+f\left(\left\{\substack{x_1\mapsto S_{x_1}\\x_p\mapsto S_{x_2}\\\cdots}\right\}\right).
 $$
+where any $x_i \in [1,n]$ represents a position in $S$, and $S_x$ represents the letter found at position $x$ in $S$.
 
@@
 $$f^{\GCb}_i(\{i\mapsto c\})=\begin{cases}-1&\text{ if }c=\Gb\text{ or }\Cb\\ 0&\text{ otherwise}.\end{cases}$$ 
-\todo{Why here $i$ and not $x_i$??}
+\todo{Why here $i$ and not $x_i$??\\Yann: Because $x_1$, $x_2$...
+was just a way to talk about the first, second...
+argument of a function having an arbitrary number of args.
+Here, there is just one, so no need for subscripts.}
 \revised{By summing $f^{\GCb}_i(\{i\mapsto S_i\})$ over the whole sequence, one simply counts the occurrences of $\Gb$ and $\Cb$.}

### 2019-01-24 RNARedPrint.tex
Clarifications in Methods

--- 2019-01-24 RNARedPrint.tex
@@
 %\paragraph{Central computational problem.}
-Let us consider $m$ (secondary) structures $R_1,\dots, R_m$, each abstracted as a set of base pairs, and \revised{$k\geq m$} features $F_1,\dots, F_k$, associated with weights $\pi_1,\dots,\pi_k$.
-Our goal is to sample from the Boltzmann distribution, defined over the set of sequences that satisfy classic base pairing rules for each structure as:
-\todo{$S$ sequence not defined!}
+Let us consider $m$ (secondary) structures $R_1,\dots, R_m$, each abstracted as a set of base pairs, and \revised{$k\geq m$} features $F_1,\dots, F_k$\revised{, typically representing the energies of the structures and additional sequence properties,} associated with weights $\pi_1,\dots,\pi_k$.
+Our goal is to sample sequences \revised{$S$ (which satisfy the base pairing rules for all strucutures) from the Boltzmann distribution defined by}
 \begin{equation}
@@
 
-The workhorse of our approach is the \revised{fixed-parameter tractable} computation of feature-dependent partition functions over sequences:
-%Given are the features $F_1,\dots,F_k$, which encode the energies functions for the fixed structures and the function for the \GCb\%, all in terms of simple atomic energy contributions.
-
-For weights $\pi_1,\dots,\pi_k$, one wants to compute the partition function
+The workhorse of our approach is the \revised{fixed-parameter tractable} computation of feature-dependent partition functions over sequences, \revised{namely partition functions of the form} 
+%
   \begin{equation}
     \label{eq:mainproblem}
-    \partfun{\pi_1,\dots,\pi_k} = \sum_{S\in\Sigma^n} \prod_{1\leq \ell\leq k} \pi_\ell^{-F_\ell(S)}.
+    \partfun{\pi_1,\dots,\pi_k} = \sum_{S\in\Sigma^n} \prod_{1\leq \ell\leq k} \pi_\ell^{-F_\ell(S)},
   \end{equation}
+  for specfic weights $\pi_1,\dots,\pi_k$.
 
 \subsection*{Expressing \GCb\%-content, sequence validity and energies as features}
-\todo{$x_1$ position(?) not defined!\\$S_x_1$ nucleotide at position(?) not defined!\\$p$ number of dependent positions(?) from $x_p$ not defined!}
 Formally, we define a \Def{feature $F$} as a function on sequences, whose value is obtained by summing over an associated set of \Def{contributions}.
@@
 $$
-
-The $\GCb\%$ can be simply expressed using $n$ contributions $f^{\GCb}_i$, $i\in[1,n]$, whose dependencies are restricted to individual nucleotides,  and such that 
+\revised{Here, since $\dep(f)=\{x_1,\ldots,x_p\}$, $\left\{\substack{x_1\mapsto S_{x_1}\\\cdots\\x_p\mapsto S_{x_p}}\right\}$ denotes the assignment, that assigns to all positions $x_k$  in $\dep(f)$ the respective nucleotide $S_{x_k}$ ($1\leq k\leq p$).}
+
+The $\GCb\%$ can be simply expressed using $n$ contributions $f^{\GCb}_i$, $i\in[1,n]$, \revised{each dependencing only on position $i$, i.e.~$\dep(f^{\GCb}_i)=\{i\}$}, such that 
 $$f^{\GCb}_i(\{i\mapsto c\})=\begin{cases}-1&\text{ if }c=\Gb\text{ or }\Cb\\ 0&\text{ otherwise}.\end{cases}$$
-\todo{Why here $i$ and not $x_i$??}
-\revised{By summing $f^{\GCb}_i(\{i\mapsto S_i\})$ over the whole sequence, one simply counts the occurrences of $\Gb$ and $\Cb$.}
+
+\revised{By summing $f^{\GCb}_i(\{i\mapsto S_i\})$ over the whole sequence ($i=1,\dots,n$), one simply counts the occurrences of $\Gb$ and $\Cb$.}
 
@@
 
-Moreover, we provide two {\tt Python} wrapper scripts. 
-The first script targets prescribed energies using multi-dimensional Boltzmann sampling.
+Moreover, we provide two {\tt Python} wrapper scripts. The first script targets prescribed energies using multi-dimensional Boltzmann sampling.
 For a given set of secondary structures, together with prescribed target energies
@@
 
-\subsection*{Practical efficacy of Boltzmann sampling for sequences}
-
-\begin{figure}
-	\includegraphicstop[width=0.8\textwidth]{Figs/Plots/doublestem_weight1-500_mean}
-	\caption{\revised{Comparison of free-energy and Boltzmann probability for 10~000 uniform ($\pi=1$; blue dots) and Boltzmann distributed ($\pi=500$; green dots) sequences, targeting a simple structure consisting of two adjacent helices.}}\label{fig:BoltzmannBenefits}
-\end{figure}
-
-\revised{First, we show how seed sequences can be generated in a Boltzmann distribution, leading to designs that are substantially more stable that those generated uniformly.
-As can be seen in Figure~\ref{fig:BoltzmannBenefits} and Supp.
-Mat.
-\ref{appsec:immediate-benefits-for-negative-design}, sequences generated in the Boltzmann distribution not only reach lower free-energies than those generated in a uniform setting, but also achieve better Boltzmann probabilities.
-While the former is expected since the Boltzmann distribution explicitly favors low-energy candidates, the latter is somewhat surprising, since the Boltzmann probability of a target structure could, in principle, decrease under Boltzmann sampling due to the partition function growing faster than the Boltzmann factor.
-The empirical superiority of Boltzmann sampling appears robust, and is at the foundation of other approaches~\cite{Reinharz2013}.
-
-\begin{figure}
-	  \begin{minipage}{.35\textwidth}\centering 
-{\sffamily {\bfseries A} -- 
-	Stacking model}\\
-	\resizebox{\textwidth}{!}{
-		\begin{tabular}{c|c@{\hspace{.5em}}c@{\hspace{.5em}}c@{\hspace{.5em}}c@{\hspace{.5em}}c@{\hspace{.5em}}c}
-		$\Delta G$ 	& \Ab\Ub & \Cb\Gb & \Gb\Cb & \Gb\Ub & \Ub\Ab & \Ub\Gb \\[.4em]\hline
-			\Ab\Ub &
-			-0.18&
-			-1.13&
-			-1.09&
-			-0.38&
-			-0.26&
-			-0.62
-			\\[.4em]
-			\Cb\Gb&
-			-1.11&
-			-2.23&
-			-1.89&
-			-1.22&
-			-1.10&
-			-1.44
-			\\[.4em]
-			\Gb\Ub&
-			-0.55&
-			-1.26&
-			-1.58&
-			-0.72&
-			-0.50&
-			-0.69
-			\\
-	\end{tabular}}\\[3em]
-{\sffamily {\bfseries C} -- Tree widths}\\
-\includegraphics[width=\textwidth,trim=350 0 0 50,clip]{Figs/td-widths}
-\end{minipage}
-	  \begin{minipage}{.6\textwidth}\centering
-	{\sffamily {\bfseries B} -- Correlation plot}\\
-	  	\includegraphicstop[width=\textwidth,trim=0 0 0 50,clip]{Figs/stackingcor} 
-  	\end{minipage}
-
-	\caption{\revised{A fitted energy model based on stacking pairs ({\sffamily\bfseries A}) leads to approximated free-energies that are highly correlated (R=0.99) with free-energies in the Turner energy model ({\sffamily\bfseries B}), yet induces tree widths that are amenable to practical sampling ({\sffamily\bfseries C})}.\label{fig:correlation}}
-\end{figure}
-
-However, while in principle feasible, sampling in a Boltzmann distribution directly using the Turner energy model may induce extreme computational demands, with tree-widths scaling at least as large as the number of nucleotides in the largest loop.
-Fortunately, we found that intricacy of the Turner energy model can be circumvented with minimal loss of precision by using a simpler stacking energy model.
-As can be shown in Figure~\ref{fig:correlation}, a simple stacking energy model, further described in Supp.
-Mat.~\ref{appsec:modelparameters} can be used to approximate the Turner energy model (correlation coefficient $R=0.99$), while greatly reducing the tree width, and thus the computational requirements of the whole method even for complex instances.
-
-}
+\subsection*{A parsimonious stacking energy model as a proxy for the Turner energy model}
+Sampling for fixed weight
+Time benchmark
+
+Figure 10 - Correlation plot -- Right
+Figure 2 -- Distribution plot -- Right (Turner is even untractable to compute the treewidth itself in benchmark instances)
+Figure 3 -- Right -- Distribution
 

### 2019-01-24 RNARedPrint.tex

--- 2019-01-24 RNARedPrint.tex
@@
 
-The workhorse of our approach is the \revised{fixed-parameter tractable} computation of feature-dependent partition functions over sequences:
-%Given are the features $F_1,\dots,F_k$, which encode the energies functions for the fixed structures and the function for the \GCb\%, all in terms of simple atomic energy contributions.
-
-For weights $\pi_1,\dots,\pi_k$, one wants to compute the partition function
+The workhorse of our approach is the \revised{fixed-parameter tractable} computation of feature-dependent partition functions over sequences, \revised{namely partition functions of the form} 
+%
   \begin{equation}
     \label{eq:mainproblem}
-    \partfun{\pi_1,\dots,\pi_k} = \sum_{S\in\Sigma^n} \prod_{1\leq \ell\leq k} \pi_\ell^{-F_\ell(S)}.
+    \partfun{\pi_1,\dots,\pi_k} = \sum_{S\in\Sigma^n} \prod_{1\leq \ell\leq k} \pi_\ell^{-F_\ell(S)},
   \end{equation}
+  for specfic weights $\pi_1,\dots,\pi_k$.
 
 \subsection*{Expressing \GCb\%-content, sequence validity and energies as features}
-\todo{$x_1$ position(?) not defined!\\$S_x_1$ nucleotide at position(?) not defined!\\$p$ number of dependent positions(?) from $x_p$ not defined!}
 Formally, we define a \Def{feature $F$} as a function on sequences, whose value is obtained by summing over an associated set of \Def{contributions}.
 Each contribution $f$ takes values in $\mathbb{R}\cup \{+\infty\}$, and depends on the nucleotides assigned to a restricted set of positions, its \emph{dependencies $\dep(f)$}, such that
+}
 $$
-F(S) =\!\!\!\!\!\!\!\sum_{\substack{\text{$f$ contribution of $F$,}\\\text{ $\dep(f)=\{x_1,\ldots,x_p\}$}}}\!\!\!\!\!\!
-f\left(\left\{\substack{x_1\mapsto S_{x_1}\\\cdots\\x_p\mapsto S_{x_p}}\right\}\right).
+F(S) =\!\!\!\!\!\!\!\sum_{\substack{\text{$f$ contribution of $F$,}\\\text{ $\dep(f)=\{x_1,x_2,\ldots\}$}}}\!\!\!\!\!\!
+f\left(\left\{\substack{x_1\mapsto S_{x_1}\\x_p\mapsto S_{x_2}\\\cdots}\right\}\right).
 $$
-
-The $\GCb\%$ can be simply expressed using $n$ contributions $f^{\GCb}_i$, $i\in[1,n]$, whose dependencies are restricted to individual nucleotides,  and such that 
+\revised{Here, since $\dep(f)=\{x_1,\ldots,x_p\}$, $\left\{\substack{x_1\mapsto S_{x_1}\\\cdots\\x_p\mapsto S_{x_p}}\right\}$ denotes the assignment, that assigns to all positions $x_k$  in $\dep(f)$ the respective nucleotide $S_{x_k}$ ($1\leq k\leq p$).}
+
+The $\GCb\%$ can be simply expressed using $n$ contributions $f^{\GCb}_i$, \revised{each dependencing only on position $i\in[1,n]$, i.e.~$\dep(f^{\GCb}_i)=\{i\}$}, such that 
 $$f^{\GCb}_i(\{i\mapsto c\})=\begin{cases}-1&\text{ if }c=\Gb\text{ or }\Cb\\ 0&\text{ otherwise}.\end{cases}$$
-\todo{Why here $i$ and not $x_i$??}
-\revised{By summing $f^{\GCb}_i(\{i\mapsto S_i\})$ over the whole sequence, one simply counts the occurrences of $\Gb$ and $\Cb$.}
+
+\revised{By summing $f^{\GCb}_i(\{i\mapsto S_i\})$ over the whole sequence ($i=1,\dots,n$), one simply counts the occurrences of $\Gb$ and $\Cb$.}
 

### 2019-01-24 RNARedPrint.tex
Updates from ShareLaTeX
fix bracket

@@
 %% document please refer to Readme.html and the instructions for   %%
-%% authors page on the biomed central website                      %%
+%% authors page on the biomed central website                      %%https://www.overleaf.com/project/5b98dc8ced90175e9717c15a
 %% http://www.biomedcentral.com/info/authors/                      %%
@@
 Each contribution $f$ takes values in $\mathbb{R}\cup \{+\infty\}$, and depends on the nucleotides assigned to a restricted set of positions, its \emph{dependencies $\dep(f)$}, such that
-}
+
 $$
@@
 $$
-\revised{Here, since $\dep(f)=\{x_1,\ldots,x_p\}$, $\left\{\substack{x_1\mapsto S_{x_1}\\\cdots\\x_p\mapsto S_{x_p}}\right\}$ denotes the assignment, that assigns to all positions $x_k$  in $\dep(f)$ the respective nucleotide $S_{x_k}$ ($1\leq k\leq p$).}
+\revised{Here, since $\dep(f)=\{x_1,\ldots,x_p\}, $\left\{\substack{x_1\mapsto S_{x_1}\\\cdots\\x_p\mapsto S_{x_p}}\right\}$ denotes the assignment, that assigns to all positions $x_k$  in $\dep(f)$ the respective nucleotide $S_{x_k}$ ($1\leq k\leq p$).}
 

### 2019-01-24 RNARedPrint.tex
typo
resolve inconsistency due to merge
Updates from ShareLaTeX

@@
 
-        Energy models for structure prediction vary considerably, yet can always be expressed as sums over contributions associated with local structural motifs (base pairs, base pair stacks, loops\ldots) under a certain nucleotide assignment.
+        Energy models for structure prediction vary considerably, yet can always be expressed as sums over contributions associated with local structural motifs (base pairs, base pair stacks, loops, \ldots) under a certain nucleotide assignment.
 Energy models can thus be captured generically by introducing, for each motif $m$ occurring in a target structure, a contribution $f_m$, taking a specific value for each assignment of nucleotides to its positions $\dep(f_m)$.
@@
 positions) connected by \Def{hyperedges}, which are arbitrary sets of
-vertices. In this way, hypergraphs generalize undirected graphs, where
+vertices. In this way, hypergraphs generalize undirected graphs where
 each edge is a set of exactly two vertices.

### 2019-01-24 RNARedPrint.tex
Cut long intro sentence for better readability

@@
 \revised{Remarkably, we could show that it is not possible to find a better, efficient method for sampling (unless ${\sf P}={\sf NP}$), since the underlying counting problem is \#{\sf P}-hard.
-The practical relevance of this theoretical result is that it rules out substantially better sampling techniques, such that there will always---even using improved sampling methods---remain an upper limit on the---in practice---tractable number and heterogeneity of structures, the complexity of the directly treatable energy model, and the number and complexity of additional constraints that could be considered in future sampling-based applications.} Technically, this result relies on a surprising bijection between valid sequences and independent sets of a bipartite graph, the latter being the object of recent breakthroughs in approximate counting complexity~\citep{Bulatov2013,Cai2016}.
+The practical relevance of this theoretical result is that it rules out substantially better sampling techniques.
+Even when using improved sampling methods, there will always remain an upper limit on the (in practice) tractable number and heterogeneity of structures, the complexity of the directly treatable energy model, and the number and complexity of additional constraints that could be considered in future sampling-based applications.} Technically, this result relies on a surprising bijection between valid sequences and independent sets of a bipartite graph, the latter being the object of recent breakthroughs in approximate counting complexity~\citep{Bulatov2013,Cai2016}.
 
@@
 $$
-F(S) =\!\!\!\!\!\!\!\sum_{\substack{\text{$f$ contribution of $F$,}\\\text{ $\dep(f)=\{x_1,x_2,\ldots\}$}}}\!\!\!\!\!\!
-f\left(\left\{\substack{x_1\mapsto S_{x_1}\\x_p\mapsto S_{x_2}\\\cdots}\right\}\right).
+F(S) =\!\!\!\!\!\!\!\sum_{\substack{\text{$f$ contribution of $F$,}\\\text{ $\dep(f)=\{x_1,\ldots,x_p\}$}}}\!\!\!\!\!\!
+f\left(\left\{\substack{x_1\mapsto S_{x_1}\\\cdots\\x_p\mapsto S_{x_p}}\right\}\right).
 $$
-\revised{Here, since $\dep(f)=\{x_1,\ldots,x_p\}, $\left\{\substack{x_1\mapsto S_{x_1}\\\cdots\\x_p\mapsto S_{x_p}}\right\}$ denotes the assignment, that assigns to all positions $x_k$  in $\dep(f)$ the respective nucleotide $S_{x_k}$ ($1\leq k\leq p$).}
+\revised{Here, since $\dep(f)=\{x_1,\ldots,x_p\}$, $\left\{\substack{x_1\mapsto S_{x_1}\\\cdots\\x_p\mapsto S_{x_p}}\right\}$ denotes the assignment, that assigns to all positions $x_k$  in $\dep(f)$ the respective nucleotide $S_{x_k}$ ($1\leq k\leq p$; $p=|\dep(f)|$).}
 

### 2019-01-24 BMC-Response-Round1.tex
some response

@@
 In particular, the reviews helped us to state the goal of this work more clearly and improve its presentation.
-While we demonstrate the applicability of our approach to negative RNA design, which (here by optimizing 'MultiDefect') aims at high and similar probability of the target structures, the focus of this work is the positive design of sequences that fold into multiple target structures with highly specific properties.
+While we demonstrate the applicability of our approach to negative RNA design, which  aims at high and similar probability of the target structures (here measured as 'MultiDefect'), the focus of this work is the positive design of sequences that fold into multiple target structures with highly specific properties.
 We discuss this positive design, in particular, for specific folding energies (for each single target structure) and GC content.
@@
 To emphasize our focus on positive design and the prototypical nature of our brief proof-of-concept study in the use of our method for negative design, we carefully rephrased the respective paragraph at the end of the introduction as well statements in the subsection about the optimization of the MultiDefect.
-(TODO: write!)
-
-\todo[inline]{Mention that our sampling could be used in combination with MODENA / not necessarily competition}
 
@@
 Thus, this manuscript is considered to be further improved in terms of intelligibility of the methods and results using fair evaluation criteria.\\[1em]}%
-\Answer{We structured and rephrased the Results section more clearly and adapted the introduction accordingly.
+
+\Answer{
+  We thank the reviewer for the criticism, which motivated us to work on clarifying the practical use of the method and tool RNARedPrint; we have carefully edited the entire manuscript having these issues in mind.
+  In the methods section, we tried to improve the general readability especially of our overview of the method and central problem.
+  Moreover, we structured and rephrased the Results section more clearly and adapted our contributions outline and conclusion accordingly.
 In Results, we introduce single subsections for the following sub-results to reflect our main results in a logical order:
   \begin{itemize}
-    \item The full-featured RNA energy model (Turner) can be well approximated by the much more efficiently tractable stacking model; in this model, controlled Boltzmann sampling for typical multi-target design instances requires little time (corresponding to small tree-width);
+    \item Boltzmann sampling of positively designed sequences for multiple target structures works effectively and accurately, where the full-featured RNA energy model (Turner) can be well approximated.
+      For this approximation, we use the much more efficiently tractable stacking model; in this model, controlled Boltzmann sampling for typical multi-target design instances requires little time (corresponding to small tree-width);
     \item Multi-dimensional Boltzmann sampling solves the positive design problem effectively, requiring only a limited number of rejections (even for multiple simultaneously considered features);
-    \item The positively designed sequences are suitable for improving negative design approaches.
+    \item Our method can be beneficially used in the setting of negative design. Here, we demonstrate a typical use case of the method. Moreover, we empirically show that 
+      the positively designed sequences can improve negative design approaches, where
+      they work better than previous approaches (for seed generation).
 \end{itemize}}
-\AnswerTODO{Restructure Results and Intro)}
 
@@
 In the scope of this study, the running time and memory comparison should be shown to produce the same number of sequences by some minor variants of RNARedPrint method, RNABluePrint, and MODENA.\\[1em]}%
-\Answer{We agree that running times are important for assessing the method; consequently, we included detailed timing information in the revision.
-The running times directly depend on the already reported treewidths (which we actively keep low due to approximating the full energy model and using state-of-the-art decomposition methods).
+\Answer{We agree that running times are important for assessing the method; consequently, we included detailed information in our supplement (Supp.
+Mat.
+C).
+Moreover, Supp.
+Mat.
+E provides detailed information on the generation of seed sequences in our application to negative design.
+
+Please note that the running times directly depend on the (for the benchmark sets already reported) treewidths, which strongly depend on the specific design instances.
+Notably, we took measures to keep these treewidths low; in particular 1) we show that it suffices to sample based on an approximation of the full energy model and 2) we tailored the approach to benefit from state-of-the-art decomposition methods.
 In consequence, the run-time requirements are generally modest for the relevant design instances; making our approach applicable in practice.
 
-Concerning the comparison: since MODENA does not allow to measure separate times for sequence sampling, we could only estimate its run-times by running the MODENA binary with parameters that turn off the optimization as far as this is possible.
-Moreover, we want to re-emphasize that MODENA, while it implements a sophisticated optimization procedure, does not put any specific focus on the sampling procedure for generating its seed sequences.
+We included further data comparing to RNABluePrint in the supplement (Supp.
+Mat.
+C) and discuss the comparison to MODENA.
+While comparing to MODENA in the same way as to RNABluePrint can admittedly appear obvious, the sampling in MODENA is of a very different nature, such that elaborate comparisons do not provide relevant insights.
+Here is a brief elaboration:
+
+While MODENA implements a sophisticated optimization procedure, does not put any specific focus on the sampling procedure for generating its seed sequences.
 Consequently, it does not sample from a specific distribution and produces random sequences without targeting any specific properties (neither good stability nor specific GC-content, etc.).
@@
 positive design).
-This crucial difference should be kept in mind, when comparing the run-times of these tools.
-\AnswerTODO{...}}
+In consequence, MODENA's sampling procedure is non-surprisingly much faster than the ones of RNARedPrint and RNABluePrint, which we openly admit in our text.
 
+Reporting data for comparison would have been difficult for one additional reason: MODENA does not allow to measure separate times for sequence sampling.
+Thus, we could only estimate its run-times by running the MODENA binary with parameters that turn off the optimization as far as this is possible.
+}
 \Comment{

### 2019-01-24 RNARedPrint.tex

@@
 To formalize our algorithms, which iteratively merge evaluations of partial solutions, we extend the idea of atomic feature contributions, which are evaluated at sets of the form $\{x_1\mapsto v_1, \dots, x_d\mapsto v_p\}$.
-Let us call the latter object a \Def{partial sequence}; such an object will help specify partial knowledge on the sequence at some point of the algorithm.
+Let us call the latter object a \Def{partial sequence}.
+Such an object will help \revised{to} specify partial knowledge on the sequence at some point of the algorithm.
 Easily, we can extend the definition of contributions $f$ to sets $\{x_1\mapsto v_1, \dots, x_p\mapsto v_p\}$, where $\{x_1\dots x_p\}$ is any super-set of $\dep(f)$ by ignoring the superfluous assignments $x\mapsto v$, where $x\not\in\dep(f)$.
@@
 on a tree decomposition of $G_{\F}$, the dependency graph associated with $\F$.
-Note that analogous algorithms could be easily derived to count valid sequences, or list sequences having minimal free-energy.
+Note, that analogous algorithms could be easily derived to count valid sequences, or list sequences having minimal free-energy.
 %This yields fixed parameter tractable (FPT) algorithms, based on the treewidth.
@@
 \end{itemize}
+\todo{$\Theta$ not defined, what is it?}
 

### 2019-01-24 RNARedPrint.tex
Updates from ShareLaTeX

@@
 
-In order to capture realistic energy models like the Turner model or pseudoknot models like {\tt HotKnots}~\citep{Ren2005}, our sampling strategy can be extended in two ways: 1) either by directly sampling based on more expressive energy models; 2) or by sampling in a simple energy model can be used to approximate sampling in more complex models.
+In order to capture realistic energy models like the Turner model or pseudoknot models like {\tt HotKnots}~\citep{Ren2005}, our sampling strategy can be extended in two ways: 1) either by directly sampling based on more expressive energy models or 2) by sampling in a simple energy model can be used to approximate sampling in more complex models.
 In practice, complex energy models has a strong influence on the treewidth (of optimal tree decompositions) of the dependency graph and, in turn the computational complexity of our approach.

### 2019-01-24 BMC-Response-Round1.tex
Typos and grammar.

@@
 \Answer{
-  We thank the reviewer for the criticism, which motivated us to work on clarifying the practical use of the method and tool RNARedPrint; we have carefully edited the entire manuscript having these issues in mind.
+  We thank the reviewer for the criticism, which motivated us to work on clarifying the practical use of the method and tool RNARedPrint in several ways; we have carefully edited the entire manuscript having these issues in mind.
   In the methods section, we tried to improve the general readability especially of our overview of the method and central problem.
@@
 C) and discuss the comparison to MODENA.
-While comparing to MODENA in the same way as to RNABluePrint can admittedly appear obvious, the sampling in MODENA is of a very different nature, such that elaborate comparisons do not provide relevant insights.
-Here is a brief elaboration:
+While comparing to MODENA in the same way as to RNABluePrint can admittedly appear obvious, the sampling in MODENA is of a very different nature, such that---in our opinion---elaborate comparisons would not provide relevant insights.
+Of course, this has to be explained:
 
-While MODENA implements a sophisticated optimization procedure, does not put any specific focus on the sampling procedure for generating its seed sequences.
-Consequently, it does not sample from a specific distribution and produces random sequences without targeting any specific properties (neither good stability nor specific GC-content, etc.).
+While MODENA implements a sophisticated optimization procedure, it does not put specific focus on the sampling procedure for generating its seed sequences.
+Consequently, it does not sample from a specific distribution and produces sequences without targeting any specific properties (neither good stability nor specific GC-content, etc.).
 In contrast, RNARedPrint produces highly specific samples, thereby solving a fundamentally different problem (i.e.
 positive design).
-In consequence, MODENA's sampling procedure is non-surprisingly much faster than the ones of RNARedPrint and RNABluePrint, which we openly admit in our text.
+As a consequence, MODENA's sampling procedure is non-surprisingly much faster than the ones of RNARedPrint and RNABluePrint, which we openly admit in our text.
 
@@
 Since RNARedPrint keeps the sequences to be under the specific constraint by a simple rejection method, such information for various target conditions would be better to estimate the practical running time of RNARedPrint.\\[1em]}%
-\AnswerTODO[Sebastian]{
-We now report the frequency of rejections measured under various conditions. +add plots (moving distributions over rounds...)
+\Answer{
+We do now explicitly report the run times of the sampling procedure for specific weights (Supp. Mat. C).
+To further the run times of the multi-dimensional Boltzmann sampling and the seed generation, which both
+
 }

### 2019-01-24 RNARedPrint.tex
Typos and grammar.

@@
 \end{itemize}
-\todo{$\Theta$ not defined, what is it?}
 
@@
 
-In order to capture realistic energy models like the Turner model or pseudoknot models like {\tt HotKnots}~\citep{Ren2005}, our sampling strategy can be extended in two ways: 1) either by directly sampling based on more expressive energy models or 2) by sampling in a simple energy model can be used to approximate sampling in more complex models.
-In practice, complex energy models has a strong influence on the treewidth (of optimal tree decompositions) of the dependency graph and, in turn the computational complexity of our approach.
-Therefore, it is interesting to consider, in addition to the base pair model, other
+In order to capture realistic energy models like the Turner model or pseudoknot models like {\tt HotKnots}~\citep{Ren2005}, our sampling strategy can be extended in two ways: 1) either by directly sampling based on more expressive energy models \revised{or 2) by sampling in a simple energy model which} can be used to approximate sampling in more complex models.
+In practice, complex energy models \revised{have} a strong influence on the treewidth (of optimal tree decompositions) of the dependency graph and \revised{thus on} the computational complexity of our approach.
+Therefore, it is interesting to consider---in addition to the base pair model---other
 stripped-down variants of the nearest neighbor model, which could offer a compromise between the low-complexity base pair energy model and the high-accuracy Turner model.
@@
 A first model, which is particularly promising, is the
-\Def{stacking energy model}. This model assigns only assigns energy
+\Def{stacking energy model}. This model only assigns energy
 contributions $\Delta G(x_i,x_j,x_{i+1},x_{j-1})$ to stacks consisting of two nested base pairs $(i,j)$ and $(i+1,j-1)$.
@@
 
-Complex \Def{loop-based}
- energy models---e.g.~the Turner model,
-which, among others, includes energy terms for special loops and dangling
+Complex \Def{loop-based} energy models---e.g.~the Turner model which, among others, includes energy terms for special loops and dangling
 ends---can also be encoded exactly as instances of our general framework.
@@
 The flexibility of our framework allows to support the advanced sampling technique called "multidimensional Boltzmann sampling"~\citep{Bodini2010}, which allows to enforce (probabilistically)  additional, complex properties of the samples through an additional rejection.
-This technique was previously used to control \GCb\%~\citep{Waldispuehl2011,Reinharz2013} and dinucleotide content~\citep{Zhang2013} of sampled RNA sequences; here, in addition to controlling \GCb\% (our feature $F_0$) we use it to target the free energies $(\TargetE_1,\ldots,\TargetE_k)$ of the individual target structures (features $F_1,\dots,F_m$).
+This technique was previously used to control \GCb\%~\citep{Waldispuehl2011,Reinharz2013} and dinucleotide content~\citep{Zhang2013} of sampled RNA sequences.
+Here, in addition to controlling \GCb\% (our feature $F_0$) we use it to target the free energies $(\TargetE_1,\ldots,\TargetE_k)$ of the individual target structures (features $F_1,\dots,F_m$).
 
@@
 %
-b\item $\Psi$ is surjective, i.e. there is a preimage for each element $I\in \IS{G}$. Define $f\in\Design{G}$ as\begin{displaymath}
+\item $\Psi$ is surjective, i.e. there is a preimage for each element $I\in \IS{G}$. Define $f\in\Design{G}$ as\begin{displaymath}
 f(v) = \left\{\begin{array}{ll} % @{\qquad}ll

### 2019-01-24 BMC-Response-Round1.tex
Updates from ShareLaTeX

@@
 Related to that problem, it is also not clear what methods (or what options or approximations) are applied for each comparison in the results.
-Thus, this manuscript is considered to be further improved in terms of intelligibility of the methods and results using fair evaluation criteria.\\[1em]}%
+Thus, this manuscript is considered to be further improved in terms of intelligibility of the methods and results using fair evaluation criteria.}%
 
@@
 \Comment{For evaluation of a method to design RNA molecules, the number of the solutions it produces and its running time are considered to be important.\\[1em]
-In the scope of this study, the running time and memory comparison should be shown to produce the same number of sequences by some minor variants of RNARedPrint method, RNABluePrint, and MODENA.\\[1em]}%
+In the scope of this study, the running time and memory comparison should be shown to produce the same number of sequences by some minor variants of RNARedPrint method, RNABluePrint, and MODENA.}
+
 \Answer{We agree that running times are important for assessing the method; consequently, we included detailed information in our supplement (Supp.
@@
 We do now explicitly report the run times of the sampling procedure for specific weights (Supp. Mat. C). 
-To further the run times of the multi-dimensional Boltzmann sampling and the seed generation, which both
+The multi-dimensional Boltzmann sampling and the seed generation, typically need several iterations of this sampling procedure.
+The numbers of required iterations for the benchmark instances are now reported in detail in Supp.
+Mat.
+E, such that a reader can get a complete picture of the computational demands.}
 
-}
 \Comment{
@@
 Moreover the final targeted free energies differ greatly across structures across our various benchmark, so one would have to report them on a per sequence basis, leading to a result that would be hard to interpret.
-Finally, note that the MODENA multidefect objective function already depends critically from achieving low, consistent, energies, and already arguably captures this notion of stability.}
+Finally, note that the MODENA MultiDefect objective function already depends critically on achieving low, consistent energies, which arguably already captures this notion of stability.}
+
 \Comment{
@@
 \end{itemize}
+In both cases, the use of weights is rather technical, which explains why weights are mentioned in the methods section.
+Weights are much rather a means to target specific energies (or other feature values like the GC content).
+
+Partially, we have problems to understand the question, since RNAblueprint does not use weights at all. We hope that our improvements to the text help to clarify.
 }
-\AnswerTODO[Yann]{
-  Introducing gently the concept of adaptive sampling and weights in intro + Diplomatic answer
-}
+
 \Comment{
-Moreover, it is not well described how we can choose the initial weights and what values they are converged in the actual cases (e.g., Fig2 and S4) after adaptive sampling.\\[1em]}%
+Moreover, it is not well described how we can choose the initial weights and what values they are converged in the actual cases (e.g., Fig2 and S4) after adaptive sampling.}
+
 \Answer{
+The initial weights are somewhat arbitrary and do not have much influence, while the final weights are automatically derived by the adaptive strategy.
+Thus, the actual value of the weights is not informative for the typical use of the method.
+That being said, we used 1 as initial weights.
+In the example of Figure 3, in the case of targeting respective energies of -20 and -30 kcal/mol for the two target structures, we arrive at respecticve weights of 229 and 117 in the last iteration.
+The entire procedure to generate 5000 samples targeted at these energies took 6m20s (CPU i7-4770).
+Note that most of this time is consumed by expensive energy evaluations by HotKnots, which we need due to the pseudoknots in this example.
+
+Some additional more technical remarks:
 The calibration of weights relies on a numerical iteration aiming at computing weights such that expected properties of sampled sequences (which depend on weights) match their targeted values.
 Associated problems are usually convex, and we describe in supplementary material a numerical iteration to compute suitable weights.
-We also added to Fig. 2 and 4 the final weights.
 }
+
 \Comment{
@@
 \\[1em]}%
-\AnswerTODO[Yann]{Our approximation only hold for a single structure.
-Long loops tend to be oblivious of their content so, while approximating the energy model for long loops would be a problem in the context of structure prediction, they are not in the context of design.
-(to be clarified/elaborated)
-}
+\Answer{
+We agree that this would be of utmost importance, if our goal would have been to estimate parameters for structure prediction.
+In our case of design, where structures are known and constant, we can simply learn the offset caused by the long loops, stacking, and other features by regression on generated samples.}
+
 \Comment{
@@
 \\[1em]}%
-\AnswerTODO[Stefan+Sebastian]{
-Compute $R^2$ (detaching plot). Write paragraph about goodness of fit
+\Answer{
+We do now include data on the quality of the fit, which show its good quality.
+Thanks for the reminder.
+The final test for this fit is of course, our ability to use this for effective positive design in reasonable time.
 }
 \Comment{
-Through the analyses, the names of energy models should be specified more clearly. The original model, simplified base paired model, and stacking model are confusing.
-\\[1em]}%
+Through the analyses, the names of energy models should be specified more clearly. The original model, simplified base paired model, and stacking model are confusing.}
+
 \Answer{While we believe that the capacity of our framework to describe various energy models is important, and demonstrates its versatility and potential for future developments, it is true that only the stacking model is absolutely necessary to understand the method.
-To avoid any further confusion, we moved the paragraph describing the integration of various energy models to the appendix.
-We now only describe our simplified stacking model, which we more explicitly show to be highly correlated with the Turner energy model.}
-\AnswerTODO[Sebastian]{Check that BP model does not appear too recurrently in manuscript}
+We thank the reviewer for pointing out this opportunity to simplify the description: we moved the paragraph describing the integration of various energy models to the appendix.
+We now describe only our simplified stacking model, which we more explicitly show to be highly correlated with the Turner energy model.}
+
 \Comment{
@@
 \\[1em]}%
-\Answer{The MultiDefect objective function is based on the rationale that target structures should achieve high stability, and that the chosen RNA cannot adopt alternative, more stable, structures.
+\Answer{We agree that MultiDefect is simply \emph{one} reasonable measure for the performance of multi-target design.
+We even believe that more research is required in this respect to define and evaluate such measures.
+However, we felt that for our purpose of demonstrating the principle use of our tool for negative design using one 
+generally accepted reasonable measure is sufficient.
+
+To elaborate: the MultiDefect objective function is based on the rationale that target structures should achieve high stability, and that the chosen RNA cannot adopt alternative, more stable, structures.
 It is a well accepted metrics introduced by other contributors, which we did not feel necessary to question or further improve.
 }
+
 \Comment{
@@
 Also, RNARedprint and RNABlueprint are optimised to increase only a Multidefect value.\\[1em]}%
-\Answer{First of all, it is perhaps worth reminding that RNAredprint is not intended as a tool for negative design, but rather a generation algorithm, not only addressing a large array of design use-cases, but also able to generate good seed sequences, amenable to further optimization using other tools.
+
+\Answer{As we tried to make clear in our general comments of this response letter (as well as by appropriate adaptions in the manuscript), RNAredprint is not intended to perform negative design.
+It is a generation algorithm, not only addressing a large array of design use-cases, but also able to generate good seed sequences, amenable to further optimization using other tools.
 For this reason, we focused our validation on this potential for better initialization, and compared the quality of sequences generated by RNARedPrint (Boltzmann weighted) to RNABlueprint (uniform).
@@
 \Comment{Hence, the distribution of Multidefect, previous criteria, and the first and second term of Multidefect should be shown in box plots or violin plots to know the robustness of RNARedPrint.
-It is also interesting what is the characteristic of the sequence designed by each method in a specific example of functional RNA molecules.\\[1em]}%
+It is also interesting what is the characteristic of the sequence designed by each method in a specific example of functional RNA molecules.}%
+
 \AnswerTODO[Yann]{Again, we do not feel the need to validate, or further analyze, the MultiDefect metrics, which has been a standard for multiple structure design.

### 2019-01-24 RNARedPrint.tex
Updates from ShareLaTeX

@@
     match \revised{(respective $R^2$ values of $0.846$ and $0.841$)} enables more efficient targeting of Turner energies based on
-    targeting stacking model energies. \textbf{(B)} Resulting D\&P energy distributions for the two target structures R1 and R2 when aiming for the respective free energies $(-30,-20)$,
+    targeting stacking model energies. \textbf{(B)} \revised{Resulting D\&P energy distributions for the two target structures R1 and R2 when aiming for the respective free energies $(-30,-20)$,
     $(-30,-30)$, $(-25,-25)$, and $(-20,-30)$. These demonstrate the effectivity of our adaptive
-    multi-dimensional Boltzmann sampling procedure, especially by comparing the distributions to those of uniform and Boltzmann sampled sequences, with homogenous weights $1$ and
+    multi-dimensional Boltzmann sampling procedure, especially by} comparing the distributions to those of uniform and Boltzmann sampled sequences, with homogenous weights $1$ and
     $e^\beta$, respectively.
@@
 \Obj{} value of these sequences against that of seed sequences,
-uniformly sampled using \RNAblueprint{}.  Moreover, for both sets, we
+uniformly sampled using \RNAblueprint{}.  Moreover, for both sets we
 used an adaptive greedy walk~\citep{Hammer2017} to \Def{minimize} the
@@
 ({\sf LE80}) to $16.05$ ({\sf 2str}) units. Remarkably, this
-improvement is observed for both of the terms in \Obj{} (see
+improvement is observed for both terms in \Obj{} (see
 Supp. Mat.~\ref{appsec:contributions-multidefect}).  This means that \ourprog{}
-produces sequences whose targets are both substantially closer to
+produces sequences whose targets are both, substantially closer to
 the sequence ensemble/minimal free energy, and more similarly stable across targets.  In fact, for
@@
 
-We introduced a novel approach, based on a general framework and efficient algorithms, to design RNA
-sequences, while targeting very specific complex properties. In particular, we describe the targeting of the free energies of multiple target strucutures and the \GCb-content.
+\revised{Based on a general framework and efficient algorithms, we introduced a novel approach to design RNA
+sequences while targeting very specific complex properties}. In particular, we describe the targeting of the free energies of multiple target strucutures and the \GCb-content.
 Our method combines an FPT stochastic sampling algorithm
@@
 constraints, like \GCb\%, while controlling the energy of
-multiple target structures. Thus it presents a major advance over
+multiple target structures. Thus, it presents a major advance over
 previously applied ad-hoc sampling and even efficient uniform sampling
@@
 Our framework is extensible to include various more complex sequence constraints, including mandatory/forbidden motifs at specific
-positions, or anywhere in the designed sequences, by adapting formal language constructs
+positions or anywhere in the designed sequences, by adapting formal language constructs
 of Zhou \emph{et al}~\citet{Zhou2013}. In future work, negative design principles could 
@@
 Indeed, the complex set of constraints required in order to achieve a desired function must be equally enforced in RNA design, and to produce random sequences that are known to be constrained at a phenotypic level.
-One could therefore use constrained design to assess whether P-values, and therefore our perception of the statistical significance, are affected by the overly simplistic simple null models (\emph{e.g.} dinucleotide shuffling) used to model random RNAs~\cite{Rivas2017}.
+\revised{Therefore, one could} use constrained design to assess whether P-values, and \revised{thus} our perception of the statistical significance, are affected by the overly simplistic simple null models (\emph{e.g.} dinucleotide shuffling) used to model random RNAs~\cite{Rivas2017}.
 

### 2019-01-24 BMC-Response-Round1.tex
Updates from ShareLaTeX

@@
 
-\Answer{As we tried to make clear in our general comments of this response letter (as well as by appropriate adaptions in the manuscript), RNAredprint is not intended to perform negative design.
-It is a generation algorithm, not only addressing a large array of design use-cases, but also able to generate good seed sequences, amenable to further optimization using other tools.
+\Answer{As we tried to make clear in our general comments of this response letter (as well as by appropriate adaptions in the manuscript), RNAredprint does not by itself perform \emph{negative} design.
+It is a generation algorithm for (positively) designing sequences with complex properties.
+In this way, it addresses a large range of design use-cases, but is also able to generate good seed sequences, amenable to further optimization using other tools.
 For this reason, we focused our validation on this potential for better initialization, and compared the quality of sequences generated by RNARedPrint (Boltzmann weighted) to RNABlueprint (uniform).
@@
 
-\AnswerTODO[Yann]{Again, we do not feel the need to validate, or further analyze, the MultiDefect metrics, which has been a standard for multiple structure design.
-The time taken by the computational design is generally negligible compared to the time taken to experimentally validate the ultimate functionality of a design.
-What we believe to matter more to the community is the expressivity of the tool which, in this case, allows to address various use-cases, including some that could not be addressed before.
-}
+\Answer{For reasons already outlined above, we think that a sophisticated analyses of the MultiDefect metrics, which has been a standard for multiple structure design, should not be part of this work.
+Nevertheless, we report and discuss a breakdown of the measure into its two components.
+The final suggestion seems more interesting for negative design, since in positive design, we can directly prescribe sequence characteristics.}
+
 \Comment{(Is RNARedPrint able to produce, for example, the combination of two long stem loops which are high and low GC-content, respectively?)\\[1em]}%
+
 \Answer{In the current implementation, it is only possible to constrain the global GC\% of the sequence.
-It is however possible to support this feature within our general framework, since one can associated individual objectives to structures (eg  structures are already associated with individual target energies).
-Different stems can be expressed as different structures by "peeling the structure", so this feature could be added to a future version, should the concrete need arise.}
+It is however possible to support this feature within our general framework, since one can associated individual objectives to structures (e.g.\ structures are already associated with individual target energies).
+Different stems can be expressed as different structures by "peeling the structure", so this feature could be added to a future version.}
+
 \Comment{
@@
 In particular, it is difficult to comprehend the relationships between the variables due to insufficient descriptions and lack of definitions.
-For example, they do not specify any relationship between R and F in this section while they should not be independent. (e.g., m should be equal or less than k?)
-\\[1em]}%
-\AnswerTODO[Yann]{We attempted to clarify it.
-}
+For example, they do not specify any relationship between R and F in this section while they should not be independent. (e.g., m should be equal or less than k?)}%
+
+\Answer{We understand that the text was still hard to read and thus rewrote large parts of the first subsections in Methods to make it more accessible.}
+
 \Comment{
@@
 Because this manuscript contains many variables and abbreviations, it is recommended to have a table or image to explain the variables and their actual examples following the sampling example in Fig1.
-The description of the brief aim and summary of each step would be a good guide for readers.\\[1em]}%
-\AnswerTODO[Yan]{
-Add method overview
-}
+The description of the brief aim and summary of each step would be a good guide for readers.}%
+\Answer{We included additional explanations and took care to define these variables. We considered the incl further advantages.}
 

### 2019-01-24 RNARedPrint.tex
Updates from ShareLaTeX

@@
 Indeed it implies that, among the $4^k$ possible assignments of nucleotides to $k$ connected positions (in the base-pair dependency graph), at most $2^k$ are compatible with base pairing rules.
-One can thus sharply reduce the complexity of Algorithm~\ref{alg:pf} by restricting the precomputations to compatible assignments.}
+One can thus sharply reduce the complexity of Algorithm~\ref{alg:pf} by restricting the precomputations to compatible assignments.
+
+For a discussion on the implications of our hardness results beyond exact counting see Supp.\ Mat.~\ref{appsec:approximate-counting}.
+}
 
@@
 Fortunately, we found that intricacy of the Turner energy model can be circumvented with minimal loss of precision by using a simpler stacking energy model.
-As can be shown in Figure~\ref{fig:correlation}, a simple stacking energy model, further described in Supp.
-Mat.~\ref{appsec:modelparameters} can be used to approximate the Turner energy model (correlation coefficient $R=0.99$), while greatly reducing the tree width, and thus the computational requirements of the whole method even for complex instances.
-
+As shown in Figure~\ref{fig:correlation}, a simple stacking energy model, further described in Supp.
+Mat.~\ref{appsec:modelparameters}, can be used to well approximate the Turner energy model (correlation coefficient $R=0.99$) for our purpose.
+This greatly reduces the treewidth, and thus the computational requirements of the whole method even for complex instances.
 }
@@
 \revised{
-
 We used our Boltzmann sampling strategy (Algorithms
@@
 %for RNA design.
-Using \ourprog{}, for
-each instance of the benchmark, we generated at least 1\,000 seed sequences with similar
-energies for all target structures. For this purpose, we determine good common target energies that
-can be succesfully targeted for all single target structures simultaneously.
-Generally, we targeted 60\% \GCb\%.  We compared the
-\Obj{} value of these sequences against that of seed sequences,
+Using \revised{multi-dimensional Boltzmann sampling based on} \ourprog{}, for
+each instance of the benchmark, we generated at least 1\,000 seed sequences with similar energies for all target structures. 
+For this purpose, we determine good common target energies that
+can be successfully targeted for all single target structures simultaneously.
+Generally, we targeted 60\% \GCb\%.
+\revised{As described in Section~\ref{sec:multiBoltzmann}, to find sequences fulfilling our energy and \GCb\% criteria, we iteratively sampled 10\,000 sequences and kept admissible ones while optimizing the weights.
+In practice, this sampling operation performs quite fast (see \ref{appsec:run-times}) and only little iterations were needed for most of the benchmark inputs as reported in  \ref{appsec:effectivity-seed-generation}.}
+We compared the
+\Obj{} value of \revised{the derived} sequences against that of seed sequences,
 uniformly sampled using \RNAblueprint{}.  Moreover, for both sets we

### 2019-01-24 BMC-Response-Round1.tex
Updates from ShareLaTeX

@@
 The description of the brief aim and summary of each step would be a good guide for readers.}%
-\Answer{We included additional explanations and took care to define these variables. We considered the incl further advantages.}
+\Answer{We included additional explanations and took care to define all variables (like the sequence $S$). We considered the inclusion of a table, but this seems to offer little further advantages.}
 

### 2019-01-24 RNARedPrint.tex
Updates from ShareLaTeX

@@
   biologically relevant multi-target Boltzmann-weighted designs for a
-  \revised{established} design benchmark. \revised{Our results demonstrate the effectivity and feasiblity of the method in practice
+  \revised{established} design benchmark. \revised{Our results demonstrate the effectivity and feasibility of the method in practice
   as well as the benefits of Boltzmann sampling over the previously best multi-target sampling strategy---even for the case of negative design of multi-stable RNAs.}\\
@@
 As well, the \Software{RNAdesign}/\RNAblueprint{} approach is specialized to uniform sampling, which limits its direct extensibility.
-\revised{Substantial improvements of multi-target sampling thus require a systematically redesigned appraoch.
+\revised{Substantial improvements of multi-target sampling thus require a systematically redesigned approach.
 To enable a fundamentally broader range of applications in extensions of the sampling method, we build our approach, from the start, on established concepts in computer science.}
@@
 Let us consider $m$ (secondary) structures $R_1,\dots, R_m$, each abstracted as a set of base pairs, and \revised{$k\geq m$} features $F_1,\dots, F_k$\revised{, typically representing the energies of the structures and additional sequence properties,} associated with weights $\pi_1,\dots,\pi_k$.
-Our goal is to sample sequences \revised{$S$ (which satisfy the base pairing rules for all strucutures) from the Boltzmann distribution defined by}
+Our goal is to sample sequences \revised{$S$ (which satisfy the base pairing rules for all structures) from the Boltzmann distribution defined by}
 \begin{equation}
@@
   \end{equation}
-  for specfic weights $\pi_1,\dots,\pi_k$.
+  for specific weights $\pi_1,\dots,\pi_k$.
 
@@
 Formally, we define a \Def{feature $F$} as a function on sequences, whose value is obtained by summing over an associated set of \Def{contributions}.
-Each contribution $f$ takes values in $\mathbb{R}\cup \{+\infty\}$, and depends on the nucleotides assigned to a restricted set of positions, its \emph{dependencies $\dep(f)$}, such that
+Each contribution $f$ takes values in $\mathbb{R}\cup \{+\infty\}$, and depends on the nucleotides assigned to a restricted set of positions, namely its \emph{dependencies $\dep(f)$}, such that
 
@@
 
-The $\GCb\%$ can be simply expressed using $n$ contributions $f^{\GCb}_i$, \revised{each dependencing only on position $i\in[1,n]$, i.e.~$\dep(f^{\GCb}_i)=\{i\}$}, such that 
+The $\GCb\%$ can be simply expressed using $n$ contributions $f^{\GCb}_i$, \revised{each depending only on position $i\in[1,n]$, i.e.~$\dep(f^{\GCb}_i)=\{i\}$}, such that 
 $$f^{\GCb}_i(\{i\mapsto c\})=\begin{cases}-1&\text{ if }c=\Gb\text{ or }\Cb\\ 0&\text{ otherwise}.\end{cases}$$
@@
 %
-\revised{Intuitively, a tree decomposition of an (hyper)graph $G$ is a tree that captures all the vertices and (hyper)edges of $G$, and properly relates dependend subproblems to ensure consistency in a recursive computation.
+\revised{Intuitively, a tree decomposition of an (hyper)graph $G$ is a tree that captures all the vertices and (hyper)edges of $G$, and properly relates dependent sub-problems to ensure consistency in a recursive computation.
 Figure~\ref{fig:treedec} shows an optimal tree decomposition for a pair of structures under the stacking energy model.}
@@
 \item
-all edges in the tree decompostion are oriented towards this root;
+all edges in the tree decomposition are oriented towards this root;
 \item
@@
 $(\gamma,\delta)$ parameters from a set of sequences generated with
-homogenous weights $w=e^{\beta}$, tuning only \GCb\% to a
+homogeneous weights $w=e^{\beta}$, tuning only \GCb\% to a
 predetermined value.  Finally, we adjusted the targeted energies
@@
 The flexibility of our framework allows to support the advanced sampling technique called "multidimensional Boltzmann sampling"~\citep{Bodini2010}, which allows to enforce (probabilistically)  additional, complex properties of the samples through an additional rejection.
-This technique was previously used to control \GCb\%~\citep{Waldispuehl2011,Reinharz2013} and dinucleotide content~\citep{Zhang2013} of sampled RNA sequences.
+This technique was previously used to control \GCb\%~\citep{Waldispuehl2011,Reinharz2013} and di-nucleotide content~\citep{Zhang2013} of sampled RNA sequences.
 Here, in addition to controlling \GCb\% (our feature $F_0$) we use it to target the free energies $(\TargetE_1,\ldots,\TargetE_k)$ of the individual target structures (features $F_1,\dots,F_m$).
@@
 
-\subsection*{Effectively targeting Turner energies using multidimensionnal sampling}
+\subsection*{Effectively targeting Turner energies using multi-dimensional sampling}
 \revised{
@@
 For the next iteration, the weights are updated by the rules $w_\ell = w_\ell \gamma^{mean(E(S;R_\ell))-E_\ell}$ and $w_{gc} = w_{gc}\frac{GC_{target}}{mean(GC)}$.
-During this procedure, we collect   in a set $P$ all the sequences exhibiting energies within $[E_1-\delta,E_1+\delta]\times [E_2-\delta,E_2+\delta]\ldots$ and and GC-content within $[GC_{target}-\idelta',GC_{target}+\delta']$.
-%If $|P| > n$ or an maximum amount of iterations is reached, the sequences in $P$ are returned.
+During this procedure, we collect   in a set $P$ all the sequences exhibiting energies within $[E_1-\delta,E_1+\delta]\times [E_2-\delta,E_2+\delta]\ldots$ and GC-content within $[GC_{target}-\idelta',GC_{target}+\delta']$.
+%If $|P| > n$ or a maximum amount of iterations is reached, the sequences in $P$ are returned.
 %
@@
     energies in the stacking model to the realistic pseudoknot energy
-    model by Dirks and Pierce (D\&P) for \revised{initially} sampled sequences and both target structures R1 and R2 (shown in subfigure B). The good
+    model by Dirks and Pierce (D\&P) for \revised{initially} sampled sequences and both target structures R1 and R2 (shown in B). The good
     match \revised{(respective $R^2$ values of $0.846$ and $0.841$)} enables more efficient targeting of Turner energies based on
@@
     $(-30,-30)$, $(-25,-25)$, and $(-20,-30)$. These demonstrate the effectivity of our adaptive
-    multi-dimensional Boltzmann sampling procedure, especially by} comparing the distributions to those of uniform and Boltzmann sampled sequences, with homogenous weights $1$ and
+    multi-dimensional Boltzmann sampling procedure, especially by} comparing the distributions to those of uniform and Boltzmann sampled sequences, with homogeneous weights $1$ and
     $e^\beta$, respectively.
@@
 %for RNA design.
-Using \revised{multi-dimensional Boltzmann sampling based on} \ourprog{}, for
-each instance of the benchmark, we generated at least 1\,000 seed sequences with similar energies for all target structures. 
+\revised{Based on} \ourprog{}, we generated at least 1\,000 seed sequences with similar energies for all target structures, for
+each instance of the benchmark. 
 For this purpose, we determine good common target energies that
@@
 Generally, we targeted 60\% \GCb\%.
-\revised{As described in Section~\ref{sec:multiBoltzmann}, to find sequences fulfilling our energy and \GCb\% criteria, we iteratively sampled 10\,000 sequences and kept admissible ones while optimizing the weights.
-In practice, this sampling operation performs quite fast (see \ref{appsec:run-times}) and only little iterations were needed for most of the benchmark inputs as reported in  \ref{appsec:effectivity-seed-generation}.}
+\revised{%
+Supp.\ Mat.~\ref{appsec:effectivity-seed-generation}, provides detailed results from this iterative procedure, which works similarily to the previously described multi-dimensional Boltzmann sampling.
+In particular, we observe that most of the benchmark inputs finish within only few iterations, where each iteration requires  little time (confer Supp.\ Mat.~\ref{appsec:run-times}).}
+
 We compared the
@@
 used an adaptive greedy walk~\citep{Hammer2017} to \Def{minimize} the
-\Obj{} function. At each step, the local search resamples (uniformly
+\Obj{} function. At each step, the local search re-samples (uniformly
 at random) the positions of a randomly selected component in the base
@@
 in a gain. We performed 500 greedy descent steps in the case of
-pseudoknot-free datasets {\sf 2str}, {\sf 3str}, and {\sf 4str}; and
+pseudoknot-free data-sets {\sf 2str}, {\sf 3str}, and {\sf 4str}; and
 200 steps for the pseudoknotted ones {\sf PK60}, {\sf PK80}, and {\sf
@@
 The results, shown in Fig.~\ref{fig:benchmark-results}, reveal that
-Boltzmann-sampled sequences outperform uniform seeds on every dataset,
+Boltzmann-sampled sequences outperform uniform seeds on every data-set,
 leading to average improvements in \Obj{} values ranging from $7.26$
@@
 \revised{Based on a general framework and efficient algorithms, we introduced a novel approach to design RNA
-sequences while targeting very specific complex properties}. In particular, we describe the targeting of the free energies of multiple target strucutures and the \GCb-content.
+sequences while targeting very specific complex properties}. In particular, we describe the targeting of the free energies of multiple target structures and the \GCb-content.
 Our method combines an FPT stochastic sampling algorithm

### 2019-01-24 supplement.tex
Updates from ShareLaTeX

@@
 For each of two small examples, a single-stem and a double-stem structure, we produce a uniform and a weighted Boltzmann
-sampling, each of 10000 sequenes. For each sampled sequence $S$, we determine
+sampling, each of 10\,000 sequences. For each sampled sequence $S$, we determine
 the Turner energy $E(S,R)$ as well as the
@@
 To illustrate the efficiency and practical feasibility of our Boltzmann sampling approach, we show the distributions of run times (Intel(R) Core(TM) i7-4770 CPU) for generating 10000 samples (over the Modena benchmark instances and grouped by benchmark sub-set).
-Since producing a controlled Boltzmann sampling requires significant pre-computation, it is informative to separately show construction and sampling times (Figure~\ref{appfig:run-times-redprint}).
+Since producing a controlled Boltzmann sampling requires significant precomputation, it is informative to separately show construction and sampling times (Figure~\ref{appfig:run-times-redprint}).
 Moreover, \ourprog{} can be used with a simpler base pair model or a rather complex stacking energy model, which introduces much stronger dependencies than required for uniform sampling.
-Technically, these stronger dependencies increase the tree width and thus the runtime, as noticeable in Figure~\ref{appfig:run-times-redprint}.
+Technically, these stronger dependencies increase the tree width and thus the run-time, as noticeable in Figure~\ref{appfig:run-times-redprint}.
 
@@
 The times are split into construction time (left) and sampling time (right).
-Construction includes all the pre-computation, while the latter is the time needed to generate 10000 sequences.
+Construction includes all the precomputation, while the latter is the time needed to generate 10000 sequences.
 }
@@
 The times are split into construction time (left) and sampling time (right).
-Construction includes all the pre-computation, while the latter is the time needed to generate 10000 sequences.}
+Construction includes all the precomputation, while the latter is the time needed to generate 10000 sequences.}
   \label{appfig:run-times-rbp}
@@
     comparison, the figure shows distributions from uniform and Boltzmann
-    sampling at respective homogenous weights $1$ and $e^\beta$.
+    sampling at respective homogeneous weights $1$ and $e^\beta$.
 
@@
         $(-20,-20,-20)$ for the target structures (annotated in red and
-        blue, resepectively), we demonstrate the
+        blue, respectively), we demonstrate the
         effectivity of our adaptive multi-dimensional Boltzmann sampling
@@
         (U) and Boltzmann (B) samples---respectively associated with
-        homogenous weights $1$ and $e^\beta$.
+        homogeneous weights $1$ and $e^\beta$.
 %    ($\delta=0.05, \delta'=0.1, \gamma=1.1, \psi=20, n=1000$)
@@
     %be even and vice versa, so the choice of a parity for any single
-    %vertex caracterizes the parity of all vertices in a bipartite
+    %vertex characterizes the parity of all vertices in a bipartite
     %connected subgraph.

### 2019-01-24 BMC-Response-Round1.tex
Tons of stuffs

@@
 
-We thank the reviewers for their encouraging comments and constructive criticisms and thoroughly addressed their points in our revised version of the manuscript.
+We thank the reviewers for their encouraging comments and constructive criticisms and thoroughly addressed their points in this revised version of our manuscript.
 
@@
 We discuss this positive design, in particular, for specific folding energies (for each single target structure) and GC content.
-Moreover, as we like to emphasize, the method is even more general, which we see as most important strength of our approach; it is prepared to handle many additional constraints and sequence features based on the same mechanism.
+Moreover, as we like to emphasize, the method is even more general, which we see as a very important strength of our approach; it is prepared to handle many additional constraints and sequence features, using the same mechanism, stemming from practical use cases.
 
 Naturally, we hope that this method turns out to be highly useful for single- and multi-target negative design (and various other applications, like generating background models for assessing significance) in future applications, but---while we present preliminary results for negative design (based on generating seed sequences for the MODENA benchmark)---a more in-depth evaluation of its use in negative design is beyond the scope of this paper.
-This seems even more acceptable, since the principal benefits of controlled sampling have been shown with RNABlueprint.
+This seems even more acceptable, since the principal benefits of controlled sampling  have been demonstrated by RNABlueprint.
 We apologize for potential misunderstandings about the objectives of the manuscript and hope to better clarify these aspects in our revision.
@@
 
-\Comment{For evaluation of a method to design RNA molecules, the number of the solutions it produces and its running time are considered to be important.\\[1em]
+\Comment{For evaluation of a method to design RNA molecules, the number of the solutions it produces and its running time are considered to be important.
 In the scope of this study, the running time and memory comparison should be shown to produce the same number of sequences by some minor variants of RNARedPrint method, RNABluePrint, and MODENA.}
@@
 It would be informative if the authors show the number of rejected sequences produced by RNARedPrint considering the specific criteria (GC\% or free energy). 
-Since RNARedPrint keeps the sequences to be under the specific constraint by a simple rejection method, such information for various target conditions would be better to estimate the practical running time of RNARedPrint.\\[1em]}%
+Since RNARedPrint keeps the sequences to be under the specific constraint by a simple rejection method, such information for various target conditions would be better to estimate the practical running time of RNARedPrint.}%
+
 \Answer{
@@
 \Comment{
-Another is the quality of their solutions in terms of the stability of each target structure.\\[1em]}%
+Another is the quality of their solutions in terms of the stability of each target structure.}%
+
 \Answer{We were unsure how to interpret this remark: RNARedPrint enforces an explicit control on target free-energies through multidimensional Boltzmann sampling, so it cannot be used as an external validation of the quality of our output.
@@
 According to their results, adaptive sampling has a strong power to improve the results. Unfortunately, the authors only introduce the existence of weights first in the first section of methods.
-It would be better that the authors introduce the concept of weights at the beginning (and whether they are completely same with those in RNAblueprint or not).\\[1em]}%
+It would be better that the authors introduce the concept of weights at the beginning (and whether they are completely same with those in RNAblueprint or not).}%
+
 \Answer{
@@
 Moreover, the number of base pairs in target structures should be considered since such approximation might be critical when the long loops appear in the target structures.
-\\[1em]}%
+}%
+
 \Answer{
@@
 $R^2$ values and $p$-values are also necessary to evaluate the correlation between two energy models. (What are the lines in Fig2A representing?)
-\\[1em]}%
+}
+
 \Answer{
@@
 }
+
 \Comment{
@@
 Multi defect can be a good measure to understand the overall performance of the sequence designing method, but there is little explanation why Multidefect is enough to evaluate the quality of sequences for the multi (target-structure) objective problem in a practical use case such as Riboswitch design.
-\\[1em]}%
+}%
+
 \Answer{We agree that MultiDefect is simply \emph{one} reasonable measure for the performance of multi-target design.
@@
 For that reason, RNARedPrint still should be compared with MODENA (in the same running time) using Multidefect and the criteria used in the previous studies such as $\delta$e1 and $\delta$e2.
-Also, RNARedprint and RNABlueprint are optimised to increase only a Multidefect value.\\[1em]}%
+Also, RNARedprint and RNABlueprint are optimised to increase only a Multidefect value.}%
 
-\Answer{As we tried to make clear in our general comments of this response letter (as well as by appropriate adaptions in the manuscript), RNAredprint does not by itself perform \emph{negative} design.
+\Answer{As we tried to clarify in our general comments of this response letter (as well as by appropriate adaptions in the manuscript), RNAredprint does not by itself perform \emph{negative} design.
 It is a generation algorithm for (positively) designing sequences with complex properties.
@@
 
-\Comment{(Is RNARedPrint able to produce, for example, the combination of two long stem loops which are high and low GC-content, respectively?)\\[1em]}%
+\Comment{(Is RNARedPrint able to produce, for example, the combination of two long stem loops which are high and low GC-content, respectively?)}%
 
@@
 The description of the brief aim and summary of each step would be a good guide for readers.}%
-\Answer{We included additional explanations and took care to define all variables (like the sequence $S$). We considered the inclusion of a table, but this seems to offer little further advantages.}
+
+\Answer{We included additional explanations and took care to define all variables (like the sequence $S$, the dependencies dep(f), etc).
+We considered the inclusion of a table, but this seems to offer little further advantages.}
 
@@
 \Comment{
-* Ensemble free energy: might be an ambiguous definition\\[1em]}%
+* Ensemble free energy: might be an ambiguous definition}
+
 \Answer{The concept of ensemble free-energy is a well-defined concept in statistical mechanics, and is used here in a way that, to the best of our knowledge, matches its usual definition.}
+
 \Comment{
-* While $\chi$ is a subset of nodes $X$, is a subset of $\chi$ an edge? (P7 L19)\\[1em]}%
-\AnswerTODO{
- $\chi(v)$ is a subset of the positions in the RNA sequence.
-For an edge/base pair to be included in this set simply means that both of its edges are part of that set.
-It is crucial that all edges/base pairs are captured by at least one node $v$, so that the sequence compatibility can be ensured.
-}
-\Comment[Sebastian :)]{
-* Is the output of algorithm 2 only one sequence? It is unclear when two (or more?) sequences are merged by $S \cup S'$.\\[1em]}%
-\AnswerTODO{Indeed. In order to simplify the notations, we use $S$ and $S'$ to mean partial assignments, which can be merged through an union.
+* While $\chi$ is a subset of nodes $X$, is a subset of $\chi$ an edge? (P7 L19)}
+
+\Answer{
+ Tree decompositions are now much better explained in the text and illustrated in Fig. 2. 
+ In the definition of tree decomposition, $\chi(v)$ is the subset of nodes in the hypergraph that is assigned to the tree node $v$.
+The condition 3 of this definition requires that every hyperedge of the dependency graph is a subset of some $\chi(v).$
 }
 \Comment{
-Is it correct no break at P10L23?\\[1em]}%
-\AnswerTODO{
-??
+* Is the output of algorithm 2 only one sequence? It is unclear when two (or more?) sequences are merged by $S \cup S'$.}%
+
+\Answer{Yes, the output is only one sequence.
+The algorithm is called repeatedly to produce an arbitrary number of sequences.
+In order to simplify the notations, we use $S$ and $S'$ to mean partial assignments, which can be merged through an union.
 }
+
+\Comment{Is it correct no break at P10L23?}%
+
+\Answer{Sorry, but we don't understand the hint.}
+
 \Comment{
-* In Fig2, the notation like $(a, b)$ is confusing with the range (of free energies or something).\\[1em]}%
-\AnswerTODO{
+* In Fig2, the notation like $(a, b)$ is confusing with the range (of free energies or something).}%
+
+\Answer{
+Thanks for pointing out; we have changed this.
 }
+
 \Comment{
 * Unfortunately, I could not run RNARedPrint (while it can be just compiled when I make bin directory).
-More detailed documentation is required such as OS dependency or the version information the authors confirmed.\\[1em]}%
-\AnswerTODO[Yann]{
-Yann will check; mention OS dependencies---not checked on MacOS. Merge pyscript branch into master branch!
+More detailed documentation is required such as OS dependency or the version information the authors confirmed.}%
+
+\Answer{
+Thanks for reporting this issue. We could resolve some problems and re-checked the installation. Moreover, we improved the documentation. 
+%Yann will check; mention OS dependencies---not checked on MacOS. Merge pyscript branch into master branch!
 }
+
 \Comment{
 * It seems not trivial that Boltzmann sampling works better to control GC contents, but the stability of structures.
-What's the reason the authors include GC contents for the criteria of Boltzmann-weighted sampling?\\[1em]}%
+What's the reason the authors include GC contents for the criteria of Boltzmann-weighted sampling?}%
+
 \Answer{GC\%, just like the free-energy of any given target structure, can be expressed uniformly in our framework in a way that is additive on base pairs.
 Control over GC\% has recurrently been identified as critical, since a good stability usually implies high GC\%, yet actual sequences usually combine good stability with reasonable GC\%.
-Modern computational objectives of design are to ensure good stability (+specificity) while producing sequences of prescribed GC\%\\[1em]}%
-\AnswerTODO{
-}
-\Comment{
-* The order of authors is different between the main manuscript and supplementary (not sure whether they used just an alphabetical order).\\[1em]}%
-\Answer{Thank you for spotting that issue, devoid of any deeper significance.
-}
+Modern computational objectives of design thus usually ensure good stability (+specificity) while producing sequences of prescribed GC\%.
+Accordingly, our benchmark shows that RNARedPrint generally manages to decorrelate those quantities.}
+
+\Comment{* The order of authors is different between the main manuscript and supplementary (not sure whether they used just an alphabetical order).}%
+
+\Answer{Thank you for pointing this out.}
+
 \Comment{
 * Is the section "\# P-hardness of counting valid designs" necessary in the main script (not supplementary) to explain the results?
-The complexity of the problem is important to consider theoretical limitations, but the assumption in that section seems not to be necessary since the authors applied a greedy sampling strategy and not in the scope of this study.\\[1em]}%
+The complexity of the problem is important to consider theoretical limitations, but the assumption in that section seems not to be necessary since the authors applied a greedy sampling strategy and not in the scope of this study.}
+
 \Answer{We respectfully disagree with that assessment.
+This result is directly relevant to the strategy that we propose in the paper.
+Without showing the \#P-hardness of the design counting problem, devising an algorithm that is efficient only for fixed treewidth would be unjustified.
+
 The core of our approach consists in generating, uniformly at random (resp.
@@
 }
-\AnswerTODO{Revamp intro for \#P result + maybe overall intro}
+
 \Comment{
-* Move the first section in the result to the method.\\[1em]}%
-\AnswerTODO[Yann]{
-Ok (but include efficiency of multidim sampling, keeping correlation analysis) TODO:move parts of the Subsection: Efficiently targeting... to Methods.
+* Move the first section in the result to the method.}%
+
+\Answer{
+In this revised version, we have largely restructured the Methods and Results section, including the suggested modification.
 }
@@
 \Comment{
-1) The "Methods" section is very long, it should be shortened.\\[1em]}%
+1) The "Methods" section is very long, it should be shortened.}%
+
 \Answer{
-As this reviewer could probably notice, the techniques used in this paper are quite divers, and stem from several subfields of computer science (graph algorithms, enumerative/analytic combinatorics...).
+As this reviewer could probably notice, the techniques used in this paper are quite diverse, and stem from several subfields of computer science (combinatorial optimization, graph algorithms, enumerative/analytic combinatorics...).
 For this reason, a more concise description of the method would be poorly self-contained, possibly to the point of irreproducibility.
 However, we share the concern of this reviewer regarding an imbalanced manuscript, and moved less original/central parts of the method to the appendix.
-This includes the expression of various energy models as constraints Turner, and the description of our naive numerical weight optimization.
+This includes the expression of various energy models as constraints, and the description of our naive numerical weight optimization.
 }
-\AnswerTODO[Yann]{
-Move material to appendix
-Better motivations for \#P-hardness
-New results : 
-  1) Validation of stacking approximation: Correlation coefficient of Stacking and Turner (+ goodness of fit)
-  2) Tree width and Time benchmark
-  3) Rejection analysis for multidim sampling
-  4) Proof-of-concept -Potential of better seeds
+%\AnswerTODO[Yann]{
+%Move material to appendix
+%Better motivations for \#P-hardness
+%New results : 
+%  1) Validation of stacking approximation: Correlation coefficient of Stacking and Turner (+ %goodness of fit)
+%  2) Tree width and Time benchmark
+%  3) Rejection analysis for multidim sampling
+%  4) Proof-of-concept -Potential of better seeds
+%}
+\Comment{
+2) Although, it was shown previously, that RNABluePrint is a leading tool for multiple design, it would be valuable to include and compare the results generated by the other tools, e.g.
+Frnakenstein.}%
+
+\Answer{Unfortunately, we could no longer locate the implementation of FRNAkenstein.
+Moreover, in the context of positive design, the strategies of MODENA and FRNAkenstein are virtually identical.
+At the time of writing this manuscript, we could not locate any more software that explicitly address the multiple design of RNAs.
+We hope the additional sections in the results section will contribute to reassure the reviewer of the quality and efficacy of our method.}
+\Comment{
+3) Please include computational times of the performed designs with the given benchmark.}
+
+\Answer{
+We understand that it is important to give an estimate of the actual performance and thus included computational times for the pre-computation, the Boltzmann sampling of 10\,000 sequences and the number of iterations for the multi-dimensional Boltzmann sampling for the complete benchmark set in Supl.
+Mat.
+Sections C and E.
 }
 \Comment{
-2) Although, it was shown previously, that RNABluePrint is a leading tool for multiple design, it would be valuable to include and compare the results generated by the other tools, e.g.
-Frnakenstein.\\[1em]}%
-\AnswerTODO[Stefan+Sebastian]{Unfortunately, we could no longer locate the implementation of FRNAkenstein.
-Moreover, in the context of positive, the strategies of MODENA and FRNAkenstein are virtually identical.}
-\Comment{
-3) Please include computational times of the performed designs with the given benchmark.\\[1em]}%
-\AnswerTODO[Stefan]{
-Will do
-}
-\Comment{
-4) The paper should be checked for English grammar and wording.\\[1em]}%
-\AnswerTODO{
-Will do
+4) The paper should be checked for English grammar and wording.}%
+
+\Answer{
+We have addressed this issue and hope to have improved the quality of the write up.
 }
@@
 In Figure 3 of Supplementary Material: more detailed description for explaining the two examples of tree decomposition is appreciated, particularly for the readers who are not familiar with tree decomposition; e.g.
-why these two examples are output, and the other examples for RF3 exist or not?\\[1em]}%
+why these two examples are output, and the other examples for RF3 exist or not?}%
+
 \Answer{
-  We addressed this comment in two ways: First, we added an intuitive explanation of tree decomposition i(after its formal definition) together with an illustration of a simple tree decomposition to the main text.
-  Second, we rewrote the corresponding text in the supplement to further clarify the purpose of the figure and provide additional explanations.
+  We addressed this comment in two ways: Firstly, we added an intuitive explanation of tree decomposition (after its formal definition) together with an illustration of a simple tree decomposition to the main text.
+  Secondly, we rewrote the corresponding text in the supplement to further clarify the purpose of the figure and provide additional explanations.
 }
+
 \Comment{
@@
 Since the Boltzmann sampling in RNARedPrint utilizes the rejection approach, I would like to know the difference between the running times of the Boltzmann sampling and the uniform sampling.
-\\[1em]}%
+}%
 \Answer{We were unsure about what was meant in this comment. Just to be precise, we distinguish the term 'Boltzmann sampling' from 'multi-dimensional Bolzmann sampling' in our paper. 
@@
 There is, of course, an added cost associated to this rejection, and the simultaneous weight optimization, which we discuss further in a new "Result" subsection, where we also report the exact time differences.}
+
 \Comment{
-In addition, I think it is better that the N/A result is indicated not by a blank but by a "N/A" in the table.\\[1em]}%
+In addition, I think it is better that the N/A result is indicated not by a blank but by a "N/A" in the table.}%
+
 \Answer{We used N/A in the table.}

### 2019-01-24 RNARedPrint.tex
Tons of stuffs

@@
 Formally, we define a \Def{feature $F$} as a function on sequences, whose value is obtained by summing over an associated set of \Def{contributions}.
-Each contribution $f$ takes values in $\mathbb{R}\cup \{+\infty\}$, and depends on the nucleotides assigned to a restricted set of positions, namely its \emph{dependencies $\dep(f)$}, such that
+Each contribution $f$ takes values in $\mathbb{R}\cup \{+\infty\}$, and depends on the nucleotides assigned to a restricted set of positions, \revised{namely its \emph{dependencies}, denoted $\dep(f)$}, such that
 
@@
     match \revised{(respective $R^2$ values of $0.846$ and $0.841$)} enables more efficient targeting of Turner energies based on
-    targeting stacking model energies. \textbf{(B)} \revised{Resulting D\&P energy distributions for the two target structures R1 and R2 when aiming for the respective free energies $(-30,-20)$,
-    $(-30,-30)$, $(-25,-25)$, and $(-20,-30)$. These demonstrate the effectivity of our adaptive
+    targeting stacking model energies. \textbf{(B)} \revised{Resulting D\&P energy distributions for the two target structures R1 and R2 when aiming for the respective free energies $-30$ and $-20$,
+    $-30$ and $-30$, $-25$ and $-25$, $-20$ and $-30$\,kcal/mol. These demonstrate the effectivity of our adaptive
     multi-dimensional Boltzmann sampling procedure, especially by} comparing the distributions to those of uniform and Boltzmann sampled sequences, with homogeneous weights $1$ and
@@
 \revised{%
-Supp.\ Mat.~\ref{appsec:effectivity-seed-generation}, provides detailed results from this iterative procedure, which works similarily to the previously described multi-dimensional Boltzmann sampling.
+Supp.\ Mat.~\ref{appsec:effectivity-seed-generation} provides detailed results from this iterative procedure, which works similarly to the previously described multi-dimensional Boltzmann sampling.
 In particular, we observe that most of the benchmark inputs finish within only few iterations, where each iteration requires  little time (confer Supp.\ Mat.~\ref{appsec:run-times}).}

### 2019-01-24 supplement.tex
Tons of stuffs

@@
         sampling. We show the Turner energy distributions for
-        three target structures R1, R2 and R3. By
-        targeting the respective free energies $(-40,-40,-20)$ and
-        $(-20,-20,-20)$ for the target structures (annotated in red and
+        three target structures R1, R2 and R3. \revised{By
+        targeting the respective free energies ($-40$, $-40$ and $-20$\,kcal/mol and $-20$,$-20$ and $-20$\,kcal/mol) for the target structures (annotated in red and
         blue, respectively), we demonstrate the
         effectivity of our adaptive multi-dimensional Boltzmann sampling
-        procedure; moreover, for comparison, the distributions for uniform
+        procedure. Moreover, }for comparison, the distributions for uniform
         (U) and Boltzmann (B) samples---respectively associated with

### 2019-01-25 BMC-Response-Round1.tex
Updates from ShareLaTeX

@@
 \Answer{
-  We thank the reviewer for the criticism, which motivated us to work on clarifying the practical use of the method and tool RNARedPrint in several ways; we have carefully edited the entire manuscript having these issues in mind.
+  We thank the reviewer for voicing her concerns, which motivated us to work on clarifying the practical use of the method/tool RNARedPrint in several ways; we have carefully edited the entire manuscript having these issues in mind.
   In the methods section, we tried to improve the general readability especially of our overview of the method and central problem.
@@
 \Answer{
-Weights are used for two different purposes:
-\begin{itemize}
-    \item In order to induce a Boltzmann distribution on the set of sequences;
-    \item As a technical artifact, to ensure the efficiency of a rejection-based strategy to control auxiliary features (free-energies for structures, GC\%\ldots)
-\end{itemize}
+Weights are used for two different purposes: 1) In order to induce a Boltzmann distribution on the set of sequences.
+2) As a technical artifact, to ensure the efficiency of a rejection-based strategy to control auxiliary features (free-energies for structures, GC\%\ldots).
+%
 In both cases, the use of weights is rather technical, which explains why weights are mentioned in the methods section.
@@
 
-Partially, we have problems to understand the question, since RNAblueprint does not use weights at all. We hope that our improvements to the text help to clarify.
-}
+We are not, however, quite certain to correctly understand the question, since RNAblueprint does not use weights at all.
+We hope that our improved description of the method helps clarify what we believe to be a misunderstanding.}
 
@@
 That being said, we used 1 as initial weights.
-In the example of Figure 3, in the case of targeting respective energies of -20 and -30 kcal/mol for the two target structures, we arrive at respecticve weights of 229 and 117 in the last iteration.
+In the example of Figure 3, in the case of targeting respective energies of -20 and -30 kcal/mol for the two target structures, we arrive at respective weights of 229 and 117 at the last iteration.
 The entire procedure to generate 5000 samples targeted at these energies took 6m20s (CPU i7-4770).
@@
 
-Some additional more technical remarks:
-The calibration of weights relies on a numerical iteration aiming at computing weights such that expected properties of sampled sequences (which depend on weights) match their targeted values.
+%Additional technical remarks: The calibration of weights relies on a numerical iteration aiming at computing weights such that expected properties of sampled sequences (which depend on weights) match their targeted values.
 Associated problems are usually convex, and we describe in supplementary material a numerical iteration to compute suitable weights.
@@
 
-To elaborate: the MultiDefect objective function is based on the rationale that target structures should achieve high stability, and that the chosen RNA cannot adopt alternative, more stable, structures.
+%To elaborate: the MultiDefect objective function is based on the rationale that target structures should achieve high stability, and that the chosen RNA cannot adopt alternative, more stable, structures.
 It is a well accepted metrics introduced by other contributors, which we did not feel necessary to question or further improve.
@@
 This result is directly relevant to the strategy that we propose in the paper.
-Without showing the \#P-hardness of the design counting problem, devising an algorithm that is efficient only for fixed treewidth would be unjustified.
+Without showing the \#P-hardness of the RNA design counting problem, one could not understand that it makes sense to devise an algorithm that is efficient only for fixed treewidth.
+We now provide more explanations in the main text that clarify the relevance of this proof.
 
-The core of our approach consists in generating, uniformly at random (resp.
+%The core of our approach consists in generating, uniformly at random (resp.
 in a Boltzmann distribution, which identifies with the uniform distribution when $T\to +\infty$).
@@
 \Comment{
-* Move the first section in the result to the method.}%
+* Move the first section in the result to the method.}
 
@@
 \end{enumerate}
+
 \subsection{Reviewer \#2}
@@
 Since the Boltzmann sampling in RNARedPrint utilizes the rejection approach, I would like to know the difference between the running times of the Boltzmann sampling and the uniform sampling.
-}%
+}
+
 \Answer{We were unsure about what was meant in this comment. Just to be precise, we distinguish the term 'Boltzmann sampling' from 'multi-dimensional Bolzmann sampling' in our paper. 

### 2019-01-25 BMC-Response-Round1.tex
sync

@@
 In particular, the reviews helped us to state the goal of this work more clearly and improve its presentation.
-While we demonstrate the applicability of our approach to negative RNA design, which  aims at high and similar probability of the target structures (here measured as 'MultiDefect'), the focus of this work is the positive design of sequences that fold into multiple target structures with highly specific properties.
+While we demonstrate the principle \emph{applicability} of our approach to \emph{negative} RNA design, which  aims at high and similar probability of the target structures (here measured as 'MultiDefect'), the \emph{focus} of this work is the \emph{positive} design of sequences that fold into multiple target structures with highly specific properties.
 We discuss this positive design, in particular, for specific folding energies (for each single target structure) and GC content.
-Moreover, as we like to emphasize, the method is even more general, which we see as a very important strength of our approach; it is prepared to handle many additional constraints and sequence features, using the same mechanism, stemming from practical use cases.
+Moreover, as we like to emphasize, the method is even more general, which we see as a very important strength of our approach; it is prepared to handle many additional constraints and sequence features, like they often are required in practical use cases, using the same mechanism.
 
-Naturally, we hope that this method turns out to be highly useful for single- and multi-target negative design (and various other applications, like generating background models for assessing significance) in future applications, but---while we present preliminary results for negative design (based on generating seed sequences for the MODENA benchmark)---a more in-depth evaluation of its use in negative design is beyond the scope of this paper.
-This seems even more acceptable, since the principal benefits of controlled sampling  have been demonstrated by RNABlueprint.
+Naturally, we hope that the method turns out to be highly useful for single- and multi-target negative design (and various other applications, like generating background models for assessing significance) in future applications, but---while we present preliminary results for negative design (based on generating seed sequences for the MODENA benchmark)---a more in-depth evaluation of its use in negative design is beyond the scope of this paper.
+This seems even more reasonable, since the principal benefits of controlled sampling  have been demonstrated by RNABlueprint.
 We apologize for potential misunderstandings about the objectives of the manuscript and hope to better clarify these aspects in our revision.
@@
     \item Boltzmann sampling of positively designed sequences for multiple target structures works effectively and accurately, where the full-featured RNA energy model (Turner) can be well approximated.
-      For this approximation, we use the much more efficiently tractable stacking model; in this model, controlled Boltzmann sampling for typical multi-target design instances requires little time (corresponding to small tree-width);
-    \item Multi-dimensional Boltzmann sampling solves the positive design problem effectively, requiring only a limited number of rejections (even for multiple simultaneously considered features);
-    \item Our method can be beneficially used in the setting of negative design. Here, we demonstrate a typical use case of the method. Moreover, we empirically show that 
-      the positively designed sequences can improve negative design approaches, where
-      they work better than previous approaches (for seed generation).
+      For this approximation, we use the much better tractable stacking energy model; in this model, controlled Boltzmann sampling for typical multi-target design instances requires little time (corresponding to small treewidth);
+    \item Multi-dimensional Boltzmann sampling solves the positive design problem effectively, requiring only a limited number of iterations (and therefore rejections)---even for multiple simultaneously considered features;
+    \item Our method can be beneficially used in the setting of negative design.
+Here, we demonstrate a typical use case of the method.
+By detailed analysis of this procedure, we demonstrate the feasibility of the approach for practically relevant design instances.
+Moreover, we empirically show that the positively designed sequences are suitable to improve negative design approaches, where they yield better performance than seeds generated by previous approaches.
 \end{itemize}}
@@
 While comparing to MODENA in the same way as to RNABluePrint can admittedly appear obvious, the sampling in MODENA is of a very different nature, such that---in our opinion---elaborate comparisons would not provide relevant insights.
-Of course, this has to be explained:
+Of course, this has to be explained in more detail:
 
@@
 \Answer{
-As this reviewer could probably notice, the techniques used in this paper are quite diverse, and stem from several subfields of computer science (combinatorial optimization, graph algorithms, enumerative/analytic combinatorics...).
-For this reason, a more concise description of the method would be poorly self-contained, possibly to the point of irreproducibility.
-However, we share the concern of this reviewer regarding an imbalanced manuscript, and moved less original/central parts of the method to the appendix.
+We share the concern of this reviewer that an unncecessarily imbalanced manuscript should be avoided, and moved less original/central parts of the method to the appendix.
 This includes the expression of various energy models as constraints, and the description of our naive numerical weight optimization.
+However, we want to emphasize that the techniques used in this paper are quite diverse, and stem from several subfields of computer science (combinatorial optimization, graph algorithms, enumerative/analytic combinatorics...).
+For this reason, we feel that the method part cannot be substantially shortened beyond the current point without making the text less accessible and poorly self-contained.
 }
@@
 
-\Answer{We were unsure about what was meant in this comment. Just to be precise, we distinguish the term 'Boltzmann sampling' from 'multi-dimensional Bolzmann sampling' in our paper. 
+\Answer{We were not completely sure what was meant by this comment. Just to be precise, we distinguish the term 'Boltzmann sampling' from 'multi-dimensional Bolzmann sampling' in our paper. 
 On the one hand, Boltzmann sampling, i.e.
-drawing sequences from the Boltzmann distribution does not involve any rejection, and is actually exactly as costly as drawing them uniformly.
-In such a uniform mode, RNARedPrint performs equally as fast as RNABluePrint, and usually faster, since the graph decomposition used in RNABluePrint implies time and memory consumptions that are provably strictly less efficient than the tree decomposition underlying RNARedPrint.
+drawing sequences from the Boltzmann distribution does not involve any rejection, and is thus exactly as fast as drawing them uniformly.
+For uniform sampling, RNARedPrint is usually even faster than RNABluePrint, due to the state-of-the-art tree decomposition underlying RNARedPrint, which typically yields lower treewidths than the graph decomposition of RNABluePrint.
 On the other hand, multi-dimensional Boltzmann sampling requires a rejection step to target specific properties that are typically difficult to obtain by chance in the uniform distribution.
-There is, of course, an added cost associated to this rejection, and the simultaneous weight optimization, which we discuss further in a new "Result" subsection, where we also report the exact time differences.}
+There is, of course, an added cost associated to this rejection, and the simultaneous weight optimization.
+We do now provide detailed overview of the seed generation procedure, which like our multi-dimensional Boltzmann sampling strategy---runs through several iterations of performing the elementary Boltzmann sampling in Suppl.\ Mat.
+E.
+The corresponding Figure (Supp.
+Fig.
+9) shows that typically only few iterations suffice to produce the sequences.
+In this way, the time for multi-dimensional Boltzmann sampling typically is a small multiple of the time required for uniform sampling.}
 

### 2019-01-25 RNARedPrint.tex
figure description includes arrows now
sync

@@
   \revised{established} design benchmark. \revised{Our results demonstrate the effectivity and feasibility of the method in practice
-  as well as the benefits of Boltzmann sampling over the previously best multi-target sampling strategy---even for the case of negative design of multi-stable RNAs.}\\
+  as well as the benefits of Boltzmann sampling over the previously best multi-target sampling strategy---even for the case of negative design of multi-stable RNAs.}
 %
@@
 The tree is then used to compute the partition function, followed by a Boltzmann sampling of valid sequences (v).
-An adaptive scheme learns weights to achieve targeted energies and \GCb\%, leading to the production of suitable designs (vi).
+An adaptive scheme learns weights to achieve targeted energies and \GCb\% \revised{(arrows)}, leading to the production of suitable designs (vi).
 \revised{Note that for simplicity, we assume in this figure that only dependencies between the ends of base pairs are considered to evaluate the energies of structures.
@@
 While the former is expected since the Boltzmann distribution explicitly favors low-energy candidates, the latter is somewhat surprising, since the Boltzmann probability of a target structure could, in principle, decrease under Boltzmann sampling due to the partition function growing faster than the Boltzmann factor.
-The empirical superiority of Boltzmann sampling appears robust, and is at the foundation of other approaches~\cite{Reinharz2013}.
+The empirical superiority of Boltzmann sampling appears robust to the target structure length and topology, as demonstrated by prior work~\cite{Reinharz2013}.
 
@@
 
-However, while in principle feasible, sampling in a Boltzmann distribution directly using the Turner energy model may induce extreme computational demands, with tree-widths scaling at least as large as the number of nucleotides in the largest loop.
+However, while in principle feasible, sampling in a Boltzmann distribution directly using the Turner energy model may induce extreme computational demands, with treewidths scaling at least as large as the number of nucleotides in the largest loop.
 Fortunately, we found that intricacy of the Turner energy model can be circumvented with minimal loss of precision by using a simpler stacking energy model.
-As shown in Figure~\ref{fig:correlation}, a simple stacking energy model, further described in Supp.
-Mat.~\ref{appsec:modelparameters}, can be used to well approximate the Turner energy model (correlation coefficient $R=0.99$) for our purpose.
-This greatly reduces the treewidth, and thus the computational requirements of the whole method even for complex instances.
+As shown in Figure~\ref{fig:correlation}, a simple stacking energy model, whose design principles are further described in Supp.
+Mat.~\ref{appsec:modelparameters}, can be used to approximate the Turner energy model very adequately (correlation coefficient $R=0.99$) in the context of sequence design.
+Using this simpler model greatly reduces the treewidth, and thus the computational requirements of the whole method even for complex instances.
 }
@@
 
-\revised{Based on a general framework and efficient algorithms, we introduced a novel approach to design RNA
+\revised{Based on a general framework and efficient algorithms, we introduce a novel approach to design RNA
 sequences while targeting very specific complex properties}. In particular, we describe the targeting of the free energies of multiple target structures and the \GCb-content.
-Our method combines an FPT stochastic sampling algorithm
+%
+Our method combines a fixed-parameter tractable (FPT) sampling algorithm
 with multi-dimensional Boltzmann sampling over distributions
 controlled by expressive RNA energy models.
-\revised{We demonstrate that the approach, despite of its theoretical hardness, performs well on typical multi-stable RNA design instances in practice, which is a direct consequence of the approximability of Turner energies as well as the systematic algorithmic framework.
+\revised{We demonstrate that the approach, despite of its theoretical hardness, performs well on typical multi-stable RNA design instances in practice.
+This good performance is a direct consequence of the approximability of Turner energies as well as the systematic algorithmic framework.
 By conducting a proof-of-concept study
   on an established benchmark set for negative multi-target RNA design (including pseudoknotted instances), we showcase a typical application of our tool \ourprog{}. In this study, 
-  the approach generated significantly better seed sequences than
-the previously best available method (uniform sampling). These results strongly suggest that the presented technique for positive design can be highly beneficial in future negative design approaches.
-}
+  the approach generates significantly better seed sequences than
+the previously best available method (uniform sampling). Our results strongly suggest that the presented technique for positive design can be highly beneficial in future negative design approaches.
+}%
+%
+% SW--we have written most of this already:
+%In this way, we as well demonstrated the practical feasibility of such controlled %sequence generation, including the case of pseudoknotted structures.
+% YP--Stick to the present tense if you write anything new :) (or move everything to past) 
+%
 %
@@
 
-The presented method enables new possibilities for the sequence generation
-in the field of RNA sequence design, allowing to enforce additional
-constraints, like \GCb\%, while controlling the energy of
-multiple target structures. Thus, it presents a major advance over
-previously applied ad-hoc sampling and even efficient uniform sampling
-strategies. We demonstrated the practical feasibility of such controlled sequence
-generation, including the case of pseudoknotted structures, and studied its use for multi-target RNA design.
-
+\revised{In this way, our framework enables new possibilities
+in the field of RNA sequence design.}
+%allowing to enforce additional
+%constraints, like \GCb\%, while controlling the energy of
+%multiple target structures.
+%Thus, it presents a major advance over
+%previously applied ad-hoc sampling and even efficient %uniform sampling
+%strategies. 
+%
 %perspectives:
-Our framework is extensible to include various more complex sequence constraints, including mandatory/forbidden motifs at specific
+\revised{As particular advantage over previous sequence generation methods,} it is extensible to include various more complex sequence constraints, including mandatory/forbidden motifs at specific
 positions or anywhere in the designed sequences, by adapting formal language constructs
@@
 penalizing a set of alternative helices/structures.
-
-One could also envision using positive design as a null model, for instance in order to assess the significance of observed properties.
-Indeed, the complex set of constraints required in order to achieve a desired function must be equally enforced in RNA design, and to produce random sequences that are known to be constrained at a phenotypic level.
-\revised{Therefore, one could} use constrained design to assess whether P-values, and \revised{thus} our perception of the statistical significance, are affected by the overly simplistic simple null models (\emph{e.g.} dinucleotide shuffling) used to model random RNAs~\cite{Rivas2017}.
+%
+\revised{We moreover envision using positive design to assess the significance of observed properties.
+Critically, our current perception of statisitical significance suffers from overly simplistic simple null models (\emph{e.g.} dinucleotide shuffling) used to model random RNAs~\cite{Rivas2017}).
+Here, our approach promises fundamental improvements by constructing null models of random sequences that satisfy multiple complex constraints.}
 
@@
 %Austrian
-FWF (ANR-14-CE34-0011; project RNALands).  SH is supported by the
+Austrian Science Fund (ANR-14-CE34-0011 and FWF-I-1804-N28; project RNALands).  SH is supported by the
 German Federal Ministry of Education and Research (BMBF support code

### 2019-01-25 BMC-Response-Round1.tex
spell check

@@
 \Comment{
-Multi defect can be a good measure to understand the overall performance of the sequence designing method, but there is little explanation why Multidefect is enough to evaluate the quality of sequences for the multi (target-structure) objective problem in a practical use case such as Riboswitch design.
+Multi defect can be a good measure to understand the overall performance of the sequence designing method, but there is little explanation why MultiDefect is enough to evaluate the quality of sequences for the multi (target-structure) objective problem in a practical use case such as Riboswitch design.
 }%
@@
 Although RNABluePrint was compared with MODENA in the previous study, RNABluePrint did not overwhelm MODENA in all of conditions (and it is unclear which version (variant) of RNABluePrint is applied in this study).
-For that reason, RNARedPrint still should be compared with MODENA (in the same running time) using Multidefect and the criteria used in the previous studies such as $\delta$e1 and $\delta$e2.
-Also, RNARedprint and RNABlueprint are optimised to increase only a Multidefect value.}%
+For that reason, RNARedPrint still should be compared with MODENA (in the same running time) using MultiDefect and the criteria used in the previous studies such as $\delta$e1 and $\delta$e2.
+Also, RNARedPrint and RNABlueprint are optimized to increase only a MultiDefect value.}%
 
-\Answer{As we tried to clarify in our general comments of this response letter (as well as by appropriate adaptions in the manuscript), RNAredprint does not by itself perform \emph{negative} design.
+\Answer{As we tried to clarify in our general comments of this response letter (as well as by appropriate adaptions in the manuscript), RNARedPrint does not by itself perform \emph{negative} design.
 It is a generation algorithm for (positively) designing sequences with complex properties.
@@
 For this reason, we focused our validation on this potential for better initialization, and compared the quality of sequences generated by RNARedPrint (Boltzmann weighted) to RNABlueprint (uniform).
-We noticed that, while RNAredprint is not explicitly tailored to optimize MultiDefect, it nevertheless produces sequences that: 1) achieve better values; and 2) are amenable to better further optimizations than the uniform version implemented by RNABluePrint.}
+We noticed that, while RNARedPrint is not explicitly tailored to optimize MultiDefect, it nevertheless produces sequences that: 1) achieve better values; and 2) are amenable to better further optimizations than the uniform version implemented by RNABluePrint.}
 
-\Comment{Hence, the distribution of Multidefect, previous criteria, and the first and second term of Multidefect should be shown in box plots or violin plots to know the robustness of RNARedPrint.
+\Comment{Hence, the distribution of MultiDefect, previous criteria, and the first and second term of MultiDefect should be shown in box plots or violin plots to know the robustness of RNARedPrint.
 It is also interesting what is the characteristic of the sequence designed by each method in a specific example of functional RNA molecules.}%
@@
 \Answer{
-We share the concern of this reviewer that an unncecessarily imbalanced manuscript should be avoided, and moved less original/central parts of the method to the appendix.
+We share the concern of this reviewer that an unnecessarily imbalanced manuscript should be avoided, and moved less original/central parts of the method to the appendix.
 This includes the expression of various energy models as constraints, and the description of our naive numerical weight optimization.
@@
 \Answer{
-We understand that it is important to give an estimate of the actual performance and thus included computational times for the pre-computation, the Boltzmann sampling of 10\,000 sequences and the number of iterations for the multi-dimensional Boltzmann sampling for the complete benchmark set in Supl.
+We understand that it is important to give an estimate of the actual performance and thus included computational times for the pre-computation, the Boltzmann sampling of 10\,000 sequences and the number of iterations for the multi-dimensional Boltzmann sampling for the complete benchmark set in Suppl.
 Mat.
@@
 
-\Answer{We were not completely sure what was meant by this comment. Just to be precise, we distinguish the term 'Boltzmann sampling' from 'multi-dimensional Bolzmann sampling' in our paper. 
+\Answer{We were not completely sure what was meant by this comment. Just to be precise, we distinguish the term 'Boltzmann sampling' from 'multi-dimensional Boltzmann sampling' in our paper. 
 On the one hand, Boltzmann sampling, i.e.

### 2019-01-25 RNARedPrint.tex
spell check

@@
   biologically relevant multi-target Boltzmann-weighted designs for a
-  \revised{established} design benchmark. \revised{Our results demonstrate the effectivity and feasibility of the method in practice
+  \revised{established} design benchmark. \revised{Our results demonstrate the efficacy and feasibility of the method in practice
   as well as the benefits of Boltzmann sampling over the previously best multi-target sampling strategy---even for the case of negative design of multi-stable RNAs.}
@@
 Moreover, to ensure a uniform algorithmic treatment of contributions, it is convenient to encode the weight $\pi$ of each feature in its contributions.
-This transformation works by multiplying all contributions with $ln(\pi)$, where $\pi$ is the weight of the corresponding feature, since then
+This transformation works by multiplying all contributions with $\ln(\pi)$, where $\pi$ is the weight of the corresponding feature, since then
 $exp(-\ln(\pi)f(S))=\pi^{-f(S)}.$
@@
 \revised{We moreover envision using positive design to assess the significance of observed properties.
-Critically, our current perception of statisitical significance suffers from overly simplistic simple null models (\emph{e.g.} dinucleotide shuffling) used to model random RNAs~\cite{Rivas2017}).
+Critically, our current perception of statistical significance suffers from overly simplistic simple null models (\emph{e.g.} dinucleotide shuffling) used to model random RNAs~\cite{Rivas2017}).
 Here, our approach promises fundamental improvements by constructing null models of random sequences that satisfy multiple complex constraints.}

### 2019-01-27 RNARedPrint.tex
cosmetics

@@
 %% document please refer to Readme.html and the instructions for   %%
-%% authors page on the biomed central website                      %%https://www.overleaf.com/project/5b98dc8ced90175e9717c15a
+%% authors page on the biomed central website                      %%
 %% http://www.biomedcentral.com/info/authors/                      %%
@@
 
-%\documentclass[doublespacing,linenumbers]{bmcart}% uncomment this for twocolumn layout and comment line below
-\documentclass[]{bmcart}
+\documentclass[doublespacing,linenumbers]{bmcart}% uncomment this for twocolumn layout and comment line below
+%\documentclass[]{bmcart}
 %\documentclass[twocolumn]{bmcart}
@@
 
+\usepackage{graphicx}
+%\usepackage{endfloat}
+%\usepackage{caption}
+%\captionsetup{labelsep=none,textformat=empty}
+
 %%% Put your definitions there:
@@
 %% Rolf's includegraphicstop
-\makeatletter
-\newsavebox{\@alignepsbox}
-\newlength{\@aligneps}
-\newcommand{\includegraphicstop}[2][]{%
-\sbox{\@alignepsbox}{\includegraphics[#1]{#2}}%
-\setlength{\@aligneps}{-\ht\@alignepsbox}%
-\addtolength{\@aligneps}{2ex}%
-\raisebox{\@aligneps}{\usebox{\@alignepsbox}}}
-\makeatother
+% \makeatletter
+% \newsavebox{\@alignepsbox}
+% \newlength{\@aligneps}
+% \newcommand{\includegraphicstop}[2][]{%
+% \sbox{\@alignepsbox}{\includegraphics[#1]{#2}}%
+% \setlength{\@aligneps}{-\ht\@alignepsbox}%
+% \addtolength{\@aligneps}{2ex}%
+% \raisebox{\@aligneps}{\usebox{\@alignepsbox}}}
+% \makeatother
 
@@
 %
-  \parttitle{Conclusion} \ourprog{} introduces a novel, flexible and---as we demonstrate---effective approach to multi-target RNA design, which promises broad applicability and extensibility.
+  \parttitle{Conclusion} \ourprog{} introduces a novel, flexible, and effective approach to multi-target RNA design, which promises broad applicability and extensibility.
 
@@
 \begin{figure}
-	\includegraphicstop[width=0.8\textwidth]{Figs/Plots/doublestem_weight1-500_mean}
-	\caption{\revised{Comparison of free-energy and Boltzmann probability for 10~000 uniform ($\pi=1$; blue dots) and Boltzmann distributed ($\pi=500$; green dots) sequences, targeting a simple structure consisting of two adjacent helices.}}\label{fig:BoltzmannBenefits}
+	\includegraphics[width=0.8\textwidth]{Figs/Plots/doublestem_weight1-500_mean}
+	\caption{\revised{Comparison of free-energy and Boltzmann probability for 10~000 uniform ($\pi=1$; blue dots) and Boltzmann distributed ($\pi=500$; green dots) sequences, targeting a simple structure consisting of two adjacent helices.}}
+        \label{fig:BoltzmannBenefits}
 \end{figure}
@@
 While the former is expected since the Boltzmann distribution explicitly favors low-energy candidates, the latter is somewhat surprising, since the Boltzmann probability of a target structure could, in principle, decrease under Boltzmann sampling due to the partition function growing faster than the Boltzmann factor.
-The empirical superiority of Boltzmann sampling appears robust to the target structure length and topology, as demonstrated by prior work~\cite{Reinharz2013}.
+The empirical superiority of Boltzmann sampling appears robust to the target structure length and topology, as demonstrated by prior work~\cite{Reinharz2013}.}
 
@@
 	{\sffamily {\bfseries B} -- Correlation plot}\\
-	  	\includegraphicstop[width=\textwidth,trim=0 0 0 50,clip]{Figs/stackingcor} 
+	  	\includegraphics[width=\textwidth,trim=0 0 0 50,clip]{Figs/stackingcor} 
   	\end{minipage}
 
-	\caption{\revised{A fitted energy model based on stacking pairs ({\sffamily\bfseries A}) leads to approximated free-energies that are highly correlated (R=0.99) with free-energies in the Turner energy model ({\sffamily\bfseries B}), yet induces tree widths that are amenable to practical sampling ({\sffamily\bfseries C})}.\label{fig:correlation}}
+        \caption{\revised{A fitted energy model based on stacking pairs ({\sffamily\bfseries A}) leads to approximated free-energies that are highly correlated (R=0.99) with free-energies in the Turner energy model ({\sffamily\bfseries B}), yet induces tree widths that are amenable to practical sampling ({\sffamily\bfseries C})}.}
+      \label{fig:correlation}
 \end{figure}
 
+\revised{%
 However, while in principle feasible, sampling in a Boltzmann distribution directly using the Turner energy model may induce extreme computational demands, with treewidths scaling at least as large as the number of nucleotides in the largest loop.
@@
     %{\sf \bfseries A}\
-    \includegraphicstop[width=0.97\textwidth]{Figs/energy_shift}\hfill
+    \includegraphics[width=0.97\textwidth]{Figs/energy_shift}\hfill
     %{\sf \bfseries B}\ \includegraphicstop[width=0.55\textwidth]{Figs/PKB00211_PKB00239_0_energy_distribution}

### 2019-03-13 RNARedPrint.tex
sync
Do not show revised markup for first revision anymore Rename \revised to \revisedOne; introduce new \revised macro

@@
 
-\newcommand{\revised}[1]{{\color{red} #1}}
+%% revised for first revision
+\newcommand{\revisedOne}[1]{#1}
+%%revised for second minor/discretionary revision
+\newcommand{\revised}[1]{{\color{red!60!black} #1}}
 
@@
   The design of multi-stable RNA molecules has important applications in biology, medicine, and biotechnology.
-Synthetic design approaches profit strongly from effective in-silico methods, \revised{which substantially reduce the need for costly wet-lab experiments.} \\
+Synthetic design approaches profit strongly from effective in-silico methods, \revisedOne{which substantially reduce the need for costly wet-lab experiments.} \\
 %
   \parttitle{Results} We devise a novel approach to a central ingredient of most in-silico
-  design methods: the generation of sequences \revised{that fold well into multiple target structures.} 
-  % 
-  \revised{Based on constraint networks, our approach \ourprog{}} supports generic
+  design methods: the generation of sequences \revisedOne{that fold well into multiple target structures.} 
+  % 
+  \revisedOne{Based on constraint networks, our approach \ourprog{}} supports generic
   Boltzmann-weighted sampling, which enables the positive design of RNA sequences
@@
   biologically relevant multi-target Boltzmann-weighted designs for a
-  \revised{established} design benchmark. \revised{Our results demonstrate the efficacy and feasibility of the method in practice
+  \revisedOne{established} design benchmark. \revisedOne{Our results demonstrate the efficacy and feasibility of the method in practice
   as well as the benefits of Boltzmann sampling over the previously best multi-target sampling strategy---even for the case of negative design of multi-stable RNAs.}
 %
-  \revised{Besides empirically studies, we finally justify the algorithmic details due to a fundamental theoretic result about multi-stable RNA design, namely the \#P-hardness of the counting of designs.}
+  \revisedOne{Besides empirically studies, we finally justify the algorithmic details due to a fundamental theoretic result about multi-stable RNA design, namely the \#P-hardness of the counting of designs.}
 %
@@
 design~\citep{Wachsmuth2013,Domin2017}, partly motivated by their
-capacity to act as biosensors~\citep{Findeiss2017}, \revised{which suggests them for biotechnological applications.}
+capacity to act as biosensors~\citep{Findeiss2017}, \revisedOne{which suggests them for biotechnological applications.}
 In particular due to the kinetic coupling of RNA folding with RNA transcription, RNA families can feature alternative,
@@
 %
-Faced with the complex constraints due to the multiple targets, early methods such as \Software{Frnakenstein}~\citep{Lyngsoe2012} and \Software{Modena}~\citep{Taneda2015} \revised{do not even attempt to sample sequences systematically from a controlled distribution}, but rely on ad-hoc generation strategies.
+Faced with the complex constraints due to the multiple targets, early methods such as \Software{Frnakenstein}~\citep{Lyngsoe2012} and \Software{Modena}~\citep{Taneda2015} \revisedOne{do not even attempt to sample sequences systematically from a controlled distribution}, but rely on ad-hoc generation strategies.
 %
@@
 target structures. \Software{RNAdesign} adopts a graph coloring perspective,
-\revised{assigning nucleotide symbols (like ``colors'') to the sequence positions, such that compatible nucleotides are assigned to the ends of each base pair.} Initially, the method decomposes the graph hierarchically and then \Def{precomputes} the number of valid sequences
+\revisedOne{assigning nucleotide symbols (like ``colors'') to the sequence positions, such that compatible nucleotides are assigned to the ends of each base pair.} Initially, the method decomposes the graph hierarchically and then \Def{precomputes} the number of valid sequences
 within each subgraph. The decomposition is then reinterpreted as a
@@
 As well, the \Software{RNAdesign}/\RNAblueprint{} approach is specialized to uniform sampling, which limits its direct extensibility.
-\revised{Substantial improvements of multi-target sampling thus require a systematically redesigned approach.
+\revisedOne{Substantial improvements of multi-target sampling thus require a systematically redesigned approach.
 To enable a fundamentally broader range of applications in extensions of the sampling method, we build our approach, from the start, on established concepts in computer science.}
@@
 The tree is then used to compute the partition function, followed by a Boltzmann sampling of valid sequences (v).
-An adaptive scheme learns weights to achieve targeted energies and \GCb\% \revised{(arrows)}, leading to the production of suitable designs (vi).
-\revised{Note that for simplicity, we assume in this figure that only dependencies between the ends of base pairs are considered to evaluate the energies of structures.
+An adaptive scheme learns weights to achieve targeted energies and \GCb\% \revisedOne{(arrows)}, leading to the production of suitable designs (vi).
+\revisedOne{Note that for simplicity, we assume in this figure that only dependencies between the ends of base pairs are considered to evaluate the energies of structures.
 Our computations based on a more complex energy model, which considers energy contributions of base pair stacks, require additional dependencies.}
@@
 
-\revised{As central contribution, we provide a systematic and flexibly extensible technique for sampling that targets multiple versatile features.
+\revisedOne{As central contribution, we provide a systematic and flexibly extensible technique for sampling that targets multiple versatile features.
 For the sake of clarity, we introduce this method specialized to the sampling of RNA sequences that have specific energies for multiple structures and specific \GCb\%.}
 
-\revised{Figure~\ref{fig:workflow} summarizes our generic framework, which enables this targeted sequence generation based on multi-dimensional Boltzmann sampling.
+\revisedOne{Figure~\ref{fig:workflow} summarizes our generic framework, which enables this targeted sequence generation based on multi-dimensional Boltzmann sampling.
 \emph{Algorithmically}, we originally contribute} dynamic programming (DP) algorithms, based on the concept of \Def{tree decomposition}, to compute partition functions and sample sequences from the Boltzmann distribution%
@@
 .
-\revised{Generally, tree decompositions are data structures that capture the specific dependencies of a problem instance (here, the dependencies between sequence positions induced by the target structures), such that they can guide the efficient processing by DP algorithms.}
-%
-\revised{Building on this principle, the complexities of our algorithms depend exponentially on a specific property of the tree decomposition, called the \Def{treewidth}.
+\revisedOne{Generally, tree decompositions are data structures that capture the specific dependencies of a problem instance (here, the dependencies between sequence positions induced by the target structures), such that they can guide the efficient processing by DP algorithms.}
+%
+\revisedOne{Building on this principle, the complexities of our algorithms depend exponentially on a specific property of the tree decomposition, called the \Def{treewidth}.
 Thus, it is essential for the applicability of our approach that---by appropriate design choices---we can keep this parameter low for typical instances.
@@
 
-\revised{Remarkably, we could show that it is not possible to find a better, efficient method for sampling (unless ${\sf P}={\sf NP}$), since the underlying counting problem is \#{\sf P}-hard.
+\revisedOne{Remarkably, we could show that it is not possible to find a better, efficient method for sampling (unless ${\sf P}={\sf NP}$), since the underlying counting problem is \#{\sf P}-hard.
 The practical relevance of this theoretical result is that it rules out substantially better sampling techniques.
@@
 %
-Due to the generality of our method, we can moreover strongly limit \revised{the treewidth} in practice by using state-of-the-art tree decomposition algorithms.
+Due to the generality of our method, we can moreover strongly limit \revisedOne{the treewidth} in practice by using state-of-the-art tree decomposition algorithms.
 %Uniform sampling is handled as a special case of Boltzmann sampling, where each valid sequence receives energy zero and---consequently---computing partition functions specializes to counting.
 By evaluating sequences in a specialized weighted constraint
-network, we support---in principle---arbitrary \revised{complex constraints and energy models},
+network, we support---in principle---arbitrary \revisedOne{complex constraints and energy models},
 notably subsuming the commonly
@@
 Turner RNA energy model
-works well by performing sampling based on \revised{a simplified RNA energy model, which induces much lower treewidth than the Turner model}.
-This result is \revised{essential for the applicability of our method, since it allows to combine high efficiency (by keeping the treewidth low) with sufficient accuracy to precisely target realistic Turner energies.}
-
-\revised{Eventually, our proof-of-concept results 
+works well by performing sampling based on \revisedOne{a simplified RNA energy model, which induces much lower treewidth than the Turner model}.
+This result is \revisedOne{essential for the applicability of our method, since it allows to combine high efficiency (by keeping the treewidth low) with sufficient accuracy to precisely target realistic Turner energies.}
+
+\revisedOne{Eventually, our proof-of-concept results 
  on a comprehensive multi-target RNA design benchmark~\citep{Taneda2015} 
@@
 \section*{Methods}
-\revised{The main computational problem addressed in this work is the positive design of RNA sequences for multiple target structures; more specifically, the generation of sequences over the alphabet $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}$, such that the sequences feature a given $\GCb\%$, and have prescribed energies for a set of target secondary structures.
+\revisedOne{The main computational problem addressed in this work is the positive design of RNA sequences for multiple target structures; more specifically, the generation of sequences over the alphabet $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}$, such that the sequences feature a given $\GCb\%$, and have prescribed energies for a set of target secondary structures.
 Here, these desired sequence properties are modeled as constraints on the values of \Def{features}, which are functions of the sequence that are expressed as sums over real-valued \Def{contributions}.
@@
 
-\revised{To generate diverse design candidates, we randomly generate sequences from a Boltzmann distribution.
+\revisedOne{To generate diverse design candidates, we randomly generate sequences from a Boltzmann distribution.
 The probability of a sequence then depends on its features (e.g.
@@
 %\paragraph{Central computational problem.}
-Let us consider $m$ (secondary) structures $R_1,\dots, R_m$, each abstracted as a set of base pairs, and \revised{$k\geq m$} features $F_1,\dots, F_k$\revised{, typically representing the energies of the structures and additional sequence properties,} associated with weights $\pi_1,\dots,\pi_k$.
-Our goal is to sample sequences \revised{$S$ (which satisfy the base pairing rules for all structures) from the Boltzmann distribution defined by}
+Let us consider $m$ (secondary) structures $R_1,\dots, R_m$, each abstracted as a set of base pairs, and \revisedOne{$k\geq m$} features $F_1,\dots, F_k$\revisedOne{, typically representing the energies of the structures and additional sequence properties,} associated with weights $\pi_1,\dots,\pi_k$.
+Our goal is to sample sequences \revisedOne{$S$ (which satisfy the base pairing rules for all structures) from the Boltzmann distribution defined by}
 \begin{equation}
@@
 
-The workhorse of our approach is the \revised{fixed-parameter tractable} computation of feature-dependent partition functions over sequences, \revised{namely partition functions of the form} 
+The workhorse of our approach is the \revisedOne{fixed-parameter tractable} computation of feature-dependent partition functions over sequences, \revisedOne{namely partition functions of the form} 
 %
@@
 Formally, we define a \Def{feature $F$} as a function on sequences, whose value is obtained by summing over an associated set of \Def{contributions}.
-Each contribution $f$ takes values in $\mathbb{R}\cup \{+\infty\}$, and depends on the nucleotides assigned to a restricted set of positions, \revised{namely its \emph{dependencies}, denoted $\dep(f)$}, such that
+Each contribution $f$ takes values in $\mathbb{R}\cup \{+\infty\}$, and depends on the nucleotides assigned to a restricted set of positions, \revisedOne{namely its \emph{dependencies}, denoted $\dep(f)$}, such that
 
@@
 $$
-\revised{Here, since $\dep(f)=\{x_1,\ldots,x_p\}$, $\left\{\substack{x_1\mapsto S_{x_1}\\\cdots\\x_p\mapsto S_{x_p}}\right\}$ denotes the assignment, that assigns to all positions $x_k$  in $\dep(f)$ the respective nucleotide $S_{x_k}$ ($1\leq k\leq p$; $p=|\dep(f)|$).}
-
-The $\GCb\%$ can be simply expressed using $n$ contributions $f^{\GCb}_i$, \revised{each depending only on position $i\in[1,n]$, i.e.~$\dep(f^{\GCb}_i)=\{i\}$}, such that 
+\revisedOne{Here, since $\dep(f)=\{x_1,\ldots,x_p\}$, $\left\{\substack{x_1\mapsto S_{x_1}\\\cdots\\x_p\mapsto S_{x_p}}\right\}$ denotes the assignment, that assigns to all positions $x_k$  in $\dep(f)$ the respective nucleotide $S_{x_k}$ ($1\leq k\leq p$; $p=|\dep(f)|$).}
+
+The $\GCb\%$ can be simply expressed using $n$ contributions $f^{\GCb}_i$, \revisedOne{each depending only on position $i\in[1,n]$, i.e.~$\dep(f^{\GCb}_i)=\{i\}$}, such that 
 $$f^{\GCb}_i(\{i\mapsto c\})=\begin{cases}-1&\text{ if }c=\Gb\text{ or }\Cb\\ 0&\text{ otherwise}.\end{cases}$$
 
-\revised{By summing $f^{\GCb}_i(\{i\mapsto S_i\})$ over the whole sequence ($i=1,\dots,n$), one simply counts the occurrences of $\Gb$ and $\Cb$.}
-
-\revised{
+\revisedOne{By summing $f^{\GCb}_i(\{i\mapsto S_i\})$ over the whole sequence ($i=1,\dots,n$), one simply counts the occurrences of $\Gb$ and $\Cb$.}
+
+\revisedOne{
   To start with a simple example of evaluating the energy of sequences by features, let us explain how they are used to count the number of \Def{valid sequences}, \emph{i.e.} sequences inducing only base pairs in $\mathcal{B}:=\{\{\Ab,\Ub\}, \{\Gb,\Cb\}, \{\Gb,\Ub\}\}$.
@@
 
-\revised{In order to compute the partition function of Eq.~\ref{eq:mainproblem}, and thus sample in a well-defined way, one must consider dependencies induced by the complete set of contributions $$\F := \bigcup_{\ell}\{f\mid f\text{ contribution of } F_\ell\}.$$}
+\revisedOne{In order to compute the partition function of Eq.~\ref{eq:mainproblem}, and thus sample in a well-defined way, one must consider dependencies induced by the complete set of contributions $$\F := \bigcup_{\ell}\{f\mid f\text{ contribution of } F_\ell\}.$$}
 
@@
 
-\revised{
+\revisedOne{
 In the simplest case, this set captures the requirement of canonical base pairing for each structure.
@@
 %
-\revised{Intuitively, a tree decomposition of an (hyper)graph $G$ is a tree that captures all the vertices and (hyper)edges of $G$, and properly relates dependent sub-problems to ensure consistency in a recursive computation.
+\revisedOne{Intuitively, a tree decomposition of an (hyper)graph $G$ is a tree that captures all the vertices and (hyper)edges of $G$, and properly relates dependent sub-problems to ensure consistency in a recursive computation.
 Figure~\ref{fig:treedec} shows an optimal tree decomposition for a pair of structures under the stacking energy model.}
@@
 {\centering \includegraphics[width=.9\linewidth]{Figs/example-treedec-stacking}\\}
-\caption{\revised{Toy example of a tree decomposition associated with two target structures in the stacking energy model (where the four positions of each base pair stack depend on each other).
+\caption{\revisedOne{Toy example of a tree decomposition associated with two target structures in the stacking energy model (where the four positions of each base pair stack depend on each other).
 Two target secondary structures ({\bfseries\sffamily A}) are merged into a joint hypergraph ({\bfseries\sffamily B}), whose hyperedges correspond to the quadruplets of positions involved in base pair stacks (colored).
@@
 Let us call the latter object a \Def{partial sequence}.
-Such an object will help \revised{to} specify partial knowledge on the sequence at some point of the algorithm.
+Such an object will help \revisedOne{to} specify partial knowledge on the sequence at some point of the algorithm.
 Easily, we can extend the definition of contributions $f$ to sets $\{x_1\mapsto v_1, \dots, x_p\mapsto v_p\}$, where $\{x_1\dots x_p\}$ is any super-set of $\dep(f)$ by ignoring the superfluous assignments $x\mapsto v$, where $x\not\in\dep(f)$.
@@
 
-In order to capture realistic energy models like the Turner model or pseudoknot models like {\tt HotKnots}~\citep{Ren2005}, our sampling strategy can be extended in two ways: 1) either by directly sampling based on more expressive energy models \revised{or 2) by sampling in a simple energy model which} can be used to approximate sampling in more complex models.
-In practice, complex energy models \revised{have} a strong influence on the treewidth (of optimal tree decompositions) of the dependency graph and \revised{thus on} the computational complexity of our approach.
+In order to capture realistic energy models like the Turner model or pseudoknot models like {\tt HotKnots}~\citep{Ren2005}, our sampling strategy can be extended in two ways: 1) either by directly sampling based on more expressive energy models \revisedOne{or 2) by sampling in a simple energy model which} can be used to approximate sampling in more complex models.
+In practice, complex energy models \revisedOne{have} a strong influence on the treewidth (of optimal tree decompositions) of the dependency graph and \revisedOne{thus on} the computational complexity of our approach.
 Therefore, it is interesting to consider---in addition to the base pair model---other
@@
 %We define the following notations: $\real^\infty:=\real\cup\{\infty\}$, $[i,j]:=\{i,\ldots,j\},$ \ldots
-\revised{While efficient, both in practice and in theory for graphs of bounded treewidth, our algorithms remain exponential in the worst case scenario, since the treewidth of a dependency graph can then become arbitrarily large.
+\revisedOne{While efficient, both in practice and in theory for graphs of bounded treewidth, our algorithms remain exponential in the worst case scenario, since the treewidth of a dependency graph can then become arbitrarily large.
 This exponential complexity in the worst case appears to be intrinsic.
@@
 
-\revised{Proposition~\ref{prop:bijection} also strongly impacts the complexity of computing the partition function.
+\revisedOne{Proposition~\ref{prop:bijection} also strongly impacts the complexity of computing the partition function.
 Indeed it implies that, among the $4^k$ possible assignments of nucleotides to $k$ connected positions (in the base-pair dependency graph), at most $2^k$ are compatible with base pairing rules.
@@
 	\includegraphics[width=0.8\textwidth]{Figs/Plots/doublestem_weight1-500_mean}
-	\caption{\revised{Comparison of free-energy and Boltzmann probability for 10~000 uniform ($\pi=1$; blue dots) and Boltzmann distributed ($\pi=500$; green dots) sequences, targeting a simple structure consisting of two adjacent helices.}}
+	\caption{\revisedOne{Comparison of free-energy and Boltzmann probability for 10~000 uniform ($\pi=1$; blue dots) and Boltzmann distributed ($\pi=500$; green dots) sequences, targeting a simple structure consisting of two adjacent helices.}}
         \label{fig:BoltzmannBenefits}
@@
 
-\revised{First, we show how seed sequences can be generated in a Boltzmann distribution, leading to designs that are substantially more stable that those generated uniformly.
+\revisedOne{First, we show how seed sequences can be generated in a Boltzmann distribution, leading to designs that are substantially more stable that those generated uniformly.
 As can be seen in Figure~\ref{fig:BoltzmannBenefits} and Supp.
@@
 
-        \caption{\revised{A fitted energy model based on stacking pairs ({\sffamily\bfseries A}) leads to approximated free-energies that are highly correlated (R=0.99) with free-energies in the Turner energy model ({\sffamily\bfseries B}), yet induces tree widths that are amenable to practical sampling ({\sffamily\bfseries C})}.}
+        \caption{\revisedOne{A fitted energy model based on stacking pairs ({\sffamily\bfseries A}) leads to approximated free-energies that are highly correlated (R=0.99) with free-energies in the Turner energy model ({\sffamily\bfseries B}), yet induces tree widths that are amenable to practical sampling ({\sffamily\bfseries C})}.}
       \label{fig:correlation}
@@
 
-\revised{%
+\revisedOne{%
 However, while in principle feasible, sampling in a Boltzmann distribution directly using the Turner energy model may induce extreme computational demands, with treewidths scaling at least as large as the number of nucleotides in the largest loop.
@@
 \subsection*{Effectively targeting Turner energies using multi-dimensional sampling}
-\revised{
+\revisedOne{
 We used our Boltzmann sampling strategy (Algorithms
@@
 pseudoknots (D\&P model) and energies in the stacking energy
-model\revised{, which is obtained for each of the two target structures (with respective $R^2$ values of $0.846$ and $0.841$)}. For the shown fits we sampled $n=10\,000$ sequences, targeting a
+model\revisedOne{, which is obtained for each of the two target structures (with respective $R^2$ values of $0.846$ and $0.841$)}. For the shown fits we sampled $n=10\,000$ sequences, targeting a
 \GCb\% of $60\%$.
@@
     energies in the stacking model to the realistic pseudoknot energy
-    model by Dirks and Pierce (D\&P) for \revised{initially} sampled sequences and both target structures R1 and R2 (shown in B). The good
-    match \revised{(respective $R^2$ values of $0.846$ and $0.841$)} enables more efficient targeting of Turner energies based on
-    targeting stacking model energies. \textbf{(B)} \revised{Resulting D\&P energy distributions for the two target structures R1 and R2 when aiming for the respective free energies $-30$ and $-20$,
+    model by Dirks and Pierce (D\&P) for \revisedOne{initially} sampled sequences and both target structures R1 and R2 (shown in B). The good
+    match \revisedOne{(respective $R^2$ values of $0.846$ and $0.841$)} enables more efficient targeting of Turner energies based on
+    targeting stacking model energies.
+\textbf{(B)} \revisedOne{Resulting D\&P energy distributions for the two target structures R1 and R2 when aiming for the respective free energies $-30$ and $-20$,
     $-30$ and $-30$, $-25$ and $-25$, $-20$ and $-30$\,kcal/mol. These demonstrate the effectivity of our adaptive
@@
 As a baseline for comparison, we considered \RNAblueprint~\citep{Hammer2017}, the current leading tool for multiple design.
-As a quality measure, we applied the objective function introduced by~\citet{Hammer2017} \revised{based on \cite{Flamm2001,HoenerzuSiederdissen2013}} for multi-stable design, defined as:
+As a quality measure, we applied the objective function introduced by~\citet{Hammer2017} \revisedOne{based on \cite{Flamm2001,HoenerzuSiederdissen2013}} for multi-stable design, defined as:
 % \begin{align}
@@
 %for RNA design.
-\revised{Based on} \ourprog{}, we generated at least 1\,000 seed sequences with similar energies for all target structures, for
+\revisedOne{Based on} \ourprog{}, we generated at least 1\,000 seed sequences with similar energies for all target structures, for
 each instance of the benchmark. 
@@
 Generally, we targeted 60\% \GCb\%.
-\revised{%
+\revisedOne{%
 Supp.\ Mat.~\ref{appsec:effectivity-seed-generation} provides detailed results from this iterative procedure, which works similarly to the previously described multi-dimensional Boltzmann sampling.
@@
 We compared the
-\Obj{} value of \revised{the derived} sequences against that of seed sequences,
+\Obj{} value of \revisedOne{the derived} sequences against that of seed sequences,
 uniformly sampled using \RNAblueprint{}.  Moreover, for both sets we
@@
 
-\revised{Based on a general framework and efficient algorithms, we introduce a novel approach to design RNA
+\revisedOne{Based on a general framework and efficient algorithms, we introduce a novel approach to design RNA
 sequences while targeting very specific complex properties}. In particular, we describe the targeting of the free energies of multiple target structures and the \GCb-content.
@@
 controlled by expressive RNA energy models.
-\revised{We demonstrate that the approach, despite of its theoretical hardness, performs well on typical multi-stable RNA design instances in practice.
+\revisedOne{We demonstrate that the approach, despite of its theoretical hardness, performs well on typical multi-stable RNA design instances in practice.
 This good performance is a direct consequence of the approximability of Turner energies as well as the systematic algorithmic framework.
@@
 
-\revised{In this way, our framework enables new possibilities
+\revisedOne{In this way, our framework enables new possibilities
 in the field of RNA sequence design.}
@@
 %perspectives:
-\revised{As particular advantage over previous sequence generation methods,} it is extensible to include various more complex sequence constraints, including mandatory/forbidden motifs at specific
+\revisedOne{As particular advantage over previous sequence generation methods,} it is extensible to include various more complex sequence constraints, including mandatory/forbidden motifs at specific
 positions or anywhere in the designed sequences, by adapting formal language constructs
@@
 %
-\revised{We moreover envision using positive design to assess the significance of observed properties.
+\revisedOne{We moreover envision using positive design to assess the significance of observed properties.
 Critically, our current perception of statistical significance suffers from overly simplistic simple null models (\emph{e.g.} dinucleotide shuffling) used to model random RNAs~\cite{Rivas2017}).

### 2019-03-15 supplement.tex
sync
supplement layout fixed
Fix typo; remove highlighting for revision 1

@@
 
-%% highlight changes for revision in red
-\newcommand{\revised}[1]{{\color{red} #1}}
+%% revised for first revision
+\newcommand{\revisedOne}[1]{#1}
+\newenvironment{revisionOne}{}{}
+%%revised for second minor/discretionary revision
+\newcommand{\revised}[1]{{\color{red!60!black} #1}}
 \newenvironment{revision}{\color{red}}{\color{black}}
@@
 weight $\pi=500$. We show results for two different small structures: (A) single stem structure (B) double stem structure. For each respective structure $R$, we plot the energies $E(S,R)$ of
-10000 sampled sequences against the probabilities $Pr(R|S)$ (in log scale).
+10\,000 sampled sequences against the probabilities $Pr(R|S)$ (in log scale).
 For illustrating the trend, we show moving averages over windows of $\pm 1$ kcal.
@@
 
-\revised{
+\revisedOne{
   Finally, for further illustration, we show concrete tree decompositions for a selected instance in the benchmark set RF3 in Figure~\ref{appfig:td-example}.
@@
 This shows in many connected components, which correspond to independently solvable sub-problems, and fewer positions per bag (i.e.
-smaller tree width), which allows far more efficient evaluation per tree node.
+smaller treewidth), which allows far more efficient evaluation per tree node.
   Note that the complexity of the tree decomposition ultimately shows the complexity of the dependencies that need to be considered for simultaneous energy evaluation for all target structures.
@@
 
-\begin{revision}
+\begin{revisionOne}
 
@@
 
-To illustrate the efficiency and practical feasibility of our Boltzmann sampling approach, we show the distributions of run times (Intel(R) Core(TM) i7-4770 CPU) for generating 10000 samples (over the Modena benchmark instances and grouped by benchmark sub-set).
-Since producing a controlled Boltzmann sampling requires significant precomputation, it is informative to separately show construction and sampling times (Figure~\ref{appfig:run-times-redprint}).
-Moreover, \ourprog{} can be used with a simpler base pair model or a rather complex stacking energy model, which introduces much stronger dependencies than required for uniform sampling.
-Technically, these stronger dependencies increase the tree width and thus the run-time, as noticeable in Figure~\ref{appfig:run-times-redprint}.
-
-\begin{figure}[h!]
+\begin{figure}[hb!]
   \centering
@@
 \end{figure}
-\begin{figure}[h!]
+\begin{figure}[hb!]
   \centering
@@
 
+To illustrate the efficiency and practical feasibility of our Boltzmann sampling approach, we show the distributions of run times (Intel(R) Core(TM) i7-4770 CPU) for generating 10000 samples (over the Modena benchmark instances and grouped by benchmark sub-set).
+Since producing a controlled Boltzmann sampling requires significant precomputation, it is informative to separately show construction and sampling times (Figure~\ref{appfig:run-times-redprint}).
+Moreover, \ourprog{} can be used with a simpler base pair model or a rather complex stacking energy model, which introduces much stronger dependencies than required for uniform sampling.
+Technically, these stronger dependencies increase the tree width and thus the run-time, as noticeable in Figure~\ref{appfig:run-times-redprint}.
+
 A fair comparison to sampling times by other tools can hardly be given, since the other tools produce samplings of very different nature.
@@
 
-\end{revision}
+\end{revisionOne}
 
@@
         sampling. We show the Turner energy distributions for
-        three target structures R1, R2 and R3. \revised{By
+        three target structures R1, R2 and R3. \revisedOne{By
         targeting the respective free energies ($-40$, $-40$ and $-20$\,kcal/mol and $-20$,$-20$ and $-20$\,kcal/mol) for the target structures (annotated in red and
@@
 
-\begin{revision}
+\begin{revisionOne}
 \section{Effectivity of the seed sequence generation}

### 2019-03-15 RNARedPrint.tex
Changes for R2; leave some todo notes

@@
 
+\newcommand{\numTargetStructures}{\revised{k}}
+\newcommand{\numFeatures}{\revised{m}}
+\newcommand{\motif}{\revised{\frak m}}
+
 %%% end macro defs
@@
   Moreover, we study general properties of our approach empirically and generate
-  biologically relevant multi-target Boltzmann-weighted designs for a
+  biologically relevant multi-target Boltzmann-weighted designs for an
   \revisedOne{established} design benchmark. \revisedOne{Our results demonstrate the efficacy and feasibility of the method in practice
@@
 For the sake of clarity, we introduce this method specialized to the sampling of RNA sequences that have specific energies for multiple structures and specific \GCb\%.}
+\revised{In this way, we address the positive design of RNA sequences. Positive design is contrasted to the often
+desirable negative design of RNAs, which optimizes the stability of the target structures \emph{in relation to all
+other potential structures}. Remarkably, the even more complex task of negative design immediately benefits from positive
+design (Additional file 1: Section~\ref{appsec:immediate-benefits-for-negative-design}), which provides an initial motivation to study the positive design problem by itself.}
 
@@
 %\paragraph{Central computational problem.}
-Let us consider $m$ (secondary) structures $R_1,\dots, R_m$, each abstracted as a set of base pairs, and \revisedOne{$k\geq m$} features $F_1,\dots, F_k$\revisedOne{, typically representing the energies of the structures and additional sequence properties,} associated with weights $\pi_1,\dots,\pi_k$.
+Let us consider \revised{a set of} $\numTargetStructures$ (secondary) structures $\R = \{R_1,\dots, R_{\numTargetStructures}\}$, each abstracted as a set of base pairs, and
+\revisedOne{$\numFeatures\geq \numTargetStructures$} features $F_1,\dots, F_{\numFeatures}$\revisedOne{, typically
+representing the energies of the structures and additional sequence properties,} associated with weights
+$\pi_1,\dots,\pi_{\numFeatures}$ \revised{in $\mathbb{R}^+$}.
 Our goal is to sample sequences \revisedOne{$S$ (which satisfy the base pairing rules for all structures) from the Boltzmann distribution defined by}
@@
 \label{eq:sample-distribution}
-\Pr(S\mid\pi_1,\dots,\pi_k) \propto \displaystyle\prod_{1\leq \ell\leq k}\!\! \pi_\ell^{-F_\ell(S)} \end{equation}
+\Pr(S\mid\pi_1,\dots,\pi_{\numFeatures}) \propto \displaystyle\prod_{1\leq \ell\leq \numFeatures}\!\! \pi_\ell^{-F_\ell(S)} \end{equation}
 
@@
     \label{eq:mainproblem}
-    \partfun{\pi_1,\dots,\pi_k} = \sum_{S\in\Sigma^n} \prod_{1\leq \ell\leq k} \pi_\ell^{-F_\ell(S)},
+    \partfun{\pi_1,\dots,\pi_{\numFeatures}} = \sum_{S\in\Sigma^n} \prod_{1\leq \ell\leq \numFeatures} \pi_\ell^{-F_\ell(S)},
   \end{equation}
-  for specific weights $\pi_1,\dots,\pi_k$.
+  for specific weights $\pi_1,\dots,\pi_{\numFeatures}$.
 
@@
 $$
-\revisedOne{Here, since $\dep(f)=\{x_1,\ldots,x_p\}$, $\left\{\substack{x_1\mapsto S_{x_1}\\\cdots\\x_p\mapsto S_{x_p}}\right\}$ denotes the assignment, that assigns to all positions $x_k$  in $\dep(f)$ the respective nucleotide $S_{x_k}$ ($1\leq k\leq p$; $p=|\dep(f)|$).}
+\revisedOne{Here, since $\dep(f)=\{x_1,\ldots,x_p\}$, $\left\{\substack{x_1\mapsto S_{x_1}\\\cdots\\x_p\mapsto
+S_{x_p}}\right\}$ denotes the assignment, that assigns the respective nucleotides $S_{x_q}$ ($1\leq q\leq p$;
+$p=|\dep(f)|$)} \revised{to the positions $x_q$  in $\dep(f)$}.
 
@@
 
-        Energy models for structure prediction vary considerably, yet can always be expressed as sums over contributions associated with local structural motifs (base pairs, base pair stacks, loops, \ldots) under a certain nucleotide assignment.
-Energy models can thus be captured generically by introducing, for each motif $m$ occurring in a target structure, a contribution $f_m$, taking a specific value for each assignment of nucleotides to its positions $\dep(f_m)$.
+Energy models for structure prediction vary considerably, yet can always be expressed as sums over contributions
+associated with local structural motifs (base pairs, base pair stacks, loops, \ldots) under a certain nucleotide
+assignment. Energy models can thus be captured generically by introducing, for each motif $\motif$ occurring in
+a target structure, a contribution $f_{\motif}$, taking a specific value for each assignment of nucleotides to its
+positions $\dep(f_{\motif})$.
 For instance, the contribution of a \Def{base pair stack}, consisting of two pairs $(i,j)$ and $(i+1,j-1)$, can be captured by the introduction of a function $f^{\textsf{Stack}}_{\substack{i,j}}$ such that $\dep(f^{\textsf{Stack}}_{\substack{i,j}}) = \{i,i+1,j-1,j\}$.
@@
 To express this, let us define the \Def{base pair dependency graph} $G_{\R}$ as the graph with nodes $\{1,\dots,n\}$
-and edges $\bigcup_{\ell\in[1,k]} R_\ell$.
-%Any edge between two nodes $i$ and $j$ materializes the existence of a base pair between positions $i$ and $j$ in some structure, inducing a direct dependency between the chosen nucleotide assignments for the two positions.
+and edges $\bigcup_{\ell\in[1,\numTargetStructures]} R_\ell$.
 
@@
 The \Def{treewidth} of a tree decomposition $(T,\chi)$ is defined as
-$\min_{u\in T} |\chi(u)| - 1.$ 
+$\max_{u\in T} |\chi(u)| - 1.$ 
 %The \Def{treewidth} of a graph $G$ is the smallest treewidth of any tree decomposition of $G$.
@@
 Let us now specify the concrete set $\F$ of contributions that we use
-for the design in the base pair model targeting \GCb\% and structures $\R$ with weights $\pi_0,\dots,\pi_k$. The set $\F$ thus consists of
+for the design in the base pair model\todo{Didn't we plan to get rid of the bp model?}{} targeting \GCb\% and structures $\R$ with weights $\pi_0,\dots,\pi_k$. The set $\F$ thus consists of
 \begin{itemize}
@@
 on a tree decomposition of $G_{\F}$, the dependency graph associated with $\F$.
-Note, that analogous algorithms could be easily derived to count valid sequences, or list sequences having minimal free-energy.
+Note, that analogous algorithms could be easily derived to count valid sequences, or list sequences having minimum free
+energy.
 %This yields fixed parameter tractable (FPT) algorithms, based on the treewidth.
@@
 Since our algorithms iterate over specific sets of sequence positions, we moreover define
-the \Def{set $\partseqs(\Y)$ of all partial sequences determining the positions of $\Y=\subseteq\{1,\dots,n\}$} in all combinations of nucleotides $\{\Ab,\Cb,\Gb,\Ub\}$,
+the \Def{set $\partseqs(\Y)$ of all partial sequences determining the positions of $\Y\subseteq\{1,\dots,n\}$} in all combinations of nucleotides $\{\Ab,\Cb,\Gb,\Ub\}$,
 i.e.~for $\Y=\{y_1,\dots,y_{q}\}$,
@@
 the correctness of the algorithm
-(Supp.\ Mat.~\ref{appsec:correctness}).  After running
+(Additional file~1: Section~\ref{appsec:correctness}).  After running
 Alg.~\ref{alg:pf}, multiplying the 0-ary messages sent to the root $r$
@@
 tree decompositions
-(Supp.\ Mat.~\ref{appsec:algcomplexity}). W.l.o.g.\ we assume that
+(Additional file 1: Section~\ref{appsec:algcomplexity}). W.l.o.g.\ we assume that
 $D=1$; note that tree decompositions can generally be transformed,
@@
 The complexity of the precomputation can be further improved to
-$\mathcal{O}(n\,k\,2^{w+1}\,2^{c})$, where $c$ ($c\le w+1$) is the maximum number of connected components represented in a node of the tree decomposition (Supp.\ Mat.~\ref{appsec:improvedComplexity}).
+$\mathcal{O}(n\,k\,2^{w+1}\,2^{c})$, where $c$ ($c\le w+1$) is the maximum number of connected components represented in a node of the tree decomposition (Additional file~1: Section~\ref{appsec:improvedComplexity}).
 
@@
 negligible compared to Alg.~\ref{alg:pf}
-(Supp.\ Mat.~\ref{appsec:treedecomp} and
+(Additional file~1: Sections~\ref{appsec:treedecomp} and
 \ref{appsec:dependency-cliques}).
@@
 $\EnergyTurner$ and the fitted stacking model $\EnergyStacking$
-(Supp.\ Mat.~\ref{appsec:modelparameters}). More precisely, we observed a
+(Additional file~1: Section~\ref{appsec:modelparameters}). More precisely, we observed a
 structure-specific affine dependency between the Turner and stacking
@@
 The flexibility of our framework allows to support the advanced sampling technique called "multidimensional Boltzmann sampling"~\citep{Bodini2010}, which allows to enforce (probabilistically)  additional, complex properties of the samples through an additional rejection.
-This technique was previously used to control \GCb\%~\citep{Waldispuehl2011,Reinharz2013} and di-nucleotide content~\citep{Zhang2013} of sampled RNA sequences.
-Here, in addition to controlling \GCb\% (our feature $F_0$) we use it to target the free energies $(\TargetE_1,\ldots,\TargetE_k)$ of the individual target structures (features $F_1,\dots,F_m$).
+This technique was previously used to control \GCb\%~\citep{Waldispuehl2011,Reinharz2013} and di-nucleotide
+content~\citep{Zhang2013} of sampled RNA sequences. Here, in addition to controlling \GCb\% (our feature $F_0$) we use
+it to target the free energies $(\TargetE_1,\ldots,\TargetE_{\numTargetStructures})$ of the individual target structures (features
+$F_1,\dots,F_{\numFeatures}$).
 
@@
 In other words,  $\mathbb{E}(F_\ell(S)\mid \pmb{\pi})=\TargetE_\ell$,  $\forall\ell\in[1,k]$ and, analogously, the expectation of $F_0(S)$ is the targeted GC content.
-The expected value of $F_\ell$ is always decreasing for increasing weights $\pi_\ell$ (see Supp.\ Mat.~\ref{appsec:weight_derivatives}).
+The expected value of $F_\ell$ is always decreasing for increasing weights $\pi_\ell$ (see Additional file~1: Section~\ref{appsec:weight_derivatives}).
 More generally, computing a suitable parameter vector $\pmb{\pi}$ can be restated as a convex optimization problem, and be efficiently solved using a wide array of methods~\citep{Denise2010,Bendkowski2017}.
@@
 
-For a discussion on the implications of our hardness results beyond exact counting see Supp.\ Mat.~\ref{appsec:approximate-counting}.
+For a discussion on the implications of our hardness results beyond exact counting see Additional file~1: Section~\ref{appsec:approximate-counting}.
 }
@@
 	\includegraphics[width=0.8\textwidth]{Figs/Plots/doublestem_weight1-500_mean}
-	\caption{\revisedOne{Comparison of free-energy and Boltzmann probability for 10~000 uniform ($\pi=1$; blue dots) and Boltzmann distributed ($\pi=500$; green dots) sequences, targeting a simple structure consisting of two adjacent helices.}}
+	\caption{\revisedOne{Comparison of free energy and Boltzmann probability for 10~000 uniform ($\pi=1$; blue dots) and Boltzmann distributed ($\pi=500$; green dots) sequences, targeting a simple structure consisting of two adjacent helices.}}
         \label{fig:BoltzmannBenefits}
@@
 \revisedOne{First, we show how seed sequences can be generated in a Boltzmann distribution, leading to designs that are substantially more stable that those generated uniformly.
-As can be seen in Figure~\ref{fig:BoltzmannBenefits} and Supp.
-Mat.
-\ref{appsec:immediate-benefits-for-negative-design}, sequences generated in the Boltzmann distribution not only reach lower free-energies than those generated in a uniform setting, but also achieve better Boltzmann probabilities.
+As can be seen in Figure~\ref{fig:BoltzmannBenefits} and Additional file~1: Section~\ref{appsec:immediate-benefits-for-negative-design}, sequences generated in the Boltzmann distribution not only reach lower free-energies than those generated in a uniform setting, but also achieve better Boltzmann probabilities.
 While the former is expected since the Boltzmann distribution explicitly favors low-energy candidates, the latter is somewhat surprising, since the Boltzmann probability of a target structure could, in principle, decrease under Boltzmann sampling due to the partition function growing faster than the Boltzmann factor.
@@
 	\end{tabular}}\\[3em]
-{\sffamily {\bfseries C} -- Tree widths}\\
+{\sffamily {\bfseries C} -- Treewidths}\\
 \includegraphics[width=\textwidth,trim=350 0 0 50,clip]{Figs/td-widths}
@@
 Fortunately, we found that intricacy of the Turner energy model can be circumvented with minimal loss of precision by using a simpler stacking energy model.
-As shown in Figure~\ref{fig:correlation}, a simple stacking energy model, whose design principles are further described in Supp.
-Mat.~\ref{appsec:modelparameters}, can be used to approximate the Turner energy model very adequately (correlation coefficient $R=0.99$) in the context of sequence design.
+As shown in Figure~\ref{fig:correlation}, a simple stacking energy model, whose design principles are further described in Additional file~1: Section~\ref{appsec:modelparameters}, can be used to approximate the Turner energy model very adequately (correlation coefficient $R=0.99$) in the context of sequence design.
 Using this simpler model greatly reduces the treewidth, and thus the computational requirements of the whole method even for complex instances.
@@
 for its targeting performance (both models:
-Supp.\ Mat.~\ref{appsec:modelparameters}).
+Additional file~1: Section~\ref{appsec:modelparameters}).
 %
@@
 to a large extent, independently shift the Turner energies of sampled
-sequences towards prescribed targets. See Supp.\ Mat.~\ref{appsec:illustrating-mdbs} for a further example with three
+sequences towards prescribed targets. See Additional file~1: Section~\ref{appsec:illustrating-mdbs} for a further example with three
 pseudoknot-free target structures.}
@@
   \label{eq:blueprintobjective}
-    \Obj(S) = \frac{1}{k} \sum_{\ell=1}^{k} (E(S, R_\ell) - \EnsE(S))\\
-    +\frac{1}{2\binom{k}{2}} \sum\limits_{1\leq\ell<j\leq k}|E(S,R_\ell) - E(S,R_j)|,
+    \Obj(S) = \frac{1}{m} \sum_{\ell=1}^{m} (E(S, R_\ell) - \EnsE(S))\\
+    +\frac{1}{2\binom{m}{2}} \sum\limits_{1\leq\ell<j\leq m}|E(S,R_\ell) - E(S,R_j)|,
 \end{multline}
@@
 pseudoknotted targets, $\EnsE(S)$ is approximated by the \Def{minimum
-  free energy} (MFE) of $S$ as estimated by {\tt HotKnots}~\citep{Ren2005} in the energy model
+  free energy} of $S$ as estimated by {\tt HotKnots}~\citep{Ren2005} in the energy model
 of~\citet{dirks-pierce-03}.  Intuitively, the first term of \Obj{}
@@
 while the second term penalizes the dispersion of targets; \Obj{} is
-best (minimal) when all targets simultaneously achieve the minimum
+best (minimized) when all targets simultaneously achieve the minimum
 free energy of the sequence.
@@
 \revisedOne{%
-Supp.\ Mat.~\ref{appsec:effectivity-seed-generation} provides detailed results from this iterative procedure, which works similarly to the previously described multi-dimensional Boltzmann sampling.
-In particular, we observe that most of the benchmark inputs finish within only few iterations, where each iteration requires  little time (confer Supp.\ Mat.~\ref{appsec:run-times}).}
+Additional file~1: Section~\ref{appsec:effectivity-seed-generation} provides detailed results from this iterative procedure, which works similarly to the previously described multi-dimensional Boltzmann sampling.
+In particular, we observe that most of the benchmark inputs finish within only few iterations, where each iteration requires  little time (confer Additional file~1: Section~\ref{appsec:run-times}).}
 
@@
 improvement is observed for both terms in \Obj{} (see
-Supp. Mat.~\ref{appsec:contributions-multidefect}).  This means that \ourprog{}
-produces sequences whose targets are both, substantially closer to
-the sequence ensemble/minimal free energy, and more similarly stable across targets.  In fact, for
+Additional file~1: Section~\ref{appsec:contributions-multidefect}).  This means that \ourprog{}
+produces sequences whose targets are substantially closer to
+the ensemble free energy \emph{and} have more similar stability across targets.  In fact, for
 every sequence in our benchmark, consisting of 332 sets of target
 structures, we observed better \Obj{} for Boltzmann sampling than for
-uniform sampling (see Supp. Mat.~\ref{appsec:detailed-benchmark-results}).  Notably,
+uniform sampling (see Additional file~1: Section~\ref{appsec:detailed-benchmark-results}).  Notably,
 \ourprog{} performs equally well in the presence of pseudoknots;
 the difficulty rather lies in the computation of the \Obj{} function, since
-free-energy minimization is costly in the presence of
+free energy minimization is costly in the presence of
 pseudoknots~\citep{Sheikh2012} and good implementations are scarce.
@@
 remain superior even after local optimizations, as shown by
-Fig.~\ref{fig:benchmark-results} and Supp. Mat.~\ref{appsec:contributions-multidefect}
+todo~\ref{fig:benchmark-results} and Additional file~1: Section~\ref{appsec:contributions-multidefect}
 and \ref{appsec:detailed-benchmark-results}. This observation is consistent with a
@@
 
+\section*{Additional Files}
+
+\subsection*{Additional file 1}
+\revised{Supplemental Material providing additional detailed information on 
+``Immediate benefits of positive design for negative design'', ``Tree decomposition for RNA design instances in
+practice'', ``Run times for Boltzmann sampling (at fixed weights)'', ``Illustrating multi-dimensional Boltzmann sampling
+for three pseudoknot-free target structures'', ``Effectivity of the seed sequence generation'', ``Parameters for the
+base pair and the stacking energy model'', ``Tree decomposition with hyper-dependencies'', ``Correctness of the FPT
+partition function algorithm'', ``General complexity of the partition function computation by cluster tree elimination
+and generation of samples'', ``Exploiting constraint consistency to reduce the complexity'', ``Monotonicity of the
+partial derivatives within weight calibration'', ``Approximate counting and random generation'', ``Contribution of terms
+to the MultiDefect score'', and ``Detailed results of the negative design benchmark''.}
+
 \begin{backmatter}
-
-\section*{Competing interests}
-  The authors declare that they have no competing interests.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                               %%
+%% Additional Files              %%
+%%                               %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section*{Abbreviations}
+\GCb\%: GC-content; \#P: complexity class ``Sharp P''; DP: dynamic programming; FPT: fixed-parameter tractable; D\&P
+model: pseudoknot energy model by Dirks and Pierce
 
@@
 The software and data of this work are available in the Github repository \url{https://github.com/yannponty/RNARedPrint}.
+\todo[inline]{This section should include a link to the most recent version of your software or code (e.g. GitHub or
+  Sourceforge) as well as a link to the archived version referenced in the manuscript. The software or code should be
+  archived in an appropriate repository with a DOI or other unique identifier. For software in GitHub, we recommend
+using Zenodo.}
 
@@
 
-\section*{Acknowledgements}
+\section*{Funding}
 YP is supported
@@
 (FET-Open grant 323987; project RiboNets).
+\revised{The funding bodies did not played any roles in the design
+of the study and collection, analysis, and interpretation of data and in writing
+the manuscript.}
+
+\section*{Acknowledgements}
 %
 We thank Leonid Chindelevitch for suggesting a drastic optimization of our FPT algorithm based on stable sets combinatorics, and Arie Koster for practical recommendations on tree decompositions.
+\revised{A preliminary version of this work was presented at the conference RECOMB 2019 in Paris; we acknowledge the
+contributions of the anonymous reviewers.}
+
+\section*{Ethics approval and consent to participate}
+Not applicable.
+
+\section*{Consent to publish}
+Not applicable.
+
+\section*{Competing interests}
+  The authors declare that they have no competing interests.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                               %%
-%% Additional Files              %%
-%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% \section*{Additional Files}
-%   \subsection*{Additional file 1 --- Sample additional file title}
-%     Additional file descriptions text (including details of how to
-%     view the file, if it is in a non-standard format or the file extension).  This might
-%     refer to a multi-page table or a figure.
-%
-%  \subsection*{Additional file 2 --- Sample additional file title}
-%    Additional file descriptions text.
-
 \end{backmatter}

### 2019-03-26 RNARedPrint.tex
Fix remaining todos for R2

@@
 
-\usepackage{todonotes}
-
 %\hypersetup{
@@
 For instance, the contribution of a \Def{base pair stack}, consisting of two pairs $(i,j)$ and $(i+1,j-1)$, can be captured by the introduction of a function $f^{\textsf{Stack}}_{\substack{i,j}}$ such that $\dep(f^{\textsf{Stack}}_{\substack{i,j}}) = \{i,i+1,j-1,j\}$.
-We refer to energy models that consider the contributions of all base pair stacks (and thus introduce the corresponding dependencies) collectively as the \Def{stacking energy model}.
+We refer to energy models that consider the contributions of all base pair stacks (and thus introduce the corresponding
+dependencies) collectively as the \Def{stacking energy model} \revised{(briefly, \Def{stacking model})}.
 }
@@
 Let us now specify the concrete set $\F$ of contributions that we use
-for the design in the base pair model\todo{Didn't we plan to get rid of the bp model?}{} targeting \GCb\% and structures $\R$ with weights $\pi_0,\dots,\pi_k$. The set $\F$ thus consists of
+for the design in the stacking energy model targeting \GCb\% and structures $\R$ with weights $\pi_0,\dots,\pi_k$. The set $\F$ thus consists of
 \begin{itemize}
 \item the transformed contributions $\ln(\pi_0)f^{\GCb}_i$ for the ~$\GCb\%$ feature ($i=1,\dots,n$);
-\item the transformed contributions $\ln(\pi_\ell)f^{R_\ell}_{ij}$for each structure $R_\ell\in\R$ and $(i,j)\in R_\ell.$
+\item the transformed contributions $\ln(\pi_\ell)f^{{\textsf{Stack}}}_{ij}$ for each structure $R_\ell\in\R$ and $(i,j)\in R_\ell.$
 \end{itemize}
-
+%
 By these definitions, the set $\F$ encodes the partition function $\partfun{\pi_0,\dots,\pi_k}$ of Eq.~(\ref{eq:mainproblem}).
@@
 \begin{theorem}[Complexities]\label{th:complexities}
-  For sequence length $n$, $k$ target structures and treewidth $w$,
+  \revised{Given are sequence length $n$, $k$ target structures, and tree\-width $w$.
   % and a base pair dependency graph having $c$ connected components%
@@
   are generated from the Boltzmann distribution  in
-  $O( n\, k \, 4^{w+1} + t\, n\, k )$ time.
+  $O( n\, k \, 4^{w+1} + t\, n\, k )$ time.}
   %$O( n\, k \min(2^{w+c+1},4^{w+1}) + t\, n\, k )$ time.
@@
 
-In order to capture realistic energy models like the Turner model or pseudoknot models like {\tt HotKnots}~\citep{Ren2005}, our sampling strategy can be extended in two ways: 1) either by directly sampling based on more expressive energy models \revisedOne{or 2) by sampling in a simple energy model which} can be used to approximate sampling in more complex models.
-In practice, complex energy models \revisedOne{have} a strong influence on the treewidth (of optimal tree decompositions) of the dependency graph and \revisedOne{thus on} the computational complexity of our approach.
-Therefore, it is interesting to consider---in addition to the base pair model---other
-stripped-down variants of the nearest neighbor model, which could offer a compromise between the low-complexity base pair energy model and the high-accuracy Turner model.
+In order to capture realistic energy models like the Turner model or pseudoknot models like {\tt
+HotKnots}~\citep{Ren2005}, our sampling strategy can be extended in two ways: 1) either by directly sampling based on
+more expressive energy models \revisedOne{or 2) by sampling in a simple energy model which} can be used to approximate
+sampling in more complex models.  In practice, complex energy models \revisedOne{have} a strong influence on the
+treewidth (of optimal tree decompositions) of the dependency graph and \revisedOne{thus on} the computational complexity
+of our approach. Therefore, it is interesting to consider---in addition to the \revised{stacking energy model}---other
+stripped-down variants of the nearest neighbor model, which could offer a compromise between \revised{low-complexity (as due to
+the stacking energy model) and the high-accuracy of the Turner model.}
 
@@
 \begin{proof}
-Consider the mapping $\Psi: \Design{G} \to \IS{G}, f \mapsto \left\{v\in V\mid f(v)\in\{\Ab,\Cb\}\right\}.$
+  Consider the mapping\\\hspace*{2em}$\Psi: \Design{G} \to \IS{G}, f \mapsto \left\{v\in V\mid f(v)\in\{\Ab,\Cb\}\right\}.$\\
 We show that $\Psi$ is bijective:%
@@
 \ourprog{}, available at: 
-
-{\centering \url{https://github.com/yannponty/RNARedPrint}\\[.3em]}%
-
+%
+{\url{https://github.com/yannponty/RNARedPrint}}.%
+
+\noindent
 \ourprog{}  takes a set of target structures, as well as
 weights for each energy feature and \GCb\%, and generates a sample set of sequences compatible with the structures in the corresponding
-Boltzmann distribution. \ourprog{} currently supports the base pair
-energy model and the stacking energy model.
+Boltzmann distribution\revised{; it currently supports the stacking energy model and the base pair energy model
+(Additional file~1: Section~\ref{appsec:modelparameters}).}
 
@@
 Turner energy model, and are targeted by efficiently sampling in the
-base pair energy model, and filtering sequences based on the {\tt RNAeval} tool from the Vienna package~\cite{Lorenz2011}.
+stacking energy model, and filtering sequences based on the {\tt RNAeval} tool from the Vienna package~\cite{Lorenz2011}.
 The second script generates high quality seed sequences suitable for negative RNA design.
@@
 
-\section*{Additional Files}
+\section*{Additional file}
 
 \subsection*{Additional file 1}
-\revised{Supplemental Material providing additional detailed information on 
-``Immediate benefits of positive design for negative design'', ``Tree decomposition for RNA design instances in
-practice'', ``Run times for Boltzmann sampling (at fixed weights)'', ``Illustrating multi-dimensional Boltzmann sampling
-for three pseudoknot-free target structures'', ``Effectivity of the seed sequence generation'', ``Parameters for the
-base pair and the stacking energy model'', ``Tree decomposition with hyper-dependencies'', ``Correctness of the FPT
-partition function algorithm'', ``General complexity of the partition function computation by cluster tree elimination
-and generation of samples'', ``Exploiting constraint consistency to reduce the complexity'', ``Monotonicity of the
-partial derivatives within weight calibration'', ``Approximate counting and random generation'', ``Contribution of terms
-to the MultiDefect score'', and ``Detailed results of the negative design benchmark''.}
+\revised{The Supplemental Material contains additional
+information on methods and parameters; elaboration of theory and proofs; and additional and detailed empirical results.}
+% 
+% \revised{Supplemental Material providing additional detailed information on 
+% ``Immediate benefits of positive design for negative design'', ``Tree decomposition for RNA design instances in
+% practice'', ``Run times for Boltzmann sampling (at fixed weights)'', ``Illustrating multi-dimensional Boltzmann sampling
+% for three pseudoknot-free target structures'', ``Effectivity of the seed sequence generation'', ``Parameters for the
+% base pair and the stacking energy model'', ``Tree decomposition with hyper-dependencies'', ``Correctness of the FPT
+% partition function algorithm'', ``General complexity of the partition function computation by cluster tree elimination
+% and generation of samples'', ``Exploiting constraint consistency to reduce the complexity'', ``Monotonicity of the
+% partial derivatives within weight calibration'', ``Approximate counting and random generation'', ``Contribution of terms
+% to the MultiDefect score'', and ``Detailed results of the negative design benchmark''.}
 
@@
 \section*{Availability of data and materials}
-The software and data of this work are available in the Github repository \url{https://github.com/yannponty/RNARedPrint}.
-\todo[inline]{This section should include a link to the most recent version of your software or code (e.g. GitHub or
-  Sourceforge) as well as a link to the archived version referenced in the manuscript. The software or code should be
-  archived in an appropriate repository with a DOI or other unique identifier. For software in GitHub, we recommend
-using Zenodo.}
+The software and data of this work are maintained in the Github repository
+\url{https://github.com/yannponty/RNARedPrint}\revised{; the acrchived version referenced in this manuscript is
+available at \href{https://dx.doi.org/10.5281/zenodo.2597571}{doi:10.5281/zenodo.2597571}.}
 

### 2019-03-26 RNARedPrint.tex
Updates from Overleaf

--- 2019-03-13 RNARedPrint.tex
@@
 
-\documentclass[doublespacing,linenumbers]{bmcart}% uncomment this for twocolumn layout and comment line below
-%\documentclass[]{bmcart}
+%\documentclass[doublespacing,linenumbers]{bmcart}% uncomment this for twocolumn layout and comment line below
+\documentclass[]{bmcart}
 %\documentclass[twocolumn]{bmcart}
@@
 
-\usepackage{graphicx}
-%\usepackage{endfloat}
-%\usepackage{caption}
-%\captionsetup{labelsep=none,textformat=empty}
-
 %%% Put your definitions there:
@@
 %% Rolf's includegraphicstop
-% \makeatletter
-% \newsavebox{\@alignepsbox}
-% \newlength{\@aligneps}
-% \newcommand{\includegraphicstop}[2][]{%
-% \sbox{\@alignepsbox}{\includegraphics[#1]{#2}}%
-% \setlength{\@aligneps}{-\ht\@alignepsbox}%
-% \addtolength{\@aligneps}{2ex}%
-% \raisebox{\@aligneps}{\usebox{\@alignepsbox}}}
-% \makeatother
+\makeatletter
+\newsavebox{\@alignepsbox}
+\newlength{\@aligneps}
+\newcommand{\includegraphicstop}[2][]{%
+\sbox{\@alignepsbox}{\includegraphics[#1]{#2}}%
+\setlength{\@aligneps}{-\ht\@alignepsbox}%
+\addtolength{\@aligneps}{2ex}%
+\raisebox{\@aligneps}{\usebox{\@alignepsbox}}}
+\makeatother
 
@@
 
-%% revised for first revision
-\newcommand{\revisedOne}[1]{#1}
-%%revised for second minor/discretionary revision
 \newcommand{\revised}[1]{{\color{red!60!black} #1}}
+%\renewcommand{\revised}[1]{{\color{red!60!black} #1}}
 
@@
   The design of multi-stable RNA molecules has important applications in biology, medicine, and biotechnology.
-Synthetic design approaches profit strongly from effective in-silico methods, \revisedOne{which substantially reduce the need for costly wet-lab experiments.} \\
+Synthetic design approaches profit strongly from effective in-silico methods, \revised{which substantially reduce the need for costly wet-lab experiments.} \\
 %
   \parttitle{Results} We devise a novel approach to a central ingredient of most in-silico
-  design methods: the generation of sequences \revisedOne{that fold well into multiple target structures.} 
-  % 
-  \revisedOne{Based on constraint networks, our approach \ourprog{}} supports generic
+  design methods: the generation of sequences \revised{that fold well into multiple target structures.} 
+  % 
+  \revised{Based on constraint networks, our approach \ourprog{}} supports generic
   Boltzmann-weighted sampling, which enables the positive design of RNA sequences
@@
   biologically relevant multi-target Boltzmann-weighted designs for a
-  \revisedOne{established} design benchmark. \revisedOne{Our results demonstrate the efficacy and feasibility of the method in practice
+  \revised{established} design benchmark. \revised{Our results demonstrate the efficacy and feasibility of the method in practice
   as well as the benefits of Boltzmann sampling over the previously best multi-target sampling strategy---even for the case of negative design of multi-stable RNAs.}
 %
-  \revisedOne{Besides empirically studies, we finally justify the algorithmic details due to a fundamental theoretic result about multi-stable RNA design, namely the \#P-hardness of the counting of designs.}
-%
-  \parttitle{Conclusion} \ourprog{} introduces a novel, flexible, and effective approach to multi-target RNA design, which promises broad applicability and extensibility.
+  \revised{Besides empirically studies, we finally justify the algorithmic details due to a fundamental theoretic result about multi-stable RNA design, namely the \#P-hardness of the counting of designs.}
+%
+  \parttitle{Conclusion} \ourprog{} introduces a novel, flexible and---as we demonstrate---effective approach to multi-target RNA design, which promises broad applicability and extensibility.
 
@@
 design~\citep{Wachsmuth2013,Domin2017}, partly motivated by their
-capacity to act as biosensors~\citep{Findeiss2017}, \revisedOne{which suggests them for biotechnological applications.}
+capacity to act as biosensors~\citep{Findeiss2017}, \revised{which suggests them for biotechnological applications.}
 In particular due to the kinetic coupling of RNA folding with RNA transcription, RNA families can feature alternative,
@@
 %
-Faced with the complex constraints due to the multiple targets, early methods such as \Software{Frnakenstein}~\citep{Lyngsoe2012} and \Software{Modena}~\citep{Taneda2015} \revisedOne{do not even attempt to sample sequences systematically from a controlled distribution}, but rely on ad-hoc generation strategies.
+Faced with the complex constraints due to the multiple targets, early methods such as \Software{Frnakenstein}~\citep{Lyngsoe2012} and \Software{Modena}~\citep{Taneda2015} \revised{do not even attempt to sample sequences systematically from a controlled distribution}, but rely on ad-hoc generation strategies.
 %
@@
 target structures. \Software{RNAdesign} adopts a graph coloring perspective,
-\revisedOne{assigning nucleotide symbols (like ``colors'') to the sequence positions, such that compatible nucleotides are assigned to the ends of each base pair.} Initially, the method decomposes the graph hierarchically and then \Def{precomputes} the number of valid sequences
+\revised{assigning nucleotide symbols (like ``colors'') to the sequence positions, such that compatible nucleotides are assigned to the ends of each base pair.} Initially, the method decomposes the graph hierarchically and then \Def{precomputes} the number of valid sequences
 within each subgraph. The decomposition is then reinterpreted as a
@@
 As well, the \Software{RNAdesign}/\RNAblueprint{} approach is specialized to uniform sampling, which limits its direct extensibility.
-\revisedOne{Substantial improvements of multi-target sampling thus require a systematically redesigned approach.
+\revised{Substantial improvements of multi-target sampling thus require a systematically redesigned approach.
 To enable a fundamentally broader range of applications in extensions of the sampling method, we build our approach, from the start, on established concepts in computer science.}
@@
 The tree is then used to compute the partition function, followed by a Boltzmann sampling of valid sequences (v).
-An adaptive scheme learns weights to achieve targeted energies and \GCb\% \revisedOne{(arrows)}, leading to the production of suitable designs (vi).
-\revisedOne{Note that for simplicity, we assume in this figure that only dependencies between the ends of base pairs are considered to evaluate the energies of structures.
+An adaptive scheme learns weights to achieve targeted energies and \GCb\% \revised{(arrows)}, leading to the production of suitable designs (vi).
+\revised{Note that for simplicity, we assume in this figure that only dependencies between the ends of base pairs are considered to evaluate the energies of structures.
 Our computations based on a more complex energy model, which considers energy contributions of base pair stacks, require additional dependencies.}
@@
 
-\revisedOne{As central contribution, we provide a systematic and flexibly extensible technique for sampling that targets multiple versatile features.
+\revised{As central contribution, we provide a systematic and flexibly extensible technique for sampling that targets multiple versatile features.
 For the sake of clarity, we introduce this method specialized to the sampling of RNA sequences that have specific energies for multiple structures and specific \GCb\%.}
 
-\revisedOne{Figure~\ref{fig:workflow} summarizes our generic framework, which enables this targeted sequence generation based on multi-dimensional Boltzmann sampling.
+\revised{Figure~\ref{fig:workflow} summarizes our generic framework, which enables this targeted sequence generation based on multi-dimensional Boltzmann sampling.
 \emph{Algorithmically}, we originally contribute} dynamic programming (DP) algorithms, based on the concept of \Def{tree decomposition}, to compute partition functions and sample sequences from the Boltzmann distribution%
@@
 .
-\revisedOne{Generally, tree decompositions are data structures that capture the specific dependencies of a problem instance (here, the dependencies between sequence positions induced by the target structures), such that they can guide the efficient processing by DP algorithms.}
-%
-\revisedOne{Building on this principle, the complexities of our algorithms depend exponentially on a specific property of the tree decomposition, called the \Def{treewidth}.
+\revised{Generally, tree decompositions are data structures that capture the specific dependencies of a problem instance (here, the dependencies between sequence positions induced by the target structures), such that they can guide the efficient processing by DP algorithms.}
+%
+\revised{Building on this principle, the complexities of our algorithms depend exponentially on a specific property of the tree decomposition, called the \Def{treewidth}.
 Thus, it is essential for the applicability of our approach that---by appropriate design choices---we can keep this parameter low for typical instances.
@@
 
-\revisedOne{Remarkably, we could show that it is not possible to find a better, efficient method for sampling (unless ${\sf P}={\sf NP}$), since the underlying counting problem is \#{\sf P}-hard.
+\revised{Remarkably, we could show that it is not possible to find a better, efficient method for sampling (unless ${\sf P}={\sf NP}$), since the underlying counting problem is \#{\sf P}-hard.
 The practical relevance of this theoretical result is that it rules out substantially better sampling techniques.
@@
 %
-Due to the generality of our method, we can moreover strongly limit \revisedOne{the treewidth} in practice by using state-of-the-art tree decomposition algorithms.
+Due to the generality of our method, we can moreover strongly limit \revised{the treewidth} in practice by using state-of-the-art tree decomposition algorithms.
 %Uniform sampling is handled as a special case of Boltzmann sampling, where each valid sequence receives energy zero and---consequently---computing partition functions specializes to counting.
 By evaluating sequences in a specialized weighted constraint
-network, we support---in principle---arbitrary \revisedOne{complex constraints and energy models},
+network, we support---in principle---arbitrary \revised{complex constraints and energy models},
 notably subsuming the commonly
@@
 Turner RNA energy model
-works well by performing sampling based on \revisedOne{a simplified RNA energy model, which induces much lower treewidth than the Turner model}.
-This result is \revisedOne{essential for the applicability of our method, since it allows to combine high efficiency (by keeping the treewidth low) with sufficient accuracy to precisely target realistic Turner energies.}
-
-\revisedOne{Eventually, our proof-of-concept results 
+works well by performing sampling based on \revised{a simplified RNA energy model, which induces much lower treewidth than the Turner model}.
+This result is \revised{essential for the applicability of our method, since it allows to combine high efficiency (by keeping the treewidth low) with sufficient accuracy to precisely target realistic Turner energies.}
+
+\revised{Eventually, our proof-of-concept results 
  on a comprehensive multi-target RNA design benchmark~\citep{Taneda2015} 
@@
 \section*{Methods}
-\revisedOne{The main computational problem addressed in this work is the positive design of RNA sequences for multiple target structures; more specifically, the generation of sequences over the alphabet $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}$, such that the sequences feature a given $\GCb\%$, and have prescribed energies for a set of target secondary structures.
+\revised{The main computational problem addressed in this work is the positive design of RNA sequences for multiple target structures; more specifically, the generation of sequences over the alphabet $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}$, such that the sequences feature a given $\GCb\%$, and have prescribed energies for a set of target secondary structures.
 Here, these desired sequence properties are modeled as constraints on the values of \Def{features}, which are functions of the sequence that are expressed as sums over real-valued \Def{contributions}.
@@
 
-\revisedOne{To generate diverse design candidates, we randomly generate sequences from a Boltzmann distribution.
+\revised{To generate diverse design candidates, we randomly generate sequences from a Boltzmann distribution.
 The probability of a sequence then depends on its features (e.g.
@@
 %\paragraph{Central computational problem.}
-Let us consider $m$ (secondary) structures $R_1,\dots, R_m$, each abstracted as a set of base pairs, and \revisedOne{$k\geq m$} features $F_1,\dots, F_k$\revisedOne{, typically representing the energies of the structures and additional sequence properties,} associated with weights $\pi_1,\dots,\pi_k$.
-Our goal is to sample sequences \revisedOne{$S$ (which satisfy the base pairing rules for all structures) from the Boltzmann distribution defined by}
+Let us consider $m$ (secondary) structures $R_1,\dots, R_m$, each abstracted as a set of base pairs, and \revised{$k\geq m$} features $F_1,\dots, F_k$\revised{, typically representing the energies of the structures and additional sequence properties,} associated with weights $\pi_1,\dots,\pi_k$.
+Our goal is to sample sequences \revised{$S$ (which satisfy the base pairing rules for all structures) from the Boltzmann distribution defined by}
 \begin{equation}
@@
 
-The workhorse of our approach is the \revisedOne{fixed-parameter tractable} computation of feature-dependent partition functions over sequences, \revisedOne{namely partition functions of the form} 
+The workhorse of our approach is the \revised{fixed-parameter tractable} computation of feature-dependent partition functions over sequences, \revised{namely partition functions of the form} 
 %
@@
 Formally, we define a \Def{feature $F$} as a function on sequences, whose value is obtained by summing over an associated set of \Def{contributions}.
-Each contribution $f$ takes values in $\mathbb{R}\cup \{+\infty\}$, and depends on the nucleotides assigned to a restricted set of positions, \revisedOne{namely its \emph{dependencies}, denoted $\dep(f)$}, such that
+Each contribution $f$ takes values in $\mathbb{R}\cup \{+\infty\}$, and depends on the nucleotides assigned to a restricted set of positions, \revised{namely its \emph{dependencies}, denoted $\dep(f)$}, such that
 
@@
 $$
-\revisedOne{Here, since $\dep(f)=\{x_1,\ldots,x_p\}$, $\left\{\substack{x_1\mapsto S_{x_1}\\\cdots\\x_p\mapsto S_{x_p}}\right\}$ denotes the assignment, that assigns to all positions $x_k$  in $\dep(f)$ the respective nucleotide $S_{x_k}$ ($1\leq k\leq p$; $p=|\dep(f)|$).}
-
-The $\GCb\%$ can be simply expressed using $n$ contributions $f^{\GCb}_i$, \revisedOne{each depending only on position $i\in[1,n]$, i.e.~$\dep(f^{\GCb}_i)=\{i\}$}, such that 
+\revised{Here, since $\dep(f)=\{x_1,\ldots,x_p\}$, $\left\{\substack{x_1\mapsto S_{x_1}\\\cdots\\x_p\mapsto S_{x_p}}\right\}$ denotes the assignment, that assigns to all positions $x_k$  in $\dep(f)$ the respective nucleotide $S_{x_k}$ ($1\leq k\leq p$; $p=|\dep(f)|$).}
+
+The $\GCb\%$ can be simply expressed using $n$ contributions $f^{\GCb}_i$, \revised{each depending only on position $i\in[1,n]$, i.e.~$\dep(f^{\GCb}_i)=\{i\}$}, such that 
 $$f^{\GCb}_i(\{i\mapsto c\})=\begin{cases}-1&\text{ if }c=\Gb\text{ or }\Cb\\ 0&\text{ otherwise}.\end{cases}$$
 
-\revisedOne{By summing $f^{\GCb}_i(\{i\mapsto S_i\})$ over the whole sequence ($i=1,\dots,n$), one simply counts the occurrences of $\Gb$ and $\Cb$.}
-
-\revisedOne{
+\revised{By summing $f^{\GCb}_i(\{i\mapsto S_i\})$ over the whole sequence ($i=1,\dots,n$), one simply counts the occurrences of $\Gb$ and $\Cb$.}
+
+\revised{
   To start with a simple example of evaluating the energy of sequences by features, let us explain how they are used to count the number of \Def{valid sequences}, \emph{i.e.} sequences inducing only base pairs in $\mathcal{B}:=\{\{\Ab,\Ub\}, \{\Gb,\Cb\}, \{\Gb,\Ub\}\}$.
@@
 
-\revisedOne{In order to compute the partition function of Eq.~\ref{eq:mainproblem}, and thus sample in a well-defined way, one must consider dependencies induced by the complete set of contributions $$\F := \bigcup_{\ell}\{f\mid f\text{ contribution of } F_\ell\}.$$}
+\revised{In order to compute the partition function of Eq.~\ref{eq:mainproblem}, and thus sample in a well-defined way, one must consider dependencies induced by the complete set of contributions $$\F := \bigcup_{\ell}\{f\mid f\text{ contribution of } F_\ell\}.$$}
 
@@
 
-\revisedOne{
+\revised{
 In the simplest case, this set captures the requirement of canonical base pairing for each structure.
@@
 %
-\revisedOne{Intuitively, a tree decomposition of an (hyper)graph $G$ is a tree that captures all the vertices and (hyper)edges of $G$, and properly relates dependent sub-problems to ensure consistency in a recursive computation.
+\revised{Intuitively, a tree decomposition of an (hyper)graph $G$ is a tree that captures all the vertices and (hyper)edges of $G$, and properly relates dependent sub-problems to ensure consistency in a recursive computation.
 Figure~\ref{fig:treedec} shows an optimal tree decomposition for a pair of structures under the stacking energy model.}
@@
 {\centering \includegraphics[width=.9\linewidth]{Figs/example-treedec-stacking}\\}
-\caption{\revisedOne{Toy example of a tree decomposition associated with two target structures in the stacking energy model (where the four positions of each base pair stack depend on each other).
+\caption{\revised{Toy example of a tree decomposition associated with two target structures in the stacking energy model (where the four positions of each base pair stack depend on each other).
 Two target secondary structures ({\bfseries\sffamily A}) are merged into a joint hypergraph ({\bfseries\sffamily B}), whose hyperedges correspond to the quadruplets of positions involved in base pair stacks (colored).
@@
 Let us call the latter object a \Def{partial sequence}.
-Such an object will help \revisedOne{to} specify partial knowledge on the sequence at some point of the algorithm.
+Such an object will help \revised{to} specify partial knowledge on the sequence at some point of the algorithm.
 Easily, we can extend the definition of contributions $f$ to sets $\{x_1\mapsto v_1, \dots, x_p\mapsto v_p\}$, where $\{x_1\dots x_p\}$ is any super-set of $\dep(f)$ by ignoring the superfluous assignments $x\mapsto v$, where $x\not\in\dep(f)$.
@@
 
-In order to capture realistic energy models like the Turner model or pseudoknot models like {\tt HotKnots}~\citep{Ren2005}, our sampling strategy can be extended in two ways: 1) either by directly sampling based on more expressive energy models \revisedOne{or 2) by sampling in a simple energy model which} can be used to approximate sampling in more complex models.
-In practice, complex energy models \revisedOne{have} a strong influence on the treewidth (of optimal tree decompositions) of the dependency graph and \revisedOne{thus on} the computational complexity of our approach.
+In order to capture realistic energy models like the Turner model or pseudoknot models like {\tt HotKnots}~\citep{Ren2005}, our sampling strategy can be extended in two ways: 1) either by directly sampling based on more expressive energy models \revised{or 2) by sampling in a simple energy model which} can be used to approximate sampling in more complex models.
+In practice, complex energy models \revised{have} a strong influence on the treewidth (of optimal tree decompositions) of the dependency graph and \revised{thus on} the computational complexity of our approach.
 Therefore, it is interesting to consider---in addition to the base pair model---other
@@
 %We define the following notations: $\real^\infty:=\real\cup\{\infty\}$, $[i,j]:=\{i,\ldots,j\},$ \ldots
-\revisedOne{While efficient, both in practice and in theory for graphs of bounded treewidth, our algorithms remain exponential in the worst case scenario, since the treewidth of a dependency graph can then become arbitrarily large.
+\revised{While efficient, both in practice and in theory for graphs of bounded treewidth, our algorithms remain exponential in the worst case scenario, since the treewidth of a dependency graph can then become arbitrarily large.
 This exponential complexity in the worst case appears to be intrinsic.
@@
 
-\revisedOne{Proposition~\ref{prop:bijection} also strongly impacts the complexity of computing the partition function.
+\revised{Proposition~\ref{prop:bijection} also strongly impacts the complexity of computing the partition function.
 Indeed it implies that, among the $4^k$ possible assignments of nucleotides to $k$ connected positions (in the base-pair dependency graph), at most $2^k$ are compatible with base pairing rules.
@@
 \begin{figure}
-	\includegraphics[width=0.8\textwidth]{Figs/Plots/doublestem_weight1-500_mean}
-	\caption{\revisedOne{Comparison of free-energy and Boltzmann probability for 10~000 uniform ($\pi=1$; blue dots) and Boltzmann distributed ($\pi=500$; green dots) sequences, targeting a simple structure consisting of two adjacent helices.}}
-        \label{fig:BoltzmannBenefits}
+	\includegraphicstop[width=0.8\textwidth]{Figs/Plots/doublestem_weight1-500_mean}
+	\caption{\revised{Comparison of free-energy and Boltzmann probability for 10~000 uniform ($\pi=1$; blue dots) and Boltzmann distributed ($\pi=500$; green dots) sequences, targeting a simple structure consisting of two adjacent helices.}}\label{fig:BoltzmannBenefits}
 \end{figure}
 
-\revisedOne{First, we show how seed sequences can be generated in a Boltzmann distribution, leading to designs that are substantially more stable that those generated uniformly.
+\revised{First, we show how seed sequences can be generated in a Boltzmann distribution, leading to designs that are substantially more stable that those generated uniformly.
 As can be seen in Figure~\ref{fig:BoltzmannBenefits} and Supp.
@@
 While the former is expected since the Boltzmann distribution explicitly favors low-energy candidates, the latter is somewhat surprising, since the Boltzmann probability of a target structure could, in principle, decrease under Boltzmann sampling due to the partition function growing faster than the Boltzmann factor.
-The empirical superiority of Boltzmann sampling appears robust to the target structure length and topology, as demonstrated by prior work~\cite{Reinharz2013}.}
+The empirical superiority of Boltzmann sampling appears robust to the target structure length and topology, as demonstrated by prior work~\cite{Reinharz2013}.
 
@@
 	{\sffamily {\bfseries B} -- Correlation plot}\\
-	  	\includegraphics[width=\textwidth,trim=0 0 0 50,clip]{Figs/stackingcor} 
+	  	\includegraphicstop[width=\textwidth,trim=0 0 0 50,clip]{Figs/stackingcor} 
   	\end{minipage}
 
-        \caption{\revisedOne{A fitted energy model based on stacking pairs ({\sffamily\bfseries A}) leads to approximated free-energies that are highly correlated (R=0.99) with free-energies in the Turner energy model ({\sffamily\bfseries B}), yet induces tree widths that are amenable to practical sampling ({\sffamily\bfseries C})}.}
-      \label{fig:correlation}
+	\caption{\revised{A fitted energy model based on stacking pairs ({\sffamily\bfseries A}) leads to approximated free-energies that are highly correlated (R=0.99) with free-energies in the Turner energy model ({\sffamily\bfseries B}), yet induces tree widths that are amenable to practical sampling ({\sffamily\bfseries C})}.\label{fig:correlation}}
 \end{figure}
 
-\revisedOne{%
 However, while in principle feasible, sampling in a Boltzmann distribution directly using the Turner energy model may induce extreme computational demands, with treewidths scaling at least as large as the number of nucleotides in the largest loop.
@@
 \subsection*{Effectively targeting Turner energies using multi-dimensional sampling}
-\revisedOne{
+\revised{
 We used our Boltzmann sampling strategy (Algorithms
@@
 pseudoknots (D\&P model) and energies in the stacking energy
-model\revisedOne{, which is obtained for each of the two target structures (with respective $R^2$ values of $0.846$ and $0.841$)}. For the shown fits we sampled $n=10\,000$ sequences, targeting a
+model\revised{, which is obtained for each of the two target structures (with respective $R^2$ values of $0.846$ and $0.841$)}. For the shown fits we sampled $n=10\,000$ sequences, targeting a
 \GCb\% of $60\%$.
@@
     %{\sf \bfseries A}\
-    \includegraphics[width=0.97\textwidth]{Figs/energy_shift}\hfill
+    \includegraphicstop[width=0.97\textwidth]{Figs/energy_shift}\hfill
     %{\sf \bfseries B}\ \includegraphicstop[width=0.55\textwidth]{Figs/PKB00211_PKB00239_0_energy_distribution}
@@
     energies in the stacking model to the realistic pseudoknot energy
-    model by Dirks and Pierce (D\&P) for \revisedOne{initially} sampled sequences and both target structures R1 and R2 (shown in B). The good
-    match \revisedOne{(respective $R^2$ values of $0.846$ and $0.841$)} enables more efficient targeting of Turner energies based on
-    targeting stacking model energies.
-\textbf{(B)} \revisedOne{Resulting D\&P energy distributions for the two target structures R1 and R2 when aiming for the respective free energies $-30$ and $-20$,
+    model by Dirks and Pierce (D\&P) for \revised{initially} sampled sequences and both target structures R1 and R2 (shown in B). The good
+    match \revised{(respective $R^2$ values of $0.846$ and $0.841$)} enables more efficient targeting of Turner energies based on
+    targeting stacking model energies. \textbf{(B)} \revised{Resulting D\&P energy distributions for the two target structures R1 and R2 when aiming for the respective free energies $-30$ and $-20$,
     $-30$ and $-30$, $-25$ and $-25$, $-20$ and $-30$\,kcal/mol. These demonstrate the effectivity of our adaptive
@@
 As a baseline for comparison, we considered \RNAblueprint~\citep{Hammer2017}, the current leading tool for multiple design.
-As a quality measure, we applied the objective function introduced by~\citet{Hammer2017} \revisedOne{based on \cite{Flamm2001,HoenerzuSiederdissen2013}} for multi-stable design, defined as:
+As a quality measure, we applied the objective function introduced by~\citet{Hammer2017} \revised{based on \cite{Flamm2001,HoenerzuSiederdissen2013}} for multi-stable design, defined as:
 % \begin{align}
@@
 %for RNA design.
-\revisedOne{Based on} \ourprog{}, we generated at least 1\,000 seed sequences with similar energies for all target structures, for
+\revised{Based on} \ourprog{}, we generated at least 1\,000 seed sequences with similar energies for all target structures, for
 each instance of the benchmark. 
@@
 Generally, we targeted 60\% \GCb\%.
-\revisedOne{%
+\revised{%
 Supp.\ Mat.~\ref{appsec:effectivity-seed-generation} provides detailed results from this iterative procedure, which works similarly to the previously described multi-dimensional Boltzmann sampling.
@@
 We compared the
-\Obj{} value of \revisedOne{the derived} sequences against that of seed sequences,
+\Obj{} value of \revised{the derived} sequences against that of seed sequences,
 uniformly sampled using \RNAblueprint{}.  Moreover, for both sets we
@@
 
-\revisedOne{Based on a general framework and efficient algorithms, we introduce a novel approach to design RNA
+\revised{Based on a general framework and efficient algorithms, we introduce a novel approach to design RNA
 sequences while targeting very specific complex properties}. In particular, we describe the targeting of the free energies of multiple target structures and the \GCb-content.
@@
 controlled by expressive RNA energy models.
-\revisedOne{We demonstrate that the approach, despite of its theoretical hardness, performs well on typical multi-stable RNA design instances in practice.
+\revised{We demonstrate that the approach, despite of its theoretical hardness, performs well on typical multi-stable RNA design instances in practice.
 This good performance is a direct consequence of the approximability of Turner energies as well as the systematic algorithmic framework.
@@
 
-\revisedOne{In this way, our framework enables new possibilities
+\revised{In this way, our framework enables new possibilities
 in the field of RNA sequence design.}
@@
 %perspectives:
-\revisedOne{As particular advantage over previous sequence generation methods,} it is extensible to include various more complex sequence constraints, including mandatory/forbidden motifs at specific
+\revised{As particular advantage over previous sequence generation methods,} it is extensible to include various more complex sequence constraints, including mandatory/forbidden motifs at specific
 positions or anywhere in the designed sequences, by adapting formal language constructs
@@
 %
-\revisedOne{We moreover envision using positive design to assess the significance of observed properties.
+\revised{We moreover envision using positive design to assess the significance of observed properties.
 Critically, our current perception of statistical significance suffers from overly simplistic simple null models (\emph{e.g.} dinucleotide shuffling) used to model random RNAs~\cite{Rivas2017}).

### 2019-03-26 RNARedPrint.tex
fix typos

--- 2019-03-26 RNARedPrint.tex
@@
 (FET-Open grant 323987; project RiboNets).
-\revised{The funding bodies did not played any roles in the design
+\revised{The funding bodies did not play any roles in the design
 of the study and collection, analysis, and interpretation of data and in writing
